---
title: "Non-ontology based trait collapsing"
author: "Isobel Beasley"
date: "2025-09-06"
output:
  workflowr::wflow_html:
    toc: true
    toc_float: true
    number_sections: true
editor_options: 
  markdown: 
    wrap: 72
---


# Set up

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r required_fns, message = F, warning= F}

library(dplyr)
library(data.table)
library(stringr)

```


```{r setup_2, message = FALSE, warning= FALSE}

gwas_study_info <- fread(here::here("output/gwas_cat/gwas_study_info_trait_cat.csv"))

```

# Initial summary

## Objectives of this analysis:

- Collapsing and standardizing disease trait terms in GWAS study metadata without relying on ontology mappings.
- Initial harmonization of disease trait terms (fixing spelling, collapsing studies that study different aspects of the same disease)


# Number of studies per trait - before any grouping 

Let's look at the number of studies associated with each unique disease/trait term before any grouping or collapsing.

```{r initial_number_sum}

n_studies_trait = gwas_study_info |>
  dplyr::filter(DISEASE_STUDY == T) |>
  dplyr::select(all_disease_terms, PUBMED_ID) |>
  dplyr::distinct() |>
  dplyr::group_by(all_disease_terms) |>
  dplyr::summarise(n_studies = dplyr::n()) |>
  dplyr::arrange(desc(n_studies))

head(n_studies_trait)

dim(n_studies_trait)

```

# Disease trait collapsing and cleaning

```{r initial_cleaning}
# Basic cleaning of disease terms
# Removes trailing commas, 
gwas_study_info$all_disease_terms = sub(",$", "", gwas_study_info$all_disease_terms)

# trims whitespace
gwas_study_info$all_disease_terms = stringr::str_trim(gwas_study_info$all_disease_terms)

# remove apostrophes
gwas_study_info$all_disease_terms = stringr::str_remove_all(gwas_study_info$all_disease_terms, "'|’")

# initializes a new column (collected_all_disease_terms) to standardized traits
# collapse more traits 
gwas_study_info = gwas_study_info |> 
  mutate(collected_all_disease_terms = all_disease_terms) 

```


## Convert seropositivity to disease terms

```{r convert}

## ---------------------------------------------------------------------------
## Recode "* seropositivity" GWAS trait terms -> infection / disease terms
##
## Design notes:
##  * Matching is on the WHOLE term, not a substring, because many terms are
##    substrings of others ("virus seropositivity", "bacillus seropositivity",
##    "seropositivity measurement", ...). Sequential str_replace_all() calls
##    on these will silently mangle each other.
##  * Cells may hold several comma-separated terms; each is mapped separately
##    and the result de-duplicated.
##  * Terms that are NOT human disease (gut-commensal and anti-phage antibody
##    panels, plant pathogens, non-specific measurements) are given
##    disease_term = NA and a term_class you can filter on.
## ---------------------------------------------------------------------------

sero_map <- tribble(
  ~term,                                                 ~disease_term,                        ~term_class,
  ## ---- viruses -----------------------------------------------------------
  "adenoviridae virus seropositivity",                   "adenovirus infection",               "viral_infection",
  "alphapapillomavirus 9 virus seropositivity",          "human papillomavirus infection",     "viral_infection",
  "bk polyomavirus seropositivity",                      "bk polyomavirus infection",          "viral_infection",
  "calicivirus seropositivity",                          "calicivirus infection",              "viral_infection",
  "chikungunya virus seropositivity",                    "chikungunya",                        "viral_infection",
  "coxsackievirus seropositivity",                       "coxsackievirus infection",           "viral_infection",
  "cytomegalovirus seropositivity",                      "cytomegalovirus infection",          "viral_infection",
  "dengue virus seropositivity",                         "dengue",                             "viral_infection",
  "epstein barr viral capsid antigen seropositivity",    "epstein-barr virus infection",       "viral_infection",
  "epstein barr virus nuclear antigen-1 seropositivity", "epstein-barr virus infection",       "viral_infection",
  "epstein-barr virus early antigen seropositivity",     "epstein-barr virus infection",       "viral_infection",
  "foot-and-mouth disease virus seropositivity",           "foot-and-mouth disease",           "viral_infection",
  "epstein-barr virus seropositivity",                   "epstein-barr virus infection",       "viral_infection",
  "hepatitis b virus (core|surface) antigen seropositivity",  "hepatitis b",                        "viral_infection",
  "anti-hepatitis b virus antibody",                     "hepatitis b",                        "viral_infection",
  "hepatitis a virus seropositivity",                    "hepatitis a",                        "viral_infection",
  "hepatitis c virus seropositivity",                    "hepatitis c",                        "viral_infection",
  "hepatitis e virus seropositivity",                    "hepatitis e",                        "viral_infection",
  "anti-hepatitis e virus antibody",                     "hepatitis e",                        "viral_infection",
  "herpes simplex virus 2 seropositivity",               "herpes simplex virus 2 infection",   "viral_infection",
  "anti-human herpes virus 7 antibody",                  "human herpesvirus 7 infection",      "viral_infection",
  "anti-herpes simplex virus 1 antibody",                "herpes simplex virus 6 infection",   "viral_infection",
  "herpes virus seropositivity",                         "herpesvirus infection",              "viral_infection",
  "hsv1 seropositivity",                                 "herpes simplex virus 1 infection",   "viral_infection",
  "hpv seropositivity",                                  "human papillomavirus infection",     "viral_infection",
  "human herpesvirus 6 seropositivity",                  "human herpesvirus 6 infection",      "viral_infection",
  "human herpesvirus 7 seropositivity",                  "human herpesvirus 7 infection",      "viral_infection",
  "human polyomavirus 6 seropositivity",                 "human polyomavirus 6 infection",     "viral_infection",
  "influenza a seropositivity",                          "influenza a",                        "viral_infection",
  "influenza b seropositivity",                          "influenza b",                        "viral_infection",
  "merkel cell virus seropositivity",                    "merkel cell polyomavirus infection", "viral_infection",
  "measles virus seropositivity",                         "measles",                           "viral_infection",
  "mumps virus seropositivity",                          "mumps",                              "viral_infection",
  "pneumoviridae seropositivity",                        "pneumovirus infection",              "viral_infection",
  "pneumoviridae virus seropositivity",                  "pneumovirus infection",              "viral_infection",
  "poliovirus seropositivity",                           "poliovirus infection",               "viral_infection",
  "polyomavirus 2 seropositivity",                       "jc polyomavirus infection",          "viral_infection",
  "rubella virus seropositivity",                        "rubella",                            "viral_infection",
  "respiratory syncytial virus seropositivity",          "respiratory syncytial virus infection", "viral_infection",
  "sars-cov-2 antibody",                                 "covid-19",                           "viral_infection",
  "sars-cov-2 neutralizing antibody",                    "covid-19",                           "viral_infection",
  "sars coronavirus seropositivity",                     "sars",                               "viral_infection",
  "varicella zoster virus seropositivity",               "varicella zoster virus infection",   "viral_infection",
  "virus seropositivity",                                "viral infectious disease",           "viral_infection",

  ## ---- bacterial pathogens ----------------------------------------------
  "bacteria seropositivity",                             "bacterial infection",                "bacterial_infection",
  "bartonella seropositivity",                           "bartonellosis",                      "bacterial_infection",
  "bordetella seropositivity",                           "pertussis",                          "bacterial_infection",
  "campylobacter seropositivity",                        "campylobacteriosis",                 "bacterial_infection",
  "chlamydia trachomatis seropositivity",                 "chlamydia trachomatis infection",    "bacterial_infection",
  "haemophilus influenzae seropositivity",               "haemophilus influenzae infection",   "bacterial_infection",
  "helicobacter pylori seropositivity",                  "helicobacter pylori infection",      "bacterial_infection",
  "listeria seropositivity",                             "listeriosis",                        "bacterial_infection",
  "legionella seropositivity",                          "legionnaires disease",                 "bacterial_infection",
  "mycobacterium tuberculosis seropositivity",           "tuberculosis",                       "bacterial_infection",
  "mycoplasma pneumoniae seropositivity",                "mycoplasma pneumoniae infection",    "bacterial_infection",
  "neisseria meningitidis seropositivity",               "meningococcal infection",            "bacterial_infection",
  "porphyromonas gingivalis seropositivity",             "porphyromonas gingivalis infection", "bacterial_infection",
  "salmonella seropositivity",                           "salmonellosis",                      "bacterial_infection",
  "shigella seropositivity",                             "shigellosis",                        "bacterial_infection",
  "staphylococcus seropositivity",                       "staphylococcal infection",           "bacterial_infection",
  "streptococcus seropositivity",                        "streptococcal infection",            "bacterial_infection",
  "yersinia seropositivity",                             "yersiniosis",                        "bacterial_infection",

  ## ---- opportunists / commensals that can also be pathogens -------------
  ## Mapped to an infection term but flagged: in microbiome serology panels
  ## these are colonisation markers, not infection. Check the source study.
  "enterobacteriaceae seropositivity",                   "enterobacteriaceae infection",       "ambiguous_bacterial",
  "enterococcus faecalis seropositivity",                "enterococcus faecalis infection",    "ambiguous_bacterial",
  "escherichia seropositivity",                          "escherichia coli infection",         "ambiguous_bacterial",
  "haemophilus parainfluenzae seropositivity",           "haemophilus parainfluenzae infection", "ambiguous_bacterial",
  "pseudomonadaceae seropositivity",                     "pseudomonas infection",              "ambiguous_bacterial",

  ## ---- fungi and parasites ----------------------------------------------
  "aspergillus seropositivity",                          "aspergillosis",                      "fungal_infection",
  "fungus seropositivity",                               "fungal infectious disease",          "fungal_infection",
  "plasmodium falciparum seropositivity",                "malaria",                            "parasitic_infection",
  "protozoan seropositivity",                            "protozoan infectious disease",       "parasitic_infection",
  "toxoplasma gondii seropositivity",                    "toxoplasmosis",                      "parasitic_infection",
  "trypanosoma cruzi seropositivity",                    "chagas disease",                     "parasitic_infection",

  ## ---- autoimmune / allergy ---------------------------------------------
  "anti-citrullinated protein antibody seropositivity",  "rheumatoid arthritis",               "autoimmune",
  "rheumatoid factor seropositivity",                    "autoimmune disease",               "autoimmune",
  "plant allergen seropositivity",                       "plant allergy",                      "allergy",
  "animal allergen seropositivity" ,                       "animal allergy",                     "allergy",

  ## ---- NOT disease: gut commensal antibody panels ------------------------
  "akkermansia seropositivity",                          NA, "microbiota_antibody",
  "alistipes seropositivity",                            NA, "microbiota_antibody",
  "bacillaceae seropositivity",                          NA, "microbiota_antibody",
  "bacilli seropositivity",                              NA, "microbiota_antibody",
  "bacillus seropositivity",                             NA, "microbiota_antibody",
  "bacteroidaceae seropositivity",                       NA, "microbiota_antibody",
  "bacteroidales seropositivity",                        NA, "microbiota_antibody",
  "bacteroides seropositivity",                          NA, "microbiota_antibody",
  "barnesiella seropositivity",                          NA, "microbiota_antibody",
  "bifidobacterium seropositivity",                      NA, "microbiota_antibody",
  "blautia seropositivity",                              NA, "microbiota_antibody",
  "butyrivibrio seropositivity",                         NA, "microbiota_antibody",
  "clostridiales seropositivity",                        NA, "microbiota_antibody",
  "coprococcus seropositivity",                          NA, "microbiota_antibody",
  "dorea seropositivity",                                NA, "microbiota_antibody",
  "eubacterium seropositivity",                          NA, "microbiota_antibody",
  "faecalibacterium seropositivity",                     NA, "microbiota_antibody",
  "firmicutes seropositivity",                           NA, "microbiota_antibody",
  "lachnospiraceae seropositivity",                      NA, "microbiota_antibody",
  "lactobacillales seropositivity",                      NA, "microbiota_antibody",
  "lactobacillus seropositivity",                        NA, "microbiota_antibody",
  "odoribacter seropositivity",                          NA, "microbiota_antibody",
  "parabacteroides seropositivity",                      NA, "microbiota_antibody",
  "prevotella seropositivity",                           NA, "microbiota_antibody",
  "roseburia seropositivity",                            NA, "microbiota_antibody",
  "ruminococcaceae seropositivity",                      NA, "microbiota_antibody",
  "flagellin seropositivity",                            NA, "microbiota_antibody",

  ## ---- NOT disease: anti-phage (gut virome) antibody panels --------------
  "bacillus phage virus seropositivity",                 NA, "phage_antibody",
  "bacteriodes phage virus seropositivity",              NA, "phage_antibody",
  "campylobacter phage virus seropositivity",            NA, "phage_antibody",
  "clostridium phage virus seropositivity",              NA, "phage_antibody",
  "enterobacter phage virus seropositivity",             NA, "phage_antibody",
  "enterococcus phage virus seropositivity",             NA, "phage_antibody",
  "escherichia phage virus seropositivity",              NA, "phage_antibody",
  "geobacillus phage virus seropositivity",              NA, "phage_antibody",
  "klebsiella phage virus seropositivity",               NA, "phage_antibody",
  "lactobacillus phage virus seropositivity",            NA, "phage_antibody",
  "lactococcus phage virus seropositivity",              NA, "phage_antibody",
  "listeria phage virus seropositivity",                 NA, "phage_antibody",
  "pelagibacter phage virus seropositivity",             NA, "phage_antibody",
  "phage seropositivity",                                NA, "phage_antibody",
  "proteus phage virus seropositivity",                  NA, "phage_antibody",
  "pseudomonas phage virus seropositivity",              NA, "phage_antibody",
  "salmonella phage virus seropositivity",               NA, "phage_antibody",
  "shigella phage virus seropositivity",                 NA, "phage_antibody",
  "staphylococcus phage virus seropositivity",           NA, "phage_antibody",
  "streptococcus virus phage virus seropositivity",      NA, "phage_antibody",

  ## ---- NOT human disease: plant pathogens --------------------------------
  "ralstonia solanacearum seropositivity",               NA, "plant_pathogen",
  "xanthomonas seropositivity",                          NA, "plant_pathogen",

  ## ---- NOT disease: non-specific / other serology ------------------------
  "fibronectin seropositivity",                          NA, "other_autoantibody"
)

## Sanity check: no duplicated source terms
stopifnot(!any(duplicated(sero_map$term)))

## ---------------------------------------------------------------------------
## Recoding helper
##
## For each cell: try a whole-cell exact match first (this preserves any
## curated multi-term string), otherwise split on commas, map each part, and
## paste the unique results back together. Unmapped parts pass through
## unchanged so nothing is silently lost.
## ---------------------------------------------------------------------------

lookup_term  <- setNames(sero_map$disease_term, sero_map$term)
lookup_class <- setNames(sero_map$term_class,   sero_map$term)

recode_sero <- function(x, what = c("term", "class")) {
  what <- match.arg(what)
  tbl  <- if (what == "term") lookup_term else lookup_class

  purrr::map_chr(x, function(cell) {
    if (is.na(cell)) return(NA_character_)
    cell <- str_squish(str_to_lower(cell))

    if (cell %in% names(tbl)) return(tbl[[cell]])

    parts <- str_squish(str_split_1(cell, ",\\s*"))
    out <- purrr::map_chr(parts, function(p) {
      if (p %in% names(tbl)) {
        tbl[[p]]
      } else if (what == "term") {
        p                      # not a seropositivity term: leave alone
      } else {
        NA_character_          # no class for non-seropositivity terms
      }
    })
    out <- unique(out[!is.na(out)])
    if (length(out) == 0) NA_character_ else paste(out, collapse = ", ")
  })
}

## ---------------------------------------------------------------------------
## Apply. Keeps the original column so the mapping stays auditable.
## ---------------------------------------------------------------------------

gwas_study_info <- gwas_study_info |>
  mutate(
    collected_all_disease_terms_raw = collected_all_disease_terms,
    sero_term_class = recode_sero(collected_all_disease_terms, "class"),
    collected_all_disease_terms = recode_sero(collected_all_disease_terms, "term")
  )

## Check what happened
gwas_study_info |>
  filter(grepl("seropos", collected_all_disease_terms_raw)) |>
  select(collected_all_disease_terms_raw) |> 
  distinct() |>
  nrow()

gwas_study_info |>
  filter(grepl("seropos", collected_all_disease_terms_raw)) |>
  select(collected_all_disease_terms, collected_all_disease_terms_raw) |> 
  distinct() |>
  head()

## Anything still containing "seropos" after mapping = a term missing from sero_map
gwas_study_info |>
  filter(grepl("seropos", collected_all_disease_terms)) |>
  distinct(collected_all_disease_terms)

```


## Where a study investigates a given aspect of of disease, reduce this to a study of this disease

### Disease susceptibility 

#### Susceptibility to X measurement, decreased susceptibility to)

```{r sus_x_measurement}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms = 
        str_replace_all(collected_all_disease_terms, 
                    "susceptibility to (.*?) measurement(?=,|$)", "\\1")

         ) |>
    mutate(collected_all_disease_terms = 
        str_replace_all(collected_all_disease_terms, 
                    "susceptibility to (.*?) (.*?) measurement(?=,|$)", "\\1 \\2")

         )


gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms = 
        str_remove(collected_all_disease_terms, 
                   pattern = "(?<=^|, )susceptibility to "
         )
  )

```

#### Decreased susceptibility to X

```{r decreased_sus_x}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms = 
        str_remove_all(collected_all_disease_terms,
                       pattern = "(?<=^|, )decreased susceptibility to")
  )

```

#### Predisposition 

```{r predisposition}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms = 
        str_remove_all(collected_all_disease_terms,
                       pattern = " predisposition measurement(?=,|$)")
  )

```

#### Specific examples / changes to make - bruisability

```{r bruisability}

gwas_study_info = gwas_study_info |>
    mutate(collected_all_disease_terms = 
        str_replace_all(collected_all_disease_terms, 
                    "(?<=^|, )bruising susceptibility(?=,|$)",
                    "bruisability")
    )
  

```

### Disease severity 

#### Severity / symptom severity / exacerbation 

```{r disease_severity_initial}

gwas_study_info = gwas_study_info |>
    mutate(collected_all_disease_terms = 
          stringr::str_remove(collected_all_disease_terms, " symptom severity measurement(?=,|$)")
         ) 

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms = 
          stringr::str_remove(collected_all_disease_terms, " severity measurement(?=,|$)")
         ) 

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms = 
          stringr::str_remove(collected_all_disease_terms, " exacerbation measurement(?=,|$)")
         ) 

```

#### Specific example: psoriasis area and severity index

```{r psoriasis_area_severity_index}

gwas_study_info = gwas_study_info |>
    mutate(collected_all_disease_terms = 
          stringr::str_remove(collected_all_disease_terms, ", psoriasis area and severity index(?=,|$)")
         )

```

#### Specific example: uveal melanoma disease severity

```{r uveal_melanoma}

gwas_study_info = gwas_study_info |>
    mutate(collected_all_disease_terms = 
          stringr::str_replace_all(collected_all_disease_terms,
                           "(?<=^|, )uveal melanoma disease severity(?=,|$)",
                           "uveal melanoma"
          )
         )

```

### Disease symptom measurement

```{r symptom_measurement_initial}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms = 
        stringr::str_remove_all(collected_all_disease_terms,
                         " symptom measurement(?=,|$)")
         )

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms = 
        stringr::str_remove_all(collected_all_disease_terms,
                         " symptoms measurement(?=,|$)")
         )


gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms = 
        stringr::str_remove_all(collected_all_disease_terms,
                         " symptom count(?=,|$)")
         )


```

### General measurement

```{r general_measurement}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms = 
        str_replace_all(collected_all_disease_terms, 
                        " measurement(?=,|$)", "")
  )

# gwas_study_info = gwas_study_info |>
#   mutate(collected_all_disease_terms = 
#         str_replace_all(collected_all_disease_terms, 
#                     "(.*?) (.*?) measurement(?=,|$) ", "\\1 \\2")
#   )

```

#### borderline personality disorder symptom -> borderline personality disorder

```{r}

gwas_study_info =
  gwas_study_info |>
  mutate(collected_all_disease_terms = 
           str_replace_all(pattern = "borderline personality disorder symptom",
                           replacement = "borderline personality disorder",
                           string = collected_all_disease_terms))


```

#### suicide behaviour measurement -> suicide 

```{r}

gwas_study_info =
  gwas_study_info |>
  mutate(collected_all_disease_terms = 
           str_replace_all(pattern = "suicide behaviour measurement|suicide ideation measurement",
                           replacement = "suicide",
                           string = collected_all_disease_terms)
         )

```

#### sars-cov-2 antibody measurement -> covid-19 

```{r}

gwas_study_info =
  gwas_study_info |>
  mutate(collected_all_disease_terms = 
           str_replace_all(pattern = "sars-cov-2 antibody measurement",
                           replacement = "covid-19",
                           string = collected_all_disease_terms))


```

#### Dependence / addiction / withdrawal measurement 

```{r dependence_addiction_withdrawal}
# dependence measurement -> dependence
gwas_study_info = gwas_study_info |>
    mutate(collected_all_disease_terms =
           str_replace_all(collected_all_disease_terms,
                           "\\bdependence\\s+measurement(?=,|$)",
                           "dependence")) 

# addiction measurement -> addiction
gwas_study_info = gwas_study_info |> 
  mutate(collected_all_disease_terms =
           str_replace_all(collected_all_disease_terms,
                           "\\baddiction\\s+measurement(?=,|$)",
                           "addiction"))

# withdrawal measurement -> withdrawal
gwas_study_info = gwas_study_info |> 
  mutate(collected_all_disease_terms =
           str_replace_all(collected_all_disease_terms,
                           "\\bwithdrawal\\s+measurement(?=,|$)",
                           "withdrawal"))

```

#### Syndrome / disorder measurement

```{r syndrome_disorder_measurement}

gwas_study_info = gwas_study_info |>
      mutate(collected_all_disease_terms =
           str_replace_all(collected_all_disease_terms,
                           "\\bsyndrome\\s+measurement(?=,|$)",
                           "syndrome"))

gwas_study_info = gwas_study_info |> 
      mutate(collected_all_disease_terms =
           str_replace_all(collected_all_disease_terms,
                           "\\bdisorder\\s+measurement(?=,|$)",
                           "disorder")) 

```

#### Allergy measurement

```{r allergy_measurement}

# allergy measurement -> allergy
gwas_study_info = gwas_study_info |> 
  mutate(collected_all_disease_terms =
           str_replace_all(collected_all_disease_terms,
                           "\\ballergy\\s+measurement(?=,|$)",
                           "allergy")) 

```

#### emphysema pattern measurement -> emphysema

```{r}

gwas_study_info =
  gwas_study_info |>
  mutate(collected_all_disease_terms = 
           str_replace_all(pattern = "emphysema pattern measurement|emphysema imaging measurement",
                           replacement = "emphysema",
                           string = collected_all_disease_terms))


```

#### 	time to remission of covid-19 symptoms -> covid-19

```{r}

gwas_study_info =
  gwas_study_info |>
  mutate(collected_all_disease_terms = 
           str_replace_all(pattern = "time to remission of covid-19 symptoms",
                           replacement = "covid-19",
                           string = collected_all_disease_terms))


```

#### vascular brain injury measurement -> vascular brain injury

```{r}

gwas_study_info =
  gwas_study_info |>
  mutate(collected_all_disease_terms = 
           str_replace_all(pattern = "vascular brain injury measurement",
                           replacement = "vascular brain injury",
                           string = collected_all_disease_terms))


```

### Symptom measurement

```{r symptom_measurement}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms = 
        stringr::str_remove_all(collected_all_disease_terms,
                         " symptom measurement(?=,|$)")
         )

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms = 
        stringr::str_remove_all(collected_all_disease_terms,
                         " symptoms measurement(?=,|$)")
         )


gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms = 
        stringr::str_remove_all(collected_all_disease_terms,
                         " symptom count(?=,|$)")
         )

```

#### Specific example: autism 

```{r autism_symptom}

gwas_study_info = gwas_study_info |> 
    mutate(collected_all_disease_terms = 
           stringr::str_replace_all(collected_all_disease_terms,
                            "(?<=^|, )autism spectrum disorder symptom(?=,|$)",
                            "autism")
         ) 

```

#### Specific example: respiratory symptom change

```{r resp_symptom_change}


gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms = 
         stringr::str_remove_all(collected_all_disease_terms,
                          ", respiratory symptom change measurement(?=,|$)"
         ))

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms = 
         stringr::str_replace_all(collected_all_disease_terms,
                          "respiratory symptom change(?=,|$)",
                          "respiratory symptom" 
         ))

```

## Age of onset of

#### Non-specific correction 

```{r age_onset}

gwas_study_info = gwas_study_info |>
      mutate(collected_all_disease_terms = 
         stringr::str_remove_all(collected_all_disease_terms,
                              pattern = "(?<=^|, )age of onset of ")
         )

```

#### Specific correction: febrile seizure

```{r febrile_seizure}

# febrile seizure (within the age range of 3 months to 6 years)
gwas_study_info = gwas_study_info |> 
 mutate(collected_all_disease_terms  = 
          stringr::str_replace_all(collected_all_disease_terms ,
                                  pattern = "febrile seizure \\(within the age range of 3 months to 6 years\\)",
                                   "febrile seizure"
                          )  
        )

```

## Family history of 

#### Non-specific example

```{r family_history}

gwas_study_info = gwas_study_info |>
      mutate(collected_all_disease_terms = 
         stringr::str_remove_all(collected_all_disease_terms,
                         pattern = "(?<=^|, )family history of "))

```


## Disease progression


### Disease progression measurement

```{r dis_progress_measure}

gwas_study_info = gwas_study_info |>
    mutate(collected_all_disease_terms = 
         stringr::str_remove_all(collected_all_disease_terms,
                          "(?<=^|, )disease progression measurement(?=,|$)"
         ))

gwas_study_info = gwas_study_info |>
    mutate(collected_all_disease_terms = 
         stringr::str_remove_all(collected_all_disease_terms,
                          "disease progression measurement"
         ))

```

## Time to remission

### COVID

```{r covid_remission}

gwas_study_info = gwas_study_info |>
    mutate(collected_all_disease_terms = 
         stringr::str_replace_all(collected_all_disease_terms,
                          "(?<=^|, )time to remission of covid-19 symptoms(?=,|$)",
                          "covid-19"
         ))

```

## Alzheimer's disease - additional examples

### Neuropathological change

```{r alz_dis_neuropath}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms = 
         stringr::str_replace_all(collected_all_disease_terms,
                          "(?<=^|, )alzheimers disease neuropathologic change(?=,|$)",
                          "alzheimers disease"
         ))

```

### Biomarker measurement

```{r alz_dis_biomarker}
gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms = 
         stringr::str_replace_all(collected_all_disease_terms,
                          "(?<=^|, )alzheimers disease biomarker measurement(?=,|$)",
                          "alzheimers disease"
         )) |>
    mutate(collected_all_disease_terms = 
         stringr::str_replace_all(collected_all_disease_terms,
                          "(?<=^|, )alzheimers disease biomarker(?=,|$)",
                          "alzheimers disease"
         ))


gwas_study_info = gwas_study_info |>
    mutate(collected_all_disease_terms = 
         stringr::str_replace_all(collected_all_disease_terms,
                          "(?<=^|, )liver disease biomarker(?=,|$)",
                          "alzheimers disease"
         ))

```



## Cancer aggressiveness measurement

```{r cancer_aggression}

gwas_study_info = 
gwas_study_info |> 
 mutate(collected_all_disease_terms  = 
          stringr::str_remove_all(collected_all_disease_terms,
                                  pattern = "(?<=^|, )cancer aggressiveness measurement(?=,|$)"
                          )
 ) 

```


## Other specific / non-defined measurement

### Anxiety

```{r anxiety_measure}

gwas_study_info = gwas_study_info |> 
    mutate(collected_all_disease_terms = 
         stringr::str_replace_all(collected_all_disease_terms,
                          "(?<=^|, )anxiety measurement(?=,|$)",
                          "anxiety"
         ))

```

### Coronary atherosclerosis

```{r coronary_athero_measure}

gwas_study_info = gwas_study_info |> 
    mutate(collected_all_disease_terms = 
         stringr::str_replace_all(collected_all_disease_terms,
                          "(?<=^|, )coronary atherosclerosis measurement(?=,|$)",
                          "coronary atherosclerosis"
         ))

```

## Cutaneous psoriasis 

```{r cutaneous_psoriasis_measure}

gwas_study_info = gwas_study_info |> 
    mutate(collected_all_disease_terms = 
         stringr::str_replace_all(collected_all_disease_terms,
                          "(?<=^|, )cutaneous psoriasis measurement(?=,|$)",
                          "cutaneous psoriasis"
         ))

```

### Depressive episode

```{r depressive_episode_measure}

gwas_study_info = gwas_study_info |> 
    mutate(collected_all_disease_terms = 
         stringr::str_replace_all(collected_all_disease_terms,
                          "(?<=^|, )depressive episode measurement(?=,|$)",
                          "depressive episode"
         ))

```

### Insomnia 

```{r insomnia_measure}

gwas_study_info = gwas_study_info |> 
    mutate(collected_all_disease_terms = 
         stringr::str_replace_all(collected_all_disease_terms,
                          "(?<=^|, )insomnia measurement(?=,|$)",
                          "insomnia"
         ))

```

### Lewy body dementia

```{r}

gwas_study_info = gwas_study_info |> 
    mutate(collected_all_disease_terms = 
         stringr::str_replace_all(collected_all_disease_terms,
                          "lewy body dementia measurement(?=,|$)",
                          "lewy body dementia"
         ))

```

### Sleep apnea

```{r sleep_apnea_measure}

gwas_study_info = gwas_study_info |> 
    mutate(collected_all_disease_terms = 
         stringr::str_replace_all(collected_all_disease_terms,
                          "(?<=^|, )sleep apnea measurement(?=,|$)",
                          "sleep apnea"
         ))

# during rem sleep
gwas_study_info = gwas_study_info |> 
    mutate(collected_all_disease_terms = 
         stringr::str_replace_all(collected_all_disease_terms,
                          "(?<=^|, )sleep apnea measurement during rem sleep(?=,|$)",
                          "sleep apnea"
         ))

# during non-rem sleep
gwas_study_info = gwas_study_info |> 
    mutate(collected_all_disease_terms = 
         stringr::str_replace_all(collected_all_disease_terms,
                          "(?<=^|, )sleep apnea measurement during non-rem sleep(?=,|$)",
                          "sleep apnea"
         ))

```


### Substance abuse measurement

### Alcohol

```{r alcohol_use_disorder}

gwas_study_info = gwas_study_info |> 
    mutate(collected_all_disease_terms = 
         stringr::str_replace_all(collected_all_disease_terms,
                          "(?<=^|, )alcohol withdrawal delirium(?=,|$)",
                          "alcohol withdrawal"
         ))


alcohol_use_disorder_terms <- c("(?<=^|, )alcohol dependence(?=,|$)",
                             "(?<=^|, )alcohol withdrawal(?=,|$)",
                             "(?<=^|, )alcohol abuse(?=,|$)",
                             "(?<=^|, )alcohol use disorder(?=,|$)",
                             "(?<=^|, )addictive alcohol use(?=,|$)"
                             )

gwas_study_info = gwas_study_info |>
      mutate(collected_all_disease_terms = 
           stringr::str_replace_all(collected_all_disease_terms,
                            paste0(alcohol_use_disorder_terms, collapse = "|"),
                            "alcohol-related disorders"
           ))

```


### Cocaine

```{r cocaine_use_disorder}

gwas_study_info = gwas_study_info |>
    mutate(collected_all_disease_terms = 
         stringr::str_replace_all(collected_all_disease_terms,
                          "cocaine use disorder(?=,|$)|cocaine dependence(?=,|$)",
                          "cocaine-related disorders"
         ))


```


### Nicotine 

```{r nicotine_use_disorder}

nicotine_use_disorder_terms <- c("(?<=^|, )nicotine dependence(?=,|$)",
                             "(?<=^|, )nicotine withdrawal(?=,|$)",
                             "(?<=^|, )nicotine addiction(?=,|$)"
                             )

gwas_study_info = gwas_study_info |>
    mutate(collected_all_disease_terms = 
         stringr::str_replace_all(collected_all_disease_terms,
                          paste0(nicotine_use_disorder_terms, collapse = "|"),
                          "nicotine-related disorders"
         ))


```


## Methamphetamine use disorders 

```{r methamphetamine_use_disorders}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          "(?<=^|, )methamphetamine dependence(?=,|$)|(?<=^|, )methamphetamine-induced psychosis(?=,|$)",
                          "methamphetamine use disorders"
         ))

```

## Deal with basic synonyms

### ADHD = attention deficit hyperactivity disorder
  
```{r adhd_synonym}

gwas_study_info = gwas_study_info |> 
    mutate(collected_all_disease_terms = 
           stringr::str_replace_all(collected_all_disease_terms,
                            "(?<=^|, )attention deficit hyperactivity disorder(?=,|$)",
                            "adhd"
         ))

```

### Anemia (phenotype) -> anemia

```{r anemia_synonym}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms =
         stringr::str_replace_all(collected_all_disease_terms,
                          "(?<=^|, )anemia \\(phenotype\\)(?=,|$)",
                          "anemia"
         ))

```

### Autism = Autism Spectrum Disorder = Asperger Syndrome

```{r autism_synonym}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms = 
        stringr::str_replace_all(collected_all_disease_terms,
                         "(?<=^|, )autism spectrum disorder(?=,|$)",
                         "autism")
        ) |>
    mutate(collected_all_disease_terms = 
        stringr::str_replace_all(collected_all_disease_terms,
                         "(?<=^|, )asperger syndrome(?=,|$)",
                         "autism")
        )

```

### Carcinoma = Cancer

```{r carcinoma_cancer_synonym}

gwas_study_info = gwas_study_info |>
      mutate(collected_all_disease_terms = 
           stringr::str_replace_all(collected_all_disease_terms,
                            "\\bcarcinoma",
                            "cancer")
         )

```

### Coronary artery disease = coronary atherosclerosis 

https://www.ebi.ac.uk/ols4/ontologies/efo/classes/http%253A%252F%252Fwww.ebi.ac.uk%252Fefo%252FEFO_0001645

```{r coronary_artery_disease_synonym}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms = 
         stringr::str_replace_all(collected_all_disease_terms,
                          "(?<=^|, )coronary atherosclerosis(?=,|$)",
                          "coronary artery disease"
         ))

```

### Head and neck cancer = head and neck malignant neoplasia

```{r head_neck_cancer_synonym}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms = 
         stringr::str_replace_all(collected_all_disease_terms,
                          "(?<=^|, )head and neck malignant neoplasia(?=,|$)",
                          "head and neck cancer"
         ))

```

### Hepatitis virus infection = hepatitis infection

```{r hepatitis_virus_infection_synonym}

gwas_study_info = gwas_study_info |>
    mutate(collected_all_disease_terms = 
         stringr::str_replace_all(collected_all_disease_terms,
                          "(?<=^|, )hepatitis a virus infection(?=,|$)",
                          "hepatitis a infection"
         )) |>
      mutate(collected_all_disease_terms = 
         stringr::str_replace_all(collected_all_disease_terms,
                          "(?<=^|, )hepatitis b virus infection(?=,|$)",
                          "hepatitis b infection"
         )) |>
        mutate(collected_all_disease_terms = 
         stringr::str_replace_all(collected_all_disease_terms,
                          "(?<=^|, )hepatitis c virus infection(?=,|$)",
                          "hepatitis c infection"
         ))

gwas_study_info = gwas_study_info |>
    mutate(collected_all_disease_terms = 
         stringr::str_replace_all(collected_all_disease_terms,
                          "(?<=^|, )chronic hepatitis c virus infection(?=,|$)",
                          "chronic hepatitis, hepatitis c infection"
         )) |>
      mutate(collected_all_disease_terms = 
         stringr::str_replace_all(collected_all_disease_terms,
                          "(?<=^|, )chronic hepatitis b virus infection(?=,|$)",
                          "chronic hepatitis, hepatitis b infection"
         ))

gwas_study_info = gwas_study_info |>
    mutate(collected_all_disease_terms = 
         stringr::str_replace_all(collected_all_disease_terms,
                          "(?<=^|, )hepatitis virus-related liver cancer(?=,|$)",
                          "hepatitis, liver cancer"
         ))


```

### Narcolepsy = Narcolepsy without cataplexy

```{r narcolepsy_synonym}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms = 
         stringr::str_replace_all(collected_all_disease_terms,
                          "(?<=^|, )narcolepsy without cataplexy(?=,|$)",
                          "narcolepsy"
         ))

```

### Mumps virus infection = mumps

```{r mumps_virus_infection_synonym}

gwas_study_info = gwas_study_info |>
    mutate(collected_all_disease_terms = 
         stringr::str_replace_all(collected_all_disease_terms,
                          "(?<=^|, )mumps virus infectious disease(?=,|$)",
                          "mumps"
         ))


```

### Renal cancer = kidney cancer

```{r renal_cancer_synonym}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms = 
         stringr::str_replace_all(collected_all_disease_terms,
                          "(?<=^|, )renal cancer(?=,|$)",
                          "kidney cancer"
         ))
  
```  

### Rubella infect = rubella

```{r rubella_infection_synonym}

gwas_study_info = gwas_study_info |>
    mutate(collected_all_disease_terms = 
         stringr::str_replace_all(collected_all_disease_terms,
                          "(?<=^|, )rubella infection(?=,|$)",
                          "rubella"
         ))


```


### Type 1 diabetes = type 1 diabetes mellitus

```{r type1_diabetes_synonym}

gwas_study_info = gwas_study_info |>
    mutate(collected_all_disease_terms = 
        stringr::str_replace_all(collected_all_disease_terms,
                         "(?<=^|, )type 1 diabetes(?=,|$)",
                         "type 1 diabetes mellitus")
        )

```

### Type 1 diabetes mellitus nephropathy = type 1 diabetes nephropathy

```{r type1_diabetes_nephropathy_synonym}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms = 
        stringr::str_replace_all(collected_all_disease_terms,
                         "(?<=^|, )type 1 diabetes mellitus nephropathy(?=,|$)",
                         "type 1 diabetes nephropathy")
        )

```




### Von willebrand disease (hereditary or acquired) = von willebrand disease

```{r vwf_synonym}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms =
         stringr::str_replace_all(collected_all_disease_terms,
                          "(?<=^|, )von willebrand disease \\(hereditary or acquired\\)(?=,|$)",
                          "von willebrand disease"
         ))

```


## Typos 

### Alzheimer -> alzheimers

```{r alzheimers_typo}
gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms =
         stringr::str_replace_all(collected_all_disease_terms,
                          "(?<=^|, )alzheimer disease(?=,|$)",
                          "alzheimers disease"
         ))
```

### Parkinson -> parkinsons

```{r parkinsons_typo}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms =
         stringr::str_replace_all(collected_all_disease_terms,
                          "(?<=^|, )parkinson disease(?=,|$)",
                          "parkinsons disease"
         ))

```

## Other

### attempted suicide -> suicide

```{r}

gwas_study_info =
  gwas_study_info |>
  mutate(collected_all_disease_terms = 
           str_replace_all(pattern = "^attempted suicide$",
                           replacement = "suicide",
                           string = collected_all_disease_terms))


```

### benign -> benign neoplasm 

```{r}

gwas_study_info =
  gwas_study_info |>
  mutate(collected_all_disease_terms = 
           str_replace_all(pattern = "^benign$",
                           replacement = "benign neoplasm",
                           string = collected_all_disease_terms))


```

### cardiovascular disease biomarker measurement -> cardiovascular disease

```{r}

gwas_study_info =
  gwas_study_info |>
  mutate(
    collected_all_disease_terms = str_replace_all(
      pattern = "cardiovascular disease biomarker(?: measurement)?",
      replacement = "cardiovascular disease",
      string = collected_all_disease_terms
    )
  )

```

#### cancer biomarker measurement -> cancer

```{r}

gwas_study_info =
  gwas_study_info |>
  mutate(collected_all_disease_terms = 
           str_replace_all(pattern = "cancer biomarker(?: measurement)?",
                           replacement = "cancer",
                           string = collected_all_disease_terms))

```

### diabetes mellitus biomarker -> diabetes mellitus

```{r}

gwas_study_info =
  gwas_study_info |>
  mutate(collected_all_disease_terms = 
           str_replace_all(pattern = "diabetes mellitus biomarker",
                           replacement = "diabetes mellitus",
                           string = collected_all_disease_terms))


```

### hiv-1 infection -> hiv infection

```{r}

gwas_study_info =
  gwas_study_info |>
  mutate(collected_all_disease_terms = 
           str_replace_all(pattern = "hiv-1 infection",
                           replacement = "hiv infection",
                           string = collected_all_disease_terms))


```

### hiv mother to child transmission -> hiv infection

```{r}

gwas_study_info =
  gwas_study_info |>
  mutate(collected_all_disease_terms = 
           str_replace_all(pattern = "hiv mother to child transmission",
                           replacement = "hiv infection",
                           string = collected_all_disease_terms))


```

#### osteoarthritis biomarker measurement -> osteoarthritis

```{r}

gwas_study_info =
  gwas_study_info |>
  mutate(collected_all_disease_terms = 
           str_replace_all(pattern = "osteoarthritis biomarker measurement",
                           replacement = "osteoarthritis",
                           string = collected_all_disease_terms))


```

### suicide behaviour -> suicide

```{r}

gwas_study_info =
  gwas_study_info |>
  mutate(collected_all_disease_terms = 
           str_replace_all(pattern = "suicide behaviour",
                           replacement = "suicide",
                           string = collected_all_disease_terms))

```


# Putting it all together

```{r, eval = F, include=F}

gwas_study_info$collected_all_disease_terms = stringr::str_trim(gwas_study_info$collected_all_disease_terms)
gwas_study_info$collected_all_disease_terms = sub(",$", "", gwas_study_info$collected_all_disease_terms)
gwas_study_info$collected_all_disease_terms = stringr::str_replace_all(gwas_study_info$collected_all_disease_terms, "^, ", "")

```


## Final summary - number of unique diseases studied

```{r}

diseases <- stringr::str_split(pattern = ", ", 
                               gwas_study_info$collected_all_disease_terms[gwas_study_info$collected_all_disease_terms != ""])  |> 
            unlist() |>
            stringr::str_trim()

length(unique(diseases))

```

## Frequency of unique disease terms after harmonization

```{r freq_table}

# make frequency table
freq <- table(as.factor(diseases))

# sort in decreasing order
freq_sorted <- sort(freq, decreasing = TRUE)

# show top N, e.g. top 10
print("Top 5 most frequent disease terms after harmonization:")
head(freq_sorted, 5)

print("Number of unique disease terms after harmonization:")
sum(freq[freq == 1])

print("Average number of studies per disease term after harmonization:")
mean(freq)

```


## Save the updated gwas_study_info with repaired / harmonized disease terms

```{r save}

data.table::fwrite(
  gwas_study_info,
   here::here("output/gwas_cat/gwas_study_info_disease_trait_simplified.csv")
  )

```

