---
title: "Extract Author provided ICD10 Codes from GWAS Catalog Studies"
author: "Isobel Beasley"
output: html_document
---

# Set up

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load_packages, message=FALSE}

library(dplyr)
library(stringr)
library(data.table)

```

### Load Data

```{r load_data}

gwas_study_info <- data.table::fread(here::here("output/gwas_cat/gwas_study_all_group.csv"))

gwas_study_info = gwas_study_info |>
    mutate(collected_all_disease_terms = stringi::stri_trans_general(collected_all_disease_terms, "Latin-ASCII")
           )

```

# Automatic mapping of GWAS traits to ICD 10

## Create a mapping table

```{r create_mapping_table}

disease_mapping <- gwas_study_info |>
  filter(DISEASE_STUDY == T) |>
  tidyr::separate_longer_delim(cols = collected_all_disease_terms, 
                               delim = ", ") |>
  select(`DISEASE/TRAIT`, 
         collected_all_disease_terms, 
         PUBMED_ID,
         YEAR,
         STUDY_ACCESSION) |>
  distinct()

disease_mapping =
  disease_mapping |>
  filter(collected_all_disease_terms != "")

print("Number of unique disease trait & study pairs")
nrow(disease_mapping)

diseases <- stringr::str_split(pattern = ", ",
                               gwas_study_info$collected_all_disease_terms[gwas_study_info$collected_all_disease_terms != ""])  |>
  unlist() |>
  stringr::str_trim()

diseases <- unique(diseases)

print("Number of unique disease terms")
print(length(diseases))

```

## Get ICD10 codes from author provided DISEASE/TRAIT column 

```{r add_icd10_from_trait}

disease_mapping =
  disease_mapping |>
  mutate(icd10_code = ifelse(grepl("ICD10", `DISEASE/TRAIT`),
                             str_extract(`DISEASE/TRAIT`, 
                                         "(?<=ICD10 )[^:]+(?=:)|(?<=ICD10 )[^ ]+"),
                             NA),
         icd10_code_origin = ifelse(grepl("ICD10", `DISEASE/TRAIT`),
                                    "Study Provided",
                                    NA)
)

# fix weird ICD10 codes: R11.103.11 -> R11
disease_mapping =
  disease_mapping |>
  mutate(icd10_code = stringr::str_replace(icd10_code, 
                                         pattern = "R11.103.11",
                                         replacement = "R11")
         )

# fix weird ICD10 codes: D63.165.8 -> D63, N18.3-N18.9
disease_mapping =
  disease_mapping |>
  mutate(icd10_code = stringr::str_replace(icd10_code, 
                                         pattern = "D63.165.8",
                                         replacement = "D63, N18.3-N18.9")
         ) 

# Additionally, some studies provide ICD10 codes in the DISEASE/TRAIT column without the "ICD10" prefix
# e.g. Source of report of K76 (other diseases of liver) (UKB data field 131671)
# let's try to capture these too
disease_mapping =
  disease_mapping |>
  mutate(icd10_code = ifelse(is.na(icd10_code) &
                             grepl("Source of report of", `DISEASE/TRAIT`),
                             str_extract(`DISEASE/TRAIT`,
                                         "(?<=Source of report of\\s)[A-Z][0-9]{1,2}(?:\\.[0-9A-Z]+)?\\b"),
                             icd10_code),
         icd10_code_origin = ifelse(grepl("Source of report of", 
                                          `DISEASE/TRAIT`),
                                    "Study Provided",
                                    icd10_code_origin)
) |>
  mutate(icd10_code = str_remove_all(icd10_code, 
                                     pattern = "Source of report of ")
         )

# Another example where  some studies provide ICD10 codes in the DISEASE/TRAIT column without the "ICD10" prefix
# e.g. Insulin-dependent diabetes mellitus (Union E10)
disease_mapping =
  disease_mapping |>
  mutate(icd10_code = ifelse(is.na(icd10_code) &
                             grepl("Union", `DISEASE/TRAIT`),
                             str_extract(`DISEASE/TRAIT`, 
                                         "Union [A-Z][0-9]{1,2}(?:\\.[0-9A-Z]+)?\\b"),
                             icd10_code),
         icd10_code_origin = ifelse(grepl("Union", 
                                          `DISEASE/TRAIT`),
                                    "Study Provided",
                                    icd10_code_origin)
) |>
  mutate(icd10_code = str_remove_all(icd10_code, 
                                     pattern = "Union ")
  )

disease_mapping =
  disease_mapping |>
  mutate(icd10_code = ifelse(is.na(icd10_code) &
                             grepl("ICD [A-Z][0-9]|ICD-10 [A-Z][0-9]", 
                                   `DISEASE/TRAIT`),
                        str_match(`DISEASE/TRAIT`,
                                  "ICD(?:-10|10)?\\s*([A-Z][0-9]{1,2}(?:\\.[0-9A-Z]+)?)")[,2],
                             icd10_code),
         icd10_code_origin = ifelse(grepl("ICD [A-Z][0-9]|ICD-10 [A-Z][0-9]",
                                          `DISEASE/TRAIT`),
                                    "Study Provided",
                                    icd10_code_origin)
)

n_study_provided = 
disease_mapping |>
  filter(icd10_code_origin == "Study Provided") |>
  nrow()

n_studies =
  disease_mapping |>
  nrow()

print("Percentage of studies that provide ICD10 codes:")
print(round(n_study_provided / n_studies * 100, 2))

```


## Get ICD10 codes from author provided PheCodes 

### Get Phecodes for diseases

```{r get_phecodes}

disease_mapping <- disease_mapping |>
  mutate(
    phecode = str_extract(`DISEASE/TRAIT`, "(?<=PheCode )[^)]+")
  ) |>
  mutate(phecode = as.numeric(phecode))

```

### Convert Phecodes to ICD10

```{r convert_phecodes_to_icd10, message=FALSE}

# phecode to ICD10 mapping from https://wei-lab.app.vumc.org/phecode-data/phecode_international_version

phecodes <- fread(here::here("data/icd/phecode_international_version_unrolled.csv"))

phecode_icd_map =
  phecodes |>
  select(icd10_code = ICD10, 
         phecode = PheCode
         ) |>
  filter(!is.na(phecode))

phecode_icd_map =
  phecode_icd_map |>
  filter(phecode %in% unique(disease_mapping$phecode))

# if more than one ICD10 code per phecode, collapse into a single row
phecode_icd_map =
  phecode_icd_map |>
group_by(phecode) |>
  summarise(icd10_code = 
            str_flatten(unique(icd10_code), 
                        collapse = ", ", 
                        na.rm = T), 
            .groups = "drop")

# label the origin of these ICD10 codes as "Study PheCode Mapping"
phecode_icd_map =
  phecode_icd_map |>
  mutate(icd10_code_origin = "Study PheCode Mapping")

disease_mapping =
  rows_patch(disease_mapping,
              phecode_icd_map,
              by = c("phecode"),
              unmatched = "ignore",
              )
print("Number of ICD10 code obtained")
disease_mapping |>
    filter(!is.na(icd10_code)) |> 
  group_by(icd10_code_origin) |> 
  summarise(n = n(), 
            percent = n()/n_studies *100)

disease_mapping |>
  filter(!is.na(icd10_code_origin)) |>
  nrow()

disease_mapping |>
  filter(!is.na(icd10_code_origin)) |>
  head()

# per pubmed id, are there multiple icd10 code origins?
disease_mapping |>
  filter(!is.na(icd10_code_origin)) |>
  group_by(PUBMED_ID) |>
  summarise(n_icd10_code_origins = n_distinct(icd10_code_origin)) |>
  pull(n_icd10_code_origins) |>
  summary()

# which pubmed ids have multiple icd10 code origins?
disease_mapping |>
  filter(!is.na(icd10_code_origin)) |>
  group_by(PUBMED_ID) |>
  summarise(n_icd10_code_origins = n_distinct(icd10_code_origin)) |>
  filter(n_icd10_code_origins > 1)

# per study accession, are there multiple icd10 code origins?
disease_mapping |>
  filter(!is.na(icd10_code_origin)) |>
  group_by(STUDY_ACCESSION) |>
  summarise(n_icd10_code_origins = n_distinct(icd10_code_origin)) |>
  pull(n_icd10_code_origins) |>
  summary()

```

### Complete PheCode to ICD10 mappings

Some Phecodes do not have ICD10 mappings in the above Phecode to ICD10 mapping file.
Here we've created a file with manual mappings for these missing Phecodes.

```{r manual_phecodes_to_icd10_mapping}

# how many missing ICD10 codes for PheCodes?
disease_mapping |>
    filter(is.na(icd10_code) & !is.na(phecode)) |> 
    nrow()

# load manual mapping file
phecode_icd_manual = 
  readxl::read_xlsx(here::here("data/icd/phecode_to_icd10_manual_mapping_v5.xlsx"))

to_add = phecode_icd_manual |>
  mutate(phecode = as.numeric(phecode),
         icd10_code_origin = "Study PheCode (Manual Mapping)") |>
  select(-phenotype) |>
  distinct()

disease_mapping =
  rows_patch(disease_mapping,
             to_add,
             unmatched = "ignore")
  
```

# (?) Extract UK Biobank Term Numbers

Possibly to add: 

```{r}

disease_mapping |>
    filter(is.na(icd10_code_origin)) |> filter(grepl("UKB", `DISEASE/TRAIT`)) |> pull(PUBMED_ID) |> unique() |> length()


disease_mapping |>
    filter(is.na(icd10_code_origin)) |> filter(grepl("[0-9+]_[0-9+]", `DISEASE/TRAIT`)) |> pull(PUBMED_ID) |> unique() |> length()

```

# How many disease terms can we map to ICD-10 codes by extracting author provided ICD-10 or PheCodes?

```{r}

n_disease_mapping = nrow(disease_mapping)

disease_mapping  |> 
group_by(icd10_code_origin) |>
  summarise(n = n(),
          percent = round(100 * n/n_disease_mapping, 
                          digits = 2)
          ) |> 
arrange(icd10_code_origin, desc(n))


# how many papers with at least one ICD-10 Code mapping
disease_mapping |>
 filter(!is.na(icd10_code_origin)) |>
  pull(PUBMED_ID) |>
  unique() |>
  length()
  

```

# Save

```{r save_disease_mapping_part}

data.table::fwrite(disease_mapping,
                   here::here("output/gwas_cat/gwas_author_provided_icd10.csv"))

```

