---
title: "Extract Author provided ICD10 Codes from GWAS Catalog Studies"
author: "Isobel Beasley"
output: html_document
---

# Set up

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load_packages, message=FALSE}

library(dplyr)
library(stringr)
library(data.table)

```

### Load Data

```{r load_data}

gwas_study_info <- data.table::fread(here::here("output/gwas_cat/gwas_study_all_group.csv"))

gwas_study_info = gwas_study_info |>
    mutate(collected_all_disease_terms = stringi::stri_trans_general(collected_all_disease_terms, "Latin-ASCII")
           )

```

# Automatic mapping of GWAS traits to ICD 10

## Create a mapping table

```{r create_mapping_table}

disease_mapping <- gwas_study_info |>
  filter(DISEASE_STUDY == T) |>
  tidyr::separate_longer_delim(cols = collected_all_disease_terms, 
                               delim = ", ") |>
  select(`DISEASE/TRAIT`, 
         collected_all_disease_terms, 
         PUBMED_ID,
         YEAR,
         STUDY_ACCESSION) |>
  distinct()

disease_mapping =
  disease_mapping |>
  filter(collected_all_disease_terms != "")

print("Number of unique disease trait & study pairs")
nrow(disease_mapping)

diseases <- stringr::str_split(pattern = ", ",
                               gwas_study_info$collected_all_disease_terms[gwas_study_info$collected_all_disease_terms != ""])  |>
  unlist() |>
  stringr::str_trim()

diseases <- unique(diseases)

print("Number of unique disease terms")
print(length(diseases))

print("Number of disease papers")
print(length(unique(disease_mapping$PUBMED_ID)))

```

## Get ICD10 codes from author provided DISEASE/TRAIT column 

```{r add_icd10_from_trait}

disease_mapping =
  disease_mapping |>
  mutate(icd10_code = ifelse(grepl("ICD10", `DISEASE/TRAIT`),
                             str_extract(`DISEASE/TRAIT`, 
                                         "(?<=ICD10 )[^:]+(?=:)|(?<=ICD10 )[^ ]+"),
                             NA)
         ) |>
    mutate(icd10_code = ifelse(is.na(icd10_code),
                             str_extract(`DISEASE/TRAIT`, 
                                         "(?<=ICD-10\\s)[A-Z]{1}[0-9]{2}"),
                             icd10_code)
         ) |>
      mutate(icd10_code = ifelse(is.na(icd10_code),
                             str_extract(`DISEASE/TRAIT`, 
                                         "(?<=ICD\\s)[A-Z]{1}[0-9]{2}"),
                             icd10_code)
         ) |>
  mutate(icd10_code = ifelse(icd10_code == "no", 
                             NA,
                             icd10_code)
         )

# Additionally, some studies provide ICD10 codes in the DISEASE/TRAIT column without the "ICD10" prefix
# e.g. Source of report of K76 (other diseases of liver) (UKB data field 131671)
# let's try to capture these too
disease_mapping =
  disease_mapping |>
  mutate(icd10_code = ifelse(is.na(icd10_code) &
                             grepl("Source of report of", `DISEASE/TRAIT`),
                             str_extract(`DISEASE/TRAIT`,
                                         "(?<=Source of report of\\s)[A-Z][0-9]{1,2}(?:\\.[0-9A-Z]+)?\\b"),
                             icd10_code)
  )

# Additionally, there are examples where  some studies provide ICD10 codes in the DISEASE/TRAIT column without the "ICD10" prefix
# e.g. Insulin-dependent diabetes mellitus (Union E10)
disease_mapping =
  disease_mapping |>
  mutate(icd10_code = ifelse(is.na(icd10_code) &
                             grepl("Union", `DISEASE/TRAIT`),
                             str_extract(`DISEASE/TRAIT`, 
                                         "(?<=Union\\s)[A-Z][0-9]{1,2}(?:\\.[0-9A-Z]+)?\\b"),
                             icd10_code)
  )

# note the source of the ICD-10 codes
disease_mapping <-
  disease_mapping |>
  mutate(icd10_source_col = ifelse(!is.na(icd10_code),
                             "DISEASE/TRAIT",
                             NA),
         icd10_source = ifelse(!is.na(icd10_code),
                                    "Study ICD10",
                                    NA)
         )

```


```{r clean_up_icd10_codes}

# cleaning up weird icd-10 codes: 
disease_mapping <-
  disease_mapping |>
  mutate(icd10_code = icd10_code |>
           # adding commas where only spaces between codes are provided
                      str_replace_all(pattern = "(?<!,) ", 
                                     replacement = ",") |>
           # putting an extra space after commas if not already present
                      str_replace_all(pattern = "\\,(?! )", 
                                     replacement = ", ") |>
           # removing extra decimals at the end of codes
                       str_remove_all(pattern = "\\.$") |>
           # removing decimals where they should be commas
                       str_replace_all(pattern = "\\.,", 
                                   replacement = ",")
         )
  

disease_mapping |>
  filter(icd10_code == "D63.165.8") |>
  distinct(`DISEASE/TRAIT`, icd10_code)

# D63.165.8 
# should be D63,8, N18.3-N18.9
disease_mapping  <-
  disease_mapping  |>
  mutate(icd10_code = ifelse(icd10_code ==  "D63.165.8", 
                              "D63.8, N18.3, N18.4, N18.5, N18.9",
                              icd10_code)
         )

# 127.0
# Should be I27.0
disease_mapping |>
  filter(icd10_code == "127.0") |>
  distinct(`DISEASE/TRAIT`, icd10_code)


disease_mapping  <-
  disease_mapping  |>
  mutate(icd10_code = ifelse(icd10_code == "127.0",
                             "I27.0",
                             icd10_code))

# K05.30-K05.31
# should be K05.3
disease_mapping |>
  filter(icd10_code == "K05.30-31") |>
  distinct(`DISEASE/TRAIT`, icd10_code)

disease_mapping  <-
  disease_mapping  |>
  mutate(icd10_code = ifelse(icd10_code == "K05.30-31",
                             "K05.3",
                             icd10_code))

# in some cases the decimal is missing, e.g. "T780" should be "T78.0"
disease_mapping |>
    filter(!is.na(icd10_code)) |>
    filter(str_detect(icd10_code, "^[A-Z]{1}[0-9]{3,5}$")) |>
    distinct(`DISEASE/TRAIT`, icd10_code)

disease_mapping <- 
disease_mapping |>
  mutate(
    icd10_code = ifelse(
      str_detect(icd10_code, "^[A-Z][0-9]{3}$"),
      str_replace(
        icd10_code,
        pattern = "^([A-Z][0-9]{2})([0-9])$",
        replacement = "\\1.\\2"
      ),
      icd10_code
    )
  )

disease_mapping <- 
disease_mapping |>
  mutate(
    icd10_code = ifelse(
      str_detect(icd10_code, "^[A-Z][0-9]{4}$"),
      str_replace(
        icd10_code,
        pattern = "^([A-Z][0-9]{2})([0-9]{2})$",
        replacement = "\\1.\\2"
      ),
      icd10_code
    )
  )

disease_mapping <- 
disease_mapping |>
  mutate(
    icd10_code = ifelse(
      str_detect(icd10_code, "^[A-Z][0-9]{5}$"),
      str_replace(
        icd10_code,
        pattern = "^([A-Z][0-9]{2})([0-9]{3})$",
        replacement = "\\1.\\2"
      ),
      icd10_code
    )
  )


# remove surplus period in ICD10 codes, e.g. J98.457.6 should be J98.457
disease_mapping |>
  filter(str_detect(icd10_code, "^[A-Z][0-9]{2}\\.[0-9]{3}\\.[0-9]{1,3}$")) |>
  distinct(`DISEASE/TRAIT`, icd10_code)

disease_mapping <- 
disease_mapping |>
  mutate(
    icd10_code = ifelse(
      str_detect(icd10_code, "^[A-Z][0-9]{2}\\.[0-9]{3}\\.[0-9]{1,3}$"),
      str_extract(
        icd10_code,
        pattern = "^[A-Z][0-9]{2}\\.[0-9]{3}",
      ),
      icd10_code
    )
  )
  

icd10_re <- "^[A-Z][0-9][0-9A-Z](\\.[0-9A-Z]{1,4})?$"

# check that all ICD10 codes are now valid
disease_mapping |>
    filter(!is.na(icd10_code)) |>
    filter(!str_detect(icd10_code, icd10_re))  |> 
    tidyr::separate_longer_delim(icd10_code,delim = ", ")  |>
    filter(!str_detect(icd10_code, icd10_re))  |> 
    distinct(`DISEASE/TRAIT`, icd10_code) 

n_study_provided = 
disease_mapping |>
  filter(icd10_source == "Study ICD10") |>
  distinct(STUDY_ACCESSION) |>
  nrow()

n_studies =
  disease_mapping |>
  distinct(STUDY_ACCESSION) |>
  nrow()

print("Percentage of studies that provide ICD10 codes:")
print(round(n_study_provided / n_studies * 100, 2))


n_paper_provided = 
disease_mapping |>
  filter(icd10_source == "Study ICD10") |>
  distinct(PUBMED_ID) |>
  nrow()

n_papers =
  disease_mapping |>
  distinct(PUBMED_ID) |>
  nrow()

print("Percentage of papers that provide ICD10 codes:")
print(round(n_paper_provided / n_papers * 100, 2))

```


## Get ICD10 codes from author provided PheCodes 

### Get Phecodes for diseases

```{r get_phecodes}

disease_mapping <- disease_mapping |>
  mutate(
    phecode = str_extract(`DISEASE/TRAIT`, "(?<=PheCode )[^)]+")
  ) |>
  mutate(phecode = as.numeric(phecode))

```

### Convert Phecodes to ICD10

```{r convert_phecodes_to_icd10, message=FALSE}

# phecode to ICD10 mapping from https://wei-lab.app.vumc.org/phecode-data/phecode_international_version
phecodes <- data.table::fread(
  here::here("data/icd/phecode_international_version_unrolled.csv"),
  encoding = "UTF-8"          # correct — the file really is UTF-8
)

# Repair malformed bytes, fold diacritics, and convert to ASCII.  This is important because the PheCode mapping file contains some non-ASCII characters (e.g., "é" in "Meniere disease") that are not present in the GWAS Catalog disease terms.  If we don't do this, we won't be able to match these terms.
ascii_key <- function(x) {
  x |>
    stringi::stri_enc_toutf8(validate = TRUE) |>   # malformed bytes -> U+FFFD
    stringi::stri_trans_nfc() |>                   # é as e+U+0301 folds like é as U+00E9
    stringi::stri_trans_general("Latin-ASCII") |>  # é->e  ö->o  ç->c  ß->ss
    stringi::stri_replace_all_regex("[\u2010-\u2015]", "-") |>  # en/em dashes
    stringi::stri_replace_all_regex("[^\\p{ASCII}]", "") |>     # U+FFFD, β, × ...
    stringi::stri_trim_both() 
}

# H81.0 cannot be recovered from its bytes — repair it explicitly.
phecodes <- phecodes |>
  mutate(ICD_DESCRIPTION = if_else(ICD10 == "H81.0", "Meniere disease", ICD_DESCRIPTION)) |>
  mutate(
         ICD_DESCRIPTION        = ascii_key(ICD_DESCRIPTION),
         Phenotype   = ascii_key(Phenotype))

# drop badly mapped conversions
# Nonrheumatic heart disorders being mapped to rheumatic heart disorders (I08.1, I08.2) in the PheCode to ICD10 mapping file
# and Type 1 diabetes mellitus complications being mapped to Type 2 diabetes mellitus (E10.4, E10.3) in the PheCode to ICD10 mapping file
bad_phecode_icd_map = 
  data.frame(
    ICD10 = c("I08.1", 
              "I08.2",
              "E10.4",
              "E10.3"),
    ICD_DESCRIPTION = c(
      "Disorders of both mitral and tricuspid valves",
      "Disorders of both aortic and tricuspid valves",
      "Type 1 diabetes mellitus, With neurological complications",
      "Type 1 diabetes mellitus, With ophthalmic complications"),
    PheCode = c(395.1, 
                395.2,
                250.24,
                250.23),
    Phenotype = c(
      "Nonrheumatic mitral valve disorders",
      "Nonrheumatic aortic valve disorders",
      "Type 2 diabetes with neurological manifestations",
      "Type 2 diabetes with ophthalmic manifestations"
    )
  )

phecodes <-
dplyr::rows_delete(phecodes,
                   bad_phecode_icd_map,
                   by = names(bad_phecode_icd_map))

# also remove Phecode 260.2 mappings 
# severe protein-calorie malnutrition mappings are weird/wrong
phecodes  <-
  phecodes  |>
  filter(PheCode != "260.2")

phecode_icd_map =
  phecodes |>
  select(icd10_code = ICD10, 
         phecode = PheCode
         ) |>
  filter(!is.na(phecode))

phecode_icd_map =
  phecode_icd_map |>
  filter(phecode %in% unique(disease_mapping$phecode))

# if more than one ICD10 code per phecode, collapse into a single row
phecode_icd_map =
  phecode_icd_map |>
group_by(phecode) |>
  summarise(icd10_code = 
            str_flatten(unique(icd10_code), 
                        collapse = ", ", 
                        na.rm = T), 
            .groups = "drop")



# label the origin of these ICD10 codes as "Study PheCode Mapping"
phecode_icd_map =
  phecode_icd_map |>
  mutate(icd10_source = "Study PheCode (PheCode Map)",
         icd10_source_col = "DISEASE/TRAIT")


disease_mapping =
  rows_patch(disease_mapping,
              phecode_icd_map,
              by = c("phecode"),
              unmatched = "ignore",
              )


print("Number of ICD10 code obtained")
disease_mapping |>
    filter(!is.na(icd10_code)) |> 
  group_by(icd10_source) |> 
  summarise(n = length(unique(STUDY_ACCESSION)), 
            percent = length(unique(STUDY_ACCESSION))/n_studies *100)

disease_mapping |>
  filter(!is.na(icd10_source)) |>
  nrow()

disease_mapping |>
  filter(!is.na(icd10_source)) |>
  head()

# per pubmed id, are there multiple icd10 code origins?
disease_mapping |>
  filter(!is.na(icd10_source)) |>
  distinct(PUBMED_ID, icd10_source) |>
  group_by(PUBMED_ID) |>
  summarise(n_icd10_sources = n()) |>
  filter(n_icd10_sources > 1)

# which pubmed ids have multiple icd10 code origins?
disease_mapping |>
  filter(!is.na(icd10_source)) |>
  group_by(PUBMED_ID) |>
  summarise(n_icd10_sources = n_distinct(icd10_source)) |>
  filter(n_icd10_sources > 1)

# per study accession, are there multiple icd10 code origins?
disease_mapping |>
  filter(!is.na(icd10_source)) |>
  group_by(STUDY_ACCESSION) |>
  summarise(n_icd10_sources = n_distinct(icd10_source)) |>
  pull(n_icd10_sources) |>
  summary()

```

### Complete PheCode to ICD10 mappings

Some Phecodes do not have ICD10 mappings in the above Phecode to ICD10 mapping file.
Here we've created a file with manual mappings for these missing Phecodes.

```{r manual_phecodes_to_icd10_mapping}

# how many missing ICD10 codes for PheCodes?
disease_mapping |>
    filter(is.na(icd10_code) & !is.na(phecode)) |> 
    pull(phecode) |> 
    unique() |> 
    length()

missing_icd_phecodes <- disease_mapping |>
    filter(is.na(icd10_code) & !is.na(phecode)) |> 
    pull(phecode) |> 
    unique() 

# load manual mapping file
phecode_icd_manual = 
  readxl::read_xlsx(here::here("data/icd/phecode_to_icd10_manual_map.xlsx")) |>
  select(-icd10_desc)

to_add = phecode_icd_manual |>
  mutate(phecode = as.numeric(phecode),
         icd10_source = "Study PheCode (Manual Map)",
         icd10_source_col = "DISEASE/TRAIT") |>
  select(-phenotype) |>
  distinct()

disease_mapping =
  rows_patch(disease_mapping,
             to_add,
             unmatched = "ignore",
             by = "phecode")

disease_mapping |>
    filter(is.na(icd10_code) & !is.na(phecode)) |> 
    pull(phecode) |>
    unique() |>
   length()
  
```

# (?) Extract UK Biobank Term Numbers

Possibly to add: 

```{r}

disease_mapping |>
    filter(is.na(icd10_source_col)) |> filter(grepl("UKB", `DISEASE/TRAIT`)) |> pull(PUBMED_ID) |> unique() |> length()


disease_mapping |>
    filter(is.na(icd10_source_col)) |> filter(grepl("[0-9+]_[0-9+]", `DISEASE/TRAIT`)) |> pull(PUBMED_ID) |> unique() |> length()

```

# How many disease terms can we map to ICD-10 codes by extracting author provided ICD-10 or PheCodes?

```{r}

disease_mapping  |> 
group_by(icd10_source) |>
  summarise(n = length(unique(STUDY_ACCESSION)),
          percent = round(100 * length(unique(STUDY_ACCESSION)) / n_studies, 
                          digits = 2)
          ) |> 
arrange(icd10_source, desc(n))


# how many papers with at least one ICD-10 Code mapping
disease_mapping |>
 filter(!is.na(icd10_source)) |>
  pull(PUBMED_ID) |>
  unique() |>
  length()

# check that all ICD10 codes are now valid
disease_mapping |>
    filter(!is.na(icd10_code)) |>
    filter(!str_detect(icd10_code, icd10_re))  |> 
    tidyr::separate_longer_delim(icd10_code,delim = ", ")  |>
    filter(!str_detect(icd10_code, icd10_re))  |> 
    distinct(`DISEASE/TRAIT`, icd10_code) 
  

```

# Save

```{r save_disease_mapping_part}

data.table::fwrite(disease_mapping,
                   here::here("output/gwas_cat/gwas_author_provided_icd10.csv"))

```

