---
title: "Extract Author provided ICD10 Codes from GWAS Catalog Studies"
author: "Isobel Beasley"
output: html_document
---

# Set up

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load_packages, message=FALSE}

library(dplyr)
library(stringr)
library(data.table)

```

### Load Data

```{r load_data}

gwas_study_info <- data.table::fread(here::here("output/gwas_cat/gwas_study_all_group.csv"))

gwas_study_info = gwas_study_info |>
    mutate(collected_all_disease_terms = stringi::stri_trans_general(collected_all_disease_terms, "Latin-ASCII")
           )

```

# Automatic mapping of GWAS traits to ICD 10

## Create a mapping table

```{r create_mapping_table}

disease_mapping <- gwas_study_info |>
  filter(DISEASE_STUDY == T) |>
  tidyr::separate_longer_delim(cols = collected_all_disease_terms, 
                               delim = ", ") |>
  select(`DISEASE/TRAIT`, 
         collected_all_disease_terms, 
         PUBMED_ID,
         YEAR,
         STUDY_ACCESSION) |>
  distinct()

disease_mapping =
  disease_mapping |>
  filter(collected_all_disease_terms != "")

print("Number of unique disease trait & study pairs")
nrow(disease_mapping)

diseases <- stringr::str_split(pattern = ", ",
                               gwas_study_info$collected_all_disease_terms[gwas_study_info$collected_all_disease_terms != ""])  |>
  unlist() |>
  stringr::str_trim()

diseases <- unique(diseases)

print("Number of unique disease terms")
print(length(diseases))

print("Number of disease papers")
print(length(unique(disease_mapping$PUBMED_ID)))

```

## Get ICD10 codes from author provided DISEASE/TRAIT column 

```{r add_icd10_from_trait}

disease_mapping =
  disease_mapping |>
  mutate(icd10_code = ifelse(grepl("ICD10", `DISEASE/TRAIT`),
                             str_extract(`DISEASE/TRAIT`, 
                                         "(?<=ICD10 )[^:]+(?=:)|(?<=ICD10 )[^ ]+"),
                             NA)
         ) |>
    mutate(icd10_code = ifelse(is.na(icd10_code),
                             str_extract(`DISEASE/TRAIT`, 
                                         "(?<=ICD-10\\s)[A-Z]{1}[0-9]{2}"),
                             icd10_code)
         ) |>
      mutate(icd10_code = ifelse(is.na(icd10_code),
                             str_extract(`DISEASE/TRAIT`, 
                                         "(?<=ICD\\s)[A-Z]{1}[0-9]{2}"),
                             icd10_code)
         ) |>
  mutate(icd10_code = ifelse(icd10_code == "no", 
                             NA,
                             icd10_code)
         )

# Additionally, some studies provide ICD10 codes in the DISEASE/TRAIT column without the "ICD10" prefix
# e.g. Source of report of K76 (other diseases of liver) (UKB data field 131671)
# let's try to capture these too
disease_mapping =
  disease_mapping |>
  mutate(icd10_code = ifelse(is.na(icd10_code) &
                             grepl("Source of report of", `DISEASE/TRAIT`),
                             str_extract(`DISEASE/TRAIT`,
                                         "(?<=Source of report of\\s)[A-Z][0-9]{1,2}(?:\\.[0-9A-Z]+)?\\b"),
                             icd10_code)
  )

# Additionally, there are examples where  some studies provide ICD10 codes in the DISEASE/TRAIT column without the "ICD10" prefix
# e.g. Insulin-dependent diabetes mellitus (Union E10)
disease_mapping =
  disease_mapping |>
  mutate(icd10_code = ifelse(is.na(icd10_code) &
                             grepl("Union", `DISEASE/TRAIT`),
                             str_extract(`DISEASE/TRAIT`, 
                                         "(?<=Union\\s)[A-Z][0-9]{1,2}(?:\\.[0-9A-Z]+)?\\b"),
                             icd10_code)
  )

# note the source of the ICD-10 codes
disease_mapping <-
  disease_mapping |>
  mutate(icd10_code_match = ifelse(!is.na(icd10_code),
                             "DISEASE/TRAIT",
                             NA),
         icd10_code_origin = ifelse(!is.na(icd10_code),
                                    "Study ICD10",
                                    NA)
         )

```


```{r clean_up_icd10_codes}

# cleaning up weird icd-10 codes: 
disease_mapping <-
  disease_mapping |>
  mutate(icd10_code = icd10_code |>
           # adding commas where only spaces between codes are provided
                      str_replace_all(pattern = "(?<!,) ", 
                                     replacement = ",") |>
           # putting an extra space after commas if not already present
                      str_replace_all(pattern = "\\,(?! )", 
                                     replacement = ", ") |>
           # removing extra decimals at the end of codes
                       str_remove_all(pattern = "\\.$") |>
           # removing decimals where they should be commas
                       str_replace_all(pattern = "\\.,", 
                                   replacement = ",")
         )
  

disease_mapping |>
  filter(icd10_code == "D63.165.8") |>
  distinct(`DISEASE/TRAIT`, icd10_code)

# D63.165.8 
# should be D63,8, N18.3-N18.9
disease_mapping  <-
  disease_mapping  |>
  mutate(icd10_code = ifelse(icd10_code ==  "D63.165.8", 
                              "D63.8, N18.3, N18.4, N18.5, N18.9",
                              icd10_code)
         )

# 127.0
# Should be I27.0
disease_mapping |>
  filter(icd10_code == "D63.165.8") |>
  distinct(`DISEASE/TRAIT`, icd10_code)


disease_mapping  <-
  disease_mapping  |>
  mutate(icd10_code = ifelse(icd10_code == "127.0",
                             "I27.0",
                             icd10_code))

# K05.30-K05.31
# should be K05.3
disease_mapping |>
  filter(icd10_code == "K05.30-31") |>
  distinct(`DISEASE/TRAIT`, icd10_code)

disease_mapping  <-
  disease_mapping  |>
  mutate(icd10_code = ifelse(icd10_code == "K05.30-31",
                             "K05.3",
                             icd10_code))

# in some cases the decimal is missing, e.g. "T780" should be "T78.0"
disease_mapping |>
    filter(!is.na(icd10_code)) |>
    filter(str_detect(icd10_code, "^[A-Z]{1}[0-9]{3,5}$")) |>
    distinct(`DISEASE/TRAIT`, icd10_code)

disease_mapping <- 
disease_mapping |>
  mutate(
    icd10_code = ifelse(
      str_detect(icd10_code, "^[A-Z][0-9]{3}$"),
      str_replace(
        icd10_code,
        pattern = "^([A-Z][0-9]{2})([0-9])$",
        replacement = "\\1.\\2"
      ),
      icd10_code
    )
  )

disease_mapping <- 
disease_mapping |>
  mutate(
    icd10_code = ifelse(
      str_detect(icd10_code, "^[A-Z][0-9]{4}$"),
      str_replace(
        icd10_code,
        pattern = "^([A-Z][0-9]{2})([0-9]{2})$",
        replacement = "\\1.\\2"
      ),
      icd10_code
    )
  )

disease_mapping <- 
disease_mapping |>
  mutate(
    icd10_code = ifelse(
      str_detect(icd10_code, "^[A-Z][0-9]{5}$"),
      str_replace(
        icd10_code,
        pattern = "^([A-Z][0-9]{2})([0-9]{3})$",
        replacement = "\\1.\\2"
      ),
      icd10_code
    )
  )


# remove surplus period in ICD10 codes, e.g. J98.457.6 should be J98.457
disease_mapping |>
  filter(str_detect(icd10_code, "^[A-Z][0-9]{2}\\.[0-9]{3}\\.[0-9]{1,3}$")) |>
  distinct(`DISEASE/TRAIT`, icd10_code)

disease_mapping <- 
disease_mapping |>
  mutate(
    icd10_code = ifelse(
      str_detect(icd10_code, "^[A-Z][0-9]{2}\\.[0-9]{3}\\.[0-9]{1,3}$"),
      str_extract(
        icd10_code,
        pattern = "^[A-Z][0-9]{2}\\.[0-9]{3}",
      ),
      icd10_code
    )
  )
  

icd10_re <- "^[A-Z][0-9][0-9A-Z](\\.[0-9A-Z]{1,4})?$"

# check that all ICD10 codes are now valid
disease_mapping |>
    filter(!is.na(icd10_code)) |>
    filter(!str_detect(icd10_code, icd10_re))  |> 
    tidyr::separate_longer_delim(icd10_code,delim = ", ")  |>
    filter(!str_detect(icd10_code, icd10_re))  |> 
    distinct(`DISEASE/TRAIT`, icd10_code) 

n_study_provided = 
disease_mapping |>
  filter(icd10_code_origin == "Study ICD10") |>
  distinct(STUDY_ACCESSION) |>
  nrow()

n_studies =
  disease_mapping |>
  distinct(STUDY_ACCESSION) |>
  nrow()

print("Percentage of studies that provide ICD10 codes:")
print(round(n_study_provided / n_studies * 100, 2))


n_paper_provided = 
disease_mapping |>
  filter(icd10_code_origin == "Study ICD10") |>
  distinct(PUBMED_ID) |>
  nrow()

n_papers =
  disease_mapping |>
  distinct(PUBMED_ID) |>
  nrow()

print("Percentage of papers that provide ICD10 codes:")
print(round(n_paper_provided / n_papers * 100, 2))

```


## Get ICD10 codes from author provided PheCodes 

### Get Phecodes for diseases

```{r get_phecodes}

disease_mapping <- disease_mapping |>
  mutate(
    phecode = str_extract(`DISEASE/TRAIT`, "(?<=PheCode )[^)]+")
  ) |>
  mutate(phecode = as.numeric(phecode))

```

### Convert Phecodes to ICD10

```{r convert_phecodes_to_icd10, message=FALSE}

# phecode to ICD10 mapping from https://wei-lab.app.vumc.org/phecode-data/phecode_international_version

phecodes <- fread(here::here("data/icd/phecode_international_version_unrolled.csv"))

phecode_icd_map =
  phecodes |>
  select(icd10_code = ICD10, 
         phecode = PheCode
         ) |>
  filter(!is.na(phecode))

phecode_icd_map =
  phecode_icd_map |>
  filter(phecode %in% unique(disease_mapping$phecode))

# if more than one ICD10 code per phecode, collapse into a single row
phecode_icd_map =
  phecode_icd_map |>
group_by(phecode) |>
  summarise(icd10_code = 
            str_flatten(unique(icd10_code), 
                        collapse = ", ", 
                        na.rm = T), 
            .groups = "drop")

# label the origin of these ICD10 codes as "Study PheCode Mapping"
phecode_icd_map =
  phecode_icd_map |>
  mutate(icd10_code_origin = "Study PheCode (PheCode Map)",
         icd10_code_match = "DISEASE/TRAIT")

disease_mapping =
  rows_patch(disease_mapping,
              phecode_icd_map,
              by = c("phecode"),
              unmatched = "ignore",
              )


print("Number of ICD10 code obtained")
disease_mapping |>
    filter(!is.na(icd10_code)) |> 
  group_by(icd10_code_origin) |> 
  summarise(n = n(), 
            percent = n()/n_studies *100)

disease_mapping |>
  filter(!is.na(icd10_code_origin)) |>
  nrow()

disease_mapping |>
  filter(!is.na(icd10_code_origin)) |>
  head()

# per pubmed id, are there multiple icd10 code origins?
disease_mapping |>
  filter(!is.na(icd10_code_origin)) |>
  distinct(PUBMED_ID, icd10_code_origin) |>
  group_by(PUBMED_ID) |>
  summarise(n_icd10_code_origins = n()) |>
  filter(n_icd10_code_origins > 1)

# which pubmed ids have multiple icd10 code origins?
disease_mapping |>
  filter(!is.na(icd10_code_origin)) |>
  group_by(PUBMED_ID) |>
  summarise(n_icd10_code_origins = n_distinct(icd10_code_origin)) |>
  filter(n_icd10_code_origins > 1)

# per study accession, are there multiple icd10 code origins?
disease_mapping |>
  filter(!is.na(icd10_code_origin)) |>
  group_by(STUDY_ACCESSION) |>
  summarise(n_icd10_code_origins = n_distinct(icd10_code_origin)) |>
  pull(n_icd10_code_origins) |>
  summary()

```

### Complete PheCode to ICD10 mappings

Some Phecodes do not have ICD10 mappings in the above Phecode to ICD10 mapping file.
Here we've created a file with manual mappings for these missing Phecodes.

```{r manual_phecodes_to_icd10_mapping}

# how many missing ICD10 codes for PheCodes?
disease_mapping |>
    filter(is.na(icd10_code) & !is.na(phecode)) |> 
    pull(phecode) |> 
    unique() |> 
    length()

missing_icd_phecodes <- disease_mapping |>
    filter(is.na(icd10_code) & !is.na(phecode)) |> 
    pull(phecode) |> 
    unique() 

# load manual mapping file
phecode_icd_manual = 
  readxl::read_xlsx(here::here("data/icd/phecode_to_icd10_manual_mapping_v5.xlsx"))

to_add = phecode_icd_manual |>
  mutate(phecode = as.numeric(phecode),
         icd10_code_origin = "Study PheCode (Manual Map)",
         icd10_code_match = "DISEASE/TRAIT") |>
  select(-phenotype) |>
  distinct()

disease_mapping =
  rows_patch(disease_mapping,
             to_add,
             unmatched = "ignore",
             by = "phecode")

disease_mapping |>
    filter(is.na(icd10_code) & !is.na(phecode)) |> 
    pull(phecode) |>
    unique() |>
   length()
  
```

# (?) Extract UK Biobank Term Numbers

Possibly to add: 

```{r}

disease_mapping |>
    filter(is.na(icd10_code_origin)) |> filter(grepl("UKB", `DISEASE/TRAIT`)) |> pull(PUBMED_ID) |> unique() |> length()


disease_mapping |>
    filter(is.na(icd10_code_origin)) |> filter(grepl("[0-9+]_[0-9+]", `DISEASE/TRAIT`)) |> pull(PUBMED_ID) |> unique() |> length()

```

# How many disease terms can we map to ICD-10 codes by extracting author provided ICD-10 or PheCodes?

```{r}

n_disease_mapping = nrow(disease_mapping)

disease_mapping  |> 
group_by(icd10_code_origin) |>
  summarise(n = n(),
          percent = round(100 * n/n_disease_mapping, 
                          digits = 2)
          ) |> 
arrange(icd10_code_origin, desc(n))


# how many papers with at least one ICD-10 Code mapping
disease_mapping |>
 filter(!is.na(icd10_code_origin)) |>
  pull(PUBMED_ID) |>
  unique() |>
  length()

# check that all ICD10 codes are now valid
disease_mapping |>
    filter(!is.na(icd10_code)) |>
    filter(!str_detect(icd10_code, icd10_re))  |> 
    tidyr::separate_longer_delim(icd10_code,delim = ", ")  |>
    filter(!str_detect(icd10_code, icd10_re))  |> 
    distinct(`DISEASE/TRAIT`, icd10_code) 
  

```

# Save

```{r save_disease_mapping_part}

data.table::fwrite(disease_mapping,
                   here::here("output/gwas_cat/gwas_author_provided_icd10.csv"))

```

