---
title: "Mapping GWAS traits to ICD 10"
author: "Isobel Beasley"
output: html_document
---

# Set up

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load_packages, message=FALSE}

library(dplyr)
library(stringr)
library(data.table)

```

### Load Data

```{r load_data}

# load GWAS study info with harmonised disease trait labels
# and author provided ICD-10 Codes 
disease_mapping <- data.table::fread(here::here("output/gwas_cat/gwas_author_provided_icd10.csv"))

disease_mapping <-
  disease_mapping |>
  mutate(icd10_code = ifelse(icd10_code == "",
                             NA,
                             icd10_code)) |>
  mutate(icd10_code_origin = ifelse(icd10_code_origin == "",
                                    NA,
                                    icd10_code_origin))

gwas_study_info <- data.table::fread(here::here("output/gwas_cat/gwas_study_all_group.csv"))

gwas_study_info = gwas_study_info |>
    mutate(collected_all_disease_terms = stringi::stri_trans_general(collected_all_disease_terms, "Latin-ASCII")
           )

```

# Tidy `DISEASE/TRAIT` Column 

Do this to enable term matching with PheCode, ICD-10 and ULS descriptions. 

### Removing model related terms 

```{r remove_model_terms, message=FALSE}

disease_mapping <- disease_mapping |>
  mutate(
    trait_clean = `DISEASE/TRAIT` |>
      str_replace_all("[\u2018\u2019]", "'") |>
      str_remove_all('["\u201C\u201D]')) 


# --- remove identifiers (PheCodes, UK Biobank data fields) ---
disease_mapping <- disease_mapping |>
  mutate(
    trait_clean = trait_clean |>
      str_remove_all("\\s*\\(UKB data field[^)]*\\)") |>
      str_remove_all("\\s*\\(\\d+_\\d+\\)\\s*$") |> 
      str_remove_all("\\s*\\(Phe[Cc]ode \\d+(?:\\.\\d+)?\\)") |>
      str_remove_all("\\s*RELEASE\\s*\\d+\\s*$") |>
      str_remove_all("\\s*\\((?:agilent|twist)_\\d+\\)") 
)


# --- remove model related terms ---
meta_paren <- paste0(
  "(?i)\\s*\\((?:",
  paste(
    c(
      "age (of|at) onset",
      "age of onset \\< 21",
      "\\d+\\s*df(?:\\s+test)?",
      "(?:SPA|Firth)\\s+correction",
      "gene[ -]burden",
      "gene-based burden",
      "does not include regular naps",
      "time to event",
      "survival",
      "\\d+-day mortality",
      "model\\s*\\d+",
      "ICD-10 coded",
      "change over time",
      "apnea hypopnea index, change over time",
      "joint analysis[^)]*",
      "trans-disease meta-analysis",
      "multivariate analysis",
      "multi-trait analysis",
      "MTAG",
      "pleiotropy",
      "CNV (?:mirror|U-shape) model",
      "case-control set\\s*\\d+",
      "adjusted for[^)]*",
      "[^)]*\\badjusted\\b[^)]*",
      "percentage of [^)]*",
      "[^)]*\bper event\b[^)]*",
      "[^)]*\bsleep episode\b[^)]*",
      "[^)]*\bChecklist[^)]*",
      "concentration drug ratio",
      "severest",
      "time to remission",
      "population-based",
      "baseline",
      "pairwise",
      "\\d+\\s*(?:month|week|year)\\s*visit",
      "self[ -]report(?:ed)?",
      "RELEASE\\s*\\d+",
      "ordinary least squares \\(OLS\\)",
      "(?:SPACox|ADuLT) model[^)]*",
      "(?:standard|weighted)\\s+GWA",
      "(?:binary|ordinal) trait",
      "[Cc]onditioned on rs\\d+",
      "conditioned\\s+on\\s+rs\\s*\\d+",
      "AlphaMissense",
      "ESRD vs. no ESRD",
      "treatment strategy interaction"
      
    ),
    collapse = "|"
  ),
  ")\\)"
)

disease_mapping <- disease_mapping |>
  mutate(
    trait_clean = trait_clean |>
      str_remove_all(meta_paren) |> 
      str_remove_all("(?i)\\s*\\([^)]*\\binteraction\\b[^)]*\\)") |>
      str_remove_all("(?i)\\s*\\([^)]*\\bvs\\.?\\b[^)]*\\)")  |> 
      str_remove_all(" adjusted for[^)]*$")  |>
      str_remove_all("(?i)(,|:) PHESANT recoding")
    )
      
# ---- remove clinical information ---- # 
disease_mapping <- disease_mapping |>
  mutate(
    trait_clean = trait_clean |>
    str_remove_all("\\s*\\([Hh]ospital (?:diagnosed|admission)\\)") |>
    str_remove_all("\\([^)]* [Cc]lassification\\)") |>
    str_remove_all("\\s*\\(\\w+(?: \\w+)? scale\\)") |>
    str_remove_all("\\s*\\(\\w+(?: \\w+)? score\\)") |>
    str_remove_all("\\s*\\(\\w+(?: \\w+)? index\\)") |>
      str_remove_all("\\s*\\(\\w+(?: \\w+)? index\\)") |>
      str_remove_all("\\s*\\(\\w+(?: \\w+)? factor\\)") |>
    str_remove_all("\\s*\\(clinical subgroup [^)]*\\)") |>
    str_remove("(?i)^(?:cancer|non-cancer illness|treatment/medication|operation) code,\\s*self-reported:\\s*") |>
    str_remove_all("Takes medication for ") |>
    str_remove_all("Viral variant \\w+ carrier status in\\s*") |>
    str_remove_all("3-month functional outcome in\\s*") |>
    str_remove_all("\\s*\\((XELOX|FOLFOX)(?:,\\s*\\w+)?\\)") |>
    str_remove_all("\\s*\\(\\d+[- ]weeks?\\)")
    
  )

# ---- remove mutation information --- # 
disease_mapping <- disease_mapping |>
  mutate(
    trait_clean = trait_clean |>
      str_remove_all("\\([^)]* mutations?\\)") |>
      str_remove_all("\\([^)]* mutation only\\)") |>
      str_remove_all("\\s*\\(\\w+ deletions?\\)") |>
      str_remove_all("\\s*\\(rs\\d+\\)") |>
      str_remove_all("\\s*\\(rs\\d+_\\d+\\)")
  )

# ---- remove expression information --- #
disease_mapping <- disease_mapping |>
  mutate(
    trait_clean = trait_clean |>
      str_remove_all("\\w+ expression at (baseline|\\d+ weeks) in (whole blood|PBMC)")  |>
      str_remove_all("upon \\w+ stimulation in\\s+") |>
      str_remove_all("^Baseline \\w+ levels in response to \\w+ \\w+ \\w+ \\w+ test in\\s+") 
  )


# --- remove family genotype effects etc.  ---- # 
disease_mapping <- disease_mapping |>
  mutate(
    trait_clean = trait_clean |>
    str_remove_all(paste0(
        "\\s*\\((?:maternal|paternal|inherited|fetal|child|recipient|donor|",
        "parent[- ]of[- ]origin|maternal genetic|",
        "(?:maternal|paternal|offspring|fetal) genotype|",
        "maternal and offspring genotype)? ?effects?\\)")) |>
      str_remove_all("(?i)(family|paternal) history of") |>
      str_remove_all("Biological Grandparent \\((?:maternal|paternal)\\):\\s*") |>
      str_remove_all("Biological Grandparent \\(maternal\\) \\(maternal\\):") |>
      str_remove_all("Biological (?:Father|Mother|Sibling):\\s*") |>
      str_remove_all("Illnesses of (?:father|mother|siblings)[: -]\\s*") |> 
      str_remove_all("\\((maternal|paternal)\\)") |>
      str_remove_all("\\s+families$") |> 
      str_remove_all("Maternal genotype effects in")
  )

# --- final cleaning ---- # 
disease_mapping <- disease_mapping |>
  mutate(
    trait_clean = trait_clean |> 
      str_remove_all("\\?") |>
      str_replace_all("\\s*/\\s*", "/") |>
      str_replace_all("\\s{2,}", " ") |> 
      str_squish() |>
      str_remove("^[\\s,;:.-]+") |>
      str_remove("[\\s,;:.-]+$")
  )

disease_mapping <- disease_mapping |>
  mutate(
trait_context_clean = str_match(
      trait_clean,
      "(?i)\\s(?:levels?|counts?|ratios?|activity|abundance|concentrations?)\\sin\\s(.+)$"
    )[, 2] |>
      str_remove("(?i)\\s+(?:with hypertension and no diabetes|patients|individuals)$") |> 
      str_remove_all("individuals with\\s+") 
  )

```

# Get ICD10 codes by matching `DISEASE/TRAIT` terms 

### Match disease/trait to ICD 10 descriptions 

#### Cleaned DISEASE/TRAIT

```{r match_disease_terms_to_icd_descriptions, message=FALSE}

# phecode to ICD10 mapping from https://wei-lab.app.vumc.org/phecode-data/phecode_international_version

phecodes <- fread(here::here("data/icd/phecode_international_version_unrolled.csv"))
n_studies <- nrow(disease_mapping)

matched_icd10_desc  = 
phecodes |>
    mutate(collected_all_disease_terms = tolower(iconv(ICD_DESCRIPTION, 
                                                       to = "UTF-8"))
           ) |>
    select(collected_all_disease_terms,    
           icd10_code = ICD10) 

# filter missing strings
matched_icd10_desc = 
  matched_icd10_desc |>
  filter(!is.na(collected_all_disease_terms) & 
         collected_all_disease_terms != "") |>
  filter(!is.na(icd10_code) & 
         icd10_code != "")

matched_icd10_desc =
  matched_icd10_desc |>
  group_by(collected_all_disease_terms) |>
  summarise(icd10_code = str_flatten(unique(icd10_code), 
                                     collapse = ", ", 
                                     na.rm = T), 
            .groups = "drop") 
  

matched_icd10_desc = 
  matched_icd10_desc |>
    mutate(icd10_code_origin = "ICD Description Match (DISEASE/TRAIT)")

# match by cleaned DISEASE/TRAIT
matched_icd10_desc =
  matched_icd10_desc |>
  rename(trait_clean = collected_all_disease_terms)

disease_mapping = 
disease_mapping |>
rows_patch(matched_icd10_desc,
           unmatched = "ignore")

print("Number of ICD10 codes obtained")
disease_mapping |>
    filter(!is.na(icd10_code)) |> 
  group_by(icd10_code_origin) |> 
  summarise(n = n(), 
            percent = n()/n_studies *100)

```

#### Extracted DISEASE from `DISEASE/TRAIT` column

```{r match_disease_context_to_icd10_descriptions, message=FALSE}

matched_icd10_desc  = 
phecodes |>
    mutate(collected_all_disease_terms = tolower(iconv(ICD_DESCRIPTION, 
                                                       to = "UTF-8"))
           ) |>
    select(collected_all_disease_terms,    
           icd10_code = ICD10) 

# filter missing strings
matched_icd10_desc = 
  matched_icd10_desc |>
  filter(!is.na(collected_all_disease_terms) & 
         collected_all_disease_terms != "") |>
  filter(!is.na(icd10_code) & 
         icd10_code != "")

matched_icd10_desc =
  matched_icd10_desc |>
  group_by(collected_all_disease_terms) |>
  summarise(icd10_code = str_flatten(unique(icd10_code), 
                                     collapse = ", ", 
                                     na.rm = T), 
            .groups = "drop") 
  
matched_icd10_desc = 
  matched_icd10_desc |>
    mutate(icd10_code_origin = "ICD Description Match (trait context)")

# match by cleaned context from DISEASE/TRAIT
matched_icd10_desc =
  matched_icd10_desc |>
  rename(trait_context_clean = collected_all_disease_terms)

disease_mapping = 
disease_mapping |>
rows_patch(matched_icd10_desc,
           unmatched = "ignore")

print("Number of ICD10 codes obtained")
disease_mapping |>
    filter(!is.na(icd10_code)) |> 
  group_by(icd10_code_origin) |> 
  summarise(n = n(), 
            percent = n()/n_studies *100)

print("Number of papers where at least one ICD-10 has been obtained so far:")
disease_mapping |>
  filter(!is.na(icd10_code)) |>
  pull(PUBMED_ID) |>  
  unique() |>
  length()

```

### Match disease/trait to phenotypes (corresponding to PheCodes)

```{r match_disease_terms_to_phenotypes, message=FALSE}

phenotype_icd_map =
  phecodes |>
group_by(Phenotype) |>
  summarise(icd10_code = 
            str_flatten(ICD10, 
                        collapse = ", ", 
                        na.rm = T), 
            .groups = "drop")

matched_phenotypes =
phenotype_icd_map 

matched_phenotypes =
matched_phenotypes |>
  mutate(collected_all_disease_terms = tolower(Phenotype)) |>
  select(collected_all_disease_terms, icd10_code) |>
  mutate(icd10_code_origin = "Phecode Phenotype Match (trait context)")

# match by DISEASE/TRAIT
matched_phenotypes =
  matched_phenotypes |>
  rename(trait_context_clean = collected_all_disease_terms)

disease_mapping =
disease_mapping |>
rows_patch(matched_phenotypes,
           unmatched = "ignore")

disease_mapping |>
    filter(!is.na(icd10_code)) |> 
  group_by(icd10_code_origin) |> 
  summarise(n = n(), 
            percent = n()/n_studies *100)

print("Number of papers where at least one ICD-10 has been obtained so far:")
disease_mapping |>
    filter(!is.na(icd10_code)) |>
  pull(PUBMED_ID) |>  
  unique() |>
  length()
         
```

### Match disease/trait to UMLS 

```{r match_disease_terms_to_umls, message=FALSE}

unmapped_disease_terms <-
  disease_mapping |>
  filter(is.na(icd10_code)) |>
  pull(trait_clean) |>
  unique() |>
  tolower()

# get UMLS CUIs with ICD10 mappings
umls_data <-
  data.table::fread(here::here("data/icd/2025AA/META/MRCONSO.RRF"),
                    sep = "|",
                    header = FALSE,
                    quote = "",
                    fill = TRUE,
                    na.strings = c("", "NA")
  )

colnames(umls_data)[1:18] <- c(
  "CUI","LAT","TS","LUI","STT","SUI","ISPREF",
  "AUI","SAUI","SCUI","SDUI","SAB","TTY","CODE",
  "STR","SRL","SUPPRESS","CVF"
)

umls_cuis_icd10 <-
  umls_data |>
  filter(SAB %in% c("ICD10", "ICD10CM")) |>
  select(CODE, CUI) |>
  group_by(CUI) |>
  summarise(CODE = str_flatten(unique(CODE),
                               collapse = ", ",
                               na.rm = T),
            .groups = "drop"
            ) |>
  rename(icd10_code = CODE) 


umls_icd10 <-
  umls_data |>
  filter(CUI %in% umls_cuis_icd10$CUI)

umls_icd10 =
  umls_icd10 |>
  left_join(umls_cuis_icd10,
            by = "CUI")

# overlap with umls terms
disease_trait_umls  <-
  umls_icd10 |>
  filter(tolower(STR) %in% unmapped_disease_terms) |>
  mutate(trait_clean = tolower(STR)) |>
  select(trait_clean,
         icd10_code) |>
  group_by(trait_clean) |>
  summarise(icd10_code = str_flatten(unique(icd10_code),
                               collapse = ", ",
                               na.rm = T),
            .groups = "drop"
            ) |>
  distinct() |>
  mutate(icd10_code_origin = "UMLS term match (DISEASE/TRAIT)")


 disease_mapping <-
  disease_mapping |>
  mutate(trait_clean = tolower(trait_clean)) |>
  rows_patch(disease_trait_umls,
             by = 'trait_clean')
 
# remove pubmed id: 27197191
# in this one, disease/trait = "cancer", but specific cancers tested
# are listed in collected_all_disease_terms - so will only
# match collected_all_disease_terms later (not disease/trait terms)

 disease_mapping  = 
  disease_mapping  |>
  mutate(
    icd10_code  = ifelse(PUBMED_ID == 27197191,
           NA,
           icd10_code
           )
         ) |>
  mutate(
   icd10_code_origin = ifelse(PUBMED_ID == 27197191,
          NA,
          icd10_code_origin
          ) 
  )

```

#### Match trait context to UMLS

```{r match_trait_context_to_umls, message=FALSE}

unmapped_disease_terms <- 
    disease_mapping |>
    filter(is.na(icd10_code)) |> 
    pull(trait_context_clean) |>
    unique() |> 
    tolower()

# overlap with umls terms
disease_trait_umls  <-
  umls_icd10 |>
  filter(tolower(STR) %in% unmapped_disease_terms) |>
  mutate(trait_context_clean = tolower(STR)) |>
  select(trait_context_clean,
         icd10_code) |>
  group_by(trait_context_clean) |>
  summarise(icd10_code = str_flatten(unique(icd10_code),
                               collapse = ", ",
                               na.rm = T),
            .groups = "drop"
            ) |>
  distinct() |>
  mutate(icd10_code_origin = "UMLS term match (trait context)")

 disease_mapping <-
  disease_mapping |>
  mutate(trait_context_clean = tolower(trait_context_clean)) |>
  rows_patch(disease_trait_umls,
             by = 'trait_context_clean')
 
 disease_mapping |>
    filter(!is.na(icd10_code)) |> 
  group_by(icd10_code_origin) |> 
  summarise(n = n(), 
            percent = n()/n_studies *100)
 
 print("Number of papers where at least one ICD-10 has been obtained so far:")
disease_mapping |>
    filter(!is.na(icd10_code)) |>
  pull(PUBMED_ID) |>  
  unique() |>
  length()

```

# Manual mapping of GWAS traits to ICD 10

```{r manual_map_icd10_disease_trait}

manual_icd10_map <-
  readxl::read_xlsx(here::here("data/icd/manual_disease_mappings_v2.xlsx"))

manual_icd10_map_trait_clean =
  manual_icd10_map |>
  select(trait_clean = mapped_trait, 
         icd10_code) |>
  mutate(trait_clean = stringr::str_squish(tolower(trait_clean))) |>
  mutate(icd10_code_origin = "Manual Mapping (DISEASE/TRAIT)") |>
  distinct()
disease_mapping =
  rows_patch(disease_mapping,
             manual_icd10_map_trait_clean,
             unmatched = "ignore")

```

### behaviour of cancer tumour: uncertain whether benign or malignant -> D37-D48

```{r manual_icd10_mapping_behaviour_gwas_cleaned_traits, eval = F}

disease_mapping =
  disease_mapping |>
  mutate(
    icd10_code = ifelse(
        str_detect(trait_clean, 
                   "behaviour of cancer tumour: uncertain whether benign or malignant"),
      paste0("D", 37:48, collapse = ", "),
      icd10_code
    )
    ) |>
    mutate(
    icd10_code_origin = ifelse(
        str_detect(trait_clean, 
                   "behaviour of cancer tumour: uncertain whether benign or malignant"),
      "Manual Mapping (DISEASE/TRAIT)",
      icd10_code_origin
    )
  )

 print("Number of papers where at least one ICD-10 has been obtained so far:")
disease_mapping |>
    filter(!is.na(icd10_code)) |>
  pull(PUBMED_ID) |>  
  unique() |>
  length()

```

# Get ICD10 codes by matching `collected_all_disease_terms` terms

### Match collected_all_disease_terms to ICD10 descriptions

```{r match_collected_icd10_desc}

matched_icd10_desc = 
  matched_icd10_desc |>
  rename(collected_all_disease_terms = trait_context_clean)

matched_icd10_desc = 
  matched_icd10_desc |>
  mutate(icd10_code_origin = "ICD Description Match (MAPPED DISEASE TRAIT)")

disease_mapping =
  rows_patch(disease_mapping,
             matched_icd10_desc,
             unmatched = "ignore")

 print("Number of papers where at least one ICD-10 has been obtained so far:")
disease_mapping |>
    filter(!is.na(icd10_code)) |>
  pull(PUBMED_ID) |>  
  unique() |>
  length()

```

### Match collected_all_disease_terms to ICD10 codes (by phenotypes)

```{r match_collected_icd10}

matched_phenotypes = 
  matched_phenotypes |>
  rename(collected_all_disease_terms = trait_context_clean)

matched_phenotypes = 
  matched_phenotypes |>
  mutate(icd10_code_origin = "Phecode Phenotype Match (MAPPED DISEASE TRAIT)")

# match by collected_all_disease_terms
disease_mapping = 
disease_mapping |>
rows_patch(matched_phenotypes,
           unmatched = "ignore")

print("Number of papers where at least one ICD-10 has been obtained so far:")
disease_mapping |>
    filter(!is.na(icd10_code)) |>
  pull(PUBMED_ID) |>  
  unique() |>
  length()


```

### UMLS match collected_all_disease_terms

```{r match_collected_umls, message=FALSE}

unmapped_terms <-
  disease_mapping |>
  filter(is.na(icd10_code)) |>
  pull(collected_all_disease_terms) |>
  unique()

# overlap with umls terms
collected_trait_umls  <-
  umls_icd10 |>
  filter(tolower(STR) %in% unmapped_terms) |>
  mutate(collected_all_disease_terms = tolower(STR)) |>
  # mutate(CODE = ifelse(SAB %in% c("ICD10", "ICD10CM"),
  #                      CODE,
  #                      NA)) |>
  select(collected_all_disease_terms,
         icd10_code) |>
  group_by(collected_all_disease_terms) |>
  summarise(icd10_code = str_flatten(unique(icd10_code),
                               collapse = ", ",
                               na.rm = T),
            .groups = "drop"
            ) |>
  distinct() |>
  mutate(icd10_code_origin = "UMLS term match (collected_all_disease_terms)")

disease_mapping <-
  disease_mapping |>
  rows_patch(collected_trait_umls,
             by = "collected_all_disease_terms")


print("Number of papers where at least one ICD-10 has been obtained so far:")
disease_mapping |>
    filter(!is.na(icd10_code)) |>
  pull(PUBMED_ID) |>  
  unique() |>
  length()


```

# Manual mapping of GWAS traits to ICD 10

### hepatitis B, where related to vaccine respone

Map to chronic hepatitis B, and acute hepatitis B

```{r, eval = F}

# collected_all_disease_terms = "hepatitis b"
# & disease/trait contains "hepatitis B vaccine"
disease_mapping =
  disease_mapping |>
  mutate(
    icd10_code = ifelse(
      collected_all_disease_terms == "hepatitis b" &
        str_detect(tolower(`DISEASE/TRAIT`), "hepatitis b vaccine"),
      "B16.0, B16.1, B16.2, B16.9, B18.0, B18.1, B18.2, B18.8, B18.9",
      icd10_code
    )
  ) |>
  mutate(
    icd10_code_origin = ifelse(
      collected_all_disease_terms == "hepatitis b" &
        str_detect(tolower(`DISEASE/TRAIT`), "hepatitis b vaccine"),
      "Manual Mapping (collected_all_disease_terms + DISEASE/TRAIT)",
      icd10_code_origin
    )
  )

```

### More manual mappings from Excel sheet

```{r create_manual_mapping_table}

manual_icd10_map_collected_trait =
  manual_icd10_map |>
  select(trait_clean = mapped_trait, 
         icd10_code) |>
  mutate(collected_all_disease_terms = stringr::str_squish(tolower(trait_clean))) |>
  mutate(icd10_code_origin = "Manual Mapping (MAPPED DISEASE TERM)") |>
  distinct()

disease_mapping =
  rows_patch(disease_mapping,
             manual_icd10_map_collected_trait,
             unmatched = "ignore")

disease_mapping =
  disease_mapping |>
  mutate(icd10_code = stringr::str_remove_all(string = icd10_code,
                                              pattern = ", $"))

```

# Inferring ICD10 codes from similar study ICD10 codes

```{r}

# similar studies:
study_icd_map =
disease_mapping |>
filter(!is.na(icd10_code)) |>
filter(icd10_code_origin == "Study Provided" | icd10_code_origin == "Study PheCode Mapping") 

study_icd_map =
  study_icd_map |>
  select(collected_all_disease_terms, icd10_code) |>
  distinct()

# remove Z95.1 and Z95.5 from coronary artery disease
study_icd_map =
  study_icd_map |>
  mutate(icd10_code = ifelse(collected_all_disease_terms == "coronary artery disease",
                             str_replace_all(icd10_code, "Z95.1|Z95.5", ""),
                             icd10_code))

# assume ovarian cancer terms more specific than 
# Malignant neoplasm of ovary and other uterine adnexa (PheCode 184.1)
# study_icd_map =
#   study_icd_map |>
#   filter(!( collected_all_disease_terms == "ovarian cancer" &&phecode == "184.10")
#   )

study_icd_map = 
  study_icd_map |>
  mutate(collected_all_disease_terms = str_trim(collected_all_disease_terms)) |>
  mutate(collected_all_disease_terms = str_remove_all(collected_all_disease_terms, "^, ")) 
  

study_icd_map =
  study_icd_map |>
  filter(icd10_code != "" & !is.na(icd10_code))


study_icd_map =
  study_icd_map |>
  tidyr::separate_longer_delim(icd10_code, delim = ", ") |>
  tidyr::separate_longer_delim(icd10_code, delim = ",")

study_icd_map =
  study_icd_map |>
  group_by(collected_all_disease_terms) |>
  summarise(icd10_code = str_flatten(unique(sort(icd10_code)),
                                     collapse = ", ", 
                                     na.rm = T), 
            .groups = "drop")

study_icd_map = 
  study_icd_map |>
  mutate(icd10_code_origin = "Inferred from similar studies")

disease_mapping = 
  rows_patch(disease_mapping,
             study_icd_map,
             unmatched = "ignore")

disease_mapping |>
  filter(is.na(icd10_code)) |>
  nrow()

```

# How was ICD-10 code inferred?

```{r how_was_icd10_inferred}

# as a percentage
n_disease_mapping = nrow(disease_mapping)

group_levels <- c(
  "UMLS Match",
  "Study Provided Info",
  "Phecode Match",
  "ICD Description Match",
  "Manual Mapping",
  "Other"
)

disease_mapping  |> 
mutate(icd10_code_origin_group = case_when(
  grepl("Study", icd10_code_origin) ~ "Study Provided Info",
  grepl("Phecode Phenotype Match", icd10_code_origin) ~ "Phecode Match",
  grepl("ICD Description Match", icd10_code_origin) ~ "ICD Description Match",
  grepl("UMLS", icd10_code_origin) ~ "UMLS Match",
  grepl("Manual Mapping", icd10_code_origin) ~ "Manual Mapping",
  TRUE ~ "Other"
)) |>
  mutate(icd10_code_origin_group = factor(icd10_code_origin_group,
                                        levels = group_levels
                                        )
       ) |>
group_by(icd10_code_origin_group, icd10_code_origin) |> 
summarise(n = n(),
          percent = round(100 * n/n_disease_mapping, 
                          digits = 2)
          ) |> 
arrange(icd10_code_origin_group, desc(n))

```



# Saving disease mapping 

## Add description of ICD10 codes to mapping table

### Prepare / fix disease mappings 

#### Fixing where single number is mapped as a range

```{r}

disease_mapping =
  disease_mapping |>
  mutate(icd10_code = sub("^([^-]+)-\\1$", "\\1", icd10_code))

disease_mapping = 
  disease_mapping |> 
  mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "C50-C50.9",
    replacement = "C50"
    )
  )

disease_mapping = 
  disease_mapping |> 
  mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "F30-F39.9",
    replacement = "F30"
    )
  )

disease_mapping = 
  disease_mapping |> 
  mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "F99-F99.9",
    replacement = "F99"
    )
  )

```

#### Fixing multiple ICD10 codes missing commas

```{r prepare_disease_mappings}

disease_mapping = 
  disease_mapping |> 
  mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "F40 F41 F42",
    replacement = "F40, F41, F42"
    )
  )

disease_mapping = 
  disease_mapping |> 
  mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "G47.8. G47.9",
    replacement = "G47.8, G47.9"
    )
  )

```

#### Correcting for letters after numbers in ICD-10 codes

```{r}
# ICD-10 cm conversion to ICD-10 WHO
# from here: https://seer.cancer.gov/tools/conversion/2026/ICD10CM-to-ICD10.FY2026.pdf
disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "D3A.8",
    replacement = "D36.7"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "F32.A",
    replacement = "F32"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "G20.A1",
    replacement = "G20"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "G20.C",
    replacement = "G20"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "G40.A",
    replacement = "G40"
    )
    ) 

# resistant hypertension to hypertension codes
disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "I1A.0",
    replacement = "I10, I11, I12, I13, I15"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "J09.X",
    replacement = "J09"
    )
    ) 


disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "S06.0X",
    replacement = "S06.0"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "S06.0XA",
    replacement = "S06.0"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "S06.0A",
    replacement = "S06.0"
    )
    ) 

```

#### Correcting ranges created by UMLS codes

```{r correct_umls_ranges}

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "A15-A19.9|A15-A19",
    replacement = paste0("A", 15:19, collapse = ", ")
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "A50-A64.9|A50-A64",
    replacement = paste0("A", 50:64, collapse = ", ")
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "B20-B24.9",
    replacement = paste0("B", 20:24, collapse = ", ")
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "C00-C97.9",
    replacement = paste0(
                  paste0("C0", 0:9, collapse = ", "),
                  ", ",
                  paste0("C", 10:97, collapse = ", "),
                  collapse = ","
    )
    )
    )

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "C00-D48.9",
    replacement = paste0(
                  paste0("C0", 0:9, collapse = ", "),
                  ", ",
                  paste0("C", 10:90, collapse = ", "),
                  ", ",
                  paste0("D", 0:9, collapse = ", "),
                  ", ",
                  paste0("D", 10:48, collapse = ", "),
                  collapse = ",")
    )
    )

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "C60-C63.9|C60-C63",
    replacement = paste0("C", 60:63, collapse = ", ")
    )
    ) 


disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "D10-D36.9",
    replacement = paste0("D", 10:36, collapse = ", ")
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "D80-D89",
    replacement = paste0("D", 80:89, collapse = ", ")
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "E00-E07.9|E00-E07",
    replacement = paste0("E0", 0:7, collapse = ", ")
    )
    ) 


disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "E08-E13",
    replacement = "E08, E09, E10, E11, E12, E13"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "E10-E14.9",
    replacement = "E10, E11, E12, E13, E14"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "E27.1-E27.4",
    replacement = "E27.1, E27.2, E27.3, E27.4"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "E70-E90.9",
    replacement = paste0("E", 70:90, collapse = ", ")
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "F30-F39",
    replacement = paste0("F", 30:39, collapse = ", ")
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "G00-G99.9|G00-G99",
    replacement = paste0(
                  paste0("G0", 0:9, collapse = ", "),
                  ", ",
                  paste0("G", 10:99, collapse = ", "),
                  collapse = ","
    )
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "H40-H42.9|H40-H42",
    replacement = paste0("H", 40:42, collapse = ", ")
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "I00-I02.9|I00-I02",
    replacement = paste0("I0", 0:2, collapse = ", ")
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "I10-I15.9",
    replacement = paste0("I", 10:15, collapse = ", ")
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "I20-I25.9|I20-I25",
    replacement = paste0("I", 20:25, collapse = ", ")
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "I60-I69.9|I60-I69",
    replacement = paste0("I", 60:69, collapse = ", ")
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "J00-J99.9|J00-J99",
    replacement = paste0(paste0("J0", 0:9, collapse = ", "),
                         ", ",
                         paste0("J", 10:99, collapse = ", "),
                         collapse = ","
    )
    )
    ) 



disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "K00-K93.9|K00-K93",
    replacement = paste0(paste0("K0", 0:9, collapse = ", "),
                         ", ",
                         paste0("K", 10:93, collapse = ", "),
                         collapse = ","
    )
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "K40-K46.9|K40-K46",
    replacement = paste0("K", 40:46, collapse = ", ")
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "K70-K77.9|K70-K77",
    replacement = paste0("K", 70:77, collapse = ", ")
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "L00-L08.9|L00-L08",
    replacement = paste0("L0", 0:8, collapse = ", ")
    )
    ) 


disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "L00-L99.9|L00-L99",
    replacement = paste0(paste0("L0", 0:9, collapse = ", "),
                         ", ",
                         paste0("L", 10:99, collapse = ", "),
                         collapse = ","
    )
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "M00-M25.9",
    replacement = paste0(paste0("M0", 0:9, collapse = ", "),
                         ", ",
                         paste0("M", 10:25, collapse = ", "),
                         collapse = ","
    )
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "M15-M19",
    replacement = paste0("M", 15:19, collapse = ", ")
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "M15-M19",
    replacement = paste0("M", 15:19, collapse = ", ")
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "M30-M36",
    replacement = paste0("M", 30:36, collapse = ", ")
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "M45-M49.9|M45-M49",
    replacement = paste0("M", 45:49, collapse = ", ")
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "M60-M79.9",
    replacement = paste0("M", 60:79, collapse = ", ")
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "N10-N16",
    replacement = paste0("N", 10:16, collapse = ", ")
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "N17-N19.9",
    replacement = paste0("N", 17:19, collapse = ", ")
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "N18.3-N18.9",
    replacement = paste0("N18.", 3:9, collapse = ", ")
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "N20-N23.9|N20-N23",
    replacement = paste0("N", 20:23, collapse = ", ")
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "N60-N64.9",
    replacement = paste0("N", 60:64, collapse = ", ")
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "N60-N65",
    replacement = paste0("N", 60:65, collapse = ", ")
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "O00-O08.9|O00-O08",
    replacement = paste0("O0", 0:8, collapse = ", ")
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "X71-X83",
    replacement = paste0("X", 71:83, collapse = ", ")
    )
    ) 


```

#### Converting ICD-10 cm codes to WHO ICD-10

```{r expand_disease_mappings}

disease_mapping =
  disease_mapping |>
  mutate(icd10_code = str_split(icd10_code, ",\\s*")) |>
  tidyr::unnest(icd10_code) 


disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "B37.49",
    replacement = "B37.4"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "E05.90",
    replacement = "E05.9"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
  mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "E11.319|E11.31|E11.329.9|E11.329.|E11.32|E11.3.",
    replacement = "E11.3"
    )
  )

disease_mapping = 
  disease_mapping |> 
  mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "E11.3.",
    replacement = "E11.3"
    )
  )

disease_mapping = 
  disease_mapping |> 
  mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "E13.621.|E13.62",
    replacement = "E11.3"
    )
  )

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "E78.00",
    replacement = "E78.0"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "F10.10",
    replacement = "F10.1"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "F17.201.|F17.20",
    replacement = "F17.2"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "F430",
    replacement = "F43.0"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "G47.00",
    replacement = "G47.0"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "H01.00|H01.09.",
    replacement = "H01.0"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "H01.09.",
    replacement = "H01.0"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "H029",
    replacement = "H02.9"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "H60.90",
    replacement = "H60.9"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "H60.90",
    replacement = "H60.9"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "H91.8X9.",
    replacement = "H91.8"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "H91.90",
    replacement = "H91.9"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "H93.299.|H93.29",
    replacement = "H01"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "I70.20",
    replacement = "I70.2"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "I82.409.|I82.40|I82.4",
    replacement = "I82"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "J30.9",
    replacement = "J30.2"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
  mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "J45.998.|J45.909.|J45.901.|J45.99|J45.90",
    replacement = "J45.9"
    )
  )

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "J98.457.6",
    replacement = "J98.4"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "K05.30-31",
    replacement = "K05.3"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "K29.70",
    replacement = "K29.7"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "K59.00",
    replacement = "K59.0"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "L08.89",
    replacement = "L08.8"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "M06.99|M06.90",
    replacement = "M06.9"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "M10.99",
    replacement = "M10.9"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "M10.99",
    replacement = "M10.9"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
  mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "M13.90|M13.94|M13.96|M13.97|M13.99",
    replacement = "M13.9"
    )
  )

disease_mapping = 
  disease_mapping |> 
  mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "M19.07",
    replacement = "M19.0"
    )
  )

disease_mapping = 
  disease_mapping |> 
  mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "M19.90|M19.91|M19.94|M19.97|M19.99",
    replacement = "M19.9"
    )
  )



disease_mapping = 
  disease_mapping |> 
  mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "M25.50|M25.51|M25.55|M25.569.|M25.56|M25.571.|M25.57",
    replacement = "M25.5"
    )
  )

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "M25.76|M25.77",
    replacement = "M25.7"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "M43.16",
    replacement = "M43.1"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
  mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "M47.80|M47.82|M47.86",
    replacement = "M47.8"
    )
  )

disease_mapping = 
  disease_mapping |> 
  mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "M47.92|M47.96",
    replacement = "M47.9"
    )
  )

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "M48.02|M48.06",
    replacement = "M48.0"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "M54.22",
    replacement = "M54.2"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "M54.30|M54.39",
    replacement = "M54.3"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "M54.56|M54.57|M54.59",
    replacement = "M54.5"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "M54.99",
    replacement = "M54.9"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "M65.34",
    replacement = "M65.3"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "M65.96",
    replacement = "M65.9"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "M72.04",
    replacement = "M72.0"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "M79.09",
    replacement = "M79.0"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "M79.65|M79.66|M79.67",
    replacement = "M79.6"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "M79.79",
    replacement = "M79.7"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "M79.86",
    replacement = "M79.8"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "M81.99",
    replacement = "M81.9"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "N390",
    replacement = "N39.0"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "N50.89",
    replacement = "N50.8"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "N52",
    replacement = "F52"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "N52.9",
    replacement = "F52.9"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "N814",
    replacement = "N81.4"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "P29.12",
    replacement = "P29.1"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "R06.00|R06.09",
    replacement = "R06.0"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "R06.83",
    replacement = "R06.8"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "R07.89",
    replacement = "R07.8"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "R07.8|R07.9",
    replacement = "R07"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "R09.89",
    replacement = "R09.8"
    )
    ) 


disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "R10.30",
    replacement = "R10.3"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "R10.8|R10.9",
    replacement = "R10"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "R11.0",
    replacement = "R11"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "R13.10",
    replacement = "R13"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "R13.1",
    replacement = "R13"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "R14.0",
    replacement = "R14"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "R19.7",
    replacement = "R19"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "R29.898.|R29.89",
    replacement = "R29.8"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "R31.29",
    replacement = "R31.2"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "R31.2|R31.9",
    replacement = "R31"
    )
    ) 


disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "R53.82|R53.83|R5382",
    replacement = "R53.8"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "R53.8",
    replacement = "R53"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "R56.9",
    replacement = "R56"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
  mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "R73.09|R73.02",
    replacement = "R73.0"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "R80.9",
    replacement = "R80"
    )
    ) 
 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "R82.998.|R82.99",
    replacement = "R82.9"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "R871",
    replacement = "R87.1"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "R87.6",
    replacement = "R87"
    )
    ) 


disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "R87.61",
    replacement = "R87.6"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "T50.905.",
    replacement = "T50.9"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "T78.40",
    replacement = "T78.4"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "T780",
    replacement = "T78.0"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "T81.149.88",
    replacement = "T81.1"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "T81.815.013.",
    replacement = "T81.8"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "T84.84",
    replacement = "T84.8"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "T887",
    replacement = "T88.7"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "U80",
    replacement = "U82"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "W44.9",
    replacement = "W44"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "Z86.79",
    replacement = "Z86.7"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "Z87.09",
    replacement = "Z87.0"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "Z87.39",
    replacement = "Z87.3"
    )
    ) 


disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "Z87.42",
    replacement = "Z87.4"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "Z87.828.",
    replacement = "Z87.8"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "Z87.891.",
    replacement = "Z87.8"
    )
    ) 




disease_mapping =
  disease_mapping |>
  distinct()




```

### Add ICD10 descriptions

#### Other ICD10 descriptions to add

```{r other_icd10_descriptions}


other_icd10_desc = 
  data.frame(
    icd10_code = c("A09.9", 
                  "A41",
                  "B95",
                  "B96",
                  "B97",
                  "B98",
                  "C44",
                  "C79",
                  "C80.0",
                  "C80.9",
                  "C90",
                  "D07",
                  "D35",
                  "D37",
                  "D47",
                  "D48",
                  "E03",
                  "E04",
                  "E21",
                  "E23",
                  "E27",
                  "E61",
                  "E80",
                  "E87",
                  "E89",
                  "F05",
                  "F17",
                  "F41",
                  "F43.0",
                  "F53",
                  "G31",
                  "G45",
                  "G46",
                  "G62",
                  "G93",
                  "G99",
                  "H47",
                  "H50",
                  "H57",
                  "H83",
                  "H91",
                  "H92",
                  "H93",
                  "I08",
                  "I27",
                  "I27.2",
                  "I44",
                  "I45",
                  "I48.0",
                  "I49",
                  "I51",
                  "I62",
                  "I71",
                  "J15",
                  "J16",
                  "J34",
                  "J38",
                  "J69",
                  "J84",
                  "J95",
                  "J96",
                  "K05",
                  "K07",
                  "K08",
                  "K09",
                  "K10",
                  "K12",
                  "K13",
                  "K22",
                  "K22.7",
                  "K52",
                  "K56",
                  "K64",
                  "K64.0",
                  "K64.9",
                  "K74",
                  "K75",
                  "K85.9",
                  "K86",
                  "K91",
                  "K91.8",
                  "K92",
                  "L02",
                  "L13",
                  "L30",
                  "L65",
                  "L73",
                  "L85",
                  "L89.1",
                  "L98",
                  "M05",
                  "M06",
                  "M07",
                  "M11",
                  "M13",
                  "M18",
                  "M19",
                  "M24",
                  "M25",
                  "M31",
                  "M35",
                  "M43",
                  "M48",
                  "M53",
                  "M62",
                  "M66",
                  "M67",
                  "M71",
                  "M77",
                  "M79",
                  "M79.7",
                  "M80",
                  "M81",
                  "M85",
                  "M89",
                  "M96",
                  "N02",
                  "N18.3",
                  "N18.4",
                  "N28",
                  "N39",
                  "N48",
                  "N73",
                  "N76",
                  "N88",
                  "N89",
                  "N91",
                  "N92",
                  "N93",
                  "N94",
                  "N99",
                  "O14",
                  "O04",
                  "O26",
                  "O32",
                  "O34",
                  "O36",
                  "O68",
                  "O75",
                  "O99",
                  "R03",
                  "R07",
                  "R09",
                  "R10",
                  "R19",
                  "R29",
                  "R29.6",
                  "R39",
                  "R40",
                  "R47",
                  "R57",
                  "T95.8",
                  "W44",
                  "Y95",
                  "Z86.3",
                  "Z87.3",
                  "Z87.4",
                  "Z87.7",
                  "Z87.8",
                  "T85.8",
                  "Z91.0",
                  "Z88.9",
                  "Z88.8",
                  "Z88",
                  "Z92.6",
                  "N90",
                  "U82"
                  ),
    icd10_description = c("Gastroenteritis and colitis of unspecified origin",
                         "Other sepsis",
                         "Streptococcus and staphylococcus as the cause of diseases classified to other chapters",
                         "Other specified bacterial agents as the cause of diseases classified to other chapters",
                         "Viral agents as the cause of diseases classified to other chapters",
                         "Other specified infectious agents as the cause of diseases classified to other chapters",
                         "Other malignant neoplasms of skin",
                         "Secondary malignant neoplasm of other and unspecified sites",
                         "Malignant neoplasm, primary site unknown, so stated",
                         "Malignant neoplasm, primary site unspecified",
                         "Multiple myeloma and malignant plasma cell neoplasms",
                         "Carcinoma in situ of other and unspecified genital organs",
                         "Benign neoplasm of other and unspecified endocrine glands",
                         "Neoplasm of uncertain or unknown behaviour of oral cavity and digestive organs",
                         "Other neoplasms of uncertain or unknown behaviour of lymphoid, haematopoietic and related tissue",
                         "Neoplasm of uncertain or unknown behaviour of other and unspecified sites",
                         "Other hypothyroidism",
                         "Other nontoxic goitre",
                         "Hyperparathyroidism and other disorders of parathyroid gland",
                         "Hypofunction and other disorders of pituitary gland",
                         "Other disorders of adrenal gland",
                         "Deficiency of other nutrient elements",
                         "Disorders of porphyrin and bilirubin metabolism",
                         "Other disorders of fluid, electrolyte and acid-base balance",
                         "Postprocedural endocrine and metabolic disorders, not elsewhere classified",
                         "Delirium, not induced by alcohol and other psychoactive substances",
                         "Mental and behavioural disorders due to use of tobacco",
                         "Other anxiety disorders",
                         "Acute stress reaction",
                         "Mental and behavioural disorders associated with the puerperium, not elsewhere classified",
                         "Other degenerative diseases of nervous system, not elsewhere classified",
                         "Transient cerebral ischaemic attacks and related syndromes",
                         "Vascular syndromes of brain in cerebrovascular diseases",
                         "Other polyneuropathies",
                         "Other disorders of brain",
                         "Other disorders of nervous system in diseases classified elsewhere",
                         "Other disorders of optic [2nd] nerve and visual pathways",
                         "Other strabismus",
                         "Other disorders of eye and adnexa",
                         "Other diseases of inner ear",
                         "Other hearing loss",
                         "Otalgia and effusion of ear",
                         "Other disorders of ear, not elsewhere classified",
                         "Multiple valve diseases",
                         "Other pulmonary heart diseases",
                         "Other secondary pulmonary hypertension",
                         "Atrioventricular and left bundle-branch block",
                         "Other conduction disorders",
                         "Paroxysmal atrial fibrillation",
                         "Ventricular fibrillation and flutter",
                         "Complications and ill-defined descriptions of heart disease",
                         "Other nontraumatic intracranial haemorrhage",
                         "Aortic aneurysm and dissection",
                         "Bacterial pneumonia, not elsewhere classified",
                         "Pneumonia due to other infectious organisms, not elsewhere classified",
                         "Other disorders of nose and nasal sinuses",
                         "Diseases of vocal cords and larynx, not elsewhere classified",
                         "Pneumonitis due to solids and liquids",
                         "Other interstitial pulmonary diseases",
                         "Postprocedural respiratory disorders, not elsewhere classified",
                         "Respiratory failure, not elsewhere classified",
                         "Gingivitis and periodontal diseases",
                         "Dentofacial anomalies [including malocclusion]",
                         "Other disorders of teeth and supporting structures",
                         "Cysts of oral region, not elsewhere classified",
                         "Other diseases of jaws",
                         "Stomatitis and related lesions",
                         "Other diseases of lip and oral mucosa",
                         "Other diseases of oesophagus",
                         "Barrett oesophagus",
                         "Other noninfective gastroenteritis and colitis",
                         "Paralytic ileus and intestinal obstruction without hernia",
                         "Haemorrhoids and perianal venous thrombosis",
                         "First degree haemorrhoids",
                         "Haemorrhoids, unspecified",
                         "Fibrosis and cirrhosis of liver",
                         "Other inflammatory liver diseases",
                         "Acute pancreatitis, unspecified",
                         "Other diseases of pancreas",
                         "Postprocedural disorders of digestive system, not elsewhere classified",
                         "Other postprocedural disorders of digestive system, not elsewhere classified",
                         "Other diseases of digestive system",
                         "Cutaneous abscess, furuncle and carbuncle",
                         "Other bullous disorders",
                         "Other dermatitis",
                         "Other nonscarring hair loss",
                         "Other follicular disorders",
                         "Other epidermal thickening",
                         "Stage II decubitus ulcer",
                         "Other disorders of skin and subcutaneous tissue, not elsewhere classified",
                         "Seropositive rheumatoid arthritis",
                         "Other rheumatoid arthritis",
                         "Psoriatic and enteropathic arthropathies",
                         "Other crystal arthropathies",
                         "Other arthritis",
                         "Arthrosis of first carpometacarpal joint",
                         "Other arthrosis",
                         "Other specific joint derangements",
                         "Other joint disorders, not elsewhere classified",
                         "Other necrotizing vasculopathies",
                         "Other systemic involvement of connective tissue",
                         "Other deforming dorsopathies",
                         "Other spondylopathies",
                         "Other dorsopathies, not elsewhere classified",
                         "Other disorders of muscle",
                         "Spontaneous rupture of synovium and tendon",
                         "Other disorders of synovium and tendon",
                         "Other bursopathies",
                         "Other enthesopathies",
                         "Other soft tissue disorders, not elsewhere classified",
                         "Fibromyalgia",
                         "Osteoporosis with pathological fracture",
                         "Osteoporosis without pathological fracture",
                         "Other disorders of bone density and structure",
                         "Other disorders of bone",
                         "Postprocedural musculoskeletal disorders, not elsewhere classified",
                         "Recurrent and persistent haematuria",
                         "Chronic kidney disease, stage 3",
                         "Chronic kidney disease, stage 4",
                         "Other disorders of kidney and ureter, not elsewhere classified",
                         "Other disorders of urinary system",
                         "Other disorders of penis",
                         "Other female pelvic inflammatory diseases",
                         "Other inflammation of vagina and vulva",
                         "Other noninflammatory disorders of cervix uteri",
                         "Other noninflammatory disorders of vagina",
                         "Other noninflammatory disorders of vulva and perineum",
                         "Excessive, frequent and irregular menstruation",
                         "Other abnormal uterine and vaginal bleeding",
                         "Pain and other conditions associated with female genital organs and menstrual cycle",
                         "Postprocedural disorders of genitourinary system, not elsewhere classified",
                         "Pre-eclampsia",
                         "Medical abortion",
                         "Maternal care for other conditions predominantly related to pregnancy",
                         "Maternal care for known or suspected malpresentation of fetus",
                         "Maternal care for known or suspected abnormality of pelvic organs",
                         "Maternal care for other known or suspected fetal problems",
                         "Labour and delivery complicated by fetal stress [distress]",
                         "Other complications of labour and delivery, not elsewhere classified",
                         "Other maternal diseases classifiable elsewhere but complicating pregnancy, childbirth and the puerperium",
                         "Abnormal blood-pressure reading, without diagnosis",
                         "Pain in throat and chest",
                         "Other symptoms and signs involving the circulatory and respiratory systems",
                         "Abdominal and pelvic pain",
                         "Other symptoms and signs involving the digestive system and abdomen",
                         "Other symptoms and signs involving the nervous and musculoskeletal systems",
                         "Tendency to fall, not elsewhere classified",
                         "Other symptoms and signs involving the urinary system",
                         "Somnolence, stupor and coma",
                         "Speech disturbances, not elsewhere classified",
                         "Shock, not elsewhere classified",
                         "Other complications of internal prosthetic devices, implants and grafts, not elsewhere classified",
                         "Foreign body entering into or through eye or natural orifice",
                         "Nosocomial condition",
                         "Personal history of endocrine, nutritional and metabolic diseases",
                         "Personal history of diseases of the musculoskeletal system and connective tissue",
                         "Personal history of diseases of the genitourinary system",
                         "Personal history of congenital malformations, deformations and chromosomal abnormalities",
                         "Personal history of other specified conditions",
                         "Other complications of internal prosthetic devices, implants and grafts, not elsewhere classified",
                         "Personal history of allergy, other than to drugs and biological substances",
                         "Personal history of allergy to unspecified drugs, medicaments and biological substances",
                         "Personal history of allergy to other drugs, medicaments and biological substances
",
"Personal history of allergy to drugs, medicaments and biological substances",
"Personal history of chemotherapy for neoplastic disease",
"Other noninflammatory disorders of vulva and perineum",
"Resistance to betalactam antibiotics")
  )

```

```{r add_icd10_descriptions}

manual_icd10_map <-
  readxl::read_xlsx(here::here("data/icd/manual_disease_icd10_mappings.xlsx"))

icd10_descriptions =
  phecodes |>
  select(icd10_code = ICD10, 
         icd10_description = ICD_DESCRIPTION
         ) |>
  distinct()

# Expand multiple ICD codes into rows
to_add_expanded <- manual_icd10_map |>
  mutate(icd10_code = str_split(icd10_code, ",\\s*")) |>
  tidyr::unnest(icd10_code)

icd10_descriptions = 
  bind_rows(
    icd10_descriptions,
    to_add_expanded |>
      select(icd10_code, 
             icd10_description = icd10_desc),
    other_icd10_desc
  ) 

icd10_descriptions = icd10_descriptions |> distinct()

icd10_descriptions = 
  icd10_descriptions |>
  group_by(icd10_code) |>
  summarise(icd10_description = 
            str_flatten(unique(icd10_description), collapse = "; ", na.rm = T), 
            .groups = "drop"
            )

disease_mapping =
  left_join(disease_mapping,
            icd10_descriptions,
            by = "icd10_code",
            relationship = "many-to-one",
            na_matches = "never"
            ) 


disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "R871",
    replacement = "R87"
    )
    ) 


disease_mapping = 
  disease_mapping |> 
    mutate(icd10_description = 
    ifelse(icd10_code == "R87",
           "Abnormal findings in specimens from female genital organs",
           icd10_description)
    ) 


```

### Missing descriptions for ICD10 codes

```{r missing_icd10_descriptions}
# check if any ICD10 codes are missing descriptions
disease_mapping |> 
  filter(is.na(icd10_description)) |>
  head()
```

### Noticing some mistakes in MAPPED trait by icd 10 codes

```{r check_mapped_trait}

gwas_study_info |> 
  filter(grepl("ICD10 F05", 
               `DISEASE/TRAIT`)) |> 
  select(`DISEASE/TRAIT`, MAPPED_TRAIT, collected_all_disease_terms, PUBMED_ID)


disease_mapping |> 
  filter(icd10_code == "F05")

# F05 refers to delirium not induced by alcohol and other psychoactive substances
# yet, the mapped trait is alcohol withdrawal delirium
# replace with delirium
disease_mapping = 
disease_mapping |>
  mutate(collected_all_disease_terms = 
         ifelse(!is.na(icd10_code) & icd10_code == "F05",
                "delirium",
                collected_all_disease_terms)
         )

gwas_study_info =
  gwas_study_info |>
  mutate(MAPPED_TRAIT = 
         ifelse(grepl("ICD10 F05", `DISEASE/TRAIT`),
                str_replace_all(collected_all_disease_terms,
                                pattern = "alcohol-related disorders",
                                replacement = "delirium"),
                collected_all_disease_terms)
         )

```

### Save disease map

```{r save_disease_mapping}

disease_mapping =
  disease_mapping |>
  mutate(icd10_code = sub("^([^-]+)-\\1$", "\\1", icd10_code))

disease_mapping = 
  disease_mapping |> 
  arrange(YEAR,
          PUBMED_ID,
          STUDY_ACCESSION,
          collected_all_disease_terms, 
          icd10_code)

fwrite(disease_mapping,
       here::here("output/icd_map/gwas_disease_to_icd10_mapping.csv")
       )

```


        


