---
title: "Mapping GWAS traits to ICD 10"
author: "Isobel Beasley"
output: html_document
---

# Set up

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load_packages, message=FALSE}

library(dplyr)
library(stringr)
library(data.table)

```

### Load Data

```{r load_data}

# load GWAS study info with harmonised disease trait labels
# and author provided ICD-10 Codes 
disease_mapping <- data.table::fread(here::here("output/gwas_cat/gwas_author_provided_icd10.csv"))

disease_mapping <-
  disease_mapping |>
  mutate(icd10_code = ifelse(icd10_code == "",
                             NA,
                             icd10_code)) |>
  mutate(icd10_source_col = ifelse(icd10_source_col == "",
                                    NA,
                                    icd10_source_col))

n_rows <- nrow(disease_mapping)

gwas_study_info <- data.table::fread(here::here("output/gwas_cat/gwas_study_all_group.csv"))

gwas_study_info = gwas_study_info |>
    mutate(collected_all_disease_terms = stringi::stri_trans_general(collected_all_disease_terms, "Latin-ASCII")
           )

```


# Tidy MAPPED TRAIT TERMS

```{r}

# replace non-specific disease terms with 'NS'
# so we don't try to map them to ICD-10 codes

ns_disease_terms <- c("disease prognosis", 
                      "disease progression", 
                      "progression free survival",
                      "disease free survival",
                      "event free survival time",
                      "survival time",
                      "overall survival",
                      "illness severity status",
                      "mild disease course",
                      "complicated disease course",
                      "disease", 
                      "chronic disease",
                      "disease risk", 
                      "disease recurrence",
                      "remission",
                      "age at assessment",
                      "age at diagnosis",
                      "age at onset",
                      "complication",
                      "cancer aggressiveness",
                      "formal thought disorder",
                      "date of diagnosis",
                      "sign or symptom",
                      "seropositivity")

disease_mapping <- disease_mapping |>
  mutate(collected_all_disease_terms = ifelse(collected_all_disease_terms %in% ns_disease_terms,
                                              "NS",
                                              collected_all_disease_terms)                 
                      )

```

# Tidy `DISEASE/TRAIT` Column 

Do this to enable term matching with PheCode, ICD-10 and UMLS descriptions. 

### Removing model related terms 

```{r remove_model_terms, message=FALSE}

source(here::here("code/clean_trait_metadata.R"))

disease_mapping <- disease_mapping |>
                   mutate(trait_clean = clean_trait_documentation(`DISEASE/TRAIT`))


disease_mapping <- disease_mapping |>
  mutate(
trait_context_clean = str_match(
      trait_clean,
      "(?i)\\s(?:levels?|counts?|ratios?|activity|abundance|concentrations?)\\sin\\s(.+)$"
    )[, 2] |>
      str_remove("(?i)\\s+(?:with hypertension and no diabetes|patients|individuals)$") |> 
      str_remove_all("individuals with\\s+") 
  )

disease_mapping <-
  disease_mapping |>
  mutate(icd10_code = ifelse(icd10_code == "",
                             NA,
                             icd10_code)) |>
  mutate(icd10_source = ifelse(icd10_source == "",
                             NA,
                             icd10_source)) |>
  mutate(icd10_source_col = ifelse(icd10_source_col == "",
                             NA,
                             icd10_source_col)
         )

```

# Get ICD10 codes by matching `DISEASE/TRAIT` terms 

### ICD 10 descriptions  - match disease/trait

#### Cleaned DISEASE/TRAIT

```{r match_disease_terms_to_icd_descriptions, message=FALSE}

# phecode to ICD10 mapping from https://wei-lab.app.vumc.org/phecode-data/phecode_international_version

phecodes <- data.table::fread(here::here("data/icd/phecode_international_version_unrolled.csv"))

bad_phecode_icd_map = 
  data.frame(
    ICD10 = c("I08.1", "I08.2"),
    ICD_DESCRIPTION = c(
      "Disorders of both mitral and tricuspid valves",
      "Disorders of both aortic and tricuspid valves"),
    PheCode = c(395.1, 395.2),
    Phenotypes = c(
      "Nonrheumatic mitral valve disorders",
      "Nonrheumatic aortic valve disorders"
    )
  )

phecodes <-
dplyr::rows_delete(phecodes,
                   bad_phecode_icd_map,
                   by = c("ICD10", "PheCode"))

icd10_desc_map  = 
phecodes |>
    mutate(trait_clean = tolower(iconv(ICD_DESCRIPTION, 
                                                       to = "UTF-8"))
           ) |>
    select(trait_clean,    
           icd10_code = ICD10) 

# filter missing strings
icd10_desc_map  = 
  icd10_desc_map |>
  filter(!is.na(trait_clean) & 
         trait_clean != "") |>
  filter(!is.na(icd10_code) & 
         icd10_code != "")

matched_icd10_desc =
  icd10_desc_map  |>
  group_by(trait_clean) |>
  summarise(icd10_code = str_flatten(unique(icd10_code), 
                                     collapse = ", ", 
                                     na.rm = T), 
            .groups = "drop") 

# diabetic retinopathy is only mapped to H36.0 
# but should also be mapped to E10.3, E11.3, E13.3, E14.3

matched_icd10_desc =
  matched_icd10_desc |>
  mutate(icd10_code = ifelse(trait_clean == "diabetic retinopathy",
                             "H36.0, E10.3, E11.3, E13.3, E14.3",
                             icd10_code))
  

# match by cleaned DISEASE/TRAIT
matched_icd10_desc = 
  matched_icd10_desc |>
    mutate(icd10_source = "ICD Description") |>
    mutate(icd10_source_col = "DISEASE/TRAIT")

disease_mapping = 
disease_mapping |>
rows_patch(matched_icd10_desc,
           unmatched = "ignore",
           by = "trait_clean")

```

#### Extracted DISEASE from `DISEASE/TRAIT` column

```{r match_extracted_disease_to_icd10_descriptions, message=FALSE}

# match by cleaned context from DISEASE/TRAIT
matched_icd10_desc = 
 matched_icd10_desc |>
 rename(trait_context_clean = trait_clean) |>
 mutate(icd10_source_col = "Extracted disease from DISEASE/TRAIT")
  
disease_mapping = 
disease_mapping |>
rows_patch(matched_icd10_desc,
           unmatched = "ignore",
           by = "trait_context_clean")

print("Number of ICD10 codes obtained")
disease_mapping |>
    filter(!is.na(icd10_code)) |> 
  group_by(icd10_source_col) |> 
  summarise(n = n(), 
            percent = n()/n_rows*100)

print("Number of papers where at least one ICD-10 has been obtained so far:")
disease_mapping |>
  filter(!is.na(icd10_code)) |>
  pull(PUBMED_ID) |>  
  unique() |>
  length()

disease_mapping |>
    filter(!is.na(icd10_code)) |> 
  group_by(icd10_source_col,
           icd10_source) |> 
  summarise(n = length(unique(PUBMED_ID)))

```

### Phenotypes (corresponding to PheCodes) - match disease/trait

#### Cleaned DISEASE/TRAIT

```{r match_disease_terms_to_phenotypes, message=FALSE}

phenotype_icd_map =
  phecodes |>
group_by(Phenotype) |>
summarise(icd10_code = 
            str_flatten(ICD10, 
                        collapse = ", ", 
                        na.rm = T), 
            .groups = "drop")

# match by cleaned DISEASE/TRAIT
matched_phenotypes =
phenotype_icd_map  |>
  mutate(trait_clean = tolower(Phenotype)) |>
  select(trait_clean, icd10_code) |>
  mutate(icd10_source_col = "DISEASE/TRAIT") |>
  mutate(icd10_source = "Phecode Phenotype")

disease_mapping =
disease_mapping |>
rows_patch(matched_phenotypes,
           unmatched = "ignore",
           by = "trait_clean")


```

#### Extracted DISEASE from `DISEASE/TRAIT` column

```{r match_extracted_disease_to_phenotypes, message=FALSE}

# match by extracted DISEASE from `DISEASE/TRAIT` column
matched_phenotypes =
matched_phenotypes |>
  rename(trait_context_clean = trait_clean) |>
  mutate(icd10_source_col = "Extracted disease from DISEASE/TRAIT")


disease_mapping =
disease_mapping |>
rows_patch(matched_phenotypes,
           unmatched = "ignore",
           by = "trait_context_clean")

disease_mapping |>
    filter(!is.na(icd10_code)) |> 
  group_by(icd10_source_col) |> 
  summarise(n = n(), 
            percent = n()/n_rows*100)

print("Number of papers where at least one ICD-10 has been obtained so far:")
disease_mapping |>
    filter(!is.na(icd10_code)) |>
  pull(PUBMED_ID) |>  
  unique() |>
  length()

disease_mapping |>
    filter(!is.na(icd10_code)) |> 
  group_by(icd10_source_col,
           icd10_source) |> 
  summarise(n = length(unique(PUBMED_ID)))
         
```

### UMLS - match disease/trait

```{bash eval =FALSE}

# extract CUIs for only ICD-10 and ICD-10CM UMLS terms
awk -F'|' '$2=="ENG" && ($12=="ICD10" || $12=="ICD10CM") {print $1}' \
  data/icd/2025AA/META/MRCONSO.RRF |
  sort -u \
  > data/icd/2025AA/META/ICD10_CUIs.txt
  
# Filter original MRCONSO using those CUIs
awk -F'|' 'NR==FNR {cui[$1]; next} $1 in cui && $2=="ENG"' \
  data/icd/2025AA/META/ICD10_CUIs.txt \
  data/icd/2025AA/META/MRCONSO.RRF \
  > data/icd/2025AA/META/MRCONSO_ICD10_ICD10CM.RRF

```

```{r make_filtered_umls_icd10, message=FALSE}

umls_data <-
  data.table::fread(here::here("data/icd/2025AA/META/MRCONSO_ICD10_ICD10CM.RRF"),
                    sep = "|",
                    header = FALSE,
                    quote = "",
                    fill = TRUE,
                    select = c(1, 12, 14, 15),
                    na.strings = c("", "NA")
  )

colnames(umls_data) <- c("CUI",  
                         "SAB",  
                         "CODE", 
                         "STR")

# what medical CUIs have ICD-10 or ICD-CM Codes? 
umls_cuis_icd10 <-
  umls_data |>
  filter(SAB %in% c("ICD10", "ICD10CM")) |>
  select(CODE, CUI, code_type = SAB) |> 
  group_by(CUI) |>
  summarise(CODE = str_flatten(unique(CODE),
                               collapse = ", ",
                               na.rm = T),
            code_type = str_flatten(unique(code_type),
                               collapse = ", ",
                               na.rm = T),
            .groups = "drop"
            ) |>
  rename(icd10_code = CODE) 

umls_icd10 =
  umls_data |>
  left_join(umls_cuis_icd10,
            by = "CUI")

umls_icd10  <- 
  umls_icd10  |>
  select(STR, icd10_code, code_type) |>
  mutate(STR = tolower(STR)) |>
  distinct()

umls_icd10 <-
  umls_icd10 |>
  filter(STR != "hypertension") |>
  filter(STR != "cancer") |>
  filter(STR != "high blood pressure")

# where K76.1 is mapped to coronary artery disease
# remove 
umls_icd10 |>
    filter(STR == "coronary artery disease")

umls_icd10 <-
  umls_icd10 |>
  mutate(icd10_code = ifelse(STR == "coronary artery disease",
                             str_remove_all(string = icd10_code,
                                            pattern = ", K76.1"),
                             icd10_code))

umls_icd10  <-
  umls_icd10  |>
  group_by(STR) |>
  summarise(icd10_code = str_flatten(unique(
                              unlist(str_split(icd10_code, ", "))
                              ),
                               collapse = ", ",
                               na.rm = T),
            code_type = str_flatten(unique(
                               unlist(str_split(code_type, ", "))
                               ),
                               collapse = ", ",
                               na.rm = T),
            .groups = "drop"
            ) 

umls_icd10 <-
  umls_icd10 |>
  mutate(icd10_source = "UMLS")

head(umls_icd10)

```

#### Cleaned DISEASE/TRAIT

```{r match_disease_terms_to_umls, message=FALSE}

unmapped_disease_terms <-
  disease_mapping |>
  filter(is.na(icd10_code)) |>
  pull(trait_clean) |>
  unique() |>
  tolower()

# overlap with umls terms
disease_trait_umls  <-
  umls_icd10 |>
  select(trait_clean = STR,
         icd10_code,
         icd10_source) |>
  distinct() |>
  mutate(icd10_source_col = "DISEASE/TRAIT")


 disease_mapping <-
  disease_mapping |>
  rows_patch(disease_trait_umls,
             by = 'trait_clean',
             unmatched = "ignore")
 
```

#### Match trait context to UMLS

```{r match_trait_context_to_umls, message=FALSE}

unmapped_disease_terms <- 
    disease_mapping |>
    filter(is.na(icd10_code)) |> 
    pull(trait_context_clean) |>
    unique() |> 
    tolower()

# overlap with umls terms
disease_trait_umls  <-
  umls_icd10 |>
  select(trait_context_clean = STR,
         icd10_code,
         icd10_source) |>
  distinct() |>
  mutate(icd10_source_col = "Extracted disease from DISEASE/TRAIT")

 disease_mapping <-
  disease_mapping |>
  mutate(trait_context_clean = tolower(trait_context_clean)) |>
  rows_patch(disease_trait_umls,
             by = 'trait_context_clean',
             unmatched = "ignore")
 
 disease_mapping |>
    filter(!is.na(icd10_code)) |> 
  group_by(icd10_source_col) |> 
  summarise(n = n(), 
            percent = n()/n_rows*100)
 
 print("Number of papers where at least one ICD-10 has been obtained so far:")
disease_mapping |>
    filter(!is.na(icd10_code)) |>
  pull(PUBMED_ID) |>  
  unique() |>
  length()

```

### Manual mapping of GWAS traits to ICD 10

```{r manual_map_icd10_disease_trait}

manual_icd10_map <-
  readxl::read_xlsx(here::here("data/icd/manual_disease_mappings_v2.xlsx"))

manual_icd10_map  <-
  manual_icd10_map  |> 
  mutate(mapped_trait = clean_trait_documentation(mapped_trait))

manual_icd10_map_trait_clean =
  manual_icd10_map |>
  select(trait_clean = mapped_trait, 
         icd10_code) |>
  mutate(icd10_source_col = "DISEASE/TRAIT") |>
  mutate(icd10_source = "Manual Map") |> 
  distinct()

disease_mapping =
  rows_patch(disease_mapping,
             manual_icd10_map_trait_clean,
             unmatched = "ignore",
             by = "trait_clean")

```

# Get ICD10 codes by matching `collected_all_disease_terms` terms

### Match collected_all_disease_terms to ICD10 descriptions

```{r match_collected_icd10_desc}

matched_icd10_desc = 
  matched_icd10_desc |>
  rename(collected_all_disease_terms = trait_context_clean)

matched_icd10_desc = 
  matched_icd10_desc |>
  mutate(icd10_source_col = "MAPPED DISEASE TERM")

disease_mapping =
  rows_patch(disease_mapping,
             matched_icd10_desc,
             unmatched = "ignore")

 print("Number of papers where at least one ICD-10 has been obtained so far:")
disease_mapping |>
    filter(!is.na(icd10_code)) |>
  pull(PUBMED_ID) |>  
  unique() |>
  length()

```

### Match collected_all_disease_terms to ICD10 codes (by phenotypes)

```{r match_collected_icd10}

matched_phenotypes = 
  matched_phenotypes |>
  rename(collected_all_disease_terms = trait_context_clean)

matched_phenotypes = 
  matched_phenotypes |>
  mutate(icd10_source_col = "MAPPED DISEASE TERM")

# match by collected_all_disease_terms
disease_mapping = 
disease_mapping |>
rows_patch(matched_phenotypes,
           unmatched = "ignore",
           by = "collected_all_disease_terms")

print("Number of papers where at least one ICD-10 has been obtained so far:")
disease_mapping |>
    filter(!is.na(icd10_code)) |>
  pull(PUBMED_ID) |>  
  unique() |>
  length()

```

# Manual mapping of GWAS traits to ICD 10

### From Excel sheet

```{r manual_mapping_mapped_disease_term}

manual_icd10_map_collected_trait =
  manual_icd10_map |>
  select(trait_clean = mapped_trait, 
         icd10_code) |>
  rename(collected_all_disease_terms = trait_clean) |>
  mutate(icd10_source_col = "MAPPED DISEASE TERM") |>
  mutate(icd10_source = "Manual Map") |>
  distinct()

disease_mapping =
  rows_patch(disease_mapping,
             manual_icd10_map_collected_trait,
             unmatched = "ignore",
             by = "collected_all_disease_terms")

disease_mapping =
  disease_mapping |>
  mutate(icd10_code = stringr::str_remove_all(string = icd10_code,
                                              pattern = ", $"))

```

### UMLS match collected_all_disease_terms

```{r match_collected_umls, message=FALSE}

unmapped_terms <-
  disease_mapping |>
  filter(is.na(icd10_code)) |>
  pull(collected_all_disease_terms) |>
  unique()

# overlap with umls terms
collected_trait_umls  <-
  umls_icd10 |>
  select(collected_all_disease_terms = STR,
         icd10_code) |>
  distinct() |>
  mutate(icd10_source_col = "MAPPED DISEASE TERM")

disease_mapping <-
  disease_mapping |>
  rows_patch(collected_trait_umls,
             by = "collected_all_disease_terms",
             unmatched = "ignore")


print("Number of papers where at least one ICD-10 has been obtained so far:")
disease_mapping |>
    filter(!is.na(icd10_code)) |>
  pull(PUBMED_ID) |>  
  unique() |>
  length()


```

# How was ICD-10 code inferred?

```{r how_was_icd10_inferred}

# number of ICD-10 codes obtained by source
disease_mapping |>
  group_by(icd10_source) |>
  summarise(n = n())

# number of papers with ICD-10 retrieved, by sources
disease_mapping |>
  group_by(icd10_source) |>
  summarise(n_papers = length(unique(PUBMED_ID)))

# number of ICD-10 codes obtained by column (i.e. whether from MAPPED TERM or DISEASE/TRAIT)
disease_mapping |>
  group_by(icd10_source_col) |>
  summarise(n = n())

disease_mapping |>
  group_by(icd10_source_col) |>
  summarise(n_papers = length(unique(PUBMED_ID)))

disease_mapping  |> 
    group_by(icd10_source, icd10_source_col) |> 
    summarise(n = n(), 
              n_papers = length(unique(PUBMED_ID))
    ) 

# number of papers with at least one ICD-10 Code retrieved
disease_mapping |>
  group_by(PUBMED_ID) |>
  summarise(at_least_one_icd10 = any(!is.na(icd10_code))) |>
  summarise(n_papers_with_icd10 = sum(at_least_one_icd10))

# total number of papers
disease_mapping |> pull(PUBMED_ID) |> unique() |> length()

```

# Saving disease mapping 

## Cleaning ICD 10 Codes

```{r cleaning_icd10_codes}


# for cleaning ICD-10 Codes, i.e. expanding code ranges to include all subcomponents, and converting 3 dp codes to 1 dp codes
source(here::here("code/paper_to_icd10_mapping/clean_icd10_codes_fns.R"))

disease_mapping <-
  disease_mapping |> 
  dplyr::mutate(icd10_code_clean = normalize_icd10_codes(icd10_code))

ICD10_CODE_RE <- "^[A-Z][0-9][0-9A-Z](\\.[0-9A-Z]{1,4})?$"

# two of those joined by a hyphen
ICD10_RANGE_RE <- paste0(
  "^([A-Z][0-9][0-9A-Z](?:\\.[0-9A-Z]{1,4})?)",
  "-",
  "([A-Z][0-9][0-9A-Z](?:\\.[0-9A-Z]{1,4})?)$"
)

disease_mapping <-
  disease_mapping |> 
  tidyr::separate_longer_delim(icd10_code_clean, delim = ", ") |>
    mutate(token_kind = case_when(
      str_detect(icd10_code_clean, ICD10_CODE_RE)  ~ "code",
      str_detect(icd10_code_clean, ICD10_RANGE_RE) ~ "range",
      is.na(icd10_code_clean)                        ~ "missing",
      TRUE                              ~ "malformed"
    ))

disease_mapping |> 
  group_by(token_kind) |>
  summarise(n = n())

# initial simplification of ranges 
# e.g. reduce C30-C30 to C30, and C30-C30.9 to C30
disease_mapping <-
  disease_mapping |> 
  rowwise() |>
  mutate(icd10_code_clean = ifelse(token_kind == "range",
                             simplify_range(icd10_code_clean),
                             icd10_code_clean)
         ) |>
  ungroup() |>
  tidyr::separate_longer_delim(icd10_code_clean, 
                               delim = ", ") 
  

disease_mapping <-
  disease_mapping |> 
    mutate(token_kind = case_when(
      str_detect(icd10_code_clean, ICD10_CODE_RE)  ~ "code",
      str_detect(icd10_code_clean, ICD10_RANGE_RE) ~ "range",
      is.na(icd10_code_clean)                        ~ "missing",
      TRUE                              ~ "malformed"
    ))

disease_mapping |> 
  group_by(token_kind) |>
  summarise(n = n())

# expand ranges to include all subcomponents
disease_mapping <-
  disease_mapping |> 
  rowwise() |>
  mutate(icd10_code_clean = ifelse(token_kind == "range",
                             expand_range(icd10_code_clean),
                             icd10_code_clean)
         ) |>
  ungroup() |>
  tidyr::separate_longer_delim(icd10_code_clean, delim = ", ") 

disease_mapping <-
  disease_mapping |> 
    mutate(token_kind = case_when(
      str_detect(icd10_code_clean, ICD10_CODE_RE)  ~ "code",
      str_detect(icd10_code_clean, ICD10_RANGE_RE) ~ "range",
      is.na(icd10_code_clean)                        ~ "missing",
      TRUE                              ~ "malformed"
    ))

disease_mapping |> 
  group_by(token_kind) |>
  summarise(n = n())

disease_mapping <-
  disease_mapping |> 
  mutate(icd10_code_clean = ifelse(icd10_code_clean == "G45-G46.8",
                                   "G45, G46",
                                   icd10_code_clean)) |>
  tidyr::separate_longer_delim(icd10_code_clean, delim = ", ") 

disease_mapping <-
  disease_mapping |> 
    mutate(token_kind = case_when(
      str_detect(icd10_code_clean, ICD10_CODE_RE)  ~ "code",
      str_detect(icd10_code_clean, ICD10_RANGE_RE) ~ "range",
      is.na(icd10_code_clean)                        ~ "missing",
      TRUE                              ~ "malformed"
    ))

disease_mapping |> 
  group_by(token_kind) |>
  summarise(n = n())

# one code range not properly expanded
disease_mapping |>
  filter(token_kind == "range") |>
  distinct(icd10_code_clean)

WHO_ICD10_CODE_RE <- "^[A-Z][0-9]{2}(\\.[0-9])?$"


# convert 3 dp codes to 1 dp - for icd10 who
disease_mapping |>
  filter(!str_detect(icd10_code_clean, WHO_ICD10_CODE_RE)) |>
  distinct(icd10_code_clean)  |>
  head()

ICD10_3dp_CODE_RE <- "^[A-Z][0-9]{2}\\.[0-9]{3}$"
ICD10_2dp_CODE_RE <- "^[A-Z][0-9]{2}\\.[0-9]{2}$"

disease_mapping |>
  filter(str_detect(icd10_code_clean, ICD10_3dp_CODE_RE) |
         str_detect(icd10_code_clean, ICD10_2dp_CODE_RE)
         ) |>
  distinct(collected_all_disease_terms, icd10_code_clean) |>
  head()

disease_mapping <- 
disease_mapping |>
  mutate(
    icd10_code_clean = ifelse(
      str_detect(icd10_code_clean, ICD10_3dp_CODE_RE) |
      str_detect(icd10_code_clean, ICD10_2dp_CODE_RE),
      stringr::str_replace(
        icd10_code_clean,
        pattern = "^([A-Z][0-9]{2}\\.[0-9]).*$",
        replacement = "\\1"
        ),
      icd10_code_clean
    )
  )

disease_mapping <- 
disease_mapping |>
  mutate(icd10_code_clean = normalize_icd10_codes(icd10_code_clean))

disease_mapping |>
    filter(!str_detect(icd10_code_clean, 
                       WHO_ICD10_CODE_RE)) |> 
  distinct(trait_clean, icd10_code_clean)

# remaining code conversion from ICDCM to ICD10 WHO
# F32.A -> F32.9
# G20.A1 -> G20
# J09.X -> J09, J10
# G40.A -> G40.3
# I1A.0 -> I10
# H91.8X9 ->  H91.8
# D3A.8 -> D36.7
disease_mapping <- 
disease_mapping |>
  mutate(icd10_code_clean = case_when(
    icd10_code_clean == "F32.A" ~ "F32.9",
    icd10_code_clean == "G20.A1" ~ "G20",
    icd10_code_clean == "J09.X" ~ "J09",
    icd10_code_clean == "G40.A" ~ "G40.3",
    icd10_code_clean == "I1A.0" ~ "I10",
    icd10_code_clean == "H91.8X9" ~ "H91.8",
    icd10_code_clean == "H54.7" ~ "H54.9",
    icd10_code_clean == "D3A.8" ~ "D36.7",
    icd10_code_clean == "B07.0" ~ "B07",
    icd10_code_clean == "B10.8" ~ "B33.8",
    icd10_code_clean == "B19.2" ~ "B17.1, B18.2", 
    icd10_code_clean == "C80.1" ~ "C80",
    icd10_code_clean %in% c("C83.3", "C83.4", "C83.6") ~ "C83",
    icd10_code_clean == "C94.5" ~ "C94.4",
    icd10_code_clean %in% c("C96.1", "C96.3") ~ "C96",
    icd10_code_clean == "R10.8" ~ "R10.4",
    icd10_code_clean == "R35.1" ~ "R35",
    icd10_code_clean == "R13.1" ~ "R13",
    icd10_code_clean == "R14.0" ~ "R14",
    icd10_code_clean == "K20.9" ~ "K20",
    icd10_code_clean ==  "I84.8" ~ "K64.5",
    icd10_code_clean %in% c("C84.2", "C84.3") ~ "C84",
    icd10_code_clean == "C88.1" ~ "C88.3",
    icd10_code_clean == "C93.2" ~ "C93",
    icd10_code_clean == "C91.2" ~ "C91",
    icd10_code_clean %in% c("C94.1", "C94.3") ~ "C94",
    icd10_code_clean %in% c("I82.4", "I82.5") ~ "I80.2",
    icd10_code_clean == "I82.6" ~ "I82.8",
    icd10_code_clean %in% c("I84.0", "I84.1", "I84.2", "I84.3",
                          "I84.4", "I84.5", "I84.6",
                          "I84.7", "I84.8", "I84.9") ~ "K64",
    icd10_code_clean == "R19.7" ~ "A09",
    icd10_code_clean == "O13.9" ~ "O13",
    icd10_code_clean == "O09" ~ "Z35",
    icd10_code_clean == "O09.9" ~ "Z35",
    icd10_code_clean == "M84.7" ~ "M84.4, M84.8",
    icd10_code_clean == "R80.9" ~ "R80",
    icd10_code_clean == "R56.9" ~ "R56.8",
    icd10_code_clean %in% c("I50.2", "I50.4") ~ "I50.0",
    icd10_code_clean %in% c("R53.1", "R53.8") ~ "R53",
    icd10_code_clean == "R39.9" ~ "R39.8",
    icd10_code_clean == "G98.8" ~ "G98",
    icd10_code_clean == "T48.9" ~ "T48",
    icd10_code_clean == "N40.0" ~ "N40",
    icd10_code_clean %in% c("O48.1", "O48.0") ~ "O48",
    icd10_code_clean == "N18.0" ~ "N18",
    icd10_code_clean %in% c("N52", "N52.9") ~ "N48.4",
    icd10_code_clean %in% c("K35.0", "K35.1") ~ "K35",
    icd10_code_clean %in% c("N44.0", "N44.1") ~ "N44",
    icd10_code_clean == "R07.8" ~ "R07.3",
    icd10_code_clean == "R07.9" ~ "R07.4",
    icd10_code_clean == "R10.9" ~ "R10.4",
    icd10_code_clean == "R15.9" ~ "R15",
    icd10_code_clean == "G56.4" ~ "G56",
    icd10_code_clean == "G90.3" ~ "G90",
    icd10_code_clean == "D75.2" ~ "D47.3", 
    icd10_code_clean == "D46.3" ~ "D46.2",
    icd10_code_clean == "F03.9" ~ "F03",
    icd10_code_clean %in% c("R50.0", "R50.1") ~ "R50",
    icd10_code_clean %in% c("R11.0", "R11.1") ~ "R11",
    icd10_code_clean == "W00.9" ~ "W00",   
    icd10_code_clean == "R31.9" ~ "R31",
    icd10_code_clean == "J30.9" ~ "J30.4",
    icd10_code_clean == "N46.1" ~ "N46",
    icd10_code_clean == "N18.6" ~ "N18",
    icd10_code_clean == "R35.0" ~ "R35",
    icd10_code_clean %in% c("W01.0", "W01.4", "W01.8", "W01.9") ~ "W01",
    icd10_code_clean %in% c("W10.0", "W10.9") ~ "W10",
    icd10_code_clean %in% c("W11.0", "W11.9") ~ "W11",
    icd10_code_clean %in% c("W18.0","W18.8", "W18.9") ~ "W18",
    icd10_code_clean %in% c("W19.0", "W19.2", "W19.4", "W19.9") ~ "W19",
    icd10_code_clean == "W44.9" ~ "W44",
    icd10_code_clean == "X50.9" ~ "X50",
    icd10_code_clean %in% c("X60.0", "X60.9") ~ "X60",
    icd10_code_clean %in% c("X61.0", "X61.9") ~ "X61",
    icd10_code_clean == "Z68.4" ~ "E66.0",

# no who equivalent:
# Y92, Y93        place / activity modifiers
# Z68             BMI observation
# Z66, Z67, Z69   DNR status, blood type, abuse counselling
# L49             exfoliation extent modifier
    
    TRUE ~ icd10_code_clean
  )
  )

disease_mapping <-
  disease_mapping |>
  tidyr::separate_longer_delim(icd10_code_clean, 
                               delim = ", ") 


disease_mapping <-
  disease_mapping |>
  mutate(icd10_who = case_when(icd10_code_clean %in% c("L49",
                                                       "Y92", "Y92.4",
                                                       "Y93", "Y93.9",
                                                       "Z66", "Z67", "Z68", "Z69") ~ NA_character_,
                               TRUE ~ icd10_code_clean)
         )
  

disease_mapping |>
    filter(!str_detect(icd10_code_clean, 
                       WHO_ICD10_CODE_RE)) |> 
  distinct(trait_clean, icd10_code_clean)

```


```{r validate_who_icd10_codes}

# for validating ICD-10 Codes
source(here::here("code/paper_to_icd10_mapping/check_icd10_who_fns.R"))

check_icd10_codes <-
validate_icd10(unique(disease_mapping$icd10_who), 
               read_icd10_meta()) 

# 31 invalid codes (including NA)
check_icd10_codes |>
  filter(!grepl("\\bok\\b", status, ignore.case = TRUE)) |>
  nrow()

head(check_icd10_codes |>
  filter(!grepl("\\bok\\b", status, ignore.case = TRUE)))

validated_codes <-
  check_icd10_codes |>
  filter(grepl("\\bok\\b", status, ignore.case = TRUE)) |>
  filter(!is.na(input)) |>
  pull(input)

# exclude ICD-10 Codes we could not validate with the 2019 ICD10 WHO: 
disease_mapping <-
  disease_mapping |>
  select(-icd10_code, -icd10_code_clean) |>
  mutate(icd10_who = ifelse(icd10_who %in% validated_codes, 
                            icd10_who, 
                            NA_character_)) |>
  rename(icd10_code = icd10_who)

```

## Save disease map

```{r save_disease_mapping}

disease_mapping = 
  disease_mapping |> 
  arrange(YEAR,
          PUBMED_ID,
          STUDY_ACCESSION,
          collected_all_disease_terms, 
          icd10_code)

disease_mapping = 
  disease_mapping |> 
  mutate(icd10_code = ifelse(icd10_code == "",
                             NA_character_,
                             icd10_code))

# make a column that displays the disease term that was used to infer the ICD-10 code
disease_mapping <-
disease_mapping |>
  mutate(icd10_term = case_when(
   icd10_source_col == "DISEASE/TRAIT" &
    grepl("Study", icd10_source) ~ "DISEASE/TRAIT",
   # icd10_source_col == "DISEASE/TRAIT" &  
   #  grepl("Study PheCode", icd10_source) & ~ phecode,
   #  icd10_source_col == "DISEASE/TRAIT" &  
   #  grepl("Study ICD10", icd10_source) & ~ icd10_code,
   # icd10_source_col == "DISEASE/TRAIT" ~ trait_clean,
   icd10_source_col == "MAPPED DISEASE TERM" ~ collected_all_disease_terms, 
   icd10_source_col == "Extracted disease from DISEASE/TRAIT" ~ trait_context_clean,
   TRUE ~ NA_character_
  )
  )

fwrite(disease_mapping,
       here::here("output/icd_map/gwas_disease_to_icd10_mapping.csv")
       )

```


        


