---
title: "Matching GWAS traits to GBD causes / diseases"
author: "Isobel Beasley"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Set up

```{r load_libraries, message=FALSE}

library(dplyr)
library(stringr)
library(purrr)
library(tidyr)
library(data.table)
library(ComplexHeatmap)
library(ggplot2)

```

## Load GWAS -> WHO ICD-10 2019 Mapping

```{r load_gwas_study_info}

disease_mapping <- data.table::fread(
here::here("output/icd_map/gwas_disease_to_icd10_mapping.csv")
)

```

## Load Lancet priority diseases

```{r load_lancet_priority_diseases}

gbd_2019 <- readxl::read_xlsx(here::here("data/icd/lancet_conditions_icd10.xlsx"))

gbd_2019 =
  gbd_2019 |>
  select(lancet_group,
         lancet_condition,
         cause = gbd_term,
         cause_category = ghe_cod_category,
         include,
         icd10_non_fatal = gbd_non_fatal_icd_10,
         icd10_cod = gbd_cod_icd_10,
         cause_hierarchy_level = gbd_level)

gbd_2019 =
  gbd_2019 |>
  mutate(include = as.logical(include))

# this ends up excluding 
# Cirrhosis and other chronic liver diseases due to hepatitis B
# Cirrhosis and other chronic liver diseases due to hepatitis C
# because these don't have specific ICD-10 Code Maps in GBD 
# this is why we use the category: Cirrhosis and other chronic liver diseases
gbd_2019 =
  gbd_2019 |>
  filter(!is.na(icd10_non_fatal) | !is.na(icd10_cod))

```

### Calculate overlap between ICD-10 code sets

I.e. between non-fatal and cause of death ICD-10 codes for the same cause.

```{r parse_icd10_code_ranges_fns}

# Convert a base-36 character to its numeric value
base36_value <- function(char) {
  # Base-36 characters
  base36_chars <- c(as.character(0:9), LETTERS)
  
  match(toupper(char), base36_chars) - 1L
}

parse_icd10_code <- function(code) {
  
  # Clean input
  raw_code <- str_trim(code)
  
  # Split ICD-10 code into components:
  #   letter + digit + alphanumeric + optional decimal + remaining characters
  parts <- str_match(
    raw_code,
    "^([A-Za-z])([0-9])([0-9A-Za-z])\\.?([0-9A-Za-z]*)$"
  )
  
  # Check: are the ICD-10 Codes formatted properly such that code an be parsed into its components?
  if (is.na(parts[1, 1])) {
    stop("Unparseable ICD10 code: '", raw_code, "'")
  }
  
  # get the separate components of the code
  # e.g. for A07.8
  
  # letter part of code, in e.g. is A
  letter <- toupper(parts[1, 2])
  
  # first and second digit of code: 
  # in e.g.is 0
  first_digit <- parts[1, 3]
  # in e.g. is 7
  second_digit <- parts[1, 4]
  
  # decimal part of code
  # in e.g. is 8
  decimal_part <- parts[1, 5]
  
  
  
  # Convert characters after the decimal to a base-36 fraction
  fraction <- 0
  
  if (nchar(decimal_part) > 0) {
    decimal_chars <- str_split(decimal_part, "")[[1]]
    
    fraction <- sum(
      base36_value(decimal_chars) /
        36^seq_along(decimal_chars)
    )
  }
  
  parsed_icd10_code <-
  list(
    letter = letter,
    value = base36_value(first_digit) * 36 +
            base36_value(second_digit) +
            fraction,
    label = raw_code,
    bare = nchar(decimal_part) == 0
  )
  
  return(parsed_icd10_code)
  
}

parse_range <- function(range_string) {
  
  # Split range into start and end codes
  range_parts <- str_trim(str_split(range_string, "-")[[1]])
  
  # get parsed start and end codes
  start_code <- parse_icd10_code(range_parts[1])
  end_code <- parse_icd10_code(range_parts[length(range_parts)])
  
  # Start and end codes must belong to the same ICD-10 category
  if (start_code$letter != end_code$letter) {
    stop("Range spans letters: '", range_string, "'")
  }
  
  # A bare 3-character END code covers the entire category
  end_value <- if (end_code$bare) {
    end_code$value + 1 - 1e-9
  } else {
    end_code$value
  }
  
  parsed_range <-
  tibble(
    letter = start_code$letter,
    start_value = start_code$value,
    end_value = end_value,
    start_label = start_code$label,
    end_label = end_code$label
  )
  
  return(parsed_range)
  
}

# to be applied rowwise to data frame
parse_rangelist <- function(range_list) {
  
  range_list <- unlist(str_split(range_list, ", "))
  
  parsed_rangelist <- purrr::map(range_list, parse_range) |>
                      purrr::list_rbind()
  
  return(parsed_rangelist)
}

# after parsing ranges, convert back to a simplified string representation
# e.g. A07.8-A07.9, A08.0-A08.1, A08.3-A08.5, A09.0-A09.9
simplify_ranges <- function(range_data) {
  
  if (nrow(range_data) == 0) {
    return(NA_character_)
  }
  
  range_labels <- if_else(
    range_data$start_label == range_data$end_label,
    range_data$start_label,
    paste0(
      range_data$start_label,
      "-",
      range_data$end_label
    )
  )
  
  paste(range_labels, collapse = ", ")
}

```


```{r overlap_gbd_icd10_ranges}

# get the overlap between two parsed ICD-10 range lists
# i.e. a & b need to be output of parse_rangelist
intersect_ranges <- function(parsed_rangelist_a, 
                             parsed_rangelist_b) {
  
  output <- list()
  
  for (i in seq_len(nrow(parsed_rangelist_a))) {
    for (j in seq_len(nrow(parsed_rangelist_b))) {
      
      # Ranges must belong to the same ICD-10 category
      if (parsed_rangelist_a$letter[i] != parsed_rangelist_b$letter[j]) {
        next
      }
      
      # The intersection starts at the later start point
      if (parsed_rangelist_a$start_value[i] >= parsed_rangelist_b$start_value[j]) {
        
        start_value <- parsed_rangelist_a$start_value[i]
        start_label <- parsed_rangelist_a$start_label[i]
        
      } else {
        
        start_value <- parsed_rangelist_b$start_value[j]
        start_label <- parsed_rangelist_b$start_label[j]
      }
      
      # The intersection ends at the earlier end point
      if (parsed_rangelist_a$end_value[i] <= parsed_rangelist_b$end_value[j]) {
        
        end_value <- parsed_rangelist_a$end_value[i]
        end_label <- parsed_rangelist_a$end_label[i]
        
      } else {
        end_value <- parsed_rangelist_b$end_value[j]
        end_label <- parsed_rangelist_b$end_label[j]
      }
      
      # Keep only valid intersections
      if (start_value <= end_value) {
        output[[length(output) + 1]] <- tibble(
          letter = parsed_rangelist_a$letter[i],
          start_value = start_value,
          end_value = end_value,
          start_label = start_label,
          end_label = end_label
        )
      }
    }
  }
  
  overlap <- bind_rows(output)
  overlap <- simplify_ranges(overlap)
  
  return(overlap)
}

compare_icd10_ranges <- function(icd10_range_a, 
                                 icd10_range_b) {
  
 parsed_range_a <- parse_rangelist(icd10_range_a)
 parsed_range_b <- parse_rangelist(icd10_range_b)
                                            
 overlap <- intersect_ranges(parsed_range_a, 
                             parsed_range_b
                             ) 

 return(overlap)

}

gbd_2019 <- 
gbd_2019 |>
  rowwise() |>
  mutate(overlap = compare_icd10_ranges(icd10_non_fatal, icd10_cod)) |>
  ungroup()

```

# Merge GWAS traits with GBD causes

```{r}

SRC <- c("overlap", 
         "icd10_non_fatal", 
         "icd10_cod")

# parse the ICD-10 ranges for each GBD cause into a data frame of intervals
cause_intervals <- gbd_2019 |>
  select(cause, 
         cause_category,
         lancet_condition,
         lancet_group,
         cause_hierarchy_level,
         include,
         all_of(SRC)) |>
  pivot_longer(all_of(SRC), 
               names_to = "cause_source", 
               values_to = "rangelist") |>
  filter(!is.na(rangelist)) |>
  mutate(parsed_rangelist = map(rangelist,
                                 ~ parse_rangelist(.x))) |>
  unnest(parsed_rangelist) |>
  filter(start_value <= end_value) |>                       # drop reversed ranges like K75.4-K75
  mutate(matched_range = if_else(start_label == end_label, 
                                 start_label, 
                                 paste0(start_label, "-", end_label))
         ) 

# parse all mapped ICD-10 codes for each GWAS study
gwas_mapped_codes <- disease_mapping |>
  distinct(icd10_code) |>
  filter(!is.na(icd10_code), 
         str_trim(icd10_code) != "") |>
  mutate(parsed_icd10code = map(icd10_code,
                                ~as.data.frame(parse_icd10_code(.x))
                                )
         ) |>
  unnest(parsed_icd10code)

code_to_cause <- gwas_mapped_codes  |>
  left_join(cause_intervals, 
            by = "letter", 
            relationship = "many-to-many") |>
  mutate(hit = !is.na(start_value) & 
         value <= end_value &  # check that gwas ICD-10 code value is less than the end value of the range
         start_value <= value) |> # check that the the gwas ICD-10 code value is greater than the start value of the range
  filter(hit == TRUE & !is.na(cause)) |>
  distinct(icd10_code, 
           cause, 
           cause_category,
           lancet_condition,
           lancet_group,
           cause_hierarchy_level,
           include,
           cause_source, 
           matched_range)

disease_mapping_annotated <- disease_mapping |>
  left_join(code_to_cause, 
            by = "icd10_code", 
            relationship = "many-to-many")


```


```{r number_of_papers_mapped_to_cause}

# number of distinct papers mapped to a cause (includes papers mapped to multiple causes,
# and papers mapped to communicable traits and other excluded diseases)
disease_mapping_annotated |>
  filter(!is.na(cause)) |>
  pull(PUBMED_ID) |>
  unique() |>
  length()

# spec_icd10_diabetes_ranges <-
#   disease_mapping_annotated |> 
#   filter(cause == "Type 1 diabetes mellitus"|
#          cause == "Type 2 diabetes mellitus") |> 
#   pull(matched_range) |> 
#   unique()
# 
# disease_mapping_annotated |>
#   filter(matched_range %in% spec_icd10_diabetes_ranges) |>
#   pull(cause) |>
#   unique()

# Diabetes mellitus is a broader category that includes both Type 1 and Type 2 diabetes mellitus
# So don't duplicate where the same study is mapped to Type 1 or Type 2 diabetes and also mapped to Diabetes mellitus
# rows_to_drop <-
#   disease_mapping_annotated |>
#   filter(matched_range %in% spec_icd10_diabetes_ranges)  |>
#   filter(cause == "Diabetes mellitus")
# 
# 
# disease_mapping_annotated <-
#   rows_delete(disease_mapping_annotated, 
#               rows_to_drop,
#               by = names(rows_to_drop)
#               )

# same but for Chronic kidney disease
# Chronic kidney disease due to diabetes mellitus type 1 and Chronic kidney disease due to diabetes mellitus type 2 are also broader categories that include Type 1 and Type 2 diabetes mellitus
# spec_icd10_ckd_ranges <-
#   disease_mapping_annotated |> 
#   filter(cause == "Chronic kidney disease due to diabetes mellitus type 1"|
#          cause == "Chronic kidney disease due to diabetes mellitus type 2") |> 
#   pull(matched_range) |> 
#   unique()
# 
# disease_mapping_annotated |>
#   filter(matched_range %in% spec_icd10_ckd_ranges) |>
#   pull(cause) |>
#   unique()
# 
# rows_to_drop <-
#   disease_mapping_annotated |>
#   filter(matched_range %in% spec_icd10_ckd_ranges)  |>
#   filter(cause == "Chronic kidney disease")

# disease_mapping_annotated <- 
#   disease_mapping_annotated |>
#   rows_delete(rows_to_drop,
#               by = names(rows_to_drop)
#               )

disease_mapping_with_cause <-
disease_mapping_annotated |>
  group_by(PUBMED_ID,
           YEAR,
           `DISEASE/TRAIT`, 
           collected_all_disease_terms,
           icd10_term,
           cause,
           cause_category,
           lancet_condition,
           lancet_group,
           cause_hierarchy_level,
           include) |>
  summarise(icd10_code = stringr::str_flatten_comma(unique(sort(icd10_code)),
                                                    na.rm = TRUE
                                                    ),
            cause_source = stringr::str_flatten_comma(unique(sort(cause_source)),
                                                      na.rm = TRUE
                                                      )
  ) |>
  mutate(cause_source = case_when(grepl("overlap", cause_source) ~ "overlap",
                                  grepl("icd10_non_fatal", cause_source) ~ "icd10_non_fatal",
                                  grepl("icd10_cod", cause_source) ~ "icd10_cod", 
                                  TRUE ~ NA_character_
                                    ) ) |>
  ungroup() |>
  distinct()

# number of papers only mapped to a cause by overlap of non-fatal and cause of death ICD-10 codes
disease_mapping_with_cause |>
  filter(!is.na(cause)) |>
  filter(cause_source == "overlap") |>
  pull(PUBMED_ID) |>
  unique() |>
  length()

# number of papers only mapped to a cause by overlap of non-fatal ICD-10 codes
disease_mapping_with_cause |>
  filter(!is.na(cause)) |>
  filter(cause_source == "icd10_non_fatal") |>
  pull(PUBMED_ID) |>
  unique() |>
  length()

# number of papers only mapped to a cause by overlap of cause of death ICD-10 codes
disease_mapping_with_cause |>
  filter(!is.na(cause)) |>
  filter(cause_source == "icd10_cod") |>
  pull(PUBMED_ID) |>
  unique() |>
  length()
                           
```

## Should we keep 'Cause of Death Specific' codes: 

Let's drop cause of death specific cause codes, as these relate less to the diseases studied. 

```{r compare_cod_mapping}
# per paper mapping
per_paper_disease_map <-
disease_mapping_with_cause |> 
  filter(include == T) |> 
  #filter(!icd10_code %in% c("B18", "B18.1", "B18.2", "D26", "D10", "D11")) |> 
  distinct(PUBMED_ID, cause, cause_source) |> 
  group_by(PUBMED_ID, cause) |> 
  summarise(cause_source = stringr::str_flatten_comma(unique(cause_source, na.rm = T))) |> 
  ungroup() 

per_paper_disease_map <-
  per_paper_disease_map |>
  mutate(cause_source = 
           ifelse(grepl("overlap", cause_source), 
                  "overlap", cause_source))

per_paper_disease_map |> distinct(cause_source)

per_paper_disease_map  |>
  filter(cause_source == "icd10_cod") |>
  group_by(cause) |>
  summarise(n_papers = length(unique(PUBMED_ID))) |>
  arrange(desc(n_papers))

# chronic viral hepatitis codes explain 
# these Cirrhosis and other chronic liver diseases causes (mapped by Cause of death codes only)

per_paper_disease_map <-
disease_mapping_with_cause |> 
  filter(include == T) |> 
  # exclude chronic viral hepatitis codes (B18, B18.1, B18.2) and benign neoplasms (D26, D10, D11) codes
  filter(!icd10_code %in% c("B18", "B18.1", "B18.2", 
                            "D26", "D10", "D10.6", "D11", "D38", "D10, D11",
                            "N08", "N08.3",
                            "I68.2")) |> 
  distinct(PUBMED_ID, cause, cause_source) |> 
  group_by(PUBMED_ID, cause) |> 
  summarise(cause_source = stringr::str_flatten_comma(unique(cause_source, na.rm = T))) |> 
  ungroup() 


per_paper_disease_map <-
  per_paper_disease_map |>
  mutate(cause_source = 
           ifelse(grepl("overlap", cause_source), 
                  "overlap", cause_source))

per_paper_disease_map  |>
  filter(cause_source == "icd10_cod") |>
  group_by(cause) |>
  summarise(n_papers = length(unique(PUBMED_ID))) |>
  arrange(desc(n_papers))


```


```{r drop_cod_only_mappings}

code_to_cause <-
  code_to_cause |> 
  filter(cause_source == "overlap" | cause_source == "icd10_non_fatal") |>
  mutate(cause_source = "icd10_non_fatal") |>
  distinct()

disease_mapping_annotated <- disease_mapping |>
  left_join(code_to_cause, 
            by = "icd10_code", 
            relationship = "many-to-many")



disease_mapping_with_cause <-
disease_mapping_annotated |>
  group_by(PUBMED_ID,
           YEAR,
           `DISEASE/TRAIT`, 
           collected_all_disease_terms,
           icd10_term,
           cause,
           cause_category,
           lancet_condition,
           lancet_group,
           cause_hierarchy_level,
           include) |>
  summarise(icd10_code = stringr::str_flatten_comma(unique(sort(icd10_code)),
                                                    na.rm = TRUE
                                                    ),
            cause_source = stringr::str_flatten_comma(unique(sort(cause_source)),
                                                      na.rm = TRUE
                                                      )
  ) |>
  ungroup() |>
  distinct()


print("Number of unique papers mapped to non-communicable diseases:")
# filter out infectious diseases
disease_mapping_with_cause |>
    dplyr::filter(include == TRUE) |>
    filter(!is.na(cause)) |> 
  pull(PUBMED_ID) |> 
  unique() |> 
  length()


print("List of non-communicable diseases mapped to GWAS Catalog studies:")
disease_mapping_with_cause |>
  dplyr::filter(include == TRUE) |>
  pull(cause) |>
  unique() |> 
  sort()
                 
```



# Plot: number of papers per cause: 

```{r}

pdf(here::here("figures/number_of_gwas_catalog_papers_per_cause.pdf"),
    width = 17, 
    height = 8
)

disease_mapping_with_cause |>
  dplyr::filter(include == TRUE) |>
  # group into broader categories 
  mutate(cause = case_when(
    grepl("diabetes", cause, ignore.case = TRUE) ~ "Diabetes",
    grepl("chronic kidney disease", cause, ignore.case = TRUE) ~ "Chronic kidney disease",
    grepl("hemorr", cause, ignore.case = TRUE) ~ "Hemorrhagic stroke",
    grepl("lip|pharynx", cause, ignore.case = TRUE) ~ "Mouth and throat cancer",
    grepl("lung cancer", cause, ignore.case = TRUE) ~ "Lung Cancer",
    grepl("cirrhosis", cause, ignore.case = TRUE) ~ "Chronic Liver Disease",
    TRUE ~ cause
  )) |>
  mutate(cause = 
           stringr::str_to_title(
           stringr::str_replace_all(cause, 
                                          pattern = "and", 
                                          replacement = "&")
           )
           ) |>
  distinct(PUBMED_ID, cause) |>
  group_by(cause) |>
  summarise(n_papers = n()) |>
  ggplot(aes(x= reorder(cause, n_papers), y=n_papers)) +
  geom_col() +
  coord_flip() +
  labs(x = "Cause", 
       y = "Number of Papers") + 
  theme_bw() +
      theme(
    axis.text.x  = element_text(size = 25),   # 8pt for axis text
    axis.text.y  = element_text(size = 25),   # 8pt for axis text
    axis.title.x = element_text(size = 25),  # 10pt for labels
    axis.title.y = element_blank(),
    #title        = element_text(size = 8),  # 10pt for labels
    strip.text   = element_text(size = 25),  # 10pt for labels
    # legend.title = element_text(size = 18),  # 10pt for labels
    # legend.text  = element_text(size = 18),   # 8pt for legend text
    # legend.position = "right",           # Bottom left corner
    # legend.justification = "left",
    # #legend.justification = c(0, 0),      # Anchor at bottom left
    # #legend.background = element_rect(fill = "white", color = "grey80"),
    # legend.margin = margin(l = -0.25, unit = "npc"),
    #legend.margin = margin(5, 5, 5, 5),
    # legend.position = "bottom"
  )

dev.off()


```


# Check overlap (upset plot) of studies mapped to different causes

## Between type 1 diabetes, type 2 diabetes, and diabetes mellitus

```{r check_overlap_diabetes}

diabetes_studies = disease_mapping_with_cause  |>
  filter(cause %in% c("Type 1 diabetes mellitus", 
                      "Type 2 diabetes mellitus", 
                      "Diabetes mellitus",
                      "Chronic kidney disease due to diabetes mellitus type 2",
                      "Chronic kidney disease due to diabetes mellitus type 1",
                      "Chronic kidney disease")) |>
  select(PUBMED_ID, cause) |>
  distinct()

# create a binary matrix indicating which studies are mapped to which causes
diabetes_studies_wide = diabetes_studies |>
  tidyr::pivot_wider(names_from = cause, values_from = cause, values_fill = NA) |>
  select(-PUBMED_ID) |>
  mutate_all(~ ifelse(is.na(.), 0, 1)) |>
  as.data.frame()

upset_matrix = make_comb_mat(diabetes_studies_wide)

# create an upset plot
UpSet(upset_matrix,
      comb_order = order(comb_size(upset_matrix), decreasing = TRUE)
      #nsets = 3, 
      #order.by = "freq", 
      # sets.bar.color = "#56B4E9", 
      # main.bar.color = "#0072B2", 
      # matrix.color = "#D55E00", 
      #text.scale = c(1.5, 1.5, 1.5, 1.5, 1.5, 1.5)
      )
  
```

## Between different types of cancer

```{r check_overlap_cancer}

cancer_studies = disease_mapping_with_cause |>
  filter(grepl("cancer|neoplasm", cause, ignore.case = TRUE)) |>
  select(PUBMED_ID, cause) |>
  distinct()

cancer_studies_wide = cancer_studies |>
  tidyr::pivot_wider(names_from = cause, values_from = cause, values_fill = NA) |>
  select(-PUBMED_ID) |>
  mutate_all(~ ifelse(is.na(.), 0, 1)) |>
  as.data.frame()

upset_matrix_cancer = make_comb_mat(cancer_studies_wide)

UpSet(upset_matrix_cancer,
      comb_order = order(comb_size(upset_matrix_cancer), decreasing = TRUE)
      #nsets = 3, 
      #order.by = "freq", 
      # sets.bar.color = "#56B4E9", 
      # main.bar.color = "#0072B2", 
      # matrix.color = "#D55E00", 
      #text.scale = c(1.5, 1.5, 1.5, 1.5, 1.5, 1.5)
      )

```

## Between different cardiovascular diseases

```{r check_overlap_cardiovascular}

cardiovascular_studies = disease_mapping_with_cause |>
  filter(grepl("heart|stroke|hemorrhage", cause, ignore.case = TRUE)) |>
  select(PUBMED_ID, cause) |>
  distinct()

cardiovascular_studies_wide = cardiovascular_studies |>
  tidyr::pivot_wider(names_from = cause, values_from = cause, values_fill = NA) |>
  select(-PUBMED_ID) |>
  mutate_all(~ ifelse(is.na(.), 0, 1)) |>
  as.data.frame()

upset_matrix_cardiovascular = make_comb_mat(cardiovascular_studies_wide)

UpSet(upset_matrix_cardiovascular,
      comb_order = order(comb_size(upset_matrix_cardiovascular), 
                         decreasing = TRUE)
      #nsets = 3, 
      #order.by = "freq", 
      # sets.bar.color = "#56B4E9", 
      # main.bar.color = "#0072B2", 
      # matrix.color = "#D55E00", 
      #text.scale = c(1.5, 1.5, 1.5, 1.5, 1.5, 1.5)
      )

```

## Across liver diseases

Cirrhosis and other chronic liver diseases, Liver cancer due to hepatitis C, Liver cancer due to hepatitis B, Diabetes 

```{r check_overlap_liver}

liver_studies = disease_mapping_with_cause |>
  filter(grepl("liver|diabetes", cause, ignore.case = TRUE)) |>
  filter(!grepl("Type 2|Type 1", cause, ignore.case = TRUE)) |>
  select(PUBMED_ID, cause) |>
  distinct()


liver_studies_wide = liver_studies |>
  tidyr::pivot_wider(names_from = cause, values_from = cause, values_fill = NA) |>
  select(-PUBMED_ID) |>
  mutate_all(~ ifelse(is.na(.), 0, 1)) |>
  as.data.frame()

upset_matrix_liver = make_comb_mat(liver_studies_wide)

UpSet(upset_matrix_liver,
      comb_order = order(comb_size(upset_matrix_liver), decreasing = TRUE)
      #nsets = 3, 
      #order.by = "freq", 
      # sets.bar.color = "#56B4E9", 
      # main.bar.color = "#0072B2", 
      # matrix.color = "#D55E00", 
      #text.scale = c(1.5, 1.5, 1.5, 1.5, 1.5, 1.5)
      )

```


## Across all relevant non-communicable diseases

```{r check_overlap_all_ncd}

non_communicable_studies =
disease_mapping_with_cause |>
  filter(include == T) |>
  filter(!is.na(cause)) |>
  select(PUBMED_ID, cause) |>
  distinct() 
  
non_communicable_studies_wide = non_communicable_studies |>
  tidyr::pivot_wider(names_from = cause, values_from = cause, values_fill = NA) |>
  select(-PUBMED_ID) |>
  mutate_all(~ ifelse(is.na(.), 0, 1)) |>
  as.data.frame()

upset_matrix_non_communicable = make_comb_mat(non_communicable_studies_wide)

UpSet(upset_matrix_non_communicable,
      comb_order = order(comb_size(upset_matrix_non_communicable), decreasing = TRUE)
      #nsets = 3, 
      #order.by = "freq", 
      # sets.bar.color = "#56B4E9", 
      # main.bar.color = "#0072B2", 
      # matrix.color = "#D55E00", 
      #text.scale = c(1.5, 1.5, 1.5, 1.5, 1.5, 1.5)
      )


```

## Between different respiratory diseases

Chronic obstructive pulmonary disease, "Tracheal, bronchus, and lung cancer

```{r check_overlap_respiratory}

respiratory_studies = disease_mapping_with_cause |>
  filter(grepl("chronic obstructive pulmonary disease|tracheal|bronchus|lung cancer", cause, ignore.case = TRUE)) |>
  select(PUBMED_ID, cause) |>
  distinct()

respiratory_studies_wide = respiratory_studies |>
  tidyr::pivot_wider(names_from = cause, values_from = cause, values_fill = NA) |>
  select(-PUBMED_ID) |>
  mutate_all(~ ifelse(is.na(.), 0, 1)) |>
  as.data.frame()


upset_matrix_respiratory = make_comb_mat(respiratory_studies_wide)

UpSet(upset_matrix_respiratory,
      comb_order = order(comb_size(upset_matrix_respiratory), decreasing = TRUE)
      #nsets = 3, 
      #order.by = "freq", 
      # sets.bar.color = "#56B4E9", 
      # main.bar.color = "#0072B2", 
      # matrix.color = "#D55E00", 
      #text.scale = c(1.5, 1.5, 1.5, 1.5, 1.5, 1.5)
      )

```

## Overlap across Lancet conditions

```{r check_overlap_lancet_conditions}

overlap_lancet_conditions =
disease_mapping_with_cause |>
  filter(!is.na(lancet_condition)) |>
  group_by(lancet_condition) |>
  summarise(PUBMED_ID = list(unique(PUBMED_ID))) |>
  tibble::deframe()


upset_matrix_lancet_conditions = make_comb_mat(overlap_lancet_conditions)

UpSet(upset_matrix_lancet_conditions,
      comb_order = order(comb_size(upset_matrix_lancet_conditions), decreasing = TRUE)
      #nsets = 3, 
      #order.by = "freq", 
      # sets.bar.color = "#56B4E9", 
      # main.bar.color = "#0072B2", 
      # matrix.color = "#D55E00", 
      #text.scale = c(1.5, 1.5, 1.5, 1.5, 1.5, 1.5)
      )

print("Number of studies mapped to one Lancet condition:")
comb_size(upset_matrix_lancet_conditions, degree = 1) |> sum()

print("All non-communicable disease studies mapped to a Lancet condition:")
overlap_lancet_conditions |> purrr::list_c() |> unique() |> length()

```

### Just for the relevant diseases 

```{r}

overlap_lancet_conditions =
disease_mapping_with_cause |>
  filter(include == TRUE) |>
  filter(!is.na(lancet_condition)) |>
  mutate(general_group = case_when(
    lancet_condition %in% c("Atherosclerotic cardiovascular diseases",
                          "Haemorrhagic stroke") ~ "Cardiovascular diseases",  
    TRUE ~ lancet_condition
  )) |>
  group_by(general_group) |>
  summarise(PUBMED_ID = list(unique(PUBMED_ID))) |>
  tibble::deframe()


upset_matrix_lancet_conditions = make_comb_mat(overlap_lancet_conditions)

UpSet(upset_matrix_lancet_conditions,
      comb_order = order(comb_size(upset_matrix_lancet_conditions), decreasing = TRUE)
      #nsets = 3, 
      #order.by = "freq", 
      # sets.bar.color = "#56B4E9", 
      # main.bar.color = "#0072B2", 
      # matrix.color = "#D55E00", 
      #text.scale = c(1.5, 1.5, 1.5, 1.5, 1.5, 1.5)
      )

# disease_mapping_with_cause |>
#   filter(include == TRUE) |>
#   filter(!is.na(lancet_condition)) |>
#   mutate(general_group = case_when(
#     lancet_condition %in% c("Atherosclerotic cardiovascular diseases",
#                           "Haemorrhagic stroke") ~ "Cardiovascular diseases",  
#     TRUE ~ lancet_condition
#   )) |>
#   group_by(PUBMED_ID) |>
#   filter(any(general_group == "Cardiovascular diseases") &
#            any(general_group == "Diabetes"))



```

# Check recall i.e. that I have identified most/all studies of the priority diseases

```{r}
# let's check against original gwas metadata
orig_gwas_meta <- fread(here::here("data/gwas_catalog/gwas-catalog-v1.0.3.1-studies-r2025-07-21.tsv"),
                         quote = "")

orig_gwas_meta = 
  orig_gwas_meta|>
  rename_all(~gsub(" ", "_", .x))

```

## Diabetes 

```{r recall_diabetes}

no_capture <- paste0("(?<!no )",
                "(?<!non )",
                "(?<!non-)",
                "(?<!no type 2 )",
                "(?<!no type 1 )"
                )

diabetes_grep = paste0(no_capture, "(?:diabetes|diabetic)")

diabetes_pmids <- disease_mapping_with_cause  |>
  group_by(PUBMED_ID) |>
  filter(any(grepl("diabetes", 
                   cause, 
                   ignore.case = TRUE))) |>
  pull(PUBMED_ID) |>
  unique()

# any missed diabetes papers?
orig_gwas_meta |>
  filter(grepl(diabetes_grep, 
               `DISEASE/TRAIT`, 
               ignore.case = TRUE, 
               perl = TRUE)|
           grepl(diabetes_grep,
                MAPPED_TRAIT,
                ignore.case = TRUE,
                perl = TRUE
                )|
           grepl(diabetes_grep,
                MAPPED_BACKGROUND_TRAIT,
                ignore.case = TRUE,
                perl = TRUE
                )
         ) |>
  filter(!PUBMED_ID %in% diabetes_pmids) |>
  distinct(PUBMED_ID, 
           `DISEASE/TRAIT`,
           MAPPED_TRAIT,
           MAPPED_BACKGROUND_TRAIT) 

# no only gestational diabetes studies
# are not captured - which is what we want  


```

## Liver cancer, and other chronic liver diseases (e.g. cirrhosis) 

Check recall for these together as they are often comorbid and may be captured by similar terms

```{r recall_liver_cancer}

liver_pmids <- disease_mapping_with_cause  |>
  group_by(PUBMED_ID) |>
  filter(any(grepl("liver", 
                   cause, 
                   ignore.case = TRUE))) |>
  pull(PUBMED_ID) |>
  unique()

liver_cancer_grep = paste0(no_capture, "(?:\\bliver\\b(?! (enzyme|level|measurement|volume))|cirrhosis|hepatocellular carcinoma|hepatoma|hepatic carcinoma)")

orig_gwas_meta |>
  filter(grepl(liver_cancer_grep, 
               `DISEASE/TRAIT`, 
               ignore.case = TRUE, 
               perl = TRUE)|
           grepl(liver_cancer_grep,
                MAPPED_TRAIT,
                ignore.case = TRUE,
                perl = TRUE
                )|
           grepl(liver_cancer_grep,
                MAPPED_BACKGROUND_TRAIT,
                ignore.case = TRUE,
                perl = TRUE
                )
         ) |>
  filter(!PUBMED_ID %in% liver_pmids) |>
  distinct(PUBMED_ID, 
           `DISEASE/TRAIT`,
           MAPPED_TRAIT,
           MAPPED_BACKGROUND_TRAIT) 

```

## Stroke

```{r recall_stroke}

stroke_pmids <-
disease_mapping_with_cause  |>
  group_by(PUBMED_ID) |>
  filter(any(grepl("stroke|hemorrhage", 
                   cause, 
                   ignore.case = TRUE))) |>
  pull(PUBMED_ID) |>
  unique()  

stroke_grep = paste0(no_capture, 
                     "(?:stroke(?! volume)|cerebrovascular|cerebral(?! (spinal fluid|cortex volume))|(?<!partum )hemorrhage)")

orig_gwas_meta |>
  filter(grepl(stroke_grep, 
               `DISEASE/TRAIT`, 
               ignore.case = TRUE, 
               perl = TRUE)|
           grepl(stroke_grep,
                MAPPED_TRAIT,
                ignore.case = TRUE,
                perl = TRUE
                )|
           grepl(stroke_grep,
                MAPPED_BACKGROUND_TRAIT,
                ignore.case = TRUE,
                perl = TRUE
                )
         ) |>
  filter(!PUBMED_ID %in% stroke_pmids) |>
  distinct(PUBMED_ID, 
           `DISEASE/TRAIT`,
           MAPPED_TRAIT,
           MAPPED_BACKGROUND_TRAIT)

```

## Stomach cancer

```{r recall_stomach_cancer}

stomach_cancer_pmids <-
disease_mapping_with_cause  |>
  group_by(PUBMED_ID) |>
  filter(any(grepl("stomach|gastric", 
                   cause, 
                   ignore.case = TRUE))) |>
  pull(PUBMED_ID) |>
  unique()  

stomach_cancer_grep = paste0(no_capture, 
                     "(?:stomach|gastric(?! (inhibitory polypeptide|bypass|triacylglycerol lipase|intrinsic factor|reflux)))")

orig_gwas_meta |>
  filter(grepl(stomach_cancer_grep, 
               `DISEASE/TRAIT`, 
               ignore.case = TRUE, 
               perl = TRUE)|
           grepl(stomach_cancer_grep,
                MAPPED_TRAIT,
                ignore.case = TRUE,
                perl = TRUE
                )|
           grepl(stomach_cancer_grep,
                MAPPED_BACKGROUND_TRAIT,
                ignore.case = TRUE,
                perl = TRUE
                )
         ) |>
  filter(!PUBMED_ID %in% stomach_cancer_pmids) |>
  distinct(PUBMED_ID, 
           `DISEASE/TRAIT`,
           MAPPED_TRAIT,
           MAPPED_BACKGROUND_TRAIT)

```

## Cervical cancer

```{r recall_cervical_cancer}

cervical_cancer_pmids <- 
disease_mapping_with_cause  |>
  group_by(PUBMED_ID) |>
  filter(any(grepl("cervical|cervix", 
                   cause, 
                   ignore.case = TRUE))) |>
  pull(PUBMED_ID) |>
  unique()  


cervical_cancer_grep = paste0(no_capture, 
                     "(?:cervical(?! (artery|spine))|cervix)")

orig_gwas_meta |>
  filter(grepl(cervical_cancer_grep, 
               `DISEASE/TRAIT`, 
               ignore.case = TRUE, 
               perl = TRUE)|
           grepl(cervical_cancer_grep,
                MAPPED_TRAIT,
                ignore.case = TRUE,
                perl = TRUE
                )|
           grepl(cervical_cancer_grep,
                MAPPED_BACKGROUND_TRAIT,
                ignore.case = TRUE,
                perl = TRUE
                )
         ) |>
  filter(!PUBMED_ID %in% cervical_cancer_pmids) |>
  distinct(PUBMED_ID, 
           `DISEASE/TRAIT`,
           MAPPED_TRAIT,
           MAPPED_BACKGROUND_TRAIT)

disease_mapping_with_cause |>
  filter(grepl(cervical_cancer_grep, 
               `DISEASE/TRAIT`, 
               ignore.case = TRUE, 
               perl = TRUE)|
           grepl(cervical_cancer_grep,
                collected_all_disease_terms,
                ignore.case = TRUE,
                perl = TRUE
                )
         ) |>
  group_by(PUBMED_ID) |>
  filter(!any(grepl("cervical", 
                   cause, 
                   ignore.case = TRUE))) |>
  distinct(PUBMED_ID, 
           `DISEASE/TRAIT`,
           collected_all_disease_terms,
           cause) 

# human papilloma virus (HPV) is a major cause of cervical cancer ...
# so do we include studies of HPV in our mapping of cervical cancer?
hpv_grep <- paste0(no_capture, 
                     "(?:hpv|human papilloma virus|pap smear)")


```

## Lip and oral cavity cancer

```{r recall_lip_oral_cavity_cancer}

lip_oral_cavity_cancer_pmids <- 
disease_mapping_with_cause  |>
  group_by(PUBMED_ID) |>
  filter(any(grepl("lip|oral cavity|mouth|tongue|pharynx|larynx", 
                   cause, 
                   ignore.case = TRUE))) |>
  pull(PUBMED_ID) |>
  unique()  

lip_oral_cavity_cancer_grep = paste0(no_capture, 
                     "(?:(?<!cleft )lip\\b(?! morphology)|oral cavity|mouth|tongue|pharynx|larynx|pharyngeal|laryngeal)")

orig_gwas_meta |>
  filter(grepl(lip_oral_cavity_cancer_grep, 
               `DISEASE/TRAIT`, 
               ignore.case = TRUE, 
               perl = TRUE)|
           grepl(lip_oral_cavity_cancer_grep,
                MAPPED_TRAIT,
                ignore.case = TRUE,
                perl = TRUE
                )|
           grepl(lip_oral_cavity_cancer_grep,
                MAPPED_BACKGROUND_TRAIT,
                ignore.case = TRUE,
                perl = TRUE
                )
         ) |>
  filter(!PUBMED_ID %in% lip_oral_cavity_cancer_pmids) |>
  distinct(PUBMED_ID, 
           `DISEASE/TRAIT`,
           MAPPED_TRAIT,
           MAPPED_BACKGROUND_TRAIT)

disease_mapping_with_cause |>
  filter(grepl(lip_oral_cavity_cancer_grep, 
               `DISEASE/TRAIT`, 
               ignore.case = TRUE, 
               perl = TRUE)|
           grepl(lip_oral_cavity_cancer_grep,
                collected_all_disease_terms,
                ignore.case = TRUE,
                perl = TRUE
                )
         ) |>
  filter(!grepl("cleft", `DISEASE/TRAIT`, ignore.case = TRUE)) |>
  group_by(PUBMED_ID) |>
  filter(!any(grepl("lip|nasopharynx", 
                   cause, 
                   ignore.case = TRUE))) |>
  distinct(PUBMED_ID, 
           `DISEASE/TRAIT`,
           collected_all_disease_terms,
           cause) 

```

## Nasopharynx cancer

## Heart disease

```{r}

heart_disease_grep = paste0(no_capture, 
                     "(?:heart disease|cardiovascular|coronary|myocardial|angina|heart failure)")

disease_mapping_with_cause |>
  filter(grepl(heart_disease_grep, 
               `DISEASE/TRAIT`, 
               ignore.case = TRUE, 
               perl = TRUE)|
           grepl(heart_disease_grep,
                collected_all_disease_terms,
                ignore.case = TRUE,
                perl = TRUE
                )
         )|>
  group_by(PUBMED_ID) |>
  filter(!any(grepl("cardiovascular", 
                   lancet_condition, 
                   ignore.case = TRUE))
         & !any(grepl("heart disease", 
                   cause, 
                   ignore.case = TRUE))) |>
  distinct(PUBMED_ID, 
           `DISEASE/TRAIT`,
           collected_all_disease_terms,
           cause) 

```

# Check precision i.e. that the studies I have mapped to the priority diseases are actually about those diseases

## Cervical cancer 

What terms caused the mapping to cervical cancer? Are they actually about cervical cancer?

```{r precision_cervical_cancer}

disease_mapping_with_cause |>
  filter(cause == "Cervical cancer") |>
  distinct(PUBMED_ID, icd10_term) |>
  group_by(icd10_term) |>
  summarise(n_papers = length(unique(PUBMED_ID))) |>
  arrange(desc(n_papers))

```

## Chronic kidney disease 

```{r precision_chronic_kidney_disease}

disease_mapping_with_cause |>
  filter(cause == "Chronic kidney disease") |>
  distinct(PUBMED_ID, icd10_term) |>
  group_by(icd10_term) |>
  summarise(n_papers = length(unique(PUBMED_ID))) |>
  arrange(desc(n_papers))

```

### Chronic kidney disease due to diabetes mellitus type 1

```{r precision_chronic_kidney_disease_type_1}

disease_mapping_with_cause |>
  filter(cause == "Chronic kidney disease due to diabetes mellitus type 1") |>
  distinct(PUBMED_ID, icd10_term) |>
  group_by(icd10_term) |>
  summarise(n_papers = length(unique(PUBMED_ID))) |>
  arrange(desc(n_papers))

```

### Chronic kidney disease due to diabetes mellitus type 2

```{r precision_chronic_kidney_disease_type_2}

disease_mapping_with_cause |>
  filter(cause == "Chronic kidney disease due to diabetes mellitus type 2") |>
  distinct(PUBMED_ID, icd10_term) |>
  group_by(icd10_term) |>
  summarise(n_papers = length(unique(PUBMED_ID))) |>
  arrange(desc(n_papers))

```

## Chronic obstructive pulmonary disease

```{r precision_copd}

disease_mapping_with_cause |>
  filter(cause == "Chronic obstructive pulmonary disease") |>
  distinct(PUBMED_ID, icd10_term) |>
  group_by(icd10_term) |>
  summarise(n_papers = length(unique(PUBMED_ID))) |>
  arrange(desc(n_papers))

```

## Cirrhosis and other chronic liver diseases

```{r precision_cirrhosis}

disease_mapping_with_cause |> 
filter(cause == "Cirrhosis and other chronic liver diseases") |>
  distinct(PUBMED_ID, icd10_term) |>
  group_by(icd10_term) |>
  summarise(n_papers = length(unique(PUBMED_ID))) |>
  arrange(desc(n_papers))

```

## Diabetes mellitus

```{r precision_diabetes}

disease_mapping_with_cause |> 
filter(cause == "Diabetes mellitus") |>
  distinct(PUBMED_ID, icd10_term) |>
  group_by(icd10_term) |>
  summarise(n_papers = length(unique(PUBMED_ID))) |>
  arrange(desc(n_papers))

# severe protein-calorie malnutrition 

```

### Type 1 diabetes mellitus

```{r precision_type_1_diabetes}

disease_mapping_with_cause |>
filter(cause == "Type 1 diabetes mellitus") |>
  distinct(PUBMED_ID, icd10_term) |>
  group_by(icd10_term) |>
  summarise(n_papers = length(unique(PUBMED_ID))) |>
  arrange(desc(n_papers))


```

### Type 2 diabetes mellitus

```{r precision_type_2_diabetes}

disease_mapping_with_cause |>
filter(cause == "Type 2 diabetes mellitus") |>
  distinct(PUBMED_ID, icd10_term) |>
  group_by(icd10_term) |>
  summarise(n_papers = length(unique(PUBMED_ID))) |>
  arrange(desc(n_papers))


```

## Haemorrhagic stroke

```{r precision_haemorrhagic_stroke}

disease_mapping_with_cause |> 
filter(cause == "Intracerebral hemorrhage" | cause == "Subarachnoid hemorrhage") |>
  distinct(PUBMED_ID, icd10_term) |>
  group_by(icd10_term) |>
  summarise(n_papers = length(unique(PUBMED_ID))) |>
  arrange(desc(n_papers))

```

## Ischemic heart disease

```{r precision_ischemic_heart_disease}

disease_mapping_with_cause |>
  filter(cause == "Ischemic heart disease") |>
  distinct(PUBMED_ID, icd10_term) |>
  group_by(icd10_term) |>
  summarise(n_papers = length(unique(PUBMED_ID))) |>
  arrange(desc(n_papers))

```

## Ischemic stroke

```{r precision_ischemic_stroke}

disease_mapping_with_cause |>
  filter(cause == "Ischemic stroke") |>
  distinct(PUBMED_ID, icd10_term) |>
  group_by(icd10_term) |>
  summarise(n_papers = length(unique(PUBMED_ID))) |>
  arrange(desc(n_papers))

```

## Intracerebral hemorrhage

```{r precision_intracerebral_hemorrhage}

disease_mapping_with_cause |>
  filter(cause == "Intracerebral hemorrhage") |>
  distinct(PUBMED_ID, icd10_term) |>
  group_by(icd10_term) |>
  summarise(n_papers = length(unique(PUBMED_ID))) |>
  arrange(desc(n_papers))

```

## Lip and oral cavity cancer

```{r precision_lip_and_oral_cavity_cancer}

disease_mapping_with_cause |>
  filter(cause == "Lip and oral cavity cancer") |>
  distinct(PUBMED_ID, icd10_term) |>
  group_by(icd10_term) |>
  summarise(n_papers = length(unique(PUBMED_ID))) |>
  arrange(desc(n_papers))


```


## Nasopharynx cancer

```{r precision_nasopharynx_cancer}

disease_mapping_with_cause |>
  filter(cause == "Nasopharynx cancer") |>
  distinct(PUBMED_ID, icd10_term) |>
  group_by(icd10_term) |>
  summarise(n_papers = length(unique(PUBMED_ID))) |>
  arrange(desc(n_papers))

```

## Other pharynx cancer

```{r precision_other_pharynx_cancer}

disease_mapping_with_cause |>
  filter(cause == "Other pharynx cancer") |>
  distinct(PUBMED_ID, icd10_term) |>
  group_by(icd10_term) |>
  summarise(n_papers = length(unique(PUBMED_ID))) |>
  arrange(desc(n_papers))

```

## Larynx cancer

```{r precision_larynx_cancer}

disease_mapping_with_cause |>
  filter(cause == "Larynx cancer") |>
  distinct(PUBMED_ID, icd10_term) |>
  group_by(icd10_term) |>
  summarise(n_papers = length(unique(PUBMED_ID))) |>
  arrange(desc(n_papers))

```

## Tracheal, bronchus, and lung cancer

```{r precision_tracheal_bronchus_lung_cancer}

disease_mapping_with_cause |>
  filter(cause == "Tracheal, bronchus, and lung cancer") |>
  distinct(PUBMED_ID, icd10_term) |>
  group_by(icd10_term) |>
  summarise(n_papers = length(unique(PUBMED_ID))) |>
  arrange(desc(n_papers))

```


## Liver cancer

```{r precision_liver_cancer}

disease_mapping_with_cause |>
filter(cause == "Liver cancer") |>
  distinct(PUBMED_ID, icd10_term) |>
  group_by(icd10_term) |>
  summarise(n_papers = length(unique(PUBMED_ID))) |>
  arrange(desc(n_papers))


```

## Rheumatic heart disease

```{r precision_rheumatic_heart_disease}

disease_mapping_with_cause |> 
filter(cause == "Rheumatic heart disease") |>
  distinct(PUBMED_ID, icd10_term) |>
  group_by(icd10_term) |>
  summarise(n_papers = length(unique(PUBMED_ID))) |>
  arrange(desc(n_papers))

```

## Stomach cancer

```{r precision_stomach_cancer}

disease_mapping_with_cause |> 
  filter(cause == "Stomach cancer") |>
  distinct(PUBMED_ID, icd10_term) |>
  group_by(icd10_term) |>
  summarise(n_papers = length(unique(PUBMED_ID))) |>
  arrange(desc(n_papers))

```


# Save Mapped GWAS Catalog Paper Mapping to GBD Causes

```{r save_final_gwas_gbd_mapping}

# remove studies not mapping to a cause
disease_mapping_with_cause <-
 disease_mapping_with_cause |>
  filter(cause != "" & !is.na(cause))
  
data.table::fwrite(
disease_mapping_with_cause,
here::here("output/icd_map/gwas_study_gbd_causes.csv")
)

```




