---
title: "Download trait ontology"
author: "Isobel Beasley"
output:
  workflowr::wflow_html:
    toc: true
    toc_float: true
    number_sections: true
editor_options: 
  markdown: 
    wrap: 72
---

# Set up

```{r setup, message=FALSE, warning=FALSE}

knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE
                      )

library(data.table)
library(dplyr)
library(ggplot2)
library(stringr)
source(here::here("code/get_term_descendants.R"))

```

```{r}

print("Version of EFO ontology used for trait ontology retrieval:")
cfg <- jsonlite::fromJSON("http://www.ebi.ac.uk/ols4/api/ontologies/efo")
cfg$config$version      
cfg$config$versionIri   
cfg$loaded; cfg$updated

```

# Get trait ontology terms relating to 'Disease' (etc.)

```{r set_up_trait_ontology_retrieval}

library(httr)
library(jsonlite)

```

## Mondo 'Human Disease' (<http://purl.obolibrary.org/obo/MONDO_0700096>)

```{r mondo_descendants, eval = T}

url <-
"http://www.ebi.ac.uk/ols4/api/ontologies/mondo/terms/http%253A%252F%252Fpurl.obolibrary.org%252Fobo%252FMONDO_0700096/descendants"

mondo_descendants <- get_descendants(url)

writeLines(mondo_descendants$terms, here::here("output/trait_ontology/mondo_0700096_descendants.txt"))

writeLines(mondo_descendants$iri_url, here::here("output/trait_ontology/mondo_0700096_descendants_iri.txt"))

```

## EFO 'Disease' Descendants (<http://www.ebi.ac.uk/efo/EFO_0000408>)

```{r efo_descendants, eval = T}

# url <-
# "http://www.ebi.ac.uk/ols4/api/ontologies/efo/terms/http%253A%252F%252Fwww.ebi.ac.uk%252Fefo%252FEFO_0000408/descendants"
# this term is now obsolete, so we use relevant archive version of EFO ontology to get descendants of EFO_0000408
# taken from: https://github.com/EBISPOT/efo/releases/tag/v3.81.0


library(ontologyIndex)   # NB: masks your own get_descendants() — qualify calls

efo <- get_ontology(here::here("data/ontology/efo_3.81.0.obo"),
                    propagate_relationships = "is_a",
                    extract_tags = "everything")

ids <- ontologyIndex::get_descendants(efo, 
                                      roots = "efo:EFO_0000408",
                                      exclude_roots = FALSE)

obo_to_iri <- function(id) {
  dplyr::case_when(
    startsWith(id, "efo:")      ~ paste0("http://www.ebi.ac.uk/efo/", sub(":", 
                                                                          "_", 
                                                                          stringr::str_remove_all(
                                                                            pattern = "^efo:",
                                                                            id
                                                                          )
                                         )),
    startsWith(id, "Orphanet:") ~ paste0("http://www.orpha.net/ORDO/", sub(":", "_", id)),
    TRUE                        ~ paste0("http://purl.obolibrary.org/obo/", sub(":", "_", id))
  )
}

efo_descendants <- tibble::tibble(
  obo_id  = ids,
  iri_url    = obo_to_iri(ids),
  terms   = stringr::str_squish(tolower(efo$name[ids]))
)


writeLines(efo_descendants$terms, here::here("output/trait_ontology/efo_0000408_descendants.txt"))
writeLines(efo_descendants$iri_url, here::here("output/trait_ontology/efo_0000408_descendants_iri.txt"))

```

## NCIT (Disease or Disorder): <http://purl.obolibrary.org/obo/NCIT_C2991>

```{r ncit_descendants, eval = T}

url <- "http://www.ebi.ac.uk/ols4/api/ontologies/ncit/terms/http%253A%252F%252Fpurl.obolibrary.org%252Fobo%252FNCIT_C2991/descendants"

ncit_descendants <- get_descendants(url)

writeLines(ncit_descendants$terms, 
           here::here("output/trait_ontology/ncit_C2991_descendants.txt"))
writeLines(ncit_descendants$iri_url, 
           here::here("output/trait_ontology/ncit_C2991_descendants_iri.txt"))

```

## Orphanet Disorder (<http://www.orpha.net/ORDO/Orphanet_557493>)

```{r orphanet, eval = T}

url <- "http://www.ebi.ac.uk/ols4/api/ontologies/ordo/terms/http%253A%252F%252Fwww.orpha.net%252FORDO%252FOrphanet_557493/descendants"

orphanet_descendants <- get_descendants(url)

writeLines(orphanet_descendants$terms, 
           here::here("output/trait_ontology/orphanet_557493_descendants.txt"))

writeLines(orphanet_descendants$iri_url, 
           here::here("output/trait_ontology/orphanet_557493_descendants_iri.txt"))

```

## Pathologic process (<http://purl.obolibrary.org/obo/OBI_1110122>)

```{r pathologic, eval = T}

url <- "http://www.ebi.ac.uk/ols4/api/ontologies/efo/terms/http%253A%252F%252Fpurl.obolibrary.org%252Fobo%252FOBI_1110122/descendants"

path_descendants <- get_descendants(url)

writeLines(path_descendants$terms, 
           here::here("output/trait_ontology/obi_1110122_descendants.txt"))
writeLines(path_descendants$iri_url, 
           here::here("output/trait_ontology/obi_1110122_descendants_iri.txt"))

```

## Disease course <http://purl.obolibrary.org/obo/OGMS_0000063>

```{r disease_course, eval = T}

url <- "http://www.ebi.ac.uk/ols4/api/ontologies/efo/terms/http%253A%252F%252Fpurl.obolibrary.org%252Fobo%252FOGMS_0000063/descendants"

disease_course_descendants <- get_descendants(url)

writeLines(disease_course_descendants$terms, here::here("output/trait_ontology/ogms_0000063_descendants.txt"))
writeLines(disease_course_descendants$iri_url, here::here("output/trait_ontology/ogms_0000063_descendants_iri.txt"))

```

## Age of onset of disease (http://purl.obolibrary.org/obo/OBA_2020000)

```{r age_of_onset, eval = T}

url <- "http://www.ebi.ac.uk/ols4/api/ontologies/oba/terms/http%253A%252F%252Fpurl.obolibrary.org%252Fobo%252FOBA_2020000/descendants"

age_of_onset_descendants <- get_descendants(url)

writeLines(age_of_onset_descendants$terms, here::here("output/trait_ontology/oba_2020000_descendants.txt"))
writeLines(age_of_onset_descendants$iri_url, here::here("output/trait_ontology/oba_2020000_descendants_iri.txt"))

```

# Get trait ontology terms relating to 'Phenotypic abnormality' (<http://purl.obolibrary.org/obo/HP_0000118>)

```{r hp_descendants, eval  = T}

url <- "http://www.ebi.ac.uk/ols4/api/ontologies/efo/terms/http%253A%252F%252Fpurl.obolibrary.org%252Fobo%252FHP_0000118/descendants"

hp_descendants <- get_descendants(url)

hp_descendants$terms = c(
                   "Abnormal aortic morphology",  
                   "Abnormal chest sounds",
                   "Abnormality of the eye", 
                   "Abnormal foot morphology",
                   "Abnormality of the gastrointestinal tract",                   
                   "Abnormality of the genital system",
                   "Abnormality of the liver",
                   "Abnormality of the nervous system",
                   "Abnormality of the skeletal system",
                   "Abnormality of the vasculature",
                   "Abnormal venous morphology",
                   "Phenotypic abnormality",
                   hp_descendants$terms)

hp_descendants$iri_url = c(
  "http://purl.obolibrary.org/obo/HP_0001679",
  "http://www.ebi.ac.uk/efo/EFO_0010670",
  "http://purl.obolibrary.org/obo/HP_0000478",
  "http://purl.obolibrary.org/obo/HP_0001760",
  "http://purl.obolibrary.org/obo/HP_0011024",
  "http://purl.obolibrary.org/obo/HP_0000078",
  "http://purl.obolibrary.org/obo/HP_0001392",
  "http://purl.obolibrary.org/obo/HP_0000707",
  "http://purl.obolibrary.org/obo/HP_0000924",
  "http://purl.obolibrary.org/obo/HP_0002597",
  "http://purl.obolibrary.org/obo/HP_0002624",
    "http://purl.obolibrary.org/obo/HP_0000118",
  hp_descendants$iri_url
)


writeLines(hp_descendants$terms, here::here("output/trait_ontology/hp_0000118_descendants.txt"))
writeLines(hp_descendants$iri_url, here::here("output/trait_ontology/hp_0000118_descendants_iri.txt"))

```

## Get trait ontology terms for sign or symptom (EFO:0003765)

```{r sign_or_symptom, eval = T}

url <- "http://www.ebi.ac.uk/ols4/api/ontologies/efo/terms/http%253A%252F%252Fwww.ebi.ac.uk%252Fefo%252FEFO_0003765/descendants"

efo_descendants <- get_descendants(url)

writeLines(efo_descendants$terms, here::here("output/trait_ontology/efo_0003765_descendants.txt"))
writeLines(efo_descendants$iri_url, here::here("output/trait_ontology/efo_0003765_descendants_iri.txt"))

```

# Get trait ontology terms relating to 'Measurement'

## Measurement terms (<http://www.ebi.ac.uk/efo/EFO_0001444>)

```{r measurement_descendants, eval = T}

base <- paste0("http://www.ebi.ac.uk/ols4/api/ontologies/efo/terms/",
               "http%253A%252F%252Fwww.ebi.ac.uk%252Fefo%252FEFO_0001444/descendants")

get_page <- function(p, size = 500) {
  res <- GET(base, query = list(page = p, size = size))
  stop_for_status(res)
  fromJSON(content(res, as = "text", encoding = "UTF-8"))
}

first <- get_page(0)
first$page          # <- look at totalElements / totalPages before anything else

out <- vector("list", first$page$totalPages)
for (p in seq_len(first$page$totalPages) - 1L) {
  d <- get_page(p)
  out[[p + 1L]] <- data.frame(
    iri        = d$`_embedded`$terms$iri,
    short_form = d$`_embedded`$terms$short_form,
    label      = d$`_embedded`$terms$label,
    stringsAsFactors = FALSE
  )
}
measurement_descendants <- do.call(rbind, out)
measurement_descendants_terms <- measurement_descendants$label
measurement_descendants_iri_url <- measurement_descendants$iri

# old measurement ids:
ids <- ontologyIndex::get_descendants(efo, 
                                      roots = "efo:EFO_0001444",
                                      exclude_roots = FALSE)

old_measurement_descendants <- tibble::tibble(
    obo_id  = ids,
    iri_url    = obo_to_iri(ids),
    terms   = stringr::str_squish(tolower(efo$name[ids]))
)

old_measurement_descendants <-
  old_measurement_descendants |> 
  filter(!c(terms %in% measurement_descendants_terms)) 

measurement_descendants_terms  <- c(measurement_descendants_terms,
                                     old_measurement_descendants$terms)

measurement_descendants_iri_url <- c(measurement_descendants_iri_url,
                                     old_measurement_descendants$iri_url)

disease_measurement_grep_terms <- c("age at",
                                    "addiction",
                                    "allergy",
                                    "allergic",
                                    "alcohol use disorder",
                                    "alzheimer's disease",
                                    "anxiety",
                                    "autoimmune disease",
                                    #"biomarker",
                                    "brain injury",
                                    "cancer biomarker",
                                    "cancer aggressiveness",
                                    "coronary atherosclerosis",
                                    "cardiovascular disease",
                                    "diagnosis",
                                    "diabetes mellitus",
                                    "dementia",
                                    "dependence",
                                    "depressive",
                                    "disease activity score",
                                    "distant metastasis free survival",
                                    "disease progression",
                                    "disease prognosis",
                                    "disease free survival",
                                    "metastasis free survival",
                                    "emphysema",
                                    "exacerbation",
                                    "event free survival time",
                                    "family history",
                                    "infection",
                                    "insomnia",
                                    "liver disease",
                                    "mental or behavioural disorder",
                                    "miscarriage",
                                    "onset",
                                    "osteoarthritis",
                                    "oppositional defiant disorder",
                                    "progression free survival",
                                    "post-traumatic stress disorder",
                                    "predisposition",
                                    "psoriasis",
                                    "respiratory disease",
                                    "severity",
                                    "susceptibility to",
                                    "suicide",
                                    "survival time",
                                    "staphylococcus aureus carrier",
                                    "stillbirth",
                                    "sleep apnea",
                                    "squamous cell carcinoma",
                                    "symptom",
                                    "withdrawal"
                                    )

seropositivity_terms <- grep("seropositivity|antibody", 
                             measurement_descendants_terms, 
                             value = T,
                             ignore.case = T)

# --- seropositivity terms to exclude: ---- #
gut_virome <-   grep("phage", 
                     seropositivity_terms, 
                     value = T) 
# gut commensal antibody panels
gut_commensal <- c(
  "akkermansia seropositivity",   
  "alistipes seropositivity",
  "bacillaceae seropositivity",   
  "bacilli seropositivity",  
  "bacillus seropositivity", 
  "bacteroidaceae seropositivity",
  "bacteroidales seropositivity", 
  "bacteroides seropositivity",   
  "barnesiella seropositivity",   
  "bifidobacterium seropositivity",
  "blautia seropositivity",  
  "butyrivibrio seropositivity",  
  "clostridiales seropositivity", 
  "coprococcus seropositivity",   
  "dorea seropositivity",    
  "eubacterium seropositivity",   
  "faecalibacterium seropositivity",
  "firmicutes seropositivity",    
  "lachnospiraceae seropositivity",
  "lactobacillales seropositivity",
  "lactobacillus seropositivity", 
  "odoribacter seropositivity",   
  "parabacteroides seropositivity",
  "prevotella seropositivity",    
  "roseburia seropositivity",
  "ruminococcaceae seropositivity",
  "flagellin seropositivity"
)

 ## ---- opportunists / commensals that can also be pathogens ------------- ###
gut_commensal_opportunists <- c(
  "enterobacteriaceae seropositivity",
  "enterococcus faecalis seropositivity",
  "escherichia seropositivity",
  "haemophilus parainfluenzae seropositivity",
  "pseudomonadaceae seropositivity"
)

other_sero_excl <- 
  c("ralstonia solanacearum seropositivity",
    "xanthomonas seropositivity",
    "fibronectin seropositivity"
    )

seropositivity_terms <- seropositivity_terms[!(seropositivity_terms %in% c(gut_virome, 
                                                     gut_commensal, 
                                                     gut_commensal_opportunists,
                                                     other_sero_excl))]

disease_measurement_terms = 
grep(pattern = paste0(disease_measurement_grep_terms, 
                      collapse = "|"),
     measurement_descendants_terms,
      value = T,
     ignore.case = T)

disease_measurement_terms <- c(disease_measurement_terms, 
                               seropositivity_terms)

is_disease_measurement <- 
  measurement_descendants_terms %in% disease_measurement_terms

disease_measurement_iri_urls <- 
  measurement_descendants_iri_url[is_disease_measurement]

measurement_descendants_terms <- 
  measurement_descendants_terms[!is_disease_measurement]

measurement_descendants_iri_urls <- 
  measurement_descendants_iri_url[!is_disease_measurement]

writeLines(measurement_descendants_terms, 
           here::here("output/trait_ontology/efo_0001444_descendants.txt"))
writeLines(measurement_descendants_iri_urls, 
           here::here("output/trait_ontology/efo_0001444_descendants_iri.txt"))

writeLines(disease_measurement_terms, 
           here::here("output/trait_ontology/efo_0001444_disease_measurement_terms.txt"))
writeLines(disease_measurement_iri_urls, 
           here::here("output/trait_ontology/efo_0001444_disease_measurement_terms_iri.txt"))

```

## Total cholesterol measurements (<http://www.ebi.ac.uk/efo/EFO_0004574>)

```{r, eval = T}

url <- "http://www.ebi.ac.uk/ols4/api/ontologies/efo/terms/http%253A%252F%252Fwww.ebi.ac.uk%252Fefo%252FEFO_0004574/descendants"

total_choles_descendants <- get_descendants(url)

total_choles_descendants$terms <- c("total cholesterol measurement", 
                                   total_choles_descendants$terms)

total_choles_descendants$iri_url <- c("http://www.ebi.ac.uk/efo/EFO_0004574",
                                      total_choles_descendants$iri_url)

writeLines(total_choles_descendants$terms, here::here("output/trait_ontology/efo_0004574_descendants.txt"))
writeLines(total_choles_descendants$iri_url, here::here("output/trait_ontology/efo_0004574_descendants_iri.txt"))

```

# Get ontology terms - response to stimulus (<http://purl.obolibrary.org/obo/GO_0050896>)

Includes:

-   Response to chemical (e.g. drug)
    (<http://purl.obolibrary.org/obo/GO_0042221>)
-   Immune response (<http://purl.obolibrary.org/obo/GO_0006955>)

## From GO

```{r, eval = T}

url <- "http://www.ebi.ac.uk/ols4/api/ontologies/go/terms/http%253A%252F%252Fpurl.obolibrary.org%252Fobo%252FGO_0050896/descendants"

go_response_descendants <- get_descendants(url)

writeLines(go_response_descendants$terms, 
           here::here("output/trait_ontology/go_0050896_descendants.txt"))
writeLines(go_response_descendants$iri_url, 
           here::here("output/trait_ontology/go_0050896_descendants_iri.txt"))


```

## From EFO

```{r, eval = T}

url <- "http://www.ebi.ac.uk/ols4/api/ontologies/efo/terms/http%253A%252F%252Fpurl.obolibrary.org%252Fobo%252FGO_0050896/descendants"

efo_response_descendants <- get_descendants(url)

writeLines(efo_response_descendants$terms, 
           here::here("output/trait_ontology/efo_go_0050896_descendants.txt"))

writeLines(efo_response_descendants$iri_url, 
           here::here("output/trait_ontology/efo_go_0050896_descendants_iri.txt"))

```

# Get ontology terms relating to 'Phenotype' (<http://www.ebi.ac.uk/efo/EFO_0000651>)

```{r trait_ontology, eval = T}

url <- "http://www.ebi.ac.uk/ols4/api/ontologies/efo/terms/http%253A%252F%252Fwww.ebi.ac.uk%252Fefo%252FEFO_0000651/descendants"

phenotype_descendants <- get_descendants(url)

writeLines(phenotype_descendants$terms, here::here("output/trait_ontology/efo_0000651_descendants.txt"))
writeLines(phenotype_descendants$iri_url, 
           here::here("output/trait_ontology/efo_0000651_descendants_iri.txt"))


```

# Mental process (<http://www.ebi.ac.uk/efo/EFO_0004323>)

```{r, eval = T}

url <- "http://www.ebi.ac.uk/ols4/api/ontologies/efo/terms/http%253A%252F%252Fwww.ebi.ac.uk%252Fefo%252FEFO_0004323/descendants"

mental_descendants <- get_descendants(url)

mental_descendants$terms <- c("mental process", 
                              "cognitive function measurement",
                              mental_descendants$terms)

mental_descendants$iri_url <- c("http://www.ebi.ac.uk/efo/EFO_0004323",
                             "http://www.ebi.ac.uk/efo/EFO_0008354",
                             mental_descendants$iri_url)


writeLines(mental_descendants$terms, 
           here::here("output/trait_ontology/efo_0004323_descendants.txt"))

writeLines(mental_descendants$iri_url, 
           here::here("output/trait_ontology/efo_0004323_descendants_iri.txt"))

```

# Behavior (<http://purl.obolibrary.org/obo/GO_0007610>)

```{r, eval = T}

url <- "http://www.ebi.ac.uk/ols4/api/ontologies/efo/terms/http%253A%252F%252Fpurl.obolibrary.org%252Fobo%252FGO_0007610/descendants"

behavior_descendants <- get_descendants(url)

behavior_descendants$terms = c("behavior", 
                               behavior_descendants$terms)

behavior_descendants$iri_url = c("http://purl.obolibrary.org/obo/GO_0007610",
                                behavior_descendants$iri_url)

writeLines(behavior_descendants$terms, here::here("output/trait_ontology/go_0007610_descendants.txt"))
writeLines(behavior_descendants$iri_url, here::here("output/trait_ontology/go_0007610_descendants_iri.txt"))


```

# Injury (<http://www.ebi.ac.uk/efo/EFO_0000546>)

```{r, eval = T}

# url <- "http://www.ebi.ac.uk/ols4/api/ontologies/efo/terms/http%253A%252F%252Fwww.ebi.ac.uk%252Fefo%252FEFO_0000546/descendants"

#url <- "http://www.ebi.ac.uk/ols4/api/ontologies/mondo/terms/http%253A%252F%252Fpurl.obolibrary.org%252Fobo%252FMONDO_0021178/descendants"
# this term is now obsolete, so we use relevant archive version of EFO ontology to get descendants

library(ontologyIndex)   # NB: masks your own get_descendants() — qualify calls

ids <- ontologyIndex::get_descendants(efo, 
                                      roots = "efo:EFO_0000546",
                                      exclude_roots = FALSE)

obo_to_iri <- function(id) {
  dplyr::case_when(
    startsWith(id, "efo:")      ~ paste0("http://www.ebi.ac.uk/efo/", sub(":", 
                                                                          "_", 
                                                                          stringr::str_remove_all(
                                                                            pattern = "^efo:",
                                                                            id
                                                                          )
                                         )),
    startsWith(id, "Orphanet:") ~ paste0("http://www.orpha.net/ORDO/", sub(":", "_", id)),
    TRUE                        ~ paste0("http://purl.obolibrary.org/obo/", sub(":", "_", id))
  )
}

injury_descendants  <- tibble::tibble(
  obo_id  = ids,
  iri_url     = obo_to_iri(ids),
  terms   = stringr::str_squish(tolower(efo$name[ids]))
)

writeLines(injury_descendants$terms, 
           here::here("output/trait_ontology/efo_0000546_descendants.txt"))
writeLines(injury_descendants$iri_url, 
           here::here("output/trait_ontology/efo_0000546_descendants_iri.txt"))

```

# Medical procedure terms (http://www.ebi.ac.uk/efo/EFO_0002571)

```{r medical_procedure_terms, eval = T}

url <- "http://www.ebi.ac.uk/ols4/api/ontologies/efo/terms/http%253A%252F%252Fwww.ebi.ac.uk%252Fefo%252FEFO_0002571/descendants"

medical_procedure_descendants <- get_descendants(url)

writeLines(medical_procedure_descendants$terms, here::here("output/trait_ontology/efo_0002571_descendants.txt"))
writeLines(medical_procedure_descendants$iri_url, here::here("output/trait_ontology/efo_0002571_descendants_iri.txt"))

```


# Surgical procedure terms (http://purl.obolibrary.org/obo/MAXO_0000004)

```{r surgical_procedure_terms, eval = T}

url <- "http://www.ebi.ac.uk/ols4/api/ontologies/maxo/terms/http%253A%252F%252Fpurl.obolibrary.org%252Fobo%252FMAXO_0000004/descendants"

surgical_procedure_descendants <- get_descendants(url)

surgical_procedure_descendants$terms <- c("surgical procedure", 
                                           surgical_procedure_descendants$terms)

surgical_procedure_descendants$iri_url = c("http://purl.obolibrary.org/obo/MAXO_0000004",
                                surgical_procedure_descendants$iri_url)

writeLines(surgical_procedure_descendants$terms, here::here("output/trait_ontology/maxo_0000004_descendants.txt"))

writeLines(surgical_procedure_descendants$iri_url, here::here("output/trait_ontology/maxo_0000004_descendants_iri.txt"))


```

 
