---
title: "Non-ontology based trait collapsing"
author: "Isobel Beasley"
date: "2025-09-06"
output:
  workflowr::wflow_html:
    toc: true
    toc_float: true
    number_sections: true
editor_options: 
  markdown: 
    wrap: 72
---


# Set up

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r required_fns, message = F, warning= F}

library(dplyr)
library(data.table)
library(stringr)

```


```{r setup_2, message = FALSE, warning= FALSE}

gwas_study_info <- fread(here::here("output/gwas_cat/gwas_study_info_trait_cat.csv"))

```

# Initial summary

## Objectives of this analysis:

- Collapsing and standardizing disease trait terms in GWAS study metadata without relying on ontology mappings.
- Initial harmonization of disease trait terms (fixing spelling, collapsing studies that study different aspects of the same disease)


# Number of studies per trait - before any grouping 

Let's look at the number of studies associated with each unique disease/trait term before any grouping or collapsing.

```{r initial_number_sum}

n_studies_trait = gwas_study_info |>
  dplyr::filter(DISEASE_STUDY == T) |>
  dplyr::select(all_disease_terms, PUBMED_ID) |>
  dplyr::distinct() |>
  dplyr::group_by(all_disease_terms) |>
  dplyr::summarise(n_studies = dplyr::n()) |>
  dplyr::arrange(desc(n_studies))

head(n_studies_trait)

dim(n_studies_trait)

```

# Disease trait collapsing and cleaning

```{r initial_cleaning}
# Basic cleaning of disease terms
# Removes trailing commas, 
gwas_study_info$all_disease_terms = sub(",$", "", gwas_study_info$all_disease_terms)

# trims whitespace
gwas_study_info$all_disease_terms = stringr::str_trim(gwas_study_info$all_disease_terms)

# remove apostrophes
gwas_study_info$all_disease_terms = stringr::str_remove_all(gwas_study_info$all_disease_terms, "'|’")

# initializes a new column (collected_all_disease_terms) to standardized traits
# collapse more traits 
gwas_study_info = gwas_study_info |> 
  mutate(collected_all_disease_terms = all_disease_terms) 

```


## Where a study investigates a given aspect of of disease, reduce this to a study of this disease

### Disease susceptibility 

#### Susceptibility to X measurement, decreased susceptibility to)

```{r sus_x_measurement}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms = 
        str_replace_all(collected_all_disease_terms, 
                    "susceptibility to (.*?) measurement(?=,|$)", "\\1")

         ) |>
    mutate(collected_all_disease_terms = 
        str_replace_all(collected_all_disease_terms, 
                    "susceptibility to (.*?) (.*?) measurement(?=,|$)", "\\1 \\2")

         )


gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms = 
        str_remove(collected_all_disease_terms, 
                   pattern = "(?<=^|, )susceptibility to "
         )
  )

```

#### Decreased susceptibility to X

```{r decreased_sus_x}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms = 
        str_remove_all(collected_all_disease_terms,
                       pattern = "(?<=^|, )decreased susceptibility to")
  )

```

#### Predisposition 

```{r predisposition}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms = 
        str_remove_all(collected_all_disease_terms,
                       pattern = " predisposition measurement(?=,|$)")
  )

```

#### Specific examples / changes to make - bruisability

```{r bruisability}

gwas_study_info = gwas_study_info |>
    mutate(collected_all_disease_terms = 
        str_replace_all(collected_all_disease_terms, 
                    "(?<=^|, )bruising susceptibility(?=,|$)",
                    "bruisability")
    )
  

```

### Disease severity 

#### Severity / symptom severity / exacerbation 

```{r disease_severity_initial}

gwas_study_info = gwas_study_info |>
    mutate(collected_all_disease_terms = 
          stringr::str_remove(collected_all_disease_terms, " symptom severity measurement(?=,|$)")
         ) 

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms = 
          stringr::str_remove(collected_all_disease_terms, " severity measurement(?=,|$)")
         ) 

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms = 
          stringr::str_remove(collected_all_disease_terms, " exacerbation measurement(?=,|$)")
         ) 

```

#### Specific example: psoriasis area and severity index

```{r psoriasis_area_severity_index}

gwas_study_info = gwas_study_info |>
    mutate(collected_all_disease_terms = 
          stringr::str_remove(collected_all_disease_terms, ", psoriasis area and severity index(?=,|$)")
         )

```

#### Specific example: uveal melanoma disease severity

```{r uveal_melanoma}

gwas_study_info = gwas_study_info |>
    mutate(collected_all_disease_terms = 
          stringr::str_replace_all(collected_all_disease_terms,
                           "(?<=^|, )uveal melanoma disease severity(?=,|$)",
                           "uveal melanoma"
          )
         )

```

### Disease symptom measurement

```{r symptom_measurement_initial}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms = 
        stringr::str_remove_all(collected_all_disease_terms,
                         " symptom measurement(?=,|$)")
         )

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms = 
        stringr::str_remove_all(collected_all_disease_terms,
                         " symptoms measurement(?=,|$)")
         )


gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms = 
        stringr::str_remove_all(collected_all_disease_terms,
                         " symptom count(?=,|$)")
         )


```

### General measurement

```{r general_measurement}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms = 
        str_replace_all(collected_all_disease_terms, 
                    "(.*?) (.*?) measurement(?=,|$) ", "\\1 \\2")
  )

```

#### Dependence / addiction / withdrawal measurement 

```{r dependence_addiction_withdrawal}
# dependence measurement -> dependence
gwas_study_info = gwas_study_info |>
    mutate(collected_all_disease_terms =
           str_replace_all(collected_all_disease_terms,
                           "\\bdependence\\s+measurement(?=,|$)",
                           "dependence")) 

# addiction measurement -> addiction
gwas_study_info = gwas_study_info |> 
  mutate(collected_all_disease_terms =
           str_replace_all(collected_all_disease_terms,
                           "\\baddiction\\s+measurement(?=,|$)",
                           "addiction"))

# withdrawal measurement -> withdrawal
gwas_study_info = gwas_study_info |> 
  mutate(collected_all_disease_terms =
           str_replace_all(collected_all_disease_terms,
                           "\\bwithdrawal\\s+measurement(?=,|$)",
                           "withdrawal"))

```

#### Syndrome / disorder measurement

```{r syndrome_disorder_measurement}

gwas_study_info = gwas_study_info |>
      mutate(collected_all_disease_terms =
           str_replace_all(collected_all_disease_terms,
                           "\\bsyndrome\\s+measurement(?=,|$)",
                           "syndrome"))

gwas_study_info = gwas_study_info |> 
      mutate(collected_all_disease_terms =
           str_replace_all(collected_all_disease_terms,
                           "\\bdisorder\\s+measurement(?=,|$)",
                           "disorder")) 

```

#### Allergy measurement

```{r allergy_measurement}

# allergy measurement -> allergy
gwas_study_info = gwas_study_info |> 
  mutate(collected_all_disease_terms =
           str_replace_all(collected_all_disease_terms,
                           "\\ballergy\\s+measurement(?=,|$)",
                           "allergy")) 

```

### Symptom measurement

```{r symptom_measurement}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms = 
        stringr::str_remove_all(collected_all_disease_terms,
                         " symptom measurement(?=,|$)")
         )

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms = 
        stringr::str_remove_all(collected_all_disease_terms,
                         " symptoms measurement(?=,|$)")
         )


gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms = 
        stringr::str_remove_all(collected_all_disease_terms,
                         " symptom count(?=,|$)")
         )

```

#### Specific example: autism 

```{r autism_symptom}

gwas_study_info = gwas_study_info |> 
    mutate(collected_all_disease_terms = 
           stringr::str_replace_all(collected_all_disease_terms,
                            "(?<=^|, )autism spectrum disorder symptom(?=,|$)",
                            "autism")
         ) 

```

#### Specific example: respiratory symptom change

```{r resp_symptom_change}


gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms = 
         stringr::str_remove_all(collected_all_disease_terms,
                          ", respiratory symptom change measurement(?=,|$)"
         ))

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms = 
         stringr::str_replace_all(collected_all_disease_terms,
                          "respiratory symptom change(?=,|$)",
                          "respiratory symptom" 
         ))

```

## Age of onset of

#### Non-specific correction 

```{r age_onset}

gwas_study_info = gwas_study_info |>
      mutate(collected_all_disease_terms = 
         stringr::str_remove_all(collected_all_disease_terms,
                              pattern = "(?<=^|, )age of onset of ")
         )

```

#### Specific correction: febrile seizure

```{r febrile_seizure}

# febrile seizure (within the age range of 3 months to 6 years)
gwas_study_info = gwas_study_info |> 
 mutate(collected_all_disease_terms  = 
          stringr::str_replace_all(collected_all_disease_terms ,
                                  pattern = "febrile seizure \\(within the age range of 3 months to 6 years\\)",
                                   "febrile seizure"
                          )  
        )

```

## Family history of 

#### Non-specific example

```{r family_history}

gwas_study_info = gwas_study_info |>
      mutate(collected_all_disease_terms = 
         stringr::str_remove_all(collected_all_disease_terms,
                         pattern = "(?<=^|, )family history of "))

```


## Disease progression


### Disease progression measurement

```{r dis_progress_measure}

gwas_study_info = gwas_study_info |>
    mutate(collected_all_disease_terms = 
         stringr::str_remove_all(collected_all_disease_terms,
                          "(?<=^|, )disease progression measurement(?=,|$)"
         ))

gwas_study_info = gwas_study_info |>
    mutate(collected_all_disease_terms = 
         stringr::str_remove_all(collected_all_disease_terms,
                          "disease progression measurement"
         ))

```

## Time to remission

### COVID

```{r covid_remission}

gwas_study_info = gwas_study_info |>
    mutate(collected_all_disease_terms = 
         stringr::str_replace_all(collected_all_disease_terms,
                          "(?<=^|, )time to remission of covid-19 symptoms(?=,|$)",
                          "covid-19"
         ))

```

## Alzheimer's disease - additional examples

### Neuropathological change

```{r alz_dis_neuropath}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms = 
         stringr::str_replace_all(collected_all_disease_terms,
                          "(?<=^|, )alzheimers disease neuropathologic change(?=,|$)",
                          "alzheimers disease"
         ))

```

### Biomarker measurement

```{r alz_dis_biomarker}
gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms = 
         stringr::str_replace_all(collected_all_disease_terms,
                          "(?<=^|, )alzheimers disease biomarker measurement(?=,|$)",
                          "alzheimers disease"
         ))

```

## Cancer aggressiveness measurement

```{r cancer_aggression}

gwas_study_info = 
gwas_study_info |> 
 mutate(collected_all_disease_terms  = 
          stringr::str_remove_all(collected_all_disease_terms,
                                  pattern = "(?<=^|, )cancer aggressiveness measurement(?=,|$)"
                          )
 ) 

```


## Other specific / non-defined measurement

### Anxiety

```{r anxiety_measure}

gwas_study_info = gwas_study_info |> 
    mutate(collected_all_disease_terms = 
         stringr::str_replace_all(collected_all_disease_terms,
                          "(?<=^|, )anxiety measurement(?=,|$)",
                          "anxiety"
         ))

```

### Coronary atherosclerosis

```{r coronary_athero_measure}

gwas_study_info = gwas_study_info |> 
    mutate(collected_all_disease_terms = 
         stringr::str_replace_all(collected_all_disease_terms,
                          "(?<=^|, )coronary atherosclerosis measurement(?=,|$)",
                          "coronary atherosclerosis"
         ))

```

## Cutaneous psoriasis 

```{r cutaneous_psoriasis_measure}

gwas_study_info = gwas_study_info |> 
    mutate(collected_all_disease_terms = 
         stringr::str_replace_all(collected_all_disease_terms,
                          "(?<=^|, )cutaneous psoriasis measurement(?=,|$)",
                          "cutaneous psoriasis"
         ))

```

### Depressive episode

```{r depressive_episode_measure}

gwas_study_info = gwas_study_info |> 
    mutate(collected_all_disease_terms = 
         stringr::str_replace_all(collected_all_disease_terms,
                          "(?<=^|, )depressive episode measurement(?=,|$)",
                          "depressive episode"
         ))

```

### Insomnia 

```{r insomnia_measure}

gwas_study_info = gwas_study_info |> 
    mutate(collected_all_disease_terms = 
         stringr::str_replace_all(collected_all_disease_terms,
                          "(?<=^|, )insomnia measurement(?=,|$)",
                          "insomnia"
         ))

```

### Lewy body dementia

```{r}

gwas_study_info = gwas_study_info |> 
    mutate(collected_all_disease_terms = 
         stringr::str_replace_all(collected_all_disease_terms,
                          "lewy body dementia measurement(?=,|$)",
                          "lewy body dementia"
         ))

```

### Sleep apnea

```{r sleep_apnea_measure}

gwas_study_info = gwas_study_info |> 
    mutate(collected_all_disease_terms = 
         stringr::str_replace_all(collected_all_disease_terms,
                          "(?<=^|, )sleep apnea measurement(?=,|$)",
                          "sleep apnea"
         ))

# during rem sleep
gwas_study_info = gwas_study_info |> 
    mutate(collected_all_disease_terms = 
         stringr::str_replace_all(collected_all_disease_terms,
                          "(?<=^|, )sleep apnea measurement during rem sleep(?=,|$)",
                          "sleep apnea"
         ))

# during non-rem sleep
gwas_study_info = gwas_study_info |> 
    mutate(collected_all_disease_terms = 
         stringr::str_replace_all(collected_all_disease_terms,
                          "(?<=^|, )sleep apnea measurement during non-rem sleep(?=,|$)",
                          "sleep apnea"
         ))

```


### Substance abuse measurement

### Alcohol

```{r alcohol_use_disorder}

gwas_study_info = gwas_study_info |> 
    mutate(collected_all_disease_terms = 
         stringr::str_replace_all(collected_all_disease_terms,
                          "(?<=^|, )alcohol withdrawal delirium(?=,|$)",
                          "alcohol withdrawal"
         ))


alcohol_use_disorder_terms <- c("(?<=^|, )alcohol dependence(?=,|$)",
                             "(?<=^|, )alcohol withdrawal(?=,|$)",
                             "(?<=^|, )alcohol abuse(?=,|$)",
                             "(?<=^|, )alcohol use disorder(?=,|$)",
                             "(?<=^|, )addictive alcohol use(?=,|$)"
                             )

gwas_study_info = gwas_study_info |>
      mutate(collected_all_disease_terms = 
           stringr::str_replace_all(collected_all_disease_terms,
                            paste0(alcohol_use_disorder_terms, collapse = "|"),
                            "alcohol-related disorders"
           ))

```


### Cocaine

```{r cocaine_use_disorder}

gwas_study_info = gwas_study_info |>
    mutate(collected_all_disease_terms = 
         stringr::str_replace_all(collected_all_disease_terms,
                          "cocaine use disorder(?=,|$)|cocaine dependence(?=,|$)",
                          "cocaine-related disorders"
         ))


```


### Nicotine 

```{r nicotine_use_disorder}

nicotine_use_disorder_terms <- c("(?<=^|, )nicotine dependence(?=,|$)",
                             "(?<=^|, )nicotine withdrawal(?=,|$)",
                             "(?<=^|, )nicotine addiction(?=,|$)"
                             )

gwas_study_info = gwas_study_info |>
    mutate(collected_all_disease_terms = 
         stringr::str_replace_all(collected_all_disease_terms,
                          paste0(nicotine_use_disorder_terms, collapse = "|"),
                          "nicotine-related disorders"
         ))


```


## Methamphetamine use disorders 

```{r methamphetamine_use_disorders}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          "(?<=^|, )methamphetamine dependence(?=,|$)|(?<=^|, )methamphetamine-induced psychosis(?=,|$)",
                          "methamphetamine use disorders"
         ))

```

## Deal with basic synonyms

### ADHD = attention deficit hyperactivity disorder
  
```{r adhd_synonym}

gwas_study_info = gwas_study_info |> 
    mutate(collected_all_disease_terms = 
           stringr::str_replace_all(collected_all_disease_terms,
                            "(?<=^|, )attention deficit hyperactivity disorder(?=,|$)",
                            "adhd"
         ))

```

### Anemia (phenotype) -> anemia

```{r anemia_synonym}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms =
         stringr::str_replace_all(collected_all_disease_terms,
                          "(?<=^|, )anemia \\(phenotype\\)(?=,|$)",
                          "anemia"
         ))

```

### Autism = Autism Spectrum Disorder = Asperger Syndrome

```{r autism_synonym}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms = 
        stringr::str_replace_all(collected_all_disease_terms,
                         "(?<=^|, )autism spectrum disorder(?=,|$)",
                         "autism")
        ) |>
    mutate(collected_all_disease_terms = 
        stringr::str_replace_all(collected_all_disease_terms,
                         "(?<=^|, )asperger syndrome(?=,|$)",
                         "autism")
        )

```

### Carcinoma = Cancer

```{r carcinoma_cancer_synonym}

gwas_study_info = gwas_study_info |>
      mutate(collected_all_disease_terms = 
           stringr::str_replace_all(collected_all_disease_terms,
                            "\\bcarcinoma",
                            "cancer")
         )

```

### Coronary artery disease = coronary atherosclerosis 

https://www.ebi.ac.uk/ols4/ontologies/efo/classes/http%253A%252F%252Fwww.ebi.ac.uk%252Fefo%252FEFO_0001645

```{r coronary_artery_disease_synonym}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms = 
         stringr::str_replace_all(collected_all_disease_terms,
                          "(?<=^|, )coronary atherosclerosis(?=,|$)",
                          "coronary artery disease"
         ))

```

### Head and neck cancer = head and neck malignant neoplasia

```{r head_neck_cancer_synonym}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms = 
         stringr::str_replace_all(collected_all_disease_terms,
                          "(?<=^|, )head and neck malignant neoplasia(?=,|$)",
                          "head and neck cancer"
         ))

```

### Hepatitis virus infection = hepatitis infection

```{r hepatitis_virus_infection_synonym}

gwas_study_info = gwas_study_info |>
    mutate(collected_all_disease_terms = 
         stringr::str_replace_all(collected_all_disease_terms,
                          "(?<=^|, )hepatitis a virus infection(?=,|$)",
                          "hepatitis a infection"
         )) |>
      mutate(collected_all_disease_terms = 
         stringr::str_replace_all(collected_all_disease_terms,
                          "(?<=^|, )hepatitis b virus infection(?=,|$)",
                          "hepatitis b infection"
         )) |>
        mutate(collected_all_disease_terms = 
         stringr::str_replace_all(collected_all_disease_terms,
                          "(?<=^|, )hepatitis c virus infection(?=,|$)",
                          "hepatitis c infection"
         ))

gwas_study_info = gwas_study_info |>
    mutate(collected_all_disease_terms = 
         stringr::str_replace_all(collected_all_disease_terms,
                          "(?<=^|, )chronic hepatitis c virus infection(?=,|$)",
                          "chronic hepatitis, hepatitis c infection"
         )) |>
      mutate(collected_all_disease_terms = 
         stringr::str_replace_all(collected_all_disease_terms,
                          "(?<=^|, )chronic hepatitis b virus infection(?=,|$)",
                          "chronic hepatitis, hepatitis b infection"
         ))

gwas_study_info = gwas_study_info |>
    mutate(collected_all_disease_terms = 
         stringr::str_replace_all(collected_all_disease_terms,
                          "(?<=^|, )hepatitis virus-related liver cancer(?=,|$)",
                          "hepatitis, liver cancer"
         ))


```

### Narcolepsy = Narcolepsy without cataplexy

```{r narcolepsy_synonym}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms = 
         stringr::str_replace_all(collected_all_disease_terms,
                          "(?<=^|, )narcolepsy without cataplexy(?=,|$)",
                          "narcolepsy"
         ))

```

### Mumps virus infection = mumps

```{r mumps_virus_infection_synonym}

gwas_study_info = gwas_study_info |>
    mutate(collected_all_disease_terms = 
         stringr::str_replace_all(collected_all_disease_terms,
                          "(?<=^|, )mumps virus infectious disease(?=,|$)",
                          "mumps"
         ))


```

### Renal cancer = kidney cancer

```{r renal_cancer_synonym}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms = 
         stringr::str_replace_all(collected_all_disease_terms,
                          "(?<=^|, )renal cancer(?=,|$)",
                          "kidney cancer"
         ))
  
```  

### Rubella infect = rubella

```{r rubella_infection_synonym}

gwas_study_info = gwas_study_info |>
    mutate(collected_all_disease_terms = 
         stringr::str_replace_all(collected_all_disease_terms,
                          "(?<=^|, )rubella infection(?=,|$)",
                          "rubella"
         ))


```


### Type 1 diabetes = type 1 diabetes mellitus

```{r type1_diabetes_synonym}

gwas_study_info = gwas_study_info |>
    mutate(collected_all_disease_terms = 
        stringr::str_replace_all(collected_all_disease_terms,
                         "(?<=^|, )type 1 diabetes(?=,|$)",
                         "type 1 diabetes mellitus")
        )

```

### Type 1 diabetes mellitus nephropathy = type 1 diabetes nephropathy

```{r type1_diabetes_nephropathy_synonym}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms = 
        stringr::str_replace_all(collected_all_disease_terms,
                         "(?<=^|, )type 1 diabetes mellitus nephropathy(?=,|$)",
                         "type 1 diabetes nephropathy")
        )

```




### Von willebrand disease (hereditary or acquired) = von willebrand disease

```{r vwf_synonym}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms =
         stringr::str_replace_all(collected_all_disease_terms,
                          "(?<=^|, )von willebrand disease \\(hereditary or acquired\\)(?=,|$)",
                          "von willebrand disease"
         ))

```


## Typos 

### Alzheimer -> alzheimers

```{r alzheimers_typo}
gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms =
         stringr::str_replace_all(collected_all_disease_terms,
                          "(?<=^|, )alzheimer disease(?=,|$)",
                          "alzheimers disease"
         ))
```

### Parkinson -> parkinsons

```{r parkinsons_typo}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms =
         stringr::str_replace_all(collected_all_disease_terms,
                          "(?<=^|, )parkinson disease(?=,|$)",
                          "parkinsons disease"
         ))

```


# Putting it all together

```{r, eval = F, include=F}

# gwas_study_info = gwas_study_info |>
#   rowwise() |>
#   mutate(
#     collected_all_disease_terms = paste0(
#       sort(unique(unlist(stringr::str_split(collected_all_disease_terms, ", ")))),
#       collapse = ", "
#     )
#   ) |>
#   ungroup()

           
gwas_study_info$collected_all_disease_terms = stringr::str_trim(gwas_study_info$collected_all_disease_terms)
gwas_study_info$collected_all_disease_terms = sub(",$", "", gwas_study_info$collected_all_disease_terms)
gwas_study_info$collected_all_disease_terms = stringr::str_replace_all(gwas_study_info$collected_all_disease_terms, "^, ", "")

```


## Final summary - number of unique diseases studied

```{r}

diseases <- stringr::str_split(pattern = ", ", 
                               gwas_study_info$collected_all_disease_terms[gwas_study_info$collected_all_disease_terms != ""])  |> 
            unlist() |>
            stringr::str_trim()

length(unique(diseases))

```

## Frequency of unique disease terms after harmonization

```{r freq_table}

# make frequency table
freq <- table(as.factor(diseases))

# sort in decreasing order
freq_sorted <- sort(freq, decreasing = TRUE)

# show top N, e.g. top 10
print("Top 5 most frequent disease terms after harmonization:")
head(freq_sorted, 5)

print("Number of unique disease terms after harmonization:")
sum(freq[freq == 1])

print("Average number of studies per disease term after harmonization:")
mean(freq)

```


## Save the updated gwas_study_info with repaired / harmonized disease terms

```{r save}

data.table::fwrite(
  gwas_study_info,
   here::here("output/gwas_cat/gwas_study_info_disease_trait_simplified.csv")
  )

```

