---
title: "Infectious Disease"
author: "Isobel Beasley"
date: "2025-09-18"
output:
  workflowr::wflow_html:
    toc: true
    toc_float: true
    number_sections: true
editor_options: 
  markdown: 
    wrap: 72
---

# Set up

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r required_fns, message = F, warning= F}

library(dplyr)
library(data.table)
library(stringr)

```

## Ontology help - for getting disease subtypes

```{r ontology_fns}

source(here::here("code/get_term_descendants.R"))

```

```{r setup_2, message = FALSE, warning= FALSE}

gwas_study_info <- fread(here::here("output/gwas_cat/gwas_study_info_disease_trait_simplified.csv"))

```

## Initial summary - number of unique study terms

```{r}

n_studies_trait = gwas_study_info |>
  dplyr::filter(DISEASE_STUDY == T) |>
  dplyr::select(collected_all_disease_terms, PUBMED_ID) |>
  dplyr::distinct() |>
  dplyr::group_by(collected_all_disease_terms) |>
  dplyr::summarise(n_studies = dplyr::n()) |>
  dplyr::arrange(desc(n_studies))

head(n_studies_trait)

dim(n_studies_trait)

```

### When separate studies with multiple terms 

```{r}

diseases <- stringr::str_split(pattern = ", ", 
 gwas_study_info$collected_all_disease_terms[gwas_study_info$collected_all_disease_terms != ""])  |> 
            unlist() |>
            stringr::str_trim()

length(unique(diseases))

```

# HIV/AIDS and sexually transmitted infections

## HIV/AIDS

### HIV subgrouping

```{r}

gwas_study_info = gwas_study_info |>
    mutate(collected_all_disease_terms = 
         stringr::str_replace_all(collected_all_disease_terms,
                          "hiv-1 infection",
                          "hiv infection"
         ))


```

```{r hiv_aids}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_remove_all(collected_all_disease_terms,
                          pattern = "(?<=^|, )hiv infection(?=,|$)"
         )
         )

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_remove_all(collected_all_disease_terms,
                          pattern = "^hiv infection$"
         )
         )

```

## Sexually transmitted infections (other than HIV)

### Syphilis

```{r syphilis}

gwas_study_info |>
  filter(grepl("syphi", collected_all_disease_terms, ignore.case = T)) |>
  select(`DISEASE/TRAIT`, collected_all_disease_terms)

```

### Clamydial infections

```{r clamydial_infections}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_remove_all(collected_all_disease_terms,
                          pattern = "(?<=^|, )chlamydophila infectious disease(?=,|$)"
         )
         )

```

### Gonorrhea

```{r gonorrhea}

gwas_study_info |>
  filter(grepl("gonorrhea", collected_all_disease_terms, ignore.case = T)) |>
  select(`DISEASE/TRAIT`, collected_all_disease_terms)

```

### Trichomoniasis

```{r trichomoniasis}

gwas_study_info |>
  filter(grepl("trichomoniasis", collected_all_disease_terms, ignore.case = T)) |>
  select(`DISEASE/TRAIT`, collected_all_disease_terms)

```

### Genital herpes

```{r genital_herpes}

gwas_study_info |>
  filter(grepl("herpes", collected_all_disease_terms, ignore.case = T)) |>
  select(`DISEASE/TRAIT`, collected_all_disease_terms)

```


### Other sexually transmitted infections

```{r other_stis}

url <- "http://www.ebi.ac.uk/ols4/api/ontologies/mesh/terms/http%253A%252F%252Fid.nlm.nih.gov%252Fmesh%252FD012749/descendants"

sti_terms <- get_descendants(url)

for(term in sti_terms){
  
  if(sum(grepl(term, gwas_study_info$collected_all_disease_terms)) > 0) {
    
  print(paste0("Term removed:", term))
    
  gwas_study_info = gwas_study_info |>
    mutate(collected_all_disease_terms  = 
           stringr::str_remove_all(collected_all_disease_terms,
                            pattern = paste0("(?<=^|, )", term, "(?=,|$)")
           )
    )
  
  }
}

```



# Respiratory infections and tuberculosis

## Tuberculosis

```{r tuberculosis}

tb_terms <- c("mycobacterium tuberculosis infection",
              "pulmonary tuberculosis",
              "extrapulmonary tuberculosis",
              "meningeal tuberculosis")


tb_terms <- paste0("(?<=^|, )", tb_terms, "(?=,|$)")

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          pattern = paste0(tb_terms, collapse = "|"),
                          "tuberculosis"
         ))

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_remove_all(collected_all_disease_terms,
                          pattern = "(?<=^|, )tuberculosis(?=,|$)"
         ))

```

## Lower respiratory infections

```{r lower_respiratory_infections}

# acute lower respiratory infections
url <- "http://www.ebi.ac.uk/ols4/api/ontologies/snomed/terms/http%253A%252F%252Fsnomed.info%252Fid%252F195742007/descendants"

lri_terms <- get_descendants(url)

for(term in lri_terms){
  
  if(sum(grepl(term, gwas_study_info$collected_all_disease_terms)) > 0) {
    
  print(paste0("Term removed:", term))
    
  gwas_study_info = gwas_study_info |>
    mutate(collected_all_disease_terms  = 
           stringr::str_remove_all(collected_all_disease_terms,
                            pattern = paste0("(?<=^|, )", term, "(?=,|$)")
           )
    )
  
  }
}

```

## Upper respiratory infections

```{r upper_respiratory_infections}

# acute upper respiratory infections
url <- "http://www.ebi.ac.uk/ols4/api/ontologies/snomed/terms/http%253A%252F%252Fsnomed.info%252Fid%252F54398005/descendants"

uri_terms <- get_descendants(url)

for(term in uri_terms){
  
  if(sum(grepl(term, gwas_study_info$collected_all_disease_terms)) > 0) {
    
  print(paste0("Term removed:", term))
    
  gwas_study_info = gwas_study_info |>
    mutate(collected_all_disease_terms  = 
           stringr::str_remove_all(collected_all_disease_terms,
                            pattern = paste0("(?<=^|, )", term, "(?=,|$)")
           )
    )
  
  }
}

```

## Otitis media

```{r otitis_media}

url <- "http://www.ebi.ac.uk/ols4/api/ontologies/doid/terms/http%253A%252F%252Fpurl.obolibrary.org%252Fobo%252FDOID_10754/descendants"

otitis_media_terms <- get_descendants(url)

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          pattern = paste0("(?<=^|, )", otitis_media_terms, "(?=,|$)", collapse = "|"),
                          "otitis media"
         ))


gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_remove_all(collected_all_disease_terms,
                          pattern = "(?<=^|, )otitis media(?=,|$)"
         )
         )


```

## Covid-19

```{r covid19}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_remove_all(collected_all_disease_terms,
                          pattern = "(?<=^|, )covid-19(?=,|$)"
         )
         )

```

## Other

### Influenza a (h1n1) (subset of influenza)

```{r}

gwas_study_info = gwas_study_info |>
    mutate(l1_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          "(?<=^|,)influenza a \\(h1n1\\)(?=,|$)",
                          "influenza"
         ))

```

# Enteric infections

## Diarrheal diseases

```{r diarrheal_diseases}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_remove_all(collected_all_disease_terms,
                          pattern = "(?<=^|, )astroviridae infectious disease, infantile diarrhea(?=,|$)"
         )
         )

# if diarrhea is caused by IBS, add ibs to collected_all_disease_terms

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
        ifelse(grepl("IBS", `DISEASE/TRAIT`, ignore.case = T) &
               grepl("diarrhea", collected_all_disease_terms, ignore.case = T),
               paste0(collected_all_disease_terms, ", irritable bowel syndrome"),
               collected_all_disease_terms
               )
         )

```


## Typhoid and paratyphoid fevers

### Typhoid fever

```{r typhoid_fever}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_remove_all(collected_all_disease_terms,
                          pattern = "(?<=^|, )typhoid fever(?=,|$)"
         )
         )

```

## Invasive Non-typhoidal salmonella

```{r invasive_non_typhoidal_salmonella}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_remove_all(collected_all_disease_terms,
                          pattern = "(?<=^|, )non-typhoidal salmonella bacteremia(?=,|$)"
         )
         )

```

## Other intestinal infectious diseases

### Gastroenteritis

```{r gastroenteritis}

# remove bacterial and viral gastroenteritis

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         ifelse(grepl("gastroenteritis", collected_all_disease_terms, ignore.case = T) &
                grepl("bacterial gastroenteritis|viral gastroenteritis", 
                       collected_all_disease_terms, 
                       ignore.case = T),
          stringr::str_remove_all(collected_all_disease_terms,
          pattern = "(?<=^|, )gastroenteritis(?=,|$)"),
          collected_all_disease_terms
         )
         )


```

### Enteritis 

```{r enteritis}

# remove viral and bacterial enteritis
gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         ifelse(grepl("enteritis", collected_all_disease_terms, ignore.case = T) &
                grepl("bacterial enteritis|viral enteritis", 
                       collected_all_disease_terms, 
                       ignore.case = T),
          stringr::str_remove_all(collected_all_disease_terms,
          pattern = "(?<=^|, )enteritis(?=,|$)"),
          collected_all_disease_terms
         )
         )


```



# Neglected tropical diseases and malaria

## Malaria

```{r malaria}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_remove_all(collected_all_disease_terms,
                          pattern = "(?<=^|, )malaria(?=,|$)"
         )
         )

```

## Chagas disease

```{r chagas_disease}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_remove_all(collected_all_disease_terms,
                          pattern = "(?<=^|, )chagas disease(?=,|$)"
         )
         )

```

## Leishmaniasis

```{r leishmaniasis}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_remove_all(collected_all_disease_terms,
                          pattern = "(?<=^|, )visceral leishmaniasis(?=,|$)"
         )
         )


gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_remove_all(collected_all_disease_terms,
                          pattern = "(?<=^|, )cutaneous leishmaniasis(?=,|$)"
         )
         )


```

## African trypanosomiasis

```{r african_trypanosomiasis}

gwas_study_info |>
  filter(grepl("trypanosomiasis", collected_all_disease_terms, ignore.case = T)) |>
  select(`DISEASE/TRAIT`, collected_all_disease_terms)

```

## Schistosomiasis

```{r schistosomiasis}

gwas_study_info |>
  filter(grepl("schistosomiasis", collected_all_disease_terms, ignore.case = T)) |>
  select(`DISEASE/TRAIT`, collected_all_disease_terms)

```

## Cysticercosis

```{r cysticercosis}

gwas_study_info |>
  filter(grepl("cysticercosis", collected_all_disease_terms, ignore.case = T)) |>
  select(`DISEASE/TRAIT`, collected_all_disease_terms)

```

## Cystic echinococcosis

```{r cystic_echinococcosis}

gwas_study_info |>
  filter(grepl("echinococcosis", collected_all_disease_terms, ignore.case = T)) |>
  select(`DISEASE/TRAIT`, collected_all_disease_terms)

```

## Lymphatic filariasis

```{r lymphatic_filariasis}

gwas_study_info |>
  filter(grepl("filariasis", collected_all_disease_terms, ignore.case = T)) |>
  select(`DISEASE/TRAIT`, collected_all_disease_terms)

```

## Onchocerciasis

```{r onchocerciasis}

gwas_study_info |>
  filter(grepl("onchocerciasis", collected_all_disease_terms, ignore.case = T)) |>
  select(`DISEASE/TRAIT`, collected_all_disease_terms)

```

## Trachoma

```{r trachoma}

gwas_study_info |>
  filter(grepl("trachoma", collected_all_disease_terms, ignore.case = T)) |>
  select(`DISEASE/TRAIT`, collected_all_disease_terms)


```

## Dengue

```{r dengue}


```

## Yellow fever

```{r yellow_fever}

gwas_study_info |>
  filter(grepl("yellow fever", collected_all_disease_terms, ignore.case = T)) |>
  select(`DISEASE/TRAIT`, collected_all_disease_terms)

```

## Rabies

```{r rabies}

gwas_study_info |>
  filter(grepl("rabies", collected_all_disease_terms, ignore.case = T)) |>
  select(`DISEASE/TRAIT`, collected_all_disease_terms)

```

## Intestinal nematode infections

```{r intestinal_nematode_infections}

# ascariasis
gwas_study_info |>
  filter(grepl("ascariasis", collected_all_disease_terms, ignore.case = T)) |>
  select(`DISEASE/TRAIT`, collected_all_disease_terms)

# trichuriasis
gwas_study_info |>
  filter(grepl("trichuriasis", collected_all_disease_terms, ignore.case = T)) |>
  select(`DISEASE/TRAIT`, collected_all_disease_terms)

# hookworm
gwas_study_info |>
  filter(grepl("hookworm", collected_all_disease_terms, ignore.case = T)) |>
  select(`DISEASE/TRAIT`, collected_all_disease_terms)


```

## Food borne trematode infections

```{r food_borne_trematode_infections}

gwas_study_info |>
  filter(grepl("trema", collected_all_disease_terms, ignore.case = T) 
         ) |>
  select(`DISEASE/TRAIT`, collected_all_disease_terms)


```


## Leprosy

```{r leprosy}

gwas_study_info |>
  filter(grepl("leprosy", collected_all_disease_terms, ignore.case = T)) |>
  select(`DISEASE/TRAIT`, collected_all_disease_terms) |>
  distinct()

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_remove_all(collected_all_disease_terms,
                          pattern = "(?<=^|, )leprosy(?=,|$)"
         )
         )

```

## Ebola

```{r ebola}

gwas_study_info |>
  filter(grepl("ebola", collected_all_disease_terms, ignore.case = T)) |>
  select(`DISEASE/TRAIT`, collected_all_disease_terms)

```

## Zika virus

```{r zika_virus}

gwas_study_info |>
  filter(grepl("zika", collected_all_disease_terms, ignore.case = T)) |>
  select(`DISEASE/TRAIT`, collected_all_disease_terms)

```


## Guinea worm disease

```{r guinea_worm_disease}

gwas_study_info |>
  filter(grepl("worm", collected_all_disease_terms, ignore.case = T)) |>
  select(`DISEASE/TRAIT`, collected_all_disease_terms)

```

## Other neglected tropical diseases

# Other infectious diseases

## Meningitis

```{r meningitis}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_remove_all(collected_all_disease_terms,
                          pattern = "(?<=^|, )infectious meningitis(?=,|$)"
         )
         )

```

## Encephalitis

```{r encephalitis}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_remove_all(collected_all_disease_terms,
                          pattern = "(?<=^|, )viral encephalitis(?=,|$)"
         )
         )


```


```{r tesrt}

# ? also remove if DISEASE/TRAIT == Encephalitis (PheCode 323)

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         ifelse(`DISEASE/TRAIT` == "Encephalitis (PheCode 323)",
         stringr::str_remove_all(collected_all_disease_terms,
                          pattern = "(?<=^|, )encephalitis(?=,|$)"),
         collected_all_disease_terms
         )
  )

```

## Diptheria

```{r diptheria}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_remove_all(collected_all_disease_terms,
                          pattern = "(?<=^|, )diphtheria(?=,|$)"
         )
         )

```

## Pertussis

```{r pertussis}

gwas_study_info |>
  filter(grepl("pertussis", collected_all_disease_terms, ignore.case = T)) |>
  select(`DISEASE/TRAIT`, collected_all_disease_terms)


```

## Tetanus

```{r tetanus}

gwas_study_info |>
  filter(grepl("tetanus", collected_all_disease_terms, ignore.case = T)) |>
  select(`DISEASE/TRAIT`, collected_all_disease_terms) 


```

## Measles

```{r measles}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_remove_all(collected_all_disease_terms,
                          pattern = "(?<=^|, )measles(?=,|$)"
         )
         )

```

## Varcella and herpes zoster

```{r varicella_herpes_zoster}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         ifelse(`DISEASE/TRAIT` == "Herpes zoster (PheCode 53)" |
                `DISEASE/TRAIT` == "Herpes zoster" |
                `DISEASE/TRAIT` == "Zoster infection" |
                grepl("ICD10 B02: Zoster \\[herpes zoster\\]", `DISEASE/TRAIT`),
         stringr::str_remove_all(collected_all_disease_terms,
                          pattern = "(?<=^|, )herpes zoster(?=,|$)"),
         collected_all_disease_terms
         )
  )


gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          pattern = "(?<=^|, )complication, herpes zoster, nervous system disease(?=,|$)",
                          "herpes zoster, nervous system disease"
                          
         )
         )

```

## Acute hepatitis

```{r acute_hepatitis}

# if acute hepatitis in DISEASE/TRAIT remove from collected_all_disease_terms
gwas_study_info |> 
  filter(grepl("acute hepat", 
                 `DISEASE/TRAIT`, 
                  ignore.case = T)) |> 
  pull(`DISEASE/TRAIT`) |>
  unique()

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         ifelse(`DISEASE/TRAIT` == "Acute hepatitis A infection",
         stringr::str_remove_all(collected_all_disease_terms,
                          pattern = "(?<=^|, )hepatitis a infection(?=,|$)"),
         collected_all_disease_terms
         )
  )

```


## Other unspecified infectious diseases

```{r other_unspecified_infectious_diseases_2}

gwas_study_info |>
  filter(grepl("epstein", collected_all_disease_terms, ignore.case = T)) |>
  pull(collected_all_disease_terms) |> unique()

```

### Infectious mononucleosis

```{r infectious_mononucleosis}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_remove_all(collected_all_disease_terms,
                          pattern = "(?<=^|, )infectious mononucleosis(?=,|$)"
         )
         )

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_remove_all(collected_all_disease_terms,
                          pattern = "(?<=^|, )epstein-barr virus infection(?=,|$)"
         )
         )

```

### Protozoa infectious disease

```{r protozoa_infectious_disease}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_remove_all(collected_all_disease_terms,
                          pattern = "(?<=^|, )protozoa infectious disease(?=,|$)"
         )
         )

```

### Herpes

```{r}


gwas_study_info = 
gwas_study_info |> 
 mutate(collected_all_disease_terms  = 
        ifelse(grepl("herpes simplex", `DISEASE/TRAIT`, ignore.case = T) &
               grepl("herpesviridae infectious disease", collected_all_disease_terms, ignore.case = T),
          stringr::str_replace_all(collected_all_disease_terms,
                                  pattern = "herpesviridae infectious disease",
                                  "herpes simplex infection"),
          collected_all_disease_terms
               )
          ) 


gwas_study_info |>
  filter(grepl("herpes", collected_all_disease_terms, ignore.case = T)) |>
  pull(`DISEASE/TRAIT`) |>
  unique()


```

## Mumps 

```{r mumps}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_remove_all(collected_all_disease_terms,
                          pattern = "(?<=^|, )mumps(?=,|$)"
         )
         )

```

```{r other_unspecified_infectious_diseases}


# parasitic infection



```


# Update and save output 

## Deal with duplicate terms created during grouping

```{r duplicate_terms}

gwas_study_info = 
 gwas_study_info |>
  rowwise() |>
  mutate(collected_all_disease_terms = paste0(sort(unique(unlist(strsplit(collected_all_disease_terms, ", ")))),
                                      collapse = ", ")
         ) |>
  ungroup()


```

## Deal with hanging commas and spaces

```{r clean_hanging_commas}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms = stringr::str_remove_all(collected_all_disease_terms, "^,|,$")
         ) |>
  mutate(collected_all_disease_terms = stringr::str_trim(collected_all_disease_terms)
         ) 

```

## Final summary - number of unique study terms

```{r}

n_studies_trait = gwas_study_info |>
  dplyr::filter(DISEASE_STUDY == T) |>
  dplyr::select(collected_all_disease_terms, PUBMED_ID) |>
  dplyr::distinct() |>
  dplyr::group_by(collected_all_disease_terms) |>
  dplyr::summarise(n_studies = dplyr::n()) |>
  dplyr::arrange(desc(n_studies))

head(n_studies_trait)

dim(n_studies_trait)

```

### When separate studies with multiple terms 

```{r}

diseases <- stringr::str_split(pattern = ", ", 
 gwas_study_info$collected_all_disease_terms[gwas_study_info$collected_all_disease_terms != ""])  |> 
            unlist() |>
            stringr::str_trim()

length(unique(diseases))

```


## Save output

```{r save_output}

fwrite(gwas_study_info,
here::here("output/gwas_cat/gwas_study_info_disease_trait_filtered.csv"))

```
