---
title: "Extra sample sizes from the GWAS Catalog"
author: "Isobel Beasley"
date: "2026-06-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load_packages}

library(dplyr)
library(stringr)
library(ggplot2)

```

# Filter dataset to relevant study accessions 

## Get relevant study accessions 

```{r get_study_accessions}

# get only relevant disease studies
gwas_study_info <- data.table::fread(here::here("output/icd_map/gwas_study_gbd_causes.csv"))

gwas_study_info = gwas_study_info |>
  dplyr::rename_with(~ gsub(" ", "_", .x))

# filter out infectious diseases
gwas_study_info <- gwas_study_info |>
    dplyr::filter(!cause %in% c("HIV/AIDS",
                             "Tuberculosis",
                             "Malaria",
                             "Lower respiratory infections",
                             "Diarrhoeal diseases",
                             "Neonatal disorders",
                             "Tetanus",
                             "Diphtheria",
                             "Pertussis" ,
                             "Measles",
                             "Maternal disorders"))

gwas_study_info <-
  gwas_study_info |>
  filter(cause != "")

selected_study_accessions <- gwas_study_info |> pull(STUDY_ACCESSION) |> 
  unique()

length(selected_study_accessions)

```


```{r load_gwas_ancest}

gwas_ancest_info <-  data.table::fread(here::here("data/gwas_catalog/gwas-catalog-v1.0.3.1-ancestries-r2025-07-21.tsv"),
                                       sep = "\t",
                                       quote = "")

gwas_ancest_info = gwas_ancest_info |>
  dplyr::rename_with(~ gsub(" ", "_", .x))

gwas_ancest_subset <-
  gwas_ancest_info |>
  filter(STUDY_ACCESSION %in% selected_study_accessions)

```

# Pre-processing GWAS Catalog ancestry dataset

## BROAD_ANCESTRAL_CATEGORY 

```{r reorder_BROAD_ANCESTRAL_CATEGORY}

# reorder BROAD_ANCESTRAL_CATEGORY alphabetically
gwas_ancest_subset  <-
  gwas_ancest_subset  |>
  mutate(BROAD_ANCESTRAL_CATEGORY = purrr::map_chr(
    str_split(BROAD_ANCESTRAL_CATEGORY,
              stringr::regex("(?<!Eastern), ")),
    ~ paste(sort(.x), collapse = ", ")))

gwas_ancest_subset |>
  filter(str_detect(BROAD_ANCESTRAL_CATEGORY,
                    "(?<!Eastern), ")) |>
  pull(BROAD_ANCESTRAL_CATEGORY) |>
  unique() 


```

```{r correct_gwas_BROAD_ANCESTRAL_CATEGORY}

gwas_ancest_subset |>
  filter(BROAD_ANCESTRAL_CATEGORY == "NR, Other")

# remove NR, Othr group
gwas_ancest_subset <-
gwas_ancest_subset |>
  mutate(BROAD_ANCESTRAL_CATEGORY =
           ifelse(BROAD_ANCESTRAL_CATEGORY ==  "NR, Other",
                  "NR",
                  BROAD_ANCESTRAL_CATEGORY)
  )


gwas_ancest_subset |>
  filter(BROAD_ANCESTRAL_CATEGORY == "Asian unspecified, East Asian")

# Japanese American
gwas_ancest_subset <-
  gwas_ancest_subset |>
  mutate(BROAD_ANCESTRAL_CATEGORY =
           ifelse(BROAD_ANCESTRAL_CATEGORY ==  "Asian unspecified, East Asian",
                  "Asian unspecified",
                  BROAD_ANCESTRAL_CATEGORY)
  )

```
```{r}

# not inferrable BROAD_ANCESTRAL
gwas_ancest_subset <-
  gwas_ancest_subset |>
  mutate(OLD_BROAD_ANCESTRAL_CATEGORY = BROAD_ANCESTRAL_CATEGORY)

#  GCST010272
gwas_ancest_subset <-
gwas_ancest_subset |>
  mutate(BROAD_ANCESTRAL_CATEGORY = ifelse(STUDY_ACCESSION == "GCST010272",
                                           "NR",
                                           BROAD_ANCESTRAL_CATEGORY))

# GCST004944
gwas_ancest_subset <-
gwas_ancest_subset |>
  mutate(BROAD_ANCESTRAL_CATEGORY = ifelse(STUDY_ACCESSION == "GCST004944",
                                           "NR",
                                           BROAD_ANCESTRAL_CATEGORY))

# GCST011660
# GCST004944 (only NR)
gwas_ancest_subset <-
  gwas_ancest_subset |>
  filter(!c(STUDY_ACCESSION %in% c('GCST011660',
                                   'GCST011659',
                                   'GCST011659')))

# GCST010557
gwas_ancest_subset <-
gwas_ancest_subset |>
  mutate(BROAD_ANCESTRAL_CATEGORY = ifelse(STUDY_ACCESSION == "GCST010557" & 
                                            BROAD_ANCESTRAL_CATEGORY == "East Asian, South Asian",
                                           "Asian unspecified",
                                           BROAD_ANCESTRAL_CATEGORY))

# GCST90000582
# 32887874 

```

## INITIAL_SAMPLE_DESCRIPTION

```{r}

# 	GCST012536
# GCST90281125

# 39620209 and East Asian in front of controls
# GCST90492696
# GCST90492681

# 36738649 for, sum all NUMBER of individuals
# GCST010078

# GCST005536 sum all N

# GCST90225531 in-inferable.

# GCST003866 ? error
# GCST004757

```

### Specific to particular studies 

```{r correcting_individual_INITIAL_SAMPLE_DESCRIPTION}

# for study accession: GCST010978
# replace "cases and 1,025" with
# cases, 1,025
gwas_ancest_subset <-
  gwas_ancest_subset |>
  mutate(INITIAL_SAMPLE_DESCRIPTION = ifelse(STUDY_ACCESSION == "GCST010978",
         str_replace_all(string = INITIAL_SAMPLE_DESCRIPTION,
                         pattern = "cases and 1,025",
                         replacement = "cases, 1,025"),
         INITIAL_SAMPLE_DESCRIPTION)
  )


# for pubmed id: 29315334
# introduce space into: "controls,15,283
gwas_ancest_subset <-
  gwas_ancest_subset |>
  mutate(INITIAL_SAMPLE_DESCRIPTION = ifelse(PUBMED_ID == "29315334",
         str_replace_all(string = INITIAL_SAMPLE_DESCRIPTION,
                         pattern = "controls,15,283",
                         replacement = "controls, 15,283"),
         INITIAL_SAMPLE_DESCRIPTION)
  )

# for study accession: GCST002298
# cases and 63
gwas_ancest_subset <-
  gwas_ancest_subset |>
  mutate(INITIAL_SAMPLE_DESCRIPTION = ifelse(STUDY_ACCESSION == "GCST002298",
                                             str_replace_all(string = INITIAL_SAMPLE_DESCRIPTION,
                                                             pattern = "cases and 63",
                                                             replacement = "cases, 63"),
                                             INITIAL_SAMPLE_DESCRIPTION)
  )


# cases and 112 Arab
gwas_ancest_subset <-
  gwas_ancest_subset |>
  mutate(INITIAL_SAMPLE_DESCRIPTION = ifelse(STUDY_ACCESSION == "GCST002123",
                                             str_replace_all(string = INITIAL_SAMPLE_DESCRIPTION,
                                                             pattern = "cases and 112",
                                                             replacement = "cases, 112"),
                                             INITIAL_SAMPLE_DESCRIPTION)
  )


#  and 10,588
gwas_ancest_subset <-
  gwas_ancest_subset |>
  mutate(INITIAL_SAMPLE_DESCRIPTION = ifelse(STUDY_ACCESSION == "GCST004086",
                                             str_replace_all(string = INITIAL_SAMPLE_DESCRIPTION,
                                                             pattern = "and 10,588",
                                                             replacement = ", 10,588"),
                                             INITIAL_SAMPLE_DESCRIPTION)
  )


# and 4,606
gwas_ancest_subset <-
  gwas_ancest_subset |>
  mutate(INITIAL_SAMPLE_DESCRIPTION = ifelse(STUDY_ACCESSION == "GCST006145",
                                             str_replace_all(string = INITIAL_SAMPLE_DESCRIPTION,
                                                             pattern = "and 4,606",
                                                             replacement = ", 4,606"),
                                             INITIAL_SAMPLE_DESCRIPTION)
  )

# individuals,16,962
gwas_ancest_subset <-
  gwas_ancest_subset |>
  mutate(INITIAL_SAMPLE_DESCRIPTION = ifelse(PUBMED_ID == "33619380",
                                             str_replace_all(string = INITIAL_SAMPLE_DESCRIPTION,
                                                             pattern = "individuals,16,962",
                                                             replacement = "individuals, 16,962"),
                                             INITIAL_SAMPLE_DESCRIPTION)
  )


```


```{r correct_cross-study_INITIAL_SAMPLE_DESCRIPTION}

# remove up to phrases
gwas_ancest_subset <-
  gwas_ancest_subset |>
  mutate(INITIAL_SAMPLE_DESCRIPTION =
           str_remove_all(
             string  = INITIAL_SAMPLE_DESCRIPTION,
             pattern = "(?<=^|, )[Uu]p to "
           )
  ) |>
  mutate(INITIAL_SAMPLE_DESCRIPTION =
           str_remove_all(
             string  = INITIAL_SAMPLE_DESCRIPTION,
             pattern = "(?<=^|, )upto "
           )
  ) |>
    mutate(INITIAL_SAMPLE_DESCRIPTION =
             str_replace_all(
               string  = INITIAL_SAMPLE_DESCRIPTION,
               pattern = ",up to",
               replacement = ","
             )
    )

# remove phrase, approximately
gwas_ancest_subset <-
  gwas_ancest_subset |>
  mutate(INITIAL_SAMPLE_DESCRIPTION =
           str_remove_all(
             string  = INITIAL_SAMPLE_DESCRIPTION,
             pattern = "approximately "
           )
  ) |>
  mutate(INITIAL_SAMPLE_DESCRIPTION =
           str_remove_all(
             string  = INITIAL_SAMPLE_DESCRIPTION,
             pattern = "\\~ "
           )
  )

# remove phrase, approximately
gwas_ancest_subset <-
  gwas_ancest_subset |>
  mutate(INITIAL_SAMPLE_DESCRIPTION =
           str_remove_all(
             string  = INITIAL_SAMPLE_DESCRIPTION,
             pattern = "\\~"
           )
  )


# remove
# at least phrases
gwas_ancest_subset <-
  gwas_ancest_subset |>
  mutate(INITIAL_SAMPLE_DESCRIPTION =
           str_remove_all(
             string  = INITIAL_SAMPLE_DESCRIPTION,
             pattern = "(?<=^|, )[Aa]t least "
           )
  )



```


## REPLICATION_SAMPLE_DESCRIPTION

```{r correct_cross-study_REPLICATION_SAMPLE_DESCRIPTION}

gwas_ancest_subset  |> filter(grepl("and [0-9]+", REPLICATION_SAMPLE_DESCRIPTION)) |> 
  pull(REPLICATION_SAMPLE_DESCRIPTION) |>
  unique()

gwas_ancest_subset <- 
gwas_ancest_subset |>
  mutate(REPLICATION_SAMPLE_DESCRIPTION = 
           str_replace_all(string = REPLICATION_SAMPLE_DESCRIPTION,
                           pattern = " and (?=[0-9])",
                           replacement = ", "))

# remove up to phrases
gwas_ancest_subset <-
  gwas_ancest_subset |>
  mutate(REPLICATION_SAMPLE_DESCRIPTION =
           str_remove_all(
             string  = REPLICATION_SAMPLE_DESCRIPTION,
             pattern = "(?<=^|, )[Uu]p to "
           )
  )

# remove
# at least phrases
gwas_ancest_subset <-
  gwas_ancest_subset |>
  mutate(REPLICATION_SAMPLE_DESCRIPTION =
           str_remove_all(
             string  = REPLICATION_SAMPLE_DESCRIPTION,
             pattern = "(?<=^|, )[Aa]t least "
           )
  )

gwas_ancest_subset  |> filter(grepl("at least", REPLICATION_SAMPLE_DESCRIPTION)) |> 
  pull(REPLICATION_SAMPLE_DESCRIPTION) |>
  unique()


```


# (INITIAL) Extract population descriptors 

```{r extract_population_desc_INITIAL}

initial_sample_desc <-
gwas_ancest_subset |>
  filter(STAGE == "initial") |>
  select(INITIAL_SAMPLE_DESCRIPTION,
         BROAD_ANCESTRAL_CATEGORY,
         NUMBER_OF_INDIVIDUALS,
         STUDY_ACCESSION,
         PUBMED_ID) |>
  tidyr::separate_longer_delim(INITIAL_SAMPLE_DESCRIPTION,
                               delim = regex(",\\s+(?=\\d)")) |>
  mutate(
      n = INITIAL_SAMPLE_DESCRIPTION |>
      str_extract("[\\d,]+\\s") |>      # leading number (may contain thousands commas)
      str_remove_all(",") |>
      as.integer(),
    sample_description = INITIAL_SAMPLE_DESCRIPTION |>
      str_remove("^(up to\\s+)?[\\d,]+\\s+")  # drop "up to" + the number
  ) 

# check where there is not comma in INITIAL_SAMPLE_DESCRIPTION
# n should equal n
initial_sample_desc |>
  filter(!str_detect(INITIAL_SAMPLE_DESCRIPTION,
                    ",\\s+(?=\\d)")) |>
  #filter(!grepl(",[0", INITIAL_SAMPLE_DESCRIPTION)) |>
  filter(grepl("^[0-9]+", INITIAL_SAMPLE_DESCRIPTION)) |>
  filter(!grepl("trios|pairs", sample_description)) |>
  filter(NUMBER_OF_INDIVIDUALS != n) |>
  nrow()

initial_sample_desc <-
  initial_sample_desc |>
  mutate(NUMBER_OF_INDIVIDUALS_ORIG = NUMBER_OF_INDIVIDUALS) |>
  mutate(NUMBER_OF_INDIVIDUALS = ifelse(!str_detect(INITIAL_SAMPLE_DESCRIPTION,
                    ",\\s+(?=\\d)") &
                                         grepl("^[0-9]+", INITIAL_SAMPLE_DESCRIPTION) &
                                         !grepl("trios|pairs", sample_description) & 
                                          NUMBER_OF_INDIVIDUALS != n,
                                        n,
                                        NUMBER_OF_INDIVIDUALS_ORIG)
  )
  

# check no missed samples
initial_sample_desc |>
  filter(str_detect(sample_description,
                    "(?<![Tt]ype )(?<!COVID-)(?<!\\< )(?<!\\> )\\b[0-9]+(?!\\%)(?!\\-)")) |>
  nrow()

initial_sample_desc |>
  filter(str_detect(sample_description,
                    "(?<![Tt]ype )(?<!COVID-)(?<!\\< )(?<!\\>)\\b[0-9]+\\b(?!\\%)")) |>
  pull(sample_description) |>
  unique()

```

### Account for sibling pairs / trios in numbers

* Add / update NUMBER_OF_INDIVIDUALS

```{r}

initial_sample_desc <- 
initial_sample_desc |>
  mutate(n_text = n)

initial_sample_desc <- 
initial_sample_desc |>
  mutate(n = ifelse(STUDY_ACCESSION == "GCST005570" & 
                    grepl("trios", sample_description),
                    3*n,
                    n
                    ))

initial_sample_desc <- 
initial_sample_desc |>
  mutate(n = ifelse(STUDY_ACCESSION == "GCST000141" & 
                    grepl("trios", sample_description),
                    3*n,
                    n
                    ))

initial_sample_desc <- 
initial_sample_desc |>
  mutate(n = ifelse(STUDY_ACCESSION == "GCST001689" & 
                    grepl("trios", sample_description),
                    3*n,
                    n
                    ))

initial_sample_desc <- 
initial_sample_desc |>
  mutate(n = ifelse(STUDY_ACCESSION == "GCST000054" & 
                    grepl("trios", sample_description) &
                    grepl("trios", INITIAL_SAMPLE_DESCRIPTION),
                    3*n,
                    n
                    ))

initial_sample_desc <- 
initial_sample_desc |>
  mutate(n = ifelse(STUDY_ACCESSION == "GCST000244" & 
                    grepl("trios", INITIAL_SAMPLE_DESCRIPTION),
                    3*n,
                    n
                    ))

# 2,601 ASP families and 69 trio families (10,796 total)
initial_sample_desc |>
  filter(STUDY_ACCESSION == "GCST005536")

initial_sample_desc <- 
initial_sample_desc |>
  mutate(n = ifelse(STUDY_ACCESSION == "GCST005536" & 
                    grepl("trios", INITIAL_SAMPLE_DESCRIPTION),
                    3*n,
                    n
                    ))

initial_sample_desc <- 
initial_sample_desc |>
  mutate(n = ifelse(STUDY_ACCESSION == "GCST005536" & 
                    grepl("families", INITIAL_SAMPLE_DESCRIPTION),
                    4.071126*n, # as (10796 - 3* 69 )/(2601)
                    n
                    ))


```

## Get broad ancestry group of each sample description


```{r get_broad_ancestry_INITIAL}

source(here::here("code/custom_plotting.R"))

initial_ancestry_sample_desc <-
  initial_sample_desc |>
  # extract population descriptors
  mutate(sample_ancestry = get_ancestry_from_sample_desc(sample_description)) |>
  distinct()

# initial_ancestry_sample_desc <-
# initial_ancestry_sample_desc |>
#    tidyr::separate_longer_delim(sample_ancestry, "; ") |>
#     mutate(sample_broad_ancestry = group_pop_labels(sample_ancestry)
#   ) 

initial_ancestry_sample_desc <-
initial_ancestry_sample_desc |>
#   filter(!str_detect(BROAD_ANCESTRAL_CATEGORY,
#                     "(?<!Eastern), ")) |> 
# tidyr::separate_longer_delim(sample_ancestry, "; ") |>
# group population descriptors
rowwise() |>
mutate(sample_broad_ancestry =
           paste0(sort(
                        unique(group_pop_labels(str_split(string = sample_ancestry,
                                                          "; ",
                                                          simplify = TRUE)
                                                )
                               )
                        ),
                  collapse = ", ")
  ) |>
  ungroup()

initial_ancestry_sample_desc <-
  initial_ancestry_sample_desc |>
  distinct()

# initial_ancestry_sample_desc |>
#   filter(str_detect(sample_broad_ancestry,
#                     "(?<!Eastern), "))

```


## Match broad ancestry groups 

```{r match_ancestry_INITIAL}

initial_ancestry_sample_desc |>
  filter(str_detect(BROAD_ANCESTRAL_CATEGORY,
                    "(?<!Eastern), "))

# where there are multiple broad ancestral categories
# separate into multiple rows
initial_ancestry_sample_desc <-
  initial_ancestry_sample_desc |>
  mutate(BROAD_ANCESTRAL_CATEGORY_ORIG = BROAD_ANCESTRAL_CATEGORY) |>
  tidyr::separate_longer_delim(BROAD_ANCESTRAL_CATEGORY,
                               delim = stringr::regex("(?<!Eastern), "))


```


## Compare estimated total number of samples

```{r compare_total_n_INITIAL}


initial_ancestry_sample_desc <-
initial_ancestry_sample_desc |>
  filter(sample_broad_ancestry == BROAD_ANCESTRAL_CATEGORY)

  initial_summed_sample_desc <-
  initial_ancestry_sample_desc |>
    # filter(str_detect(BROAD_ANCESTRAL_CATEGORY_ORIG,
    #                 "(?<!Eastern), ")) |>
    group_by(PUBMED_ID,
           STUDY_ACCESSION,
           NUMBER_OF_INDIVIDUALS,
           BROAD_ANCESTRAL_CATEGORY_ORIG,
           INITIAL_SAMPLE_DESCRIPTION,
           sample_description,
           n_text) |>
  summarise(n = paste0(unique(n), collapse = ", "),
            BROAD_ANCESTRAL_CATEGORY = paste0(sort(unique(BROAD_ANCESTRAL_CATEGORY)),
                                              collapse = ", "),
            sample_broad_ancestry = paste0(sort(unique(sample_broad_ancestry)),
                                              collapse = ", "))
              

# NA's introduced for GCST90449028
# fine - number of individuals is NA for this STUDY_ACCESSION
  initial_summed_sample_desc <-
  initial_ancestry_sample_desc |>
  ungroup() |>
  mutate(n = as.integer(n)) |>
  #filter(sample_broad_ancestry == BROAD_ANCESTRAL_CATEGORY) |>
  group_by(PUBMED_ID,
           STUDY_ACCESSION,
           NUMBER_OF_INDIVIDUALS,
           INITIAL_SAMPLE_DESCRIPTION,
           BROAD_ANCESTRAL_CATEGORY) |>
  summarise(n_sample_desc = sum(n))


  initial_summed_sample_desc |>
  filter(!between(NUMBER_OF_INDIVIDUALS,
                 n_sample_desc,
                 n_sample_desc + 10)) |>
  nrow()
  
  initial_summed_sample_desc |>
  filter(n_sample_desc != NUMBER_OF_INDIVIDUALS) |>
  nrow()
  
    initial_summed_sample_desc |>
    filter(n_sample_desc != NUMBER_OF_INDIVIDUALS)  |>
    pull(PUBMED_ID) |>
    unique() |> 
      length()


```
