---
title: "Get supplemental files"
author: "Isobel Beasley"
date: "2026-03-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Ideas/help for downloading supplemental files: 
- https://pmc.ncbi.nlm.nih.gov/articles/PMC12371329/

What file types are likely to be relevant?
- pdf
- Excel spreadsheet: (xls, xlsx)
- Data files (csv, txt)
- Word documents (docx, doc) 

## Required packages 

```{r, message=FALSE}

library(httr)
library(xml2)
library(stringr)
library(here)
library(dplyr)
library(data.table)

```


# Get PMCIDs and PUBMIDs for GWAS Catalog studies of diseases

## Pubmed ids from GWAS catalog

```{r get_pmids}

## Step 1: 
# get only relevant disease studies
gwas_study_info <- data.table::fread(here::here("output/icd_map/gwas_study_gbd_causes.csv"))

gwas_study_info = gwas_study_info |>
  dplyr::rename_with(~ gsub(" ", "_", .x))

# filter out infectious diseases
gwas_study_info <- gwas_study_info |>
    dplyr::filter(!cause %in% c("HIV/AIDS",
                             "Tuberculosis",
                             "Malaria",
                             "Lower respiratory infections",
                             "Diarrhoeal diseases",
                             "Neonatal disorders",
                             "Tetanus",
                             "Diphtheria",
                             "Pertussis" ,
                             "Measles",
                             "Maternal disorders"))

print("Number of disease studies to get full texts for:")
all_pmids <- unique(gwas_study_info$PUBMED_ID)
length(all_pmids)

```

### PMCIDs for these PMIDS from Europe PMC mapping file

```{r pmids_to_pmcids_stats, warning=FALSE}

converted_ids <- data.table::fread(here::here("output/fulltexts/pmid_to_pmcid_mapping.csv"))

print("Head of pmid to pmcid mapping data.frame:")
head(converted_ids)

print("Dimensions of pmid to pmcid mapping data.frame:")
dim(converted_ids)
length(all_pmids)

print("All pmids are in this data.frame, but some don't have pmcid mapping")

not_converted_pmids <-
converted_ids |>
  filter(pmcids == "")  |>
  pull(PMID)

print("Number of pmids without pmcid mapping:")
length(not_converted_pmids)

pmcids <-
converted_ids$pmcids |>
  unique()

pmcids <- pmcids[pmcids != ""]

print("Number of pmids with pmcid mapping:")
length(pmcids)

print("Percentage of pmids with pmcid:")
round(100 * length(pmcids) / length(all_pmids), digits = 2)

```


# Download from European PMC

```{r download_supplemental_files, eval = F}

library(httr)
library(here)
library(tools)

download_pmc_supplements <- function(pmcid,
                                     out_dir = here::here("output/supplement")) {
  
  # Create PMCID-specific subdirectory
  pmcid_dir <- file.path(out_dir, pmcid)
  
  # Check if directory already exists and has files
  if (dir.exists(pmcid_dir) && length(list.files(pmcid_dir)) > 0) {
    message("Supplementary files for ", pmcid, " already exist. Skipping.")
    return(invisible(list(pmcid = pmcid, status = "skipped", files = list.files(pmcid_dir))))
  }
  
  # Create directory if it doesn't exist
  dir.create(pmcid_dir, recursive = TRUE, showWarnings = FALSE)
  
  # Build URL
  sup_url <- paste0(
    "https://www.ebi.ac.uk/europepmc/webservices/rest/",
    pmcid,
    "/supplementaryFiles"
  )
  
  message("Fetching supplementary files for ", pmcid, "...")
  
  # Save zip to a temp file
  zip_path <- file.path(pmcid_dir, 
                        paste0(pmcid, "_supplements.zip"))
  resp <- GET(sup_url, 
              write_disk(zip_path, overwrite = TRUE))
  
  if (status_code(resp) != 200) {
    warning("Failed to retrieve supplements for ", pmcid,
            ". HTTP status: ", status_code(resp))
    unlink(pmcid_dir, recursive = TRUE)
    return(invisible(list(pmcid = pmcid, status = "failed", files = NULL)))
  }
  
  # Unzip into the PMCID folder
  unzip_result <- tryCatch({
    unzip(zip_path, exdir = pmcid_dir)
  }, error = function(e) {
    warning("Failed to unzip supplements for ", pmcid, ": ", e$message)
    NULL
  })
  
  # Remove the zip file after extraction
  file.remove(zip_path)
  
  if (is.null(unzip_result)) {
    return(invisible(list(pmcid = pmcid, status = "unzip_failed", files = NULL)))
  }
  
  extracted_files <- list.files(pmcid_dir, recursive = TRUE, full.names = TRUE)
  message("Extracted ", length(extracted_files), " file(s) to ", pmcid_dir)
  
  return(invisible(list(
    pmcid  = pmcid,
    status = "success",
    dir    = pmcid_dir,
    files  = extracted_files
  )))
}


# Batch across multiple PMCIDs
results <- lapply(pmcids, function(id) {
  tryCatch(
    download_pmc_supplements(id),
    error = function(e) {
      message("Error processing ", id, ": ", e$message)
      list(pmcid = id, status = "error", files = NULL)
    }
  )
})

```

```{bash unzip_europe_pmc, eval = F}

find output/supplement -name "*.zip" | while read f; do
  unzip -o "$f" -d "$(dirname "$f")"
done

```

```{bash delete_empty_dirs_europe_pmc, eval = F}

find output/supplement -type d -empty -delete

```


# Download from NCBI Cloud AWS

```{r pmcids_to_get_aws}

# check how many pmcids I have downloaded supplementary materials for
sup_dir <- here::here("output/supplement")

folders <- list.dirs(sup_dir, 
                     full.names = TRUE, 
                     recursive = FALSE)

folders_with_files <- folders[
  sapply(folders, function(d) length(list.files(d, recursive = TRUE)) > 0)
]

not_retrieved_pmcids <- setdiff(pmcids, basename(folders_with_files))

print("PMCIDs for which I could not retrieve supplements from European PMC:")
print(length(not_retrieved_pmcids))

writeLines(
  not_retrieved_pmcids,
  here::here("output/supplement/selected_pmcids.txt")
)

```


```{bash, eval = FALSE}

bash code/extract_text/download_pmc_supplements_aws.sh --file output/supplement/selected_pmcids.txt

```


```{bash delete_empty_dirs_ncbi_aws, eval = FALSE}

find output/supplement -type d -empty -delete

```

## Check how many supplements I have downloaded, and what file types they are

```{r check_supplemental_files}

# check how many pmcids I have downloaded supplementary materials for
sup_dir <- here::here("output/supplement")

folders <- list.dirs(sup_dir, 
                     full.names = TRUE, 
                     recursive = FALSE)

folders_with_files <- folders[
  sapply(folders, function(d) length(list.files(d, recursive = TRUE)) > 0)
]

message(length(folders_with_files), 
        " / ", 
        length(folders), 
        " folders contain at least one file")


# check extensions of files downloaded
all_files <- unlist(lapply(folders_with_files, 
                           list.files, 
                           recursive = TRUE, 
                           full.names = TRUE))

file_extensions <- tools::file_ext(all_files) |>
  table() |>
  as.data.frame() |>
  dplyr::arrange(desc(Freq))

print("File extensions of downloaded supplementary files:")
print(file_extensions)

folders_with_relevant_files <- folders[
  sapply(folders, function(d) length(list.files(d, 
                                                recursive = TRUE,
                                                pattern = "*.pdf|*.docx|*.doc|*.xls|*.XLSX")) > 0)
]

message(length(folders_with_relevant_files), 
        " / ", length(folders), 
        " folders contain at least one relevant file type (pdf, docx, doc, xls, xlsx)")

```

# Convert sup materials

```{r}

folders <- list.dirs(sup_dir, 
                     full.names = TRUE, 
                     recursive = FALSE)

folders_with_relevant_files <- folders[
  sapply(folders, function(d) length(list.files(d, 
                                                recursive = TRUE,
                                                pattern = "*.pdf|*.docx|*.doc|*.DOCX|*.txt|*.xls|*.XLSX")) > 0)
]

print("Number of folders with at least one relevant file type (pdf, docx, doc, xls, xlsx):")
print(length(folders_with_relevant_files))

no_relevant_files <- folders[!c(folders %in% folders_with_relevant_files)]

print("File extensions of folders without relevant file types:")
list.files(no_relevant_files, 
           recursive = TRUE, 
           full.names = TRUE) |>
  tools::file_ext()  |>
  unique() 


```

```{r convert_supplemental_materials}

# what sup files to convert?

sup_files <- list.files(sup_dir,
                        recursive = TRUE,
                        pattern = "*.pdf|*.docx|*.doc|*.xls|*.XLSX"
                        )


sup_files <- paste0("output/supplement/", sup_files)

writeLines(sup_files, 
           here::here("output/supplement/supplemental_files_to_convert.txt"))


```

```{bash, eval = FALSE}

./code/extract_text/convert_supplemental_materials.sh 

```

# Archived / not run

## FAIR-SMART links from NCBI

```{r get_supplemental_files_ncbi, eval = F}

get_fair_smart_links <- function(pmid) {
  
  url <- paste0("https://www.ncbi.nlm.nih.gov/research/bionlp/RESTful/supplmat_query_request.cgi/bioc_xml/", 
                pmid,
                "/CPC",)
  
  response <- GET(url)
  
  if (status_code(response) == 200) {
    
    collected_response <- content(response, as = "text", encoding = "UTF-8")
    
    sup_link = xml2::xml_text(xml2:::read_xml(collected_response))
    
    if(grepl("No result can be found", sup_link)) {
      warning("No relevant supplements found for PMID: ", pmid)
      return(NULL)
    }
    
    supplement_links <- stringr::str_extract_all(sup_link, "https?://[^\\s]+")
    
    supplement_links <- unlist(supplement_links)
    
    # print("Supplement links found:")
    # print(supplement_links)
    
    return(supplement_links)
    
  } else {
    warning("Failed to retrieve supplements for PMID: ", pmid, 
            ". HTTP status: ", status_code(response))
    return(NULL)
  }
}

sup_links <- purrr::map(all_pmids, 
                        get_fair_smart_links)

names(sup_links) <- all_pmids


download_relevant_sup <- function(sup_link,
                                  out_dir,
                                  pmid) {
  
  if(is.null(sup_link)) {
    return(NULL)
  }
  
  out_path <- here::here(paste0(out_dir,"/", 
                                pmid, 
                                "_supplement_", basename(sup_link))
                         )
  
  resp <- GET(sup_link, 
              write_disk(out_path, overwrite = TRUE))
  
  if (status_code(resp) != 200) {
    warning("Failed to download supplement from ", sup_link,
            ". HTTP status: ", status_code(resp))
    return(NULL)
  }
  
  message("Downloaded supplement from ", sup_link, " to ", out_path)
  return(out_path)
}

purrr::iwalk(sup_links,
             ~download_relevant_sup(.x, 
                                    out_dir = here::here("output/supplement/ncbi"), 
                                    pmid = .y)
             )


```

## Download supplements from Elsevier

```{r elsevier_supplements, eval = F}


elseiver_xmls <- list.files(here::here("output/fulltexts/elsevier/elsevier_xml/"), full.names = TRUE)


xml <- xml2::read_xml(here::here("output/fulltexts/elsevier/elsevier_xml/17223258.xml")) 

use_local <- function(x) {
  paste0(".//*[local-name()='", x, "']")
}

get_elsevier_supplement_links <- function(xml_file_path,
                                          api_key    = Sys.getenv("ELSEVIER_API_KEY"),
                                          out_dir = here::here("output/supplement/elsevier")) {
  
  
  # Build auth headers
  headers <- c("X-ELS-APIKey" = api_key)
  
  xml <- xml2::read_xml(xml_file_path)
  doc <- xml_find_first(xml, use_local("xocs:doc"))
  meta <- xml_find_first(doc, use_local("xocs:meta"))
  
  attachments <- xml_find_all(meta, 
                                use_local("xocs:attachments")
                                )
  supplement_eids <- xml_find_all(attachments, 
                                   use_local("xocs:attachment-eid")
                                   ) |> 
    xml_text()
  
  supplement_eids <- supplement_eids[!grepl("main", 
                                              supplement_eids, 
                                              ignore.case = TRUE)]
  
# sup_names <- xml_text(xml_find_all(attachments, use_local("attachment-eid")))
  # track results for all files
  results <- list()

if(length(supplement_eids) > 0) {
  
  #@browser()
  
   out_dir = paste0(out_dir, 
                    "/",
                    tools::file_path_sans_ext(basename(xml_file_path))
                    )
   
   # make directory for this article
   dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)
  
  for(eid in supplement_eids) {
    
    message("Supplement link: ", eid)
    
    out_path  <- paste0(out_dir, "/", basename(eid))
    
    api_url  = paste0("https://api.elsevier.com/content/object/eid/", eid)
    
    file_resp <- GET(api_url,
                     add_headers(.headers = headers),  # <-- auth headers
                write_disk(out_path,
                           overwrite = TRUE)
                )
    
    http_status  <- status_code(file_resp)
    file_size_kb <- file.size(out_path) / 1024
      
    is_html_error <- tryCatch({
        first_bytes <- readLines(out_path, n = 1, warn = FALSE)
        grepl("^<!DOCTYPE|^<html", first_bytes, ignore.case = TRUE)
      }, error = function(e) FALSE)
    
    status <- dplyr::case_when(
        http_status == 403               ~ "permission_denied",
        http_status == 401               ~ "unauthorized",
        http_status != 200               ~ paste0("http_error_", http_status),
        is_html_error                    ~ "permission_denied_html_response",
        file_size_kb < 5                 ~ "suspiciously_small",
        TRUE                             ~ "success"
      )
    
    if (status != "success") {
        warning(sprintf("  [%s] %s (%.1f KB, HTTP %d)",
                        status, basename(eid), file_size_kb, http_status))
        # Remove the bad file so it doesn't look like a successful download
        file.remove(out_path)
      } else {
        message(sprintf("  [OK] %s (%.1f KB)", basename(eid), file_size_kb))
      }
    
    # results[[basename(sup)]] <- list(
    #     url         = sup,
    #     out_path    = out_path,
    #     http_status = http_status,
    #     file_size_kb = round(file_size_kb, 2),
    #     status      = status
    #   )
    # 
    
  }
  

} else {
  message("No supplement links found in ", xml_file_path)
}

}

purrr::map(elseiver_xmls, get_elsevier_supplement_links)

```


#### More

