---
title: "Get dbGAP ids for studies"
author: "Isobel Beasley"
date: "2025-10-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r setup-libraries, message=FALSE}

library(dplyr)
library(data.table)
library(rentrez)
library(purrr)
library(stringr)

```

# Get from full text 

## Functions to extract dbGaP (and other archive) ids from full text

```{r get_full_text_files_fns}

# patterns to extract: 
# get dbgap ids
# get EGA ids
# get JGAS
# get PRJEB / PRJNA ids
patterns <- list(
  dbgap = "\\bphs\\d+(?:\\.v\\d+)?(?:\\.p\\d+)?",
  ega   = "\\bEGA[CDFS]\\d+",
  jgas  = "\\bJGAS\\d+|JGAD\\d+",
  prj   = "\\bPRJEB\\d+|PRJNA\\d+"
)

grep_pattern <- paste(patterns, collapse = "|")

library(tokenizers)

# get sentences with dbgap / ega ids etc. in them
get_grep_sentences <- function(file_path,
                               grep_pattern) {
  
  #browser()
  
  
  txt_in <- suppressWarnings(xml2::read_xml(file_path))
  
  # txt_in <- tryCatch(
  #   xml2::read_xml(file_path),
  #   error = function(e) {
  #     print(paste0("error with ", file_path));
  #     return(NULL)
  #   }
  # )
  
  txt_in <- xml2::read_xml(file_path)
  
  get_patterns <- c(".//*[local-name()='body']//*[local-name()='p']",
                    ".//*[local-name()='abstract']//*[local-name()='p']",
                    ".//*[local-name()='back']//*[local-name()='p']",
                    ".//*[local-name()='fig']//*[local-name()='caption']",
                    ".//*[local-name()='table-wrap']//*[local-name()='p']",
                    ".//*[local-name()='sec' and @sec-type='data-availability']//*[local-name()='p']",
                    ".//*[local-name()='supplementary-material']//*[local-name()='p']",
                    ".//passage/text")
  
  canidate_text <- txt_in |>
                   xml2::xml_find_all(paste(get_patterns, collapse = " | ")) |>
                   # xml2::xml_find_all("//body//p | 
                   #                     //abstract//p |
                   #                    //back//p  |
                   #                    //fig//caption |
                   #                    //table-wrap//p |
                   #                    //sec[@sec-type='data-availability']//p
                   #                    //supplementary-material") |>
                   #xml2::xml_find_all("//body|//back") |>
                   xml2::xml_text()
  
  
  if(length(canidate_text) == 0) {
    print(paste0("No candidate text found in ", file_path))
    return(character(0))
  }
  
  # txt_in <- readLines(file_path,
  #                     warn = FALSE,
  #                     encoding = "UTF-8")
  # tokenize into sentences
  
  
  sentences <- unlist(
  tokenize_sentences(paste(canidate_text, collapse = " "))
)

grep_sentences <- sentences[grepl(grep_pattern, sentences)]

  # sentences <- tokenize_sentences(paste(canidate_text, collapse = "\n"))
  # 
  # grep_sentences <- sentences[grepl(grep_pattern, sentences)]

  # grep_sentences <- tokenize_sentences(paste(candidate_lines, 
  #                                            collapse = "\n"))
  # 
  # sentences <- tokenize_sentences(paste(txt_in, collapse = " "))
  # 
  # sentences <- unlist(tokenize_sentences(txt_in))
  # 
  # dbgap_sentences = sentences[grepl(grep_pattern, sentences)]

  return(grep_sentences)
}


```

## Text file paths to load 

```{r get_full_text_file_paths}

full_text_files <-
  list.files(here::here("output/fulltexts"), 
             recursive = T, 
             pattern = "\\.xml$") 

# some duplicates come from converted xml files 
# i.e. from propriety wiley and elsevier xml
# let's keep only the converted files:
full_text_files <- 
  grep("elsevier/elsevier_xml|wiley/wiley_xml", 
       full_text_files, 
       value = TRUE,
       invert = TRUE)


# get pmcids of these files
all_ids <-
  full_text_files |>
  gsub(pattern = ".*/", replacement = "") |>
  gsub(pattern = "\\.xml$", replacement = "") |>
  gsub(pattern = "_bioc", replacement = "") |>
  gsub(pattern = "\\.pdf.tei", replacement = "")

# some duplicates come from ncbi and europe pmc
# let's keep the europe pmc files
duplicate_ids <-
data.frame(ids = all_ids) |>
  group_by(ids) |>
  summarise(count = n()) |>
  filter(count > 1) |>
  pull(ids)

grep(full_text_files, 
     pattern = duplicate_ids[1],#paste(duplicate_ids, collapse = "|"),
     value = T)

duplicates_to_remove <-
grep(full_text_files, 
     pattern = paste(duplicate_ids, collapse = "|"),
     value = T) |>
  grep(pattern = "ncbi", value = T)

full_text_files <- full_text_files[!c(full_text_files %in% duplicates_to_remove)]

# convert to file paths 
full_text_files <- here::here(paste0("output/fulltexts/", 
                                     full_text_files)
                              )

# get pmids and pmcids of relevant files to load
relevant_ids <- data.table::fread(here::here("output/fulltexts/pmid_to_pmcid_mapping.csv")) 

relevant_ids <- c(relevant_ids$PMID, relevant_ids$pmcids)

all_text_ids <-   full_text_files |>
  gsub(pattern = ".*/", replacement = "") |>
  gsub(pattern = "\\.xml$", replacement = "") |>
  gsub(pattern = "_bioc", replacement = "") |>
  gsub(pattern = "\\.pdf.tei", replacement = "")

text_files_to_get <-
data.frame(full_text_files = full_text_files,
           all_text_ids = all_text_ids) |>
  filter(all_text_ids %in% relevant_ids)

full_text_files <- text_files_to_get$full_text_files

print("Number of full text files to load:")
print(length(full_text_files))

```

## Extract dbGaP accessions from full text

```{r get_dbgap_from_full_text, warning=FALSE}

pmcid_dbgap_sentences <- 
purrr::map(full_text_files,
           ~get_grep_sentences(.x, 
                               grep_pattern = grep_pattern
                               )
           ) 

all_text_ids <-   full_text_files |>
  gsub(pattern = ".*/", replacement = "") |>
  gsub(pattern = "\\.xml$", replacement = "")  |>
  gsub(pattern = "_bioc", replacement = "") |>
  gsub(pattern = "\\.pdf.tei", replacement = "")

names(pmcid_dbgap_sentences) <- all_text_ids

keep_pmcid_dbgap_sentences <-
  purrr::discard(
pmcid_dbgap_sentences,
~rlang::is_empty(.x)
) 


# Convert to data frame
sentences_df <- 
  purrr::imap(keep_pmcid_dbgap_sentences,
              ~data.frame(pmcid = .y,
                          sentence = unique(.x),
                        stringsAsFactors = FALSE)
              )|>
  dplyr::bind_rows()

# Extract all dbGaP IDs from the sentences
sentences_df <- sentences_df |>
  mutate(
    dbgap_id = str_extract_all(sentence, patterns[["dbgap"]]),
                               #"phs\\d+(\\.v\\d+)?(\\.p\\d+)?"),
    ega_id = str_extract_all(sentence, patterns[["ega"]]),
                             #"EGA[CDFS]\\d+"),
    jgas_id = str_extract_all(sentence, patterns[["jgas"]]),
                              #"JGAS\\d+|JGAD"),
    prj_id = str_extract_all(sentence, patterns[["prj"]]),
                             #"PRJEB\\d+|PRJNA\\d+")
  )
             
sentences_df = 
  sentences_df |>
  distinct()

print("Number of unique PMCIDs with dbgap / ega ids found in full text:")
sentences_df$pmcid |> unique() |> length()

# number of pubmed ids with dbgap ids found
print("Number of unique PMCIDs with dbgap ids found in full text:")
sentences_df %>%
  rowwise() %>%
  filter(!rlang::is_empty(dbgap_id)) %>%
  filter(length(dbgap_id) > 0) %>%
  ungroup() %>%
  pull(pmcid) %>%
  unique() %>%
  length()

print("Number of unique PMCIDs with EGA ids found in full text:")
sentences_df %>%
  rowwise() %>%
  filter(!rlang::is_empty(ega_id)) %>%
  filter(length(ega_id) > 0) %>%
  pull(pmcid) %>%
  unique() %>%
  length()

convert_pmid_df  <- data.table::fread(here::here("output/fulltexts/pmid_to_pmcid_mapping.csv")) |>
  rename(PMCID = pmcids)

# add pmid column
sentences_df =
  left_join(sentences_df,
            convert_pmid_df |> 
              dplyr::select(pmcid = PMCID,
                            pmid = PMID),
            by = "pmcid"
            ) |>
  mutate(pmid = ifelse(is.na(pmid), 
                      pmcid,
                      pmid
                      )
  ) |> 
  mutate(pmcid = ifelse(grepl("PMC", pmid),
                        pmid,
                        pmcid)
         ) |>
  mutate(pmcid = ifelse(grepl("PMC", pmcid), 
                        pmcid, 
                        ""
                        )
  )


sentences_df = 
  left_join(sentences_df |> 
              select(-pmcid) |> 
              mutate(pmid = as.integer(pmid)),
            convert_pmid_df |> 
              dplyr::select(pmcid = PMCID,
                            pmid = PMID),
            by = "pmid"
            ) 

# typo in text for PMC8486151
# replace  "phs000" with 
# phs000.56.v1.p1

sentences_df |> filter(pmcid == "PMC8486151") |> pull(dbgap_id)

fix_dbgap_ids <- 
  c(sentences_df |> filter(pmcid == "PMC8486151") |> pull(dbgap_id),
    "phs00056.v1.p1"
    ) |>
  unlist()

fix_dbgap_ids <- fix_dbgap_ids[fix_dbgap_ids != "phs000"]

sentences_df <-
  sentences_df |>
  mutate(dbgap_id = ifelse(pmcid == "PMC8486151",
                          list(fix_dbgap_ids),
                          dbgap_id
                          )
         )
         

data.table::fwrite(sentences_df,
                   here::here("output/gwas_cat/gwas_study_dbgap_ega_sentences.csv")
                   )
                   
                   
```

                  
# Get from Europe PMC list

```{r get-dbgap-ids-europe-pmc}

dbgap_info <-
data.table::fread(here::here("data/europe_pmc/dbgap.csv")) 

pmids <- convert_pmid_df$PMID 

pmcids <- convert_pmid_df$PMCID

relevant_pmid_dbgaps = dbgap_info |>
  filter(EXTID %in% pmids) |>
  distinct() 

print("Number of unique PMIDs with dbgap ids found in Europe PMC:")
relevant_pmid_dbgaps |>
  pull(EXTID) |>
  unique() |>
  length()

relevant_pmcids_dbgaps = dbgap_info |>
  filter(PMCID %in% pmcids) |>
  filter(PMCID != "") |>
  distinct()

print("Number of unique PMCIDs with dbgap ids found in Europe PMC:")
relevant_pmcids_dbgaps |>
  pull(PMCID) |>
  unique() |>
  length()

# overlap between pmid and pmcid dbgap annotations in europe pmc
print("Number of unique PMIDs with dbgap ids found in Europe PMC that also have dbgap ids found in full text:")
unique(sentences_df$pmid) %in% unique(relevant_pmid_dbgaps$EXTID) |>
  sum() 

# overlap between pmid and pmcid dbgap annotations in europe pmc
# 49 / 57 = 86% of pmids with dbgap ids in europe pmc annotations are also found in full text

print("Number of unique PMIDs with dbgap ids found in full text, but not in Europe PMC:")
setdiff(unique(sentences_df$pmid), 
        unique(relevant_pmid_dbgaps$EXTID)) |> length()

# 101 / 158 = 64% of pmids with dbgap ids found in full text are not found in europe pmc annotations

print("Number of unique PMIDs with dbgap ids found in Europe PMC that do not have dbgap ids found in full text:")
setdiff(unique(relevant_pmid_dbgaps$EXTID), 
        unique(sentences_df$pmid)) |> length()

```

```{r compare-europe-pmc-full-text}

full_text_dbgaps = 
sentences_df  |>
  select(dbgap_id, pmid, pmcid) |>
  tidyr::unnest(dbgap_id) |>
  mutate(dbgap_id = str_trim(dbgap_id)) |>
  mutate(dbgap_id = str_extract(dbgap_id, "^phs\\d+")) |>
  # mutate(dbgap_id = str_remove(dbgap_id, "\\.v\\d+\\.p\\d+.*$")) |> 
  group_by(pmid, pmcid) |>
  summarise(dbgap_full_text = str_flatten(unique(sort(dbgap_id)), 
                          collapse = ", ", 
                          na.rm = TRUE)) |>
#  tidyr::nest(dbgap_full_text = dbgap_id) |>
  mutate(pmid = as.character(pmid))


europe_pmc_dbgaps =
relevant_pmid_dbgaps |>
  select(dbgap_id = dbgap,
         pmcid = PMCID,
         pmid = EXTID) |>
    group_by(pmid, pmcid) |>
  summarise(dbgap_epmc = str_flatten(unique(sort(dbgap_id)), 
                          collapse = ", ", 
                          na.rm = TRUE)) 
 # tidyr::nest(dbgap_epmc = dbgap_id) 

joined <-
full_join(full_text_dbgaps,
          europe_pmc_dbgaps,
          by = c("pmid", "pmcid")
          )

# includes more dbgap ids from full text than europe pmc annotations
joined |> 
  filter(!is.na(dbgap_full_text) & !is.na(dbgap_epmc)) |> 
  filter(dbgap_full_text != dbgap_epmc) 

joined |>
  filter(!is.na(dbgap_full_text) & is.na(dbgap_epmc))

# seems good extraction: 
# pubmed id: 39024449
# pubmed id: 26959717
# pubmed id: 31537701
# pubmed id: 37167549
# pubmed id: 30207284
# pubmed id: 33135175
# pubmed id: 37069358

joined |>
  filter(is.na(dbgap_full_text) & !is.na(dbgap_epmc))
```

# Get from Entrez

## Extract internal dbGAP ids

```{r get-dbgap-ids, eval = F}

Sys.getenv("ENTREZ_KEY")

get_internal_dbgap_ids = function(pubmed_id) {
  
  Sys.sleep(0.1)

dbgap_links <- entrez_link(
            dbfrom = "pubmed",
            db = "gap",
            id = pubmed_id
)

dbgap_ids <- unlist(dbgap_links$links$pubmed_gap)

if(is.null(dbgap_ids)){
  dbgap_ids = NA
}

return(data.frame(PUBMED_ID = pubmed_id,
                  DBGAP_ID = str_flatten(dbgap_ids,
                                         collapse = ",",
                                         na.rm = TRUE)
                  )
)

}

pubmed_dbgap_mapping <-
purrr::map(pmids,
           get_internal_dbgap_ids,
           .progress = TRUE
           ) |>
  dplyr::bind_rows()


```

### Save dbGaP id mapping

```{r save-dbgap-ids, eval = F}

data.table::fwrite(pubmed_dbgap_mapping,
                   here::here("output/gwas_cohorts/gwas_study_dbgap_ids.csv")
                   )

```

### Percentage of studies with dbGAP ids

```{r percent-with-dbgap-ids}

# load previously saved mapping
pubmed_dbgap_mapping <- data.table::fread(
  here::here("output/gwas_cohorts/gwas_study_dbgap_ids.csv")
)

n_with_dbgap = pubmed_dbgap_mapping |>
  dplyr::filter(DBGAP_ID != "") |>
  nrow()

print("Number of unique PMIDs with dbgap ids found via Entrez:")
n_with_dbgap

percent_with_dbgap = n_with_dbgap / nrow(pubmed_dbgap_mapping) * 100

print("Percentage of PMIDs with dbgap ids found via Entrez:")
percent_with_dbgap

```

## Extract dbGAP study accessions

```{r extract-dbgap-accessions, warning=FALSE, message=FALSE, eval = F}

entrez_dbgap_ids = pubmed_dbgap_mapping$DBGAP_ID

entrez_dbgap_ids = entrez_dbgap_ids[entrez_dbgap_ids != ""]

entrez_dbgap_ids = entrez_dbgap_ids |>
  strsplit(",") |>
  unlist() |>
  unique()

print("Number of unique internal dbgap ids to retrieve accessions for:")
print(length(entrez_dbgap_ids))


get_dbgap_accession = function(internal_dbgap_id) {
  
  
  summary = entrez_summary(db = "gap", 
                           id = internal_dbgap_id)
  
  
  accession = summary$d_study_results$d_study_id
  
  return(data.frame(
    INTERNAL_DBGAP_ID = internal_dbgap_id,
    DBGAP_ACCESSION = str_flatten(accession, 
                                  collapse = ",", 
                                  na.rm = TRUE)
  )
  )

} 

dbgap_accession_mapping <-
  purrr::map(entrez_dbgap_ids,
             get_dbgap_accession,
             .progress = TRUE
             ) |>
  dplyr::bind_rows()

```


### Save dbGaP accession mapping

```{r save-dbgap-accessions, eval = F}

  data.table::fwrite(
    dbgap_accession_mapping,
     here::here("output/gwas_cohorts/gwas_study_dbgap_accessions.csv")
                     )
  

```

### Percentage of internal dbGaP ids for which we could not retrieve accessions

```{r percent-with-dbgap-accessions}

dbgap_accession_mapping <- data.table::fread(
  here::here("output/gwas_cohorts/gwas_study_dbgap_accessions.csv")
)

# a lot of internal dbgap ids are not able be retrieve study accessions
# sad
# ? perhaps because they are old dbgap versions? 
print("Number of internal dbgap ids for which we could not retrieve accessions:")
dbgap_accession_mapping |> 
  filter(DBGAP_ACCESSION == "") |>
  nrow()


pubmed_dbgap_mapping =
  pubmed_dbgap_mapping |>
  tidyr::separate_longer_delim(cols = "DBGAP_ID", 
                               delim = ",")

dbgap_accession_mapping = 
  dbgap_accession_mapping |>
  tidyr::separate_longer_delim(cols = "INTERNAL_DBGAP_ID", 
                               delim = ",")

dbgap_accession_mapping = 
left_join(pubmed_dbgap_mapping, 
     dbgap_accession_mapping,
     by = c("DBGAP_ID" = "INTERNAL_DBGAP_ID")
     ) 

print("Number of unique PMIDs with dbgap accessions found via Entrez:")
dbgap_accession_mapping |>
  filter(DBGAP_ACCESSION != "" & DBGAP_ID != "") |>
  pull(PUBMED_ID) |>
  unique() |>
  length()

```

### Save combined dbGaP mapping

```{r save-combined-dbgap-mapping, eval = F}

combined_dbgap_pubmed_mapping = 
  dbgap_accession_mapping |>
  filter(DBGAP_ACCESSION != "" & DBGAP_ID != "") |>
  select(PUBMED_ID, DBGAP_ACCESSION) |>
  distinct()

data.table::fwrite(
  combined_dbgap_pubmed_mapping,
  here::here("output/gwas_cohorts/gwas_study_combined_dbgap_pubmed_mapping.csv")
)

```

## Compare dbgap retrieved via extracting full text vs via entrez

```{r compare-dbgap-sources}

combined_dbgap_pubmed_mapping <- data.table::fread(
  here::here("output/gwas_cohorts/gwas_study_combined_dbgap_pubmed_mapping.csv")
)

entrez_dbgaps <- 
  combined_dbgap_pubmed_mapping |>
  rename(pmid = PUBMED_ID,
         dbgap_id = DBGAP_ACCESSION
         ) |>
  mutate(dbgap_id = str_extract(dbgap_id, "^phs\\d+")) |>
  filter(dbgap_id != "") |>
  group_by(pmid) |>
  summarise(dbgap_annotation = str_flatten(unique(sort(dbgap_id)), 
                          collapse = ", ", 
                          na.rm = TRUE)) |>
  mutate(pmid = as.character(pmid))


joined_dbgaps =
full_join(entrez_dbgaps,
          full_text_dbgaps,
          by = "pmid")

print("Number of PMIDs with dbgap annotations from Entrez that do not match dbgap ids found in full text:")
joined_dbgaps |>
  filter(!is.na(dbgap_annotation) & !is.na(dbgap_full_text)) |>
  filter(dbgap_annotation != dbgap_full_text)
 
print("Number of PMIDs with dbgap annotations from Entrez that match dbgap ids found in full text:")
joined_dbgaps |>
  filter(!is.na(dbgap_annotation) & !is.na(dbgap_full_text)) |>
  filter(dbgap_annotation == dbgap_full_text) |>
  nrow()

print("Number of PMIDs with dbgap annotations from Entrez but no dbgap ids found in full text:")
joined_dbgaps |>
  filter(!is.na(dbgap_annotation) & is.na(dbgap_full_text)) |>
  nrow()

print("Number of PMIDs with dbgap found in full text but not in Entrez annotations:")
joined_dbgaps |>
  filter(is.na(dbgap_annotation) & !is.na(dbgap_full_text)) |>
  nrow()

```

# Combine and save

```{r combine-and-save}

# europe_pmc_dbgaps
# entrez_dbgaps 
# full_text_dbgaps

full_text_dbgaps = 
sentences_df  |>
  select(dbgap_id, pmid, pmcid) |>
  tidyr::unnest(dbgap_id) |>
  mutate(dbgap_id = str_trim(dbgap_id)) |>
  # mutate(dbgap_id = str_extract(dbgap_id, "^phs\\d+")) |>
  # mutate(dbgap_id = str_remove(dbgap_id, "\\.v\\d+\\.p\\d+.*$")) |> 
  group_by(pmid, pmcid) |>
  summarise(dbgap_full_text = str_flatten(unique(sort(dbgap_id)), 
                          collapse = ", ", 
                          na.rm = TRUE)) |>
#  tidyr::nest(dbgap_full_text = dbgap_id) |>
  mutate(pmid = as.character(pmid))


europe_pmc_dbgaps =
relevant_pmid_dbgaps |>
  select(dbgap_id = dbgap,
         pmcid = PMCID,
         pmid = EXTID) |>
    group_by(pmid, pmcid) |>
  summarise(dbgap_epmc = str_flatten(unique(sort(dbgap_id)), 
                          collapse = ", ", 
                          na.rm = TRUE)) 

entrez_dbgaps <- 
  combined_dbgap_pubmed_mapping |>
  rename(pmid = PUBMED_ID,
         dbgap_id = DBGAP_ACCESSION
         ) |>
  # mutate(dbgap_id = str_extract(dbgap_id, "^phs\\d+")) |>
  filter(dbgap_id != "") |>
  group_by(pmid) |>
  summarise(dbgap_annotation = str_flatten(unique(sort(dbgap_id)), 
                          collapse = ", ", 
                          na.rm = TRUE)) |>
  mutate(pmid = as.character(pmid))

combined_annotations <-
full_join(full_text_dbgaps,
          europe_pmc_dbgaps,
          by = c("pmid", "pmcid")
          ) |>
  full_join(entrez_dbgaps,
            by = c("pmid")
            ) |>
  tidyr::separate_longer_delim(cols = c(dbgap_full_text),
                               delim = ", ")|>
  tidyr::separate_longer_delim(cols = c(dbgap_epmc),
                               delim = ", ")|>
  tidyr::separate_longer_delim(cols = c(dbgap_annotation),
                               delim = ", ")|>
  tidyr::pivot_longer(cols = c(dbgap_full_text, dbgap_epmc, dbgap_annotation),
               names_to = "source",
               values_to = "dbgap_id") |>
  filter(!is.na(dbgap_id) & dbgap_id != "") |>
  mutate(source = stringr::str_remove_all("dbgap_", string = source)) |>
  mutate(source = ifelse(source == "annotation", "entrez", source)) |>
  distinct()


combined_annotations$pmid |> unique() |> length()


n_pubmeds_per_dbgap =
combined_annotations |>
  mutate(broad_dbgap = str_extract(dbgap_id, "^phs\\d+")) |>
  select(pmid, broad_dbgap) |>
  distinct() |>
  group_by(broad_dbgap) |>
  summarise(n_pubmeds = n()) |>
  arrange(desc(n_pubmeds))

n_pubmeds_per_dbgap |>
  head(
  )

n_pubmeds_per_dbgap$n_pubmeds |> summary()


```

### Are there any dbgap ids not in our cohort description mapping?

```{r new_dbgap_ids}

cohort_dbgap_mapping <- readxl::read_xlsx(here::here("data/cohort/cohort_desc.xlsx"),
                                 sheet = 1)
annotated_dbgap_ids =
  combined_annotations |>
  pull(dbgap_id) |>
  str_trim() |>
  str_extract(pattern = "^phs\\d+") |>
  unique()

cohort_desc_dbgap_ids = 
  cohort_dbgap_mapping |>
  pull(dbGaP) |>
  strsplit(",") |>
  unlist() |>
  str_trim() |>
  str_extract(pattern = "^phs\\d+") |> 
  unique()

cohort_desc_dbgap_ids_p2 = 
    cohort_dbgap_mapping |>
  pull(old_dbGaP) |>
  strsplit(",") |>
  unlist() |>
  str_trim() |>
  str_extract(pattern = "^phs\\d+") |> 
  unique()

cohort_desc_dbgap_ids <- unique(c(cohort_desc_dbgap_ids, cohort_desc_dbgap_ids_p2))

not_found_dbgap_ids <- annotated_dbgap_ids[!c(annotated_dbgap_ids %in% cohort_desc_dbgap_ids)]

not_found_dbgap_ids  <- sort(not_found_dbgap_ids)

not_found_dbgap_ids |> length()

not_found_dbgap_ids

```


```{r get_cohort_names}

get_dbgap_cohort_names <- 
cohort_dbgap_mapping |>
  tidyr::separate_longer_delim(cols = c(dbGaP),
                               delim = ",")  |>
  tidyr::separate_longer_delim(cols = c(old_dbGaP),
                               delim = ",")  |>
  tidyr::pivot_longer(cols = c(dbGaP, old_dbGaP),
               names_to = "source",
               values_to = "dbgap_id") |>
  select(cohort, dbgap_id) |>
  mutate(dbgap_id = str_trim(dbgap_id)) |>
  mutate(broad_dbgap = str_extract(dbgap_id, "^phs\\d+")) |>
  filter(!is.na(dbgap_id)) |>
  select(-dbgap_id) |>
  distinct()

get_dbgap_cohort_names |>
  group_by(broad_dbgap) |>
  summarise(n = n(),
            cohorts = stringr::str_flatten(cohort, collapse = ", ", na.rm = T)) |>
  filter(n > 1)

get_dbgap_cohort_names <- 
  get_dbgap_cohort_names |>
  group_by(broad_dbgap) |>
  summarise(cohort = stringr::str_flatten(cohort, collapse = "|", na.rm = T))

dbgap_cohort_names =
left_join(
combined_annotations |> 
  mutate(broad_dbgap = str_extract(dbgap_id, "^phs\\d+")),
get_dbgap_cohort_names,
by = c("broad_dbgap")
) 


```


```{save-dbgap-cohort-names}

data.table::fwrite(dbgap_cohort_names,
                   here::here("output/gwas_cohorts/gwas_study_dbgap_cohort_names.csv")
                   )

```




