---
title: "Grep search patterns for annotations"
author: "Isobel Beasley"
date: "2026-06-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}

library(dplyr)
library(stringr)
library(httr)
library(jsonlite)

```

```{r gwas_cat_load_preprocess, message=FALSE}

# get only relevant disease studies
gwas_study_info <- data.table::fread(here::here("output/icd_map/gwas_study_gbd_causes.csv"))

gwas_study_info = gwas_study_info |>
  dplyr::rename_with(~ gsub(" ", "_", .x))

# filter out infectious diseases
gwas_study_info <- gwas_study_info |>
    dplyr::filter(!cause %in% c("HIV/AIDS",
                             "Tuberculosis",
                             "Malaria",
                             "Lower respiratory infections",
                             "Diarrhoeal diseases",
                             "Neonatal disorders",
                             "Tetanus",
                             "Diphtheria",
                             "Pertussis" ,
                             "Measles",
                             "Maternal disorders"))

gwas_study_info <-
  gwas_study_info |>
  filter(cause != "")

```


# Cohort context terms 

Combine multiple sources of terms for cohort context matching. 

#### Source of terms: GWAS Catalog Ancestry Metadata

GWAS ancestry data.frame contains: country of recruitment, ancestry label, population descriptors

```{r gwas_cat_load_ncestry_info}

gwas_ancest_info <-  data.table::fread(here::here("data/gwas_catalog/gwas-catalog-v1.0.3.1-ancestries-r2025-07-21.tsv"),
                           sep = "\t",
                           quote = "")

gwas_ancest_info = gwas_ancest_info |>
  dplyr::rename_with(~ gsub(" ", "_", .x))

```

##### Countries 

```{r get_countries_gwas_cat}

countries <- 
gwas_ancest_info$COUNTRY_OF_RECRUITMENT |>
  stringr::str_split("\\,") |>
  unlist() |>
  str_trim() |>
  unique() 
  
countries <- countries[countries != "NR"]
countries <- c(countries, 
               "United States", 
               "United Kingdom",
               "Korea",
               "America",
               "USA",
               "United States of America",
               "Latin America")


```

##### Population descriptors

```{r get_population_descriptors_gwas_cat}

population_descriptors <- 
  c(gwas_ancest_info$BROAD_ANCESTRAL_CATEGORY, gwas_ancest_info$ADDITIONAL_ANCESTRY_DESCRIPTION) |>
  str_replace_all("Recruitment in North, Central, and South America, Asia, and Europe", 
                  "North America, Central America, Asia, Europe") |>
  #stringr::str_split("\\,|;|heritage:\\|") |>
  stringr::str_split("\\,|;|heritage:|\\(|\\)|\\/|\\ - |\\: |\\band\\b") |>
  unlist() |>
  str_trim() 

# processing population descriptors to remove extra information and get more general terms to match cohort descriptions in abstracts (e.g. "European" instead of "European (non-Finnish)")
population_descriptors <- population_descriptors |>
  str_remove_all("\\.$")  |> # Escape the dot to make it literal
  str_remove_all("^and ") |> 
  str_remove_all(" \\(founder/genetic isolate\\)$|\\(founder/genetic isolate$") |>
  #str_remove_all("^[0-9+]% ") |>
  str_remove_all("^\\d+% ") |>  # \d is shorthand for [0-9]
  str_remove_all("^[‡†*]+") |> # Remove dagger symbols at start 
  str_remove_all(" - Korea Association Resource \\(KARE\\)$") |>
  str_remove_all("or unknown ancestry$|\\(middle eastern$") |>
  str_remove_all("\\(see Springelkamp 2017\\)") |>
  str_remove_all("parents & grandparents born in ") |>
  str_remove_all(" cell lines$| cell line$") |>
  str_remove_all(" cohort$") |>
  str_remove_all(" population$") |> 
  str_remove_all("^Erasmus Rucphen in |^Erasmus Rucphen") |>
  str_remove_all("^including ") |>
  str_remove_all("\\(Middle Eastern$") |>
  unique() 

# exclude population descriptors containing numbers 
population_descriptors <- population_descriptors[!str_detect(population_descriptors, "\\d")]

# exclude uninformative population descriptors 
uninform_pop_desc <- c("NR", "N.R", 
                       "Other", "other", 
                       "", "unspecified", 
                       "Mixed", "mixed", "Multiple", "multiple", 
                       "Split", "split") 

population_descriptors <- population_descriptors[!c(population_descriptors %in% uninform_pop_desc)]

# exclude disease descriptors that are not valid to match cohort descriptions in abstracts (e.g. "Asthma", "Type 2 diabetes")
disease_descriptors <- grep("cases|Bipolar|controls|Schizophrenia|disorder", 
                            population_descriptors, 
                            ignore.case = T,
                            value = T)

population_descriptors <- population_descriptors[!(population_descriptors %in% disease_descriptors
                                                     )]

# exclude cohort names 
cohort_names <- c("UKB", "UKBB", "DECODE", "Controls from UKBiobank", "UKBB and deCODE", 
                  "NHAPC", "GeneID-I", " Family", "KARE", "Mayo")

population_descriptors <- population_descriptors[!(population_descriptors %in% cohort_names)]

# exclude weird paper related terms 
grep("for|see|et al", population_descriptors, ignore.case = T, value = T)

population_descriptors <- population_descriptors[!grepl(population_descriptors, "for|see", ignore.case = T)]

# add directions
directions <- c("North", 
                "South",
                "East", 
                "West",
                "Central",
                "Southeast",
                "Northeast",
                "Northwest",
                "Southwest")

population_descriptors <- c(population_descriptors, 
                            "multiracial")

population_descriptors <- unique(c(population_descriptors, directions))

population_descriptors <- population_descriptors |> str_trim() |> unique()

```

#### Source of terms: hancestro ontology of human ancestry terms

```{r get_hancestro_terms}

# Get all terms in one go (if total < 500)
response <- GET("https://www.ebi.ac.uk/ols4/api/ontologies/hancestro/terms?size=1000")
data <- fromJSON(content(response, "text"), flatten = TRUE)
hancestro_terms <- data$`_embedded`$terms$label

# Remove terms with specific biobanks / datasets in brackets, as these are unlikely to be used in cohort descriptions in abstracts / texts
hancestro_terms <- str_remove_all(hancestro_terms, 
                                  " \\(SGDP\\)$| \\(1KGP\\)$| \\(HGDP\\)$| \\(GGVP\\)$")

# Remove obsolete terms
hancestro_terms <- grep("obsolete|obsolescence", 
                        hancestro_terms, 
                        value = TRUE, 
                        invert = TRUE)

# remove terms with specific regions in brackets, as these are unlikely to be used in cohort descriptions in abstracts / texts
hancestro_terms <- str_remove_all(hancestro_terms, pattern = "\\(Carmel\\)$|\\(pre1989\\)$|\\(Bergamo\\)$|\\(Negev\\)$|\\(Central\\)|\\(Caucasus\\)")

# replace underscores with spaces
hancestro_terms <- str_replace_all(hancestro_terms, 
                                   pattern = "_", 
                                   replacement = " ")

hancestro_terms <- unique(hancestro_terms)

# specific hancestro terms to exclude: 
terms_to_exclude <- c("ancestry category",
                      "ancestry status",
                      "BFO 0000006",
                      "Country",
                      "continent",
                      "continuant",
                      "curation status specification",
                      "denotator type",
                      "entity",
                      "ethnicity category",
                      "ethnicity descriptor",
                      "geographic location",
                      "geographic descriptor",
                      "geography-based population category",
                      "immaterial entity",
                      "independent continuant",
                      "!Kung",
                      "material entity",
                      "quality",
                      "reference population",
                      "region",
                      "organization",
                      "population",
                      "specifically dependent continuant",
                      "Thing",
                      "uncategorised population",
                      "undefined ancestry population")

hancestro_terms <- hancestro_terms[!c(hancestro_terms %in% terms_to_exclude)]

hancestro_terms <- 
hancestro_terms |>
  str_split("\\(|\\)") |>
  unlist() |>
  str_trim() |>
  unique() 

```

#### ? Other possible sources

Country R packages 
Population descriptors in: https://pmc.ncbi.nlm.nih.gov/articles/PMC8715140/#sec2


#### Terms to identify sentence identifying / describing cohort

```{r get_cohort_descriptor_terms}

participant_words <- c("adults", "adolescents", "children", "infants", "neonates",
                       "women", "men", "female", "male",
                       "trios", "families", "twins", "siblings", 
                        "cases", "probands", "affected", 
                        "controls", "unaffected", "healthy",                        
                       "genomes",  "individuals", "patients",  "participants", 
                       "subjects", "samples", "donors", "volunteers")

# words that indicate sample or cohort is being discussed:
sample_terms <- c(
  participant_words,
  # demographic / population descriptor terms
  "ancestry", "ancestral", "origin", "descent",  "lineage",  "heritage", 
  "admixed", "admixture",
  "ethnic", "ethnicity",  "multi-ethnic",
  "racial", "multi-racial", "multiracial", "race",
  "demographic", "self-identified", "self-reported",
  "age", "aged", "years",
  "founder", "isolate", "isolated", "population", "populations",
  
  # cohort/sample related terms
  "biobank", "cohort", "cohorts", "consortium", "consortia",
  "registry", "sample", "study", "trial",
  
  # enrollment related terms
   "enroll", "enrolled", "enrollment", "enrollees",
   "recruitment", "recruited", 
   "volunteer", "volunteered", 
   "sampling", "sampled", 
   "ascertained", "residing",
  
  # Types of study
  "case-control", "prospective", "longitudinal", "cross-sectional",  "nationwide",
  "meta-analysis", "pooled",
  "replication", "discovery", "validation",
  
  # Subsets 
  "subgroup", "subset", "subcohort"
  
)


```

#### Combine sources 

```{r combine_cohort_context_terms}

# combine terms from different sources, and add some extra terms based on looking at cohort descriptions in abstracts
cohort_context_terms <-
c(countries, 
  hancestro_terms, 
  population_descriptors,
  sample_terms) |> 
  unique() |>
  sort()


# more terms to exclude:
cohort_context_terms <- cohort_context_terms[!cohort_context_terms %in% c("San", "Split", "Han")]

# terms to remove ... "Qatar", "UK", "Taiwan", "Korean"
# cohort_context_terms <- cohort_context_terms[!cohort_context_terms %in% c("population", 
#                                                                           "mixed",
#                                                                           "split")]

# replace brackets with \\( & and \\)
cohort_context_terms <- cohort_context_terms |> 
  str_replace_all(pattern = "\\(", replacement = "\\\\(") |>
  str_replace_all(pattern = "\\)", replacement = "\\\\)")

# cohort_context_terms <- c(cohort_context_terms, 
#                           "UK\\b(?! Biobank)",
#                           "UK\\b\\(?! Biobank\\)",
#                           "Qatar\\b\\(?! Biobank\\)",
#                           "Taiwan\\b\\(?! Biobank\\)",
#                           "Korean\\b\\(?! Biobank\\)"
#                           )


# remove single character terms
cohort_context_terms <- cohort_context_terms[nchar(cohort_context_terms) > 2]

# sort by length of term (longest first) to match longest names first
cohort_context_terms <- cohort_context_terms[order(-nchar(cohort_context_terms))]

print("Number of cohort context terms:")
length(cohort_context_terms)

writeLines(cohort_context_terms, 
           here::here("output/gwas_cat/ancestry_population_terms.txt"))

```

# Cohort names

### GWAS Catalog Cohort information

```{r gwas_cat_add_cohort_info}

gwas_study_info_cohort =
  data.table::fread(here::here("output/gwas_cohorts/gwas_cohort_name_corrected.csv"))

gwas_study_info_cohort =
  gwas_study_info_cohort |>
  dplyr::rename_with(~ gsub(" ", "_", .x))

gwas_study_info_cohort =
  gwas_study_info_cohort |>
  select(PUBMED_ID,
         COHORT) |>
  distinct()

print("Check adding cohort information has only added columns, not rows:")
print("Before adding:")
dim(gwas_study_info)

gwas_study_info =
  left_join(gwas_study_info,
            gwas_study_info_cohort,
            by = "PUBMED_ID"
  )

print("After adding:")
dim(gwas_study_info)


print("Check adding cohort information has not increaed number of unique pubmed ids:")
gwas_study_info |>
  pull(PUBMED_ID) |>
  unique() |>
  length()

print("Number of unique pubmed ids for disease studies with cohort info:")
gwas_study_info |>
  filter(COHORT != "") |>
  pull(PUBMED_ID) |>
  unique() |>
  length()

```

### Dictionary of cohort names 

```{r get_cohort_names}

# this xlsx was built from looking at acrynyms / cohort names in the gwas catalog
# and finding the corresponding full names / details of cohorts

cohort_names <- readxl::read_xlsx(here::here("data/cohort/cohort_desc.xlsx"),
                                 sheet = 1) |>
  mutate(across(everything(), 
                ~stringr::str_replace_all(.x,
                                          pattern = "\u00A0",
                                          replacement = " "))
         ) 


cohort_full_names = cohort_names$full_name[!is.na(cohort_names$full_name)]
cohort_full_names <- str_trim(cohort_full_names)
cohort_full_names <- iconv(cohort_full_names, to = "UTF-8")
cohort_full_names <- gsub("[\u00A0\r\n]", " ", cohort_full_names)  # replace non-breaking spaces, CR, LF with space
cohort_full_names <- str_squish(cohort_full_names)  # trims and removes extra spaces
# sort by length of name (longest first) to match longest names first
cohort_full_names <- cohort_full_names[order(-nchar(cohort_full_names))]
cohort_full_names <- cohort_full_names[cohort_full_names != "Not Reported"]
cohort_full_names <- unique(cohort_full_names)

cohort_full_names <- cohort_full_names[cohort_full_names != "Japan"]
cohort_full_names <- cohort_full_names[cohort_full_names != "Coronary Artery Disease"]

print("Number of unique cohort full names:")
length(cohort_full_names)


cohort_abbr_names = cohort_names$cohort[!is.na(cohort_names$cohort)]
cohort_abbr_names <- str_trim(cohort_abbr_names)
cohort_abbr_names <- iconv(cohort_abbr_names, to = "UTF-8")
cohort_abbr_names <- gsub("[\u00A0\r\n]", " ", cohort_abbr_names)  # replace non-breaking spaces, CR, LF with space
cohort_abbr_names <- str_squish(cohort_abbr_names)  # trims and removes extra spaces

# remove abbreviations that are too short
# cohort_abbr_names <- cohort_abbr_names[nchar(cohort_abbr_names) >= 4]
# small_abbr_to_keep <- c("C4D", 
#                         "BBJ", 
#                         "UKB", 
#                         "MVP", 
#                         "TWB", 
#                         "QBB",
#                         "MEC",
#                         "WHI"
#                         )

# cohort_abbr_names <- unique(c(cohort_abbr_names, 
#                               small_abbr_to_keep
#                               ))

cohort_abbr_names <- cohort_abbr_names[!str_detect(pattern = "\\?", 
                                                   cohort_abbr_names)]
# sort by length of name (longest first) to match longest names first
cohort_abbr_names <- cohort_abbr_names[order(-nchar(cohort_abbr_names))]


# add cohort names from GWAS catalog not yet added to data-dictionary
gwas_cat_cohorts = gwas_study_info_cohort$COHORT
gwas_cat_cohorts = unlist(strsplit(gwas_cat_cohorts, "\\|"))
gwas_cat_cohorts = gwas_cat_cohorts[!(gwas_cat_cohorts %in% c("", "other", "multiple"))]

cohort_abbr_names = unique(c(cohort_abbr_names,
                           gwas_cat_cohorts))

# remove small abbreviations that are likely to be false positives
small_abbrev = c("CFRD",
                 "ALS",
                 "UK", 
                 "US",
                 "NZ",
                 "DN",
                 "CHB",
                 "FG",
                 "HTN",
                 "ICH",
                 "GD",
                 "MD",
                 "UC",
                 "DC",
                 "ACE",
                 "Japan",
                 "Nigeria",
                 "Ghana",
                 "YRI",
                 "CEU",
                 "Columbia",
                 "York",
                 "Icelandic")


cohort_abbr_names <- cohort_abbr_names[!c(cohort_abbr_names %in% small_abbrev)]

print("Number of unique cohort abbreviation names:")
length(cohort_abbr_names)

```

```{r save_cohort_terms}

writeLines(cohort_abbr_names,
           here::here("output/gwas_cat/cohort_abbrev_names.txt"))

writeLines(cohort_full_names,
           here::here("output/gwas_cat/cohort_full_names.txt"))

```


