---
title: "group_non_cancer_diseases"
author: "Isobel Beasley"
date: "2025-09-28"
output: html_document
---

# Set up

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load_packages, message=FALSE}

library(dplyr)
library(stringr)
library(data.table)

```

## Ontology help - for getting disease subtypes

```{r ontology_fns}

source(here::here("code/get_term_descendants.R"))

```

### Load Data

```{r load_data}

gwas_study_info <- fread(here::here("output/gwas_cat/gwas_study_info_group.csv"))

```

## Typo: spine psteoarthritis

```{r}

gwas_study_info =
  gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern(c(
                            "spine psteoarthritis"
                            )
                            ),
                          "spine osteoarthritis"
         )
  )

```

## Intracranial artery stenosis

```{r icas}
# "arterial stenosis"  -> "Intracranial artery stenosis"
# where DISEASE/TRAIT contains "Intracranial artery stenosis"
# replace "arterial stenosis" with "intracranial artery stenosis"

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         ifelse(grepl("Intracranial artery stenosis", 
                     `DISEASE/TRAIT`,
                     ignore.case = TRUE),
                stringr::str_replace_all(collected_all_disease_terms,
                          pattern = "arterial stenosis",
                          replacement = "intracranial artery stenosis"
         ),
         collected_all_disease_terms
         )
  )

```

# Pubmed id: 37271218 study of complications in Behcet's disease

Taking definitions etc. from: https://pmc-ncbi-nlm-nih-gov.ucsf.idm.oclc.org/articles/PMC8584626/

```{r behcets_disease}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         ifelse(PUBMED_ID == "37271218",
                stringr::str_replace_all(collected_all_disease_terms,
                          pattern = "abnormality of the eye",
                          replacement = "uveitis, retinal vasculitis, chorioretinitis, papillitis"
         ),
         collected_all_disease_terms
         )
  )

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         ifelse(PUBMED_ID == "37271218",
                stringr::str_replace_all(collected_all_disease_terms,
                          pattern = "abnormality of the genital system",
                          replacement = "genital aphthosis"
         ),
         collected_all_disease_terms
         )
  )

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         ifelse(PUBMED_ID == "37271218",
                stringr::str_replace_all(collected_all_disease_terms,
                          pattern = "abnormal aortic morphology",
                          replacement = "arterial thrombosis, aortic aneurysm, aneurysm of pulmonary artery"
         ),
         collected_all_disease_terms
         )
  )

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         ifelse(PUBMED_ID == "37271218",
                stringr::str_replace_all(collected_all_disease_terms,
                          pattern = "abnormal venous morphology",
                          replacement = "venous thromboembolism, superficial thrombophlebitis"
         ),
         collected_all_disease_terms
         )
  )


gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         ifelse(PUBMED_ID == "37271218",
                stringr::str_replace_all(collected_all_disease_terms,
                          pattern = "abnormality of the vasculature",
                          replacement = "arterial thrombosis, aortic aneurysm, aneurysm of pulmonary artery, venous thromboembolism, superficial thrombophlebitis"
         ),
         collected_all_disease_terms
         )
  )

```

## Pigmentary iris degeneration 

```{r pigmentary_iris_degeneration}

# where Pigmentary iris degeneration in DISEASE/TRAIT & pubmed id 39024449
# replace abnormality of the eye with pigmentary iris degeneration

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         ifelse(grepl("Pigmentary iris degeneration", 
                     `DISEASE/TRAIT`,
                     ignore.case = TRUE) & PUBMED_ID == "39024449",
                stringr::str_replace_all(collected_all_disease_terms,
                          pattern = "abnormality of the eye",
                          replacement = "pigmentary iris degeneration"
         ),
         collected_all_disease_terms
         )
  )


```

## Coronary restenosis <- add inclusion criteria 

```{r coronary_restenosis}

# for pubmed id 21878436 
# measuring Coronary restenosis, which is an outcome of Percutaneous coronary intervention (PCI)
# included / background traits for the relevant study GENetic DEterminants of Restenosis project (GENDER) 
# are found:
# https://academic.oup.com/eurheartj/article/25/13/1163/465324?login=true#89258510

# stable angina
# silent ischaemia -> silent myocardial ischaemia
# non-ST-elevation acute coronary syndromes excl. acute ST-elevation myocardial infarction
# this includes: 
# unstable angina, 
# non-ST-elevation myocardial infarction (NSTEMI) within acute subendocardial myocardial infarction
# Acute ischemic heart disease, unspecified: 

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         ifelse(PUBMED_ID == "21878436",
                stringr::str_replace_all(collected_all_disease_terms,
                          pattern = "coronary restenosis",
                          replacement = "coronary restenosis, stable angina, silent myocardial ischaemia, unstable angina, acute subendocardial myocardial infarction, acute ischemic heart disease"
         ),
         collected_all_disease_terms
         )
  )

```


## Acute coronary syndrome

```{r acute_coronary_syndrome}

# for pubmed id 28753643
# acute coronary syndrome (ACS) 
# comes from the SOLID-TIMI 52 cohort: experienced acute coronary syndrome (ACS; unstable angina, non–ST-elevation or ST-elevation MI)

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         ifelse(PUBMED_ID == "28753643",
                stringr::str_replace_all(collected_all_disease_terms,
                          pattern = "acute coronary syndrome",
                          replacement = "unstable angina, acute subendocardial myocardial infarction, acute transmural myocardial infarction of unspecified site"
         ),
         collected_all_disease_terms
         )
  )

# define similarly in pubmed id: 24952865, 25583994, 25935875
gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         ifelse(PUBMED_ID %in% c("28753643","24952865", "28753645", "25583994", "25935875", "34158603"),
                stringr::str_replace_all(collected_all_disease_terms,
                          pattern = "acute coronary syndrome",
                          replacement = "unstable angina, acute subendocardial myocardial infarction, acute transmural myocardial infarction of unspecified site"
         ),
         collected_all_disease_terms
         )
  )

# for pubmed id 35356681, acute coronary syndrome (ACS) - is actually non-ST elevation acute coronary syndrome
gwas_study_info |> 
  filter(grepl(vec_to_grep_pattern("acute coronary syndrome"), 
               collected_all_disease_terms, 
               perl = T)) |> 
  filter(PUBMED_ID == "35356681") |> 
  pull(`DISEASE/TRAIT`) |>
  unique()

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         ifelse(PUBMED_ID == "35356681",
                stringr::str_replace_all(collected_all_disease_terms,
                          pattern = "acute coronary syndrome",
                          replacement = "acute subendocardial myocardial infarction"
         ),
         collected_all_disease_terms
         )
  )

```

## Amebiasis

```{r amebiasis}

# If `DISEASE/TRAIT` contains Diarrhoea-associated Entamoeba histolytica infection
# then replace amebiasis with amoeboma of intestine 

gwas_study_info =
  gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         ifelse(grepl("Diarrhoea-associated Entamoeba histolytica infection", 
                     `DISEASE/TRAIT`,
                     ignore.case = TRUE),
                stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern("amebiasis"),
                          "amoeboma of intestine"
         ),
         collected_all_disease_terms
         )
  )


```

## Congenital anomalies of intestine

```{r congenital_anomalies_of_intestine}

# anorectal malformation to congenital anomalies of intestine
gwas_study_info =
  gwas_study_info |>
  mutate(collected_all_disease_terms  = 
                stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern("anorectal malformation"),
                          "congenital anomalies of intestine"
         )
  )

```

## Vasculitis

```{r vasculitis}

gwas_study_info =
  gwas_study_info |>
  mutate(collected_all_disease_terms  = 
                stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern("anti-neutrophil antibody associated vasculitis"),
                          "vasculitis"
         )
  )


```




## Hepatic steatosis

```{r hepatic_steatosis}

# if DISEASE/TRAIT contains hepatic steatosis in non-alcoholic fatty liver disease
# replace hepatic steatosis with nonalcoholic steatohepatitis 

gwas_study_info =
  gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         ifelse(grepl("hepatic steatosis in non-alcoholic fatty liver disease", 
                     `DISEASE/TRAIT`,
                     ignore.case = TRUE),
                stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern("hepatic steatosis"),
                          "nonalcoholic steatohepatitis"
         ),
         collected_all_disease_terms
         )
  )


gwas_study_info =
  gwas_study_info |>
  mutate(collected_all_disease_terms  = 
                stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern("non-alcoholic steatohepatitis"),
                          "nonalcoholic steatohepatitis"
  )
  )


```

## Thrombophilia

```{r thrombophilia}

# if DISEASE/TRAIT contains PheCode 286.8
# then replce thrombophilia with primary thrombophilia
gwas_study_info =
  gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         ifelse(grepl("PheCode 286.8", 
                     `DISEASE/TRAIT`,
                     ignore.case = TRUE),
                stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern("thrombophilia"),
                          "primary thrombophilia"
         ),
         collected_all_disease_terms
         )
  )

```

## Replace abnormal blistering of the skin -> blister


```{r blistering_of_skin}

gwas_study_info =
  gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         ifelse(`DISEASE/TRAIT` == "Blister (PheCode 911)",
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern("abnormal blistering of the skin"),
                          "blister"
         ),
         collected_all_disease_terms
         )
  )


```

## abnormal circulating lipid concentration -> disorders of lipid metabolism

```{r abnormal_circulating_lipid_concentration}

gwas_study_info =
  gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         ifelse(grepl("Disorders of lipid metabolism", 
                     `DISEASE/TRAIT`,
                     ignore.case = TRUE),
                stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern("abnormal circulating lipid concentration"),
                          "disorders of lipid metabolism"
         ),
         collected_all_disease_terms
         )
  )

```

## Abnormal thrombosis

```{r abnormal_thrombosis}

# interesting things going on with abnormal thrombosis
gwas_study_info |> 
  filter(grepl(vec_to_grep_pattern("abnormal thrombosis"), 
               collected_all_disease_terms, perl = T)) |> 
  select(`DISEASE/TRAIT`, collected_all_disease_terms, PUBMED_ID)

# if DISEASE/TRAIT contains Blood clot in the lung \\(UKB data field 6152\\) 
# replace abnormal thrombosis in collected_all_disease_terms with pulmonary embolism
gwas_study_info =
  gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         ifelse(grepl("Blood clot in the lung \\(UKB data field 6152\\)", 
                     `DISEASE/TRAIT`,
                     ignore.case = TRUE),
                stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern("abnormal thrombosis"),
                          "pulmonary embolism"
         ),
         collected_all_disease_terms
         )
  )


gwas_study_info =
  gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         ifelse(grepl("Blood clot in the lung \\(UKB data field 6152_7\\)", 
                     `DISEASE/TRAIT`,
                     ignore.case = TRUE),
                stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern("abnormal thrombosis"),
                          "pulmonary embolism"
         ),
         collected_all_disease_terms
         )
  )

## PHESANT recoding (UKB data field 6152_100)
gwas_study_info =
  gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         ifelse(grepl("PHESANT recoding \\(UKB data field 6152_100\\)", 
                     `DISEASE/TRAIT`,
                     ignore.case = TRUE),
                stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern("abnormal thrombosis, allergic disease, respiratory system disease"),
                          "pulmonary embolism, deep vein thrombosis, chronic bronchitis, emphysema, asthma, rhinitis, eczema"
         ),
         collected_all_disease_terms
         )
  )


# for pubmed id: 26908601
# remove abnormal thrombosis
# because this is included in the other provided terms 

gwas_study_info =
  gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         ifelse(PUBMED_ID == "26908601",
                stringr::str_remove_all(collected_all_disease_terms,
                          vec_to_grep_pattern("abnormal thrombosis")
         ),
         collected_all_disease_terms
         )
  )

```


### How many unique traits are there now?

```{r n_unique_traits_start}

diseases <- stringr::str_split(pattern = ", ",
                               gwas_study_info$collected_all_disease_terms[gwas_study_info$collected_all_disease_terms != ""])  |>
  unlist() |>
  stringr::str_trim()

diseases <- unique(diseases)

print(length(diseases))


```

# Filtering some traits 

## Abnormality of the digestive system

```{r abnormality_of_digestion}

gwas_study_info |> 
  filter(grepl(vec_to_grep_pattern("abnormality of the digestive system"), 
               collected_all_disease_terms, 
               perl = T)) |>
  pull(`DISEASE/TRAIT`) |>
  unique()

gwas_study_info =
  gwas_study_info |>
  mutate(collected_all_disease_terms =
           str_remove_all(collected_all_disease_terms,
                          pattern = vec_to_grep_pattern("abnormality of the digestive system")))


```

## Chronic disease

```{r chronic_disease}

gwas_study_info |> 
  filter(grepl(vec_to_grep_pattern("chronic disease"), 
               collected_all_disease_terms, 
               perl = T)) |>
  pull(`DISEASE/TRAIT`) |>
  unique()

gwas_study_info =
  gwas_study_info |>
  mutate(collected_all_disease_terms =
           str_remove_all(collected_all_disease_terms,
                          pattern = vec_to_grep_pattern("chronic disease")))


```

## Fibrosis

```{r fibrosis}

gwas_study_info |> 
  filter(grepl(vec_to_grep_pattern("fibrosis"), 
               collected_all_disease_terms, 
               perl = T)) |>
  pull(`DISEASE/TRAIT`) |>
  unique()


gwas_study_info =
  gwas_study_info |>
  mutate(collected_all_disease_terms =
           str_replace_all(collected_all_disease_terms,
                           pattern = vec_to_grep_pattern("fibrosis, head and neck cancer"),
                          replacement = "head and neck cancer"
           )
  )

gwas_study_info =
  gwas_study_info |>
  mutate(collected_all_disease_terms =
           ifelse(PUBMED_ID == "38601445",
           str_replace_all(collected_all_disease_terms,
                           pattern = vec_to_grep_pattern("fibrosis"),
                          replacement = "lung cancer"
           ),
           collected_all_disease_terms
           )
  )

```

## Hyperplasia 

```{r hyperplasia}

gwas_study_info |> 
  filter(grepl(vec_to_grep_pattern("hyperplasia"), 
               collected_all_disease_terms, 
               perl = T)) |>
  pull(`DISEASE/TRAIT`) |>
  unique()

gwas_study_info =
  gwas_study_info |>
  mutate(collected_all_disease_terms =
           ifelse(grepl("Bilateral adrenal hyperplasia",
                        `DISEASE/TRAIT`,
                       ignore.case = TRUE),
                  str_replace_all(collected_all_disease_terms,
                                  pattern = vec_to_grep_pattern("hyperplasia"),
                                  replacement = "bilateral adrenal hyperplasia"
                  ),
                  collected_all_disease_terms
           )
)

```

## Cognitive decline (age-related)

```{r cognitive_decline}

gwas_study_info =
  gwas_study_info |>
  mutate(collected_all_disease_terms =
           ifelse(grepl("Cognitive decline \\(age-related\\)",
                        `DISEASE/TRAIT`,
                       ignore.case = TRUE) |
                  grepl("cognitive ageing",
                        STUDY,
                        ignore.case = TRUE),
                  str_replace_all(collected_all_disease_terms,
                                  pattern = vec_to_grep_pattern("mental deterioration"),
                                  replacement = "cognitive decline (age-related)"
                  ),
                  collected_all_disease_terms
           )
)


gwas_study_info =
  gwas_study_info |>
  mutate(collected_all_disease_terms =
           ifelse(grepl("cognitive decline in schizophrenia",
                        STUDY,
                       ignore.case = TRUE),  
                  str_replace_all(collected_all_disease_terms,
                                  pattern = vec_to_grep_pattern("mental deterioration"),
                                  replacement = "schizophrenia"
                  ),
                  collected_all_disease_terms
           )
)
```

## nasopharyngeal neoplasm

```{r nasopharyngeal_neoplasm}

gwas_study_info =
  gwas_study_info |>
  mutate(collected_all_disease_terms =
           ifelse(grepl("nasopharyngeal carcinoma",
                        STUDY,
                       ignore.case = TRUE),  
                  str_replace_all(collected_all_disease_terms,
                                  pattern = vec_to_grep_pattern("nasopharyngeal neoplasm"),
                                  replacement = "nasopharyngeal cancer"
                  ),
                  collected_all_disease_terms
           )
)


```

## Formal thought disorder

```{r formal_though_disorder}

gwas_study_info |> 
  filter(grepl(vec_to_grep_pattern("formal thought disorder"), 
               collected_all_disease_terms, 
               perl = T)) |>
  pull(`DISEASE/TRAIT`) |>
  unique()

# as formal thought disorder is within other psychiatric traits
# remove
gwas_study_info =
  gwas_study_info |>
  mutate(collected_all_disease_terms =
           str_remove_all(collected_all_disease_terms,
                          pattern = vec_to_grep_pattern("formal thought disorder")))


```

## Hypersensitivity reaction disease

```{r hypersensitivity reaction disease}

gwas_study_info |> 
  filter(grepl(vec_to_grep_pattern("hypersensitivity reaction disease"), 
               collected_all_disease_terms, 
               perl = T)) |>
  pull(`DISEASE/TRAIT`) |>
  unique()

gwas_study_info =
  gwas_study_info |>
  mutate(collected_all_disease_terms =
           str_remove_all(collected_all_disease_terms,
                          pattern = vec_to_grep_pattern("hypersensitivity reaction disease")))


```

## inflammatory skin disease

```{r inflammatory_skin_disease}

gwas_study_info |> 
  filter(grepl(vec_to_grep_pattern("inflammatory skin disease"), 
               collected_all_disease_terms, 
               perl = T)) |>
  pull(`DISEASE/TRAIT`) |>
  unique()


gwas_study_info =
  gwas_study_info |>
  mutate(collected_all_disease_terms =
           str_replace_all(collected_all_disease_terms,
                           pattern = vec_to_grep_pattern("inflammatory skin disease"),
                          replacement = "atopic dermatitis, psoriasis"
           )
  )

```

## intestinal cancer

```{r intestinal_cancer}

# if DISEASE?/TRAIT contains "Bowel cancer"
# replace intestinal cancer with colorectal cancer

gwas_study_info =
  gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         ifelse(grepl("Bowel cancer", 
                     `DISEASE/TRAIT`,
                     ignore.case = TRUE),
                stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern("intestinal cancer"),
                          "colorectal cancer"
         ),
         collected_all_disease_terms
         )
  )



```

## Abnormality of skin pigmentation

```{r abnormality_of_skin_pigmentation}

gwas_study_info |> 
  filter(grepl("abnormality of skin pigmentation", collected_all_disease_terms)) |>
  pull(`DISEASE/TRAIT`) |>
  unique()

# as can see, most of these traits are not about true skin pigmentation abnormalities
# just about skin pigmentation in general

# if `DISEASE/TRAIT` contains "UKB data field 1717" (which is just a skin colour measurment field)
# then remove "abnormality of skin pigmentation" from collected_all_disease_terms
gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         ifelse(grepl("UKB data field 1717", 
                     `DISEASE/TRAIT`,
                     ignore.case = TRUE),
                stringr::str_remove_all(collected_all_disease_terms,
                          pattern = "abnormality of skin pigmentation"
         ),
         collected_all_disease_terms
         )
  )

# also remove if `DISEASE/TRAIT` contains "Tattoo pigmentation"
gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         ifelse(grepl("tattoo pigmentation|tatto pigmentation", 
                     `DISEASE/TRAIT`,
                     ignore.case = TRUE),
                stringr::str_remove_all(collected_all_disease_terms,
                          pattern = "abnormality of skin pigmentation"
         ),
         collected_all_disease_terms
         )
  )

# also remove if `DISEASE/TRAIT` is any of the following:
# - Perceived skin darkness 
# - Skin red/green component
# - Skin yellow/blue component
# - Skin pigmentation traits
# - Skin pigmentation 
# - Skin luminance

remove_terms <- c("Perceived skin darkness",
                  "Skin red/green component",
                  "Skin yellow/blue component",
                  "Skin pigmentation traits",
                  "Skin pigmentation",
                  "Skin luminance",
                  "Skin colour saturation"
                  )

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         ifelse(`DISEASE/TRAIT` %in% remove_terms,
                stringr::str_remove_all(collected_all_disease_terms,
                          pattern = "abnormality of skin pigmentation"
         ),
         collected_all_disease_terms
         )
  )

# this leaves PUBMED_ID 23548203, which according to suplementary table 1
# is `Number of non-melanoma skin cancer (5+)` 
# so change to "non-melanoma skin cancer"

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         ifelse(PUBMED_ID == "23548203",
                stringr::str_replace_all(collected_all_disease_terms,
                          pattern = "abnormality of skin pigmentation",
                          replacement = "non-melanoma skin cancer"
         ),
         collected_all_disease_terms
         )
  )


```

## Handedness 

```{r handedness}

gwas_study_info |> 
  filter(grepl("handedness", collected_all_disease_terms)) |>
  pull(`DISEASE/TRAIT`) |>
  unique()

# remove handedness
gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_remove_all(collected_all_disease_terms,
                          pattern = "handedness"
         ))

# also remove functional laterality if Handedness chirality laterality is in DISEASE/TRAIT
gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         ifelse(grepl("Handedness", 
                     `DISEASE/TRAIT`,
                     ignore.case = TRUE),
                stringr::str_remove_all(collected_all_disease_terms,
                          pattern = "functional laterality"
         ),
         collected_all_disease_terms
         )
  )

# also remove functional laterality if Usual side of head for mobile phone use is in DISEASE/TRAIT
gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         ifelse(grepl("Usual side of head for mobile phone use", 
                     `DISEASE/TRAIT`,
                     ignore.case = TRUE),
                stringr::str_remove_all(collected_all_disease_terms,
                          pattern = "functional laterality"
         ),
         collected_all_disease_terms
         )
  )

# also remove functional laterality if PUBMED_ID=20585627
gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         ifelse(PUBMED_ID == "20585627",
                stringr::str_remove_all(collected_all_disease_terms,
                          pattern = "functional laterality"
         ),
         collected_all_disease_terms
         )
  ) 


```


## Attached earlobe

```{r attached_earlobe}

gwas_study_info |> 
  filter(grepl("attached earlobe", collected_all_disease_terms)) |>
  pull(`DISEASE/TRAIT`) |>
  unique()

# remove attached earlobe
gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_remove_all(collected_all_disease_terms,
                          pattern = "attached earlobe"
         ))

```

## Facial wrinkling 

```{r facial_wrinkling}

gwas_study_info |> 
  filter(grepl("facial wrinkling", collected_all_disease_terms)) |>
  pull(`DISEASE/TRAIT`) |>
  unique()

# remove facial wrinkling
gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_remove_all(collected_all_disease_terms,
                          pattern = "facial wrinkling"
         ))

```

## Gender identity disorder

```{r gender_identity_disorder}

gwas_study_info = 
  gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern("childhood gender nonconformity"),
                          "gender identity disorder"
         ))

gwas_study_info =
    gwas_study_info |>
    mutate(collected_all_disease_terms  = 
         stringr::str_remove_all(collected_all_disease_terms,
                          pattern = "gender identity disorder"
         ))

gwas_study_info =
  gwas_study_info |>
    mutate(collected_all_disease_terms  = 
         stringr::str_remove_all(collected_all_disease_terms,
                          pattern = "sexual and gender identity disorders"
         ))

```


## Sneezing in response to bright light (autosomal dominant compelling helio-ophthalmic outburst syndrome)

```{r sneezing_in_response_to_bright_light}

gwas_study_info |> 
  filter(grepl("autosomal dominant compelling helio-ophthalmic outburst syndrome", collected_all_disease_terms)) |>
  pull(`DISEASE/TRAIT`) |>
  unique()

# remove autosomal dominant compelling helio-ophthalmic outburst syndrome
gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_remove_all(collected_all_disease_terms,
                          pattern = "autosomal dominant compelling helio-ophthalmic outburst syndrome"
         ))


```

## Skin sensitivity to sun

```{r skin_sensitivity_to_sun}

gwas_study_info |> 
  filter(grepl("skin sensitivity to sun", collected_all_disease_terms)) |>
  pull(`DISEASE/TRAIT`) |>
  unique()

# remove skin sensitivity to sun
gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_remove_all(collected_all_disease_terms,
                          pattern = vec_to_grep_pattern("skin sensitivity to sun")
         ))

# also remove suntan
gwas_study_info |> 
  filter(grepl("suntan", collected_all_disease_terms)) |>
  pull(`DISEASE/TRAIT`) |>
  unique()


gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_remove_all(collected_all_disease_terms,
                          pattern = vec_to_grep_pattern("suntan"
                          )
         ))

```


# Group some non-cancer diseases together

## Abnormal brain morphology

```{r abnormal_brain_morphology}

# if DISEASE/TRAIT contains Unidentified bright object on brain MRI, 
# then replace abnormal brain morphology with Other abnormal findings on diagnostic imaging of central nervous system 
gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         ifelse(grepl("Unidentified bright object on brain MRI", 
                     `DISEASE/TRAIT`,
                     ignore.case = TRUE),
                stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern("abnormal brain morphology"),
                          "other abnormal findings on diagnostic imaging of central nervous system"
         ),
         collected_all_disease_terms
         )
  )

```

## Abnormal ecg 

```{r abnormal_ecg}
# if abnormal ecg in disease/trait, then change abnormal ekg to abnormal ecg
gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         ifelse(grepl("abnormal ecg", 
                     `DISEASE/TRAIT`,
                     ignore.case = TRUE),
                stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern("abnormal ekg"),
                          "abnormal ecg"
         ),
         collected_all_disease_terms
         )
  )

```

## Abnormality of gait

```{r abnormality_of_gait}

gwas_study_info = 
  gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern(c(
                            "gait imbalance",
                            "decreased walking ability",
                            "postural instability")
                            ),
                          "abnormality of gait"
         ))

```

## Abnormality of the nervous system, abnormality of the musculature, abnormality of the skeletal system

```{r abnormality_of_nervous_system_musculature_skeletal_system}

gwas_study_info = 
  gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern(c(
                            "abnormality of the musculature, abnormality of the nervous system, abnormality of the skeletal system")
                            ),
                          "other symptoms and signs involving the nervous and musculoskeletal systems"
         ))


```

## Other demyelinating diseases of central nervous system

```{r other_demyelinating_diseases_of_cns}

# if DISEASE/TRAIT contains PheCode 341
# replace abnormality of the nervous system, with other demyelinating diseases of central nervous system

gwas_study_info = 
  gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         ifelse(grepl("PheCode 341", 
                     `DISEASE/TRAIT`,
                     ignore.case = TRUE),
                stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern("abnormality of the nervous system"),
                          "other demyelinating diseases of central nervous system"
         ),
         collected_all_disease_terms
         )
  )


gwas_study_info = 
  gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         ifelse(grepl("PheCode 345", 
                     `DISEASE/TRAIT`,
                     ignore.case = TRUE),
                stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern("abnormality of the nervous system"),
                          "epilepsy"
         ),
         collected_all_disease_terms
         )
  )


gwas_study_info = 
  gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         ifelse(grepl("Neurological involvement in Behcet's disease", 
                     `DISEASE/TRAIT`,
                     ignore.case = TRUE),
                stringr::str_remove_all(collected_all_disease_terms,
                          vec_to_grep_pattern("abnormality of the nervous system")
         ),
         collected_all_disease_terms
         )
  )


```

## Abdominal abscess

```{r abdominal_abscess}

gwas_study_info |>
  filter(grepl(vec_to_grep_pattern("abdominal abscess"), 
               collected_all_disease_terms, perl = T)) 

# just one study - 35173190, has abdominal abscess
# they define unusually as: D73.3, K35-37, K57, K61, K63.0, K65, K75.0, K81, K83.0
# 'Abdominal infections'

gwas_study_info = 
  gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         ifelse(PUBMED_ID == "35173190",
                stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern("abdominal abscess"),
                          "abdominal infections code"
         ),
         collected_all_disease_terms
         )
  )

```

## Abnormal mammogram

```{r abnormal_mammogram}

# if DISEASE/TRAIT contains Abnormal mammogram, then change Abnormality of the breast to abnormal mammogram

gwas_study_info = 
  gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         ifelse(grepl("Abnormal findings on mammogram or breast exam", 
                     `DISEASE/TRAIT`,
                     ignore.case = TRUE),
                stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern("abnormality of the breast"),
                          "abnormal mammogram"
         ),
         collected_all_disease_terms
         )
  )

```

## Achalasia of cardia

```{r achalasia_of_cardia}

gwas_study_info = 
  gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern("achalasia"),
                          "achalasia of cardia"
         ))

```

## Acute pancreatitis

```{r acute_pancreatitis}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern("asparaginase-induced acute pancreatitis"),
                          "acute pancreatitis"
         ))

```

## Altitude sickness

```{r altitude_sickness}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern("chronic mountain sickness"),
                          "altitude sickness"
         ))


```

## Alopecia

```{r alopecia}


alopecia_terms <- c("frontal fibrosing alopecia" 
                   )


gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          pattern = vec_to_grep_pattern(alopecia_terms),
                          "alopecia"
         )
  )
         
         
drug_induced_alopecia_terms <- c(
                               "chemotherapy-induced alopecia"
                               )


gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          pattern = vec_to_grep_pattern(drug_induced_alopecia_terms),
                          "drug-induced androgenic alopecia"
         )
  )


```

## Alcohol and nicotine codependence

```{r alcohol_and_nicotine_codependence}
# alcohol and nicotine codependence -> alcohol dependence, nicotine dependence

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern("alcohol and nicotine codependence"),
                          "alcohol dependence, nicotine dependence"
         )) 

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern("alcohol dependence"),
                          "alcohol-related disorders"
         ))

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern("nicotine dependence"),
                          "tobacco use disorder"
         ))

```

## Amyloidosis

```{r amyloidosis}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern("al amyloidosis"),
                          "amyloidosis"
         ))

```

## Anxiety disorder

```{r anxiety}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern("anxiety"),
                          "anxiety disorder"
         )) 

```

## Aplastic anemia

```{r aplastic_anemia}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          pattern = vec_to_grep_pattern("severe aplastic anemia"),
                          "aplastic anemia"
         ))


```

## Androgenic alopecia 

```{r androgenic_alopecia}

gwas_study_info = 
  gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          pattern = vec_to_grep_pattern("androgenetic alopecia"),
                          "androgenic alopecia"
         ))  

```

## angioedema -> Angioneurotic oedema

```{r angioedema}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern("angioedema"),
                          "angioneurotic oedema"
         ))


```

## Astigmatism

```{r astigmatism}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern("corneal astigmatism"),
                          "astigmatism"
         ))

```

## Asthma

```{r asthma}

gwas_study_info = gwas_study_info |>
      mutate(collected_all_disease_terms  = 
           stringr::str_replace_all(collected_all_disease_terms,
                            vec_to_grep_pattern(c("atopic asthma", "chronic obstructive asthma")),
                            "asthma"
           )) |>
    mutate(collected_all_disease_terms  = 
           stringr::str_replace_all(collected_all_disease_terms,
                            vec_to_grep_pattern(c("childhood onset asthma",
                                                    "adult onset asthma",
                                                    "aspirin-induced asthma"
                                                  )
                            ),
                            "asthma"
           ))

```

## Atrial fibrillation and flutter

```{r atrial_fibrillation_and_flutter}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern("atrial flutter"),
                          "atrial fibrillation and flutter"
         ))

```

## Atopic dermatitis

```{r atopic_dermatitis}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern("recalcitrant atopic dermatitis"),
                          "atopic dermatitis"
         ))

```

## Eczema

```{r eczema}

# if DISEASE/STUDY contains eczema, change eczematoid dermatitis to eczema
gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         ifelse(grepl("eczema", 
                     `DISEASE/TRAIT`,
                     ignore.case = TRUE),
                stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern("eczematoid dermatitis"),
                          "eczema"
         ),
         collected_all_disease_terms
         )
  )


gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
                stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern("hand eczema"),
                          "eczema"
         )
  )


```

## Post-operative

```{r post_op}

gwas_study_info = gwas_study_info |> 
    mutate(collected_all_disease_terms  = 
         stringr::str_remove_all(collected_all_disease_terms,
                         pattern = "^post-operative |^postoperative"
         ))


```

## Bipolar disorder

```{r bipolar_disorder}

gwas_study_info = gwas_study_info |> 
    mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern(
                            c("bipolar ii disorder",
                              "bipolar i disorder"
                              )
                            ),
                          "bipolar disorder"
         ))

```

## Bacterial infection

```{r bacterial_infection}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern(c(
                             "bacterial infection"
                             )
                             ),
                          "bacterial infection nos"
         ))


```

## Benign mammary dysplasia

```{r benign_mammary_dysplasia}

# if `DISEASE/TRAIT contains Benign mammary dysplasia`, then change abnormality of the breast to benign mammary dysplasia
gwas_study_info = 
  gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         ifelse(grepl("Benign mammary dysplasia", 
                     `DISEASE/TRAIT`,
                     ignore.case = TRUE),
                stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern("abnormality of the breast"),
                          "benign mammary dysplasia"
         ),
         collected_all_disease_terms
         )
  )


```

## Bronchopneumonia and lung abscess

```{r bronchopneumonia_and_lung_abscess}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern(c(
                             "bronchopneumonia, lung abscess"
                             )
                             ),
                          "bronchopneumonia and lung abscess"
         ))

```

## Blindness and low vision

```{r blindness_and_low_vision}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern(c(
                             "blindness",
                             "progressive visual loss",
                             "visual loss"
                             )
                             ),
                          "blindness and low vision"
         ))

# Disorders of optic nerve and visual pathways
gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern(c(
                             "visual pathway disorder",
                             "optic nerve disorder"
                             )
                             ),
                          "disorders of optic nerve and visual pathways"
         ))

# visuospatial impairment
gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern(c(
                             "visuospatial impairment"
                             )
                             ),
                          "other and unspecified symptoms and signs involving cognitive functions and awareness"
         ))

```

## Cafe-au-lait spot

```{r cafe_au_lait_spot}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern("cafe-au-lait spot"),
                          "café au lait spots"
         ))

```

## Cancer of eye

```{r cancer_of_eye}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern(c(
                             "ocular melanoma")
                             ),
                          "cancer of eye"
         ))

```

## Candidiasis of vulva and vagina

```{r candid_vagina}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern(c(
                             "vaginal yeast infection")
                             ),
                          "candidiasis of vulva and vagina"
         ))


```

## Carbuncle and furuncle

```{r carbuncle_and_furuncle}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern(c(
                             "carbuncle", "furuncle")
                             ),
                          "carbuncle and furuncle"
         ))


```

## Cardiac arrhythmia

```{r cardiac_arrhythmia}

arrhythmia_terms <-
c("ventricular arrhythmia",
  "torsades de pointes"
  )

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          pattern = vec_to_grep_pattern(arrhythmia_terms),
                          "cardiac arrhythmia"
         ))


```


## Cardiomyopathy

```{r cardiomyopathy}

cardiomyopathy_terms <- c("nonischemic cardiomyopathy")

gwas_study_info =
gwas_study_info |> 
 mutate(collected_all_disease_terms  = 
          stringr::str_replace_all(collected_all_disease_terms,
                                  pattern = vec_to_grep_pattern(cardiomyopathy_terms),
                                   "cardiomyopathy"
                          )  
        )

```

## Celiac disease

```{r celiac_disease}

gwas_study_info = 
gwas_study_info |> 
 mutate(collected_all_disease_terms  = 
          stringr::str_replace_all(collected_all_disease_terms,
                                  pattern = vec_to_grep_pattern("refractory celiac disease"),
                                   "celiac disease"
                          )  
        )


```

## Central and perpheral ertigo

```{r central_origin_vertigo}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern("central nervous system origin vertigo"),
                          "central origin vertigo"
         ))


gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern("peripheral vertigo"),
                          "peripheral or central vertigo"
         ))
  

```

## Cerebral atherosclerosis

```{r cerebral_atherosclerosis}

# if Brain vascular atherosclerosis in DISEASE/TRAIT, then change vascular brain injury to  cerebral atherosclerosis 

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         ifelse(grepl("Brain vascular atherosclerosis", 
                     `DISEASE/TRAIT`,
                     ignore.case = TRUE),
                stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern("vascular brain injury"),
                          "cerebral atherosclerosis"
         ),
         collected_all_disease_terms
         )
  )

```

## Cerebrovascular disease

```{r cerebrovascular_disease}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
        ifelse(grepl("Vascular brain injury", 
                     `DISEASE/TRAIT`,
                     ignore.case = TRUE),
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern(c(
                             "vascular brain injury"
                             )
                             ),
                          "cerebrovascular disease"
         ),
         collected_all_disease_terms)
  )

```

## Occlusion of cerebral arteries

```{r occlusion_of_cerebral_arteries}

# if Brain vascular stenosis in DISEASE/TRAIT, then change vascular brain injury to  occlusion of cerebral arteries

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         ifelse(grepl("Brain vascular stenosis", 
                     `DISEASE/TRAIT`,
                     ignore.case = TRUE),
                stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern("vascular brain injury"),
                          "occlusion of cerebral arteries"
         ),
         collected_all_disease_terms
         )
  )

```

## Cerebral infarction

```{r cerebral_infarction}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
          ifelse(grepl("Ischemic stroke \\(cardioembolic\\)", 
                     `DISEASE/TRAIT`,
                     ignore.case = TRUE),
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern(c(
                             "cardiac embolism"
                             )),
                          "cardioembolic stroke"
         ),
         collected_all_disease_terms
         )
  )

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern(c(
                             "mri defined brain infarct",
                             "cardioembolic stroke")
                             ),
                          "cerebral infarction"
         ))

```

## cervical cancer, dysplasia -> cervical cancer, cervical dysplasia

```{r cervical_cancer_dysplasia}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern("cervical cancer, dysplasia"),
                          "cervical cancer, cervical dysplasia"
         ))


```

## Chronic kidney disease

```{r chronic_kidney_disease}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern("stage 5 chronic kidney disease"),
                          "chronic kidney disease"
         ))

```

## Chronic non-alcoholic pancreatitis

```{r chronic_non_alcoholic_pancreatitis}

# if Non-alcoholic chronic pancreatitis  in `DISEASE/TRAIT`,
# then non-alcoholic pancreatitis to chronic pancreatitis

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         ifelse(grepl("Non-alcoholic chronic pancreatitis", 
                     `DISEASE/TRAIT`,
                     ignore.case = TRUE),
                stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern("non-alcoholic pancreatitis"),
                          "chronic pancreatitis"
         ),
         collected_all_disease_terms
         )
  )

```


## Chronic rhinosinusitis 

```{r chronic_rhinosinusitis}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern("chronic rhinosinusitis with nasal polyps"),
                          "chronic rhinosinusitis"
         ))

```

## Chronic sinusitis

```{r chronic_sinusitis}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern("chronic sinus infection"),
                          "chronic sinusitis"
         ))

```

## Crohn's Disease

```{r crohns_disease}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern("perianal crohns disease"),
                          "crohns disease"
         ))

```

## Cholangitis

```{r cholangitis}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern("sclerosing cholangitis"),
                          "cholangitis"
         ))

```

## Cluster headache syndrome

```{r cluster_headache_syndrome}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern("cluster headache"),
                          "cluster headache syndrome"
         ))


```

## Circumscribed brain atrophy

```{r circumscribed_brain_atrophy}

brain_atropy <- c("frontotemporal dementia",
                  "grn-related frontotemporal lobar degeneration with tdp43 inclusions",
                  "pick disease")

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern(brain_atropy),
                          "circumscribed brain atrophy"
         ))


```

## Contact dermatitis

```{r}

pattern <- vec_to_grep_pattern("contact dermatitis due to nickel")


gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms = 
         stringr::str_replace_all(collected_all_disease_terms,
                          pattern = pattern,
                          "contact dermatitis"
         ))

```

## Congenital heart disease/s

```{r congenital_heart_disease}

congenital_heart_disease_terms <- c(
  "heart septal defect",  
  "atrial heart septal defect",
  
  "congenital left-sided heart lesions",
  "congenital right-sided heart lesions",
  
  "congenital anomaly of the great arteries", # equiv = "Congenital malformation of great arteries, unspecified"
  
  # malformation of cardiac septum, 
    "abnormal cardiac septum morphology",
    "atrioventricular canal defect",
  "ventricular outflow obstruction"
)
  

url <- "http://www.ebi.ac.uk/ols4/api/ontologies/efo/terms/http%253A%252F%252Fwww.ebi.ac.uk%252Fefo%252FEFO_0005207/descendants"

congenital_heart_disease_terms <- c(congenital_heart_disease_terms,
                                  get_descendants(url)
                                  ) |>
  unique()

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          pattern = vec_to_grep_pattern(congenital_heart_disease_terms),
                          "congenital heart disease"
         )
  )

# if DISEASE/TRAIT contains Congenital heart disease 
# then replace abnormal cardiovascular system morphology with congenital heart disease
gwas_study_info = 
  gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         ifelse(grepl("Congenital heart disease", 
                     `DISEASE/TRAIT`,
                     ignore.case = TRUE),
                stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern("abnormal cardiovascular system morphology"),
                          "congenital heart disease"
         ),
         collected_all_disease_terms
         )
  )


url <- "http://www.ebi.ac.uk/ols4/api/ontologies/mondo/terms/http%253A%252F%252Fpurl.obolibrary.org%252Fobo%252FMONDO_0024239/descendants"

more_congenital_heart_disease_terms <- get_descendants(url)

congenital_heart_disease_terms <- c(congenital_heart_disease_terms,
                                  more_congenital_heart_disease_terms
                                  ) |>
  unique() |>
  str_length_sort()


gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          pattern = vec_to_grep_pattern(congenital_heart_disease_terms),
                          "congenital anomaly of cardiovascular system"
         )
  )



```

## Cardiac congenital anomalies

```{r cardiac_congenital_anomalies}

cardiac_congenital_anomalies_terms <- c(
  "bicuspid aortic valve"
)

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          pattern = vec_to_grep_pattern(cardiac_congenital_anomalies_terms),
                          "cardiac congenital anomalies"
         )
  )


```

## Congenital hypertrophic pyloric stenosis

```{r congenital_hypertrophic_pyloric_stenosis}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern("infantile hypertrophic pyloric stenosis"),
                          "congenital hypertrophic pyloric stenosis"
         ))

```

## Congenital hypothyroidism

```{r congenital_hypothyroidism}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern("congenital hypothyroidism due to developmental anomaly"),
                          "congenital hypothyroidism"
         ))

```

## Congestive heart failure (CHF) NOS

```{r congestive_heart_failure}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern(c(
                            "diastolic heart failure",
                            "systolic heart failure")
                            ),
                          "congestive heart failure \\(chf\\) nos"
         ))

```

## Coronary artery aneurysm and dissection

```{r coronary_artery_aneurysm_and_dissection}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern(c(
                            "coronary aneurysm",
                            "coronary artery dissection")
                            ),
                          "coronary artery aneurysm and dissection"
         ))

# Aortic aneurysm

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern(c(
                            "thoracic aortic aneurysm")
                            ),
                          "aortic aneurysm"
         ))

```

## Creutzfeldt jacob disease

```{r creutzfeldt_jacob_disease}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern(
                          c("sporadic creutzfeld jacob disease",
                            "creutzfeldt jacob disease",
                            "creutzfeldt-jacob disease")
                          ),
                          "creutzfeldt-jakob disease"
         ))


```

## Cryoglobulinemia

```{r cryoglobulinemia}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern("cryoglobulinemia"),
                          "cryoglobulinaemia"
         ))

```

## Chronic pain

```{r chronic_pain}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern(c(
                            "chronic widespread pain",
                            "multisite chronic pain",
                            "chronic musculoskeletal pain",
                            "chronic pain syndrome")
                            ),
                          "chronic pain"
         ))

```

## Cystic fibrosis with intestinal manifestations

```{r cystic_fibrosis_with_intestinal_manifestations}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern("cystic fibrosis associated meconium ileus"),
                          "cystic fibrosis with intestinal manifestations"
         ))

```

## Dental caries

```{r dental_caries}


dental_caries_terms <- c("pit and fissure surface dental caries",
                         "smooth surface dental caries",
                         "primary dental caries",
                         "enamel caries",
                         "permanent dental caries"
                         )

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          pattern = vec_to_grep_pattern(dental_caries_terms),
                          "dental caries"
         )
)

```

## Dermatomyositis

```{r dermatomyositis}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern("clinically amyopathic dermatomyositis"),
                          "dermatomyositis"
         ))

```

## Dilated cardiomyopathy

```{r dilated_cardiomyopathy}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern("idiopathic dilated cardiomyopathy"),
                          "dilated cardiomyopathy"
         ))

```

## Disturbances of sensation of smell and taste

```{r disturbances_of_sensation_of_smell_and_taste}

terms <- c("ageusia",
           "abnormality of the sense of smell")

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          pattern = vec_to_grep_pattern(terms),
                          "disturbances of sensation of smell and taste"
         )
  )


```

## Disorders of tooth development

```{r disorders_of_tooth_development}

tooth_dev_terms <- c("dental enamel hypoplasia",
                     "tooth agenesis",
                     "molar-incisor hypomineralization"
                     )

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          pattern = vec_to_grep_pattern(tooth_dev_terms),
                          "disorders of tooth development"
         )
  )

```

## Depressive episode

```{r depressive_episode}

depress_epi <- c("depressive",
                 "depression",
                 "depressive disorder"
                 )

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern(depress_epi),
                          "depressive episode"
         ))

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern(depress_epi),
                          "depressive episode"
         ))

```

## Disorders of purine and pyrimidine metabolism 

```{r disorders_of_purine_and_pyrimidine_metabolism}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern("hyperuricemia"),
                          "disorders of purine and pyrimidine metabolism"
         ))

```

## Disorders of refraction and accommodation

```{r disorders_of_refraction_and_accommodation}

refractive_terms <- c("hyperopia",
                      "refractive error"
                      )

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          pattern = vec_to_grep_pattern(refractive_terms),
                          "disorders of refraction and accommodation"
         )
  )

```

## Diseases of white blood cells

```{r diseases_of_white_blood_cells}

wbc_terms <- c("leukopenia")

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          pattern = vec_to_grep_pattern(wbc_terms),
                          "diseases of white blood cells"
         )
  )

```

## Disorders of amino-acid metabolism

```{r disorders_of_amino_acid_metabolism}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern("amino acid metabolism disease"),
                          "disorders of amino-acid metabolism"
         ))

```

## Drug allergy

```{r drug_allergy}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern("beta-lactam allergy"),
                          "drug allergy"
         ))

```

## Degeneration of macula and posterior pole

```{r degeneration_of_macula_and_posterior_pole}

macular <- c("macular degeneration",
             "atrophic macular degeneration",
             "retinal drusen"
             )

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern(macular),
                          "degeneration of macula and posterior pole"
         ))

```

## Essential hypertension

```{r essential_hypertension}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern(
                            c(
                              "early onset hypertension",
                              "treatment-resistant hypertension")),
                          "essential hypertension"
         ))


```

## Exstrophy of urinary bladder

```{r exstrophy_of_urinary_bladder}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern("bladder exstrophy"),
                          "exstrophy of urinary bladder"
         ))

```

## Febrile seizures

```{r febrile_seizures}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern(c(
                                                "mmr-related febrile seizures",
                                                "febrile seizure")
                                              ),
                          "febrile convulsions"
         ))


```

## Food allergy

```{r food_allergy}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                                  vec_to_grep_pattern(
                                    c("peanut allergy",
                                      "milk allergy",
                                      "egg allergy",
                                      "wheat allergic reaction"
                                     )),
                          "food allergy"
         ))
```

## Fuchs endothelial corneal dystrophy

```{r fuchs_endothelial_corneal_dystrophy}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern("fuchs endothelial dystrophy"),
                          "fuchs endothelial corneal dystrophy"
         ))

```

## Glaucoma

```{r glaucoma}

url <- "http://www.ebi.ac.uk/ols4/api/ontologies/efo/terms/http%253A%252F%252Fpurl.obolibrary.org%252Fobo%252FMONDO_0005041/descendants"

glaucoma_terms <- get_descendants(url)

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          pattern = vec_to_grep_pattern(glaucoma_terms),
                          "glaucoma"
         ))


```

## Hordeolum and other deep inflammation of eyelid

```{r hordeolum_and_other_deep_inflammation_of_eyelid}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern("hordeolum"),
                          "hordeolum and other deep inflammation of eyelid"
         ))

```

## Graft vs host disease

```{r graft_vs_host_disease}

graft_vs_host_terms <- c("chronic graft versus host disease",
                         "chronic graft vs. host disease",
                         "acute graft versus host disease",
                         "acute graft vs. host disease"
                         )

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern(graft_vs_host_terms),
                          "graft versus host disease"
         ))

```

## Gingival and periodontal diseases

```{r gingival_and_periodontal_diseases}

ginival_and_periodontal_terms <- c("periodontal pocket",
                                   "periodontal disorder",
                                   "gingival disease",
                                   "gingival bleeding"
                               )

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          pattern = vec_to_grep_pattern(ginival_and_periodontal_terms),
                          "gingival and periodontal diseases"
         )
  )


```

## Hemiplegia

```{r hemiplegia}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern(c("hemiparesis")),
                          "hemiplegia"
         ))

```

## Hearing loss

```{r hearing_loss}

gwas_study_info  = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern("age-related hearing impairment"),
                          "presbycusis"
         ))

gwas_study_info  = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern(c(
                            "deafness",
                            "noise-induced hearing loss")
                            ),
                          "hearing loss"
         ))

```

## Hereditary hemochromatosis

```{r}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern("hereditary hemochromatosis type 1"),
                          "hereditary hemochromatosis"
         ))

```

## Hypercholesterolemia

```{r hypercholesterolemia}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern(c(
                            "familial hypercholesterolemia"
                            )),
                          "hypercholesterolemia"
         ))

```

## Hyperlipidemia

```{r hyperlipidemia}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern(c(
                            "familial hyperlipidemia"
                            )),
                          "hyperlipidemia"
         ))

```

## HIV disease resulting in encephalopathy

```{r hiv_disease_resulting_in_encephalopathy}

gwas_study_info = 
  gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern(c(
                                               "aids dementia",
                                               "hiv-associated neurocognitive disorder")
                                              ),
                          "hiv disease resulting in encephalopathy"
         ))

```

## Inherited retinal dystrophy

```{r inherited_retinal_dystrophy}

url <-  "http://www.ebi.ac.uk/ols4/api/ontologies/mondo/terms/http%253A%252F%252Fpurl.obolibrary.org%252Fobo%252FMONDO_0019118/descendants"

inherited_retinal_dystrophy_terms <- get_descendants(url)

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          pattern = vec_to_grep_pattern(inherited_retinal_dystrophy_terms),
                          "hereditary retinal dystrophy"
         ))

```

## Intentional self-harm by unspecified means

```{r intentional_self_harm_by_unspecified_means}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern(c("self-injurious behavior",
                                                "self-injurious ideation")),
                          "intentional self-harm by unspecified means"
         ))

```

## Infertility (male)

```{r}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
                  stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern(
                            c("azoospermia",
                              "sertoli cell-only syndrome")
                            ),
                          "male infertility"
                          )  )

```

## Induratio penis plastica

```{r indura}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
                  stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern(
                            c("peyronie disease")
                            ),
                          "induratio penis plastica"
                          )  )



```

## Intracerebral hemorrhage

```{r intra_hemo}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
                  stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern(
                            c("non-lobar intracerebral hemorrhage",
                              "lobar intracerebral hemorrhage")
                            ),
                          "intracerebral hemorrhage"
                          )  )


```

## Idiopathic generalized epilepsy

```{r idiopathic_generalized_epilepsy}

url <- "http://www.ebi.ac.uk/ols4/api/ontologies/snomed/terms/http%253A%252F%252Fsnomed.info%252Fid%252F36803009/descendants"

idiopathic_generalized_epilepsy_terms <- get_descendants(url)

idiopathic_generalized_epilepsy_terms = stringr::str_remove_all(
                                        idiopathic_generalized_epilepsy_terms, 
                                        " \\(disorder\\)$"
                                        )

idiopathic_generalized_epilepsy_terms = c("epilepsy with generalized tonic-clonic seizures",
                                          idiopathic_generalized_epilepsy_terms
                                         )


gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          pattern = vec_to_grep_pattern(idiopathic_generalized_epilepsy_terms),
                          "generalized idiopathic epilepsy and epileptic syndromes"
         ))


```

## Idiopathic thrombocytopenic purpura

```{r idiopathic_thrombocytopenic_purpura}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern("autoimmune thrombocytopenic purpura"),
                          "idiopathic thrombocytopenic purpura"
         ))


```

## Juvenile idiopathic arthritis

```{r juvenile_idiopathic_arthritis}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern("rheumatoid factor-negative juvenile idiopathic arthritis"),
                          "juvenile idiopathic arthritis"
         ))

```

## Keratoconjunctivitis

```{r keratoconjunctivitis}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern("keratoconjunctivitis sicca"),
                          "keratoconjunctivitis"
         ))

```

## Learning disorder

```{r learning_disorder}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern(c("dyslexia",
                                                "mathematics disorder",
                                                "disorder of written expression"
                                                )
                                              ),
                          "learning disorder"
         ))

```

## Lewy body dementia

```{r lewy_body_dementia}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern("lewy body attribute"),
                          "lewy body dementia"
         ))
```



## Rheumatoid arthritis

```{r}

ra_terms <- c("acpa-positive rheumatoid arthritis",
               "acpa-negative rheumatoid arthritis",
              "adult-onset stills disease")

gwas_study_info = 
gwas_study_info |> 
 mutate(collected_all_disease_terms  = 
          stringr::str_replace_all(collected_all_disease_terms,
                                  pattern = vec_to_grep_pattern(ra_terms),
                                   "rheumatoid arthritis"
                          )  
        )

```


## Migraine 

```{r migraine}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern(
                            c("migraine with aura",
                              "migraine without aura",
                              "migraine disorder"
                              )
                            ),
                          "migraine"
         ))

```

## Multiple sclerosis

```{r multiple_sclerosis}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern("relapsing-remitting multiple sclerosis"),
                          "multiple sclerosis"
         ))

```

## Myasthenia gravis

```{r}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern("late-onset myasthenia gravis"),
                          "myasthenia gravis"
         ))


```

## narcolepsy-cataplexy syndrome -> Narcolepsy and cataplexy

```{r narcolepsy_and_cataplexy}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern("narcolepsy-cataplexy syndrome"),
                          "narcolepsy and cataplexy"
         ))


```

## Nephrolithiasis

```{r nephrolithiasis}

nephro_terms <- c("uric acid nephrolithiasis",
                   "calcium phosphate nephrolithiasis",
                    "calcium oxalate nephrolithiasis",
                   "struvite nephrolithiasis")


gwas_study_info =
gwas_study_info |> 
 mutate(collected_all_disease_terms  = 
          stringr::str_replace_all(collected_all_disease_terms,
                                  pattern = vec_to_grep_pattern(nephro_terms),
                                   "nephrolithiasis"
                          )  
        )


```

## Nephritis and nephropathy with pathological lesion

```{r nephritis_and_nephropathy_with_pathological_lesion}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern(c("acute tubulointerstitial nephritis",
                                              "iga glomerulonephritis",
                                              "membranous glomerulonephritis",
                                              "lupus nephritis")
                                              ),
                          "nephritis and nephropathy with pathological lesion"
         ))

```

## Neurofibromatosis

```{r}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern(
                            c("neurofibromatosis type 1",
                              "neurofibromatosis type 2")
                          ),
                          "neurofibromatosis"
         ))

```

## Neuromyelitis optica 

```{r}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern(
                            c("aquaporin-4 antibody positive neuromyelitis optica",
                              "aquaporin-4 antibody negative neuromyelitis optica",
                              "aqp4-igg-positive neuromyelitis optica",
                              "aqp4-igg-negative neuromyelitis optica"
                            )
                          ),
                          "neuromyelitis optica"
         ))

```

## Noninflammatory disorders of vagina

```{r non_inflam_vagina}

non_inflam_terms <- c("abnormal vaginal discharge itching",
                      "abnormal vaginal discharge smell",
                      "vaginal discharge")

gwas_study_info = 
  gwas_study_info |>
   mutate(collected_all_disease_terms  = 
          stringr::str_replace_all(collected_all_disease_terms,
                                  pattern = 
                                    vec_to_grep_pattern(
                                      non_inflam_terms
                                      ),
                                   "noninflammatory disorders of vagina"
                          )  
        )
  

```

## Obesity

```{r obesity}

gwas_study_info = 
gwas_study_info |> 
 mutate(collected_all_disease_terms  = 
          stringr::str_replace_all(collected_all_disease_terms,
                                  pattern = 
                                    vec_to_grep_pattern(
                                      c("morbid obesity",
                                         "metabolically healthy obesity"
                                        )
                                      ),
                                   "obesity"
                          )  
        )

```

## Obsessive-compulsive disorder

```{r obsessive_compulsive_disorder}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern(c
                                              ("obsessive-compulsive trait",
                                                "obsessive-compulsive")
                                              ),
                          "obsessive-compulsive disorder"
         ))

```

## Osteonecrosis

```{r osteonecrosis}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern("idiopathic osteonecrosis of the femoral head"),
                          "osteonecrosis"
         ))

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern("osteoradionecrosis"),
                          "osteonecrosis"
         ))

```

## Other epilepsy 

```{r other_epilepsy}

gwas_study_info =
  gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern(
                            c("mesial temporal lobe epilepsy with hippocampal sclerosis",
                              "rolandic epilepsy")
                            ),
                          "epilepsy"
         )) 

```

## Chromosomal anomalies

```{r other_deletions_from_the_autosomes}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern(c(
                            "22q11.2 deletion syndrome",
                            "abnormality of chromosome segregation",
                            "fragile x syndrome")
                            ),
                          "chromosomal anomalies"
         ))

```

## Opiod dependence

```{r opioid_dependence}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern(c("heroin dependence",
                                              "opioid use disorder")),
                          "opioid dependence"
         ))

```

## Other and unspecified cirrhosis of liver 

```{r other_and_unspecified_cirrhosis_of_liver}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern(c("hepatitis c induced liver cirrhosis",
                                                "biliary liver cirrhosis")
                                              ),
                          "other and unspecified cirrhosis of liver"
         ))


```

## Other cardiac conduction disorders

```{r other_cardiac_conduction_disorder}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern(c("familial long qt syndrome")
                                              ),
                          "other cardiac conduction disorders"
         ))

```

## Other cerebral degenerations

```{r other_cerebral_degeneration}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern(c("brain atrophy",
                                                "hippocampal atrophy"
                                                )
                                              ),
                          "other cerebral degenerations"
         ))

```

## Other specified degenerative diseases of nervous system

```{r other_specified_degenerative_diseases_of_nervous_system}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern(c("corticobasal degeneration disorder"
                                                )
                                              ),
                          "other specified degenerative diseases of nervous system"
         ))

# progressive supranuclear palsy -> Dementia with cerebral degenerations

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern("progressive supranuclear palsy"),
                          "dementia with cerebral degenerations"
         ))

```

## Other chronic nonalcoholic liver disease

```{r other_chronic_nonalcoholic_liver_disease}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern(c("non-alcoholic fatty liver disease"
                                                )),
                          "other chronic nonalcoholic liver disease"
         ))

```

## Other disorders of bone and cartilage

```{r other_disorders_of_bone_and_cartilage}

other_bone_cartilage_terms <- c("tietze syndrome"
                               )

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          pattern = vec_to_grep_pattern(other_bone_cartilage_terms),
                          "other disorders of bone and cartilage"
         )
  )

```

## Other eating disorders

```{r other_eating_disorders}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern(c("binge eating")
                                              ),
                          "other eating disorders"
         ))

```

## Other specified inflammatory liver diseases

```{r other_specified_inflammatory_liver_diseases}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern(c("primary biliary cholangitis",
                                                "primary sclerosing cholangitis"
                                                )
                                              ),
                          "non-alcoholic steatohepatitis"
         ))

```

## Other haemoglobinopathies

```{r other_haemoglobinopathies}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern("hemoglobin e disease"),
                          "other haemoglobinopathies"
         ))

```

## Other paralytic syndromes

```{r other_paralytic_syndromes}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern(
                            c("paraplegia",
                              "quadriplegia")
                            ),
                          "other paralytic syndromes"
         ))


```

## Parkinsons disease

```{r parkinsons_disease}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern("young adult-onset parkinsonism"),
                          "parkinsons disease"
         ))

```

## Perinatal jaundice

```{r perinatal_jaundice}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern("neonatal jaundice"),
                          "perinatal jaundice"
         ))


```

## Periodontitis

```{r periodontitis}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern("aggressive periodontitis"),
                          "periodontitis"
         ))

```

## Phlebitis and thrombophlebitis

```{r phlebitis_and_thrombophlebitis}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern(c("thrombophlebitis")),
                          "phlebitis and thrombophlebitis"
         ))

```

## Phobias

```{r phobias}

# social anxiety disorder -> social phobias
gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern("social anxiety disorder"),
                          "social phobias"
         ))

# specific phobia -> Specific \\(isolated\\) phobias

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern("specific phobia"),
                          "specific \\(isolated\\) phobias"
         ))

```

## Primary ovarian failure

```{r primary_ovarian_failure}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern("primary ovarian insufficiency"),
                          "primary ovarian failure"
         ))

```

## Premature menopause and other ovarian failure

```{r premature_menopause_and_other_ovarian_failure}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern("premature menopause"),
                          "premature menopause and other ovarian failure"
         ))


```

## Primary hyperaldosteronism

```{r primary_hyperaldosteronism}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern("primary aldosteronism"),
                          "primary hyperaldosteronism"
         ))

```

## Proteinuria

```{r proteinuria}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern(c("albuminuria",
                                                "moderate albuminuria")
                                              ),
                          "proteinuria"
         ))

```

## Prurigo

```{r prurigo}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern("prurigo"),
                          "other prurigo"
         ))

```

## Precocious puberty

```{r precocious_puberty}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern("central precocious puberty"),
                          "precocious puberty"
         ))


```

## Psoriasis

```{r}

psoriasis_terms <- c("cutaneous psoriasis")


gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          pattern = vec_to_grep_pattern(psoriasis_terms),
                          "psoriasis")
  )

```

## Psychosis (pyschotic, pyschotic symptoms)

```{r psychosis}

psychosis_terms <- c("psychotic",
                     "psychotic symptoms"
                     )

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          pattern = vec_to_grep_pattern(psychosis_terms),
                          "psychosis")
  )

```

## Pulmonary fibrosis

```{r pulmonary_fibrosis}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern("idiopathic pulmonary fibrosis"),
                          "pulmonary fibrosis"
         ))

```

## Retinoschisis and retinoschisis and retinal cysts

```{r retinoschisis_and_retinal_cysts}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern("retinoschisis"),
                          "retinoschisis and retinal cysts"
         ))

```

## Separation of retinal layers

```{r separation_of_retinal_layers}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern(c("central serous retinopathy",
                                              "chronic central serous retinopathy")
                                              ),
                          "separation of retinal layers"
         ))

```

## Rash and other nonspecific skin eruption

```{r rash_and_other_nonspecific_skin_eruption}

rash_terms <- c(
                "maculopapular eruption"
                )

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          pattern = vec_to_grep_pattern(rash_terms),
                          "rash and other nonspecific skin eruption"
         )
  )

```

## Rhinitis

```{r rhinitis}

rhinitis_terms <- c("non-allergic rhinitis")

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          pattern = vec_to_grep_pattern(rhinitis_terms),
                          "rhinitis"
         )
  )

```

## Sciatica

```{r sciatica}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern(c("ldh-related sciatica")
                                              ),
                          "sciatica"
         )) 

```

	
## Schizoaffective disorder

```{r schizoaffective_disorder}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern("schizoaffective disorder-bipolar type"),
                          "schizoaffective disorder")
         )

```

## Schizophrenia

```{r schizophrenia}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern("treatment refractory schizophrenia"),
                          "schizophrenia")
         )

```

## Scoliosis

```{r scoliosis}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern("adolescent idiopathic scoliosis"),
                          "scoliosis")
         )

```

## Sickle cell disease and related diseases

```{r sickle_cell_disease_and_related_diseases}

# if DISEASE/TRAIT contains sickle cell disease, change sickle cell disease and related diseases to sickle cell anemia
gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         ifelse(grepl("sickle cell disease", 
                     `DISEASE/TRAIT`,
                     ignore.case = TRUE),
                stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern("sickle cell disease and related diseases"),
                          "sickle cell anemia"
         ),
         collected_all_disease_terms
         )
  )

```

## Sleep apnea 

```{r}

sleep_apnea_terms <- c("sleep apnea during non-rem sleep",
                      "sleep apnea during rem sleep",
                      "obstructive sleep apnea")

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          pattern = vec_to_grep_pattern(sleep_apnea_terms),
                          "sleep apnea")
  )


```

## Sleep disorders

```{r sleep_disorders}

sleep_disorder_terms <- c("sleepiness",
                          "somnambulism",
                          "rem sleep behavior disorder",
                          "periodic limb movement disorder",
                          "bruxism"
                         )


gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          pattern = vec_to_grep_pattern(sleep_disorder_terms),
                          "sleep disorders"
         )
  )

```

## Speech and language disorder

```{r speech_and_language_disorder}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern(c("specific language impairment",
                                                "language impairment",
                                                "social communication impairment"
                                                )
                                              ),
                          "speech and language disorder"
         ))

# if DISEASE/TRAIT contains Developmental stuttering
# then replace stuttering with speech and language disorder
gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         ifelse(grepl("Developmental stuttering", 
                     `DISEASE/TRAIT`,
                     ignore.case = TRUE),
                stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern("stuttering"),
                          "speech and language disorder"
         ),
         collected_all_disease_terms
         )
  )


```

## chronic pharyngitis, nasopharyngitis

```{r chronic_pharyngitis_nasopharyngitis}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern(c("chronic pharyngitis, nasopharyngitis")
                                              ),
                          "chronic pharyngitis, chronic nasopharyngitis"
         ))

```

## Staphylococcus infections

```{r staphylococcus_infections}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern(
                            c("staphylococcus aureus infection",
                              "skin and soft tissue staphylococcus aureus infection",
                              "methicillin-resistant staphylococcus aureus infection")),
                          "staphylococcus infections"
         ))


```

## Strabismus

```{r}

url <-  "http://www.ebi.ac.uk/ols4/api/ontologies/doid/terms/http%253A%252F%252Fpurl.obolibrary.org%252Fobo%252FDOID_540/descendants"

strabismus_terms <- get_descendants(url)

strabismus_terms = c("non-accomodative esotropia",
                     strabismus_terms)

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          pattern = vec_to_grep_pattern(strabismus_terms),
                          "strabismus"
         ))


```


## Stroke

```{r stroke}

other_nonspec_stroke <- c("large artery stroke",
                         "small vessel stroke",
                         "stroke outcome",
                         "stroke disorder"
                         )

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern(other_nonspec_stroke),
                          "stroke"
         ))

```

## Stuttering, tics -> Tics and stuttering 

```{r tics_and_stuttering}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         ifelse(grepl("PheCode 313.2", 
                     `DISEASE/TRAIT`,
                     ignore.case = TRUE),
                stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern(c("stuttering, tics")),
                          "tics and stuttering"
                          ),
         collected_all_disease_terms
)
)


gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         ifelse(grepl("PheCode 333.3", 
                     `DISEASE/TRAIT`,
                     ignore.case = TRUE),
                stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern(c("stuttering, tics")),
                          "tics and choreas"
         ),
         collected_all_disease_terms
))


```

## Systemic sclerosis

```{r systemic_sclerosis}

terms <- c("diffuse scleroderma",
           "limited scleroderma",
           "anti-centromere-antibody-positive systemic scleroderma",
           "anti-topoisomerase-i-antibody-positive systemic scleroderma")

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern(terms),
                          "systemic sclerosis"
         ))

```

## Thyrotoxicosis with or without goiter

```{r thyrotoxicosis_with_or_without_goiter}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern("hyperthyroidism"),
                          "thyrotoxicosis with or without goiter"
         ))

```

## Tinea 

```{r tinea}

# if DISEASE/TRAIT contains Ringworm, replace tinea with tinea corporis
gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         ifelse(grepl("Ringworm", 
                     `DISEASE/TRAIT`,
                     ignore.case = TRUE),
                stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern("tinea"),
                          "tinea corporis"
         ),
         collected_all_disease_terms
         )
  )


```

## Type 1 diabetes

```{r}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern("latent autoimmune diabetes in adults"),
                          "type 1 diabetes mellitus"
         ))

```

##  diabetic foot, neuropathy, type 1 diabetes mellitus, type 2 diabetes mellitus

```{r diabetic_foot_neuropathy}


icd_terms <- c("type 1 diabetes with foot ulcer",
               "type 2 diabetes with foot ulcer",
               "type 1 diabetes with diabetic neuropathy",
               "type 2 diabetes with diabetic neuropathy")

icd_replacement <- paste0(icd_terms, collapse = ", ")

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern("diabetic foot, neuropathy, type 1 diabetes mellitus, type 2 diabetes mellitus"),
                          icd_replacement
         ))


```

## Type 2 diabetes with a ophthalmic manifestations

```{r type_2_eye_terms}

type_2_eye_terms <- c("diabetes mellitus type 2 associated cataract",
                      "diabetic maculopathy, type 2 diabetes mellitus",
                      "diabetic macular edema, type 2 diabetes mellitus",
                      "proliferative diabetic retinopathy, type 2 diabetes mellitus",
                      "macrovascular complications of diabetes, type 2 diabetes mellitus"
                     )

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern(type_2_eye_terms),
                          "type 2 diabetes with ophthalmic manifestations"
         ))

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern(type_2_eye_terms),
                          "type 2 diabetes with ophthalmic manifestations"
         ))

# for 30487263, all discovery samples are type 2 diabetes - so 
# replace proliferative diabetic retinopathy with type 2 diabetes with ophthalmic manifestations
gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         ifelse(PUBMED_ID == 30487263,
                stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern("proliferative diabetic retinopathy"),
                          "type 2 diabetes with ophthalmic manifestations"
         ),
         collected_all_disease_terms
         )
  )

# for pubmed id: 31482010
# all samples are type 2 diabetes
# so replace diabetic retinopathy with type 2 diabetes with ophthalmic manifestations
gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         ifelse(PUBMED_ID == 31482010,
                stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern("proliferative diabetic retinopathy"),
                          "type 2 diabetes with ophthalmic manifestations"
         ),
         collected_all_disease_terms
         )
)

# neuropathic pain, type 2 diabetes mellitus
# change to type 2 diabetes with diabetic neuropathy
gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
                stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern("neuropathic pain, type 2 diabetes mellitus"),
                          "type 2 diabetes with ophthalmic manifestations"
         )
  )
         
```

## Rash, Petechiae

```{r rash_petechiae}

# for pubmed id: 37469131
# and DISEASE/TRAIT contains 'rush' (a tyop of rash)
# replace purpura with rash, petechiae

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         ifelse(PUBMED_ID == 37469131 & grepl("rush", `DISEASE/TRAIT`, ignore.case = TRUE),
                stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern("purpura"),
                          "rash, petechiae"
         ),
         collected_all_disease_terms
         )
  )

```



## Treatment resistant 

```{r treatment_resistant}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_remove_all(collected_all_disease_terms,
                       "^treatment-resistant |^treatment resistant |^treatment-resistant "
         ))

```

## Unspecified condition associated with female genital organs and menstrual cycle

```{r menstrual}

gwas_study_info = 
  gwas_study_info |>
  mutate(collected_all_disease_terms = 
    ifelse(grepl("Menstruation", 
                `DISEASE/TRAIT`,
                 ignore.case = TRUE),
           str_replace_all(collected_all_disease_terms,
                           pattern = vec_to_grep_pattern("decreased attention"),
                           "unspecified condition associated with female genital organs and menstrual cycle"),
           collected_all_disease_terms
    )
    )

```


## Uveitis

```{r uveitis}

uveitis_terms <- c("anterior uveitis",
                   "iritis",
                   "vogt-koyanagi-harada disease",
                   "birdshot chorioretinopathy",
                   "multifocal choroiditis"
                   )

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          pattern = vec_to_grep_pattern(uveitis_terms),
                          "uveitis"
         ))

```

## Other specified retinal disorders 

```{r other_specified_retinal_disorders}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern(c("retinal edema"
                                              )),
                          "other specified retinal disorders"
         ))

```

## Other disorders of eyelids

```{r other_disorders_of_eyelids}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern(c("dermatochalasis",
                                                "filarial elephantiasis"
                                                )
                                              ),
                          "other disorders of eyelids"
         ))

```

## Other disorders of iris and ciliary body

```{r other_disorders_of_iris_and_ciliary_body}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern(c("iris disorder"
                                                )
                                              ),
                          "other disorders of iris and ciliary body"
         ))

```

## Background retinopathy and retinal vascular changes

```{r background_retinopathy_and_retinal_vascular_changes}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern(c("macular telangiectasia type 2"
                                                )
                                              ),
                          "background retinopathy and retinal vascular changes"
         ))

```

## Treatment resistant depression

```{r treatment_resistant_depression}

# if DISEASE/TRAIT contains Treatment resistant depression
# then replace depression with treatment resistant depression
gwas_study_info =
  gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         ifelse(grepl("Treatment resistant depression", 
                     `DISEASE/TRAIT`,
                     ignore.case = TRUE),
                stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern("depression"),
                          "treatment resistant depression"
         ),
         collected_all_disease_terms
         )
  )

gwas_study_info =
  gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         ifelse(grepl("treatment resistant depression", 
                     MAPPED_BACKGROUND_TRAIT,
                     ignore.case = TRUE),
                stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern("depression"),
                          "treatment resistant depression"
         ),
         collected_all_disease_terms
         )
  )

```


# Save

## How many unique traits are there now?

```{r n_unique_traits_end}

diseases <- stringr::str_split(pattern = ", ",
                               gwas_study_info$collected_all_disease_terms[gwas_study_info$collected_all_disease_terms != ""])  |>
  unlist() |>
  stringr::str_trim()

diseases <- unique(diseases)

print(length(diseases))


```

```{r save_data}

fwrite(
  gwas_study_info,
  here::here("output/gwas_cat/gwas_study_info_group_v2.csv")
  )

```
