---
title: "Matching GWAS traits to GBD causes / diseases"
author: "Isobel Beasley"
date: "2025-10-07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Set up

```{r load_libraries, message=FALSE}

library(dplyr)
library(stringr)
library(purrr)
library(tidyr)
library(data.table)

```

## Load Lancet priority diseases

```{r load_lancet_priority_diseases}

gbd_2019 <- readxl::read_xlsx(here::here("data/icd/lancet_conditions_icd10.xlsx"))

gbd_2019 =
  gbd_2019 |>
  select(cause = mapped_gbd_term,
         icd10 = icd_10_non_fatal,
         cause_hierarchy_level = gbd_level)


gbd_2019 =
  gbd_2019 |>
  filter(!is.na(icd10))

```


# Prepare GBD 2019 data for merging with GWAS traits

## Tidy ICD codes

### Make long format, separate multiple ICD code ranges into rows

```{r make_long_icd10_2019}

# expand multiple ICD codes into rows
gbd_2019 = 
  gbd_2019 |>
  mutate(icd10 = str_split(icd10, ",\\s*")) |>
  tidyr::unnest(icd10) 

# separate ICD code ranges into start and end codes
gbd_2019 =
  gbd_2019 |>
  tidyr::separate_wider_delim(col = "icd10",
                        delim = "-", 
                       names = c("start_icd10_code", "end_icd10_code"),
                       too_few = "align_end"
                       ) 

# if only one ICD code (i.e. no range), set start and end to be the same
gbd_2019 =
  gbd_2019 |>
  mutate(start_icd10_code = 
         ifelse(is.na(start_icd10_code),
                end_icd10_code,
                start_icd10_code)
         )

gbd_2019 =
  gbd_2019 |>
  distinct()

head(gbd_2019)

```

### Where icd10 code ranges are the same, but missing decimal places, add .9 to end code (to help with filtering ranges)

```{r}

gbd_2019 |>
  filter(start_icd10_code == end_icd10_code &
         !grepl(".", start_icd10_code, fixed = TRUE)
         ) |>
  arrange(start_icd10_code) |>
  head()

gbd_2019 = 
gbd_2019 |>
  mutate(end_icd10_code =
         ifelse(
         start_icd10_code == end_icd10_code &
         !grepl(".", start_icd10_code, fixed = TRUE),
         paste0(end_icd10_code, ".9"),
         end_icd10_code
         ) 
  )

```

### Seperate numeric and letter parts of ICD codes (to help with filtering ranges)

```{r separate_numeric_letter_icd10_2019}

# replace O9A513 with O99.9
gbd_2019 =
  gbd_2019 |>
  mutate(end_icd10_code =
        ifelse(end_icd10_code == "O9A513",
               "O99.9",
               end_icd10_code)
         )

# replace Z3A49 with Z39.2
gbd_2019 =
  gbd_2019 |>
  mutate(end_icd10_code =
        ifelse(end_icd10_code == "Z3A49",
               "Z39.2",
               end_icd10_code)
         )

# check start_icd10_code and end_icd10_code start with the same letter
# if not, need to fix these rows
  gbd_2019 |>
  filter(str_extract(start_icd10_code, "^[A-Z]") != 
         str_extract(end_icd10_code, "^[A-Z]")
         )

gbd_2019 =
  gbd_2019 |>
  mutate(icd10_code_letter = str_extract(start_icd10_code, "^[A-Z]")) 

gbd_2019 =
  gbd_2019 |>
  mutate(start_icd10_code_num = as.numeric(str_remove(start_icd10_code, 
                                                      "^[A-Z]")
                                           ),
         end_icd10_code_num = as.numeric(str_remove(end_icd10_code, 
                                                    "^[A-Z]")
                                         )
         )

```



# Merge GWAS traits with GBD causes

## Load Disease Mapping 

```{r load_gwas_study_info}

disease_mapping <- data.table::fread(
here::here("output/icd_map/gwas_disease_to_icd10_mapping.csv")
)

disease_mapping =
  disease_mapping |>
  mutate(icd10_code = sub("^([^-]+)-\\1$", "\\1", icd10_code))

disease_mapping =
  disease_mapping |>
    mutate(icd10_code_letter = str_extract(icd10_code, "^[A-Z]")) |>
    mutate(icd10_code_num = as.numeric(str_remove(icd10_code, "^[A-Z]")))

disease_mapping |>
  filter(is.na(icd10_code_num)) |>
  nrow()

disease_mapping =
  disease_mapping |>
  filter(!is.na(icd10_code_num)) 

disease_mapping = 
  disease_mapping |>
  mutate(icd10_code_num = as.numeric(icd10_code_num))

```

## Join disease mapping with GBD 2019 data

```{r join_gwas_gbd_2019}

disease_mapping_with_cause <- disease_mapping |>
  rowwise() |>
  mutate(
    cause = 
      list(
      gbd_2019$cause[
      which(icd10_code_num >= gbd_2019$start_icd10_code_num &
            icd10_code_num <= gbd_2019$end_icd10_code_num & 
            icd10_code_letter == gbd_2019$icd10_code_letter)
      ]),
    cause_hierarchy_level  = 
       list(
            gbd_2019$cause_hierarchy_level[
      which(icd10_code_num >= gbd_2019$start_icd10_code_num &
            icd10_code_num <= gbd_2019$end_icd10_code_num & 
            icd10_code_letter == gbd_2019$icd10_code_letter)
      ])
  ) |>
  ungroup()


disease_mapping_with_cause = 
  disease_mapping_with_cause |>
  tidyr::unnest(c(cause, cause_hierarchy_level),
                keep_empty = TRUE
                )


```

```{r check_mapping}

# K76.1 ULMS mapping to chronic liver disease, but is actually about downstream consequences of
# heart failure/disease 
disease_mapping_with_cause |> 
  filter(icd10_code == "K76.1" & grepl("UMLS", icd10_code_origin)) |> 
  select(`DISEASE/TRAIT`, cause) |> 
  distinct()  |> 
  head()

# remove this mapping 
disease_mapping_with_cause = 
  disease_mapping_with_cause |>
  filter(!(icd10_code == "K76.1" & grepl("UMLS", icd10_code_origin)))

```

```{r check_mapping_results}

print("Number of unique studies mapped to a cause:")
disease_mapping_with_cause |> 
  filter(!is.na(cause)) |> 
  pull(PUBMED_ID) |> 
  unique() |> 
  length()

print("Number of unique studies mapped to non-communicable diseases:")
# filter out infectious diseases
disease_mapping_with_cause |>
    dplyr::filter(!cause %in% c("HIV/AIDS",
                             "Tuberculosis",
                             "Malaria",
                             "Lower respiratory infections",
                             "Diarrhoeal diseases",
                             "Neonatal disorders",
                             "Tetanus",
                             "Diphtheria",
                             "Pertussis" ,
                             "Measles",
                             "Maternal disorders")) |>
    filter(!is.na(cause)) |> 
  pull(PUBMED_ID) |> 
  unique() |> 
  length()

```

# Check overlap (upset plot) of studies mapped to different causes

## Between type 1 diabetes, type 2 diabetes, and diabetes mellitus

```{r check_overlap_diabetes}

diabetes_studies = disease_mapping_with_cause |>
  filter(cause %in% c("Type 1 diabetes mellitus", 
                      "Type 2 diabetes mellitus", 
                      "Diabetes mellitus",
                      "Chronic kidney disease due to diabetes mellitus type 2",
                      "Chronic kidney disease due to diabetes mellitus type 1")) |>
  select(PUBMED_ID, cause) |>
  distinct()

library(ComplexHeatmap)

# create a binary matrix indicating which studies are mapped to which causes

diabetes_studies_wide = diabetes_studies |>
  tidyr::pivot_wider(names_from = cause, values_from = cause, values_fill = NA) |>
  select(-PUBMED_ID) |>
  mutate_all(~ ifelse(is.na(.), 0, 1)) |>
  as.data.frame()

upset_matrix = make_comb_mat(diabetes_studies_wide)

# create an upset plot
UpSet(upset_matrix,
      comb_order = order(comb_size(upset_matrix), decreasing = TRUE)
      #nsets = 3, 
      #order.by = "freq", 
      # sets.bar.color = "#56B4E9", 
      # main.bar.color = "#0072B2", 
      # matrix.color = "#D55E00", 
      #text.scale = c(1.5, 1.5, 1.5, 1.5, 1.5, 1.5)
      )
  
```

## Between different types of cancer

```{r check_overlap_cancer}

cancer_studies = disease_mapping_with_cause |>
  filter(grepl("cancer|neoplasm", cause, ignore.case = TRUE)) |>
  select(PUBMED_ID, cause) |>
  distinct()

cancer_studies_wide = cancer_studies |>
  tidyr::pivot_wider(names_from = cause, values_from = cause, values_fill = NA) |>
  select(-PUBMED_ID) |>
  mutate_all(~ ifelse(is.na(.), 0, 1)) |>
  as.data.frame()

upset_matrix_cancer = make_comb_mat(cancer_studies_wide)

UpSet(upset_matrix_cancer,
      comb_order = order(comb_size(upset_matrix_cancer), decreasing = TRUE)
      #nsets = 3, 
      #order.by = "freq", 
      # sets.bar.color = "#56B4E9", 
      # main.bar.color = "#0072B2", 
      # matrix.color = "#D55E00", 
      #text.scale = c(1.5, 1.5, 1.5, 1.5, 1.5, 1.5)
      )

```

## Between different cardiovascular diseases

```{r check_overlap_cardiovascular}

cardiovascular_studies = disease_mapping_with_cause |>
  filter(grepl("heart|stroke|hemorrhage", cause, ignore.case = TRUE)) |>
  select(PUBMED_ID, cause) |>
  distinct()

cardiovascular_studies_wide = cardiovascular_studies |>
  tidyr::pivot_wider(names_from = cause, values_from = cause, values_fill = NA) |>
  select(-PUBMED_ID) |>
  mutate_all(~ ifelse(is.na(.), 0, 1)) |>
  as.data.frame()

upset_matrix_cardiovascular = make_comb_mat(cardiovascular_studies_wide)

UpSet(upset_matrix_cardiovascular,
      comb_order = order(comb_size(upset_matrix_cardiovascular), decreasing = TRUE)
      #nsets = 3, 
      #order.by = "freq", 
      # sets.bar.color = "#56B4E9", 
      # main.bar.color = "#0072B2", 
      # matrix.color = "#D55E00", 
      #text.scale = c(1.5, 1.5, 1.5, 1.5, 1.5, 1.5)
      )

```

## Across liver diseases

Cirrhosis and other chronic liver diseases, Liver cancer due to hepatitis C, Liver cancer due to hepatitis B, Diabetes 

```{r check_overlap_liver}

liver_studies = disease_mapping_with_cause |>
  filter(grepl("liver|diabetes", cause, ignore.case = TRUE)) |>
  filter(!grepl("Type 2|Type 1", cause, ignore.case = TRUE)) |>
  select(PUBMED_ID, cause) |>
  distinct()


liver_studies_wide = liver_studies |>
  tidyr::pivot_wider(names_from = cause, values_from = cause, values_fill = NA) |>
  select(-PUBMED_ID) |>
  mutate_all(~ ifelse(is.na(.), 0, 1)) |>
  as.data.frame()

upset_matrix_liver = make_comb_mat(liver_studies_wide)

UpSet(upset_matrix_liver,
      comb_order = order(comb_size(upset_matrix_liver), decreasing = TRUE)
      #nsets = 3, 
      #order.by = "freq", 
      # sets.bar.color = "#56B4E9", 
      # main.bar.color = "#0072B2", 
      # matrix.color = "#D55E00", 
      #text.scale = c(1.5, 1.5, 1.5, 1.5, 1.5, 1.5)
      )

```


## Across all non-communicable diseases

```{r check_overlap_all_ncd}

non_communicable_studies =
disease_mapping_with_cause |>
    dplyr::filter(!cause %in% c("HIV/AIDS",
                             "Tuberculosis",
                             "Malaria",
                             "Lower respiratory infections",
                             "Diarrhoeal diseases",
                             "Neonatal disorders",
                             "Tetanus",
                             "Diphtheria",
                             "Pertussis" ,
                             "Measles",
                             "Maternal disorders")) |>
  filter(!is.na(cause)) |>
  select(PUBMED_ID, cause) |>
  distinct() 
  
non_communicable_studies_wide = non_communicable_studies |>
  tidyr::pivot_wider(names_from = cause, values_from = cause, values_fill = NA) |>
  select(-PUBMED_ID) |>
  mutate_all(~ ifelse(is.na(.), 0, 1)) |>
  as.data.frame()

upset_matrix_non_communicable = make_comb_mat(non_communicable_studies_wide)

UpSet(upset_matrix_non_communicable,
      comb_order = order(comb_size(upset_matrix_non_communicable), decreasing = TRUE)
      #nsets = 3, 
      #order.by = "freq", 
      # sets.bar.color = "#56B4E9", 
      # main.bar.color = "#0072B2", 
      # matrix.color = "#D55E00", 
      #text.scale = c(1.5, 1.5, 1.5, 1.5, 1.5, 1.5)
      )


```

## Between different respiratory diseases

Chronic obstructive pulmonary disease, "Tracheal, bronchus, and lung cancer

```{r check_overlap_respiratory}

respiratory_studies = disease_mapping_with_cause |>
  filter(grepl("chronic obstructive pulmonary disease|tracheal|bronchus|lung cancer", cause, ignore.case = TRUE)) |>
  select(PUBMED_ID, cause) |>
  distinct()

respiratory_studies_wide = respiratory_studies |>
  tidyr::pivot_wider(names_from = cause, values_from = cause, values_fill = NA) |>
  select(-PUBMED_ID) |>
  mutate_all(~ ifelse(is.na(.), 0, 1)) |>
  as.data.frame()


upset_matrix_respiratory = make_comb_mat(respiratory_studies_wide)

UpSet(upset_matrix_respiratory,
      comb_order = order(comb_size(upset_matrix_respiratory), decreasing = TRUE)
      #nsets = 3, 
      #order.by = "freq", 
      # sets.bar.color = "#56B4E9", 
      # main.bar.color = "#0072B2", 
      # matrix.color = "#D55E00", 
      #text.scale = c(1.5, 1.5, 1.5, 1.5, 1.5, 1.5)
      )

```

## Overlap across Lancet conditions

```{r check_overlap_lancet_conditions}

lancet_cause_mapping <- readxl::read_xlsx(here::here("data/icd/lancet_conditions_icd10.xlsx"),
                                     sheet = 1) |>
  dplyr::rename_with(~ gsub(" ", "_", .x)) |>
  filter(!is.na(mapped_gbd_term)) |>
  filter(lancet_group == "NCD-7") |>
  select(cause = mapped_gbd_term, lancet_condition) |>
  distinct()

disease_mapping_with_cause = 
  disease_mapping_with_cause |>
  left_join(lancet_cause_mapping,
            by = "cause"
            )

overlap_lancet_conditions =
disease_mapping_with_cause |>
  group_by(lancet_condition) |>
  filter(!is.na(lancet_condition)) |>
  summarise(PUBMED_ID = list(unique(PUBMED_ID))) |>
  tibble::deframe()


upset_matrix_lancet_conditions = make_comb_mat(overlap_lancet_conditions)

UpSet(upset_matrix_lancet_conditions,
      comb_order = order(comb_size(upset_matrix_lancet_conditions), decreasing = TRUE)
      #nsets = 3, 
      #order.by = "freq", 
      # sets.bar.color = "#56B4E9", 
      # main.bar.color = "#0072B2", 
      # matrix.color = "#D55E00", 
      #text.scale = c(1.5, 1.5, 1.5, 1.5, 1.5, 1.5)
      )

print("Number of studies mapped to one Lancet condition:")
comb_size(upset_matrix_lancet_conditions, degree = 1) |> sum()

print("All non-communicable disease studies mapped to a Lancet condition:")
overlap_lancet_conditions |> purrr::list_c() |> unique() |> length()

```

# Check recall i.e. that I have identified most/all studies of the priority diseases

## Diabetes 

```{r recall_diabetes}

no_capture <- paste0("(?<!no )",
                "(?<!non )",
                "(?<!no type 2 )"
                )

diabetes_grep = paste0(no_capture, "diabetes")

disease_mapping_with_cause |>
  filter(grepl(diabetes_grep, 
               `DISEASE/TRAIT`, 
               ignore.case = TRUE, 
               perl = TRUE)|
           grepl(diabetes_grep,
                collected_all_disease_terms,
                ignore.case = TRUE,
                perl = TRUE
                )
         ) |>
  group_by(PUBMED_ID) |>
  filter(all(is.na(cause))) |>
  select(PUBMED_ID, 
         `DISEASE/TRAIT`,
         collected_all_disease_terms, 
         icd10_code,
         cause) |>
  distinct() 


```

## Liver cancer, and other chronic liver diseases (e.g. cirrhosis) 

Check recall for these together as they are often comorbid and may be captured by similar terms

```{r recall_liver_cancer}

liver_cancer_grep = paste0(no_capture, "liver|cirhosis|hepatocellular carcinoma|hepatoma|hepatic carcinoma")

disease_mapping_with_cause |>
  filter(grepl(liver_cancer_grep, 
               `DISEASE/TRAIT`, 
               ignore.case = TRUE, 
               perl = TRUE)|
           grepl(liver_cancer_grep,
                collected_all_disease_terms,
                ignore.case = TRUE,
                perl = TRUE
                )
         ) |>
  group_by(PUBMED_ID) |>
  filter(all(is.na(cause))) |>
  group_by(PUBMED_ID, 
           `DISEASE/TRAIT`,
           collected_all_disease_terms,
           cause) |>
  summarise(icd10_code = str_flatten(unique(icd10_code), 
                                     collapse = ", ",
                                     na.rm = T),
            cause = str_flatten(unique(icd10_code), 
                                     collapse = ", ",
                                     na.rm = T)) |>
  ungroup() |>
  select(`DISEASE/TRAIT`,
         collected_all_disease_terms) |>
  distinct()

```

## Stroke

```{r recall_stroke}

stroke_grep = paste0(no_capture, 
                     "stroke|cerebrovascular|cerebral|hemorrhage")

disease_mapping_with_cause |>
  filter(grepl(stroke_grep, 
               `DISEASE/TRAIT`, 
               ignore.case = TRUE, 
               perl = TRUE)|
           grepl(stroke_grep,
                collected_all_disease_terms,
                ignore.case = TRUE,
                perl = TRUE
                )
         ) |>
  group_by(PUBMED_ID) |>
  filter(all(is.na(cause))) |>
  group_by(PUBMED_ID, 
           `DISEASE/TRAIT`,
           collected_all_disease_terms,
           cause) |>
  summarise(icd10_code = str_flatten(unique(icd10_code), 
                                     collapse = ", ",
                                     na.rm = T),
            cause = str_flatten(unique(icd10_code), 
                                collapse = ", ",
                                na.rm = T)) |>
  ungroup() |>
  select(`DISEASE/TRAIT`,
         collected_all_disease_terms) |>
  distinct()
  

```

## Stomach cancer

```{r recall_stomach_cancer}

stomach_cancer_grep = paste0(no_capture, 
                     "stomach|gastric")

disease_mapping_with_cause |>
  filter(grepl(stomach_cancer_grep, 
               `DISEASE/TRAIT`, 
               ignore.case = TRUE, 
               perl = TRUE)|
           grepl(stomach_cancer_grep,
                collected_all_disease_terms,
                ignore.case = TRUE,
                perl = TRUE
                )
         ) |>
  group_by(PUBMED_ID) |>
  filter(all(is.na(cause))) |>
  group_by(PUBMED_ID, 
           `DISEASE/TRAIT`,
           collected_all_disease_terms,
           cause) |>
  summarise(icd10_code = str_flatten(unique(icd10_code), 
                                     collapse = ", ",
                                     na.rm = T),
            cause = str_flatten(unique(icd10_code), 
                                collapse = ", ",
                                na.rm = T)) |>
  ungroup() |>
  select(`DISEASE/TRAIT`,
         collected_all_disease_terms) |>
  distinct()


```

## Cervical cancer

```{r recall_cervical_cancer}

cervical_cancer_grep = paste0(no_capture, 
                     "cervical|cervix")

disease_mapping_with_cause |>
  filter(grepl(cervical_cancer_grep, 
               `DISEASE/TRAIT`, 
               ignore.case = TRUE, 
               perl = TRUE)|
           grepl(cervical_cancer_grep,
                collected_all_disease_terms,
                ignore.case = TRUE,
                perl = TRUE
                )
         ) |>
  group_by(PUBMED_ID) |>
  filter(all(is.na(cause))) |>
  group_by(PUBMED_ID, 
           `DISEASE/TRAIT`,
           collected_all_disease_terms,
           cause) |>
  summarise(icd10_code = str_flatten(unique(icd10_code), 
                                     collapse = ", ",
                                     na.rm = T),
            cause = str_flatten(unique(icd10_code), 
                                collapse = ", ",
                                na.rm = T)) |>
  ungroup() |>
  select(`DISEASE/TRAIT`,
         collected_all_disease_terms) |>
  distinct()

# pubmed 35729236 studies 
# hormone-sensitive cancers: breast, ovarian, thyroid, prostate, and uterine cancer
# but GWAS Catalog mapped uterine cancer to cervical cancer, which is
# why it is picked up in this recall, but correctly not 
# mapped to cervical cancer in ICD-10 mapping!


# human papilloma virus (HPV) is a major cause of cervical cancer ...
# so do we include studies of HPV in our mapping of cervical cancer?
hpv_grep <- paste0(no_capture, 
                     "hpv|human papilloma virus|pap smear")

disease_mapping_with_cause |>
  filter(grepl(hpv_grep, 
               `DISEASE/TRAIT`, 
               ignore.case = TRUE, 
               perl = TRUE)|
           grepl(hpv_grep,
                collected_all_disease_terms,
                ignore.case = TRUE,
                perl = TRUE
                )
         ) |>
  group_by(PUBMED_ID) |>
  filter(all(is.na(cause))) |>
  group_by(PUBMED_ID, 
           `DISEASE/TRAIT`,
           collected_all_disease_terms,
           cause) |>
  summarise(icd10_code = str_flatten(unique(icd10_code), 
                                     collapse = ", ",
                                     na.rm = T),
            cause = str_flatten(unique(icd10_code), 
                                collapse = ", ",
                                na.rm = T)) |>
  ungroup() |>
  select(`DISEASE/TRAIT`,
         collected_all_disease_terms) |>
  distinct()

disease_mapping_with_cause <- 
  disease_mapping_with_cause |>
  mutate(icd10_code = str_replace(icd10_code, "R879", "R87.9")) 

```

## Lip and oral cavity cancer

```{r recall_lip_oral_cavity_cancer}

lip_oral_cavity_cancer_grep = paste0(no_capture, 
                     "lip\\b|oral cavity|mouth|tongue|pharynx|larynx|pharyngeal|laryngeal")

disease_mapping_with_cause |>
  filter(grepl(lip_oral_cavity_cancer_grep, 
               `DISEASE/TRAIT`, 
               ignore.case = TRUE, 
               perl = TRUE)|
           grepl(lip_oral_cavity_cancer_grep,
                collected_all_disease_terms,
                ignore.case = TRUE,
                perl = TRUE
                )
         ) |>
  filter(!grepl("cleft", `DISEASE/TRAIT`, ignore.case = TRUE)) |>
  group_by(PUBMED_ID) |>
  filter(all(is.na(cause))) |>
  group_by(PUBMED_ID, 
           `DISEASE/TRAIT`,
           collected_all_disease_terms,
           cause) |>
  summarise(icd10_code = str_flatten(unique(icd10_code), 
                                     collapse = ", ",
                                     na.rm = T),
            cause = str_flatten(unique(icd10_code), 
                                collapse = ", ",
                                na.rm = T))


# !!!!!!!!!!!! Maybe caught one! 

```

## Heart disease

```{r}

heart_disease_grep = paste0(no_capture, 
                     "heart disease|cardiovascular|coronary|myocardial|angina|heart failure")

disease_mapping_with_cause |>
  filter(grepl(heart_disease_grep, 
               `DISEASE/TRAIT`, 
               ignore.case = TRUE, 
               perl = TRUE)|
           grepl(heart_disease_grep,
                collected_all_disease_terms,
                ignore.case = TRUE,
                perl = TRUE
                )
         ) |>
  group_by(PUBMED_ID) |>
  filter(all(is.na(cause))) |>
  group_by(PUBMED_ID, 
           `DISEASE/TRAIT`,
           collected_all_disease_terms,
           cause) |>
  summarise(icd10_code = str_flatten(unique(icd10_code), 
                                     collapse = ", ",
                                     na.rm = T),
            cause = str_flatten(unique(icd10_code), 
                                collapse = ", ",
                                na.rm = T))

```

# Check precision i.e. that the studies I have mapped to the priority diseases are actually about those diseases

## Type 2 diabetes 

```{r precision_type_2_diabetes}

disease_mapping_with_cause |>
  filter(cause == "Type 2 diabetes mellitus") |>
  filter(!grepl("type 2 diabetes|diabetes type 2|e11|type ii diabetes", `DISEASE/TRAIT`, ignore.case = TRUE) 
         ) |>
 group_by(PUBMED_ID,
          STUDY_ACCESSION,
          `DISEASE/TRAIT`,
          collected_all_disease_terms) |>
  summarise(icd10_code = str_flatten(unique(icd10_code), 
                                     collapse = ", ",
                                     na.rm = T),
            cause = str_flatten(unique(cause), 
                                collapse = ", ",
                                na.rm = T)) |>
  select(PUBMED_ID, 
         `DISEASE/TRAIT`,
         collected_all_disease_terms, 
         icd10_code,
         cause) |>
  distinct()


# problems with GCST90079743, GCST90083729 study provided mapping

# pubmed id 29703844 - should map to type 2 diabetes
#  genome-wide association study (GWAS) of DKD in type 2 diabetes (T2D)


# ? idk: 27670767

```


### Checking hierarchy levels

```{r check_hierarchy_levels}

# disease_mapping_with_cause = 
#   disease_mapping_with_cause |>
#   filter(!is.na(cause))

# filtering na to deal with pivoting 
# disease_mapping_with_cause_grouped = 
# disease_mapping_with_cause |>
# select(-icd10_code_num) |>
# tidyr::pivot_wider(
#             id_cols = c("collected_all_disease_terms", 
#                         "icd10_code", 
#                         "icd10_description"),
#             names_from = cause_hierarchy_level, 
#             names_glue = "l{.name}_{.value}",
#             values_from = cause,
#             values_fn = ~str_flatten(unique(.x), collapse = ", ", na.rm = TRUE)
#             )


```

```{r fill_in_l2_causes}

# check which l2 causes are NA but l3 causes are not NA
# can use l3 causes therefore to fill in l2 causes
# disease_mapping_with_cause_grouped |> 
#   rowwise() |> 
#   filter(is.na(l2_cause) & !is.na(l3_cause)) |>
#   head()
# 
# neoplasms = c("Other neoplasms", 
#                                     "Hodgkin lymphoma",
#                                     "Leukemia",
#                                     "Non-Hodgkin lymphoma",
#               "Multiple myeloma")
# 
# disease_mapping_with_cause_grouped = 
#   disease_mapping_with_cause_grouped |>
#   rowwise() |>
#   mutate(l2_cause = 
#            list(ifelse(is.na(l2_cause),
#                   case_when(map_lgl(l3_cause, ~ .x %in% neoplasms) ~ "Neoplasms",
#                             map_lgl(l3_cause, ~ .x %in% "Neonatal disorders") ~ "Maternal and neonatal disorders",
#                             map_lgl(l3_cause, ~ .x %in% "Other cardiovascular and circulatory diseases") ~ "Cardiovascular diseases"
#                             ),
#                   l2_cause
#            )
#            )) |>
#   ungroup()
          

# to_map_manually <- 
# disease_mapping_with_cause_grouped |> 
#   rowwise() |> 
#   filter(is.null(l2_cause)) |> 
#   pull(collected_all_disease_terms) |> 
#   unique()
# 
# 
# update_unmapped_cause = function(df, 
#                                  unmapped_term, 
#                                  l2_cause_rep, 
#                                  l3_cause_rep){
#   
#   df = 
#   df |>
#     mutate(l3_cause = 
#            ifelse(collected_all_disease_terms == unmapped_term,
#                   l3_cause_rep,
#                   l3_cause
#            )) |>
#     mutate(l2_cause = 
#            ifelse(collected_all_disease_terms == unmapped_term,
#                   l2_cause_rep,
#                   l2_cause
#            ))    
#            
#     return(df)
#            
# }
# 
# disease_mapping_with_cause_grouped = 
# update_unmapped_cause(disease_mapping_with_cause_grouped, 
#                       "bone cancer",
#                       "Neoplasms",
#                       "Malignant neoplasms of bone and articular cartilage")


```

# Save + combine with GWAS data

```{r save_final_gwas_gbd_mapping}

# disease_mapping_orig <- data.table::fread(
# here::here("output/icd_map/gwas_disease_to_icd10_mapping.csv")
# )
# 
# disease_mapping_with_cause_grouped =
#   disease_mapping_with_cause_grouped |>
#   tidyr::unnest_longer(c("l3_cause", "l4_cause")) 

disease_mapping_final =  disease_mapping_with_cause
  # left_join(
  #   disease_mapping_orig,
  #   disease_mapping_with_cause,
  #   by = c("collected_all_disease_terms", 
  #          "icd10_code", 
  #          "icd10_description")
  # )

# gwas_study_info <- fread(here::here("output/gwas_cat/gwas_study_info_group_v2.csv"))
# 
# gwas_study_info =
#   gwas_study_info |>
#   select(`DISEASE/TRAIT`, PUBMED_ID, STUDY_ACCESSION, COHORT)
# 
# disease_mapping_final =
#   left_join(
#     disease_mapping_final,
#     gwas_study_info
#   )


# disease_mapping_final = 
#   disease_mapping_final |>
#   group_by(STUDY_ACCESSION, collected_all_disease_terms) |>
#   summarise(
#     PUBMED_ID = paste0(unique(PUBMED_ID), collapse = "; "),
#     COHORT = paste0(unique(COHORT), collapse = "; "),
#     icd10_code = paste0(unique(icd10_code), collapse = "; "),
#     icd10_description = paste0(unique(icd10_description), collapse = "; "),
#     # l2_cause = str_flatten(unique(str_split(l2_cause, ", ")), # add separation step here ... 
#     #                        collapse = "; ", na.rm = TRUE),
#     l3_cause = str_flatten(unique(str_split(l3_cause, ", ")), 
#                            collapse = "; ", na.rm = TRUE),
#     l4_cause = str_flatten(unique(str_split(l4_cause, ", ")), 
#                            collapse = "; ", na.rm = TRUE)
#   ) 

data.table::fwrite(
# disease_mapping_final,
disease_mapping_with_cause,
here::here("output/icd_map/gwas_study_gbd_causes.csv")
)

```


