---
title: "GWAS Meta-science ideas exploration"
site: workflowr::wflow_site
output:
  workflowr::wflow_html:
    toc: true
    toc_float: true
    number_sections: true
editor_options: 
  markdown: 
    wrap: 72
---

<b> [github repository
link](https://github.com/IJbeasley/genomics_ancest_disease_dispar) </b>

<br>

<br>

# Find GWAS studies of Lancet Commission Priority Diseases

```{r include = FALSE}

# function to find and print out number of traits
n_unique_traits <- function(trait_col){
  
  trait_col = stringr::str_split(pattern = ", ", 
                                 trait_col[trait_col != ""]
                                 )
  
  trait_col = unlist(trait_col) |>
              stringr::str_trim()
  
  n_unique = length(unique(trait_col))
  
  print(paste0("Number of unique disease terms: ", n_unique))
  
}

```

## Identify GWAS studies of diseases in general

I find GWAS Catalog studies that of diseases, by looking for
disease-related EFO ontology mapped terms in the 'MAPPED_TRAIT' column
of the GWAS Catalog metadata. I categorize traits into disease,
response, measurement etc. using the EFO ontology.

-   [Download trait ontology information](download_ontology.html)

-   [Categorising GWAS traits into disease, response, measurement
    etc.](trait_ontology_categorization.html)

```{r trait_ontology, echo=FALSE}

gwas_study_info <- data.table::fread(here::here("output/gwas_cat/gwas_study_info_trait_cat.csv"))

gwas_study_info <-
  gwas_study_info |>
  dplyr::filter(DISEASE_STUDY == T)

print("Number of pubmed ids studying at least one disease")
gwas_study_info |>
  dplyr::pull(PUBMED_ID) |>
  unique() |>
  length()

n_unique_traits(gwas_study_info$all_disease_terms)

```

## Mapping traits to ICD 10 Codes

Next, for GWAS studies of disease I harmonise disease trait labels to
reduce redundancy due to typos, synonyms etc.

-   [Harmonising disease trait labels - typos, synonyms
    etc.](disease_trait_terms_simplify.html)

```{r, echo=FALSE}

gwas_study_info <- data.table::fread(here::here("output/gwas_cat/gwas_study_info_disease_trait_simplified.csv"))

n_unique_traits(gwas_study_info$collected_all_disease_terms)

```

-   [Grouping cancer disease traits (removing disease subtypes
    etc)](group_cancer_diseases.html)

```{r, echo=FALSE}

gwas_study_info <- data.table::fread(here::here("output/gwas_cat/gwas_study_cancer_group.csv"))

n_unique_traits(gwas_study_info$collected_all_disease_terms)

```

-   [Grouping non-cancer disease traits (removing disease subtypes
    etc)](group_non_cancer_diseases.html)

```{r, echo=FALSE}

gwas_study_info <- data.table::fread(here::here("output/gwas_cat/gwas_study_all_group.csv"))

n_unique_traits(gwas_study_info$collected_all_disease_terms)

```

### Now: Mapping traits to ICD 10 Codes

Then, I map the harmonised disease trait labels to ICD-10 codes, by the
following step-wise procedure:

1.  Where available, extract author provided ICD-10 Codes in GWAS
    Catalog DISEASE/TRAIT metadata

2.  Where available, extract author provided PheCodes in GWAS Catalog
    DISEASE/TRAIT metadata. These PheCodes are converted to ICD-10 codes
    using the PheWAS R package international mapping file, and checked
    against PheCode to ICD-10cm mappings (from
    <https://phenomics.va.ornl.gov/phecodemap/>).

3.  If author provided ICD-10 or PheCodes are not available, match
    `DISEASE/TRAIT` labels to ICD-10 code descriptions

4.  If not able to match `DISEASE/TRAIT` labels to ICD-10 code
    descriptions, try match `DISEASE/TRAIT` labels to PheCode
    descriptions

5.  If not able to match `DISEASE/TRAIT` labels to PheCode descriptions,
    try match `DISEASE/TRAIT` labels to UMLS terms

6.  Use manually created excel mapping file (ICD-10 to `DISEASE/TRAIT`
    terms, from WHO ICD-10 2019 index:
    <https://icd.who.int/browse10/2019>).

7.  Repeat Steps 3-5 for but instead of trying to match terms to
    `DISEASE/TRAIT` try to match terms to `collected_all_disease_terms`
    labels (these are processed EFO and other ontology terms)

8.  Use author provide ICD-10 codes from studies of the same disease
    (same `collected_all_disease_terms` )

The code creating the final map of disease traits to ICD-10 codes is
below:

-   [Mapping all traits to ICD-10 codes](map_trait_to_icd10.html)

-   [Manually mapped traits ](manual_trait_map_icd10.html)

<br>

## Find GWAS publications of Lancet Commission Priority Diseases

Use the above mapping of disease traits to ICD-10 codes to find GWAS Catalog studies of the Lancet Commission Priority Diseases. This is done by matching the ICD-10 codes of the diseases to the ICD-10 codes of the GWAS Catalog disease traits.

-   [Matching GWAS traits to GBD causes / diseases](gwas_to_gbd.html)

<br> 

# Extracting cohort labels

Allowing us to deal with overlapping samples.

-   [Fixing & harmonising GWAS Catalog cohort metadata](correcting_cohort_names.html)

    Mostly just correcting for typos and different ways of writing the
    same cohort name (e.g. Finland vs FINLAND). 

-   [Missing cohort metadata in the GWAS Catalog](missing_cohort_info.html)

     Investigation into missingness of GWAS Catalog cohort metadata, is it more common in older studies, or more recent studies? 

-   [Distribution of cohort information exploration](cohort_dist.html)
     
     Explore the structure of GWAS Catalog metadata and how cohort labels are distributed across studies.

## Fine tune text mining model 

### Get text to fine tune model

-   [Getting abstracts from GWAS Catalog](get_abstracts.html)

-   [Getting full-text of GWAS Catalog articles](get_full_text.html)

-   [Getting supplement of GWAS Catalog articles](get_supplement.html) 

Pre-process article text for text mining:

-   Convert text to sentences: code/extract_text/spacy_obtain_sentences.pu

-   [Get initial annotated training sentences](test_for_cohort_labels.html)

### Other investigations

-   [Extracting dbGAP accessions from GWAS Catalog studies](get_dbgap_ids.html)

<br>

## Calculating ancestry biases

-   [Replicate observed ancestry biases figures in GWAS
    dataset](replication_ancestry_bias.html)

    Replicating Martin et al. 2019 Figure

-   [Disease investigated by ancestry](disease_inves_by_ancest.html)

<br> <br> <br>

# Exploratory analyses

Not included in analysis/study. 

## Disease burden statistics

-   [Global Burden of Disease Study data - DALYs, deaths
    etc.](gbd_data_plots.html)

    Global Burden of Disease Study (GBD) global statistics on incidence,
    prevalence, DALYs (Disability-Adjusted Life Years) and PAF
    (population Attributable Fraction) for diseases and risk factors.

<br> <br>

### Archived / not currently being used:

-   [Looking into: can identify UK Biobank studies from citations?](collapse_cohorts.html)

#### Paper citation metrics

-   [iCite - citation metrics for GWAS Catalog Papers](icite_rcr.html)

    Replicating Reales & Wallace, 2023 - sharing gwas data results in
    more citations

#### Old grouping / filtering disease traits pages:

-   [Excluding infectious diseases](grouping_infectious_diseases.html)

-   [Filtering of non-disease, symptom phenotypes
    etc.](other_disease_filtering.html)

-   [Grouping cancer disease traits (removing disease subtypes
    etc)](level_1_disease_group_cancer.html)

-   [Grouping non-cancer disease traits (removing disease subtypes
    etc)](level_1_disease_group_non_cancer.html)

-   [Match GWAS studies to Global Burden of Disease Study
    Groupings](level_2_disease_group.html)
