---
title: "Level 1 Disease Trait Grouping - Non-cancer"
author: "Isobel Beasley"
date: "2025-09-08"
output:
  workflowr::wflow_html:
    toc: true
    toc_float: true
    number_sections: true
editor_options: 
  markdown: 
    wrap: 72
---

# Set up

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r required_fns, message = F, warning= F}

library(dplyr)
library(data.table)
library(stringr)

```

## Ontology help - for getting disease subtypes

```{r ontology_fns}

source(here::here("code/get_term_descendants.R"))

```

```{r setup_2, message = FALSE, warning= FALSE}

gwas_study_info <- fread(here::here("output/gwas_cat/gwas_study_info_disease_trait_simplified.csv"))

```

# Initial summary

## Objectives of this analysis:

Level 1 disease grouping: 
-   collapse disease causes, e.g. contact dermatitis due to nickel to contact dermatitis
-   collapse disease subtypes, e.g. bipolar I and bipolar II to bipolar disorder
-   collapse disease onset times, e.g. early-onset alzheimers disease and late-onset alzheimers disease to alzheimers disease
-   collapse disease stages
-   collapse disease complications, e.g. device complication, trauma complication to complication

## Grouping - level 1 set up 

```{r group_l1}

gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms = collected_all_disease_terms)

```

## Initial summary - number of unique study terms

```{r}

n_studies_trait = gwas_study_info |>
  dplyr::filter(DISEASE_STUDY == T) |>
  dplyr::select(collected_all_disease_terms, PUBMED_ID) |>
  dplyr::distinct() |>
  dplyr::group_by(collected_all_disease_terms) |>
  dplyr::summarise(n_studies = dplyr::n()) |>
  dplyr::arrange(desc(n_studies))

head(n_studies_trait)

dim(n_studies_trait)

```

### When separate studies with multiple terms 

```{r}

diseases <- stringr::str_split(pattern = ", ", 
 gwas_study_info$collected_all_disease_terms[gwas_study_info$collected_all_disease_terms != ""])  |> 
            unlist() |>
            stringr::str_trim()

length(unique(diseases))

```


# Disturbances to senses 

```{r disturbances_to_senses}

disturb_senses_terms <- c("disturbances of sensation of smell and taste",
                          "abnormality of the sense of smell",
                          "ageusia")

```


# Disease stage grouping

## Chronic kidney disease (CKD)

```{r}

gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms = 
         stringr::str_replace_all(l1_all_disease_terms,
                          "chronic kidney disease stage 5|chronic kidney disease stage 4",
                          "chronic kidney disease"
         ))


gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms = 
         stringr::str_replace_all(l1_all_disease_terms,
                          "stage 5 chronic kidney disease",
                          "chronic kidney disease"
         ))

```

# Disease cause grouping

## Contact dermatitis

```{r}


gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms = 
         stringr::str_replace_all(l1_all_disease_terms,
                          "contact dermatitis due to nickel",
                          "contact dermatitis"
         ))



```

# Disease complication

```{r disease_complication}

disease_complication_terms = c("device complication",
                             "trauma complication",
                             "adverse effect",
                             "aseptic loosening"
                             )

gwas_study_info = gwas_study_info |>
 mutate(l1_all_disease_terms  = 
          stringr::str_replace_all(l1_all_disease_terms,
                                  pattern = paste0(disease_complication_terms, collapse = "(?=,|$)|\\b"),
                                   "complication"
                          )  
        )

```


# Cardiovascular diseases

## Hypertension

### Pulmonary arterial hypertension

```{r pulmonary_arterial_hypertension}

url <- "http://www.ebi.ac.uk/ols4/api/ontologies/efo/terms/http%253A%252F%252Fwww.ebi.ac.uk%252Fefo%252FEFO_0001361/descendants"

pulmonary_arterial_hypertension_terms <- get_descendants(url)

gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms  = 
         stringr::str_replace_all(l1_all_disease_terms,
                          pattern = paste0(pulmonary_arterial_hypertension_terms, collapse = "(?=,|$)|\\b"),
                          "pulmonary arterial hypertension"
         ))

```

### Atrial fibrillation & flutter

### Atrial fibrillation

```{r atrial_fibrillation}

gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms  = 
         stringr::str_replace_all(l1_all_disease_terms,
                          "post-operative atrial fibrillation",
                          "atrial fibrillation, post-operative"
         ))



url <- "http://www.ebi.ac.uk/ols4/api/ontologies/snomed/terms/http%253A%252F%252Fsnomed.info%252Fid%252F49436004/descendants"

atrial_fibrillation_terms <- get_descendants(url)

gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms  = 
         stringr::str_replace_all(l1_all_disease_terms,
                          pattern = paste0(atrial_fibrillation_terms, collapse = "(?=,|$)|\\b"),
                          "atrial fibrillation"
         ))

```

#### Atrial flutter

```{r atrial_flutter}

url <- "http://www.ebi.ac.uk/ols4/api/ontologies/snomed/terms/http%253A%252F%252Fsnomed.info%252Fid%252F5370000/descendants"

atrial_flutter_terms <- get_descendants(url)

gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms  = 
         stringr::str_replace_all(l1_all_disease_terms,
                          pattern = paste0(atrial_flutter_terms, collapse = "(?=,|$)|\\b"),
                          "atrial flutter"
         ))

```

## Heart disease

### Hypertensive heart disease

```{r hypertensive_heart_disease}

url <- "http://www.ebi.ac.uk/ols4/api/ontologies/snomed/terms/http%253A%252F%252Fsnomed.info%252Fid%252F64715009/descendants"

hypertensive_heart_disease_terms <- get_descendants(url)

gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms  = 
         stringr::str_replace_all(l1_all_disease_terms,
                          pattern = paste0(hypertensive_heart_disease_terms, collapse = "(?=,|$)|\\b"),
                          "hypertensive heart disease"
         ))

```

### Ischemic heart disease

```{r ischemic_heart_disease}

url <- "http://www.ebi.ac.uk/ols4/api/ontologies/cvdo/terms/http%253A%252F%252Fpurl.obolibrary.org%252Fobo%252FDOID_3394/descendants"

ischemic_heart_disease_terms <- get_descendants(url)

url <- "http://www.ebi.ac.uk/ols4/api/ontologies/snomed/terms/http%253A%252F%252Fsnomed.info%252Fid%252F414545008/descendants"

ischemic_heart_disease_terms_snomed <- get_descendants(url)

ischemic_heart_disease_terms = c(ischemic_heart_disease_terms,
                                 ischemic_heart_disease_terms_snomed
                                 ) |>
                                str_length_sort()

gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms  = 
         stringr::str_replace_all(l1_all_disease_terms,
                          pattern = paste0(ischemic_heart_disease_terms, collapse = "(?=,|$)|\\b"),
                          "ischemic heart disease"
         ))

```

#### Coronary artery disease

Likely also add, as ICD10 I25.1, seems to be coronary artery disease
And gbd includes this in ischemic heart disease

```{r coronary_artery_disease}

url <- "http://www.ebi.ac.uk/ols4/api/ontologies/efo/terms/http%253A%252F%252Fwww.ebi.ac.uk%252Fefo%252FEFO_0001645/descendants"

coronary_artery_disease_terms <- get_descendants(url)

gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms  = 
         stringr::str_replace_all(l1_all_disease_terms,
                          pattern = paste0(coronary_artery_disease_terms, collapse = "(?=,|$)|\\b"),
                          "coronary artery disease"
         ))

gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms  = 
         stringr::str_replace_all(l1_all_disease_terms,
                          "coronary artery disease",
                          "ischemic heart disease"
         ))

```

#### Myocardial infarction

```{r myocardial_infarction}

url <- "http://www.ebi.ac.uk/ols4/api/ontologies/efo/terms/http%253A%252F%252Fwww.ebi.ac.uk%252Fefo%252FEFO_0000612/descendants"

myocardial_infarction_terms <- get_descendants(url)

myocardial_infarction_terms <- c("subsequent st elevation \\(stemi\\) and non-st elevation \\(nstemi\\) myocardial infarction",
                              myocardial_infarction_terms
                              )

gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms  = 
         stringr::str_replace_all(l1_all_disease_terms,
                          pattern = paste0(myocardial_infarction_terms, collapse = "(?=,|$)|\\b"),
                          "myocardial infarction"
         ))


gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms  = 
         stringr::str_replace_all(l1_all_disease_terms,
                          "post-operative myocardial infarction",
                          "myocardial infarction, post-operative"
         ))

gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms  = 
         stringr::str_replace_all(l1_all_disease_terms,
                          "myocardial infarction",
                          "ischemic heart disease"
         ))

```
#### Fixing

```{r fix_ischemic_heart_disease}


to_fix <- c("non-st elevation ischemic heart disease",
            "post-operative ischemic heart disease",
            "st elevation ischemic heart disease",
            "subsequent st elevation (stemi) and non-st elevation (nstemi) ischemic heart disease")

gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms  = 
         stringr::str_replace_all(l1_all_disease_terms,
                          paste0(to_fix, collapse = "(?=,|$)|\\b"),
                          "ischemic heart disease"
         ))


```


### (Non-rheumatic) Heart valvular disease

```{r heart_valvular_disease}

url <- "http://www.ebi.ac.uk/ols4/api/ontologies/efo/terms/http%253A%252F%252Fwww.ebi.ac.uk%252Fefo%252FEFO_0009551/descendants"

heart_valve_disease_terms <- get_descendants(url)

# make non-rheumatic 
rheumatic_terms <- c("rheumatic pulmonary valve disease",
                     "rheumatic disease of mitral valve")

heart_valve_disease_terms <- 
  heart_valve_disease_terms[!heart_valve_disease_terms %in% rheumatic_terms]

gwas_study_info = gwas_study_info |> 
          mutate(l1_all_disease_terms  = 
          stringr::str_replace_all(l1_all_disease_terms ,
                                  pattern = paste0(heart_valve_disease_terms, 
                                                   collapse = "(?=,|$)|\\b"),
                                   "heart valve disease"
                                  )  
          )

```

## Other cardiovascular diseases (GBD)

### Aortic aneurysm

```{r aortic_aneurysm}

aortic_aneurysm_terms <- c("thoracic aortic aneurysm",
                          "abdominal aortic aneurysm"
                          )

gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms  = 
         stringr::str_replace_all(l1_all_disease_terms,
                          pattern = paste0(aortic_aneurysm_terms, collapse = "(?=,|$)|\\b"),
                          "aortic aneurysm"
         ))

```


### Endocarditis

```{r endocarditis}

gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms  = 
         stringr::str_replace_all(l1_all_disease_terms,
                          "bacterial endocarditis",
                          "endocarditis"
         ))

```

### Cardiomyopathy

```{r cardiomyopathy}

url <- "http://www.ebi.ac.uk/ols4/api/ontologies/hp/terms/http%253A%252F%252Fpurl.obolibrary.org%252Fobo%252FHP_0001638/descendants"

cardiomyopathy_terms <- get_descendants(url)

cardiomyopathy_terms <- c("chagas cardiomyopathy",
                          "ischemic cardiomyopathy",
                          "idiopathic cardiomyopathy",
                          "nonischemic cardiomyopathy",
                          "peripartum cardiomyopathy", #? maybe include in pregnancy 
                          cardiomyopathy_terms
                          )

gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms  = 
         stringr::str_replace_all(l1_all_disease_terms,
                          pattern = paste0(cardiomyopathy_terms, collapse = "(?=,|$)|\\b"),
                          "cardiomyopathy"
         ))

gwas_study_info =
gwas_study_info |> 
 mutate(l1_all_disease_terms  = 
          stringr::str_replace_all(l1_all_disease_terms ,
                                  pattern = "\\bidiopathic cardiomyopathy\\b",
                                   "cardiomyopathy"
                          )  
        )


```


### Stroke

https://www.ebi.ac.uk/ols4/ontologies/efo/classes/http%253A%252F%252Fwww.ebi.ac.uk%252Fefo%252FEFO_0000712?lang=en

```{r}


hemorrhage_terms <- c("intracerebral hemorrhage",
                      "non-lobar intracerebral hemorrhage",
                      "lobar intracerebral hemorrhage")

gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms  = 
         stringr::str_replace_all(l1_all_disease_terms,
                          pattern = paste0(hemorrhage_terms, collapse = "(?=,|$)|\\b"),
                          "hemorrhagic stroke")
  )

stroke_terms <- c("stroke outcome",
                  "large artery stroke",
                  "small vessel stroke",
                  "ischemic stroke",
                  "stroke disorder",
                  "cardioembolic stroke",
                  "hemorrhagic stroke",
                  "intracranial hemorrhage",
                  "subdural hemorrhage"
                  )

gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms  = 
         stringr::str_replace_all(l1_all_disease_terms,
                          paste0(stroke_terms, collapse = "(?=,|$)|\\b"),
                          "stroke")
  )


gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms  = 
         stringr::str_replace_all(l1_all_disease_terms,
                          "post-operative stroke",
                          "post-operative, stroke")
  )

```

### Lower extremity peripheral artery disease

```{r peripheral_artery_disease}

# studies in gwas catalog don't specify lower or upper extremity
# but most peripheral artery disease is in lower extremities

gwas_study_info =
 gwas_study_info |>
  mutate(l1_all_disease_terms  = 
         stringr::str_replace_all(l1_all_disease_terms,
                          "\\bperipheral artery disease|\\bperipheral arterial disease",
                          "lower extremity peripheral artery disease"
         ))

```


## Other (non-GBD)

### Carotid artery disease

```{r carotid_artery_disease}

carotid_artery_disease_terms <- c("carotid artery thrombosis",
                             "carotid atherosclerosis"
                             )

gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms  = 
         stringr::str_replace_all(l1_all_disease_terms,
                          pattern = paste0(carotid_artery_disease_terms, collapse = "(?=,|$)|\\b"),
                          "carotid artery disease"
         ))

```


### Other hypertension

```{r}

url <-  "http://www.ebi.ac.uk/ols4/api/ontologies/efo/terms/http%253A%252F%252Fwww.ebi.ac.uk%252Fefo%252FEFO_0000537/descendants"

hypertension_terms <- get_descendants(url)

hypertension_terms = c("primary hypertension",
                       "chronic venous hypertension",
                       "portal hypertension",
                      hypertension_terms
                      )

hypertension_terms = hypertension_terms[!hypertension_terms %in% 
                                        c(pulmonary_arterial_hypertension_terms,
                                          hypertensive_heart_disease_terms,
                                          "pulmonary arterial hypertension",
                                          "hypertensive heart disease"
                                          )
                                        ]

gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms  = 
         stringr::str_replace_all(l1_all_disease_terms,
                          pattern = paste0(hypertension_terms, collapse = "(?=,|$)|\\b"),
                          "hypertension"
         ))

gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms  = 
         stringr::str_replace_all(l1_all_disease_terms,
                          pattern = paste0(hypertension_terms, collapse = "(?=,|$)|\\b"),
                          "hypertension"
         ))

```

### Heart block 

```{r heart_block}

heart_block_terms <- c("first degree atrioventricular block",
                      "second degree atrioventricular block",
                      "third-degree atrioventricular block",
                      "atrioventricular block",
                      "bundle branch block",
                      "complete right heart block",
                      "left heart block"
                      )

gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms  = 
         stringr::str_replace_all(l1_all_disease_terms,
                          pattern = paste0(heart_block_terms, collapse = "(?=,|$)|\\b"),
                          "heart block"
         ))

```

### Tachycardia

```{r tacycardia}

tacycardia_terms <- c("paroxysmal tachycardia",
                     "ventricular tachycardia",
                     "supraventricular tachycardia",
                     "paroxysmal supraventricular tachycardia",
                     "paroxysmal ventricular tachycardia"
                     )


gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms  = 
         stringr::str_replace_all(l1_all_disease_terms,
                          pattern = paste0(tacycardia_terms, collapse = "(?=,|$)|\\b"),
                          "tachycardia"
         )
  )

```

### Cardiac arrest

```{r cardiac_arrest}

gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms  = 
         stringr::str_replace_all(l1_all_disease_terms,
                          "sudden cardiac arrest",
                          "cardiac arrest"
         )
  )

```

### Other cardiac arrhythmias

```{r other_cardiac_arrhythmias}

other_cardiac_arrhythmia_terms <- c("ventricular fibrillation",
                                   "ventricular arrhythmia",
                                   "cardiac arrest",
                                   "bradycardia",
                                   "cardiac arrhythmia"
                                   )


gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms  = 
         stringr::str_replace_all(l1_all_disease_terms,
                          pattern = paste0(other_cardiac_arrhythmia_terms, collapse = "(?=,|$)|\\b"),
                          "other cardiac arrhythmias"
         )
  )


```

### Congenital heart disease/s

```{r congenital_heart_disease}

congenital_heart_disease_terms <- c(
  "congenital anomaly of cardiovascular system",
  "congenital anomaly of the great arteries",
  "congenital heart malformation",
  "congenital left-sided heart lesions",
  "congenital right-sided heart lesions",
  "heart septal defect",
  "atrial heart septal defect",
  "atrioventricular canal defect",
  "conotruncal heart malformations",
  "abnormal aortic morphology",
  "abnormal cardiovascular system physiology",
  "abnormal cardiovascular system morphology",
  "abnormality of the vasculature",
  "cerebral arteriovenous malformation"
  )


gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms  = 
         stringr::str_replace_all(l1_all_disease_terms,
                          pattern = paste0(congenital_heart_disease_terms, collapse = "(?=,|$)|\\b"),
                          "congenital heart disease"
         )
  )


```

### Thrombosis

```{r thrombosis}

url <- "http://www.ebi.ac.uk/ols4/api/ontologies/mondo/terms/http%253A%252F%252Fpurl.obolibrary.org%252Fobo%252FMONDO_0000831/descendants"

thrombosis_terms <- get_descendants(url)

thrombosis_terms <- c("abnormal thrombosis",
                      thrombosis_terms)

gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms  = 
         stringr::str_replace_all(l1_all_disease_terms,
                          pattern = paste0(thrombosis_terms, collapse = "(?=,|$)|\\b"),
                          "thrombotic disease"
         )
  )

```

### Arterial occlusion

```{r arterial_occlusion}

arterial_occlusion_terms <- c("cerebral artery occlusion",
                              "occlusion precerebral artery",
                              "small artery occlusion"
                              )

gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms  = 
         stringr::str_replace_all(l1_all_disease_terms,
                          pattern = paste0(arterial_occlusion_terms, collapse = "(?=,|$)|\\b"),
                          "arterial occlusion"
         )
  )


```

### Other vascular disorders / diseases

```{r other_vascular_disorders}

other_vascular_disorder_terms <- c("skin vascular disease",
                                   "peripheral vascular disease",
                                   "vascular disease",
                                   "vasculitis",
                                   "arteritis")


gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms  = 
         stringr::str_replace_all(l1_all_disease_terms,
                          pattern = paste0(other_vascular_disorder_terms, collapse = "(?=,|$)|\\b"),
                          "other vascular disorders"
         )
  )


```

## Congestive heart failure

```{r congestive_heart_failure}

congestive_heart_failure_terms <- c("systolic heart failure",
                                 "diastolic heart failure",
                                 "cor pulmonale",
                                 "congenital heart disease",
                                 "chronic pulmonary heart disease"
                                 )


gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms  = 
         stringr::str_replace_all(l1_all_disease_terms,
                          pattern = paste0(congestive_heart_failure_terms, collapse = "(?=,|$)|\\b"),
                          "congestive heart failure"
         ))

```

## Heart failure

```{r heart_failure}




```

## Brain infarction

```{r brain_infarction}

brain_infarction_terms <- c("cerebral infarction",
                          "mri defined brain infarct",
                          "small cerebral infarction"
                          )

gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms  = 
         stringr::str_replace_all(l1_all_disease_terms,
                          pattern = paste0(brain_infarction_terms, collapse = "(?=,|$)|\\b"),
                          "brain infarction"
         ))

```

## Vascular insufficiency

```{r vascular_insufficiency}

vascular_insufficiency_terms <- c("vascular insufficiency disorder",
                                 "chronic venous insufficiency",
                                  "chronic intestinal vascular insufficiency"
                                 )

gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms  = 
         stringr::str_replace_all(l1_all_disease_terms,
                          pattern = paste0(vascular_insufficiency_terms, collapse = "(?=,|$)|\\b"),
                          "vascular insufficiency"
         )
  )

```

# Disease subtype grouping (non-cancer)

## Abornmal total eosinophil count

```{r abnormal_total_eosinophil_count}

url <- "http://www.ebi.ac.uk/ols4/api/ontologies/hp/terms/http%253A%252F%252Fpurl.obolibrary.org%252Fobo%252FHP_0020064/descendants"

abnormal_total_eosinophil_count_terms <- get_descendants(url)

gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms  = 
         stringr::str_replace_all(l1_all_disease_terms,
                          pattern = paste0(abnormal_total_eosinophil_count_terms, collapse = "(?=,|$)|\\b"),
                          "abnormal total eosinophil count"
         ))


```

## Abortion

```{r pregnancy_loss}

pregnancy_loss_terms <- c("habitual abortion",
                          "incomplete abortion",
                          "spontaneous abortion",
                          "abortion",
                          "spontaneous loss of pregnancy",
                          "incomplete loss of pregnancy"
                          )

gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms = 
         stringr::str_replace_all(l1_all_disease_terms,
                          paste0(pregnancy_loss_terms, collapse = "(?=,|$)|\\b"),
                          "loss of pregnancy"
         ))

# exact synonyms (to avoid partial matches
# https://www.ebi.ac.uk/ols4/ontologies/efo/classes/http%253A%252F%252Fwww.ebi.ac.uk%252Fefo%252FEFO_1001491?lang=en

```

## Acute pancreatitis

```{r acute_pancreatitis}

gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms  = 
         stringr::str_replace_all(l1_all_disease_terms,
                          "asparaginase-induced acute pancreatitis",
                          "acute pancreatitis"
         ))

```

## Age-related macular degeneration

```{r age_related_macular_degeneration}

# from: http://purl.obolibrary.org/obo/HP_0011141
age_related_macular_degeneration_terms <- c("age-related cortical cataract",
                                             "age-related nuclear cataract",
                                             "age-related posterior subcapsular cataract"
                                         )

url <- "http://www.ebi.ac.uk/ols4/api/ontologies/efo/terms/http%253A%252F%252Fwww.ebi.ac.uk%252Fefo%252FEFO_0001365/descendants"

age_related_macular_degeneration_terms <- c(age_related_macular_degeneration_terms,
                                             get_descendants(url)
                                             )

gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms  = 
         stringr::str_replace_all(l1_all_disease_terms,
                          pattern = paste0(age_related_macular_degeneration_terms, collapse = "(?=,|$)|\\b"),
                          "age-related macular degeneration"
         ))


```

## Alcohol-use disorders

```{r alcohol_use_disorders}

alcohol_use_disorder_terms <- c("alcohol dependence",
                             "alcohol withdrawal",
                             "alcohol withdrawal delirium",
                             "alcohol abuse",
                             "alcohol-related disorders delirium",
                             "alcohol use disorder",
                             "addictive alcohol use"
                             )

gwas_study_info = gwas_study_info |>
      mutate(l1_all_disease_terms = 
           stringr::str_replace_all(l1_all_disease_terms,
                            paste0(alcohol_use_disorder_terms, collapse = "(?=,|$)|\\b"),
                            "alcohol-related disorders"
           ))

gwas_study_info = gwas_study_info |>
      mutate(l1_all_disease_terms = 
           stringr::str_replace_all(l1_all_disease_terms,
                            paste0(alcohol_use_disorder_terms, collapse = "(?=,|$)|\\b"),
                            "alcohol-related disorders"
           ))


```


## Alcoholic liver disease

```{r alcoholic_liver_disease}

alcoholic_liver_disease_terms <- c("alcoholic liver cirrhosis",
                                   "alcoholic fatty liver disease",
                                   "alcoholic hepatitis"
                                   )


gwas_study_info = gwas_study_info |>
      mutate(l1_all_disease_terms = 
           stringr::str_replace_all(l1_all_disease_terms,
                             paste0(alcoholic_liver_disease_terms, collapse = "(?=,|$)|\\b"),
                            "alcoholic liver disease"
           ))

```

## Alzheimer's disease

```{r alzheimers_disease}

gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms  = 
         stringr::str_replace_all(l1_all_disease_terms,
                          "early-onset alzheimers disease|late-onset alzheimers disease",
                          "alzheimers disease"
         ))

```

## Allergic contact dermatitis

```{r}

gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms  = 
         stringr::str_replace_all(l1_all_disease_terms,
                          "allergic contact dermatitis of eyelid",
                          "allergic contact dermatitis"
         ))

```

## Allergic rhinitis

```{r allergic_rhinitis}

gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms  = 
         stringr::str_replace_all(l1_all_disease_terms,
                          "seasonal allergic rhinitis",
                          "allergic rhinitis"
         ))

```

## Aplastic anemia

```{r aplastic_anemia}

url <- "http://www.ebi.ac.uk/ols4/api/ontologies/mondo/terms/http%253A%252F%252Fpurl.obolibrary.org%252Fobo%252FMONDO_0015909/descendants"

aplastic_anemia_terms <- get_descendants(url)

aplastic_anemia_terms <- c("severe aplastic anemia",
                           aplastic_anemia_terms
                          )

gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms  = 
         stringr::str_replace_all(l1_all_disease_terms,
                          pattern = paste0(aplastic_anemia_terms, collapse = "(?=,|$)|\\b"),
                          "aplastic anemia"
         ))

```

## Astigmatism

```{r astigmatism}

gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms  = 
         stringr::str_replace_all(l1_all_disease_terms,
                          "corneal astigmatism",
                          "astigmatism"
         ))

```

## Asthma

```{r asthma}

gwas_study_info = gwas_study_info |>
      mutate(l1_all_disease_terms  = 
           stringr::str_replace_all(l1_all_disease_terms,
                            "atopic asthma|chronic obstructive asthma",
                            "asthma"
           )) |>
    mutate(l1_all_disease_terms  = 
           stringr::str_replace_all(l1_all_disease_terms,
                            "childhood onset asthma|adult onset asthma|aspirin-induced asthma",
                            "asthma"
           ))

```



## Autoimmune pancreatitis

```{r autoimmune_pancreatitis}

gwas_study_info = gwas_study_info |>
      mutate(l1_all_disease_terms  = 
           stringr::str_replace_all(l1_all_disease_terms,
                            "autoimmune pancreatitis type 1",
                            "autoimmune pancreatitis"
           )) 


```

## Bipolar disorder

```{r bipolar_disorder}

gwas_study_info = gwas_study_info |> 
    mutate(l1_all_disease_terms  = 
         stringr::str_replace_all(l1_all_disease_terms,
                          "bipolar ii disorder|bipolar i disorder",
                          "bipolar disorder"
         ))

```

## Bone fracture

```{r bone_fracture}

url <- "http://www.ebi.ac.uk/ols4/api/ontologies/efo/terms/http%253A%252F%252Fwww.ebi.ac.uk%252Fefo%252FEFO_0003931/descendants"

bone_fracture_terms <- get_descendants(url)

gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms  = 
         stringr::str_replace_all(l1_all_disease_terms,
                          pattern = paste0(bone_fracture_terms, collapse = "(?=,|$)|\\b"),
                          "bone fracture"
         ))

```

## Bundle branch block 

```{r bundle_branch_block}

bundle_branch_block_terms <- c("left bundle branch block",
                             "right bundle branch block",
                             "complete right bundle branch block"
                             )

gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms  = 
         stringr::str_replace_all(l1_all_disease_terms,
                          pattern = paste0(bundle_branch_block_terms, collapse = "(?=,|$)|\\b"),
                          "bundle branch block"
         )
  )

```


## Cataract

```{r cataract}

url <- "http://www.ebi.ac.uk/ols4/api/ontologies/mondo/terms/http%253A%252F%252Fpurl.obolibrary.org%252Fobo%252FMONDO_0005129/descendants"

cataract_terms <- get_descendants(url)

gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms  = 
         stringr::str_replace_all(l1_all_disease_terms,
                          pattern = paste0(cataract_terms, collapse = "(?=,|$)|\\b"),
                          "cataract"
         ))


```

## Celiac disease

```{r celiac_disease}

gwas_study_info = 
gwas_study_info |> 
 mutate(l1_all_disease_terms  = 
          stringr::str_replace_all(l1_all_disease_terms,
                                  pattern = "refractory celiac disease",
                                   "celiac disease"
                          )  
        )


```

## Charcot-marie-tooth disease type 1a

```{r charcot_marie_tooth_disease_type_1a}

gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms  = 
         stringr::str_replace_all(l1_all_disease_terms,
                          "charcot-marie-tooth disease type 1a, decreased fine motor function",
                          "charcot-marie-tooth disease type 1a"
         ))


gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms  = 
         stringr::str_replace_all(l1_all_disease_terms,
                          "charcot-marie-tooth disease type 1a, decreased walking ability",
                          "charcot-marie-tooth disease type 1a"
         ))

gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms  = 
         stringr::str_replace_all(l1_all_disease_terms,
                          "charcot-marie-tooth disease type 1a, gait imbalance",
                          "charcot-marie-tooth disease type 1a, gait disturbance"
         ))



```

## Chronic pain

```{r chronic_pain}

url <- "http://www.ebi.ac.uk/ols4/api/ontologies/efo/terms/http%253A%252F%252Fpurl.obolibrary.org%252Fobo%252FHP_0012532/descendants"

chronic_pain_terms <- get_descendants(url)

gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms  = 
         stringr::str_replace_all(l1_all_disease_terms,
                          pattern = paste0(chronic_pain_terms, collapse = "(?=,|$)|\\b"),
                          "chronic pain"
         ))

```

## Chronic rhinosinusitis 

```{r chronic_rhinosinusitis}

gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms  = 
         stringr::str_replace_all(l1_all_disease_terms,
                          "chronic rhinosinusitis with nasal polyps",
                          "chronic rhinosinusitis"
         ))

```

## Cholecystitis

```{r cholecystitis}

gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms  = 
         stringr::str_replace_all(l1_all_disease_terms,
                          "acute cholecystitis",
                          "cholecystitis"
         ))

```

## Communication disorder

```{r communication_disorder}

url <- "http://www.ebi.ac.uk/ols4/api/ontologies/mondo/terms/http%253A%252F%252Fpurl.obolibrary.org%252Fobo%252FMONDO_0002182/descendants"

communication_disorder_terms <- get_descendants(url)

gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms  = 
         stringr::str_replace_all(l1_all_disease_terms,
                          pattern = paste0(communication_disorder_terms, collapse = "(?=,|$)|\\b"),
                          "communication disorder"
         ))


```



## Creutzfeldt jacob disease

```{r creutzfeldt_jacob_disease}

gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms  = 
         stringr::str_replace_all(l1_all_disease_terms,
                          "sporadic creutzfeld jacob disease",
                          "creutzfeldt jacob disease"
         ))

```

## Delirium

```{r delirium}

gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms  = 
         stringr::str_replace_all(l1_all_disease_terms,
                          "post-operative delirium",
                          "delirium, post-operative"
         ))

```

## Dental caries

```{r dental_caries}


dental_caries_terms <- c("pit and fissure surface dental caries",
                         "smooth surface dental caries",
                         "primary dental caries",
                         "enamel caries",
                         "permanent dental caries"
                         )

gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms  = 
         stringr::str_replace_all(l1_all_disease_terms,
                          pattern = paste0(dental_caries_terms, collapse = "(?=,|$)|\\b"),
                          "dental caries"
         )
)

```

## Dermatitis

```{r dermatitis}

url <- "http://www.ebi.ac.uk/ols4/api/ontologies/doid/terms/http%253A%252F%252Fpurl.obolibrary.org%252Fobo%252FDOID_2723/descendants"

dermatitis_terms <- get_descendants(url)

gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms  = 
         stringr::str_replace_all(l1_all_disease_terms,
                          pattern = paste0(dermatitis_terms, collapse = "(?=,|$)|\\b"),
                          "dermatitis"
         ))

# see: http://www.ebi.ac.uk/efo/EFO_0000274
eczema_terms <- c("atopic eczema",
                 "hand eczema",
                 "eczematoid dermatitis", # see: http://purl.obolibrary.org/obo/HP_0000964
                 "recalcitrant dermatitis" # assuming same as recalcitrant atopic dermatitis http://www.ebi.ac.uk/efo/EFO_1000651
                 )

gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms  = 
         stringr::str_replace_all(l1_all_disease_terms,
                          pattern = paste0(eczema_terms, collapse = "(?=,|$)|\\b"),
                          "dermatitis"
         ))

```

## Drug allergy

```{r drug_allergy}

gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms  = 
         stringr::str_replace_all(l1_all_disease_terms,
                          "beta-lactam allergy",
                          "drug allergy"
         ))

```

## Diabetic eye disease

```{r}

diabetic_eye_terms <- c("diabetic maculopathy",
                        "diabetic macular edema",
                        "diabetic retinopathy",
                        "proliferative diabetic retinopathy",
                        "non-proliferative diabetic retinopathy",
                        "diabetes mellitus type 2 associated cataract")


gwas_study_info = 
  gwas_study_info |> 
  mutate(l1_all_disease_terms  = 
           stringr::str_replace_all(l1_all_disease_terms ,
                                    pattern = paste0(diabetic_eye_terms, collapse = "(?=,|$)|\\b"),
                                    "diabetic eye disease"
           )  
  )

```


## Diabetic neuropathy

```{r diabetic_nephropathy}

gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms  = 
         stringr::str_replace_all(l1_all_disease_terms,
                          "diabetic polyneuropathy",
                          "diabetic neuropathy"
         ))

```

## Deficiency anemia

```{r deficiency_anemia}

gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms  = 
         stringr::str_replace_all(l1_all_disease_terms,
                          "iron deficiency anemia",
                          "deficiency anemia"
         ))

```

## Encephalopathy

```{r encephalopathy}

gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms  = 
         stringr::str_replace_all(l1_all_disease_terms,
                          "delayed encephalopathy after acute poisoning",
                          "encephalopathy, poisoning"
         ))


gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms  = 
         stringr::str_replace_all(l1_all_disease_terms,
                          "encephalopathy acute infection-induced",
                          "encephalopathy"
         ))

```

]

## Food allergy

```{r food_allergy}

gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms  = 
         stringr::str_replace_all(l1_all_disease_terms,
                          "peanut allergy|milk allergy|egg allergy|wheat allergic reaction",
                          "food allergy"
         ))
```

## Glaucoma

```{r glaucoma}

url <- "http://www.ebi.ac.uk/ols4/api/ontologies/efo/terms/http%253A%252F%252Fpurl.obolibrary.org%252Fobo%252FMONDO_0005041/descendants"

glaucoma_terms <- get_descendants(url)

gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms  = 
         stringr::str_replace_all(l1_all_disease_terms,
                          pattern = paste0(glaucoma_terms, collapse = "(?=,|$)|\\b"),
                          "glaucoma"
         ))


```

## Gout

```{r gout}

# from: http://www.ebi.ac.uk/efo/EFO_0004274
gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms  = 
         stringr::str_replace_all(l1_all_disease_terms,
                          "renal overload-type gout",
                          "gout"
         ))

```


## Graft vs host disease

```{r graft_vs_host_disease}

graft_vs_host_terms <- c("chronic graft versus host disease",
                         "chronic graft vs. host disease",
                         "acute graft versus host disease",
                         "acute graft vs. host disease"
                         )

gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms  = 
         stringr::str_replace_all(l1_all_disease_terms,
                          paste0(graft_vs_host_terms, collapse = "(?=,|$)|\\b"),
                          "graft versus host disease"
         ))

```



## Hemolytic anemia

```{r hemolytic_anemia}

url <- "http://www.ebi.ac.uk/ols4/api/ontologies/mondo/terms/http%253A%252F%252Fpurl.obolibrary.org%252Fobo%252FMONDO_0003664/descendants"


hemolytic_anemia_terms <- get_descendants(url)

gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms  = 
         stringr::str_replace_all(l1_all_disease_terms,
                          pattern = paste0(hemolytic_anemia_terms, collapse = "(?=,|$)|\\b"),
                          "hemolytic anemia"
         ))

```

## Hernia of the abdominal wall 

```{r hernia_abdominal_wall}

url <- "http://www.ebi.ac.uk/ols4/api/ontologies/hp/terms/http%253A%252F%252Fpurl.obolibrary.org%252Fobo%252FHP_0004299/children"

hernia_abdominal_wall_terms <- get_descendants(url)

gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms  = 
         stringr::str_replace_all(l1_all_disease_terms,
                          pattern = paste0(hernia_abdominal_wall_terms, collapse = "(?=,|$)|\\b"),
                          "hernia of the abdominal wall"
         ))


```

## Herpes

```{r}


gwas_study_info = 
gwas_study_info |> 
 mutate(l1_all_disease_terms  = 
          stringr::str_replace_all(l1_all_disease_terms,
                                  pattern = "herpes simplex infection",
                                   "herpesviridae infectious disease"
                          )  
        )


```

## HIV 

```{r}

gwas_study_info = gwas_study_info |>
    mutate(l1_all_disease_terms = 
         stringr::str_replace_all(l1_all_disease_terms,
                          "hiv-1 infection",
                          "hiv infection"
         ))


```

## Hyperlipidemia

```{r}

hyperlipidemia_terms <- c("hypercholesterolemia",
                         "familial hypercholesterolemia",
                         "familial hyperlipidemia"
                         )

gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms  = 
         stringr::str_replace_all(l1_all_disease_terms,
                          pattern = paste0(hyperlipidemia_terms, collapse = "(?=,|$)|\\b"),
                          "hyperlipidemia"
         ))
  
```

## Idiopathic generalized epilepsy

```{r idiopathic_generalized_epilepsy}

url <- "http://www.ebi.ac.uk/ols4/api/ontologies/snomed/terms/http%253A%252F%252Fsnomed.info%252Fid%252F36803009/descendants"

idiopathic_generalized_epilepsy_terms <- get_descendants(url)

idiopathic_generalized_epilepsy_terms = stringr::str_remove_all(
                                        idiopathic_generalized_epilepsy_terms, 
                                        " \\(disorder\\)$"
                                        )


gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms  = 
         stringr::str_replace_all(l1_all_disease_terms,
                          pattern = paste0(idiopathic_generalized_epilepsy_terms, collapse = "(?=,|$)|\\b"),
                          "idiopathic generalized epilepsy"
         ))


```

## Inborn errors of metabolism

```{r}

inborn_error_metab <- c("inborn disorder of amino acid metabolism",
                        "inborn disorder of amino acid transport",
                        "inborn disorder of porphyrin metabolism",
                        "inborn carbohydrate metabolic disorder",
                        "familial lipoprotein lipase deficiency",
                        "lactose intolerance",
                        "hereditary hemochromatosis",
                        "alpha 1-antitrypsin deficiency",
                        "urea cycle disorder",
                        "gaucher disease",
                        "plasma protein metabolism disease",
                        "disorder of metabolite absorption and transport")

gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms  = 
         stringr::str_replace_all(l1_all_disease_terms,
                          pattern = paste0(inborn_error_metab, collapse = "(?=,|$)|\\b"),
                          "inborn errors of metabolism"
         ))


```

## Infectious meningitis

```{r}

gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms  = 
         stringr::str_replace_all(l1_all_disease_terms,
                          "bacterial meningitis|pneumococcal meningitis|viral meningitis",
                          "infectious meningitis"
         ))

```

## Infertility (female & male)

```{r}

gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms  = 
         stringr::str_replace_all(l1_all_disease_terms,
                          "tubal factor infertility",
                          "female infertility"
                          )) |>
  mutate(l1_all_disease_terms  = 
                  stringr::str_replace_all(l1_all_disease_terms,
                          "azoospermia",
                          "male infertility"
                          )  )

```

## Influenza a (h1n1) (subset of influenza)

```{r}

gwas_study_info = gwas_study_info |>
    mutate(l1_all_disease_terms  = 
         stringr::str_replace_all(l1_all_disease_terms,
                          "influenza a \\(h1n1\\)",
                          "influenza"
         ))

```

## Intestinal obstruction

```{r intestinal_obstruction}

url <- "http://www.ebi.ac.uk/ols4/api/ontologies/mondo/terms/http%253A%252F%252Fpurl.obolibrary.org%252Fobo%252FMONDO_0004565/descendants"

intestinal_obstruction_terms <- get_descendants(url)

# also add: http://purl.obolibrary.org/obo/DOID_8437

intestinal_obstruction_terms = c(intestinal_obstruction_terms,
                                 "intestinal impaction"
                                 )

gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms  = 
         stringr::str_replace_all(l1_all_disease_terms,
                          pattern = paste0(intestinal_obstruction_terms, collapse = "(?=,|$)|\\b"),
                          "intestinal obstruction"
         ))


```



## Juvenile idiopathic arthritis

```{r juvenile_idiopathic_arthritis_onset_group}

jia_terms <- c("systemic juvenile idiopathic arthritis",
                "oligoarticular juvenile idiopathic arthritis",
                "polyarticular juvenile idiopathic arthritis"
                )

gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms  = 
        stringr::str_replace_all(l1_all_disease_terms,
                         paste0(jia_terms, collapse = "(?=,|$)|\\b"),
                         "juvenile idiopathic arthritis")
        )

```

## Kidney disease 

```{r kidney_disease}

url <- "http://www.ebi.ac.uk/ols4/api/ontologies/doid/terms/http%253A%252F%252Fpurl.obolibrary.org%252Fobo%252FDOID_1074/descendants"

kidney_disease_terms <- get_descendants(url)

gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms  = 
         stringr::str_replace_all(l1_all_disease_terms,
                          pattern = paste0(kidney_disease_terms, collapse = "(?=,|$)|\\b"),
                          "kidney disease"
         ))


```

## Laryngitis

```{r laryngitis}

laryngitis_terms <- c("acute laryngitis",
                      "chronic laryngitis"
                      )

gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms  = 
         stringr::str_replace_all(l1_all_disease_terms,
                          pattern = paste0(laryngitis_terms, collapse = "(?=,|$)|\\b"),
                          "laryngitis"
         ))


```

## Lewy body dementia

```{r lewy_body_dementia}

gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms  = 
         stringr::str_replace_all(l1_all_disease_terms,
                          "lewy body attribute",
                          "lewy body dementia"
         ))


```
## Malaria

```{r malaria}

gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms  = 
         stringr::str_replace_all(l1_all_disease_terms,
                          "plasmodium falciparum malaria|plasmodium vivax malaria",
                          "malaria"
         ))

```

## Methamphetamine use disorders 

```{r methamphetamine_use_disorders}

gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms  = 
         stringr::str_replace_all(l1_all_disease_terms,
                          "methamphetamine dependence|methamphetamine-induced psychosis",
                          "methamphetamine use disorders"
         ))

```

## Migraine 

```{r migraine}

gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms  = 
         stringr::str_replace_all(l1_all_disease_terms,
                          "migraine disorder|migraine with aura|migraine without aura",
                          "migraine"
         ))

```

## Multiple sclerosis

```{r multiple_sclerosis}

gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms  = 
         stringr::str_replace_all(l1_all_disease_terms,
                          "relapsing-remitting multiple sclerosis",
                          "multiple sclerosis"
         ))

```



## Photosensitivity disease

```{r photosensitivity_disease}

gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms  = 
        stringr::str_replace_all(l1_all_disease_terms,
                         "phototoxic dermatitis|skin sensitivity to sun",
                         "photosensitivity disease")
        )

```

## Phobic disorder

```{r phobic_disorder}

url <- "http://www.ebi.ac.uk/ols4/api/ontologies/efo/terms/http%253A%252F%252Fwww.ebi.ac.uk%252Fefo%252FEFO_1001908/descendants"

phobic_disorder_terms <- get_descendants(url)

gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms  = 
         stringr::str_replace_all(l1_all_disease_terms,
                          pattern = paste0(phobic_disorder_terms, collapse = "(?=,|$)|\\b"),
                          "phobic disorder"
         ))

```

## Hereditary hemochromatosis

```{r}

gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms  = 
         stringr::str_replace_all(l1_all_disease_terms,
                          "hereditary hemochromatosis type 1",
                          "hereditary hemochromatosis"
         ))

```




## Myasthenia gravis

```{r}

gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms  = 
         stringr::str_replace_all(l1_all_disease_terms,
                          "late-onset myasthenia gravis",
                          "myasthenia gravis"
         ))


```



## Neuromyelitis optica 

```{r}

gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms  = 
         stringr::str_replace_all(l1_all_disease_terms,
                          "aqp4-igg-negative neuromyelitis optica|aqp4-igg-positive neuromyelitis optica",
                          "neuromyelitis optica"
         ))

```

## Neurofibromatosis

```{r}

gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms  = 
         stringr::str_replace_all(l1_all_disease_terms,
                          "neurofibromatosis type 1|neurofibromatosis type 2",
                          "neurofibromatosis"
         ))

```

## Non-alcoholic fatty liver disease (NAFLD)

```{r}

gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms  = 
         stringr::str_replace_all(l1_all_disease_terms,
                          "non-alcoholic steatohepatitis|non-alcoholic liver disease",
                          "non-alcoholic fatty liver disease"
         ))

```


## Obesity

```{r}

gwas_study_info = 
gwas_study_info |> 
 mutate(l1_all_disease_terms  = 
          stringr::str_replace_all(l1_all_disease_terms,
                                  pattern = "morbid obesity|metabolically healthy obesity",
                                   "obesity"
                          )  
        )

```


## Obsessive-compulsive disorder

```{r obsessive_compulsive_disorder}

gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms  = 
         stringr::str_replace_all(l1_all_disease_terms,
                          "obsessive-compulsive trait",
                          "obsessive-compulsive disorder"
         ))

```


## Osteoarthritis

```{r osteoarthritis}

osteoarthritis_terms <- c("knee osteoarthritis",
                         "hip osteoarthritis",
                         "hand osteoarthritis",
                         "spine osteoarthritis",
                         "toe osteoarthritis"
                         )

gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms  = 
         stringr::str_replace_all(l1_all_disease_terms,
                          pattern = paste0(osteoarthritis_terms, collapse = "(?=,|$)|\\b"),
                          "osteoarthritis"
         )
  )

```


## Osteonecrosis

```{r osteonecrosis}

gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms  = 
         stringr::str_replace_all(l1_all_disease_terms,
                          "idiopathic osteonecrosis of the femoral head",
                          "osteonecrosis"
         ))

```

## Pancreatitis 

```{r pancreatitis}

url <- "http://www.ebi.ac.uk/ols4/api/ontologies/efo/terms/http%253A%252F%252Fwww.ebi.ac.uk%252Fefo%252FEFO_0000278/descendants"

pancreatitis_terms <- get_descendants(url)

gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms  = 
         stringr::str_replace_all(l1_all_disease_terms,
                          pattern = paste0(pancreatitis_terms, collapse = "(?=,|$)|\\b"),
                          "pancreatitis"
         ))


```

## Peridontal disease

```{r periodontal}

url <- "http://www.ebi.ac.uk/ols4/api/ontologies/doid/terms/http%253A%252F%252Fpurl.obolibrary.org%252Fobo%252FDOID_3388/descendants"

periodontal_terms <- get_descendants(url)

gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms  = 
         stringr::str_replace_all(l1_all_disease_terms,
                          pattern = paste0(periodontal_terms, collapse = "(?=,|$)|\\b"),
                          "periodontal disease"
         ))

```

## Poisoning

```{r poisoning}

gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms  = 
         stringr::str_replace_all(l1_all_disease_terms,
                          "carbon monoxide poisoning",
                          "poisoning"
         ))

```

## Polycythemia

```{r polycythemia}

url <- "http://www.ebi.ac.uk/ols4/api/ontologies/efo/terms/http%253A%252F%252Fwww.ebi.ac.uk%252Fefo%252FEFO_0005804/descendants"

polycythemia_terms <- get_descendants(url)

gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms  = 
         stringr::str_replace_all(l1_all_disease_terms,
                          pattern = paste0(polycythemia_terms, collapse = "(?=,|$)|\\b"),
                          "polycythemia"
         ))

```


## Pulmonary embolism

```{r pulmonary_embolism}

gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms  = 
         stringr::str_replace_all(l1_all_disease_terms,
                          "pulmonary embolism, pulmonary infarction",
                          "pulmonary embolism"
         ))

```

## Pulmonary fibrosis

```{r pulmonary_fibrosis}

gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms  = 
         stringr::str_replace_all(l1_all_disease_terms,
                          "idiopathic pulmonary fibrosis",
                          "pulmonary fibrosis"
         ))

```

## Psoriasis

```{r}

psoriasis_terms <- c("psoriasis vulgaris",
                      "psoriasis area and severity index",
                      "cutaneous psoriasis",
                      "psoriatic arthritis",
                     "parapsoriasis")


gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms  = 
         stringr::str_replace_all(l1_all_disease_terms,
                          pattern = paste0(psoriasis_terms, collapse = "(?=,|$)|\\b"),
                          "psoriasis")
  )

```


## Psychosis

```{r psychosis}

gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms  = 
         stringr::str_replace_all(l1_all_disease_terms,
                          "psychosis predisposition",
                          "psychosis")
         )

```

## Retinopathy

```{r retinopathy}

retinopathy_terms <- c("chronic central serous retinopathy",
                       "central serous retinopathy"
                       )

gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms = 
           stringr::str_replace_all(l1_all_disease_terms,
                          pattern = paste0(retinopathy_terms, collapse = "(?=,|$)|\\b"),
                          "retinopathy")
  )

```


## Rheumatoid arthritis

```{r}

ra_terms <- c("acpa-positive rheumatoid arthritis",
               "acpa-negative rheumatoid arthritis",
              "adult-onset stills disease")

gwas_study_info = 
gwas_study_info |> 
 mutate(l1_all_disease_terms  = 
          stringr::str_replace_all(l1_all_disease_terms,
                                  pattern = paste0(ra_terms, collapse = "(?=,|$)|\\b"),
                                   "rheumatoid arthritis"
                          )  
        )

```

## Schizoaffective disorder

```{r schizoaffective_disorder}

gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms  = 
         stringr::str_replace_all(l1_all_disease_terms,
                          "schizoaffective disorder-bipolar type",
                          "schizoaffective disorder")
         )

```

## Schizophrenia

```{r schizophrenia}

gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms  = 
         stringr::str_replace_all(l1_all_disease_terms,
                          "treatment refractory schizophrenia",
                          "schizophrenia")
         )

```

## Sciatica

```{r sciatica}

gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms  = 
         stringr::str_remove_all(l1_all_disease_terms,
                          "^ldh-related sciatica, |, ldh-related sciatica$"
         )
  )

```

## Scoliosis

```{r}

gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms  = 
         stringr::str_replace_all(l1_all_disease_terms,
                          "adolescent idiopathic scoliosis",
                          "scoliosis")
         )

```

## Sickle cell disease 

```{r}

gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms  = 
         stringr::str_replace_all(l1_all_disease_terms,
                          "sickle cell anemia",
                          "sickle cell disease and related diseases")
         )

```

## Sleep apnea 

```{r}

sleep_apnea_terms <- c("sleep apnea measurement during non-rem sleep",
                      "sleep apnea measurement during rem sleep",
                      "obstructive sleep apnea")

gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms  = 
         stringr::str_replace_all(l1_all_disease_terms,
                          pattern = paste0(sleep_apnea_terms, collapse = "(?=,|$)|\\b"),
                          "sleep apnea")
  )


```

## Staphylococcus aureus infection

```{r staphylococcus_aureus_infection}

gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms  = 
         stringr::str_replace_all(l1_all_disease_terms,
                          "methicillin-resistant staphylococcus aureus infection",
                          "staphylococcus aureus infection"
         ))

```




## Systemic scleroderma

```{r}

url <- "http://www.ebi.ac.uk/ols4/api/ontologies/efo/terms/http%253A%252F%252Fwww.ebi.ac.uk%252Fefo%252FEFO_0000717/descendants"

scleroderma_terms <- get_descendants(url)

gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms  = 
         stringr::str_replace_all(l1_all_disease_terms,
                          pattern = paste0(scleroderma_terms, collapse = "(?=,|$)|\\b"),
                          "systemic scleroderma"
         ))

```

## Tenosynovitis

```{r}

gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms  = 
        stringr::str_replace_all(l1_all_disease_terms,
                         "stenosing tenosynovitis",
                         "tenosynovitis")
        )

```

## Tuberculosis

```{r tuberculosis}

tb_terms <- c("mycobacterium tuberculosis infection",
              "\\bpulmonary tuberculosis",
              "extrapulmonary tuberculosis",
              "meningeal tuberculosis")

gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms  = 
         stringr::str_replace_all(l1_all_disease_terms,
                          pattern = paste0(tb_terms, collapse = "(?=,|$)|\\b"),
                          "tuberculosis"
         ))

```

## Type 1 diabetes

```{r}

gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms  = 
         stringr::str_replace_all(l1_all_disease_terms,
                          "latent autoimmune diabetes in adults",
                          "type 1 diabetes mellitus"
         ))

```

## Thrombocytopenia

```{r thrombocytopenia}

thrombocytopenia_terms <- c("thrombocytopenia 4",
                           "acquired thrombocytopenia",
                           "primary thrombocytopenia"
                           )

```

## Tinea 

```{r tinea}

tinea_terms <- c("tinea pedis",
                 "tinea unguium",
                 "dermatophytosis"
                 )


gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms  = 
         stringr::str_replace_all(l1_all_disease_terms,
                          pattern = paste0(tinea_terms, collapse = "(?=,|$)|\\b"),
                          "tinea"
         ))

```

## Uveitis

```{r uveitis}

uveitis_terms <- c("anterior uveitis"
                   )

gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms  = 
         stringr::str_replace_all(l1_all_disease_terms,
                          pattern = paste0(uveitis_terms, collapse = "(?=,|$)|\\b"),
                          "uveitis"
         ))

```

## Vitamin b deficiency

```{r}

gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms  = 
         stringr::str_replace_all(l1_all_disease_terms,
                          "vitamin b12 deficiency",
                          "vitamin b deficiency"
         ))

```


## Visual impairment

```{r visual_impairment}

url <- "http://www.ebi.ac.uk/ols4/api/ontologies/efo/terms/http%253A%252F%252Fpurl.obolibrary.org%252Fobo%252FHP_0000505/descendants"

visual_impairment_terms <- get_descendants(url)

gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms  = 
         stringr::str_replace_all(l1_all_disease_terms,
                          pattern = paste0(visual_impairment_terms, collapse = "(?=,|$)|\\b"),
                          "visual impairment"
         ))

```


# Further grouping of terms

## Abnormal refraction

```{r abnormal_refraction}

url <- "http://www.ebi.ac.uk/ols4/api/ontologies/upheno/terms/http%253A%252F%252Fpurl.obolibrary.org%252Fobo%252FHP_0000539/descendants"

abnormal_refraction_terms <- get_descendants(url)

gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms  = 
         stringr::str_replace_all(l1_all_disease_terms,
                          pattern = paste0(abnormal_refraction_terms, collapse = "(?=,|$)|\\b"),
                          "abnormality of refraction"
         ))

```

# Final summary - number of unique study terms

## Deal with duplicate terms created during grouping

```{r duplicate_terms}

gwas_study_info = 
 gwas_study_info |>
  rowwise() |>
  mutate(l1_all_disease_terms = paste0(sort(unique(unlist(strsplit(l1_all_disease_terms, ", ")))),
                                      collapse = ", ")
         ) |>
  ungroup()


```

## Deal with hanging commas and spaces

```{r clean_hanging_commas}

gwas_study_info = gwas_study_info |>
  mutate(l1_all_disease_terms = stringr::str_remove_all(l1_all_disease_terms, "^,|,$")
         ) |>
  mutate(l1_all_disease_terms = stringr::str_trim(l1_all_disease_terms)
         ) 

```

## Final summary - number of unique study terms pairs 

```{r final_summary}

n_studies_trait = gwas_study_info |>
  dplyr::filter(DISEASE_STUDY == T) |>
  dplyr::select(l1_all_disease_terms, PUBMED_ID) |>
  dplyr::distinct() |>
  dplyr::group_by(l1_all_disease_terms) |>
  dplyr::summarise(n_studies = dplyr::n()) |>
  dplyr::arrange(desc(n_studies))


head(n_studies_trait)

dim(n_studies_trait)

```

### When separate studies with multiple terms

```{r final_diseases}

diseases <- stringr::str_split(pattern = ", ", 
                               gwas_study_info$l1_all_disease_terms[gwas_study_info$l1_all_disease_terms != ""])  |> 
            unlist() |>
            stringr::str_trim()


test <- data.frame(trait = unique(diseases))

length(unique(diseases))

# make frequency table
freq <- table(as.factor(diseases))

# sort in decreasing order
freq_sorted <- sort(freq, decreasing = TRUE)

# show top N, e.g. top 10
head(freq_sorted, 10)

```

### Save the updated gwas_study_info with harmonized disease terms

```{r}

fwrite(gwas_study_info,
        here::here("output/gwas_cat/gwas_study_info_group_l1.csv")
         )

```



