---
title: "More disease grouping to match GBD"
author: "Isobel Beasley"
date: "2025-09-08"
output:
  workflowr::wflow_html:
    toc: true
    toc_float: true
    number_sections: true
editor_options: 
  markdown: 
    wrap: 72
---

# Set up

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r required_fns, message = F, warning= F}

library(dplyr)
library(data.table)
library(ggplot2)
library(stringr)

```

## Ontology help - for getting disease subtypes

```{r ontology_fns}

source(here::here("code/get_term_descendants.R"))

```

```{r setup_2, message = FALSE, warning= FALSE}

gwas_study_info <- fread(here::here("output/gwas_cat/gwas_study_info_group_l1_v2.csv"))

```


# Objectives: 

- Further group disease terms (level 2 categories) to match GBD (globalqburden of disease) categories more closely.

## Grouping - level 2 set up 

```{r group_l2}

gwas_study_info = gwas_study_info |>
  mutate(l2_all_disease_terms = l1_all_disease_terms)

```


# Rename disease terms to match GBD categories more closely

## Acne -> acne vulgaris

```{r acne}

gwas_study_info = gwas_study_info |> 
 mutate(l2_all_disease_terms  = 
          stringr::str_replace_all(l2_all_disease_terms ,
                                  pattern = "acne",
                                   "acne vulgaris"
                          )  
        )



```

## ADHD

```{r adhd}


gwas_study_info = gwas_study_info |> 
 mutate(l2_all_disease_terms  = 
          stringr::str_replace_all(l2_all_disease_terms ,
                                  pattern = "adhd",
                                   "attention-deficit/hyperactivity disorder"
                          )  
        )


```

## Age-related and other hearing loss 

```{r hearing_loss}


hearing_loss_terms <- c("sensorineural hearing loss",
                        "age-related hearing impairment",
                        "noise-induced hearing loss")

pattern = paste0(hearing_loss_terms, collapse = "|")


gwas_study_info = 
gwas_study_info |> 
 mutate(l2_all_disease_terms  = 
          stringr::str_replace_all(l2_all_disease_terms ,
                                  pattern = pattern,
                                   "age-related and other hearing loss"
                          )  
        )



gwas_study_info = 
gwas_study_info |> 
 mutate(l2_all_disease_terms  = 
          stringr::str_replace_all(l2_all_disease_terms ,
                                  pattern = "hearing loss",
                                   "age-related and other hearing loss"
                          )  
        )

```

## Alcohol use disorders

```{r alcohol_use_disorder}

gwas_study_info = 
gwas_study_info |> 
 mutate(l2_all_disease_terms  = 
          stringr::str_replace_all(l2_all_disease_terms ,
                                  pattern = "alcohol-related disorders|alcohol and nicotine codependence",
                                   "alcohol use disorders"
                          )  
        )

gwas_study_info = 
gwas_study_info |> 
 mutate(l2_all_disease_terms  = 
          stringr::str_replace_all(l2_all_disease_terms ,
                                  pattern = "alcohol use disorder",
                                   "alcohol use disorders"
                          )  
        )

```


## Alzheimer's disease and other dementias 

```{r dementia}

dementia <- c("alzheimers disease biomarker measurement",
              "alzheimers disease neuropathologic change",
              "aids dementia",
              "dementia",
              "frontotemporal dementia",
              "lewy body dementia",
              "vascular dementia",
              "alzheimers disease"
)

gwas_study_info = 
gwas_study_info |> 
 mutate(l2_all_disease_terms  = 
          stringr::str_replace_all(l2_all_disease_terms ,
                                  pattern = paste0(dementia, collapse = "|"),
                                   "alzheimer's disease and other dementias"
                          )  
        )




```


## Anxiety disorders 

```{r anxiety_disordders}

url <- "http://www.ebi.ac.uk/ols4/api/ontologies/mesh/terms/http%253A%252F%252Fid.nlm.nih.gov%252Fmesh%252FD001008/descendants"

anxiety_terms <- get_descendants(url)

anxiety_terms <- c(anxiety_terms, "obsessive-compulsive symptom measurement")

gwas_study_info = 
gwas_study_info |> 
 mutate(l2_all_disease_terms  = 
          stringr::str_replace_all(l2_all_disease_terms ,
                                  pattern = paste0(anxiety_terms, collapse = "|"),
                                   "anxiety disorders"
                          )  
        ) |>
   mutate(l2_all_disease_terms  = 
          stringr::str_replace_all(l2_all_disease_terms ,
                                  pattern = "anxiety disorder|anxiety measurement",
                                   "anxiety disorders"
                          )  
        ) |>
     mutate(l2_all_disease_terms  = 
          stringr::str_replace_all(l2_all_disease_terms ,
                                  pattern = "anxiety",
                                   "anxiety disorders"
                          )  
        )



```

## Atrial fibrillation & flutter

```{r afib_terms}

afib_terms <- c("atrial fibrillation",
                "atrial flutter",
                "post-operative atrial fibrillation")

gwas_study_info = 
gwas_study_info |> 
 mutate(l2_all_disease_terms  = 
          stringr::str_replace_all(l2_all_disease_terms ,
                                  pattern = paste0(afib_terms, collapse = "|"),
                                   "atrial fibrillation and flutter"
                          )  
        )


```

## Amphetamine use disorders

```{r amphetamine_use_disorders}

gwas_study_info = 
gwas_study_info |> 
 mutate(l2_all_disease_terms  = 
          stringr::str_replace_all(l2_all_disease_terms ,
                                  pattern = "methamphetamine",
                                   "amphetamine"
                          )  
        )


```

## Aortic aneurysm

```{r aortic_aneurysm_terms}

url <- "http://www.ebi.ac.uk/ols4/api/ontologies/cvdo/terms/http%253A%252F%252Fpurl.obolibrary.org%252Fobo%252FDOID_3627/descendants"

aortic_aneurysm_terms <- get_descendants(url)

gwas_study_info = gwas_study_info |> 
 mutate(l2_all_disease_terms  = 
          stringr::str_replace_all(l2_all_disease_terms ,
                                  pattern = paste0(aortic_aneurysm_terms, collapse = "|"),
                                   "aortic aneurysm"
                          )  
        )

```

## Autism spectrum disorders

```{r autism}


gwas_study_info = gwas_study_info |> 
 mutate(l2_all_disease_terms  = 
          stringr::str_replace_all(l2_all_disease_terms,
                                  pattern = "autism",
                                   "autism spectrum disorders"
                          )  
        )


```

## Blindness and vision loss 

```{r blindness}

gwas_study_info = gwas_study_info |> 
 mutate(l2_all_disease_terms  = 
          stringr::str_replace_all(l2_all_disease_terms ,
                                  pattern = "blindness|color vision disorder|vision disorder|myopia|refractive error|hyperopia|astigmatism|corneal astigmatism|presbyopia|anisometropia|esotropia|on-accomodative esotropia|ccommodative esotropia",
                                   "blindness and vision loss"
                          )  
        )

```

## Cannabis use disorders
 
```{r cannabis_use}

gwas_study_info = 
gwas_study_info |> 
 mutate(l2_all_disease_terms  = 
          stringr::str_replace_all(l2_all_disease_terms ,
                                  pattern = "cannabis dependence",
                                   "cannabis use disorders"
                          )  
        )

```

### Cancers

#### Bladder cancer 

```{r}


gwas_study_info = 
gwas_study_info |> 
 mutate(l2_all_disease_terms  = 
          stringr::str_replace_all(l2_all_disease_terms ,
                                  pattern = "urinary bladder cancer, disease progression measurement|b;adder tumor",
                                   "bladder cancer"
                          )  
        )


```

#### Colon and rectum cancer

```{r colorectal_cancer_terms}


gwas_study_info = gwas_study_info |> 
 mutate(l2_all_disease_terms  = 
          stringr::str_replace_all(l2_all_disease_terms ,
                                  pattern = "colorectal cancer",
                                   "colon and rectum cancer"
                          )  
        )


```

#### Eye cancer

```{r eye_cancer}


gwas_study_info = 
gwas_study_info |> 
 mutate(l2_all_disease_terms  = 
          stringr::str_replace_all(l2_all_disease_terms ,
                                  pattern = "ocular melanoma|ocular cancer",
                                   "eye cancer"
                          )  
        )

```



#### Hodgkins lymphoma 

```{r}


gwas_study_info = 
gwas_study_info |> 
 mutate(l2_all_disease_terms  = 
          stringr::str_replace_all(l2_all_disease_terms ,
                                  pattern = "hodgkins lymphoma",
                                   "hodgkin lymphoma"
                          )  
        )

```

## Malignant skin melanoma 

```{r}

skin_melanoma <- c("cutaneous melanoma",
                   "melanoma")

```

#### Non-hodgkin lymphoma 

```{r}


gwas_study_info = 
gwas_study_info |> 
 mutate(l2_all_disease_terms  = 
          stringr::str_replace_all(l2_all_disease_terms ,
                                  pattern = "non-hodgkins lymphoma",
                                   "non-hodgkin lymphoma"
                          )  
        )

```


### Tracheal, bronchus, and lung cancer

```{r group_respiratory_cancer}

resp_cancer_terms = c("lung cancer",
                      "bronchus cancer",
                      "respiratory system cancer"
                        )

gwas_study_info = 
gwas_study_info |> 
 mutate(l2_all_disease_terms  = 
          stringr::str_replace_all(l2_all_disease_terms ,
                                  pattern = paste0(resp_cancer_terms, collapse = "|"),
                                   "tracheal bronchus and lung cancer"
                          )  
        )

```

## Cirrhosis & other chronic liver diseases

```{r chronic_liver_disease_terms}

url <- "http://www.ebi.ac.uk/ols4/api/ontologies/snomed/terms/http%253A%252F%252Fsnomed.info%252Fid%252F328383001/descendants"

chronic_liver_disease_terms <- get_descendants(url)

chronic_liver_disease_terms <- c("primary biliary cirrhosis",
                                 "alcoholic liver cirrhosis",
                                 "chronic hepatitis B virus infection", 
                                 "acute-on-chronic liver failure",
                                 chronic_liver_disease_terms)

gwas_study_info = gwas_study_info |> 
 mutate(l2_all_disease_terms  = 
          stringr::str_replace_all(l2_all_disease_terms,
                                  pattern = paste0(chronic_liver_disease_terms, collapse = "|"),
                                   "cirrhosis and other chronic liver diseases"
                          )  
        )

gwas_study_info = gwas_study_info |> 
 mutate(l2_all_disease_terms  = 
          stringr::str_replace_all(l2_all_disease_terms,
                                  pattern = "liver disease",
                                   "cirrhosis and other chronic liver diseases"
                          )  
        )

```


## Cocaine use disorders

```{r cocaine_use_terms}

gwas_study_info = gwas_study_info |> 
 mutate(l2_all_disease_terms  = 
          stringr::str_replace_all(l2_all_disease_terms ,
                                  pattern = "cocaine-related disorders",
                                   "cocaine use disorders"
                          )  
        )

```


## Depressive disorders

```{r depressive_disorders}

gwas_study_info = gwas_study_info |> 
 mutate(l2_all_disease_terms  = 
          stringr::str_replace_all(l2_all_disease_terms ,
                                  pattern = "depressive symptom measurement|major depressive disorder",
                                   "depressive disorders"
                          )  
        ) |>
   mutate(l2_all_disease_terms  = 
          stringr::str_replace_all(l2_all_disease_terms ,
                                  pattern = "depressive disorder",
                                   "depressive disorders"
                          )  
        )

```

## Gallbladder and biliary diseases

```{r}

gal_bile_terms = c("gallbladder disease",
                   "bile duct disorder",
                   "biliary tract disease")


gwas_study_info = gwas_study_info |> 
 mutate(l2_all_disease_terms  = 
          stringr::str_replace_all(l2_all_disease_terms,
                                  pattern = paste0(gal_bile_terms, collapse = "|"),
                                   "gallbladder and biliary diseases"
                          )  
        )



```

## Gynecological diseases

```{r}


diseases <- stringr::str_split(pattern = ", ", 
                               gwas_study_info$l2_all_disease_terms)  |> 
            unlist() |>
            stringr::str_trim()


pregnancy_terms <- grep("pregnancy", diseases, value = T)

gyno_terms <- c("endometriosis","placenta disease", pregnancy_terms)


gwas_study_info = gwas_study_info |> 
 mutate(l2_all_disease_terms  = 
          stringr::str_replace_all(l2_all_disease_terms,
                                  pattern = paste0(gyno_terms, collapse = "|"),
                                   "gynecological diseases"
                          )  
        )


```

## Eating disorders 

```{r}


gwas_study_info = gwas_study_info |> 
 mutate(l2_all_disease_terms  = 
          stringr::str_replace_all(l2_all_disease_terms ,
                                  pattern = "bulimia nervosa|anorexia nervosa|binge eating|eating disorder",
                                  "eating disorders"
                          )  
        ) |>
   mutate(l2_all_disease_terms  = 
          stringr::str_replace_all(l2_all_disease_terms ,
                                  pattern = "anorexia",
                                  "eating disorders"
                          )  
        )

```

## Headache disorders

```{r headache}

gwas_study_info = gwas_study_info |> 
 mutate(l2_all_disease_terms  = 
          stringr::str_replace_all(l2_all_disease_terms,
                                  pattern = "headache disorder|cluster headache|migraine",
                                   "headache disorders"
                          )  
        )


```

## Ischemic heart disease

== coronary artery disease (https://www.ncbi.nlm.nih.gov/books/NBK209964/)

```{r}

gwas_study_info = gwas_study_info |> 
 mutate(l2_all_disease_terms  = 
          stringr::str_replace_all(l2_all_disease_terms,
                                  pattern = "coronary artery disease",
                                   "ischemic heart disease"
                          )  
        )


```

## Opioid use disorders

```{r opioid_use_terms}


gwas_study_info = 
gwas_study_info |> 
 mutate(l2_all_disease_terms  = 
          stringr::str_replace_all(l2_all_disease_terms ,
                                  pattern = "opioid dependence|opioid use disorder",
                                   "opioid use disorders"
                          )  
        )


```

## Parkinsons disease

```{r pd}

gwas_study_info = 
gwas_study_info |> 
 mutate(l2_all_disease_terms  = 
          stringr::str_replace_all(l2_all_disease_terms ,
                                  pattern = "parkinsons disease",
                                   "parkinson's disease"
                          )  
        )


```

## Personality disorders

```{r}

url <- "http://www.ebi.ac.uk/ols4/api/ontologies/mondo/terms/http%253A%252F%252Fpurl.obolibrary.org%252Fobo%252FMONDO_0002028/descendants"

personality_disorders <- get_descendants(url)

gwas_study_info = gwas_study_info |> 
 mutate(l2_all_disease_terms  = 
          stringr::str_replace_all(l2_all_disease_terms,
                                  pattern = paste0(personality_disorders, collapse = "|"),
                                   "personality disorders"
                          )  
        )


```

## Other drug use disorders

```{r}


gwas_study_info = 
gwas_study_info |> 
 mutate(l2_all_disease_terms  = 
          stringr::str_replace_all(l2_all_disease_terms ,
                                  pattern = "heroin dependence|drug dependence|nictone dependence|substance abuse|drug misuse|alcohol use disorders delirium",
                                   "other drug use disorders"
                          )  
        )


```

## Sleep disorders

```{r sleep disorders}

url <- "http://www.ebi.ac.uk/ols4/api/ontologies/doid/terms/http%253A%252F%252Fpurl.obolibrary.org%252Fobo%252FDOID_535/descendants"

sleep_disorders <- get_descendants(url)

url <- "http://www.ebi.ac.uk/ols4/api/ontologies/efo/terms/http%253A%252F%252Fwww.ebi.ac.uk%252Fefo%252FEFO_0008568/descendants"

other_sleep_disorders <- get_descendants(url)

sleep_disorders <- c(sleep_disorders,
                    other_sleep_disorders)

gwas_study_info = gwas_study_info |> 
 mutate(l2_all_disease_terms  = 
          stringr::str_replace_all(l2_all_disease_terms,
                                  pattern = paste0(sleep_disorders, collapse = "|"),
                                   "sleep disorders"
                          )  
        )


```

# Other neurodegenerative disorder (to do)

take url: "http://www.ebi.ac.uk/ols4/api/ontologies/snomed/terms/http%253A%252F%252Fsnomed.info%252Fid%252F52522001/descendants"

# Malignant skin melanoma

and remove alz, parks, dementia 

# Other mental disorders

```{r}

other_mental_disorders <- c("schizophrenia",
                            "manic or hypomanic episode",
                            "mental or behavioural disorder",
                            "mental disorder"
)

```

# Other neurological disorders

```{r}

other_neuro <- c("mild neurocognitive disorder",
                 "hiv-associated neurocognitive disorder")

```

# Other sensory 

disturbances of sensation of smell and taste

# Check compatibility with gbd data 

## Weird fix 

```{r}

gwas_study_info = 
gwas_study_info |> 
 mutate(l2_all_disease_terms  = 
          stringr::str_replace_all(l2_all_disease_terms ,
                                  pattern = "disorderss",
                                   "disorders"
                          )  
        )


gwas_study_info = 
gwas_study_info |> 
 mutate(l2_all_disease_terms  = 
          stringr::str_replace_all(l2_all_disease_terms ,
                                  pattern = "anxiety disorders disorderss",
                                   "anxiety disorders"
                          )  
        )

```

```{r}

gbd_data <- data.table::fread(here::here("data/gbd/IHME-GBD_2021_DATA-aa22a7fd-1.csv"))

gbd_data$cause <- stringr::str_remove_all(gbd_data$cause, ",")

diseases <- stringr::str_split(pattern = ", ", 
                               gwas_study_info$l2_all_disease_terms[gwas_study_info$l2_all_disease_terms != ""])  |> 
            unlist() |>
            stringr::str_trim()

gbd_data$cause[!tolower(gbd_data$cause) %in% unique(diseases)] |> sort()

```

```{r}

gbd_data =
  gbd_data |>
  mutate(cause = tolower(cause))

gwas_disease_traits = data.frame(cause = diseases)
  # gwas_study_info |>
  # filter(DISEASE_STUDY == T) |>
  # select(all_disease_terms, l1_all_disease_terms, cause = l2_all_disease_terms) |>
  # distinct()

left_join(gwas_disease_traits, 
          gbd_data) |>
  head()

gwas_study_info |> select(cause = l2_all_disease_terms) |>
  distinct() |>
  left_join(gbd_data) |>
  head()

```

```{r}

diseases <- stringr::str_split(pattern = ", ", 
                               gwas_study_info$l2_all_disease_terms[gwas_study_info$l2_all_disease_terms != ""])  |> 
            unlist() |>
            stringr::str_trim()

length(unique(diseases))

# make frequency table
freq <- table(as.factor(diseases))

# sort in decreasing order
freq_sorted <- sort(freq, decreasing = TRUE)

# show top N, e.g. top 10
head(freq_sorted, 10)

```

### Save the updated gwas_study_info with harmonized disease terms

```{r}

gwas_study_info <- fwrite(gwas_study_info,
                          here::here("output/gwas_cat/gwas_study_info_trait_group_l2.csv"))

```

