---
title: "Mapping GWAS traits to ICD 10"
author: "Isobel Beasley"
date: "2025-09-26"
output: html_document
---

# Set up

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load_packages, message=FALSE}

library(dplyr)
library(stringr)
library(data.table)

```

## Ontology help - for getting disease subtypes

```{r ontology_fns}

source(here::here("code/get_term_descendants.R"))

```

### Load Data

```{r load_data}

# gwas_study_info <- fread(here::here("output/gwas_cat/gwas_study_info_group_v2.csv"))

gwas_study_info <- data.table::fread(here::here("output/gwas_cat/gwas_study_all_group.csv"))

gwas_study_info = gwas_study_info |>
    mutate(collected_all_disease_terms = stringi::stri_trans_general(collected_all_disease_terms, "Latin-ASCII")
           )

# study accession = GCST003209
# replace benign neoplasm with colorectal cancer, endometrial cancer

gwas_study_info = 
  gwas_study_info |>
  mutate(collected_all_disease_terms = ifelse(STUDY_ACCESSION == "GCST003209" |
                                              STUDY_ACCESSION == "GCST003208" ,
                                              "colorectal cancer, endometrial cancer",
                                              collected_all_disease_terms))

# for study accession = GCST90133383
# replace benign neoplasm with testicular cancer, hearing loss

```

# Automatic mapping of GWAS traits to ICD 10

## Create a mapping table

```{r create_mapping_table}

disease_mapping <- gwas_study_info |>
  filter(DISEASE_STUDY == T) |>
  tidyr::separate_longer_delim(cols = collected_all_disease_terms, 
                               delim = ", ") |>
  select(`DISEASE/TRAIT`, 
         collected_all_disease_terms, 
         PUBMED_ID,
         STUDY_ACCESSION) |>
  distinct()


disease_mapping =
  disease_mapping |>
  filter(collected_all_disease_terms != "")

print("Number of unique disease trait & study pairs")
nrow(disease_mapping)


diseases <- stringr::str_split(pattern = ", ",
                               gwas_study_info$collected_all_disease_terms[gwas_study_info$collected_all_disease_terms != ""])  |>
  unlist() |>
  stringr::str_trim()

diseases <- unique(diseases)

print("Number of unique disease terms")
print(length(diseases))

```

## Get ICD10 codes from author provided DISEASE/TRAIT column 

```{r add_icd10_from_trait}

disease_mapping =
  disease_mapping |>
  mutate(icd10_code = ifelse(grepl("ICD10", `DISEASE/TRAIT`),
                             str_extract(`DISEASE/TRAIT`, 
                                         "(?<=ICD10 )[^:]+(?=:)|(?<=ICD10 )[^ ]+"),
                             NA),
         icd10_code_origin = ifelse(grepl("ICD10", `DISEASE/TRAIT`),
                                    "Study Provided",
                                    NA)
)

# fix weird ICD10 codes: R11.103.11 -> R11
disease_mapping =
  disease_mapping |>
  mutate(icd10_code = stringr::str_replace(icd10_code, 
                                         pattern = "R11.103.11",
                                         replacement = "R11")
         )

# fix weird ICD10 codes: D63.165.8 -> D63, N18.3-N18.9
disease_mapping =
  disease_mapping |>
  mutate(icd10_code = stringr::str_replace(icd10_code, 
                                         pattern = "D63.165.8",
                                         replacement = "D63, N18.3-N18.9")
         ) 

# Additionally, some studies provide ICD10 codes in the DISEASE/TRAIT column without the "ICD10" prefix
# e.g. Source of report of K76 (other diseases of liver) (UKB data field 131671)
# let's try to capture these too
disease_mapping =
  disease_mapping |>
  mutate(icd10_code = ifelse(is.na(icd10_code) &
                             grepl("Source of report of", `DISEASE/TRAIT`),
                             str_extract(`DISEASE/TRAIT`,
                                         "(?<=Source of report of\\s)[A-Z][0-9]{1,2}(?:\\.[0-9A-Z]+)?\\b"),
                             icd10_code),
         icd10_code_origin = ifelse(grepl("Source of report of", 
                                          `DISEASE/TRAIT`),
                                    "Study Provided",
                                    icd10_code_origin)
) |>
  mutate(icd10_code = str_remove_all(icd10_code, 
                                     pattern = "Source of report of ")
         )

# Another example where  some studies provide ICD10 codes in the DISEASE/TRAIT column without the "ICD10" prefix
# e.g. Insulin-dependent diabetes mellitus (Union E10)
disease_mapping =
  disease_mapping |>
  mutate(icd10_code = ifelse(is.na(icd10_code) &
                             grepl("Union", `DISEASE/TRAIT`),
                             str_extract(`DISEASE/TRAIT`, 
                                         "Union [A-Z][0-9]{1,2}(?:\\.[0-9A-Z]+)?\\b"),
                             icd10_code),
         icd10_code_origin = ifelse(grepl("Union", 
                                          `DISEASE/TRAIT`),
                                    "Study Provided",
                                    icd10_code_origin)
) |>
  mutate(icd10_code = str_remove_all(icd10_code, 
                                     pattern = "Union ")
  )

disease_mapping =
  disease_mapping |>
  mutate(icd10_code = ifelse(is.na(icd10_code) &
                             grepl("ICD [A-Z][0-9]|ICD-10 [A-Z][0-9]", 
                                   `DISEASE/TRAIT`),
                        str_match(`DISEASE/TRAIT`,
                                  "ICD(?:-10|10)?\\s*([A-Z][0-9]{1,2}(?:\\.[0-9A-Z]+)?)")[,2],
                             icd10_code),
         icd10_code_origin = ifelse(grepl("ICD [A-Z][0-9]|ICD-10 [A-Z][0-9]",
                                          `DISEASE/TRAIT`),
                                    "Study Provided",
                                    icd10_code_origin)
)

n_study_provided = 
disease_mapping |>
  filter(icd10_code_origin == "Study Provided") |>
  nrow()

n_studies =
  disease_mapping |>
  nrow()

print("Percentage of studies that provide ICD10 codes:")
print(round(n_study_provided / n_studies * 100, 2))

```

## Get ICD10 codes from author provided PheCodes 

### Get Phecodes for diseases

```{r get_phecodes}

disease_mapping <- disease_mapping |>
  mutate(
    phecode = str_extract(`DISEASE/TRAIT`, "(?<=PheCode )[^)]+")
  ) |>
  mutate(phecode = as.numeric(phecode))

```

### Convert Phecodes to ICD10

```{r convert_phecodes_to_icd10, message=FALSE}

# phecode to ICD10 mapping from https://wei-lab.app.vumc.org/phecode-data/phecode_international_version

phecodes <- fread(here::here("data/icd/phecode_international_version_unrolled.csv"))

phecode_icd_map =
  phecodes |>
  select(icd10_code = ICD10, 
         phecode = PheCode
         ) |>
  filter(!is.na(phecode))

phecode_icd_map =
  phecode_icd_map |>
  filter(phecode %in% unique(disease_mapping$phecode))

# if more than one ICD10 code per phecode, collapse into a single row
phecode_icd_map =
  phecode_icd_map |>
group_by(phecode) |>
  summarise(icd10_code = 
            str_flatten(unique(icd10_code), 
                        collapse = ", ", 
                        na.rm = T), 
            .groups = "drop")

# label the origin of these ICD10 codes as "Study PheCode Mapping"
phecode_icd_map =
  phecode_icd_map |>
  mutate(icd10_code_origin = "Study PheCode Mapping")

disease_mapping =
  rows_patch(disease_mapping,
              phecode_icd_map,
              by = c("phecode"),
              unmatched = "ignore",
              )
# left_join(disease_mapping,
#           phecode_icd_map,
#           by = c("phecode","icd10_code_origin", "icd10_code"),
#           relationship = "many-to-one",
#           na_matches = "never")

# disease_mapping =
#   disease_mapping |>
#   mutate(icd10_code_origin = "Study PheCode")

print("Number of ICD10 code obtained")
disease_mapping |>
    filter(!is.na(icd10_code)) |> 
  group_by(icd10_code_origin) |> 
  summarise(n = n(), 
            percent = n()/n_studies *100)

disease_mapping |>
  filter(!is.na(icd10_code_origin)) |>
  nrow()

disease_mapping |>
  filter(!is.na(icd10_code_origin)) |>
  head()

# per pubmed id, are there multiple icd10 code origins?
disease_mapping |>
  filter(!is.na(icd10_code_origin)) |>
  group_by(PUBMED_ID) |>
  summarise(n_icd10_code_origins = n_distinct(icd10_code_origin)) |>
  pull(n_icd10_code_origins) |>
  summary()

# which pubmed ids have multiple icd10 code origins?
disease_mapping |>
  filter(!is.na(icd10_code_origin)) |>
  group_by(PUBMED_ID) |>
  summarise(n_icd10_code_origins = n_distinct(icd10_code_origin)) |>
  filter(n_icd10_code_origins > 1)

# per study accession, are there multiple icd10 code origins?
disease_mapping |>
  filter(!is.na(icd10_code_origin)) |>
  group_by(STUDY_ACCESSION) |>
  summarise(n_icd10_code_origins = n_distinct(icd10_code_origin)) |>
  pull(n_icd10_code_origins) |>
  summary()

```

### Complete PheCode to ICD10 mappings

Some Phecodes do not have ICD10 mappings in the above Phecode to ICD10 mapping file.
Here we've created a file with manual mappings for these missing Phecodes.

```{r}

# how many missing ICD10 codes for PheCodes?
disease_mapping |>
    filter(is.na(icd10_code) & !is.na(phecode)) |> 
    nrow()

# load manual mapping file
phecode_icd_manual = 
  readxl::read_xlsx(here::here("data/icd/phecode_to_icd10_manual_mapping.xlsx"))

to_add = phecode_icd_manual |>
  mutate(phecode = as.numeric(phecode),
         icd10_code_origin = "Study PheCode (Manual Mapping)") |>
  select(-phenotype) |>
  distinct()

disease_mapping =
  rows_patch(disease_mapping,
             to_add,
             unmatched = "ignore")
  
# 
# nrow(phecode_icd_manual)
# 
# phecode = c("38.30",
#             "79.90",
#             "110.00",
#             "530.13",
#             "562.20",
#             "580.10",
#             "174.00",
#             "174.20",
#             "189.10",
#             "218.00",
#             "228.10",
#             "224.00",
#             "250.15",
#             "250.25",
#             "250.40",
#             "285.21",
#             "292.11",
#             "362.26",
#             "362.23",
#             "362.27",
#             "362.50",
#             "452.20",
#             "528.10",
#             "535.90",
#             "724.10",
#             "724.22",
#             "740.00",
#             "743.10",
#             "743.12",
#             "743.13",
#             "172.21",
#             "274.00",
#             "286.00",
#             "282.00",
#             "280.00",
#             "272.00",
#             "276.00",
#             "338.00",
#             "350.00",
#             "401.00",
#             "411.00",
#             "414.20",
#             "415.1",
#             "427.21",
#             "429.00",
#             "585.00",
#             "735.22",
#             "724.00",
#             "722.00",
#             "716.10",
#             "709.00",
#             "706.00",
#             "592.00",
#             "571.00",
#             "580.00",
#             "170.00",
#             "172.00",
#             "264.00",
#             "291.00",
#             "555.00",
#             "562.00",
#             "578.00",
#             "783.1",
#             "536.80",
#             "427.40",
#             "443.00",
#             "41.80",
#             "41.90",
#             "244.00",
#             "250.00",
#             "427.40",
#             "526.40",
#             "977.00",
#             "840.20",
#             "823.00",
#             "751.00",
#             "743.00",
#             "270.38",
#             "504.10",
#             "253.40",
#             "279.20",
#             "426.22",
#             "537.1",
#             "707.20",
#             "736.10",
#             "789.10",
#             "290.13",
#             "327.70",
#             "433.60",
#             "695.00",
#             "602.30",
#             "375.10",
#             "560.00",
#             "586.10",
#             "593.20",
#             "620.10",
#             "475.90",
#             "799.00",
#             "662.10",
#             "855.00",
#             "536.70",
#             "856.00",
#             "519.10",
#             "771.20",
#             "792.10",
#             "525.10",
#             "724.20",
#             "611.11",
#             "556.11",
#             "767.00",
#             "736.50",
#             "526.42")
# 
# icd10_code = c("A49.9",
#                "B34.9",
#                "B35, B36",
#                "K22.7",
#                "K57",
#                "N00, N01, N02, N03, N04, N05, N06, N07",
#                "Z85.3",
#                "C50",
#                "C64, C65",
#                "D25, D26",
#                "D18.0",
#                "E00, E00.0, E00.1, E00.2, E00.9, E01.8, E02, E03.0, E03.1, E03.2, E03.3, E03.8, E03.9, E89.0",
#                "E10.5",
#                "E11.5",
#                "R73.0",
#                "D63",
#                "R47.0",
#                "H35.3",
#                "H35.3",
#                "H35.3",
#                "H35.5",
#                "I80",
#                "K12.30",
#                "K29.7, K29.8, K29.9",
#                "M43.2",
#                "M21.5",
#                "M13.9, M15.0, M15.1, M15.2, M15.3, M15.4, M16, M16.0, M16.1, M16.3, M16.6, M16.7, M16.9, M17.1, M17.4, M17.5, M18.0, M18.1, M18.5, M18.9, M19.0, M19.2",
#                "M81",
#                "M81.8",
#                "M81.8",
#                "C44",
#                "M10, M10.0, M10.1, M10.2, M10.4, M10.9, M11.0, M11.1, M11.2, M11.8, M11.9, M67.9",
#                "D65, D66, D67, D68, D68, D68.0, D68.1, D68.2, D68.3, D68.4, D68.8, D68.9, O72.3, O99.1",
#                "D55, D55.0, D55.1, D55.2, D55.3, D55.8, D55.9, D56, D56.0, D56.1, D56.2, D56.3, D56.4, D56.8, D56.9, D57, D57.0, D57.1, D57.2, D57.3, D57.8, D58, D58.0, D58.1, D58.2, D58.8, D58.9, M90.4",
#                "D50, D50.0, D50.1, D50.8, D50.9",
#                "E78.0, E78.1, E78.2, E78.3, E78.4, E78.5, E78.9",
#                "E86, E87.0, E87.1, E87.2, E87.3, E87.4, E87.5, E87.6, E87.7, E87.8, R63.1",
#                "R52.0, R52.2, R52.9",
#                "R25, R25.0, R25.1, R25.2, R25.3, R25.8, R26, R26.0, R26.1, R26.8, R27, R27.0, R27.8, R29.0, R29.2, R43, R43.0, R43.1, R43.2",
#                "I10, I11, I11.0, I11.9, I12, I12.0, I12.9, I13, I13.0, I13.1, I13.2, I13.9, I15, I15.0, I15.1, I15.2, I15.8, I15.9, I67.4",
#                "I20, I20.0, I20.1, I20.8, I20.9, I21, I21.0, I21.1, I21.2, I21.3, I21.4, I21.9, I22, I22.0, I22.1, I22.8, I22.9, I23, I23.0, I23.1, I23.2, I23.3, I23.6, I23.8, I24, I24.0, I24.1, I24.8, I24.9, I25, I25.1, I25.2, I25.3, I25.4, I25.5, I25.6, I25.8, I25.9, I34.1, I51.0, I51.3, Z95.1, Z95.5",
#                "I25.10",
#                "I26, I26.0",
#                "I48",
#                "I51.8",
#                "N17, N17.0, N17.1, N17.2, N17.8, N17.9, N18, N18.0, N18.9, N19, Y60.2, Y61.2, Y62.0, Y84.1, Z49.1, Z49.2, Z99.2",
#                "M21.5",
#                "M40.2, M43.2, M43.8, M48.8, M49.8, M50.0, M99.6",
#                "G55.1, M46.4, M50, M50.0, M50.0, M50.1, M50.2, M50.3, M50.8, M50.9, M51.3, M51.4, M96.1",
#                "M13.0",
#                "L94.3, M33, M33.0, M33.1, M33.2, M33.9, M34, M34.0, M34.1, M34.2, M34.8, M34.9, M35.0, M35.1, M35.5, M35.8, M35.9, M36.0, M36.8, M65.3, N16.4",
#                "K09.8, L70, L70.0, L70.1, L70.2, L70.3, L70.4, L70.5, L70.8, L70.9, L72, L72.0, L72.1, L72.2, L72.8, L72.9, L73.0, L85.3",
#                "N30, N30.0, N30.1, N30.2, N30.3, N30.8, N30.9, N34, N34.0, N34.2, N34.3, N35.1, N37",
#                 "K70.4, K72, K72.1, K72.9, K74.0, K74.1, K74.2, K74.3, K74.4, K74.5, K74.6, K75.0, K75.1, K76.0, K76.6, K76.7",
#                "B52.0, N00.0, N00.1, N00.2, N00.3, N00.4, N00.5, N00.6, N00.7, N01, N01.0, N01.1, N01.2, N01.3, N01.4, N01.5, N01.6, N01.7, N01.9, N02.0, N02.1, N02.2, N02.3, N02.4, N02.5, N02.6, N02.7, N03, N03.0, N03.1, N03.2, N03.3, N03.4, N03.5, N03.6, N03.7, N03.9, N04, N04.0, N04.1, N04.2, N04.3, N04.4, N04.5, N04.6, N04.7, N04.8, N04.9, N05, N05.0, N05.1, N05.2, N05.3, N05.4, N05.5, N05.6, N05.7, N05.9, N06.0, N06.1, N06.2, N06.3, N06.4, N06.5, N06.6, N06.7, N07.0, N07.1, N07.2, N07.3, N07.4, N07.5, N07.6, N07.7, N08, N08.1, N08.2, N08.3, N08.4, N08.5, N08.8, N14, N14.0, N14.1, N14.2, N14.3, N14.4, N15.0, N15.8, N16.1, N16.2, N16.3, N16.4, N16.5",
#                "C40, C40.0, C40.1, C40.2, C40.3, C40.8, C40.9, C41, C41.0, C41.1, C41.2, C41.3, C41.4, C41.9, C47, C47.0, C47.1, C47.2, C47.3, C47.4, C47.5, C47.6, C47.8, C47.9, C49, C49.0, C49.1, C49.2, C49.3, C49.4, C49.5, C49.6, C49.8, C49.9",
# "C43, C43.0, C43.1, C43.2, C43.3, C43.4, C43.5, C43.6, C43.7, C43.8, C43.9, C44.0, C44.1, C44.2, C44.3, C44.4, C44.5, C44.6, C44.7, C44.8, C44.9, D03, D03.0, D03.1, D03.2, D03.3, D03.4, D03.5, D03.6, D03.7, D03.8, D03.9, D04, D04.0, D04.1, D04.2, D04.3, D04.4, D04.5, D04.6, D04.7, D04.8, D04.9",
# "R62.0, R62.8, R62.9",
# "F06, F06.1, F07.0, F07.1, F07.2, F07.8, F07.9, F23, F23.0, F23.1, F23.8, F23.9, G47.1, R40.0, R40.1",
# "K50, K50.0, K50.1, K50.8, K50.9, K51, K51.0, K51.1, K51.2, K51.3, K51.4, K51.5, K51.8, K51.9",
# "K57, K57.0, K57.1, K57.2, K57.3, K57.4, K57.5, K57.8, K57.9",
# "K62.5, K92.0, K92.1, K92.2",
# "R50.8",
# "K30",
# "I46, I46.0, I46.9, I49.0",
# "E10.5, E11.5, E14.5, I73, I73.0, I73.8, I73.9, I79.1, I79.2, I79.8",
# "B96.8",
# "U82, U83, U84",
# "E00, E00.0, E00.1, E00.2, E00.9, E01.8, E02, E03.0, E03.1, E03.2, E03.3, E03.8, E03.9, E89.0",
# "E10, E10.0, E10.1, E10.2, E10.3, E10.3, E10.3, E10.4, E10.4, E10.6, E10.7, E10.8, E10.9, E11, E11.0, E11.1, E11.2, E11.3, E11.4, E11.6, E11.7, E11.8, E11.9, E12.3, E13, E13.1, E13.3, E13.4, E13.5, E13.6, E13.7, E13.8, E13.9, E14.9, G59.0, G63.2, H36.0, R73.0, R73.9, R81, R82.4, Z96.4",
# "I46, I46.0, I46.9, I49.0",
# "K07.6",
# "Z88.9",
# "S43.4",
# "S82.1, S82.3, S82.8",
#  "Q50.0, Q50.1, Q50.2, Q50.3, Q50.4, Q50.5, Q50.6, Q51, Q51.0, Q51.1, Q51.2, Q51.3, Q51.4, Q51.5, Q51.6, Q51.7, Q51.8, Q51.9, Q52.0, Q52.1, Q52.2, Q52.3, Q52.4, Q52.5, Q52.6, Q52.7, Q52.8, Q52.9, Q53, Q53.0, Q53.1, Q53.2, Q53.9, Q54, Q54.0, Q54.1, Q54.2, Q54.3, Q54.4, Q54.8, Q54.9, Q55, Q55.0, Q55.1, Q55.2, Q55.3, Q55.4, Q55.5, Q55.6, Q55.8, Q55.9, Q56, Q56.0, Q56.1, Q56.2, Q56.3, Q56.4, Q60, Q60.0, Q60.1, Q60.2, Q60.3, Q60.4, Q60.5, Q60.6, Q61, Q61.0, Q61.1, Q61.2, Q61.3, Q61.4, Q61.5, Q61.8, Q61.9, Q62, Q62.0, Q62.1, Q62.3, Q62.4, Q62.5, Q62.6, Q62.7, Q62.8, Q63, Q63.0, Q63.1, Q63.2, Q63.3, Q63.8, Q63.9, Q64.0, Q64.1, Q64.2, Q64.3, Q64.4, Q64.5, Q64.6, Q64.7, Q64.8, Q64.9",
# "M48.4, M48.5, M80.5, M80.8, M81.6, M81.9, M84.4, M85.9, M89.9",
# "E88.0",
# "J84.1, J84.2",
# "E23.6",
# "D89.8",
# "I44.1",
# "K31.8, K31.9",
# "L97",
# "M21.0, M21.1, M21.9",
# "R11.10",
# "F03",
# "G47.6, G25.8",
# "G43.6",
# "L49",
# "N42.3",
# "H04.1",
# "K56.6",
# "N28.8",
# "R31.2",
# "N87",
# "R09.8",
# "R53",
# "T38.0, T50.0",
# "T85.1, T85.8",
# "K91.4, K91.8, Y83.3",
# "I97.8",
# "J95.0",
# "R25.2",
# "R87.6",
# "K08.1",
# "M53.2, M53.3",
# "R92",
# "K55.2",
# "M53.0, M53.1",
# "M21.1, M21.8",
# "K07.6")
# 
# writexl::write_xlsx(data.frame(phecode, icd10_code) |> arrange(phecode),
#                     here::here("data/icd/phecode_to_icd10_manual_mapping.xlsx")
#                     )
# 
# icd10_code_origin = rep("Study PheCode (Manual Mapping)", 
#                         length(phecode))
# 
# to_add = data.frame(phecode, 
#                     icd10_code, 
#                     icd10_code_origin)
# 
# to_add = to_add |> distinct()
# 
# to_add = 
#   to_add |>
#   mutate(phecode = as.numeric(phecode))
# 
# disease_mapping =
#   rows_patch(disease_mapping,
#              to_add,
#              unmatched = "ignore")



```

### How many diseases are not mapped yet?

```{r how_many_not_mapped_phecodes_icd}

print("Number of ICD10 codes obtained")
disease_mapping |>
    filter(!is.na(icd10_code)) |> 
  group_by(icd10_code_origin) |> 
  summarise(n = n(), 
            percent = n()/n_studies *100)
# we were able to get PheCodes or ICD10 codes directly for roughly 25% of studies 

disease_mapping_matched =
  disease_mapping |>
  filter(icd10_code != "" | !is.na(icd10_code))

not_found_diseases <- diseases[!diseases %in% disease_mapping_matched$collected_all_disease_terms] 
not_found_diseases <- not_found_diseases[not_found_diseases != ""]

print("Number of diseases not mapped to a single ICD10 code yet:")
print(length(not_found_diseases))

```

## Get ICD10 codes by matching `DISEASE/TRAIT` terms 

### Tidy `DISEASE/TRAIT` Column to better match terms 

#### Removing model related terms 

```{r remove_model_terms, message=FALSE}

# removing '(Gene-based burden)' in `DISEASE/TRAIT` column
disease_mapping =
  disease_mapping |>
  mutate(`DISEASE/TRAIT` = str_remove_all(`DISEASE/TRAIT`, 
                                          "\\s*\\(Gene-based burden\\)"))

# removing '(1df test)'
disease_mapping =
  disease_mapping |>
  mutate(`DISEASE/TRAIT` = str_remove_all(`DISEASE/TRAIT`, 
                                          "\\s*\\(1df test\\)|\\s*\\(1df\\)"))

# removing `(2df test)`
disease_mapping =
  disease_mapping |>
  mutate(`DISEASE/TRAIT` = str_remove_all(`DISEASE/TRAIT`, 
                                          "\\s*\\(2df test\\)||\\s*\\(2df\\)"))

# removing '(3df test)'
disease_mapping =
  disease_mapping |>
  mutate(`DISEASE/TRAIT` = str_remove_all(`DISEASE/TRAIT`, 
                                          "\\s*\\(3df test\\)|\\s*\\(3df\\)"))

# removing '(multivariate analysis)'
disease_mapping =
  disease_mapping |>
  mutate(`DISEASE/TRAIT` = str_remove_all(`DISEASE/TRAIT`, 
                                          "\\s*\\(multivariate analysis\\)"))

# removing '(ordinary least squares (OLS))'
disease_mapping =
  disease_mapping |>
  mutate(`DISEASE/TRAIT` = str_remove_all(`DISEASE/TRAIT`, 
                                          "\\s*\\(ordinary least squares \\(OLS\\)\\)"))

# remove '(CNV mirror model)'
disease_mapping =
  disease_mapping |>
  mutate(`DISEASE/TRAIT` = str_remove_all(`DISEASE/TRAIT`, 
                                          "\\s*\\(CNV mirror model\\)"))

# remove '(CNV U-shape model)'
disease_mapping =
  disease_mapping |>
  mutate(`DISEASE/TRAIT` = str_remove_all(`DISEASE/TRAIT`, 
                                          "\\s*\\(CNV U-shape model\\)"))

# removing '(MTAG)'
disease_mapping =
  disease_mapping |>
  mutate(`DISEASE/TRAIT` = str_remove_all(`DISEASE/TRAIT`, "\\s*\\(MTAG\\)"))

# removing '(pleiotropy)'
disease_mapping =
  disease_mapping |>
  mutate(`DISEASE/TRAIT` = str_remove_all(`DISEASE/TRAIT`, "\\s*\\(pleiotropy\\)"))

# removing '(multi-trait analysis)'


# remove 'SNP x SNP interaction'
disease_mapping =
  disease_mapping |>
  mutate(`DISEASE/TRAIT` = str_remove_all(`DISEASE/TRAIT`, 
                                          "\\s*\\(SNP x SNP interaction\\)"))


```

#### Removing study related phenotype IDs

```{r tidy_disease_terms_for_matching, message=FALSE}

# removing references to UKB data fields in `DISEASE/TRAIT` column
disease_mapping =
  disease_mapping |>
  mutate(`DISEASE/TRAIT` = str_remove_all(`DISEASE/TRAIT`, 
                                          "\\s*\\(UKB data field \\d+\\)")) |>
  mutate(`DISEASE/TRAIT` = str_remove_all(`DISEASE/TRAIT`, 
                                          "\\s*\\(UKB data field \\d+_\\d+\\)"))


# removing Phecode references (e.g. (Phecode 250.2))
disease_mapping =
  disease_mapping |>
  mutate(`DISEASE/TRAIT` = str_remove_all(`DISEASE/TRAIT`, "\\s*\\(Phe[Cc]ode [0-9]+\\.[0-9]+\\)")) |>
  mutate(`DISEASE/TRAIT` = str_remove_all(`DISEASE/TRAIT`, "\\s*\\(Phe[Cc]ode [0-9]+\\)"))

# remove any substring of the form (Number_Number) — e.g. (13013_41)
disease_mapping =
  disease_mapping |>
  mutate(`DISEASE/TRAIT` = 
           str_remove_all(`DISEASE/TRAIT`, 
                          "\\s*\\(\\d+_\\d+\\)")
         )


```


#### Remove patient genotype effects

```{r}

# removing genotype effect (e.g. (fetal genotype effect))
disease_mapping =
  disease_mapping |>
  mutate(`DISEASE/TRAIT` = 
           str_remove_all(`DISEASE/TRAIT`, 
                          "\\s*\\([^)]*genotype effect\\)")
         ) |>
    mutate(`DISEASE/TRAIT` = 
           str_remove_all(`DISEASE/TRAIT`, 
                          "\\s*\\([^)]* effect\\)")
         )

# remove '(adjusted for APOE e4 dosage)'
disease_mapping =
  disease_mapping |>
  mutate(`DISEASE/TRAIT` = str_remove_all(`DISEASE/TRAIT`, 
                                          "\\(adjusted for APOE e4 dosage\\)"))

# remove '(maternal):' & '(paternal):
disease_mapping =
  disease_mapping |>
  mutate(`DISEASE/TRAIT` = str_remove_all(`DISEASE/TRAIT`, "\\s*\\(maternal\\):")) |>
  mutate(`DISEASE/TRAIT` = str_remove_all(`DISEASE/TRAIT`, "\\s*\\(paternal\\):"))


# remove 'Biological Grandparent '
disease_mapping =
  disease_mapping |>
  mutate(`DISEASE/TRAIT` = str_remove_all(`DISEASE/TRAIT`, "Biological Grandparent "))

# remove 'Biological Father: ', 'Biological Sibling: ', 'Biological Mother: '
disease_mapping =
  disease_mapping |>
  mutate(`DISEASE/TRAIT` = str_remove_all(`DISEASE/TRAIT`, "Biological Father: ")) |>
  mutate(`DISEASE/TRAIT` = str_remove_all(`DISEASE/TRAIT`, "Biological Sibling: ")) |>
  mutate(`DISEASE/TRAIT` = str_remove_all(`DISEASE/TRAIT`, "Biological Mother: "))

```


#### Removing trait specific terms 

```{r remove_trait_specific_terms, message=FALSE}
# remove  (apnea hypopnea index), 
# (average respiratory event duration), 
#  (micro-arousal index)
#  (percentage of N3 sleep time during total sleep time)
# (percentage of N3 sleep time during sleep period time)
# (average oxyhemoglobin desaturation per event)
#  (average oxyhemoglobin saturation across sleep episode)
# (percentage sleep with oxyhemoglobin saturation less than 90%)
#  (wake time during sleep period time)
# (minimum oxyhemoglobin saturation across sleep episode)
#  (oxygen desaturation index)
# (average oxygen saturation during sleep)
# (apnea hypopnea index, change over time)

disease_mapping =
  disease_mapping |>
  mutate(`DISEASE/TRAIT` = str_remove_all(`DISEASE/TRAIT`, 
                                          "\\s*\\(apnea hypopnea index\\)")) |>
  mutate(`DISEASE/TRAIT` = str_remove_all(`DISEASE/TRAIT`, 
                                          "\\s*\\(average respiratory event duration\\)")) |>
  mutate(`DISEASE/TRAIT` = str_remove_all(`DISEASE/TRAIT`,
                                          "\\s*\\(micro-arousal index\\)")) |>
  mutate(`DISEASE/TRAIT` = str_remove_all(`DISEASE/TRAIT`,
                                          "\\s*\\(percentage of N3 sleep time during total sleep time\\)")) |>
  mutate(`DISEASE/TRAIT` = str_remove_all(`DISEASE/TRAIT`,
                                          "\\s*\\(percentage of N3 sleep time during sleep period time\\)")) |>
  mutate(`DISEASE/TRAIT` = str_remove_all(`DISEASE/TRAIT`,
                                          "\\s*\\(average oxyhemoglobin desaturation per event\\)")) |>
  mutate(`DISEASE/TRAIT` = str_remove_all(`DISEASE/TRAIT`,
                                          "\\s*\\(average oxyhemoglobin saturation across sleep episode\\)")) |>
  mutate(`DISEASE/TRAIT` = str_remove_all(`DISEASE/TRAIT`,
                                          "\\s*\\(percentage sleep with oxyhemoglobin saturation less than 90%\\)")) |>
  mutate(`DISEASE/TRAIT` = str_remove_all(`DISEASE/TRAIT`,
                                          "\\s*\\(wake time during sleep period time\\)")) |>
  mutate(`DISEASE/TRAIT` = str_remove_all(`DISEASE/TRAIT`,
                                          "\\s*\\(minimum oxyhemoglobin saturation across sleep episode\\)")) |>
  mutate(`DISEASE/TRAIT` = str_remove_all(`DISEASE/TRAIT`,
                                          "\\s*\\(oxygen desaturation index\\)")) |>
  mutate(`DISEASE/TRAIT` = str_remove_all(`DISEASE/TRAIT`,
                                          "\\s*\\(average oxygen saturation during sleep\\)")) |>
  mutate(`DISEASE/TRAIT` = str_remove_all(`DISEASE/TRAIT`,
                                          "\\s*\\(apnea hypopnea index, change over time\\)"))

disease_mapping =
  disease_mapping |>
  mutate(`DISEASE/TRAIT` = str_remove_all(`DISEASE/TRAIT`, 
                                          " during REM sleep$| during non-REM sleep$"))


# ends in  levels in coronary artery disease, make it just coronary artery disease
disease_mapping =
  disease_mapping |>
  mutate(`DISEASE/TRAIT` = 
           ifelse(str_detect(`DISEASE/TRAIT`,
                             "(?i)\\s*levels in coronary artery disease$"),
                  "coronary artery disease",
                  `DISEASE/TRAIT`))

# ends in levels in chronic kidney disease, make it just chronic kidney disease
# ends in  levels in coronary artery disease, make it just coronary artery disease
disease_mapping =
  disease_mapping |>
  mutate(`DISEASE/TRAIT` = 
           ifelse(str_detect(`DISEASE/TRAIT`,
                             "(?i)\\s*levels in chronic kidney disease$"),
                  "chronic kidney disease",
                  `DISEASE/TRAIT`))

# ends in levels in type 2 diabetes, make it just type 2 diabetes
disease_mapping =
  disease_mapping |>
  mutate(`DISEASE/TRAIT` = 
           ifelse(str_detect(`DISEASE/TRAIT`,
                             "(?i)\\s*levels in type 2 diabetes$"),
                  "type 2 diabetes",
                  `DISEASE/TRAIT`))

# ends in levels in prediabetes, make it just prediabetes
disease_mapping =
  disease_mapping |>
  mutate(`DISEASE/TRAIT` = 
           ifelse(str_detect(`DISEASE/TRAIT`,
                             "(?i)\\s*levels in prediabetes$"),
                  "prediabetes",
                  `DISEASE/TRAIT`))


# remove str_extract(x, "(?<=Takes medication for )\\w+")
disease_mapping =
  disease_mapping |>
  mutate(`DISEASE/TRAIT` =
           ifelse(str_detect(`DISEASE/TRAIT`,
                             "(?i)Takes medication for \\w+"),
                  str_extract(`DISEASE/TRAIT`,
                              "(?<=Takes medication for )\\w+"),
                  `DISEASE/TRAIT`))
                                          

# remove BMI adjustments 
# '(BMI adjusted)', or '(adjusted for BMI)'
disease_mapping =
  disease_mapping |>
  mutate(`DISEASE/TRAIT` = str_remove_all(`DISEASE/TRAIT`, 
                                          "\\s*\\(BMI adjusted\\)|\\s*\\(adjusted for BMI\\)"))

# remove 'trait'
disease_mapping =
  disease_mapping |>
  mutate(`DISEASE/TRAIT` = str_remove_all(`DISEASE/TRAIT`, 
                                          "\\s*trait$"))


# remove  (slight), (severe), (generalised), (localised)
disease_mapping =
  disease_mapping |>
  mutate(`DISEASE/TRAIT` = str_remove_all(`DISEASE/TRAIT`, 
                                          "\\s*\\(slight\\)|\\s*\\(severe\\)|\\s*\\(generalised\\)|\\s*\\(localised\\)"))
```

#### Removing patient condition related terms

```{r}

# removing '(time to event)'
disease_mapping =
  disease_mapping |>
  mutate(`DISEASE/TRAIT` = str_remove_all(`DISEASE/TRAIT`, 
                                          "\\s*\\(time to event\\)"))

# removing age of onset
disease_mapping =
  disease_mapping |>
  mutate(`DISEASE/TRAIT` = str_remove_all(`DISEASE/TRAIT`, 
                                          "\\s*\\(age of onset\\)|\\s*\\(age at onset\\)")) |>
  mutate(`DISEASE/TRAIT` = str_remove_all(`DISEASE/TRAIT`, 
                                          "^Age of onset of "))


# remove visit times
disease_mapping =
  disease_mapping |>
  mutate(`DISEASE/TRAIT` =
  str_remove_all(`DISEASE/TRAIT`, "\\s*\\(\\d+ month visit\\)")
  )

# remove '(baseline)'
disease_mapping =
  disease_mapping |>
  mutate(`DISEASE/TRAIT` =
  str_remove_all(`DISEASE/TRAIT`, "\\s*\\(baseline\\)")
  )

# remove '(hospitalized vs population)', '(hospitalized vs not hospitalized)', '(severe vs population)'
disease_mapping =
  disease_mapping |>
  mutate(`DISEASE/TRAIT` =
  str_remove_all(`DISEASE/TRAIT`, "\\s*\\(hospitalized vs population\\)|\\s*\\(hospitalized vs not hospitalized\\)|\\s*\\(severe vs population\\)")
  )

# remove (RELEASE [Number])
disease_mapping = 
  disease_mapping |>
  mutate(`DISEASE/TRAIT` =
  str_remove_all(`DISEASE/TRAIT`, "\\s*\\(RELEASE \\d+\\)")
  )


```

#### Remove white space

```{r remove_whitespace, message=FALSE}

disease_mapping =
  disease_mapping |>
  mutate(`DISEASE/TRAIT` = str_squish(`DISEASE/TRAIT`))

```

### Match disease/trait to ICD 10 descriptions 

```{r match_disease_terms_to_icd_descriptions, message=FALSE}

matched_icd10_desc  = 
phecodes |>
    mutate(collected_all_disease_terms = tolower(iconv(ICD_DESCRIPTION, 
                                                       to = "UTF-8"))
           ) |>
    select(collected_all_disease_terms,    
           icd10_code = ICD10) 

# filter missing strings
matched_icd10_desc = 
  matched_icd10_desc |>
  filter(!is.na(collected_all_disease_terms) & 
         collected_all_disease_terms != "") |>
  filter(!is.na(icd10_code) & 
         icd10_code != "")


matched_icd10_desc =
  matched_icd10_desc |>
  group_by(collected_all_disease_terms) |>
  summarise(icd10_code = str_flatten(unique(icd10_code), 
                                     collapse = ", ", 
                                     na.rm = T), 
            .groups = "drop") 
  

matched_icd10_desc = 
  matched_icd10_desc |>
    mutate(icd10_code_origin = "ICD Description Match (DISEASE/TRAIT)")

# match by DISEASE/TRAIT
matched_icd10_desc =
  matched_icd10_desc |>
  rename(`DISEASE/TRAIT` = collected_all_disease_terms)

disease_mapping = 
disease_mapping |>
rows_patch(matched_icd10_desc,
           unmatched = "ignore")

print("Number of ICD10 codes obtained")
disease_mapping |>
    filter(!is.na(icd10_code)) |> 
  group_by(icd10_code_origin) |> 
  summarise(n = n(), 
            percent = n()/n_studies *100)

```

### Match disease/trait to phenotypes (corresponding to PheCodes)

```{r match_disease_terms_to_phenotypes, message=FALSE}

phenotype_icd_map =
  phecodes |>
group_by(Phenotype) |>
  summarise(icd10_code = 
            str_flatten(ICD10, 
                        collapse = ", ", 
                        na.rm = T), 
            .groups = "drop")

matched_phenotypes =
phenotype_icd_map #|>
#filter(tolower(Phenotype) %in% not_found_diseases)

matched_phenotypes =
matched_phenotypes |>
  mutate(collected_all_disease_terms = tolower(Phenotype)) |>
  select(collected_all_disease_terms, icd10_code) |>
  mutate(icd10_code_origin = "Phecode Phenotype Match (DISEASE/TRAIT)")

# match by DISEASE/TRAIT
matched_phenotypes =
  matched_phenotypes |>
  rename(`DISEASE/TRAIT` = collected_all_disease_terms)

disease_mapping =
disease_mapping |>
rows_patch(matched_phenotypes,
           unmatched = "ignore")

disease_mapping |>
  filter(icd10_code_origin == "Phecode Phenotype Match (DISEASE/TRAIT)") |>
  nrow()

disease_mapping |>
    filter(!is.na(icd10_code)) |> 
  group_by(icd10_code_origin) |> 
  summarise(n = n(), 
            percent = n()/n_studies *100)
         
```

### How many diseases are not mapped yet?

```{r how_many_not_mapped_matched_phenotypes}

matched =
  disease_mapping |>
  filter(icd10_code != "") |>
  pull(collected_all_disease_terms)

not_found_diseases <- diseases[!diseases %in% 
                               matched
                               ] 

not_found_diseases <- not_found_diseases[not_found_diseases != ""]

print(length(not_found_diseases))

```


### Match disease/trait to UMLS 

```{r match_disease_terms_to_umls, message=FALSE}

unmapped_disease_terms <-
  disease_mapping |>
  filter(is.na(icd10_code)) |>
  pull(`DISEASE/TRAIT`) |>
  unique() |>
  tolower()

# get UMLS CUIs with ICD10 mappings
umls_data <-
  data.table::fread(here::here("data/icd/2025AA/META/MRCONSO.RRF"),
                    sep = "|",
                    header = FALSE,
                    quote = "",
                    fill = TRUE,
                    na.strings = c("", "NA")
  )

colnames(umls_data)[1:18] <- c(
  "CUI","LAT","TS","LUI","STT","SUI","ISPREF",
  "AUI","SAUI","SCUI","SDUI","SAB","TTY","CODE",
  "STR","SRL","SUPPRESS","CVF"
)

umls_cuis_icd10 <-
  umls_data |>
  filter(SAB %in% c("ICD10", "ICD10CM")) |>
  select(CODE, CUI) |>
  group_by(CUI) |>
  summarise(CODE = str_flatten(unique(CODE),
                               collapse = ", ",
                               na.rm = T),
            .groups = "drop"
            ) |>
  rename(icd10_code = CODE) 



umls_icd10 <-
  umls_data |>
  filter(CUI %in% umls_cuis_icd10$CUI)

umls_icd10 =
  umls_icd10 |>
  left_join(umls_cuis_icd10,
            by = "CUI")

# overlap with umls terms
disease_trait_umls  <-
  umls_icd10 |>
  filter(tolower(STR) %in% unmapped_disease_terms) |>
  mutate(`DISEASE/TRAIT` = tolower(STR)) |>
  # mutate(CODE = ifelse(SAB %in% c("ICD10", "ICD10CM"),
  #                      CODE,
  #                      NA)) |>
  select(`DISEASE/TRAIT`,
         icd10_code) |>
  group_by(`DISEASE/TRAIT`) |>
  summarise(icd10_code = str_flatten(unique(icd10_code),
                               collapse = ", ",
                               na.rm = T),
            .groups = "drop"
            ) |>
  distinct() |>
  mutate(icd10_code_origin = "UMLS term match")

 disease_mapping <-
  disease_mapping |>
  mutate(`DISEASE/TRAIT` = tolower(`DISEASE/TRAIT`)) |>
  rows_patch(disease_trait_umls,
             by = "DISEASE/TRAIT")

```

# Get ICD10 codes by matching `collected_all_disease_terms` terms

### Match collected_all_disease_terms to ICD10 descriptions

```{r match_collected_icd10_desc}

matched_icd10_desc = 
  matched_icd10_desc |>
  rename(collected_all_disease_terms = `DISEASE/TRAIT`)

matched_icd10_desc = 
  matched_icd10_desc |>
  mutate(icd10_code_origin = "ICD Description Match (collected_all_disease_terms)")

disease_mapping =
  rows_patch(disease_mapping,
             matched_icd10_desc,
             unmatched = "ignore")

```

### How many diseases are not mapped yet?

```{r how_many_not_mapped_matched_phenotypes_collected_all_disease_terms_icd_desc}

print("Number of ICD10 codes obtained")
disease_mapping |>
    filter(!is.na(icd10_code)) |> 
    group_by(icd10_code_origin) |> 
    summarise(n = n(), 
              percent = n()/n_studies *100)

matched =
  disease_mapping |>
  filter(icd10_code != "") |>
  pull(collected_all_disease_terms)

not_found_diseases <- diseases[!diseases %in% 
                               matched
                               ] 

not_found_diseases <- not_found_diseases[not_found_diseases != ""]

print(length(not_found_diseases))

```



### Match collected_all_disease_terms to ICD10 codes (by phenotypes)

```{r match_collected_icd10}

matched_phenotypes = 
  matched_phenotypes |>
  rename(collected_all_disease_terms = `DISEASE/TRAIT`)

matched_phenotypes = 
  matched_phenotypes |>
  mutate(icd10_code_origin = "Phecode Phenotype Match (collected_all_disease_terms)")

# match by collected_all_disease_terms
disease_mapping = 
disease_mapping |>
rows_patch(matched_phenotypes,
           unmatched = "ignore")


```

### How many diseases are not mapped yet?

```{r how_many_not_mapped_matched_phenotypes_collected_all_disease_terms}

print("Number of ICD10 codes obtained")
disease_mapping |>
    filter(!is.na(icd10_code)) |> 
    group_by(icd10_code_origin) |> 
    summarise(n = n(), 
              percent = n()/n_studies *100)

matched =
  disease_mapping |>
  filter(icd10_code != "") |>
  pull(collected_all_disease_terms)

not_found_diseases <- diseases[!diseases %in% 
                               matched
                               ] 

not_found_diseases <- not_found_diseases[not_found_diseases != ""]

print(length(not_found_diseases))

```
### UMLS match collected_all_disease_terms

```{r match_collected_umls, message=FALSE}

unmapped_terms <-
  disease_mapping |>
  filter(is.na(icd10_code)) |>
  pull(collected_all_disease_terms) |>
  unique()

# overlap with umls terms
collected_trait_umls  <-
  umls_icd10 |>
  filter(tolower(STR) %in% unmapped_terms) |>
  mutate(collected_all_disease_terms = tolower(STR)) |>
  # mutate(CODE = ifelse(SAB %in% c("ICD10", "ICD10CM"),
  #                      CODE,
  #                      NA)) |>
  select(collected_all_disease_terms,
         icd10_code) |>
  group_by(collected_all_disease_terms) |>
  summarise(icd10_code = str_flatten(unique(icd10_code),
                               collapse = ", ",
                               na.rm = T),
            .groups = "drop"
            ) |>
  distinct() |>
  mutate(icd10_code_origin = "UMLS term match (collected_all_disease_terms)")

disease_mapping <-
  disease_mapping |>
  rows_patch(collected_trait_umls,
             by = "collected_all_disease_terms")


```

# Manual mapping of GWAS traits to ICD 10

```{r create_manual_mapping_table}

manual_icd10_map <-
  readxl::read_xlsx(here::here("data/icd/manual_disease_icd10_mappings.xlsx"))

manual_icd10_map =
  manual_icd10_map |>
  select(collected_all_disease_terms = mapped_trait, 
         icd10_code) |>
  mutate(collected_all_disease_terms = stringr::str_squish(tolower(collected_all_disease_terms))) |>
  mutate(icd10_code_origin = "Manual Mapping (collected_all_disease_terms)")

# disease_mapping =
#   bind_rows(disease_mapping, to_add) |>
#   distinct()
disease_mapping =
  rows_patch(disease_mapping,
             manual_icd10_map,
             unmatched = "ignore")

disease_mapping |>
  filter(icd10_code_origin == "Manual Mapping (collected_all_disease_terms)") |>
  nrow()

# repeat for `DISEASE/TRAIT`
manual_icd10_map =
  manual_icd10_map |>
  select(`DISEASE/TRAIT` = collected_all_disease_terms,
         icd10_code) |>
  mutate(icd10_code_origin = "Manual Mapping (DISEASE/TRAIT)")

disease_mapping =
  rows_patch(disease_mapping |> mutate(`DISEASE/TRAIT` = tolower(`DISEASE/TRAIT`)),
             manual_icd10_map,
             unmatched = "ignore")

```

## Additional manual mapping 

```{r additional_manual_mappings, eval = F}

disease_mapping =
  disease_mapping |>
  mutate(icd10_code = ifelse(collected_all_disease_terms == "toxicity" & 
                            grepl("Anthracycline-induced cardiotoxicity|Trastuzumab-induced cardiotoxicity", `DISEASE/TRAIT`, ignore.case = T),
                            "I42.7, T45.1",
                             icd10_code)) |>
mutate(icd10_code_origin = 
         ifelse(collected_all_disease_terms == "toxicity" & 
                  grepl("Anthracycline-induced cardiotoxicity|Trastuzumab-induced cardiotoxicity", `DISEASE/TRAIT`, ignore.case = T),
                "Manual Mapping (from DISEASE/TRAIT)",
                icd10_code_origin)
         )

disease_mapping =
  disease_mapping |>
  mutate(icd10_code = ifelse(collected_all_disease_terms == "toxicity" & 
                            grepl("Induration at the injection site after COVID-19", `DISEASE/TRAIT`, ignore.case = T),
                            "R23.4",
                             icd10_code)) |>
  mutate(collected_all_disease_terms = 
           ifelse(collected_all_disease_terms == "toxicity" & 
                    grepl("Induration at the injection site after COVID-19", `DISEASE/TRAIT`, ignore.case = T),
                  "induration of skin",
                  collected_all_disease_terms)) |>
  mutate(icd10_code_origin =
           ifelse(collected_all_disease_terms == "induration of skin",
                  "Manual Mapping (from DISEASE/TRAIT)",
                  icd10_code_origin)
         )


# N64.5
disease_mapping =
  disease_mapping |>
  mutate(icd10_code = ifelse(collected_all_disease_terms == "toxicity" & 
                            grepl("Induration \\(>= grade 2\\) in breast cancer treated with radiotherapy", 
                                  `DISEASE/TRAIT`, ignore.case = T),
                            "N64.5",
                             icd10_code)) |>
  mutate(collected_all_disease_terms = 
           ifelse(collected_all_disease_terms == "toxicity" & 
                    grepl("Induration \\(>= grade 2\\) in breast cancer treated with radiotherapy", 
                          `DISEASE/TRAIT`, ignore.case = T),
                  "induration of breast",
                  collected_all_disease_terms)) |>
  mutate(icd10_code_origin =
           ifelse(collected_all_disease_terms == "induration of breast",
                  "Manual Mapping (from DISEASE/TRAIT)",
                  icd10_code_origin)
         )


# T45.1
disease_mapping =
  disease_mapping |>
  mutate(icd10_code = ifelse(collected_all_disease_terms == "toxicity" & 
                            grepl("Nivolumab-induced immune-related adverse events in cancer|Response to immune checkpoint inhibitors in melanoma", 
                                  `DISEASE/TRAIT`, ignore.case = T),
                            "T45.1",
                             icd10_code)) |>
  mutate(icd10_code_origin =
           ifelse(collected_all_disease_terms == "toxicity" & 
                    grepl("Nivolumab-induced immune-related adverse events in cancer|Response to immune checkpoint inhibitors in melanoma", 
                          `DISEASE/TRAIT`, ignore.case = T),
                  "Manual Mapping (from DISEASE/TRAIT)",
                  icd10_code_origin)
         )


disease_mapping =
  disease_mapping |>
  mutate(icd10_code = ifelse(collected_all_disease_terms == "toxicity" & 
                            grepl("Methotrexate-related central neurotoxicity in children treated for acute lymphoblastic leukemia", 
                                  `DISEASE/TRAIT`, ignore.case = T),
                            "G92, T45.1",
                             icd10_code)) |>
  mutate(icd10_code_origin =
           ifelse(collected_all_disease_terms == "toxicity" & 
                    grepl("Methotrexate-related central neurotoxicity in children treated for acute lymphoblastic leukemia", 
                          `DISEASE/TRAIT`, ignore.case = T),
                  "Manual Mapping (from DISEASE/TRAIT)",
                  icd10_code_origin)
         )

## if DISEASE/TRAIT == 	Abnormalities of forces of labour
## and collected_all_disease_terms == abnormal delivery
## if icd10_code is missing, map to
disease_mapping =
  disease_mapping |>
  mutate(icd10_code = ifelse(collected_all_disease_terms == "abnormal delivery" & 
                            grepl("^Abnormalities of forces of labour$", 
                                  `DISEASE/TRAIT`, 
                                  ignore.case = T) &
                            icd10_code == "",
                            "O62",
                             icd10_code)) |>
  mutate(icd10_code_origin = 
         ifelse(collected_all_disease_terms == "abnormal delivery" & 
                  grepl("^Abnormalities of forces of labour$", 
                        `DISEASE/TRAIT`, 
                        ignore.case = T),
                "Manual Mapping (from DISEASE/TRAIT)",
                icd10_code_origin)
         )

## if DISEASE/TRAIT == Ischemic heart disease
## and collected_all_disease_terms == heart disease
## if icd10_code is missing, map to "I20-I21.6, I21.9-I25.9, Z82.4-Z82.49"
disease_mapping =
  disease_mapping |>
  mutate(icd10_code = ifelse(collected_all_disease_terms == "heart disease" & 
                            grepl("^Ischemic heart disease$", 
                                  `DISEASE/TRAIT`, 
                                  ignore.case = T) &
                            icd10_code == "",
                            "I20-I21.6, I21.9-I25.9, Z82.4-Z82.49",
                             icd10_code)) |>
  mutate(icd10_code_origin = 
         ifelse(collected_all_disease_terms == "heart disease" & 
                  grepl("^Ischemic heart disease$", 
                        `DISEASE/TRAIT`, 
                        ignore.case = T),
                "Manual Mapping (from DISEASE/TRAIT)",
                icd10_code_origin)
         )


# if collected_all_disease_terms == "heart disease"
# and icd10_code is missing
disease_mapping =
  disease_mapping |>
  mutate(icd10_code = ifelse(collected_all_disease_terms == "heart disease" &
                            icd10_code == "",
                            "C34",
                             icd10_code)) |>
  mutate(icd10_code_origin = 
         ifelse(collected_all_disease_terms == "heart disease" &
                  icd10_code == "C34",
                "Manual Mapping (collected_all_disease_terms)",
                icd10_code_origin)
         )

```

### How many diseases are not mapped yet?

```{r how_many_not_mapped_manual}

# disease_mapping =
#   disease_mapping |>
#   filter(icd10_code != "")

matched <- c(disease_mapping_matched$collected_all_disease_terms,
             matched_phenotypes$collected_all_disease_terms,
             to_add$collected_all_disease_terms,
             manual_icd10_map$collected_all_disease_terms)

not_found_diseases <- diseases[!diseases %in% matched] 
not_found_diseases <- not_found_diseases[not_found_diseases != ""]

print(length(not_found_diseases))

```

# Inferring ICD10 codes from similar study ICD10 codes

```{r}

# similar studies:
study_icd_map =
disease_mapping |>
filter(!is.na(icd10_code)) |>
filter(icd10_code_origin == "Study Provided" | icd10_code_origin == "Study PheCode Mapping") 

study_icd_map =
  study_icd_map |>
  select(collected_all_disease_terms, icd10_code) |>
  distinct()

# remove Z95.1 and Z95.5 from coronary artery disease
study_icd_map =
  study_icd_map |>
  mutate(icd10_code = ifelse(collected_all_disease_terms == "coronary artery disease",
                             str_replace_all(icd10_code, "Z95.1|Z95.5", ""),
                             icd10_code))

# assume ovarian cancer terms more specific than 
# Malignant neoplasm of ovary and other uterine adnexa (PheCode 184.1)
# study_icd_map =
#   study_icd_map |>
#   filter(!( collected_all_disease_terms == "ovarian cancer" &&phecode == "184.10")
#   )

study_icd_map = 
  study_icd_map |>
  mutate(collected_all_disease_terms = str_trim(collected_all_disease_terms)) |>
  mutate(collected_all_disease_terms = str_remove_all(collected_all_disease_terms, "^, ")) 
  

study_icd_map =
  study_icd_map |>
  filter(icd10_code != "" & !is.na(icd10_code))


study_icd_map =
  study_icd_map |>
  tidyr::separate_longer_delim(icd10_code, delim = ", ") |>
  tidyr::separate_longer_delim(icd10_code, delim = ",")

study_icd_map =
  study_icd_map |>
  group_by(collected_all_disease_terms) |>
  summarise(icd10_code = str_flatten(unique(sort(icd10_code)),
                                     collapse = ", ", 
                                     na.rm = T), 
            .groups = "drop")

study_icd_map = 
  study_icd_map |>
  mutate(icd10_code_origin = "Inferred from similar studies")

disease_mapping = 
  rows_patch(disease_mapping,
             study_icd_map,
             unmatched = "ignore")

disease_mapping |>
  filter(is.na(icd10_code)) |>
  nrow()

```


# How was ICD-10 code inferred?

```{r how_was_icd10_inferred}

disease_mapping  |> 
group_by(icd10_code_origin) |> 
summarise(n = n()) |> 
arrange(desc(n))

```



# Saving disease mapping 

## Add description of ICD10 codes to mapping table

### Prepare / fix disease mappings 

#### Fixing multiple ICD10 codes missing commas

```{r prepare_disease_mappings}

disease_mapping = 
  disease_mapping |> 
  mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "F40 F41 F42",
    replacement = "F40, F41, F42"
    )
  )

disease_mapping = 
  disease_mapping |> 
  mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "G47.8. G47.9",
    replacement = "G47.8, G47.9"
    )
  )

```

#### Converting ICD-10 cm codes to WHO ICD-10

```{r expand_disease_mappings}

disease_mapping =
  disease_mapping |>
  mutate(icd10_code = str_split(icd10_code, ",\\s*")) |>
  tidyr::unnest(icd10_code) 


disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "B37.49",
    replacement = "B37.4"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "E05.90",
    replacement = "E05.9"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
  mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "E11.319|E11.31|E11.329.9|E11.329.|E11.32|E11.3.",
    replacement = "E11.3"
    )
  )

disease_mapping = 
  disease_mapping |> 
  mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "E11.3.",
    replacement = "E11.3"
    )
  )

disease_mapping = 
  disease_mapping |> 
  mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "E13.621.|E13.62",
    replacement = "E11.3"
    )
  )

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "E78.00",
    replacement = "E78.0"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "F10.10",
    replacement = "F10.1"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "F17.201.|F17.20",
    replacement = "F17.2"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "F430",
    replacement = "F43.0"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "G47.00",
    replacement = "G47.0"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "H01.00|H01.09.",
    replacement = "H01.0"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "H01.09.",
    replacement = "H01.0"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "H029",
    replacement = "H02.9"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "H60.90",
    replacement = "H60.9"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "H60.90",
    replacement = "H60.9"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "H91.8X9.",
    replacement = "H91.8"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "H91.90",
    replacement = "H91.9"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "H93.299.|H93.29",
    replacement = "H01"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "I70.20",
    replacement = "I70.2"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "I82.409.|I82.40|I82.4",
    replacement = "I82"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "J30.9",
    replacement = "J30.2"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
  mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "J45.998.|J45.909.|J45.901.|J45.99|J45.90",
    replacement = "J45.9"
    )
  )

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "J98.457.6",
    replacement = "J98.4"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "K05.30-31",
    replacement = "K05.3"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "K29.70",
    replacement = "K29.7"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "K59.00",
    replacement = "K59.0"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "L08.89",
    replacement = "L08.8"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "M06.99|M06.90",
    replacement = "M06.9"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "M10.99",
    replacement = "M10.9"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "M10.99",
    replacement = "M10.9"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
  mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "M13.90|M13.94|M13.96|M13.97|M13.99",
    replacement = "M13.9"
    )
  )

disease_mapping = 
  disease_mapping |> 
  mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "M19.07",
    replacement = "M19.0"
    )
  )

disease_mapping = 
  disease_mapping |> 
  mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "M19.90|M19.91|M19.94|M19.97|M19.99",
    replacement = "M19.9"
    )
  )



disease_mapping = 
  disease_mapping |> 
  mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "M25.50|M25.51|M25.55|M25.569.|M25.56|M25.571.|M25.57",
    replacement = "M25.5"
    )
  )

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "M25.76|M25.77",
    replacement = "M25.7"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "M43.16",
    replacement = "M43.1"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
  mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "M47.80|M47.82|M47.86",
    replacement = "M47.8"
    )
  )

disease_mapping = 
  disease_mapping |> 
  mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "M47.92|M47.96",
    replacement = "M47.9"
    )
  )

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "M48.02|M48.06",
    replacement = "M48.0"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "M54.22",
    replacement = "M54.2"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "M54.30|M54.39",
    replacement = "M54.3"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "M54.56|M54.57|M54.59",
    replacement = "M54.5"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "M54.99",
    replacement = "M54.9"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "M65.34",
    replacement = "M65.3"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "M65.96",
    replacement = "M65.9"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "M72.04",
    replacement = "M72.0"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "M79.09",
    replacement = "M79.0"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "M79.65|M79.66|M79.67",
    replacement = "M79.6"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "M79.79",
    replacement = "M79.7"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "M79.86",
    replacement = "M79.8"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "M81.99",
    replacement = "M81.9"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "N390",
    replacement = "N39.0"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "N50.89",
    replacement = "N50.8"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "N52",
    replacement = "F52"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "N52.9",
    replacement = "F52.9"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "N814",
    replacement = "N81.4"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "P29.12",
    replacement = "P29.1"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "R06.00|R06.09",
    replacement = "R06.0"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "R06.83",
    replacement = "R06.8"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "R07.89",
    replacement = "R07.8"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "R07.8|R07.9",
    replacement = "R07"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "R09.89",
    replacement = "R09.8"
    )
    ) 


disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "R10.30",
    replacement = "R10.3"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "R10.8|R10.9",
    replacement = "R10"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "R11.0",
    replacement = "R11"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "R13.10",
    replacement = "R13"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "R13.1",
    replacement = "R13"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "R14.0",
    replacement = "R14"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "R19.7",
    replacement = "R19"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "R29.898.|R29.89",
    replacement = "R29.8"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "R31.29",
    replacement = "R31.2"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "R31.2|R31.9",
    replacement = "R31"
    )
    ) 


disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "R53.82|R53.83|R5382",
    replacement = "R53.8"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "R53.8",
    replacement = "R53"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "R56.9",
    replacement = "R56"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
  mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "R73.09|R73.02",
    replacement = "R73.0"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "R80.9",
    replacement = "R80"
    )
    ) 
 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "R82.998.|R82.99",
    replacement = "R82.9"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "R871",
    replacement = "R87.1"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "R87.6",
    replacement = "R87"
    )
    ) 


disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "R87.61",
    replacement = "R87.6"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "T50.905.",
    replacement = "T50.9"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "T78.40",
    replacement = "T78.4"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "T780",
    replacement = "T78.0"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "T81.149.88",
    replacement = "T81.1"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "T81.815.013.",
    replacement = "T81.8"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "T84.84",
    replacement = "T84.8"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "T887",
    replacement = "T88.7"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "Z86.79",
    replacement = "Z86.7"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "Z87.09",
    replacement = "Z87.0"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "Z87.39",
    replacement = "Z87.3"
    )
    ) 


disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "Z87.42",
    replacement = "Z87.4"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "Z87.891.",
    replacement = "Z87.8"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "W44.9",
    replacement = "W44"
    )
    ) 

disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "U80",
    replacement = "U82"
    )
    ) 

disease_mapping =
  disease_mapping |>
  distinct()




```

### Add ICD10 descriptions

#### Other ICD10 descriptions to add

```{r other_icd10_descriptions}


other_icd10_desc = 
  data.frame(
    icd10_code = c("A09.9", 
                  "A41",
                  "B95",
                  "B96",
                  "B97",
                  "B98",
                  "C44",
                  "C79",
                  "C80.0",
                  "C80.9",
                  "C90",
                  "D07",
                  "D35",
                  "D37",
                  "D47",
                  "D48",
                  "E03",
                  "E04",
                  "E21",
                  "E23",
                  "E27",
                  "E61",
                  "E80",
                  "E87",
                  "E89",
                  "F05",
                  "F17",
                  "F41",
                  "F43.0",
                  "F53",
                  "G31",
                  "G45",
                  "G46",
                  "G62",
                  "G93",
                  "G99",
                  "H47",
                  "H50",
                  "H57",
                  "H83",
                  "H91",
                  "H92",
                  "H93",
                  "I08",
                  "I27",
                  "I27.2",
                  "I44",
                  "I45",
                  "I48.0",
                  "I49",
                  "I51",
                  "I62",
                  "I71",
                  "J15",
                  "J16",
                  "J34",
                  "J38",
                  "J69",
                  "J84",
                  "J95",
                  "J96",
                  "K05",
                  "K07",
                  "K08",
                  "K09",
                  "K10",
                  "K12",
                  "K13",
                  "K22",
                  "K22.7",
                  "K52",
                  "K56",
                  "K64",
                  "K64.0",
                  "K64.9",
                  "K74",
                  "K75",
                  "K85.9",
                  "K86",
                  "K91",
                  "K91.8",
                  "K92",
                  "L02",
                  "L13",
                  "L30",
                  "L65",
                  "L73",
                  "L85",
                  "L89.1",
                  "L98",
                  "M05",
                  "M06",
                  "M07",
                  "M11",
                  "M13",
                  "M18",
                  "M19",
                  "M24",
                  "M25",
                  "M31",
                  "M35",
                  "M43",
                  "M48",
                  "M53",
                  "M62",
                  "M66",
                  "M67",
                  "M71",
                  "M77",
                  "M79",
                  "M79.7",
                  "M80",
                  "M81",
                  "M85",
                  "M89",
                  "M96",
                  "N02",
                  "N18.3",
                  "N18.4",
                  "N28",
                  "N39",
                  "N48",
                  "N73",
                  "N76",
                  "N88",
                  "N89",
                  "N91",
                  "N92",
                  "N93",
                  "N94",
                  "N99",
                  "O14",
                  "O04",
                  "O26",
                  "O32",
                  "O34",
                  "O36",
                  "O68",
                  "O75",
                  "O99",
                  "R03",
                  "R07",
                  "R09",
                  "R10",
                  "R19",
                  "R29",
                  "R29.6",
                  "R39",
                  "R40",
                  "R47",
                  "R57",
                  "T95.8",
                  "W44",
                  "Y95",
                  "Z86.3",
                  "Z87.3",
                  "Z87.4",
                  "Z87.7",
                  "Z87.8",
                  "T85.8",
                  "Z91.0",
                  "Z88.9",
                  "Z88.8",
                  "Z88",
                  "Z92.6",
                  "N90",
                  "U82"
                  ),
    icd10_description = c("Gastroenteritis and colitis of unspecified origin",
                         "Other sepsis",
                         "Streptococcus and staphylococcus as the cause of diseases classified to other chapters",
                         "Other specified bacterial agents as the cause of diseases classified to other chapters",
                         "Viral agents as the cause of diseases classified to other chapters",
                         "Other specified infectious agents as the cause of diseases classified to other chapters",
                         "Other malignant neoplasms of skin",
                         "Secondary malignant neoplasm of other and unspecified sites",
                         "Malignant neoplasm, primary site unknown, so stated",
                         "Malignant neoplasm, primary site unspecified",
                         "Multiple myeloma and malignant plasma cell neoplasms",
                         "Carcinoma in situ of other and unspecified genital organs",
                         "Benign neoplasm of other and unspecified endocrine glands",
                         "Neoplasm of uncertain or unknown behaviour of oral cavity and digestive organs",
                         "Other neoplasms of uncertain or unknown behaviour of lymphoid, haematopoietic and related tissue",
                         "Neoplasm of uncertain or unknown behaviour of other and unspecified sites",
                         "Other hypothyroidism",
                         "Other nontoxic goitre",
                         "Hyperparathyroidism and other disorders of parathyroid gland",
                         "Hypofunction and other disorders of pituitary gland",
                         "Other disorders of adrenal gland",
                         "Deficiency of other nutrient elements",
                         "Disorders of porphyrin and bilirubin metabolism",
                         "Other disorders of fluid, electrolyte and acid-base balance",
                         "Postprocedural endocrine and metabolic disorders, not elsewhere classified",
                         "Delirium, not induced by alcohol and other psychoactive substances",
                         "Mental and behavioural disorders due to use of tobacco",
                         "Other anxiety disorders",
                         "Acute stress reaction",
                         "Mental and behavioural disorders associated with the puerperium, not elsewhere classified",
                         "Other degenerative diseases of nervous system, not elsewhere classified",
                         "Transient cerebral ischaemic attacks and related syndromes",
                         "Vascular syndromes of brain in cerebrovascular diseases",
                         "Other polyneuropathies",
                         "Other disorders of brain",
                         "Other disorders of nervous system in diseases classified elsewhere",
                         "Other disorders of optic [2nd] nerve and visual pathways",
                         "Other strabismus",
                         "Other disorders of eye and adnexa",
                         "Other diseases of inner ear",
                         "Other hearing loss",
                         "Otalgia and effusion of ear",
                         "Other disorders of ear, not elsewhere classified",
                         "Multiple valve diseases",
                         "Other pulmonary heart diseases",
                         "Other secondary pulmonary hypertension",
                         "Atrioventricular and left bundle-branch block",
                         "Other conduction disorders",
                         "Paroxysmal atrial fibrillation",
                         "Ventricular fibrillation and flutter",
                         "Complications and ill-defined descriptions of heart disease",
                         "Other nontraumatic intracranial haemorrhage",
                         "Aortic aneurysm and dissection",
                         "Bacterial pneumonia, not elsewhere classified",
                         "Pneumonia due to other infectious organisms, not elsewhere classified",
                         "Other disorders of nose and nasal sinuses",
                         "Diseases of vocal cords and larynx, not elsewhere classified",
                         "Pneumonitis due to solids and liquids",
                         "Other interstitial pulmonary diseases",
                         "Postprocedural respiratory disorders, not elsewhere classified",
                         "Respiratory failure, not elsewhere classified",
                         "Gingivitis and periodontal diseases",
                         "Dentofacial anomalies [including malocclusion]",
                         "Other disorders of teeth and supporting structures",
                         "Cysts of oral region, not elsewhere classified",
                         "Other diseases of jaws",
                         "Stomatitis and related lesions",
                         "Other diseases of lip and oral mucosa",
                         "Other diseases of oesophagus",
                         "Barrett oesophagus",
                         "Other noninfective gastroenteritis and colitis",
                         "Paralytic ileus and intestinal obstruction without hernia",
                         "Haemorrhoids and perianal venous thrombosis",
                         "First degree haemorrhoids",
                         "Haemorrhoids, unspecified",
                         "Fibrosis and cirrhosis of liver",
                         "Other inflammatory liver diseases",
                         "Acute pancreatitis, unspecified",
                         "Other diseases of pancreas",
                         "Postprocedural disorders of digestive system, not elsewhere classified",
                         "Other postprocedural disorders of digestive system, not elsewhere classified",
                         "Other diseases of digestive system",
                         "Cutaneous abscess, furuncle and carbuncle",
                         "Other bullous disorders",
                         "Other dermatitis",
                         "Other nonscarring hair loss",
                         "Other follicular disorders",
                         "Other epidermal thickening",
                         "Stage II decubitus ulcer",
                         "Other disorders of skin and subcutaneous tissue, not elsewhere classified",
                         "Seropositive rheumatoid arthritis",
                         "Other rheumatoid arthritis",
                         "Psoriatic and enteropathic arthropathies",
                         "Other crystal arthropathies",
                         "Other arthritis",
                         "Arthrosis of first carpometacarpal joint",
                         "Other arthrosis",
                         "Other specific joint derangements",
                         "Other joint disorders, not elsewhere classified",
                         "Other necrotizing vasculopathies",
                         "Other systemic involvement of connective tissue",
                         "Other deforming dorsopathies",
                         "Other spondylopathies",
                         "Other dorsopathies, not elsewhere classified",
                         "Other disorders of muscle",
                         "Spontaneous rupture of synovium and tendon",
                         "Other disorders of synovium and tendon",
                         "Other bursopathies",
                         "Other enthesopathies",
                         "Other soft tissue disorders, not elsewhere classified",
                         "Fibromyalgia",
                         "Osteoporosis with pathological fracture",
                         "Osteoporosis without pathological fracture",
                         "Other disorders of bone density and structure",
                         "Other disorders of bone",
                         "Postprocedural musculoskeletal disorders, not elsewhere classified",
                         "Recurrent and persistent haematuria",
                         "Chronic kidney disease, stage 3",
                         "Chronic kidney disease, stage 4",
                         "Other disorders of kidney and ureter, not elsewhere classified",
                         "Other disorders of urinary system",
                         "Other disorders of penis",
                         "Other female pelvic inflammatory diseases",
                         "Other inflammation of vagina and vulva",
                         "Other noninflammatory disorders of cervix uteri",
                         "Other noninflammatory disorders of vagina",
                         "Other noninflammatory disorders of vulva and perineum",
                         "Excessive, frequent and irregular menstruation",
                         "Other abnormal uterine and vaginal bleeding",
                         "Pain and other conditions associated with female genital organs and menstrual cycle",
                         "Postprocedural disorders of genitourinary system, not elsewhere classified",
                         "Pre-eclampsia",
                         "Medical abortion",
                         "Maternal care for other conditions predominantly related to pregnancy",
                         "Maternal care for known or suspected malpresentation of fetus",
                         "Maternal care for known or suspected abnormality of pelvic organs",
                         "Maternal care for other known or suspected fetal problems",
                         "Labour and delivery complicated by fetal stress [distress]",
                         "Other complications of labour and delivery, not elsewhere classified",
                         "Other maternal diseases classifiable elsewhere but complicating pregnancy, childbirth and the puerperium",
                         "Abnormal blood-pressure reading, without diagnosis",
                         "Pain in throat and chest",
                         "Other symptoms and signs involving the circulatory and respiratory systems",
                         "Abdominal and pelvic pain",
                         "Other symptoms and signs involving the digestive system and abdomen",
                         "Other symptoms and signs involving the nervous and musculoskeletal systems",
                         "Tendency to fall, not elsewhere classified",
                         "Other symptoms and signs involving the urinary system",
                         "Somnolence, stupor and coma",
                         "Speech disturbances, not elsewhere classified",
                         "Shock, not elsewhere classified",
                         "Other complications of internal prosthetic devices, implants and grafts, not elsewhere classified",
                         "Foreign body entering into or through eye or natural orifice",
                         "Nosocomial condition",
                         "Personal history of endocrine, nutritional and metabolic diseases",
                         "Personal history of diseases of the musculoskeletal system and connective tissue",
                         "Personal history of diseases of the genitourinary system",
                         "Personal history of congenital malformations, deformations and chromosomal abnormalities",
                         "Personal history of other specified conditions",
                         "Other complications of internal prosthetic devices, implants and grafts, not elsewhere classified",
                         "Personal history of allergy, other than to drugs and biological substances",
                         "Personal history of allergy to unspecified drugs, medicaments and biological substances",
                         "Personal history of allergy to other drugs, medicaments and biological substances
",
"Personal history of allergy to drugs, medicaments and biological substances",
"Personal history of chemotherapy for neoplastic disease",
"Other noninflammatory disorders of vulva and perineum",
"Resistance to betalactam antibiotics")
  )

```

```{r add_icd10_descriptions}

manual_icd10_map <-
  readxl::read_xlsx(here::here("data/icd/manual_disease_icd10_mappings.xlsx"))

icd10_descriptions =
  phecodes |>
  select(icd10_code = ICD10, 
         icd10_description = ICD_DESCRIPTION
         ) |>
  distinct()

# Expand multiple ICD codes into rows
to_add_expanded <- manual_icd10_map |>
  mutate(icd10_code = str_split(icd10_code, ",\\s*")) |>
  tidyr::unnest(icd10_code)

icd10_descriptions = 
  bind_rows(
    icd10_descriptions,
    to_add_expanded |>
      select(icd10_code, 
             icd10_description = icd10_desc),
    other_icd10_desc
  ) 

icd10_descriptions = icd10_descriptions |> distinct()

icd10_descriptions = 
  icd10_descriptions |>
  group_by(icd10_code) |>
  summarise(icd10_description = 
            str_flatten(unique(icd10_description), collapse = "; ", na.rm = T), 
            .groups = "drop"
            )

disease_mapping =
  left_join(disease_mapping,
            icd10_descriptions,
            by = "icd10_code",
            relationship = "many-to-one",
            na_matches = "never"
            ) 


disease_mapping = 
  disease_mapping |> 
    mutate(icd10_code = str_replace_all(
    icd10_code,
    pattern = "R871",
    replacement = "R87"
    )
    ) 


disease_mapping = 
  disease_mapping |> 
    mutate(icd10_description = 
    ifelse(icd10_code == "R87",
           "Abnormal findings in specimens from female genital organs",
           icd10_description)
    ) 


```

### Missing descriptions for ICD10 codes

```{r missing_icd10_descriptions}
# check if any ICD10 codes are missing descriptions
disease_mapping |> 
  filter(is.na(icd10_description)) |>
  head()
```

### Noticing some mistakes in MAPPED trait by icd 10 codes

```{r check_mapped_trait}

gwas_study_info |> 
  filter(grepl("ICD10 F05", 
               `DISEASE/TRAIT`)) |> 
  select(`DISEASE/TRAIT`, MAPPED_TRAIT, collected_all_disease_terms, PUBMED_ID)


disease_mapping |> 
  filter(icd10_code == "F05")

# F05 refers to delirium not induced by alcohol and other psychoactive substances
# yet, the mapped trait is alcohol withdrawal delirium
# replace with delirium
disease_mapping = 
disease_mapping |>
  mutate(collected_all_disease_terms = 
         ifelse(icd10_code == "F05",
                "delirium",
                collected_all_disease_terms)
         )

gwas_study_info =
  gwas_study_info |>
  mutate(MAPPED_TRAIT = 
         ifelse(grepl("ICD10 F05", `DISEASE/TRAIT`),
                str_replace_all(collected_all_disease_terms,
                                pattern = "alcohol-related disorders",
                                replacement = "delirium"),
                collected_all_disease_terms)
         )

```

### Save disease map

```{r save_disease_mapping}

disease_mapping = 
  disease_mapping |> 
  arrange(collected_all_disease_terms, icd10_code)

fwrite(disease_mapping,
       here::here("output/icd_map/gwas_disease_to_icd10_mapping.csv")
       )

```


        


