---
title: "Non-ontology based trait collapsing"
author: "Isobel Beasley"
date: "2025-09-06"
output:
  workflowr::wflow_html:
    toc: true
    toc_float: true
    number_sections: true
editor_options: 
  markdown: 
    wrap: 72
---


# Set up

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r required_fns, message = F, warning= F}

library(dplyr)
library(data.table)
library(ggplot2)
library(stringr)

```

```{r setup_2, message = FALSE, warning= FALSE}

gwas_study_info <- fread(here::here("output/gwas_study_info_trait_ontology_info.csv"))


```


# Number of studies per trait - before any grouping 

```{r}

# gwas_study_info |> 
#   dplyr::filter(DISEASE_STUDY == T) |>
#   dplyr::filter(all_disease_terms == "") |>
#   View()

n_studies_trait = gwas_study_info |>
  dplyr::filter(DISEASE_STUDY == T) |>
  dplyr::select(all_disease_terms, PUBMED_ID) |>
  dplyr::distinct() |>
  dplyr::group_by(all_disease_terms) |>
  dplyr::summarise(n_studies = dplyr::n()) |>
  dplyr::arrange(desc(n_studies))

head(n_studies_trait)

dim(n_studies_trait)

```


# Collapse more traits



```{r}

gwas_study_info$all_disease_terms = sub(",$", "", gwas_study_info$all_disease_terms)

gwas_study_info$all_disease_terms = stringr::str_trim(gwas_study_info$all_disease_terms)

# collapse more traits 
gwas_study_info = gwas_study_info |> 
  mutate(collected_all_disease_terms = all_disease_terms) 


```


## Susceptibility to X measurement, X disorder measurement etc.

```{r}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms = 
        str_replace(collected_all_disease_terms, 
                    "^susceptibility to (.*?) measurement$", "\\1")

         ) 

gwas_study_info = gwas_study_info |>
    mutate(collected_all_disease_terms = 
          stringr::str_remove(collected_all_disease_terms, " symptom severity measurement")
         ) |>
  mutate(collected_all_disease_terms = 
          stringr::str_remove(collected_all_disease_terms, " severity measurement| exacerbation measurement")
         ) 



gwas_study_info = gwas_study_info |>
    mutate(collected_all_disease_terms =
           str_replace_all(collected_all_disease_terms,
                           "\\bdependence\\s+measurement\\b",
                           "dependence")) |> 
  mutate(collected_all_disease_terms =
           str_replace_all(collected_all_disease_terms,
                           "\\ballergy\\s+measurement\\b",
                           "allergy")) |> 
  mutate(collected_all_disease_terms =
           str_replace_all(collected_all_disease_terms,
                           "\\baddiction\\s+measurement\\b",
                           "addiction")) |> 
    mutate(collected_all_disease_terms =
           str_replace_all(collected_all_disease_terms,
                           "\\baddiction\\s+measurement\\b",
                           "addiction")) |> 
      mutate(collected_all_disease_terms =
           str_replace_all(collected_all_disease_terms,
                           "\\bsyndrome\\s+measurement\\b",
                           "syndrome")) |> 
      mutate(collected_all_disease_terms =
           str_replace_all(collected_all_disease_terms,
                           "\\bdisorder\\s+measurement\\b",
                           "disorder")) 


gwas_study_info = gwas_study_info |>
      mutate(collected_all_disease_terms = 
         stringr::str_remove(collected_all_disease_terms,
                         pattern = "age of onset of"))

```


## Cancer

```{r}
  

gwas_study_info = gwas_study_info |>
      mutate(collected_all_disease_terms = 
           stringr::str_replace_all(collected_all_disease_terms,
                            "\\bcarcinoma",
                            "cancer")
         )

gwas_study_info = gwas_study_info |>
      mutate(collected_all_disease_terms = 
           stringr::str_replace_all(collected_all_disease_terms,
                            "triple-negative breast cancer",
                            "breast cancer"
           ))


```

### Melanoma

```{r}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms = 
         stringr::str_replace_all(collected_all_disease_terms,
                          "cutaneous melanoma",
                          "melanoma"
         ))

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms = 
         stringr::str_replace_all(collected_all_disease_terms,
                          "uveal melanoma|uveal melanoma disease severity|epithelioid cell uveal melanoma|choroidal melanoma",
                          "ocular melanoma"
         ))


```


## Substance abuse

### Alcohol 



```{r}

gwas_study_info = gwas_study_info |> 
    mutate(collected_all_disease_terms = 
         stringr::str_replace_all(collected_all_disease_terms,
                          "alcohol dependence measurement",
                          "alcohol dependence"
         ))

gwas_study_info = gwas_study_info |>
      mutate(collected_all_disease_terms = 
           stringr::str_replace_all(collected_all_disease_terms,
                            "alcohol dependence|alcohol withdrawal|alcohol withdrawal delirium",
                            "alcohol-related disorders"
           ))


```

```{r}
  
gwas_study_info = gwas_study_info |>
    mutate(collected_all_disease_terms = 
         stringr::str_replace_all(collected_all_disease_terms,
                          "cannabis dependence measurement",
                          "cannabis dependence"
         ))

```

## Asthma

```{r}


gwas_study_info = gwas_study_info |> 
    mutate(collected_all_disease_terms = 
         stringr::str_replace_all(collected_all_disease_terms,
                          "asthma exacerbation measurement",
                          "asthma"
         )) |>
    mutate(collected_all_disease_terms = 
           stringr::str_replace_all(collected_all_disease_terms,
                            "adult onset asthma|childhood onset asthma",
                            "asthma"
           )) |>
  mutate(collected_all_disease_terms = 
           stringr::str_replace_all(collected_all_disease_terms,
                            "asthma symptoms measurement",
                            "asthma"
           )) |> 
    mutate(collected_all_disease_terms = 
           stringr::str_replace_all(collected_all_disease_terms,
                            "atopic asthma",
                            "asthma"
           )) |>
    mutate(collected_all_disease_terms = 
           stringr::str_replace_all(collected_all_disease_terms,
                            "asthma, asthma",
                            "asthma"
           )) |>
  mutate(collected_all_disease_terms = 
           stringr::str_replace_all(collected_all_disease_terms,
                            "chronic obstructive asthma",
                            "asthma"
           ))
           




```


## Psychiatric 

```{r}

gwas_study_info = gwas_study_info |> 
    mutate(collected_all_disease_terms = 
         stringr::str_replace_all(collected_all_disease_terms,
                          "bipolar ii disorder|bipolar i disorder",
                          "bipolar disorder"
         ))


gwas_study_info = gwas_study_info |> 
    mutate(collected_all_disease_terms = 
         stringr::str_replace_all(collected_all_disease_terms,
                          "insomnia measurement",
                          "insomnia"
         ))


gwas_study_info = gwas_study_info |> 
    mutate(collected_all_disease_terms = 
         stringr::str_replace_all(collected_all_disease_terms,
                          "post-traumatic stress disorder symptom measurement",
                          "post-traumatic stress disorder"
         )) |>
    mutate(collected_all_disease_terms = 
         stringr::str_replace_all(collected_all_disease_terms,
                          "multiple sclerosis symptom measurement",
                          "multiple sclerosis"
         ))  


gwas_study_info = gwas_study_info |> 
    mutate(collected_all_disease_terms = 
         stringr::str_replace_all(collected_all_disease_terms,
                          "anxiety disorder measurement",
                          "anxiety disorder"
         ))




gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms = 
         stringr::str_replace_all(collected_all_disease_terms,
                          "agoraphobia symptom measurement",
                          "agoraphobia"
         ))

```

### ADHD & autism

```{r}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms = 
        stringr::str_replace_all(collected_all_disease_terms,
                         "autism spectrum disorder",
                         "autism")
        )


gwas_study_info = gwas_study_info |> 
    mutate(collected_all_disease_terms = 
           stringr::str_replace_all(collected_all_disease_terms,
                            "attention deficit hyperactivity disorder",
                            "adhd"
         )) |>
  mutate(collected_all_disease_terms = 
           stringr::str_replace_all(collected_all_disease_terms,
                            "adhd symptom measurement",
                            "adhd")
         )

```

### Dementia, Alzheimer's, Parkinson's etc.

```{r}


gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms = 
         stringr::str_replace_all(collected_all_disease_terms,
                          "early-onset alzheimers disease|late-onset alzheimers disease",
                          "alzheimers disease"
         )) |>
    mutate(collected_all_disease_terms = 
         stringr::str_replace_all(collected_all_disease_terms,
                          "alzheimer's disease",
                          "alzheimers disease"
         )) |>
  mutate(collected_all_disease_terms = 
         stringr::str_replace_all(collected_all_disease_terms,
                          "family history of alzheimer’s disease",
                          "alzheimers disease"
         )) |>
  mutate(collected_all_disease_terms =
         stringr::str_replace_all(collected_all_disease_terms,
                          "alzheimer disease",
                          "alzheimers disease"
         ))

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms = 
         stringr::str_replace_all(collected_all_disease_terms,
                          "parkinson's disease symptom measurement",
                          "parkinsons disease"
         ))


```

### Sleep Apnea 

```{r}

gwas_study_info = gwas_study_info |> 
    mutate(collected_all_disease_terms = 
         stringr::str_replace_all(collected_all_disease_terms,
                          "lewy body dementia measurement",
                          "lewy body dementia"
         )) |>
      mutate(collected_all_disease_terms = 
         stringr::str_replace_all(collected_all_disease_terms,
                          "sleep apnea measurement",
                          "sleep apnea"
         )) |>
  mutate(collected_all_disease_terms = 
         stringr::str_replace_all(collected_all_disease_terms,
                          "sleep apnea during non-rem sleep|sleep apnea during rem sleep",
                          "sleep apnea")
  )

```


```{r}


gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms = 
         stringr::str_replace_all(collected_all_disease_terms,
                          "contact dermatitis due to nickel",
                          "contact dermatitis"
         ))


gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms = 
         stringr::str_replace_all(collected_all_disease_terms,
                          "covid-19 symptoms measurement",
                          "covid-19")
  )

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms = 
         stringr::str_replace_all(collected_all_disease_terms,
                          "coronary atherosclerosis measurement",
                          "coronary atherosclerosis")
  ) 

```


## Allergies 


```{r}         

         
gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms = 
         stringr::str_replace_all(collected_all_disease_terms,
                          "allergic contact dermatitis of eyelid",
                          "allergic contact dermatitis"
         ))


gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms = 
         stringr::str_replace_all(collected_all_disease_terms,
                          "seasonal allergic rhinitis",
                          "allergic rhinitis"
         ))


```



## Infertility


```{r}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms = 
         stringr::str_replace_all(collected_all_disease_terms,
                          "tubal factor infertility",
                          "female infertility"
                          )) |>
  mutate(collected_all_disease_terms = 
                  stringr::str_replace_all(collected_all_disease_terms,
                          "azoospermia",
                          "male infertility"
                          )  )



```


## Putting it all together



```{r}

gwas_study_info = gwas_study_info |>
  rowwise() |>
  mutate(
    collected_all_disease_terms = paste0(
      unique(unlist(stringr::str_split(collected_all_disease_terms, ", "))),
      collapse = ", "
    )
  ) |>
  ungroup()

           
gwas_study_info$collected_all_disease_terms = stringr::str_trim(gwas_study_info$collected_all_disease_terms)
gwas_study_info$collected_all_disease_terms = sub(",$", "", gwas_study_info$collected_all_disease_terms)

n_studies_trait = gwas_study_info |>
  dplyr::filter(DISEASE_STUDY == T) |>
  dplyr::select(collected_all_disease_terms, PUBMED_ID) |>
  dplyr::distinct() |>
  dplyr::group_by(collected_all_disease_terms) |>
  dplyr::summarise(n_studies = dplyr::n()) |>
  dplyr::arrange(desc(n_studies))

head(n_studies_trait)

dim(n_studies_trait)

```


# To do

"asperger syndrome" 


```{r}

diseases <- stringr::str_split(pattern = ", ", 
                               n_studies_trait$collected_all_disease_terms)  |> 
            unlist() |>
            stringr::str_trim()

length(unique(diseases))



```


# (?) Even more grouping to maybe add later:
- Cancer
- Cardiovascular disease
- Autoimmune disease
- Allergy
- Dementia
- Substance abuse
