---
title: "Other disease/disorder filtering + fixing"
author: "Isobel Beasley"
date: "2025-09-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Set up

```{r setup_2, message=FALSE, warning=FALSE}

library(data.table)
library(dplyr)
library(stringr)

gwas_study_info <- fread(here::here("output/gwas_cat/gwas_study_info_disease_trait_filtered.csv"))

```

## Ontology help - for getting disease subtypes

```{r ontology_fns}

source(here::here("code/get_term_descendants.R"))

```

# Injuries 

## Poisoning

### Combining Poisoning

```{r poisoning}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          vec_to_grep_pattern("carbon monoxide poisoning"),
                          "poisoning"
         ))

```

```{r remove_poisoning}

gwas_study_info = gwas_study_info |>
           mutate(collected_all_disease_terms =
           stringr::str_remove_all(collected_all_disease_terms,
                          pattern = vec_to_grep_pattern("poisoning")
                          )
           )
```



## Bone fracture

```{r bone_fracture}

url <- "http://www.ebi.ac.uk/ols4/api/ontologies/efo/terms/http%253A%252F%252Fwww.ebi.ac.uk%252Fefo%252FEFO_0003931/descendants"

bone_fracture_terms <- get_descendants(url)

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          pattern = vec_to_grep_pattern(bone_fracture_terms),
                          "bone fracture"
         ))

gwas_study_info = gwas_study_info |>
           mutate(collected_all_disease_terms =
           stringr::str_remove_all(collected_all_disease_terms,
                          pattern = vec_to_grep_pattern("bone fracture")
                          )
           )

```


## Toxicity 

```{r toxicity}

toxicity_terms = c("cardiotoxicity",
                   "dermatological toxicity",
                   "immune system toxicity",  
                   "neurotoxicity"
                   )

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          pattern = vec_to_grep_pattern(toxicity_terms),
                          "toxicity"
         ))

gwas_study_info = gwas_study_info |>
           mutate(collected_all_disease_terms =
           stringr::str_remove_all(collected_all_disease_terms,
                          pattern = vec_to_grep_pattern("toxicity")
                          )
           )

```


# Naming Fixing: 

## Nausea, vomiting

### Pregancy nausea 

```{r pregnancy_nausea}

gwas_study_info |> 
  dplyr::filter(PUBMED_ID == "38509478") |> 
  select(`DISEASE/TRAIT`, MAPPED_TRAIT, collected_all_disease_terms)


# replace nausea and vomiting, with vomiting of pregnancy

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
          ifelse(PUBMED_ID == "38509478",
         stringr::str_replace_all(collected_all_disease_terms,
                          pattern = vec_to_grep_pattern(c("nausea and vomiting")),
                          "vomiting of pregnancy"),
         collected_all_disease_terms
         )
  )

# where MAPPED_TRAIT is "nausea and vomiting of pregnancy" add vomiting of pregnancy

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
          ifelse(MAPPED_TRAIT == "nausea and vomiting of pregnancy severity measurement" & 
                 PUBMED_ID == "38509478",
         paste0(collected_all_disease_terms, ", vomiting of pregnancy"),
         collected_all_disease_terms)
  )


```


### Post-operative nausea and vomiting

```{r post_op_nausea}


gwas_study_info |> 
  filter(grepl("nausea", collected_all_disease_terms))  |>
  select(`DISEASE/TRAIT`, MAPPED_TRAIT, collected_all_disease_terms) |>
  distinct()

# for given pubmed id, remove post operative nausea and vomiting

gwas_study_info = 
  gwas_study_info |>
  mutate(collected_all_disease_terms = 
           ifelse(PUBMED_ID %in% c(21694509, 32928300, 37156489),
                  stringr::str_remove_all(
                    collected_all_disease_terms,
                    pattern = vec_to_grep_pattern(c("post operative nausea and vomiting", 
                                                    "post-operative nausea and vomiting"))
                  ),
                  collected_all_disease_terms
           )
  )
  
```

### Unspecified nausea 

```{r unspecified_nausea}


# for given pubmed id, remove unspecified nausea
gwas_study_info  = 
 gwas_study_info |>
  mutate(collected_all_disease_terms  = 
          ifelse(PUBMED_ID %in% c(30104761, 34737426, 39024449, 34662886),
         stringr::str_remove_all(
           collected_all_disease_terms,
            pattern = vec_to_grep_pattern(c("nausea and vomiting"))
                          ),
         collected_all_disease_terms)
  )

```

### Unspecified vomiting

```{r unspecified_vomiting}

# where `DISEASE/TRAIT' contains ICD10 R1, or PheCode 789.1 remove vomiting

gwas_study_info  = 
 gwas_study_info |>
  mutate(collected_all_disease_terms  = 
          ifelse(grepl("ICD10 R11", `DISEASE/TRAIT`) | grepl("PheCode 789.1", `DISEASE/TRAIT`),
         stringr::str_remove_all(
           collected_all_disease_terms,
            pattern = vec_to_grep_pattern(c("vomiting"))
                          ),
         collected_all_disease_terms)
  )

# where `DISEASE/TRAIT' contains BNT162b2 vaccine, remove vomiting (and covid-19 implicitly)

gwas_study_info  = 
 gwas_study_info |>
  mutate(collected_all_disease_terms  = 
          ifelse(grepl("BNT162b2 vaccine", `DISEASE/TRAIT`, ignore.case = TRUE),
         stringr::str_remove_all(
           collected_all_disease_terms,
            pattern = vec_to_grep_pattern(c("vomiting"))
                          ),
         collected_all_disease_terms)
  )

```

## Fever

```{r fever}

# remove fever where `DISEASE/TRAIT' contains BNT162b2 vaccine (and covid-19 implicitly)

gwas_study_info  = 
 gwas_study_info |>
  mutate(collected_all_disease_terms  = 
          ifelse(grepl("BNT162b2 vaccine|COVID-19 booster vaccination", `DISEASE/TRAIT`, ignore.case = TRUE),
         stringr::str_remove_all(
           collected_all_disease_terms,
            pattern = vec_to_grep_pattern(c("fever"))
                          ),
         collected_all_disease_terms)
  )


# remove fever where `DISEASE/TRAIT' contains "Postprocedural fever"


gwas_study_info  = 
 gwas_study_info |>
  mutate(collected_all_disease_terms  = 
          ifelse(grepl("Postprocedural fever", `DISEASE/TRAIT`, ignore.case = TRUE),
         stringr::str_remove_all(
           collected_all_disease_terms,
            pattern = vec_to_grep_pattern(c("fever"))
                          ),
         collected_all_disease_terms)
  )


# where vec_to_grep_pattern(c("fever")) is in collected_all_disease_terms, and Fever when IBS symptoms first began
# is in `DISEASE/TRAIT`, replace fever with irritable bowel syndrome
gwas_study_info  = 
 gwas_study_info |>
  mutate(collected_all_disease_terms  = 
          ifelse(grepl("Fever when IBS symptoms first began",
                       `DISEASE/TRAIT`, 
                       ignore.case = TRUE) &
                 grepl(vec_to_grep_pattern(c("fever")), 
                       collected_all_disease_terms,
                       perl = T),
         stringr::str_replace_all(
           collected_all_disease_terms,
            pattern = vec_to_grep_pattern(c("fever")),
            replacement = "irritable bowel syndrome"
                          ),
         collected_all_disease_terms)
  )


```


## Sneeze

```{r sneeze}

# remove sneeze from collected_all_disease_terms 

gwas_study_info  = 
 gwas_study_info |>
  mutate(collected_all_disease_terms  = 
          ifelse(grepl("sneeze", collected_all_disease_terms, ignore.case = TRUE),
         stringr::str_remove_all(
           collected_all_disease_terms,
            pattern = vec_to_grep_pattern(c("sneeze"))
                          ),
         collected_all_disease_terms)
  )

```


# Unspecified disease

## Asthma, and other disease

```{r unspecified_disease}

# if DISEASE/TRAIT contains Asthma in any disease, remove disease from collected_all_disease_terms

gwas_study_info  = 
 gwas_study_info |>
  mutate(collected_all_disease_terms  = 
          ifelse(grepl("Asthma in any disease", `DISEASE/TRAIT`, ignore.case = TRUE),
         stringr::str_remove_all(
           collected_all_disease_terms,
            pattern = vec_to_grep_pattern(c("disease"))
                          ),
         collected_all_disease_terms)
  )

```

### ICD10 R69: Illness, unspecified 

```{r icd10_r69}

# if DISEASE/TRAIT contains ICD10 R69, remove disease from collected_all_disease_terms

gwas_study_info  = 
 gwas_study_info |>
  mutate(collected_all_disease_terms  = 
          ifelse(grepl("ICD10 R69", `DISEASE/TRAIT`, ignore.case = TRUE) & 
                   collected_all_disease_terms == "disease",
         stringr::str_remove_all(
           collected_all_disease_terms,
            pattern = vec_to_grep_pattern(c("illness", "unspecified", "disease"))
                          ),
         collected_all_disease_terms)
  )


```

## PheCode 1019 - unspecified illness 

```{r phecode_1019}

# if DISEASE/TRAIT contains PheCode 1019, remove disease from collected_all_disease_terms
gwas_study_info  = 
 gwas_study_info |>
  mutate(collected_all_disease_terms  = 
          ifelse(grepl("PheCode 1019", `DISEASE/TRAIT`, ignore.case = TRUE) &
                   collected_all_disease_terms == "disease",
         stringr::str_remove_all(
           collected_all_disease_terms,
            pattern = vec_to_grep_pattern(c("illness", "unspecified", "disease"))
                          ),
         collected_all_disease_terms)
  )



```

## Number of non-cancer illnesses

```{r n_non_cancer_illnesses}


gwas_study_info  = 
 gwas_study_info |>
  mutate(collected_all_disease_terms  = 
          ifelse(grepl("Number of non-cancer illnesses", `DISEASE/TRAIT`, ignore.case = TRUE) & 
                   collected_all_disease_terms == "disease",
         stringr::str_remove_all(
           collected_all_disease_terms,
            pattern = vec_to_grep_pattern(c("illness", "unspecified", "disease"))
                          ),
         collected_all_disease_terms)
  )


```

## Other unspecified disease terms

```{r other_unspecified_disease_terms}

# Unknown and unspecified causes of morbidity
gwas_study_info  = 
 gwas_study_info |>
  mutate(collected_all_disease_terms  = 
          ifelse(grepl("Unknown and unspecified causes of morbidity", `DISEASE/TRAIT`, ignore.case = TRUE) & 
                   collected_all_disease_terms == "disease",
         stringr::str_remove_all(
           collected_all_disease_terms,
            pattern = vec_to_grep_pattern(c("illness", "unspecified", "disease"))
                          ),
         collected_all_disease_terms)
  )


# Non cancer illness - year age first occurred
gwas_study_info  = 
 gwas_study_info |>
  mutate(collected_all_disease_terms  = 
          ifelse(grepl("Non cancer illness - year age first occurred", `DISEASE/TRAIT`, ignore.case = TRUE) & 
                   collected_all_disease_terms == "disease",
         stringr::str_remove_all(
           collected_all_disease_terms,
            pattern = vec_to_grep_pattern(c("illness", "unspecified", "disease"))
                          ),
         collected_all_disease_terms)
  )


# ICD10 Z87.8: Personal history of other specified conditions
gwas_study_info  = 
 gwas_study_info |>
  mutate(collected_all_disease_terms  = 
          ifelse(grepl("ICD10 Z87.8: Personal history of other specified conditions", `DISEASE/TRAIT`, ignore.case = TRUE) & 
                   collected_all_disease_terms == "disease",
         stringr::str_remove_all(
           collected_all_disease_terms,
            pattern = vec_to_grep_pattern(c("illness", "unspecified", "disease"))
                          ),
         collected_all_disease_terms)
  )

# Number of non-cancer illnesses
gwas_study_info  = 
 gwas_study_info |>
  mutate(collected_all_disease_terms  = 
          ifelse(grepl("Number of non-cancer illnesses", `DISEASE/TRAIT`, ignore.case = TRUE) & 
                   collected_all_disease_terms == "disease",
         stringr::str_remove_all(
           collected_all_disease_terms,
            pattern = vec_to_grep_pattern(c("illness", "unspecified", "disease"))
                          ),
         collected_all_disease_terms)
  )

#  Disease burden
gwas_study_info  = 
 gwas_study_info |>
  mutate(collected_all_disease_terms  = 
          ifelse(grepl("Disease burden", `DISEASE/TRAIT`, ignore.case = TRUE) & 
                   collected_all_disease_terms == "disease",
         stringr::str_remove_all(
           collected_all_disease_terms,
            pattern = vec_to_grep_pattern(c("illness", "unspecified", "disease"))
                          ),
         collected_all_disease_terms)
  )

# Diagnosed with life threatening illness 
gwas_study_info  = 
 gwas_study_info |>
  mutate(collected_all_disease_terms  = 
          ifelse(grepl("Diagnosed with life threatening illness", `DISEASE/TRAIT`, ignore.case = TRUE) & 
                   collected_all_disease_terms == "disease",
         stringr::str_remove_all(
           collected_all_disease_terms,
            pattern = vec_to_grep_pattern(c("illness", "unspecified", "disease"))
                          ),
         collected_all_disease_terms)
  )

# Interpolated Age of participant when non cancer illness 
gwas_study_info  = 
 gwas_study_info |>
  mutate(collected_all_disease_terms  = 
          ifelse(grepl("Interpolated Age of participant when non cancer illness", `DISEASE/TRAIT`, ignore.case = TRUE) & 
                   collected_all_disease_terms == "disease",
         stringr::str_remove_all(
           collected_all_disease_terms,
            pattern = vec_to_grep_pattern(c("illness", "unspecified", "disease"))
                          ),
         collected_all_disease_terms)
  )

# illness, injury, bereavement, stress in last 2 years
gwas_study_info  = 
 gwas_study_info |>
  mutate(collected_all_disease_terms  = 
          ifelse(grepl("Illness, injury, bereavement, stress in last 2 years", `DISEASE/TRAIT`, ignore.case = TRUE) & 
                   collected_all_disease_terms == "disease",
         stringr::str_remove_all(
           collected_all_disease_terms,
            pattern = vec_to_grep_pattern(c("illness", "unspecified", "disease"))
                          ),
         collected_all_disease_terms)
  )

# ICD10 Z87.89: Personal history of other specified conditions
gwas_study_info  = 
 gwas_study_info |>
  mutate(collected_all_disease_terms  = 
          ifelse(grepl("ICD10 Z87.89: Personal history of other specified conditions", `DISEASE/TRAIT`, ignore.case = TRUE) & 
                   collected_all_disease_terms == "disease",
         stringr::str_remove_all(
           collected_all_disease_terms,
            pattern = vec_to_grep_pattern(c("illness", "unspecified", "disease"))
                          ),
         collected_all_disease_terms)
  )

# ICD10 Z86: Personal history of certain other diseases
gwas_study_info  = 
 gwas_study_info |>
  mutate(collected_all_disease_terms  = 
          ifelse(grepl("ICD10 Z86: Personal history of certain other diseases", `DISEASE/TRAIT`, ignore.case = TRUE) & 
                   collected_all_disease_terms == "disease",
         stringr::str_remove_all(
           collected_all_disease_terms,
            pattern = vec_to_grep_pattern(c("illness", "unspecified", "disease"))
                          ),
         collected_all_disease_terms)
  )


# ICD10 Z87.828.: Personal history of other (healed) physical injury and trauma
gwas_study_info  = 
 gwas_study_info |>
  mutate(collected_all_disease_terms  = 
          ifelse(grepl("Personal history of other \\(healed\\) physical injury and trauma", `DISEASE/TRAIT`, ignore.case = TRUE) & 
                   collected_all_disease_terms == "disease",
         stringr::str_remove_all(
           collected_all_disease_terms,
            pattern = vec_to_grep_pattern(c("illness", "unspecified", "disease"))
                          ),
         collected_all_disease_terms)
  )

# Other serious medical condition/disability diagnosed by doctor
gwas_study_info  = 
 gwas_study_info |>
  mutate(collected_all_disease_terms  = 
          ifelse(grepl("Other serious medical condition/disability diagnosed by doctor", `DISEASE/TRAIT`, ignore.case = TRUE) & 
                   collected_all_disease_terms == "disease",
         stringr::str_remove_all(
           collected_all_disease_terms,
            pattern = vec_to_grep_pattern(c("illness", "unspecified", "disease"))
                          ),
         collected_all_disease_terms)
  )

```


### ? ICD10 Z86.6: Personal history of diseases of the nervous system and sense organs (? doesn't fit nicely in gbd so remove)

```{r icd10_z86.6}

# if DISEASE/TRAIT contains ICD10 Z86.6, remove disease from collected_all_disease_terms
gwas_study_info  = 
 gwas_study_info |>
  mutate(collected_all_disease_terms  = 
          ifelse(grepl("ICD10 Z86.6", `DISEASE/TRAIT`, ignore.case = TRUE) & 
                   collected_all_disease_terms == "disease",
         stringr::str_remove_all(
           collected_all_disease_terms,
            pattern = vec_to_grep_pattern(c("illness", "unspecified", "disease"))
                          ),
         collected_all_disease_terms)
  )

```


### Fix diseases that should be specified: 

```{r fix_diseases}

diseases_to_fix = c("calculus of lower urinary tract" = "lower urinary tract calculus",
                    "urinary calculus" = "urinary calculus",
                    "miscarriage; stillbirth" = "miscarriage, stillbirth",
                    "chronic ischemic heart disease" = "chronic ischemic heart disease",
                    "pericarditis" = "pericarditis",
                    "retention of urine" = "urinary retention",
                    "back pain" = "back pain",
                    "prostatitis" = "prostatitis",
                    "dysuria" = "dysuria",
                    "renal colic" = "renal colic",
                    "cervicitis and endocervicitis" = "cervicitis, endocervicitis",
                    "edema" = "edema",
                    "hypertensive chronic kidney disease" = "hypertensive chronic kidney disease",
                    "substance abuse" = "substance abuse",
                    "pyogenic arthritis" = "pyogenic arthritis",
                    "hematuria" = "hematuria",
                    "pilonidal cyst" = "pilonidal cyst",
                    "atopic/contact dermatitis" = "atopic dermatitis, contact dermatitis",
                    "arterial embolism and thrombosis" = "arterial embolism, arterial thrombosis",
                    "functional disorders of bladder" = "urinary disorder",
                    "spondylosis" = "spondylosis",
                    "abnormal involuntary movements" = "abnormal involuntary movements",
                    "degenerative disease of the spinal cord" = "degenerative disease of the spinal cord",
                    "voice disturbance" = "voice disturbance",
                    "diseases of nail" = "nail disorder",
                    "visual disturbances" = "visual disturbances",
                    "late pregnancy and failed induction" = "late pregnancy, failed induction",
                    "other diseases of respiratory system" = "respiratory system disease",
                    "malposition and malpresentation of fetus or obstruction" = "obstructed labor",
                    "Hemorrhage in early pregnancy" = "early pregnancy hemorrhage",
                    "Chronic pharyngitis and nasopharyngitis" = "chronic pharyngitis, nasopharyngitis",
                    "Chronic tonsillitis and adenoiditis" = "chronic tonsillitis, adenoiditis",
                    "fetal abnormality affecting management of mother" = "fetal abnormality",
                    "Other disorders of synovium and tendon" = "synovium and tendon disorder",
                    "Noninfectious disorders of lymphatic channels" = "lymphatic system disorder",
                    "Noninflammatory disorders of cervix" = "cervical disorder",
                    "Noninflammatory disorders of ovary, fallopian tube, and broad ligament" = "noninflammatory disorder of ovary fallopian tube and broad ligament",
                    "Other inflammatory spondylopathies" = "inflammatory spondylopathy",
                    "Inflammatory disease of breast" = "inflammatory breast disease",
                    "Lump or mass in breast" = "lump or mass in breast",
                    "Corns and callosities" = "keratosis",
                    "Other acquired musculoskeletal deformity" = "acquired musculoskeletal deformity",
                    "Other derangement of joint" = "joint derangement",
                    "Other and unspecified disc disorder" = "disc disorder",
                    "Internal derangement of knee" = "joint derangement",
                    "Diseases of the musculoskeletal system and connective tissue" = "musculoskeletal diseases",
                    "diseases of the blood and blood-forming organs and certain disorders involving the immune mechanism" = "blood disorder",
                    "endocrine, nutritional and metabolic diseases" = "endocrine and metabolic disease",
                    "congenital malformations" = "congenital malformations",
                    "Localized superficial swelling" = "localized superficial swelling or mass"
                    )

# if collected_all_disease_terms == disease,
# but grep one of diseases_to_fix in DISEASE/TRAIT, replace disease with specific disease
# in collected_all_disease_terms


for(specifc_disease in 1:length(diseases_to_fix)){
  
  grep_pattern = names(diseases_to_fix[specifc_disease])
  replacement = diseases_to_fix[specifc_disease]
  
  gwas_study_info  = 
   gwas_study_info |>
    mutate(collected_all_disease_terms  = 
            ifelse(grepl(grep_pattern, `DISEASE/TRAIT`, ignore.case = TRUE) &
                   collected_all_disease_terms == "disease",
           stringr::str_replace_all(
             collected_all_disease_terms,
              pattern = vec_to_grep_pattern(c("disease")),
              replacement = replacement
                            ),
           collected_all_disease_terms)
    )
  
}


```



## Foot deformity

```{r foot_abnormality}

url <- "http://www.ebi.ac.uk/ols4/api/ontologies/mesh/terms/http%253A%252F%252Fid.nlm.nih.gov%252Fmesh%252FD005530/descendants"

foot_abnormality_terms <- get_descendants(url)

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_replace_all(collected_all_disease_terms,
                          pattern = vec_to_grep_pattern(foot_abnormality_terms),
                          "foot deformity"
         ))

```



# Other phenotypes to remove

## Widows peak 

```{r widows_peak}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_remove_all(collected_all_disease_terms,
                          pattern = vec_to_grep_pattern(c("widows peak"))
         ))


```

## Cough (non-specific)

```{r cough}

gwas_study_info = gwas_study_info |>
  mutate(collected_all_disease_terms  = 
         stringr::str_remove_all(collected_all_disease_terms,
                          pattern = vec_to_grep_pattern(c("cough"))
         ))

```


## Final summary - number of unique study terms

```{r}

n_studies_trait = gwas_study_info |>
  dplyr::filter(DISEASE_STUDY == T) |>
  dplyr::select(collected_all_disease_terms, PUBMED_ID) |>
  dplyr::distinct() |>
  dplyr::group_by(collected_all_disease_terms) |>
  dplyr::summarise(n_studies = dplyr::n()) |>
  dplyr::arrange(desc(n_studies))

head(n_studies_trait)

dim(n_studies_trait)

```

### When separate studies with multiple terms 

```{r}

diseases <- stringr::str_split(pattern = ", ", 
 gwas_study_info$collected_all_disease_terms[gwas_study_info$collected_all_disease_terms != ""])  |> 
            unlist() |>
            stringr::str_trim()

length(unique(diseases))

```

## Save output

```{r save_output}

fwrite(gwas_study_info,
here::here("output/gwas_cat/gwas_study_info_disease_trait_filtered_v2.csv"))

```
