---
title: "Final preparation of sentence training set"
author: "Isobel Beasley"
date: "2026-04-22"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

```{r setup_packages, message=FALSE, warning=FALSE}

library(dplyr)
library(jsonlite)
library(tidyr)
library(purrr)
library(stringr)

```

## Processing and cleaning doccano annotations

```{r load_doccano_annotations}

df <- stream_in(file(here::here("output/doccano/doccano_test_group.jsonl")))

# in case of multiple uploads of the same sentences to docanno, deduplicate by pubmed_id and text
df <-
 df |>
  select(-id) |>
  distinct()

```


```{r clean_doccano_annotations}

# make each entity annotation a new line
df_long <- df |>
  mutate(label = map(label, ~{
    if (length(.x) == 0) return(NULL)
    as.data.frame(.x) |>
      setNames(c("start", "end", "label"))
  })) |>
  unnest(label, keep_empty = T) |>
  mutate(across(c(start, end), as.integer))

df_entities <-
df_long |> select(-Comments) |>
  mutate(entity_text = stringr::str_sub(text, 
                                        start + 1, 
                                        end)) |>
  mutate(entity_text = ifelse(is.na(entity_text), 
                              "", 
                              entity_text))

df_entities <-
  df_entities |>
  distinct()

# fix for whitespace problems 
# if white-space at the end of the entity, adjust the end position
# if white-space at the start of the entity, adjust the start position
df_entities <- df_entities |>
  mutate(
    end = if_else(grepl(" $", entity_text), end - 1L, end),
    start = if_else(grepl("^ ", entity_text), start + 1L, start)
  ) |>
  mutate(entity_text = stringr::str_sub(text, start + 1, end)) |>
  mutate(entity_text = ifelse(is.na(entity_text), "", entity_text))

```

## Selecting negative (control sentences)

### Number of positives (sentences / annotated entities)


```{r number_of_positives}

df_entities |>
  group_by(text) |>
  summarise(has_entity = any(entity_text != "")) |>
  distinct() |>
  group_by(has_entity) |>
  summarise(n = n())

entity_texts <- df_entities |>
  filter(entity_text != "") |>
  pull(text) |>
  unique()

n_has_entities <- length(entity_texts)

print("Number of sentences containing an entity")
n_has_entities

print('Number of papers with at least one annotated entities')
df_entities |>
  filter(entity_text != "") |>
  select(pubmed_id, text) |>
  group_by(pubmed_id) |>
  summarise(n = n())

```


### Updated approach 

#### Sampling non-entity sentences that contain the same text as entities

Start with sampling sentences that contain the same text as entities, but not labelled entity

```{r}

training_entities <-
df_entities |>
  pull(entity_text) |> 
  unique()

training_entities <- training_entities[training_entities != ""]

same_as_entities <-
df_entities |>
 group_by(text) |>
  summarise(has_entity = any(entity_text != "")) |> 
  filter(has_entity == F) |>
  mutate(new_entity_text = stringr::str_extract_all(pattern = paste0(training_entities, 
                                                                     collapse = "|"),
                                                    text)
  ) 


same_as_entities <-
  same_as_entities |>
  rowwise() |>
  mutate(new_entity_text = stringr::str_flatten(unique(unlist(new_entity_text)), 
                                                collapse = "")
  )

same_as_entities_sentences <-
same_as_entities |>
  filter(new_entity_text != "") |>
  pull(text)

print("Number of null sentences that contain same words as positive sentences")
length(same_as_entities_sentences)

```

#### Sampling the most semantically similar sentences

```{r}

entity_json_list <- 
entity_texts |>
  as.list()

# Write to JSON file
writeLines(toJSON(entity_json_list,
                  auto_unbox = TRUE,
                  pretty = TRUE),
           here::here("output/doccano/training_cohort_sentences.json"))

non_entity_json_list <-
df_entities |>
  filter(entity_text == "") |>
  pull(text) |>
  unique() |>
  as.list()

writeLines(toJSON(non_entity_json_list,
                  auto_unbox = TRUE,
                  pretty = TRUE),
           here::here("output/doccano/training_non_cohort_sentences.json"))


```


```{bash, eval = F}

source  spacyr_venv/bin/activate

python3 code/extract_text/sentence_embeddings.py \
        --label "training" \
        --cohort_path "output/doccano/training_cohort_sentences.json" \
        --non_cohort_path "output/doccano/training_non_cohort_sentences.json" \
        --figure_save_path "output/doccano/training_sentence_similarities.png" \
        --p 75
        
```


```{r}

sem_similar_sentences <-
jsonlite::fromJSON(here::here("output/doccano/hard_negatives_training.json")) 

print("Nuber of non-cohort sentences selected for training based on semantic similarity")
length(sem_similar_sentences)

```

### Add number of participants labelling: 

```{r}

participant_words <- c("adults", "adolescents", "children", "infants", 
                       "neonates",
                       "women", "men", "female", "male","trios", 
                       "families", "twins", "siblings", "pairs",
                        "cases", "proband", 
                        "controls",                        
                        "individuals", "patients",  "participants", 
                       "subjects", "samples", "donors")

# sort by length 
participant_words <- participant_words[order(-nchar(participant_words))]

sample_size_words <- c("n", 
                       "N", 
                       "Neffective", "Neff",
                       "Ncontrol", "Ncase", "ncase", "n_total")

sample_size_words <- sample_size_words[order(-nchar(sample_size_words ))]


#num <- "[0-9]{1,3}(?:[,\\s][0-9]{3})*"
num <- "([0-9]+\\s+[0-9]+|[0-9,]+)"

participant_regex <- paste0(
  "\\b",
  num,
  "\\s+(",
  paste(participant_words, collapse="|"),
  ")\\b"
)

n_participants_regex <- paste0(
  "\\b", 
  "(",
  paste(sample_size_words, collapse="|"),
  ")",
  "\\s*=\\s*",
  num
)
  

all_participant_regex <- paste0(
  "(",
  paste(c(
    n_participants_regex,
    participant_regex
  ), collapse = "|"),
  ")"
)

df_n_entities <-
df_entities |>
  select(-start, -end, entity_text) |>
  #filter(grepl("Briefly, samples with call-rate", text)) |>
  distinct() |>
  #head() |>
  mutate(participant_positions = str_locate_all(text, all_participant_regex)) |>
  unnest_longer(participant_positions)

df_n_entities$start <- df_n_entities$participant_positions[,1]
df_n_entities$end <- df_n_entities$participant_positions[,2]

df_n_entities <-
  df_n_entities |>
  select(-participant_positions) |>
  mutate(label = "N_SAMPLE") |>
  mutate(entity_text = stringr::str_sub(text, 
                                        start, 
                                        end)) |>
  mutate(entity_text = ifelse(is.na(entity_text), 
                              "", 
                              entity_text)) |>
  filter(grepl("[0-9]", entity_text))

df_n_entities <- 
df_n_entities |>
  mutate(entity_number = str_extract(entity_text, num)) |>
  mutate(entity_number = str_remove(entity_number, ",$")) |>
  rowwise() |>
  mutate(participant_positions = str_locate_all(text, entity_number)) |>
  unnest_longer(participant_positions)
         
df_n_entities$start <- df_n_entities$participant_positions[,1]
df_n_entities$end <- df_n_entities$participant_positions[,2]

df_n_entities <-
  df_n_entities |>
  select(-participant_positions, -entity_number) |>
  mutate(label = "N_SAMPLE") |>
  mutate(entity_text = stringr::str_sub(text, 
                                        start, 
                                        end)) |>
  mutate(entity_text = ifelse(is.na(entity_text), 
                              "", 
                              entity_text)) 

too_low_values <- c(0:10,
                    "2,")

df_n_entities <-
  df_n_entities |>
  filter(entity_text != ",") |>
  filter(!c(entity_text %in% too_low_values)) |>
  mutate(start = start - 1) |>
  distinct()

df_entities <-
  bind_rows(df_n_entities,
            df_entities)

```

```{r}


n_non_cohort_sentences <-
df_entities |>
  filter(text %in% c(sem_similar_sentences, same_as_entities_sentences)) |>
  pull(text) |>
  unique() |>
  length()

n_cohort_sentences <- 
df_entities |> 
  filter(entity_text != "") |>
  pull(text) |>
  unique() |>
  length() 

print(paste0("Number of non-cohort sentences selected for training: ", n_non_cohort_sentences))
print(paste0("Number of cohort sentences selected for training: ", n_cohort_sentences))

print("Percentage of non-cohort sentences selected for training:")
print(round(100 * n_non_cohort_sentences / (n_non_cohort_sentences + n_cohort_sentences), 2))


```

### Save training set

```{r}

df_entities <-
  df_entities |>
  filter(text %in% c(sem_similar_sentences, same_as_entities_sentences) | entity_text != "")

# save the cleaned entities to a new jsonl file
build_label_list <- function(start, end, label) {
  keep <- !is.na(start) & !is.na(end) & !is.na(label)
  if (!any(keep)) return(list())
  pmap(list(start[keep], end[keep], label[keep]),
       \(s, e, l) list(s, e, l))
}

df_doc <- df_entities |>
  group_by(#id, 
           text, pubmed_id, date, country, gwas_cat_cohort_label) |>
  summarise(label = list(build_label_list(start, end, label)),
            .groups = "drop") |>
  mutate(
    # Doccano always has these as strings; cast defensively
    pubmed_id             = as.character(pubmed_id),
    date                  = as.character(date),
    country               = as.character(country),
    gwas_cat_cohort_label = as.character(gwas_cat_cohort_label),
    Comments              = list(list())
  )


write_doccano_jsonl <- function(df_doc, out_path) {
  dir.create(dirname(out_path), recursive = TRUE, showWarnings = FALSE)

  con <- file(out_path, "w", encoding = "UTF-8")
  on.exit(close(con), add = TRUE)   # now correctly scoped to this function

  for (i in seq_len(nrow(df_doc))) {
    tryCatch({
      obj <- list(
        text                  = df_doc$text[[i]],
        pubmed_id             = df_doc$pubmed_id[[i]],
        date                  = df_doc$date[[i]],
        country               = df_doc$country[[i]],
        gwas_cat_cohort_label = df_doc$gwas_cat_cohort_label[[i]],
        label                 = df_doc$label[[i]]
      )

      json <- jsonlite::toJSON(obj,
                               auto_unbox = TRUE,
                               null       = "null",
                               na         = "null")

      # jsonlite serialises empty list() as {}; force [] for these two fields
      json <- gsub('"label":{}',    '"label":[]',    json, fixed = TRUE)
      json <- gsub('"Comments":{}', '"Comments":[]', json, fixed = TRUE)

      writeLines(as.character(json), con, useBytes = TRUE)
    }, error = function(e) {
      message(sprintf("row %d (id=%s) failed: %s",
                      i, df_doc$id[[i]], conditionMessage(e)))
    })
  }

  invisible(out_path)
}




write_doccano_jsonl(df_doc, 
                    here::here("output/doccano/abstracts_with_cohort_info_reconstructed.jsonl"))

```


## Save cohort annotations per pubmed id:

```{r}

# save cohort annotations per pubmed id:
cohort_annotations_per_pubmed <-
df_entities |>
  filter(entity_text != "") |>
  select(pubmed_id, text, entity_text)


source(here::here("code/merge_cohort_names.R"))
# standardize cohort names
cohort_group <-
  matched |>
  select(cohort, full_name, synonyms, group_id) |>
  tidyr::pivot_longer(cols = c(cohort, full_name, synonyms),
                      names_to = "field",
                      values_to = "COHORT") |>
  filter(!is.na(COHORT)) |>
  select(-field) |>
  distinct()

cohort_group <-
  cohort_group |>
  group_by(group_id) |>
  # group name is the shortest cohort name in the group (after trimming whitespace)
  mutate(group_name = COHORT[which.min(nchar(str_trim(COHORT)))][1]) |>
  ungroup() |>
  select(COHORT, group_name) |>
  #mutate(across(everything(), ~ tolower(.))) |>
  mutate(across(everything(), ~stringr::str_replace_all(., "\\s*-\\s*", "-"))) |>
  distinct()


cohort_annotations_per_pubmed <-
  left_join(cohort_annotations_per_pubmed,
            cohort_group |> rename(entity_text = COHORT),
            by = "entity_text") |>
  mutate(group_name = ifelse(is.na(group_name),
                             entity_text, group_name)) |>
  select(-entity_text) |>
  rename(cohort = group_name)


cohort_annotations_per_pubmed <-
  cohort_annotations_per_pubmed |>
  group_by(pubmed_id) |>
  summarise(cohort_annotations = str_flatten(sort(unique(cohort)), collapse = "; "))

out_annotations <- "output/gwas_cat/training_set_cohort_annotations.csv"

data.table::fwrite(cohort_annotations_per_pubmed,
                   file = out_annotations,
                   sep = ",")


```
