---
title: "Final preparation of sentence training set"
author: "Isobel Beasley"
date: "2026-04-22"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r setup_packages}

library(dplyr)
library(jsonlite)
library(tidyr)
library(purrr)

```

## Processing and cleaning doccano annotations

```{r load_doccano_annotations}

df <- stream_in(file(here::here("output/doccano/doccano_pt7_v5.jsonl")))

# in case of multiple uploads of the same sentences to docanno, deduplicate by pubmed_id and text
df  <- df |>
  group_by(pubmed_id, text) |>
  slice_sample(n = 1) |>
  ungroup()

```


```{r clean_doccano_annotations}

# make each entity annotation a new line
df_long <- df |>
  mutate(label = map(label, ~{
    if (length(.x) == 0) return(NULL)
    as.data.frame(.x) |>
      setNames(c("start", "end", "label"))
  })) |>
  unnest(label, keep_empty = T) |>
  mutate(across(c(start, end), as.integer))

df_entities <-
df_long |> select(-Comments) |>
  mutate(entity_text = stringr::str_sub(text, 
                                        start + 1, 
                                        end)) |>
  mutate(entity_text = ifelse(is.na(entity_text), 
                              "", 
                              entity_text))

df_entities <-
  df_entities |>
  distinct()

# fix for whitespace problems 
# if white-space at the end of the entity, adjust the end position
# if white-space at the start of the entity, adjust the start position
df_entities <- df_entities |>
  mutate(
    end = if_else(grepl(" $", entity_text), end - 1L, end),
    start = if_else(grepl("^ ", entity_text), start + 1L, start)
  ) |>
  mutate(entity_text = stringr::str_sub(text, start + 1, end)) |>
  mutate(entity_text = ifelse(is.na(entity_text), "", entity_text))

```


## Selecting negative (control sentences)

### Number of positives (sentences / annotated entities)


```{r number_of_positives}

df_entities |>
  group_by(text) |>
  summarise(has_entity = any(entity_text != "")) |>
  distinct() |>
  group_by(has_entity) |>
  summarise(n = n())

entity_texts <- df_entities |>
  filter(entity_text != "") |>
  pull(text) |>
  unique()

n_has_entities <- length(entity_texts)

print("Number of sentences containing an entity")
n_has_entities

print('Number of papers with at least one annotated entities')
df_entities |>
  filter(entity_text != "") |>
  select(pubmed_id, text) |>
  group_by(pubmed_id) |>
  summarise(n = n())

```


#### Initial approach - just random sampling

```{r random_sample_train, eval=F}

set.seed(40)

keep_non_entity_text <-
df_entities |>
  group_by(text) |>
  summarise(has_entity = any(entity_text != "")) |>
  filter(!has_entity) |>
  slice_sample(n = ceiling(0.2 * n_has_entities)) |>
  pull(text)

df_entities <-
  df_entities |>
  filter(text %in% keep_non_entity_text | entity_text != "")


```


### Updated approach 

#### Sampling non-entity sentences that contain the same text as entities

Start with sampling sentences that contain the same text as entities, but not labelled entity

```{r}

training_entities <-
df_entities |>
  pull(entity_text) |> 
  unique()

training_entities <- training_entities[training_entities != ""]

same_as_entities <-
df_entities |>
 group_by(text) |>
  summarise(has_entity = any(entity_text != "")) |> 
  filter(has_entity == F) |>
  mutate(new_entity_text = stringr::str_extract_all(pattern = paste0(training_entities, 
                                                                     collapse = "|"),
                                                    text)
  ) 


same_as_entities <-
  same_as_entities |>
  rowwise() |>
  mutate(new_entity_text = stringr::str_flatten(unique(unlist(new_entity_text)), 
                                                collapse = "")
  )

same_as_entities_sentences <-
same_as_entities |>
  filter(new_entity_text != "") |>
  pull(text)

print("Number of null sentences that contain same words as positive sentences")
length(same_as_entities_sentences)

```

#### Sampling the most semantically similar sentences

```{r}

entity_json_list <- 
entity_texts |>
  as.list()

# Write to JSON file
writeLines(toJSON(entity_json_list,
                  auto_unbox = TRUE,
                  pretty = TRUE),
           here::here("output/doccano/training_cohort_sentences.json"))

non_entity_json_list <-
df_entities |>
  filter(entity_text == "") |>
  pull(text) |>
  unique() |>
  as.list()

writeLines(toJSON(non_entity_json_list,
                  auto_unbox = TRUE,
                  pretty = TRUE),
           here::here("output/doccano/training_non_cohort_sentences.json"))


```


```{bash, eval = F}

source  spacyr_venv/bin/activate

python3 code/extract_text/sentence_embeddings.py \
        --label "training" \
        --cohort_path "output/doccano/training_cohort_sentences.json" \
        --non_cohort_path "output/doccano/training_non_cohort_sentences.json" \
        --figure_save_path "output/doccano/training_sentence_similarities.png" \
        --p 75
        
```

