---
title: "Stats for specific aims"
author: "Isobel Beasley"
date: "2025-10-23"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load_libraries, message=FALSE, warning=FALSE}

library(data.table)
library(dplyr)
library(ggplot2)

```

# Proportion of all published studies that have cohort meta-data

```{r original_cohort_info}

orig_gwas_study_info <- data.table::fread(here::here("data/gwas_catalog/gwas-catalog-v1.0.3.1-studies-r2025-07-21.tsv"))

orig_gwas_study_info = orig_gwas_study_info |>
  dplyr::rename_with(~ tolower(gsub(" ", "_", .x)))

# how many published studies (i.e. number of unique pubmed ids)
n_published_studies_orig <- orig_gwas_study_info |>
  dplyr::filter(!is.na(pubmed_id)) |>
  dplyr::pull(pubmed_id) |>
  unique() |>
  length()

print("Number of published studies in original GWAS Catalog:")
print(n_published_studies_orig)

# how many published studies include any cohort information?
n_studies_with_cohort_info_orig <- orig_gwas_study_info |>
  dplyr::filter(!is.na(cohort) & cohort != "") |>
  dplyr::pull(pubmed_id) |>
  unique() |>
  length()

print("Number of published studies with cohort information in original GWAS Catalog:")
print(n_studies_with_cohort_info_orig)

print("Proportion of published studies with cohort information in original GWAS Catalog:")
100 * n_studies_with_cohort_info_orig / n_published_studies_orig

studies_with_true_cohort_info_orig <- orig_gwas_study_info |>
  dplyr::mutate(cohort = stringr::str_remove_all(pattern = "(^NR$)|(^NR\\|)|(\\|NR$)|(\\|NR\\|)|(not reported)", 
                                                 cohort)) |>
  dplyr::mutate(cohort = stringr::str_remove_all(pattern = "other|multiple|Other|Multiple|Others|\\(Multiple cohorts\\)|OTHER", 
                                                 cohort)) |>
  dplyr::filter(!is.na(cohort) & 
                  cohort != "") |>
  dplyr::pull(pubmed_id) |>
  unique()

# how many published studies include any true cohort information 
# (i.e. not 'NR', or 'not reported', 'other', 'multiple')
n_studies_with_true_cohort_info_orig <- studies_with_true_cohort_info_orig |> 
  unique() |>
  length()

print("Number of published studies with true cohort information in original GWAS Catalog:")
print(n_studies_with_true_cohort_info_orig)

print("Proportion of published studies with true cohort information in original GWAS Catalog:")
100 * n_studies_with_true_cohort_info_orig / n_published_studies_orig

data.frame(n_studies = c(n_published_studies_orig - n_studies_with_cohort_info_orig, 
                         n_studies_with_cohort_info_orig - n_studies_with_true_cohort_info_orig,
                         n_studies_with_true_cohort_info_orig),
           type = c("all",
                    "studies_with_cohort_info",
                    "studies_with_true_cohort_info"),
           group = c("GWAS Catalog Published Studies")) |>
ggplot(aes(x = group, y = n_studies, fill = type)) +
  geom_col(position = "stack") +
  labs(x = "",
       y = "Number of studies",
       title = "Published GWAS Catalog Studies") +
  scale_y_continuous(
    breaks = seq(0, max(n_published_studies_orig), by = 1000),
    expand = expansion(mult = c(0, 0.05))
  ) + 
  scale_fill_manual(
    name = "Cohort metadata availability",
    values = c(
      all = "grey",
      studies_with_cohort_info = "black",
      studies_with_true_cohort_info = "#1b9e77"
    ),
    labels = c(
      all = "None",
      studies_with_cohort_info = "Uninformative metadata\n(NR, other, multiple)",
      studies_with_true_cohort_info = "At least one valid cohort name"
    )
  ) + 
  theme_minimal() +
  theme(
    axis.text.x  = element_blank(),
    axis.ticks.x = element_blank(),
    axis.text.y  = element_text(size = 20),
    axis.title.y = element_text(size = 24),
    legend.title = element_text(size = 24),
    legend.text  = element_text(size = 23),
    title        = element_text(size = 24)
  )



```

# Number of papers - for 15 conditions

```{r n_total_papers, message=FALSE, warning=FALSE}

#gwas_study_info <- data.table::fread(here::here("data/gwas_catalog/gwas-catalog-v1.0.3.1-studies-r2025-07-21.tsv"))
gwas_study_info <- data.table::fread(here::here("output/icd_map/gwas_study_gbd_causes.csv"))

gwas_study_info = gwas_study_info |>
  dplyr::rename_with(~ tolower(gsub(" ", "_", .x)))

# how many published studies (i.e. number of unique pubmed ids)
n_published_studies <- gwas_study_info |>
  dplyr::filter(!is.na(pubmed_id)) |>
  dplyr::pull(pubmed_id) |>
  unique() |>
  length()

print("Number of published studies in GWAS Catalog:")
print(n_published_studies)


```

# Number of papers (excluding infectious diseases)

```{r n_non_com_total_papers, message=FALSE, warning=FALSE}

#gwas_study_info <- data.table::fread(here::here("data/gwas_catalog/gwas-catalog-v1.0.3.1-studies-r2025-07-21.tsv"))
gwas_study_info <- data.table::fread(here::here("output/icd_map/gwas_study_gbd_causes.csv"))

gwas_study_info = gwas_study_info |>
  dplyr::rename_with(~ tolower(gsub(" ", "_", .x)))


gwas_non_comm_study_info = gwas_study_info |>
  dplyr::filter(!cause %in% c("HIV/AIDS",
                             "Tuberculosis",
                             "Malaria",
                             "Lower respiratory infections",
                             "Diarrhoeal diseases",
                             "Neonatal disorders",
                             "Tetanus",
                             "Diphtheria",
                             "Pertussis" ,
                             "Measles",
                             "Maternal disorders"))


gwas_non_comm_study_info$cause |> unique() |> sort()

# how many published studies (i.e. number of unique pubmed ids)
n_published_non_comm_studies <- gwas_non_comm_study_info |>
  dplyr::filter(!is.na(pubmed_id)) |>
  dplyr::pull(pubmed_id) |>
  unique() |>
  length()

n_published_non_comm_studies

```

# Number of studies for each condition

```{r n_papers_per_condition}

condition_counts <-
gwas_study_info |>
  dplyr::select(pubmed_id, cause) |>
  dplyr::distinct() |> 
  dplyr::group_by(cause) |>
  dplyr::summarise(n_studies = length(unique(pubmed_id))) |>
  dplyr::arrange(desc(n_studies))

lancet_grouping <- readxl::read_xlsx(here::here("data/icd/lancet_conditions_icd10.xlsx"))

lancet_grouping = lancet_grouping |>
  dplyr::select(cause = mapped_gbd_term, lancet_group)

condition_counts = condition_counts |>
  dplyr::left_join(lancet_grouping, by = "cause") |>
  dplyr::arrange(desc(n_studies))

# combine Liver cancer due to hepatitis B and Liver cancer due to hepatitis C
condition_counts = condition_counts |>
  dplyr::mutate(cause = ifelse(cause %in% c("Liver cancer due to hepatitis B",
                                           "Liver cancer due to hepatitis C"),
                              "Liver cancer",
                              cause)) |>
  dplyr::group_by(cause, lancet_group) |>
  dplyr::summarise(n_studies = sum(n_studies)) |>
  dplyr::arrange(desc(n_studies))

# exclude "Diabetes mellitus"
condition_counts = condition_counts |>
  dplyr::filter(cause != "Diabetes mellitus")

# rename Chronic kidney disease due to diabetes mellitus type 2
# Chronic kidney disease (type 2 diabetes)
condition_counts = condition_counts |>
  dplyr::mutate(cause = ifelse(cause == "Chronic kidney disease due to diabetes mellitus type 2",
                              "Chronic kidney disease (type 2 diabetes)",
                              cause))

# repeat for Chronic kidney disease due to diabetes mellitus type 1
condition_counts = condition_counts |>
  dplyr::mutate(cause = ifelse(cause == "Chronic kidney disease due to diabetes mellitus type 1",
                              "Chronic kidney disease (type 1 diabetes)",
                              cause))


# change lancent group names, for better display
# I-8 = Infectious diseases
# NCD-7 = Non-communicable diseases
condition_counts = condition_counts |>
  dplyr::mutate(lancet_group = ifelse(lancet_group == "I-8",
                                     "Infectious diseases",
                                     lancet_group)) |>
  dplyr::mutate(lancet_group = ifelse(lancet_group == "NCD-7",
                                     "Non-communicable diseases",
                                     lancet_group))

ggplot(aes(x = reorder(cause, n_studies), 
           fill = lancet_group,
           y = n_studies),
       data = condition_counts) +
  geom_col() +
  coord_flip() +
  labs(#x = "Condition",
       y = "Number of published studies") +
       #title = "Number of published GWAS Catalog studies per condition") +
  # facet_wrap(~lancet_group,
  #            nrow = 2,
  #            scales = "free_y") +
  facet_grid(space = "free_y",
             scale = "free_y",
             rows = vars(lancet_group)) + 
  scale_fill_discrete(type = c("skyblue", "orange")) + 
  theme_minimal() +
  theme(
    axis.text.x  = element_text(size = 20),
    axis.text.y  = element_text(size = 20),
    axis.title.x = element_text(size = 24),
    axis.title.y = element_blank(),
    title        = element_text(size = 24), 
    strip.text = element_text(size = 22),
    legend.position = "none"
  )

```

# Number of studies per condition - non_communicable diseases only

```{r n_papers_per_non-communicable_condition}

condition_counts <-
gwas_non_comm_study_info |>
  dplyr::select(pubmed_id, cause) |>
  dplyr::distinct() |> 
  dplyr::group_by(cause) |>
  dplyr::summarise(n_studies = length(unique(pubmed_id))) |>
  dplyr::arrange(desc(n_studies))

lancet_grouping <- readxl::read_xlsx(here::here("data/icd/lancet_conditions_icd10.xlsx"))

lancet_grouping = lancet_grouping |>
  dplyr::select(cause = mapped_gbd_term, lancet_group)

condition_counts = condition_counts |>
  dplyr::left_join(lancet_grouping, by = "cause") |>
  dplyr::arrange(desc(n_studies))

# combine Liver cancer due to hepatitis B and Liver cancer due to hepatitis C
condition_counts = condition_counts |>
  dplyr::mutate(cause = ifelse(cause %in% c("Liver cancer due to hepatitis B",
                                           "Liver cancer due to hepatitis C"),
                              "Liver cancer",
                              cause)) |>
  dplyr::group_by(cause, lancet_group) |>
  dplyr::summarise(n_studies = sum(n_studies)) |>
  dplyr::arrange(desc(n_studies))

# exclude "Diabetes mellitus"
condition_counts = condition_counts |>
  dplyr::filter(cause != "Diabetes mellitus")

# rename Chronic kidney disease due to diabetes mellitus type 2
# Chronic kidney disease (type 2 diabetes)
condition_counts = condition_counts |>
  dplyr::mutate(cause = ifelse(cause == "Chronic kidney disease due to diabetes mellitus type 2",
                              "Chronic kidney disease (type 2 diabetes)",
                              cause))

# repeat for Chronic kidney disease due to diabetes mellitus type 1
condition_counts = condition_counts |>
  dplyr::mutate(cause = ifelse(cause == "Chronic kidney disease due to diabetes mellitus type 1",
                              "Chronic kidney disease (type 1 diabetes)",
                              cause))


# change lancent group names, for better display
# I-8 = Infectious diseases
# NCD-7 = Non-communicable diseases
condition_counts = condition_counts |>
  dplyr::mutate(lancet_group = ifelse(lancet_group == "I-8",
                                     "Infectious diseases",
                                     lancet_group)) |>
  dplyr::mutate(lancet_group = ifelse(lancet_group == "NCD-7",
                                     "Non-communicable diseases",
                                     lancet_group))

ggplot(aes(x = reorder(cause, n_studies), 
           fill = lancet_group,
           y = n_studies),
       data = condition_counts) +
  geom_col() +
  coord_flip() +
  labs(#x = "Condition",
       y = "Number of published studies") +
       #title = "Number of published GWAS Catalog studies per condition") +
  # facet_wrap(~lancet_group,
  #            nrow = 2,
  #            scales = "free_y") +
  facet_grid(space = "free_y",
             scale = "free_y",
             rows = vars(lancet_group)) + 
  scale_fill_discrete(type = c("skyblue", "orange")) + 
  theme_minimal() +
  theme(
    axis.text.x  = element_text(size = 20),
    axis.text.y  = element_text(size = 20),
    axis.title.x = element_text(size = 24),
    axis.title.y = element_blank(),
    title        = element_text(size = 24), 
    strip.text = element_text(size = 22),
    legend.position = "none"
  )

```

# Relationship between causes and mapped traits

```{r causes_mapped_traits}

cause_mapped_trait_counts <-
gwas_study_info |>
  dplyr::select(cause, collected_all_disease_terms, pubmed_id) |>
  dplyr::distinct() |> 
  dplyr::group_by(cause, collected_all_disease_terms) |>
  dplyr::summarise(n_studies = length(unique(pubmed_id))) |>
  dplyr::arrange(cause, desc(n_studies))
  

cause_mapped_trait_counts |>
  dplyr::filter(cause == "Cervical cancer")

```

# Proportion include cohort information ( published papers in the GWAS catalog that )

```{r prop_cohort_info}

gwas_cohort <- data.table::fread(here::here("output/gwas_cohorts/gwas_cohort_name_corrected.csv"))

gwas_cohort = gwas_cohort |>
  dplyr::filter(PUBMED_ID %in% unique(gwas_study_info$pubmed_id))

# how many published studies include cohort information? 
n_studies_with_cohort_info <- gwas_cohort |>
  dplyr::filter(COHORT != "") |>
  dplyr::pull(PUBMED_ID) |>
  unique() |>
  length()

print("Number of published studies with cohort information:")
print(n_studies_with_cohort_info)

print("Proportion of published studies with cohort information:")
100 * n_studies_with_cohort_info / n_published_studies

```

# Proportion of non-communicable disease studies that include cohort information

```{r prop_cohort_info_non_comm}

gwas_non_comm_study_info = gwas_study_info |>
  dplyr::filter(!cause %in% c("HIV/AIDS",
                             "Tuberculosis",
                             "Malaria",
                             "Lower respiratory infections",
                             "Diarrhoeal diseases",
                             "Neonatal disorders",
                             "Tetanus",
                             "Diphtheria",
                             "Pertussis" ,
                             "Measles",
                             "Maternal disorders"))


gwas_cohort_non_comm = gwas_cohort |>
  dplyr::filter(PUBMED_ID %in% unique(gwas_non_comm_study_info$pubmed_id))

# how many published studies include cohort information?
n_studies_with_cohort_info_non_comm <- gwas_cohort_non_comm |>
  dplyr::filter(COHORT != "") |>
  dplyr::pull(PUBMED_ID) |>
  unique() |>
  length()

print("Number of published non-communicable disease studies with cohort information:")
n_studies_with_cohort_info_non_comm

print("Proportion of published non-communicable disease studies with cohort information:")
100 * n_studies_with_cohort_info_non_comm / n_published_non_comm_studies

```

# Proportion of studies with abstracts

```{r prop_abstracts}

downloaded_abstracts <-
 list.files(here::here("output/abstracts")) |> 
  stringr::str_remove_all(".txt$") 

n_studies_with_abstracts <-
gwas_study_info |>
  dplyr::filter(pubmed_id %in% downloaded_abstracts) |>
  dplyr::pull(pubmed_id) |>
  unique() |>
  length()

print("Number of published studies with abstracts downloaded:")
print(n_studies_with_abstracts)

print("Proportion of published studies with abstracts downloaded:")
100 * n_studies_with_abstracts / n_published_studies

print("Number of published studies without abstracts downloaded:")
print(n_published_studies - n_studies_with_abstracts)

```

# Year number of papers per condition

```{r papers_per_year}

gwas_study_info |>
  select(cause, year, pubmed_id) |> 
  distinct() |> 
  ggplot(aes(x = year)) +
  geom_histogram() + 
  facet_wrap(~cause)


gwas_study_info |>
  select(cause, year, pubmed_id) |> 
  distinct() |>
  group_by(cause) |>
  summarise(min_year = min(year),
            first_quartile_year = quantile(year, 0.25),
            median_year = median(year),
            average_year = mean(year),
            third_quartile_year = quantile(year, 0.75),
            max_year = max(year)
            ) 

```

# Proportion of studies with abstracts - non-communicable diseases only

```{r prop_abstracts_non_comm}

n_studies_with_abstracts_non_comm <-
gwas_non_comm_study_info   |>
   dplyr::rename_with(~ tolower(gsub(" ", "_", .x))) |>
  dplyr::filter(pubmed_id %in% downloaded_abstracts) |>
  dplyr::pull(pubmed_id) |>
  unique() |>
  length()
  
print("Number of published non-communicable disease studies with abstracts downloaded:")
print(n_studies_with_abstracts_non_comm)

print("Proportion of published non-communicable disease studies with abstracts downloaded:")
100 * n_studies_with_abstracts_non_comm / n_published_non_comm_studies


```

# Proportion of studies with full text: 

```{r prop_full_text}

downloaded_fulltexts <-
list.files(c(here::here("output/fulltexts/ncbi_cloud"),
              here::here("output/fulltexts/europe_pmc")
             
             )
) |>
  stringr::str_remove_all(".xml|.pdf")

convert_pmid_df <- data.table::fread(here::here("data/europe_pmc/PMID_PMCID_DOI.csv"))

downloaded_fulltexts <-
convert_pmid_df |>
  dplyr::filter(PMCID %in% downloaded_fulltexts) |>
  dplyr::pull(PMID) |> 
  unique() 

other_downloaded_fulltexts <-
  list.files(c(here::here("output/fulltexts/springer_nature"),
             here::here("output/fulltexts/wiley"),
             here::here("output/fulltexts/elsevier"),
             here::here("output/fulltexts/oxford_academic/txt"),
             here::here("output/fulltexts/sage")
             )
             ) |>
  stringr::str_remove_all(".xml|.txt|.pdf")

downloaded_fulltexts <-
c(downloaded_fulltexts,
  other_downloaded_fulltexts
) |>
  unique()

n_studies_with_fulltext <-
gwas_study_info |>
  dplyr::filter(pubmed_id %in% downloaded_fulltexts) |>
  dplyr::pull(pubmed_id) |>
  unique() |>
  length()

print("Number of published studies with full text downloaded:")
print(n_studies_with_fulltext)

print("Proportion of published studies with full text downloaded:")
100 * n_studies_with_fulltext / n_published_studies

print("Number of published studies without full text downloaded:")
print(n_published_studies - n_studies_with_fulltext)

```


# Proportion of studies with full text - non-communicable diseases only

```{r prop_full_text_non_comm}

studies_with_fulltext_non_comm <-
gwas_non_comm_study_info |>
   dplyr::rename_with(~ tolower(gsub(" ", "_", .x))) |>
  dplyr::filter(pubmed_id %in% downloaded_fulltexts) |>
  dplyr::pull(pubmed_id) |>
  unique()

n_studies_with_fulltext_non_comm <-
gwas_non_comm_study_info |>
   dplyr::rename_with(~ tolower(gsub(" ", "_", .x))) |>
  dplyr::filter(pubmed_id %in% downloaded_fulltexts) |>
  dplyr::pull(pubmed_id) |>
  unique() |>
  length()

print("Number of published non-communicable disease studies with full text downloaded:")
print(n_studies_with_fulltext_non_comm)

print("Proportion of published non-communicable disease studies with full text downloaded:")
100 * n_studies_with_fulltext_non_comm / n_published_non_comm_studies

```

# Of not full-text, how many open acess by OpenAlex? 

```{r open_access_no_fulltext}

library(openalexR)

studies_without_fulltext <-
  gwas_non_comm_study_info |>
   dplyr::rename_with(~ tolower(gsub(" ", "_", .x))) |>
  filter(!(pubmed_id %in% downloaded_fulltexts)) |>
  pull(pubmed_id) |>
  unique()

doi_information <-
convert_pmid_df |>
  filter(PMID %in% studies_without_fulltext) |>
  pull(DOI)

doi_information = 
doi_information[doi_information != ""]

# get open alex works for pmids
open_alex_works <- oa_fetch(
  doi = doi_information,
  entity = "works",
  options = list(select = c("doi", 
                            "open_access"))
)

```

# Plot number of studies with abstracts or full text 

```{r plot_abstracts_fulltext}

data.frame(n_studies = c(n_published_studies - n_studies_with_abstracts,
                         n_studies_with_abstracts - n_studies_with_fulltext,
                         n_studies_with_fulltext),
           type = c("all",
                    "studies_with_abstracts",
                    "studies_with_fulltext"),
           group = c("GWAS Catalog Published Studies")) |>
ggplot(aes(x = group, y = n_studies, fill = type)) +
  geom_col(position = "stack") +
  labs(x = "",
       y = "Number of studies",
       title = "Published GWAS Catalog Studies") +
  # scale_y_continuous(
  #   breaks = seq(0, max(n_published_studies), by = 1000),
  #   expand = expansion(mult = c(0, 0.05))
  # ) + 
  scale_fill_manual(
    name = "Text availability",
    values = c(
      all = "grey",
      studies_with_abstracts = "black",
      studies_with_fulltext = "#7570b3"
    ),
    labels = c(
      all = "None",
      studies_with_abstracts = "Abstract only",
      studies_with_fulltext = "Full text + abstract"
    )
  ) + 
  theme_minimal() +
  theme(
    axis.text.x  = element_blank(),
    axis.ticks.x = element_blank(),
    axis.text.y  = element_text(size = 20),
    axis.title.y = element_text(size = 24),
    legend.title = element_text(size = 24),
    legend.text  = element_text(size = 23),
    title        = element_text(size = 24)
  )

```


# Studies without full text, what proportion have cohort info?

```{r no_fulltext_cohort_info}

studies_without_fulltext <-
setdiff(gwas_study_info$pubmed_id, downloaded_fulltexts)

n_studies_without_fulltext_with_cohort_info <-
gwas_cohort |>
  dplyr::filter(!c(PUBMED_ID %in% downloaded_fulltexts) &
                COHORT != "") |>
  dplyr::pull(PUBMED_ID) |>
  unique() |>
  length()

print("Number of published studies without full text downloaded that have cohort information:")
print(n_studies_without_fulltext_with_cohort_info)

print("Proportion of published studies without full text downloaded that have cohort information:")
100 * n_studies_without_fulltext_with_cohort_info / length(studies_without_fulltext)

```

# Plot % deaths explained by diseases (California, 2019)

```{r deaths_by_condition}


gbd_2019 <-
  data.table::fread(here::here("data/gbd/gbd_2019_california_percent_deaths.csv"))

gbd_2019 = gbd_2019 |>
  dplyr::mutate(val = 100 * val) |>
  dplyr::mutate(lower = 100 * lower) |>
  dplyr::mutate(upper = 100 * upper)

gbd_2019 =
  gbd_2019 |>
  dplyr::mutate(cause = ifelse(cause == "Diabetes mellitus type 2",
                               "Type 2 diabetes mellitus",
                               cause)) |>
  dplyr::mutate(cause = ifelse(cause == "Diabetes mellitus type 1",
                               "Type 1 diabetes mellitus",
                               cause)) |>
    dplyr::mutate(cause = ifelse(cause == "Chronic kidney disease due to diabetes mellitus type 2",
                               "Chronic kidney disease (type 2 diabetes)",
                               cause)) |>
      dplyr::mutate(cause = ifelse(cause == "Chronic kidney disease due to diabetes mellitus type 1",
                               "Chronic kidney disease (type 1 diabetes)",
                               cause)) 

# add row summing all conditions
gbd_2019 |>
  dplyr::summarise(total_val = sum(val))


ggplot(aes(x = reorder(cause, val), 
           y = val),
       data = gbd_2019) +
  geom_col(fill = "firebrick") +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.3) +
  coord_flip() +
  labs(x = "",
       y = "% Percentage of all deaths (California, 2019)",
       #title = "Percentage of deaths by condition in California (2019)"
       ) +
    theme_minimal(base_family = "Arial") +
  # theme(
  #   axis.text.x  = element_text(size = 8),   # 8pt for axis text
  #   axis.text.y  = element_text(size = 8),   # 8pt for axis text
  #   axis.title.x = element_text(size = 9),  # 10pt for labels
  #   axis.title.y = element_blank(),
  #   title        = element_text(size = 9),  # 10pt for labels
  #   strip.text   = element_text(size = 9),  # 10pt for labels
  #   legend.title = element_text(size = 8),  # 10pt for labels
  #   legend.text  = element_text(size = 8),   # 8pt for legend text
  #   legend.position = "bottom",           # Bottom left corner
  #   legend.justification = "left",
  #   #legend.justification = c(0, 0),      # Anchor at bottom left
  #   #legend.background = element_rect(fill = "white", color = "grey80"),
  #   legend.margin = margin(l = -0.25, unit = "npc"),
  #   #legend.margin = margin(5, 5, 5, 5),
  #   # legend.position = "bottom"
  # )
    theme(
    axis.text.x  = element_text(size = 8),   # 8pt for axis text
    axis.text.y  = element_text(size = 8),   # 8pt for axis text
    axis.title.x = element_text(size = 8),  # 10pt for labels
    axis.title.y = element_blank(),
    title        = element_text(size = 8),  # 10pt for labels
    strip.text   = element_text(size = 8),  # 10pt for labels
    legend.title = element_text(size = 8),  # 10pt for labels
    legend.text  = element_text(size = 8),   # 8pt for legend text
    legend.position = "right",           # Bottom left corner
    legend.justification = "left",
    #legend.justification = c(0, 0),      # Anchor at bottom left
    #legend.background = element_rect(fill = "white", color = "grey80"),
    legend.margin = margin(l = -0.25, unit = "npc"),
    #legend.margin = margin(5, 5, 5, 5),
    # legend.position = "bottom"
  )


cause_levels <- factor(gbd_2019$cause,
                        levels = gbd_2019$cause[order(gbd_2019$val)])

ggsave(
  filename = "california_rank_studies.pdf",
  width = 7.5 * 0.6,   # Letter width
  height = 11 * 0.3,   # Letter height
  units = "in",
  device = cairo_pdf  # Better font rendering
)

```

# Plot: Proportion of full text, across disease conditions

```{r fulltext_by_condition}

fulltext_by_condition <-
  gwas_non_comm_study_info |>
  dplyr::mutate(HAS_FULLTEXT = ifelse(pubmed_id %in% downloaded_fulltexts,
                                     T,
                                     F)) |>
# gwas_study_info |>
  # dplyr::mutate(HAS_FULLTEXT = ifelse(pubmed_id %in% downloaded_fulltexts,
  #                                    T,
  #                                    F)) |>
  # dplyr::mutate(HAS_COHORT = ifelse(pubmed_id %in% studies_with_true_cohort_info_orig,
  #                                 T,
  #                                F)) |>
  # dplyr::mutate(HAS_ARCHIVE = ifelse(pubmed_id %in% studies_with_archive_info,
  #                                 T,
  #                                 F)) |>
  # dplyr::mutate(cohort_info = case_when( HAS_ARCHIVE == T &
  #                                        HAS_COHORT == T & 
  #                                        HAS_FULLTEXT == T ~ "All",
  #                                        
  #                                      HAS_ARCHIVE == T &
  #                                      HAS_COHORT == T & 
  #                                     HAS_FULLTEXT == F ~ "Cohort + Archive",
  #                                     
  #                                      HAS_ARCHIVE == F &
  #                                      HAS_COHORT == T & 
  #                                        HAS_FULLTEXT == T ~ "Cohort + Full text",
  #                                     
  #                                        HAS_COHORT == T & 
  #                                        HAS_FULLTEXT == F & 
  #                                        HAS_ARCHIVE == F ~ "Cohort only",
  #                                     
  #                                      HAS_ARCHIVE == T &
  #                                      HAS_COHORT == F & 
  #                                        HAS_FULLTEXT == T ~ "Archive + Full text",
  #                                     
  #                                     
  #                                     HAS_ARCHIVE == F &
  #                                      HAS_COHORT == F & 
  #                                        HAS_FULLTEXT == T ~ "Full text only",
  #                                     
  #                                     HAS_ARCHIVE == T &
  #                                      HAS_COHORT == F & 
  #                                        HAS_FULLTEXT == F ~ "Archive only",
  #                                        
  #                                        HAS_ARCHIVE == F &
  #                                        HAS_COHORT == F & 
  #                                        HAS_FULLTEXT == F ~ "None")) |>
  dplyr::group_by(cause, 
                  HAS_FULLTEXT) |>
  dplyr::summarise(n_studies = length(unique(pubmed_id))) |>
  dplyr::ungroup() |>
  tidyr::pivot_wider(names_from = HAS_FULLTEXT,
                     values_from = n_studies,
                     values_fill = 0) |>
  dplyr::mutate(TOTAL_STUDIES = `FALSE` + `TRUE`) |>
  dplyr::mutate(PROP_FULLTEXT = 100 * `TRUE` / TOTAL_STUDIES) |>
  dplyr::arrange(desc(PROP_FULLTEXT))

# add lancet grouping
lancet_grouping <- readxl::read_xlsx(here::here("data/icd/lancet_conditions_icd10.xlsx"))

lancet_grouping = lancet_grouping |>
  dplyr::select(cause = mapped_gbd_term, lancet_group)

fulltext_by_condition = fulltext_by_condition |>
  dplyr::left_join(lancet_grouping, by = "cause") |>
  dplyr::arrange(desc(PROP_FULLTEXT))


# group liver cancer due to hepatitis B and liver cancer due to hepatitis C
fulltext_by_condition = fulltext_by_condition |>
  dplyr::mutate(cause = ifelse(cause %in% c("Liver cancer due to hepatitis B",
                                           "Liver cancer due to hepatitis C"),
                              "Liver cancer",
                              cause)) |>
  dplyr::group_by(cause, lancet_group) |>
  dplyr::summarise(
    `False` = sum(`FALSE`),
    `True` = sum(`TRUE`)
  ) |>
  dplyr::mutate(TOTAL_STUDIES = `False` + `True`) |>
  dplyr::mutate(PROP_FULLTEXT = 100 * `True` / TOTAL_STUDIES) |>
  dplyr::arrange(desc(PROP_FULLTEXT))

# exclude "Diabetes mellitus"
fulltext_by_condition = fulltext_by_condition |>
  dplyr::filter(cause != "Diabetes mellitus")

# rename Chronic kidney disease due to diabetes mellitus type 2
# Chronic kidney disease (type 2 diabetes)
fulltext_by_condition = fulltext_by_condition |>
  dplyr::mutate(cause = ifelse(cause == "Chronic kidney disease due to diabetes mellitus type 2",
                              "Chronic kidney disease (type 2 diabetes)",
                              cause))

# repeat for Chronic kidney disease due to diabetes mellitus type 1
fulltext_by_condition = fulltext_by_condition |>
  dplyr::mutate(cause = ifelse(cause == "Chronic kidney disease due to diabetes mellitus type 1",
                              "Chronic kidney disease (type 1 diabetes)",
                              cause))


fulltext_by_condition <-
  tidyr::pivot_longer(data = fulltext_by_condition,
                      cols = !c("cause", 
                               "lancet_group", 
                               "TOTAL_STUDIES", 
                               "PROP_FULLTEXT"),
                      names_to = "FULLTEXT_STATUS",
                      values_to = "N_STUDIES")

# I-8 = Infectious diseases
# NCD-7 = Non-communicable diseases
fulltext_by_condition  = fulltext_by_condition  |>
  dplyr::mutate(lancet_group = ifelse(lancet_group == "I-8",
                                     "Infectious diseases",
                                     lancet_group)) |>
  dplyr::mutate(lancet_group = ifelse(lancet_group == "NCD-7",
                                     "Non-communicable diseases",
                                     lancet_group))


# fulltext_by_condition =
#   fulltext_by_condition |>
#   mutate(cause = factor(cause,
#                         levels = gbd_2019$cause[order(gbd_2019$val)]))

ggplot(aes(x = reorder(cause, N_STUDIES),
           fill = FULLTEXT_STATUS,
           y = N_STUDIES),
       data = fulltext_by_condition) +
  geom_col() +
  coord_flip() +
  labs(x = "",
       y = "Number of published studies") +
       #title = "Number of published GWAS Catalog studies per condition") +
  # facet_grid(space = "free_y",
  #            scale = "free_y",
  #            rows = vars(lancet_group)) + 
  scale_fill_manual(
    name = "Full text downloaded",
    values = c(
      `True` = "dodgerblue",
      `False` = "grey"
    ),
    labels = c(
      `True` = "Y",
      `False` = "N"
    )
  ) + 
    theme_minimal(base_family = "Arial") +
  theme(
    #axis.text.y  = #element_text(size = 20), 
    #element_text(size = 8),   # 8pt for axis text
    axis.text.x  = #element_text(size = 20),   # 8pt for axis text
    element_text(size = 8),   
    axis.title.x = #element_text(size = 20),  # 10pt for labels
    element_text(size = 8),   
    axis.title.y = #element_text(size = 18), #element_blank(),
    element_blank(),
    #title        = element_text(size = 9),  # 10pt for labels
    strip.text   = #element_text(size = 18),  # 10pt for labels
    element_text(size = 8),
    legend.title = #element_text(size = 22),  # 10pt for labels
    element_text(size = 8),
    legend.text  = #element_text(size = 20),   # 8pt for legend text
    element_text(size = 8),
    legend.position = "bottom",#, #"left", # "right", # "bottom",           # Bottom left corner
    legend.margin = margin(t = 0, unit = "pt"),
    legend.box.spacing = unit(0, "pt"),
    plot.margin = margin(t = 5, 
                         r = 5, 
                         b = 5, 
                         l = 0, 
                         unit = "pt"),
    axis.text.y = element_text(size = 8, margin = margin(r = 0, unit = "pt")),
    #legend.justification = "left"
    legend.justification = c(-5, 2)  # Anchor at bottom left
    #legend.background = element_rect(fill = "white", color = "grey80"),
    #legend.margin = margin(l = -0.25, unit = "npc"),
    #legend.margin = margin(5, 5, 5, 5),
    # legend.position = "bottom"
  )

  # theme_minimal() +
  # theme(
  #   axis.text.x  = element_text(size = 20),
  #   axis.text.y  = element_text(size = 20),
  #   axis.title.x = element_text(size = 24),
  #   axis.title.y = element_blank(),
  #   title        = element_text(size = 24), 
  #   strip.text = element_text(size = 22),
  #   legend.title = element_text(size = 24),
  #   legend.text  = element_text(size = 23)
  # )

# Export to PDF with letter size (8.5 x 11 inches)
# Adjust width to maintain readability based on number of categories
ggsave(
  filename = "fulltext_by_condition.pdf",
  width = 7.5 * 0.7,   # Letter width
  height = 11 * 0.28,   # Letter height
  units = "in",
  device = cairo_pdf  # Better font rendering
)


```

# Overlap between studies with cohort info and full text 

```{r overlap_cohort_fulltext}

cohort_pubmed_ids <-
gwas_cohort |>
  dplyr::filter(COHORT != "") |>
  dplyr::pull(PUBMED_ID) |>
  unique() 

fulltext_pubmed_ids <- downloaded_fulltexts

n_studies_with_cohort_and_fulltext <-
length(intersect(cohort_pubmed_ids, fulltext_pubmed_ids))

print("Number of published studies with cohort information and full text downloaded:")
print(n_studies_with_cohort_and_fulltext)

print("Proportion of published studies with cohort information that also have full text downloaded:")
100 * n_studies_with_cohort_and_fulltext / n_studies_with_cohort_info

print("Proportion of published studies with full text downloaded that also have cohort information:")
100 * n_studies_with_cohort_and_fulltext / n_studies_with_fulltext

print("Proportion of published studies with cohort information or full text downloaded:")
n_studies_with_cohort_or_fulltext <-
length(union(cohort_pubmed_ids, fulltext_pubmed_ids))

100 * n_studies_with_cohort_or_fulltext / n_published_studies

```

# Proportion of studies with dbGAP information (Europe PMC)

```{r dbgap_info}

dbgap_info <-
data.table::fread(here::here("data/europe_pmc/dbgap.csv")) 

all_pmids <- gwas_study_info$pubmed_id |> unique()

pmids_with_dbgap_info <-
dbgap_info |>
  dplyr::filter(EXTID %in% all_pmids) 

# check no different with pmcids
all_pmcids <-
convert_pmid_df |>
  dplyr::filter(PMID %in% gwas_study_info$pubmed_id) |>
  dplyr::pull(PMCID)

all_pmcids <- all_pmcids[all_pmcids != ""]

pmcids_with_dbgap_info <-
dbgap_info |>
  dplyr::filter(PMCID %in% all_pmcids) 

n_studies_with_dbgap_info <-
  dplyr::bind_rows(pmids_with_dbgap_info,
            pmcids_with_dbgap_info)  |>
  dplyr::select(EXTID, PMCID) |>
  dplyr::distinct() |>
  dplyr::pull(EXTID) |>
  unique() |>
  length()

print("Number of published studies with dbGAP information:")
print(n_studies_with_dbgap_info)

print("Proportion of published studies with dbGAP information:")
100 * n_studies_with_dbgap_info / n_published_studies

```

# Proportion of studies with EGA information (Europe PMC)

```{r ega_info}

data.table::fread(here::here("data/europe_pmc/ega.csv")) -> ega_info

all_pmids <- gwas_study_info$pubmed_id |> unique()

pmids_with_ega_info <-
ega_info |>
  dplyr::filter(EXTID %in% all_pmids) |>
  dplyr::pull(EXTID) 

n_studies_with_ega_info <-
pmids_with_ega_info |>
  unique() |>
  length()

print("Number of published studies with EGA information:")
print(n_studies_with_ega_info)

print("Proportion of published studies with EGA information:")
100 * n_studies_with_ega_info / n_published_studies

```

# Studies with EGA, dbGAP, or JGA information

```{r ega_dbgap_jga_info}

epmc_dbgap_info <- pmids_with_dbgap_info |> pull(EXTID) |> unique()

text_extracted <- data.table::fread(
                   here::here("output/gwas_cat/gwas_study_dbgap_ega_sentences.csv")
)

pmcids_with_archive_info <- 
text_extracted |>
  pull(pmcid) |>
  unique()

pmids_with_archive_info <-
convert_pmid_df |>
  dplyr::filter(PMCID %in% pmcids_with_archive_info) |>
  dplyr::pull(PMID) |>
  unique()

epmc_ega_info <- pmids_with_ega_info |> unique()

studies_with_archive_info <- 
  Reduce("union", 
      list(epmc_dbgap_info,
      epmc_ega_info,
      pmids_with_archive_info))

```

# Overlap in dbgap and full text

```{r overlap_dbgap_fulltext}

dbgap_pubmed_ids <-
  dplyr::bind_rows(pmids_with_dbgap_info,
            pmcids_with_dbgap_info) |>
  dplyr::select(EXTID, PMCID) |>
  dplyr::distinct() |>
  dplyr::pull(EXTID) |>
  unique()
  
n_studies_with_dbgap_and_fulltext <-
length(intersect(dbgap_pubmed_ids, fulltext_pubmed_ids))

print("Number of published studies with dbGAP information and full text downloaded:")
print(n_studies_with_dbgap_and_fulltext)

print("Proportion of published studies with dbGAP information that also have full text downloaded:")
100 * n_studies_with_dbgap_and_fulltext / n_studies_with_dbgap_info

print("Proportion of published studies with full text downloaded that also have dbGAP information:")
100 * n_studies_with_dbgap_and_fulltext / n_studies_with_fulltext

print("Proportion of published studies with dbGAP information or full text downloaded:")
n_studies_with_dbgap_or_fulltext <-
length(union(dbgap_pubmed_ids, fulltext_pubmed_ids))

100 * n_studies_with_dbgap_or_fulltext / n_published_studies

print("Proportion of published studies that have dbGAP, cohort information or full text downloaded:")
n_studies_with_dbgap_cohort_or_fulltext <-
length(Reduce(union, list(dbgap_pubmed_ids,
                                 cohort_pubmed_ids,
                                 fulltext_pubmed_ids
                                 )
                      )
       )

100 * n_studies_with_dbgap_cohort_or_fulltext / n_published_studies

```

# Studies without full text, what proportion have dbGAP info?

```{r no_fulltext_dbgap_info}

n_studies_without_fulltext_with_dbgap_info <-
sum(studies_without_fulltext %in% pmids_with_dbgap_info$EXTID)

print("Number of published studies without full text downloaded that have dbGAP information:")
print(n_studies_without_fulltext_with_dbgap_info)

print("Proportion of published studies without full text downloaded that have dbGAP information:")
100 * n_studies_without_fulltext_with_dbgap_info / length(studies_without_fulltext)

```

# Percentage of cohort information that includes 'other' or 'multiple'

```{r, eval = F}

studies_with_other_or_multiple_cohort_info <- gwas_study_info |>
  dplyr::mutate(COHORT = stringr::str_remove_all(pattern = "(^NR$)|(^NR\\|)|(\\|NR$)|(\\|NR\\|)", 
                                                 COHORT)) |> 
  dplyr::filter(grepl("other|multiple", tolower(COHORT))) |>
  dplyr::pull(pubmed_id) |>
  unique() |>
  length()

print("Number of published studies with 'other' or 'multiple' cohort information:")
print(studies_with_other_or_multiple_cohort_info)

print("Proportion of published studies with 'other' or 'multiple' cohort information:")
100 * studies_with_other_or_multiple_cohort_info / n_studies_with_cohort_info 


```




