---
title: "Stats for specific aims"
author: "Isobel Beasley"
date: "2025-10-23"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load_libraries, message=FALSE, warning=FALSE}

library(data.table)
library(dplyr)
library(ggplot2)

```

# Proportion of all published studies that have cohort meta-data

```{r original_cohort_info}

orig_gwas_study_info <- data.table::fread(here::here("data/gwas_catalog/gwas-catalog-v1.0.3.1-studies-r2025-07-21.tsv"))

orig_gwas_study_info = orig_gwas_study_info |>
  dplyr::rename_with(~ tolower(gsub(" ", "_", .x)))

# how many published studies (i.e. number of unique pubmed ids)
n_published_studies_orig <- orig_gwas_study_info |>
  dplyr::filter(!is.na(pubmed_id)) |>
  dplyr::pull(pubmed_id) |>
  unique() |>
  length()

print("Number of published studies in original GWAS Catalog:")
print(n_published_studies_orig)

# how many published studies include any cohort information?
n_studies_with_cohort_info_orig <- orig_gwas_study_info |>
  dplyr::filter(!is.na(cohort) & cohort != "") |>
  dplyr::pull(pubmed_id) |>
  unique() |>
  length()

print("Number of published studies with cohort information in original GWAS Catalog:")
print(n_studies_with_cohort_info_orig)

print("Proportion of published studies with cohort information in original GWAS Catalog:")
100 * n_studies_with_cohort_info_orig / n_published_studies_orig

studies_with_true_cohort_info_orig <- orig_gwas_study_info |>
  dplyr::mutate(cohort = stringr::str_remove_all(pattern = "(^NR$)|(^NR\\|)|(\\|NR$)|(\\|NR\\|)|(not reported)", 
                                                 cohort)) |>
  dplyr::mutate(cohort = stringr::str_remove_all(pattern = "other|multiple|Other|Multiple|Others|\\(Multiple cohorts\\)|OTHER", 
                                                 cohort)) |>
  dplyr::filter(!is.na(cohort) & 
                  cohort != "") |>
  dplyr::pull(pubmed_id) |>
  unique()

# how many published studies include any true cohort information 
# (i.e. not 'NR', or 'not reported', 'other', 'multiple')
n_studies_with_true_cohort_info_orig <- studies_with_true_cohort_info_orig |> 
  unique() |>
  length()

print("Number of published studies with true cohort information in original GWAS Catalog:")
print(n_studies_with_true_cohort_info_orig)

print("Proportion of published studies with true cohort information in original GWAS Catalog:")
100 * n_studies_with_true_cohort_info_orig / n_published_studies_orig

data.frame(n_studies = c(n_published_studies_orig - n_studies_with_cohort_info_orig, 
                         n_studies_with_cohort_info_orig - n_studies_with_true_cohort_info_orig,
                         n_studies_with_true_cohort_info_orig),
           type = c("all",
                    "studies_with_cohort_info",
                    "studies_with_true_cohort_info"),
           group = c("GWAS Catalog Published Studies")) |>
ggplot(aes(x = group, y = n_studies, fill = type)) +
  geom_col(position = "stack") +
  labs(x = "",
       y = "Number of studies",
       title = "Published GWAS Catalog Studies") +
  scale_y_continuous(
    breaks = seq(0, max(n_published_studies_orig), by = 1000),
    expand = expansion(mult = c(0, 0.05))
  ) + 
  scale_fill_manual(
    name = "Cohort metadata availability",
    values = c(
      all = "grey",
      studies_with_cohort_info = "black",
      studies_with_true_cohort_info = "#1b9e77"
    ),
    labels = c(
      all = "None",
      studies_with_cohort_info = "Uninformative metadata\n(NR, other, multiple)",
      studies_with_true_cohort_info = "At least one valid cohort name"
    )
  ) + 
  theme_minimal() +
  theme(
    axis.text.x  = element_blank(),
    axis.ticks.x = element_blank(),
    axis.text.y  = element_text(size = 20),
    axis.title.y = element_text(size = 24),
    legend.title = element_text(size = 24),
    legend.text  = element_text(size = 23),
    title        = element_text(size = 24)
  )



```

# Number of papers

## For all priority conditions

Non-communicable and communicable

```{r n_total_papers, message=FALSE, warning=FALSE}

gwas_study_info <- data.table::fread(here::here("output/icd_map/gwas_study_gbd_causes.csv"))

gwas_study_info = gwas_study_info |>
  dplyr::rename_with(~ tolower(gsub(" ", "_", .x))) |>
  dplyr::filter(cause != "")

# how many published studies (i.e. number of unique pubmed ids)
n_published_studies <- gwas_study_info |>
  dplyr::filter(!is.na(pubmed_id)) |>
  dplyr::pull(pubmed_id) |>
  unique() |>
  length()

print("Number of published studies in GWAS Catalog:")
print(n_published_studies)

```

## For non-communicable diseases only (i.e. excluding infectious disease studies)

```{r n_non_com_total_papers, message=FALSE, warning=FALSE}

gwas_non_comm_study_info = gwas_study_info |>
  dplyr::filter(!cause %in% c("HIV/AIDS",
                             "Tuberculosis",
                             "Malaria",
                             "Lower respiratory infections",
                             "Diarrhoeal diseases",
                             "Neonatal disorders",
                             "Tetanus",
                             "Diphtheria",
                             "Pertussis" ,
                             "Measles",
                             "Maternal disorders")) 

print("Non-communicable disease causes in the dataset:")
gwas_non_comm_study_info$cause |> unique() |> sort()

# how many published studies (i.e. number of unique pubmed ids)
n_published_non_comm_studies <- gwas_non_comm_study_info |>
  dplyr::filter(!is.na(pubmed_id)) |>
  dplyr::pull(pubmed_id) |>
  unique() |>
  length()

print("Number of published non-communicable disease studies in GWAS Catalog:")
n_published_non_comm_studies

```

## For each condition

```{r n_papers_per_condition}

condition_counts <-
gwas_study_info |>
  dplyr::select(pubmed_id, cause) |>
  dplyr::distinct() |> 
  dplyr::group_by(cause) |>
  dplyr::summarise(n_studies = length(unique(pubmed_id))) |>
  dplyr::arrange(desc(n_studies))

lancet_grouping <- readxl::read_xlsx(here::here("data/icd/lancet_conditions_icd10.xlsx"))

lancet_grouping = lancet_grouping |>
  dplyr::select(cause = mapped_gbd_term, lancet_group) |>
  distinct()

condition_counts = condition_counts |>
  dplyr::left_join(lancet_grouping, by = "cause") |>
  dplyr::arrange(desc(n_studies))

# combine Liver cancer due to hepatitis B and Liver cancer due to hepatitis C
condition_counts = condition_counts |>
  dplyr::mutate(cause = ifelse(cause %in% c("Liver cancer due to hepatitis B",
                                           "Liver cancer due to hepatitis C"),
                              "Liver cancer",
                              cause)) |>
  dplyr::group_by(cause, lancet_group) |>
  dplyr::summarise(n_studies = sum(n_studies)) |>
  dplyr::arrange(desc(n_studies))

# exclude "Diabetes mellitus"
condition_counts = condition_counts |>
  dplyr::filter(cause != "Diabetes mellitus")

# rename Chronic kidney disease due to diabetes mellitus type 2
# Chronic kidney disease (type 2 diabetes)
condition_counts = condition_counts |>
  dplyr::mutate(cause = ifelse(cause == "Chronic kidney disease due to diabetes mellitus type 2",
                              "Chronic kidney disease (type 2 diabetes)",
                              cause))

# repeat for Chronic kidney disease due to diabetes mellitus type 1
condition_counts = condition_counts |>
  dplyr::mutate(cause = ifelse(cause == "Chronic kidney disease due to diabetes mellitus type 1",
                              "Chronic kidney disease (type 1 diabetes)",
                              cause))


# change lancent group names, for better display
# I-8 = Infectious diseases
# NCD-7 = Non-communicable diseases
condition_counts = condition_counts |>
  dplyr::mutate(lancet_group = ifelse(lancet_group == "I-8",
                                     "Infectious diseases",
                                     lancet_group)) |>
  dplyr::mutate(lancet_group = ifelse(lancet_group == "NCD-7",
                                     "Non-communicable diseases",
                                     lancet_group))

ggplot(aes(x = reorder(cause, n_studies), 
           fill = lancet_group,
           y = n_studies),
       data = condition_counts) +
  geom_col() +
  coord_flip() +
  labs(#x = "Condition",
       y = "Number of published studies") +
       #title = "Number of published GWAS Catalog studies per condition") +
  # facet_wrap(~lancet_group,
  #            nrow = 2,
  #            scales = "free_y") +
  facet_grid(space = "free_y",
             scale = "free_y",
             rows = vars(lancet_group)) + 
  scale_fill_discrete(type = c("skyblue", "orange")) + 
  theme_minimal() +
  theme(
    axis.text.x  = element_text(size = 20),
    axis.text.y  = element_text(size = 20),
    axis.title.x = element_text(size = 24),
    axis.title.y = element_blank(),
    title        = element_text(size = 24), 
    strip.text = element_text(size = 22),
    legend.position = "none"
  )

```

# Number of studies per condition - non_communicable diseases only

```{r n_papers_per_non-communicable_condition}

condition_counts <-
gwas_non_comm_study_info |>
  dplyr::select(pubmed_id, cause) |>
  dplyr::distinct() |> 
  dplyr::group_by(cause) |>
  dplyr::summarise(n_studies = length(unique(pubmed_id))) |>
  dplyr::arrange(desc(n_studies))

lancet_grouping <- readxl::read_xlsx(here::here("data/icd/lancet_conditions_icd10.xlsx"))

lancet_grouping = lancet_grouping |>
  dplyr::select(cause = mapped_gbd_term, lancet_group) |>
  distinct()

condition_counts = condition_counts |>
  dplyr::left_join(lancet_grouping, by = "cause") |>
  dplyr::arrange(desc(n_studies))

# combine Liver cancer due to hepatitis B and Liver cancer due to hepatitis C
condition_counts = condition_counts |>
  dplyr::mutate(cause = ifelse(cause %in% c("Liver cancer due to hepatitis B",
                                           "Liver cancer due to hepatitis C"),
                              "Liver cancer",
                              cause)) |>
  dplyr::group_by(cause, lancet_group) |>
  dplyr::summarise(n_studies = sum(n_studies)) |>
  dplyr::arrange(desc(n_studies))

# exclude "Diabetes mellitus"
condition_counts = condition_counts |>
  dplyr::filter(cause != "Diabetes mellitus")

# rename Chronic kidney disease due to diabetes mellitus type 2
# Chronic kidney disease (type 2 diabetes)
condition_counts = condition_counts |>
  dplyr::mutate(cause = ifelse(cause == "Chronic kidney disease due to diabetes mellitus type 2",
                              "Chronic kidney disease (type 2 diabetes)",
                              cause))

# repeat for Chronic kidney disease due to diabetes mellitus type 1
condition_counts = condition_counts |>
  dplyr::mutate(cause = ifelse(cause == "Chronic kidney disease due to diabetes mellitus type 1",
                              "Chronic kidney disease (type 1 diabetes)",
                              cause))


# change lancent group names, for better display
# I-8 = Infectious diseases
# NCD-7 = Non-communicable diseases
condition_counts = condition_counts |>
  dplyr::mutate(lancet_group = ifelse(lancet_group == "I-8",
                                     "Infectious diseases",
                                     lancet_group)) |>
  dplyr::mutate(lancet_group = ifelse(lancet_group == "NCD-7",
                                     "Non-communicable diseases",
                                     lancet_group))

ggplot(aes(x = reorder(cause, n_studies), 
           fill = lancet_group,
           y = n_studies),
       data = condition_counts) +
  geom_col() +
  coord_flip() +
  labs(#x = "Condition",
       y = "Number of published studies") +
       #title = "Number of published GWAS Catalog studies per condition") +
  # facet_wrap(~lancet_group,
  #            nrow = 2,
  #            scales = "free_y") +
  facet_grid(space = "free_y",
             scale = "free_y",
             rows = vars(lancet_group)) + 
  scale_fill_discrete(type = c("skyblue", "orange")) + 
  theme_minimal() +
  theme(
    axis.text.x  = element_text(size = 20),
    axis.text.y  = element_text(size = 20),
    axis.title.x = element_text(size = 24),
    axis.title.y = element_blank(),
    title        = element_text(size = 24), 
    strip.text = element_text(size = 22),
    legend.position = "none"
  )

```

# Examples

## Relationship between causes and mapped traits

```{r causes_mapped_traits}

cause_mapped_trait_counts <-
gwas_study_info |>
  dplyr::select(cause, collected_all_disease_terms, pubmed_id) |>
  dplyr::distinct() |> 
  dplyr::group_by(cause, collected_all_disease_terms) |>
  dplyr::summarise(n_studies = length(unique(pubmed_id))) |>
  dplyr::arrange(cause, desc(n_studies))
  

cause_mapped_trait_counts |>
  dplyr::filter(cause == "Cervical cancer")

```

# Proportion of papers

## That have cohort information/metadata that includes 'other' or 'multiple'

```{r, eval = F}

studies_with_other_or_multiple_cohort_info <- gwas_study_info |>
  dplyr::mutate(COHORT = stringr::str_remove_all(pattern = "(^NR$)|(^NR\\|)|(\\|NR$)|(\\|NR\\|)", 
                                                 COHORT)) |> 
  dplyr::filter(grepl("other|multiple", tolower(COHORT))) |>
  dplyr::pull(pubmed_id) |>
  unique() |>
  length()

print("Number of published studies with 'other' or 'multiple' cohort information:")
print(studies_with_other_or_multiple_cohort_info)

print("Proportion of published studies with 'other' or 'multiple' cohort information:")
100 * studies_with_other_or_multiple_cohort_info / n_studies_with_cohort_info 


```

## That include cohort information (of all papers studying priority conditions)

```{r prop_cohort_info}

gwas_cohort <- data.table::fread(here::here("output/gwas_cohorts/gwas_cohort_name_corrected.csv")) 

gwas_cohort = gwas_cohort |>
  dplyr::filter(PUBMED_ID %in% unique(gwas_study_info$pubmed_id))

# how many published studies include cohort information? 
n_studies_with_cohort_info <- gwas_cohort |>
  dplyr::filter(COHORT != "") |>
  dplyr::pull(PUBMED_ID) |>
  unique() |>
  length()

print("Number of published studies with cohort information:")
print(n_studies_with_cohort_info)

print("Proportion of published studies with cohort information:")
100 * n_studies_with_cohort_info / n_published_studies

```

## That include cohort information/metadata (of non-communicable disease studies)

```{r prop_cohort_info_non_comm}

gwas_cohort_non_comm = gwas_cohort |>
  dplyr::filter(PUBMED_ID %in% unique(gwas_non_comm_study_info$pubmed_id))

# how many published studies include cohort information?
n_studies_with_cohort_info_non_comm <- gwas_cohort_non_comm |>
  dplyr::filter(COHORT != "") |>
  dplyr::pull(PUBMED_ID) |>
  unique() |>
  length()

print("Number of published non-communicable disease studies with cohort information:")
n_studies_with_cohort_info_non_comm

print("Proportion of published non-communicable disease studies with cohort information:")
100 * n_studies_with_cohort_info_non_comm / n_published_non_comm_studies

```

## That have downloaded abstracts (of all papers studying non-communicable conditions)

```{r prop_abstracts}

downloaded_abstracts <-
 list.files(here::here("output/abstracts")) |> 
  stringr::str_remove_all(".txt$") 

studies_with_abstracts <-
gwas_non_comm_study_info |>
  dplyr::filter(pubmed_id %in% downloaded_abstracts) |>
  dplyr::pull(pubmed_id) |>
  unique() 

n_studies_with_abstracts <-
  studies_with_abstracts  |>
  length()

print("Number of published studies with abstracts downloaded:")
print(n_studies_with_abstracts)

print("Proportion of published studies with abstracts downloaded:")
100 * n_studies_with_abstracts / n_published_non_comm_studies

print("Number of published studies without abstracts downloaded:")
print(n_published_non_comm_studies - n_studies_with_abstracts)

```

## That have downloaded full text (.xml) (of all papers studying non-communicable conditions)

```{r prop_full_text}

full_text_files <-
  list.files(here::here("output/fulltexts"), 
             recursive = T, 
             pattern = "\\.xml$") 

all_ids <-
  full_text_files |>
  gsub(pattern = ".*/", replacement = "") |>
  gsub(pattern = "\\.xml$", replacement = "") |>
  gsub(pattern = "_bioc", replacement = "") |>
  gsub(pattern = "\\.pdf.tei", replacement = "")


converted_ids <- data.table::fread(here::here("output/fulltexts/pmid_to_pmcid_mapping.csv")) 

# convert pmcids to pmids
converted_pmids <-
converted_ids |> 
  filter(pmcids %in% all_ids) |>
  pull(PMID) |>
  unique() 

all_fulltext_pmids <- c(grep("PMC", all_ids, value = T, invert = T), 
               converted_pmids) |> 
  unique()

studies_with_fulltext <-
gwas_non_comm_study_info |>
  dplyr::filter(pubmed_id %in% all_fulltext_pmids) |>
  dplyr::pull(pubmed_id) |>
  unique() 

studies_without_fulltext <-
  gwas_non_comm_study_info |>
  dplyr::filter(!c(pubmed_id %in% all_fulltext_pmids)) |>
  dplyr::pull(pubmed_id) |>
  unique() 

n_studies_with_fulltext <-
studies_with_fulltext |>
  length()

print("Number of published studies with full text downloaded:")
print(n_studies_with_fulltext)

print("Proportion of published studies with full text downloaded:")
100 * n_studies_with_fulltext / n_published_non_comm_studies

print("Number of published studies without full text downloaded:")
print(n_published_non_comm_studies - n_studies_with_fulltext)

```

## Proportion of papers without full text downloaded 

### That have cohort information

```{r no_fulltext_cohort_info}

n_studies_without_fulltext_with_cohort_info <-
gwas_cohort |>
  dplyr::filter(PUBMED_ID %in% studies_without_fulltext) |>
  dplyr::filter(COHORT != "") |>
  dplyr::pull(PUBMED_ID) |>
  unique() |>
  length()

print("Number of published studies without full text downloaded that have cohort information:")
print(n_studies_without_fulltext_with_cohort_info)

print("Proportion of published studies without full text downloaded that have cohort information:")
100 * n_studies_without_fulltext_with_cohort_info / length(studies_without_fulltext)

```

### That have dbGAP info

```{r no_fulltext_dbgap_info}

combined_dbgap_pubmed_mapping <- data.table::fread(
  here::here("output/gwas_cohorts/gwas_study_combined_dbgap_pubmed_mapping.csv")
)

n_studies_without_fulltext_with_dbgap_info <-
sum(studies_without_fulltext %in% combined_dbgap_pubmed_mapping$PUBMED_ID)

print("Number of published studies without full text downloaded that have dbGAP information:")
print(n_studies_without_fulltext_with_dbgap_info)

print("Proportion of published studies without full text downloaded that have dbGAP information:")
100 * n_studies_without_fulltext_with_dbgap_info / length(studies_without_fulltext)

```


## Proportion of papers with full text

### That have cohort info 

```{r overlap_cohort_fulltext}

cohort_pubmed_ids <-
gwas_cohort |>
  dplyr::filter(PUBMED_ID %in% unique(gwas_non_comm_study_info$pubmed_id)) |>
  dplyr::filter(COHORT != "") |>
  dplyr::pull(PUBMED_ID) |>
  unique() 

n_studies_with_cohort_and_fulltext <-
length(intersect(cohort_pubmed_ids, all_fulltext_pmids))

print("Number of published studies with cohort information and full text downloaded:")
print(n_studies_with_cohort_and_fulltext)

print("Proportion of published studies with cohort information that also have full text downloaded:")
100 * n_studies_with_cohort_and_fulltext / n_studies_with_cohort_info

print("Proportion of published studies with full text downloaded that also have cohort information:")
100 * n_studies_with_cohort_and_fulltext / n_studies_with_fulltext

print("Proportion of published studies of non-communicable traits with cohort information or full text downloaded:")
n_studies_with_cohort_or_fulltext <-
length(unique(c(cohort_pubmed_ids, all_fulltext_pmids)))

100 * n_studies_with_cohort_or_fulltext / n_published_non_comm_studies

```


### That have dbgap info 

```{r overlap_dbgap_fulltext}
  
n_studies_with_dbgap_and_fulltext <-
length(intersect(combined_dbgap_pubmed_mapping$PUBMED_ID, all_fulltext_pmids))

n_studies_with_dbgap_info <- 
combined_dbgap_pubmed_mapping$PUBMED_ID |>
  unique() |>
  length()

print("Number of published studies with dbGAP information and full text downloaded:")
print(n_studies_with_dbgap_and_fulltext)

print("Proportion of published studies with dbGAP information that also have full text downloaded:")
100 * n_studies_with_dbgap_and_fulltext / n_studies_with_dbgap_info

print("Proportion of published studies with full text downloaded that also have dbGAP information:")
100 * n_studies_with_dbgap_and_fulltext / n_studies_with_fulltext

print("Proportion of published studies with dbGAP information or full text downloaded:")
n_studies_with_dbgap_or_fulltext <-
length(union(combined_dbgap_pubmed_mapping$PUBMED_ID, all_fulltext_pmids))

100 * n_studies_with_dbgap_or_fulltext / n_published_non_comm_studies

print("Proportion of published studies that have dbGAP, cohort information or full text downloaded:")
n_studies_with_dbgap_cohort_or_fulltext <-
length(Reduce(union, list(combined_dbgap_pubmed_mapping$PUBMED_ID,
                                 cohort_pubmed_ids,
                                 all_fulltext_pmids
                                 )
                      )
       )

100 * n_studies_with_dbgap_cohort_or_fulltext / n_published_non_comm_studies

```



# Plots



## Number of studies with abstracts or full text 

```{r plot_abstracts_fulltext}

got_text_df <- 
gwas_non_comm_study_info |>
  select(pubmed_id) |>
  distinct() |>
  mutate(has_abstract = ifelse(pubmed_id %in% studies_with_abstracts, "Yes", "No"),
         has_fulltext = ifelse(pubmed_id %in% studies_with_fulltext, "Yes", "No")) 


got_text_df <-
  got_text_df  |>
  mutate(text_availability = case_when(
    has_fulltext == "Yes" & has_abstract == "Yes" ~ "Full text + abstract",
    has_abstract == "Yes" & has_abstract == "No" ~ "Abstract only",
    has_fulltext == "Yes" & has_abstract == "No" ~ "Full text only",
    TRUE ~ "None"
  ))


got_text_df <-
  got_text_df |> 
  group_by(text_availability) |>
  summarise(n_studies = n()) 

got_text_df |>
ggplot(aes(x = text_availability, 
           y = n_studies, 
           fill = text_availability)) +
  geom_col(position = "stack") +
  labs(x = "",
       y = "Number of studies",
       title = "Published GWAS Catalog Studies") +
  # scale_y_continuous(
  #   breaks = seq(0, max(n_published_studies), by = 1000),
  #   expand = expansion(mult = c(0, 0.05))
  # ) + 
  scale_fill_manual(
    name = "Text availability",
    values = c(
      "Full text + abstract" = "grey",
      "Full text only" = "black",
      "None" = "#7570b3"
    ),
    labels = c(
      all = "None",
      studies_with_abstracts = "Abstract only",
      studies_with_fulltext = "Full text + abstract"
    )
  ) + 
  theme_minimal() +
  theme(
    axis.text.x  = element_blank(),
    axis.ticks.x = element_blank(),
    axis.text.y  = element_text(size = 20),
    axis.title.y = element_text(size = 24),
    legend.title = element_text(size = 24),
    legend.text  = element_text(size = 23),
    title        = element_text(size = 24)
  )

```

## Year number of papers per condition

```{r papers_per_year}

gwas_study_info |>
  select(cause, year, pubmed_id) |> 
  distinct() |> 
  ggplot(aes(x = year)) +
  geom_histogram() + 
  facet_wrap(~cause) + 
  theme_minimal() 


gwas_study_info |>
  select(cause, year, pubmed_id) |> 
  distinct() |>
  group_by(cause) |>
  summarise(min_year = min(year),
            first_quartile_year = quantile(year, 0.25),
            median_year = median(year),
            average_year = mean(year),
            third_quartile_year = quantile(year, 0.75),
            max_year = max(year)
            ) 

```

## % deaths explained by diseases (California, 2019)

```{r deaths_by_condition}

gbd_2019 <-
  data.table::fread(here::here("data/gbd/gbd_2019_california_percent_deaths.csv"))

gbd_2019 = gbd_2019 |>
  dplyr::mutate(val = 100 * val) |>
  dplyr::mutate(lower = 100 * lower) |>
  dplyr::mutate(upper = 100 * upper)

gbd_2019 =
  gbd_2019 |>
  dplyr::mutate(cause = ifelse(cause == "Diabetes mellitus type 2",
                               "Type 2 diabetes mellitus",
                               cause)) |>
  dplyr::mutate(cause = ifelse(cause == "Diabetes mellitus type 1",
                               "Type 1 diabetes mellitus",
                               cause)) |>
    dplyr::mutate(cause = ifelse(cause == "Chronic kidney disease due to diabetes mellitus type 2",
                               "Chronic kidney disease (type 2 diabetes)",
                               cause)) |>
      dplyr::mutate(cause = ifelse(cause == "Chronic kidney disease due to diabetes mellitus type 1",
                               "Chronic kidney disease (type 1 diabetes)",
                               cause)) 

# add row summing all conditions
gbd_2019 |>
  dplyr::summarise(total_val = sum(val))


ggplot(aes(x = reorder(cause, val), 
           y = val),
       data = gbd_2019) +
  geom_col(fill = "firebrick") +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.3) +
  coord_flip() +
  labs(x = "",
       y = "% Percentage of all deaths (California, 2019)",
       #title = "Percentage of deaths by condition in California (2019)"
       ) +
    theme_minimal(base_family = "Arial") +
  # theme(
  #   axis.text.x  = element_text(size = 8),   # 8pt for axis text
  #   axis.text.y  = element_text(size = 8),   # 8pt for axis text
  #   axis.title.x = element_text(size = 9),  # 10pt for labels
  #   axis.title.y = element_blank(),
  #   title        = element_text(size = 9),  # 10pt for labels
  #   strip.text   = element_text(size = 9),  # 10pt for labels
  #   legend.title = element_text(size = 8),  # 10pt for labels
  #   legend.text  = element_text(size = 8),   # 8pt for legend text
  #   legend.position = "bottom",           # Bottom left corner
  #   legend.justification = "left",
  #   #legend.justification = c(0, 0),      # Anchor at bottom left
  #   #legend.background = element_rect(fill = "white", color = "grey80"),
  #   legend.margin = margin(l = -0.25, unit = "npc"),
  #   #legend.margin = margin(5, 5, 5, 5),
  #   # legend.position = "bottom"
  # )
    theme(
    axis.text.x  = element_text(size = 8),   # 8pt for axis text
    axis.text.y  = element_text(size = 8),   # 8pt for axis text
    axis.title.x = element_text(size = 8),  # 10pt for labels
    axis.title.y = element_blank(),
    title        = element_text(size = 8),  # 10pt for labels
    strip.text   = element_text(size = 8),  # 10pt for labels
    legend.title = element_text(size = 8),  # 10pt for labels
    legend.text  = element_text(size = 8),   # 8pt for legend text
    legend.position = "right",           # Bottom left corner
    legend.justification = "left",
    #legend.justification = c(0, 0),      # Anchor at bottom left
    #legend.background = element_rect(fill = "white", color = "grey80"),
    legend.margin = margin(l = -0.25, unit = "npc"),
    #legend.margin = margin(5, 5, 5, 5),
    # legend.position = "bottom"
  )


cause_levels <- factor(gbd_2019$cause,
                        levels = gbd_2019$cause[order(gbd_2019$val)])

ggsave(
  filename = here::here("figures/california_rank_studies.pdf"),
  width = 7.5 * 0.6,   # Letter width
  height = 11 * 0.3,   # Letter height
  units = "in",
  device = cairo_pdf  # Better font rendering
)

```

## Proportion of papers with full text, across disease conditions

```{r fulltext_by_condition}

fulltext_by_condition <-
  gwas_non_comm_study_info |>
  dplyr::filter(cause != "") |>
  select(cause, pubmed_id) |>
  distinct() |>
  dplyr::mutate(HAS_FULLTEXT = ifelse(pubmed_id %in% all_fulltext_pmids,
                                     T,
                                     F)) |>
# gwas_study_info |>
  # dplyr::mutate(HAS_FULLTEXT = ifelse(pubmed_id %in% downloaded_fulltexts,
  #                                    T,
  #                                    F)) |>
  # dplyr::mutate(HAS_COHORT = ifelse(pubmed_id %in% studies_with_true_cohort_info_orig,
  #                                 T,
  #                                F)) |>
  # dplyr::mutate(HAS_ARCHIVE = ifelse(pubmed_id %in% studies_with_archive_info,
  #                                 T,
  #                                 F)) |>
  # dplyr::mutate(cohort_info = case_when( HAS_ARCHIVE == T &
  #                                        HAS_COHORT == T & 
  #                                        HAS_FULLTEXT == T ~ "All",
  #                                        
  #                                      HAS_ARCHIVE == T &
  #                                      HAS_COHORT == T & 
  #                                     HAS_FULLTEXT == F ~ "Cohort + Archive",
  #                                     
  #                                      HAS_ARCHIVE == F &
  #                                      HAS_COHORT == T & 
  #                                        HAS_FULLTEXT == T ~ "Cohort + Full text",
  #                                     
  #                                        HAS_COHORT == T & 
  #                                        HAS_FULLTEXT == F & 
  #                                        HAS_ARCHIVE == F ~ "Cohort only",
  #                                     
  #                                      HAS_ARCHIVE == T &
  #                                      HAS_COHORT == F & 
  #                                        HAS_FULLTEXT == T ~ "Archive + Full text",
  #                                     
  #                                     
  #                                     HAS_ARCHIVE == F &
  #                                      HAS_COHORT == F & 
  #                                        HAS_FULLTEXT == T ~ "Full text only",
  #                                     
  #                                     HAS_ARCHIVE == T &
  #                                      HAS_COHORT == F & 
  #                                        HAS_FULLTEXT == F ~ "Archive only",
  #                                        
  #                                        HAS_ARCHIVE == F &
  #                                        HAS_COHORT == F & 
  #                                        HAS_FULLTEXT == F ~ "None")) |>
  dplyr::group_by(cause, 
                  HAS_FULLTEXT) |>
  dplyr::summarise(n_studies = length(unique(pubmed_id))) |>
  dplyr::ungroup() |>
  tidyr::pivot_wider(names_from = HAS_FULLTEXT,
                     values_from = n_studies,
                     values_fill = 0) |>
  dplyr::mutate(TOTAL_STUDIES = `FALSE` + `TRUE`) |>
  dplyr::mutate(PROP_FULLTEXT = 100 * `TRUE` / TOTAL_STUDIES) |>
  dplyr::arrange(desc(PROP_FULLTEXT))

# add lancet grouping
lancet_grouping <- readxl::read_xlsx(here::here("data/icd/lancet_conditions_icd10.xlsx"))

lancet_grouping = lancet_grouping |>
  dplyr::select(cause = mapped_gbd_term, lancet_group)

fulltext_by_condition = fulltext_by_condition |>
  dplyr::left_join(lancet_grouping, by = "cause") |>
  dplyr::arrange(desc(PROP_FULLTEXT)) |>
  distinct()


# group liver cancer due to hepatitis B and liver cancer due to hepatitis C
fulltext_by_condition = fulltext_by_condition |>
  dplyr::mutate(cause = ifelse(cause %in% c("Liver cancer due to hepatitis B",
                                           "Liver cancer due to hepatitis C"),
                              "Liver cancer",
                              cause)) |>
  dplyr::group_by(cause, lancet_group) |>
  dplyr::summarise(
    `False` = sum(`FALSE`),
    `True` = sum(`TRUE`)
  ) |>
  dplyr::mutate(TOTAL_STUDIES = `False` + `True`) |>
  dplyr::mutate(PROP_FULLTEXT = 100 * `True` / TOTAL_STUDIES) |>
  dplyr::arrange(desc(PROP_FULLTEXT))

# exclude "Diabetes mellitus"
fulltext_by_condition = fulltext_by_condition |>
  dplyr::filter(cause != "Diabetes mellitus")

# rename Chronic kidney disease due to diabetes mellitus type 2
# Chronic kidney disease (type 2 diabetes)
fulltext_by_condition = fulltext_by_condition |>
  dplyr::mutate(cause = ifelse(cause == "Chronic kidney disease due to diabetes mellitus type 2",
                              "Chronic kidney disease (type 2 diabetes)",
                              cause))

# repeat for Chronic kidney disease due to diabetes mellitus type 1
fulltext_by_condition = fulltext_by_condition |>
  dplyr::mutate(cause = ifelse(cause == "Chronic kidney disease due to diabetes mellitus type 1",
                              "Chronic kidney disease (type 1 diabetes)",
                              cause))


fulltext_by_condition <-
  tidyr::pivot_longer(data = fulltext_by_condition,
                      cols = !c("cause", 
                               "lancet_group", 
                               "TOTAL_STUDIES", 
                               "PROP_FULLTEXT"),
                      names_to = "FULLTEXT_STATUS",
                      values_to = "N_STUDIES")

# I-8 = Infectious diseases
# NCD-7 = Non-communicable diseases
fulltext_by_condition  = fulltext_by_condition  |>
  dplyr::mutate(lancet_group = ifelse(lancet_group == "I-8",
                                     "Infectious diseases",
                                     lancet_group)) |>
  dplyr::mutate(lancet_group = ifelse(lancet_group == "NCD-7",
                                     "Non-communicable diseases",
                                     lancet_group))


# fulltext_by_condition =
#   fulltext_by_condition |>
#   mutate(cause = factor(cause,
#                         levels = gbd_2019$cause[order(gbd_2019$val)]))

ggplot(aes(x = reorder(cause, N_STUDIES), 
           fill = FULLTEXT_STATUS,
           y = N_STUDIES),
       data = fulltext_by_condition) +
  geom_col() +
  coord_flip() +
  labs(x = "",
       y = "Number of published studies") +
       #title = "Number of published GWAS Catalog studies per condition") +
  # facet_grid(space = "free_y",
  #            scale = "free_y",
  #            rows = vars(lancet_group)) + 
  scale_fill_manual(
    name = "Full text downloaded",
    values = c(
      `True` = "dodgerblue",
      `False` = "grey"
    ),
    labels = c(
      `True` = "Y",
      `False` = "N"
    )
  ) + 
    theme_minimal(base_family = "Arial") +
  theme(
    #axis.text.y  = #element_text(size = 20), 
    #element_text(size = 8),   # 8pt for axis text
    axis.text.x  = #element_text(size = 20),   # 8pt for axis text
    element_text(size = 8),   
    axis.title.x = #element_text(size = 20),  # 10pt for labels
    element_text(size = 8),   
    axis.title.y = #element_text(size = 18), #element_blank(),
    element_blank(),
    #title        = element_text(size = 9),  # 10pt for labels
    strip.text   = #element_text(size = 18),  # 10pt for labels
    element_text(size = 8),
    legend.title = #element_text(size = 22),  # 10pt for labels
    element_text(size = 8),
    legend.text  = #element_text(size = 20),   # 8pt for legend text
    element_text(size = 8),
    legend.position = "bottom",#, #"left", # "right", # "bottom",           # Bottom left corner
    legend.margin = margin(t = 0, unit = "pt"),
    legend.box.spacing = unit(0, "pt"),
    plot.margin = margin(t = 5, 
                         r = 5, 
                         b = 5, 
                         l = 0, 
                         unit = "pt"),
    axis.text.y = element_text(size = 8, margin = margin(r = 0, unit = "pt")),
    #legend.justification = "left"
    legend.justification = c(-5, 2)  # Anchor at bottom left
    #legend.background = element_rect(fill = "white", color = "grey80"),
    #legend.margin = margin(l = -0.25, unit = "npc"),
    #legend.margin = margin(5, 5, 5, 5),
    # legend.position = "bottom"
  )

  # theme_minimal() +
  # theme(
  #   axis.text.x  = element_text(size = 20),
  #   axis.text.y  = element_text(size = 20),
  #   axis.title.x = element_text(size = 24),
  #   axis.title.y = element_blank(),
  #   title        = element_text(size = 24), 
  #   strip.text = element_text(size = 22),
  #   legend.title = element_text(size = 24),
  #   legend.text  = element_text(size = 23)
  # )

# Export to PDF with letter size (8.5 x 11 inches)
# Adjust width to maintain readability based on number of categories
ggsave(
  filename = here::here("figures/fulltext_by_condition.png"),
  width = 7.5 * 0.6,   # Letter width
  height = 11 * 0.27275,   # Letter height
  units = "in",
  device = "png" #cairo_pdf  # Better font rendering
)

```







