---
title: "Get abstract text for GWAS Catalog studies"
author: "Isobel Beasley"
date: "2025-10-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Set up 

## Required packages

```{r, message=FALSE}

library(stringr)
library(readxl)
library(dplyr)
library(stringi)
library(httr)
library(rentrez)
library(xml2)
library(jsonlite)
library(tokenizers)

```

# Getting required information from GWAS catalog

### Step 1: Get only disease studies, and select relevant columns

```{r gwas_cat_load_preprocess, message=FALSE}

## Step 1: 
# get only relevant disease studies
gwas_study_info <- data.table::fread(here::here("output/icd_map/gwas_study_gbd_causes.csv"))

gwas_study_info = gwas_study_info |>
  dplyr::rename_with(~ gsub(" ", "_", .x))

# filter out infectious diseases
gwas_study_info <- gwas_study_info |>
    dplyr::filter(!cause %in% c("HIV/AIDS",
                             "Tuberculosis",
                             "Malaria",
                             "Lower respiratory infections",
                             "Diarrhoeal diseases",
                             "Neonatal disorders",
                             "Tetanus",
                             "Diphtheria",
                             "Pertussis" ,
                             "Measles",
                             "Maternal disorders"))

gwas_study_info <-
  gwas_study_info |>
  filter(cause != "")

```

### What training set to use? 

```{r get_lancet_conditions}

# group conditions into broader categories to ensure we have enough papers in each category for sampling
# lancet_cause_mapping <- readxl::read_xlsx(here::here("data/icd/lancet_conditions_icd10.xlsx"),
#                                      sheet = 1) |>
#   dplyr::rename_with(~ gsub(" ", "_", .x))
# 
# gwas_study_info <- gwas_study_info |>
#   left_join(lancet_cause_mapping |> select(cause = mapped_gbd_term,
#                                            lancet_condition), 
#             by = "cause",
#             relationship = "many-to-many") 

gwas_study_info <- gwas_study_info |>
  select(PUBMED_ID, YEAR, lancet_condition) |>
  distinct() |>
  group_by(PUBMED_ID) |>
  summarise(
    YEAR = paste(unique(YEAR), collapse = "|"),
    lancet_condition = paste(sort(unique(lancet_condition)), collapse = " & ")
  ) |>
  mutate(lancet_condition_group = case_when(
    str_detect(pattern = "&", lancet_condition) ~ "Multiple",
    TRUE ~ lancet_condition
  ))

```

```{r training_set_features}

# get medcpt article embeddings for all disease studies
embeddings <- data.table::fread(here::here("output/clustering/medcpt_embeddings.csv"))

PCA_test <- prcomp(embeddings, center = TRUE, scale. = TRUE) 

pca_table <-
data.frame(PC1 = PCA_test$x[,1], 
           PC2 = PCA_test$x[,2], 
           PC3 = PCA_test$x[,3], 
           PC4 = PCA_test$x[,4], 
           PC5 = PCA_test$x[,5], 
           PC6 = PCA_test$x[,6], 
           PC7 = PCA_test$x[,7], 
           PC8 = PCA_test$x[,8], 
           PC9 = PCA_test$x[,9], 
           PC10 = PCA_test$x[,10],
           PUBMED_ID = embeddings$PUBMED_ID) 


library(ggplot2)

var_exp <- PCA_test$sdev^2 / sum(PCA_test$sdev^2)

data.frame(PC = paste0("PC", 1:length(var_exp)),
           PC_n = 1:length(var_exp),
           Variance_Explained = var_exp) |>
  filter(PC_n <= 10) |>
  mutate(PC_n = factor(PC_n, levels = 1:10)) |>
  ggplot(aes(x = PC_n, y = Variance_Explained)) +
  geom_col() +
  labs(x = "Principal Component", y = "Proportion of Variance Explained", title = "Variance explained by each principal component") + 
  theme_bw()

  pca_table |>
  left_join(gwas_study_info |> 
  select(PUBMED_ID, YEAR, lancet_condition), 
  by = "PUBMED_ID") |>
  mutate(YEAR = as.factor(YEAR)) |>
  ggplot(aes(x = PC2, 
  y = PC1, 
  fill = YEAR, 
  color = YEAR)) +
  geom_point() +
  scale_color_manual(values = viridisLite::plasma(n = 19)) +
  #geom_smooth(method = "loess") +
  labs(x = "PC2", y = "PC1", title = "PCA of GWAS Catalog abstract embeddings, colored by publication year") + 
  theme_bw()
  
  pca_table |>
  left_join(gwas_study_info |> 
  select(PUBMED_ID, YEAR, lancet_condition_group), 
  by = "PUBMED_ID") |>
  mutate(YEAR = as.factor(YEAR)) |>
  ggplot(aes(x = PC2, 
  y = PC1, 
  color = lancet_condition_group)) +
  geom_point() +
  scale_color_brewer(palette = "Set2") + 
  labs(x = "PC2", 
  y = "PC1", title = "PCA of GWAS Catalog abstract embeddings, colored by lancet condition") + theme_bw()

pca_table |>
    left_join(gwas_study_info |> 
                  select(PUBMED_ID, YEAR, lancet_condition_group), 
              by = "PUBMED_ID") |>
    mutate(YEAR = as.factor(YEAR)) |>
    ggplot(aes(x = PC3, 
               y = PC2, 
               color = lancet_condition_group)) +
    geom_point() +
    scale_color_brewer(palette = "Set2") + 
    labs(x = "PC3", 
         y = "PC2",
         title = "PCA of GWAS Catalog abstract embeddings, colored by lancet condition") + 
    theme_bw()
  

pca_table |>
    left_join(gwas_study_info |> 
                  select(PUBMED_ID, YEAR, lancet_condition), 
              by = "PUBMED_ID") |>
    mutate(YEAR = as.factor(YEAR)) |>
    ggplot(aes(x = PC3, 
               y = PC2, 
               color = YEAR)) +
    geom_point() +
    scale_color_manual(values = viridisLite::plasma(n = 19)) +
    labs(x = "PC3", 
         y = "PC2",
         title = "PCA of GWAS Catalog abstract embeddings, colored by lancet condition") + 
    theme_bw()
  
```

```{r get_training_sample}

# set.seed(500)

# training_sample = gwas_study_info |>
#   filter(!c(PUBMED_ID %in% original_training)) |>
#   select(YEAR, PUBMED_ID,
#          lancet_condition) |>
#   tidyr::separate_longer_delim(lancet_condition, delim = " & ") |>
#   distinct() |>
#   dplyr::group_by(YEAR,lancet_condition) |>
#   dplyr::slice_sample(n = 1) |>
#   pull(PUBMED_ID) |>
#   unique()

# pmids <- training_sample

pmids <- gwas_study_info$PUBMED_ID |> unique()

# don't bother with already annotated data: 
annotated <- stream_in(file(here::here("output/doccano/doccano_pt7_v5.jsonl")))
annotated_pmids <- annotated$pubmed_id |> unique()
pmids <- pmids[!c(pmids %in% annotated_pmids)]

gwas_study_info =
  gwas_study_info |>
  filter(PUBMED_ID %in% pmids)

print("Number of selected training pubmed ids for disease studies:")
gwas_study_info$PUBMED_ID |> unique() |> length()

```

#### How much overlap with abstracts and full-texts retrieved? 

```{r check_abstract_overlap}

converted_ids <- data.table::fread(here::here("output/fulltexts/pmid_to_pmcid_mapping.csv"))

full_text_files <-
  list.files(here::here("output/fulltexts"), 
             recursive = T, 
             pattern = "\\.html$|\\.xml$") 

full_text_files <- basename(full_text_files) |> 
  stringr::str_remove_all("\\.html$|\\.xml$") |>
  unique()

# convert pmcids to pmids 
converted_fulltext_pmcids <-
  converted_ids |>
  filter(pmcids %in% full_text_files) |>
  pull(PMID) |>
  unique() 

full_text_files  <- c(full_text_files, 
                      converted_fulltext_pmcids)

full_text_pmids <- grep("PMC", 
                        full_text_files, 
                        invert = T, 
                        value = T)


print("Number of training sample papers with full text retrieved:")
sum(pmids %in% full_text_pmids)

```

### Step 2: Get list of terms to use for cohort context matching

#### Source of terms: GWAS Catalog Ancestry Metadata

GWAS ancestry data.frame contains: country of recruitment, ancestry label, population descriptors

```{r gwas_cat_load_ncestry_info}

gwas_ancest_info <-  data.table::fread(here::here("data/gwas_catalog/gwas-catalog-v1.0.3.1-ancestries-r2025-07-21.tsv"),
                           sep = "\t",
                           quote = "")

gwas_ancest_info = gwas_ancest_info |>
  dplyr::rename_with(~ gsub(" ", "_", .x))

```

##### Countries 

```{r get_countries_gwas_cat, eval = F}

countries <- 
gwas_ancest_info$COUNTRY_OF_RECRUITMENT |>
  stringr::str_split("\\,") |>
  unlist() |>
  str_trim() |>
  unique() 
  
countries <- countries[countries != "NR"]
countries <- c(countries, 
               "United States", 
               "United Kingdom",
               "Korea",
               "America",
               "USA",
               "United States of America",
               "Latin America")


```

##### Population descriptors

```{r get_population_descriptors_gwas_cat, eval = F}

population_descriptors <- 
  c(gwas_ancest_info$BROAD_ANCESTRAL_CATEGORY, gwas_ancest_info$ADDITIONAL_ANCESTRY_DESCRIPTION) |>
  str_replace_all("Recruitment in North, Central, and South America, Asia, and Europe", 
                  "North America, Central America, Asia, Europe") |>
  #stringr::str_split("\\,|;|heritage:\\|") |>
  stringr::str_split("\\,|;|heritage:|\\(|\\)|\\/|\\ - |\\: |\\band\\b") |>
  unlist() |>
  str_trim() 

# processing population descriptors to remove extra information and get more general terms to match cohort descriptions in abstracts (e.g. "European" instead of "European (non-Finnish)")
population_descriptors <- population_descriptors |>
  str_remove_all("\\.$")  |> # Escape the dot to make it literal
  str_remove_all("^and ") |> 
  str_remove_all(" \\(founder/genetic isolate\\)$|\\(founder/genetic isolate$") |>
  #str_remove_all("^[0-9+]% ") |>
  str_remove_all("^\\d+% ") |>  # \d is shorthand for [0-9]
  str_remove_all("^[‡†*]+") |> # Remove dagger symbols at start 
  str_remove_all(" - Korea Association Resource \\(KARE\\)$") |>
  str_remove_all("or unknown ancestry$|\\(middle eastern$") |>
  str_remove_all("\\(see Springelkamp 2017\\)") |>
  str_remove_all("parents & grandparents born in ") |>
  str_remove_all(" cell lines$| cell line$") |>
  str_remove_all(" cohort$") |>
  str_remove_all(" population$") |> 
  str_remove_all("^Erasmus Rucphen in |^Erasmus Rucphen") |>
  str_remove_all("^including ") |>
  str_remove_all("\\(Middle Eastern$") |>
  unique() 

# exclude population descriptors containing numbers 
population_descriptors <- population_descriptors[!str_detect(population_descriptors, "\\d")]

# exclude uninformative population descriptors 
uninform_pop_desc <- c("NR", "N.R", 
                       "Other", "other", 
                       "", "unspecified", 
                       "Mixed", "mixed", "Multiple", "multiple", 
                       "Split", "split") 

population_descriptors <- population_descriptors[!c(population_descriptors %in% uninform_pop_desc)]

# exclude disease descriptors that are not valid to match cohort descriptions in abstracts (e.g. "Asthma", "Type 2 diabetes")
disease_descriptors <- grep("cases|Bipolar|controls|Schizophrenia|disorder", 
                            population_descriptors, 
                            ignore.case = T,
                            value = T)

population_descriptors <- population_descriptors[!(population_descriptors %in% disease_descriptors
                                                     )]

# exclude cohort names 
cohort_names <- c("UKB", "UKBB", "DECODE", "Controls from UKBiobank", "UKBB and deCODE", 
                  "NHAPC", "GeneID-I", " Family", "KARE", "Mayo")

population_descriptors <- population_descriptors[!(population_descriptors %in% cohort_names)]

# exclude weird paper related terms 
grep("for|see|et al", population_descriptors, ignore.case = T, value = T)

population_descriptors <- population_descriptors[!grepl(population_descriptors, "for|see", ignore.case = T)]

# add directions
directions <- c("North", 
                "South",
                "East", 
                "West",
                "Central",
                "Southeast",
                "Northeast",
                "Northwest",
                "Southwest")

population_descriptors <- c(population_descriptors, 
                            "multiracial")

population_descriptors <- unique(c(population_descriptors, directions))

population_descriptors <- population_descriptors |> str_trim() |> unique()

```

#### Source of terms: hancestro ontology of human ancestry terms

```{r get_hancestro_terms, eval = F}

# Get all terms in one go (if total < 500)
response <- GET("https://www.ebi.ac.uk/ols4/api/ontologies/hancestro/terms?size=1000")
data <- fromJSON(content(response, "text"), flatten = TRUE)
hancestro_terms <- data$`_embedded`$terms$label

# Remove terms with specific biobanks / datasets in brackets, as these are unlikely to be used in cohort descriptions in abstracts / texts
hancestro_terms <- str_remove_all(hancestro_terms, 
                                  " \\(SGDP\\)$| \\(1KGP\\)$| \\(HGDP\\)$| \\(GGVP\\)$")

# Remove obsolete terms
hancestro_terms <- grep("obsolete|obsolescence", 
                        hancestro_terms, 
                        value = TRUE, 
                        invert = TRUE)

# remove terms with specific regions in brackets, as these are unlikely to be used in cohort descriptions in abstracts / texts
hancestro_terms <- str_remove_all(hancestro_terms, pattern = "\\(Carmel\\)$|\\(pre1989\\)$|\\(Bergamo\\)$|\\(Negev\\)$|\\(Central\\)|\\(Caucasus\\)")

# replace underscores with spaces
hancestro_terms <- str_replace_all(hancestro_terms, 
                                   pattern = "_", 
                                   replacement = " ")

hancestro_terms <- unique(hancestro_terms)

# specific hancestro terms to exclude: 
terms_to_exclude <- c("ancestry category",
                      "ancestry status",
                      "BFO 0000006",
                      "Country",
                      "continent",
                      "continuant",
                      "curation status specification",
                      "denotator type",
                      "entity",
                      "ethnicity category",
                      "ethnicity descriptor",
                      "geographic location",
                      "geographic descriptor",
                      "geography-based population category",
                      "immaterial entity",
                      "independent continuant",
                      "!Kung",
                      "material entity",
                      "quality",
                      "reference population",
                      "region",
                      "organization",
                      "population",
                      "specifically dependent continuant",
                      "Thing",
                      "uncategorised population",
                      "undefined ancestry population")

hancestro_terms <- hancestro_terms[!c(hancestro_terms %in% terms_to_exclude)]

hancestro_terms <- 
hancestro_terms |>
  str_split("\\(|\\)") |>
  unlist() |>
  str_trim() |>
  unique() 

```

#### ? Other possible sources

Country R packages 
Population descriptors in: https://pmc.ncbi.nlm.nih.gov/articles/PMC8715140/#sec2


#### Terms to identify sentence identifying / describing cohort

```{r get_cohort_descriptor_terms,}

participant_words <- c("adults", "adolescents", "children", "infants", "neonates",
                       "women", "men", "female", "male",
                       "trios", "families", "twins", "siblings", 
                        "cases", "probands", "affected", 
                        "controls", "unaffected", "healthy",                        
                       "genomes",  "individuals", "patients",  "participants", 
                       "subjects", "samples", "donors", "volunteers")

# words that indicate sample or cohort is being discussed:
sample_terms <- c(
  participant_words,
  # demographic / population descriptor terms
  "ancestry", "ancestral", "origin", "descent",  "lineage",  "heritage", 
  "admixed", "admixture",
  "ethnic", "ethnicity",  "multi-ethnic",
  "racial", "multi-racial", "multiracial", "race",
  "demographic", "self-identified", "self-reported",
  "age", "aged", "years",
  "founder", "isolate", "isolated", "population", "populations",
  
  # cohort/sample related terms
  "biobank", "cohort", "cohorts", "consortium", "consortia",
  "registry", "sample", "study", "trial",
  
  # enrollment related terms
   "enroll", "enrolled", "enrollment", "enrollees",
   "recruitment", "recruited", 
   "volunteer", "volunteered", 
   "sampling", "sampled", 
   "ascertained", "residing",
  
  # Types of study
  "case-control", "prospective", "longitudinal", "cross-sectional",  "nationwide",
  "meta-analysis", "pooled",
  "replication", "discovery", "validation",
  
  # Subsets 
  "subgroup", "subset", "subcohort"
  
)


```

#### Combine sources 

```{r combine_cohort_context_terms, eval = F}

# combine terms from different sources, and add some extra terms based on looking at cohort descriptions in abstracts
cohort_context_terms <-
c(countries, 
  hancestro_terms, 
  population_descriptors,
  sample_terms) |> 
  unique() |>
  sort()


# more terms to exclude:
cohort_context_terms <- cohort_context_terms[!cohort_context_terms %in% c("San", "Split", "Han")]

# terms to remove ... "Qatar", "UK", "Taiwan", "Korean"
# cohort_context_terms <- cohort_context_terms[!cohort_context_terms %in% c("population", 
#                                                                           "mixed",
#                                                                           "split")]

# replace brackets with \\( & and \\)
cohort_context_terms <- cohort_context_terms |> 
  str_replace_all(pattern = "\\(", replacement = "\\\\(") |>
  str_replace_all(pattern = "\\)", replacement = "\\\\)")

# cohort_context_terms <- c(cohort_context_terms, 
#                           "UK\\b(?! Biobank)",
#                           "UK\\b\\(?! Biobank\\)",
#                           "Qatar\\b\\(?! Biobank\\)",
#                           "Taiwan\\b\\(?! Biobank\\)",
#                           "Korean\\b\\(?! Biobank\\)"
#                           )


# remove single character terms
cohort_context_terms <- cohort_context_terms[nchar(cohort_context_terms) > 2]

# sort by length of term (longest first) to match longest names first
cohort_context_terms <- cohort_context_terms[order(-nchar(cohort_context_terms))]

writeLines(cohort_context_terms, 
           here::here("output/gwas_cat/ancestry_population_terms.txt"))

```

```{r load_ancestry_population_terms}

cohort_context_terms <- readLines(here::here("output/gwas_cat/ancestry_population_terms.txt"))

print("Number of cohort context terms:")
length(cohort_context_terms)

```

### Step 3: Get cohort information

```{r gwas_cat_add_cohort_info}

gwas_study_info_cohort =
  data.table::fread(here::here("output/gwas_cohorts/gwas_cohort_name_corrected.csv"))

gwas_study_info_cohort =
  gwas_study_info_cohort |>
  dplyr::rename_with(~ gsub(" ", "_", .x))

gwas_study_info_cohort =
  gwas_study_info_cohort |>
  select(PUBMED_ID,
         COHORT) |>
  distinct()

print("Check adding cohort information has only added columns, not rows:")
print("Before adding:")
dim(gwas_study_info)

gwas_study_info =
  left_join(gwas_study_info,
            gwas_study_info_cohort,
            by = "PUBMED_ID"
  )

print("After adding:")
dim(gwas_study_info)


print("Check adding cohort information has not increaed number of unique pubmed ids:")
gwas_study_info |>
  pull(PUBMED_ID) |>
  unique() |>
  length()

print("Number of unique pubmed ids for disease studies with cohort info:")
gwas_study_info |>
  filter(COHORT != "") |>
  pull(PUBMED_ID) |>
  unique() |>
  length()

```


### Step 4: Add ancestry/population info

```{r gwas_cat_add_ancestry_info}

gwas_ancest_info =
  gwas_ancest_info |>
  select(PUBMED_ID,
         BROAD_ANCESTRAL_CATEGORY,
         COUNTRY_OF_RECRUITMENT) |>
  distinct() |>
  group_by(PUBMED_ID) |>
  summarise(
    BROAD_ANCESTRAL_CATEGORY = paste(
      unique(
        unlist(strsplit(BROAD_ANCESTRAL_CATEGORY, split = "\\|"))
      ),
      collapse = "|"
    ),
    COUNTRY_OF_RECRUITMENT = paste(
      unique(
        unlist(strsplit(COUNTRY_OF_RECRUITMENT, split = "\\|"))
      ),
      collapse = "|"
    )
  )

print("Check adding ancestry information has only added columns, not rows:")
print("Before adding:")
dim(gwas_study_info)

gwas_study_info =
  left_join(gwas_study_info,
            gwas_ancest_info,
            by = "PUBMED_ID"
  )

print("After adding:")
dim(gwas_study_info)

```

### Step 5: Extract information required to get paper text & abstracts

```{r gwas_cat_extract_info}

gwas_study_info <-
  gwas_study_info |>
  #filter(COHORT != "") |>
  select(PUBMED_ID,
         COHORT,
         YEAR,
         BROAD_ANCESTRAL_CATEGORY,
         COUNTRY_OF_RECRUITMENT) |>
  distinct() |>
  group_by(PUBMED_ID,
           YEAR,
           BROAD_ANCESTRAL_CATEGORY,
           COUNTRY_OF_RECRUITMENT) |>
  summarise(
    COHORT = paste(
      unique(
        unlist(strsplit(COHORT, split = "\\|"))
      ),
      collapse = "|"
    )
  )

gwas_study_info = 
  gwas_study_info |>
  ungroup() |>
  arrange(PUBMED_ID)

pmids = gwas_study_info$PUBMED_ID

cohort = gwas_study_info$COHORT
names(cohort) = pmids

date = gwas_study_info$YEAR
names(date) = pmids

country = gwas_study_info$COUNTRY_OF_RECRUITMENT
names(country) = pmids

ancestry = gwas_study_info$BROAD_ANCESTRAL_CATEGORY
names(ancestry) = pmids

print("Number of papers without cohort information:")
gwas_study_info |> 
  filter(COHORT == "") |>
  nrow()

print("Number of papers with cohort information:")
gwas_study_info |>
  filter(COHORT != "") |>
  nrow()

```

### Step 6: Get data dictionary of cohort names 

```{r get_cohort_names}

# this xlsx was built from looking at acrynyms / cohort names in the gwas catalog
# and finding the corresponding full names / details of cohorts

cohort_names <- readxl::read_xlsx(here::here("data/cohort/cohort_desc.xlsx"),
                                 sheet = 1) |>
  mutate(across(everything(), 
                ~stringr::str_replace_all(.x,
                                          pattern = "\u00A0",
                                          replacement = " "))
         ) 


cohort_full_names = cohort_names$full_name[!is.na(cohort_names$full_name)]
cohort_full_names <- str_trim(cohort_full_names)
cohort_full_names <- iconv(cohort_full_names, to = "UTF-8")
cohort_full_names <- gsub("[\u00A0\r\n]", " ", cohort_full_names)  # replace non-breaking spaces, CR, LF with space
cohort_full_names <- str_squish(cohort_full_names)  # trims and removes extra spaces
# sort by length of name (longest first) to match longest names first
cohort_full_names <- cohort_full_names[order(-nchar(cohort_full_names))]
cohort_full_names <- cohort_full_names[cohort_full_names != "Not Reported"]
cohort_full_names <- unique(cohort_full_names)

cohort_full_names <- cohort_full_names[cohort_full_names != "Japan"]

print("Number of unique cohort full names:")
length(cohort_full_names)


cohort_abbr_names = cohort_names$cohort[!is.na(cohort_names$cohort)]
cohort_abbr_names <- str_trim(cohort_abbr_names)
cohort_abbr_names <- iconv(cohort_abbr_names, to = "UTF-8")
cohort_abbr_names <- gsub("[\u00A0\r\n]", " ", cohort_abbr_names)  # replace non-breaking spaces, CR, LF with space
cohort_abbr_names <- str_squish(cohort_abbr_names)  # trims and removes extra spaces

# remove abbreviations that are too short
# cohort_abbr_names <- cohort_abbr_names[nchar(cohort_abbr_names) >= 4]
# small_abbr_to_keep <- c("C4D", 
#                         "BBJ", 
#                         "UKB", 
#                         "MVP", 
#                         "TWB", 
#                         "QBB",
#                         "MEC",
#                         "WHI"
#                         )

# cohort_abbr_names <- unique(c(cohort_abbr_names, 
#                               small_abbr_to_keep
#                               ))

cohort_abbr_names <- cohort_abbr_names[!str_detect(pattern = "\\?", 
                                                   cohort_abbr_names)]
# sort by length of name (longest first) to match longest names first
cohort_abbr_names <- cohort_abbr_names[order(-nchar(cohort_abbr_names))]


# add cohort names from GWAS catalog not yet added to data-dictionary
gwas_cat_cohorts = gwas_study_info_cohort$COHORT
gwas_cat_cohorts = unlist(strsplit(gwas_cat_cohorts, "\\|"))
gwas_cat_cohorts = gwas_cat_cohorts[!(gwas_cat_cohorts %in% c("", "other", "multiple"))]

cohort_abbr_names = unique(c(cohort_abbr_names,
                           gwas_cat_cohorts))

# remove small abbreviations that are likely to be false positives
cohort_abbr_names <- cohort_abbr_names[cohort_abbr_names != "UK"]
cohort_abbr_names <- cohort_abbr_names[cohort_abbr_names != "US"]
cohort_abbr_names <- cohort_abbr_names[cohort_abbr_names != "NZ"]
cohort_abbr_names <- cohort_abbr_names[cohort_abbr_names != "DN"]
cohort_abbr_names <- cohort_abbr_names[cohort_abbr_names != "CHB"]
cohort_abbr_names <- cohort_abbr_names[cohort_abbr_names != "FG"]
cohort_abbr_names <- cohort_abbr_names[cohort_abbr_names != "HTN"]
cohort_abbr_names <- cohort_abbr_names[cohort_abbr_names != "GD"]
cohort_abbr_names <- cohort_abbr_names[cohort_abbr_names != "MD"]
cohort_abbr_names <- cohort_abbr_names[cohort_abbr_names != "UC"]
cohort_abbr_names <- cohort_abbr_names[cohort_abbr_names != "DC"]
cohort_abbr_names <- cohort_abbr_names[cohort_abbr_names != "ACE"]
cohort_abbr_names <- cohort_abbr_names[cohort_abbr_names != "Japan"]


print("Number of unique cohort abbreviation names:")
length(cohort_abbr_names)

```

# Get cohort-relevant abstract sentences

## Load relevant abstracts

### Convert abstract sections to sentences

```{bash, eval = F}

source spacyr_venv/bin/activate

python3 code/extract_text/spacy_obtain_sentences.py \
--input_dir output/abstracts \
--output_dir output/abstracts \
--hf_model pritamdeka/S-PubMedBert-MS-MARCO

```

```{r read_abstracts_from_files}

abstract_files <- list.files(here::here("output/abstracts/"), 
                             pattern = "*.json", 
                             full.names = FALSE
                             ) |>
                   sort()

abstract_pmids = str_remove_all(abstract_files, "_sentences.json$")
abstract_pmids = abstract_pmids[abstract_pmids %in% pmids]


abstracts <- sapply(abstract_pmids,
                    function(file) {
                      json_data <- fromJSON(here::here(paste0("output/abstracts/", file, "_sentences.json")))
                      
                      abstract_lines <- json_data[json_data != ""]
                      
                      # readLines(here::here(paste0("output/abstracts/",file, ".txt")), 
                      #           warn = FALSE) |> 
                      #   paste(collapse = " ")
                    }
                    )


# check lengths of these vectors are the same
print("Check lengths of vectors are the same:")
print(paste("Length of pmids:", length(pmids)))
print(paste("Length of abstract pmids:", length(abstract_pmids)))
print(paste("Length of abstracts:", length(abstracts)))
print(paste("Length of cohort:", length(cohort)))
print(paste("Length of date:", length(date)))

```

###  Check: What abstracts are missing? 

```{r check_missing_abstracts, warning=FALSE, message=FALSE}

missing_abstracts = pmids[!(pmids %in% abstract_pmids)] 

print("Number of missing abstracts:")
length(missing_abstracts)

#pmids <- pmids[pmids %in% abstract_pmids]

```

## Cohort name matching in abstract sentences

### Remove title sentences and small sentences that are likely to contain cohort names but are not sentences describing the cohort

```{r remove_title_and_small_sentences}

# remove title from sentences (as this is likely to contain cohort names but is not a sentence describing the cohort)

has_balanced_parens <- function(s) {
  stringr::str_count(s, fixed("(")) == stringr::str_count(s, fixed(")"))
}

get_title_sentences <- function(x, n_words_threshold = 5) {
                 
                 # get short sentences < 6 words
                 sentences <- x[stringr::str_count(x, " ") < n_words_threshold]
                 
                 # catch / inspect title sentences 
                 # upper_case <- "(?m)^\\s*[A-Z][A-Z\\s]+$"
                 
                 # get sentences that start with capital letter
                 sentences <- sentences[grepl("^(\\d+(\\.\\d+)*\\s+)?[A-Za-z]", sentences)]
                  #sentences[grepl("^[A-Z]", sentences)]
                 
                 # exclude sentences containing unmatched bracket as these are more likely to be complex sentences that are less likely to be titles or headings
                 #sentences <- sentences[!grepl("\\(|\\)", sentences)]
                 sentences <- sentences[has_balanced_parens(sentences)]
                 
                 # exclude sentences that end with a comma, as these are more likely to be incomplete sentences that are part of a larger sentence, and less likely to be titles or headings
                 sentences <- sentences[!grepl(",$|\\($", sentences)] 
                 
                 # exclude sentences that contain urls
                 sentences <- sentences[!grepl("www.|http", sentences)]
                 
                 return(sentences)
                 
}

get_non_title_sentences <- function(x) {
  
                title_sentences <- get_title_sentences(x)
                 
                keep_non_title_sentences <- x[!c(x %in% title_sentences)]
                 
                return(keep_non_title_sentences)
}

title_sentences <- purrr::map(abstracts,
                              get_title_sentences
                              ) |>
                              # ~get_title_sentences(.x, 
                              #                     n_words_threshold = 3)) |>
                   purrr::list_c() |>
                   sort() |>
                   unique()


print("Title sentences identified:")
print(title_sentences)


abstracts <-
purrr::map(abstracts,
           get_non_title_sentences        
                 )


```

### Function to extract sentences with cohort names

```{r fn_extract_cohort_sentences}

extract_cohort_sentences <- function(text_list, # list of texts to search through (e.g. abstracts, or full texts)
                                     terms, # terms to grep search for
                                     column_name = "COHORT",
                                     tokenize = FALSE,
                                     ignore_case) {
  
  #word boundary or preceded by ( [ ,
  terms_grep <- paste0("(?<=\\b|\\(|\\[)",
                            terms,
                            "(?=\\b|,|\\(|\\])")
  
  results <- lapply(seq_along(text_list), function(i) {
    
    # Split text into sentences, if not done already
    if(tokenize) {
      sentences <- tokenizers::tokenize_sentences(test_list[[i]])[[1]]
    } else {
      sentences <-  text_list[[i]]
    }
    
    # For each sentence, find all matching cohort names
    lapply(seq_along(sentences), function(s) {
      
      sentence <- sentences[s]
      
      # Identify cohort names present in this sentence
      matched_cohorts <- terms[str_detect(sentence, 
                                          regex(terms_grep, 
                                                ignore_case = ignore_case)
                                          )
                               ]
      term_df <-
      data.frame(
        article_id = i,
        sentence_id = s,
        sentence = sentence,
        stringsAsFactors = FALSE
      )
      
      # Add column with matched term names (if any)
      term_df[[column_name]] <- ifelse(length(matched_cohorts) > 0,
                                       str_flatten(unique(matched_cohorts), 
                                                   collapse = "|", 
                                                   na.rm = T),
                                       ""
      )
      
      # Add column with binary label for whether any terms were matched
      term_df[[paste0("has_", tolower(column_name))]] <- ifelse(length(matched_cohorts) > 0, 
                                                                TRUE, 
                                                                FALSE)
      
      return(term_df)
      
    }) |> bind_rows()
  })

 results <- bind_rows(results)
 
 results$pubmed_id <- names(text_list)[results$article_id]
 
 return(results) #returns a df of sentences, with labelled columns for whether they contain cohort names, and which cohort names they contain (if any)
 # id of sentence, and abstract / article
}

```

### Get cohort relevant sentences 

#### Get cohort name sentences

```{r extract_cohort_sentences_from_abstracts}

# extract sentences with cohort full names (case-insensitive matching, as full names are less likely to be ambiguous)
cohort_sentences_df_p1 <- extract_cohort_sentences(abstracts,
                                                cohort_full_names,
                                                ignore_case = TRUE
                                                )

# extract sentences with cohort abbreviation names (case-sensitive matching, as abbreviations are more likely to be ambiguous)
cohort_sentences_df_p2 = extract_cohort_sentences(abstracts,
                                                  cohort_abbr_names,
                                                  ignore_case = FALSE
                                                  )

cohort_sentences_df =
  bind_rows(cohort_sentences_df_p1,
            cohort_sentences_df_p2
            )


cohort_sentences_df = 
  cohort_sentences_df |>
  distinct()

separate_cohorts <- function(COHORT) {
  
  if (any(grepl("\\|", COHORT))) {
    return(unlist(strsplit(COHORT, "\\|")))
  } else {
    return(COHORT)
  }
  
}

cohort_sentences_df =
  cohort_sentences_df |>
  group_by(pubmed_id,
           article_id, 
           sentence_id, 
           sentence) |>
  summarise(
    COHORT = str_flatten(unique(separate_cohorts(COHORT)), 
                         collapse = "|", 
                         na.rm = T),
    has_cohort = ifelse(any(COHORT != ""), TRUE, FALSE)
  ) |>
  ungroup() 

cohort_sentences_df =
  cohort_sentences_df |>
  mutate(COHORT = str_remove_all(COHORT, 
                               pattern = "\\|$|^\\|"
                               )
         ) 

```

```{r cohort_sentence_summary}

print("Number of abstracts containing probable cohort reference")
abstracts_cohorts <- cohort_sentences_df |>
  filter(COHORT != "") |>
  pull(article_id) |>
  unique() 

n_abstracts_cohorts =  abstracts_cohorts |>
  length()

print(n_abstracts_cohorts)

print("Percentage of sampled abstracts containing probable cohort reference:")
100 * n_abstracts_cohorts / length(unique(pmids))

print("Number of sentences containing probable cohort reference")
  cohort_sentences_df |>
  filter(COHORT != "") |>
  select(article_id, sentence_id) |>
  distinct() |>
  nrow()
  
print("Number of entities")
cohort_sentences_df |>
  filter(COHORT != "") |>
  pull(COHORT) |>
  str_split("\\|") |>
  unlist() |>
  length()
  
print("Number of unique cohorts referenced (not unique cohort names, but unique strings in the COHORT column):")
detected_cohorts <- cohort_sentences_df |>
  filter(COHORT != "") |>
  pull(COHORT) |>
  str_split("\\|") |>
  unlist() 

n_cohorts <- detected_cohorts |>
  unique()  |>
  length()

print(n_cohorts)

print("Most common cohort names detected (by number of sentences they are mentioned in):")
data.frame(cohort = detected_cohorts) |>
  group_by(cohort) |>
  summarise(n = n()) |>
  arrange(desc(n)) |>
  head(10)

print("Most common cohort names detected (by number of abstracts they are mentioned in):")
cohort_sentences_df |>
  filter(COHORT != "") |>
  select(COHORT, article_id) |>
  tidyr::separate_rows(COHORT, sep = "\\|") |>
  distinct() |>
  group_by(COHORT) |>
  summarise(n_abstracts = n()) |>
  arrange(desc(n_abstracts)) |>
  head(5)
  
```

#### Get cohort-related context sentences

```{r extract_cohort_context_sentences_from_abstracts}

all_cohort_terms <- c(cohort_full_names, cohort_abbr_names)

# mask cohort names in methods text
# so that extract population descriptors etc. are not masked by cohort name matching
cohort_patterns <- 
  all_cohort_terms |>
  unique()

cohort_patterns <- cohort_patterns[order(-nchar(cohort_patterns))] # order by length, to avoid shorter patterns matching within longer patterns (e.g. "UK Biobank" and "Biobank")

cohort_patterns  <- cohort_patterns |> 
  stringr::str_escape()   # or: gsub("([][{}()+*^$|\\\\?.])", "\\\\\\1", x)

cohort_patterns <- paste0("(?<=\\b|\\(|\\[)",
                            cohort_patterns,
                            "(?=\\b|,|\\(|\\])")

mask_re <- paste0("\\b(?:",
                  paste(cohort_patterns,
                        collapse = "|"),
                  ")\\b")


masked_abstracts <- purrr::map(abstracts,
           function(x) {
             stringr::str_replace_all(x, 
                                      stringr::regex(mask_re, ignore_case = TRUE), 
                                      "\u241F"
                                      )
           }
           )

# extract sentences with cohort context terms (case-sensitive matching, as some context terms are common words that could be ambiguous)
cohort_context_df = extract_cohort_sentences(masked_abstracts,
                                            cohort_context_terms,
                                            column_name = "CONTEXT",
                                            ignore_case = FALSE
                                            )

cohort_sentences_df <-
left_join(cohort_sentences_df,
          cohort_context_df |> select(pubmed_id,
                                      article_id, 
                                      sentence_id, 
                                      CONTEXT,
                                      has_context),
          by = c("pubmed_id",
                 "article_id", 
                 "sentence_id")
          ) 

print("Number of sentences with cohort names and/or cohort-related context:")
cohort_sentences_df |> 
  group_by(has_cohort, has_context) |> 
  summarise(n = n())

filtered_cohort_sentences_df =
  cohort_sentences_df |>
  filter(COHORT != "" | CONTEXT != "")

print("Number of sentences with cohort names or cohort-related context (population, ancestry, country terms):")
nrow(filtered_cohort_sentences_df)

```


#### Get hard negatives 

These are 'semanitically similar' sentences to cohort names. 

```{r save_sentences}

json_list <- 
cohort_sentences_df |>
  filter(has_cohort) |>
  pull(sentence) |>
  as.list()

# Write to JSON file
writeLines(toJSON(json_list,
                  auto_unbox = TRUE,
                  pretty = TRUE),
           here::here("output/doccano/abstract_cohort_sentences.json"))


json_list <- 
cohort_sentences_df |>
  filter(!has_cohort) |>
  pull(sentence) |>
  as.list()

# Write to JSON file
writeLines(toJSON(json_list,
                  auto_unbox = TRUE,
                  pretty = TRUE),
           here::here("output/doccano/abstract_non_cohort_sentences.json"))

```

```{bash, eval = F}

source  spacyr_venv/bin/activate

python3 code/extract_text/sentence_embeddings.py \
        --label "abstracts" \
        --cohort_path "output/doccano/abstract_cohort_sentences.json" \
        --non_cohort_path "output/doccano/abstract_non_cohort_sentences.json" \
        --figure_save_path "output/doccano/abstract_sentence_similarity.png" \
        --p 75
        
```

#### Make training set

```{r add_hard_negs_to_abstract_sentences}

abstract_hard_negatives <- jsonlite::read_json(here::here("output/doccano/hard_negatives_abstracts.json"),
                                               simplifyVector = TRUE)

cohort_sentences_df  <- 
cohort_sentences_df |>
  dplyr::mutate(hard_negative = ifelse(sentence %in% abstract_hard_negatives,
                                TRUE,
                                FALSE)
  )

print("Overlap between hard negatives and sentences with cohort names or cohort-related context:")
cohort_sentences_df |>
  filter(hard_negative == TRUE) |>
  group_by(has_cohort, has_context) |>
  summarise(n = n())

# most hard negatives overlap with context sentences! (As expected, as 
# as context sentences should also be semantically similar to cohort sentences

filtered_cohort_sentences_df <-
cohort_sentences_df |>
  filter(COHORT != "" | CONTEXT != "" | hard_negative == TRUE)

print("Size of training set:")
print("Number of sentences with cohort names or cohort-related context (population, ancestry, country terms) or hard negatives:")
nrow(filtered_cohort_sentences_df)

```

## Prepare data in Doccano JSON format

```{r prepare_doccano_json_abstracts}

# file path for intermediate json file output
json_file = here::here("output/doccano/abstracts_with_cohort_info.json")

# file path for final jsonl file output
jsonl_file = here::here("output/doccano/abstracts_with_cohort_info.jsonl")


convert_to_doccano_json_sentence_level <- function(cohort_sentences_df) {
  
  # set up json list
  doccano_list <- list()
  example_id <- 1

  for(current_sentence in cohort_sentences_df$sentence) {

      # Filter cohort sentences that match this sentence (safe matching)
      df <- cohort_sentences_df |>
             dplyr::filter(sentence == current_sentence) 
      
      for (i in seq_len(nrow(df))) {
        
        matched_cohort <- df$COHORT[i]
        pubmed_id_val  <- df$pubmed_id[i]
        date_val       <- df$date[i]
        cohort_val     <- df$cohort[i]
        country_val    <- df$country[i]
      
        if(matched_cohort == ""){
        
          doccano_list[[example_id]] <- list(
            text = current_sentence,
            pubmed_id = pubmed_id_val,
            date = date_val,
            country = country_val,
            gwas_cat_cohort_label = cohort_val,
            label = list()
          )
        
        } else {
        
          # Find the location of all matches of the cohort name in the sentence
          if(grepl("\\|", matched_cohort)) {
             
             # If multiple cohort names, separate
             matched_cohort <- unlist(strsplit(matched_cohort, split = "\\|"))
          
             match_locations <- list()
          
             for(current_matched_cohort in matched_cohort){
            
                 matches <- str_locate_all(current_sentence,
                                           fixed(current_matched_cohort, 
                                                 ignore_case = TRUE)
                                           )[[1]]
            
                 match_locations <- append(match_locations, list(matches))
             }
          
             # Combine all match locations into a single matrix
             matches <- do.call(rbind, match_locations)
          
          } else {
          
             matches <- str_locate_all(current_sentence,
                                       fixed(matched_cohort, 
                                             ignore_case = TRUE)
                                       )[[1]]
          }
          
          # Convert matches to 0-based indexing (for doccano)
          matches[, "start"] <- matches[, "start"] - 1  
          
          # Turn match locations into entity list
          entities <- list()
          
          for(k in seq_len(nrow(matches))) {
            entities <- append(entities, list(list(
              start_offset = matches[k, "start"],
              end_offset = matches[k, "end"],
              label = "COHORT"
            )))
          }
          
          # Create Doccano JSON entry
          doccano_list[[example_id]] <- list(
            text = current_sentence,
            pubmed_id = pubmed_id_val,
            date = date_val,
            country = country_val,
            gwas_cat_cohort_label = cohort_val,
            label = entities
          )
        }
      }

      example_id <- example_id + 1
  }
  
  # Convert named list label format to vector format [start, end, label]
  doccano_list <- lapply(doccano_list, function(x) {
    x$label <- lapply(x$label, function(l) {
      c(l$start_offset, l$end_offset, l$label)
    })
    x
  })

  return(doccano_list)
}


# Join metadata onto sentences before passing to function
meta_data_df <- data.frame(
  pubmed_id = as.character(pmids), 
  cohort    = cohort,
  date      = date,
  country   = country
)

filtered_cohort_sentences_df <- left_join(
  filtered_cohort_sentences_df,
  meta_data_df,
  by = "pubmed_id"
)

# Create JSON as a list
json_list <- convert_to_doccano_json_sentence_level(
  cohort_sentences_df = filtered_cohort_sentences_df
)

# Write to JSON file
writeLines(toJSON(json_list,
                  auto_unbox = TRUE,
                  pretty = TRUE),
           json_file)


json_data <- fromJSON(json_file,
                      simplifyVector = FALSE)

# Open connection to JSONL file
con <- file(jsonl_file, "w")

# Loop over each element (object) and write as one line
for (i in seq_along(json_data)) {
  writeLines(toJSON(json_data[[i]], auto_unbox = TRUE), con)
}

# Close connection
close(con)

cat("JSONL saved to:", jsonl_file, "\n")

```

# Get cohort-revelevant methods section sentence extraction

## Convert methods sections to sentences

```{bash, eval = F}

source spacyr_venv/bin/activate

python3 code/extract_text/spacy_obtain_sentences.py \
--input_dir output/methods \
--output_dir  output/methods_sentences 

```

## Load relevant methods text

```{r}

methods_sections <- list.files(here::here("output/methods_sentences/"), 
                             pattern = "*.json", 
                             full.names = TRUE
                             )

methods_sections <- methods_sections[!c(methods_sections %in% 
                                          grep("failed_files|no_methods", 
                                                methods_sections, 
                                                value = TRUE)
                                      )
                                     ]

pmcids_with_methods <- methods_sections |>
  gsub(pattern = ".*/", replacement = "") |>
  gsub(pattern = "_methods_sentences\\.json$", replacement = "") |>
  gsub(pattern = "_methods_main_sentences\\.json$", replacement = "") |>
  gsub(pattern = "_bioc", replacement = "") 

converted_pmcids <-  converted_ids |>
  filter(pmcids %in% pmcids_with_methods) |>
  pull(PMID)

pubmeds_with_methods <- c(converted_pmcids, 
                          grep("PMC", pmcids_with_methods, value = TRUE, invert = TRUE)
                          )

print("Number of papers with methods sections extracted out of training sample")
sum(pubmeds_with_methods %in% pmids) 

100 * sum(pubmeds_with_methods %in% pmids) / length(pmids)

# get corresponding files to pubmed ids 
pubmeds_to_get <- pubmeds_with_methods[pubmeds_with_methods %in% pmids]

get_file_names_df <- 
   converted_ids |>
   filter(PMID %in% pubmeds_to_get) |>
   select(-DOI) |>
   rename(pmid = PMID,
          pmcid = pmcids
          ) |>
  rowwise() |>
  
  # data.frame(pmid = pubmeds_to_get) |>
  #                    #rowwise() |>
  #                    mutate(pmcid = converted_ids |> 
  #                                    filter(PMID %in% pubmeds_to_get) |> 
  #                                       pull(pmcids)
  #                               ) |>
  mutate(file_name = ifelse(!is.na(pmcid) | pmcid != "",
                            grep(methods_sections, 
                                 pattern = pmcid, 
                                 value = TRUE),
                            grep(methods_sections,
                                 pattern = pmid, 
                                 value = TRUE)
                            )
  ) |>
  mutate(file_name = ifelse(is.na(file_name),
                            grep(methods_sections,
                                 pattern = pmid, 
                                 value = TRUE),
                            file_name
                            )) |>
  filter(!is.na(file_name)) |> 
  ungroup()

methods_texts <- sapply(get_file_names_df$file_name,
                    function(file) {
                      
                      json_data <- fromJSON(file)
                      
                      text_lines <- json_data[json_data != ""]

                                          }
                    )


names(methods_texts) <- get_file_names_df$pmid

# order methods texts by pmids
methods_texts <- methods_texts[order(names(methods_texts))]

```

### Remove title sentences

```{r remove_title_sentences}


title_sentences <- purrr::map(methods_texts,
                              get_title_sentences) |>
                   purrr::list_c() |>
                   sort() |>
                   unique()


print("Number of title sentences extracted from methods sections:")
length(title_sentences)
           

methods_texts <-
purrr::map(methods_texts,
           get_non_title_sentences        
                 )

# what small sentences are left
purrr::map(
methods_texts,
~.x[stringr::str_count(.x, " ") < 5]
) |>
  purrr::list_c()

```

### Manual checks of methods text sentences 

```{r check_methods_texts, eval = F}

# to do when more settled with methods text extraction, to check the quality of the sentences extracted from methods sections, and whether they are likely to be useful for cohort name extraction

# Check 1. 
# Do all sentences start:
# with a capital,
# Or a number 
# Or a word that is lower case, and then upper case (e.g. eQTl)
check_1_regex <- "^([A-Z]|[0-9]+|[a-z]+[A-Z])"

check_1 <- purrr::keep(methods_texts,
            ~any(!grepl(check_1_regex,
                       .x)
                 )
                 ) 

check_1 <- purrr::imap(check_1, 
           function(x, idx) {
             
             sentences_idx = grep(check_1_regex,
                                  x,
                                  invert = TRUE,
                                  value = F
                                  )
             
             just_broken_sentences_idx <- sentences_idx
             
             sentences_idx <- sort(unique(c(sentences_idx - 1,
                               sentences_idx, 
                               sentences_idx + 1)))
             
             sentences_idx <- sentences_idx[sentences_idx > 0 & sentences_idx <= length(x)]
             
             
             # idx of nearby sentences
             # sentences_idx <- sort(unique(c(sentences_idx - 2,
             #                                sentences_idx, 
             #                                sentences_idx + 2))
             #                       )
             # 
             sentences = x[sentences_idx]
             
             df = data.frame(sentence = sentences,
                        pubmed_id = idx,
                        sentence_id = sentences_idx
                        ) |>
               dplyr::mutate(broken_sentence = ifelse(sentence_id %in% just_broken_sentences_idx, TRUE, FALSE))
             
             return(df)
           }
           ) |>
  bind_rows()

print("\n \n Check the starts of sentences seem reasonable")
print("Number of papers with sentences that don't start with capital, number, or lower-upper case word:")
check_1  |> pull(pubmed_id) |> unique() |> length()

print("Number of sentences that don't start with capital, number, or lower-upper case word:")
check_1 |> filter(broken_sentence) |> nrow()

# Check 2: 
# Sentence ends with a full stop: 
# "\\.$"
check_2_regex <- "\\.$|\\?$"

check_2 <- purrr::keep(methods_texts,
            ~any(!grepl(check_2_regex,
                       .x)
                 )
                 )

check_2 <- purrr::imap(check_2,
           function(x, idx) {
             
             sentences_idx = grep(check_2_regex,
                                  x,
                                  invert = TRUE,
                                  value = F
                                  )
             
             just_broken_sentences_idx <- sentences_idx
             
             sentences_idx <- sort(unique(c(sentences_idx - 1,
                               sentences_idx, 
                               sentences_idx + 1)))
             
             sentences_idx <- sentences_idx[sentences_idx > 0 & sentences_idx <= length(x)]
             
             
             # idx of nearby sentences
             # sentences_idx <- sort(unique(c(sentences_idx - 2,
             #                                sentences_idx, 
             #                                sentences_idx + 2))
             #                       )
             # 
             sentences = x[sentences_idx]
             
             df = data.frame(sentence = sentences,
                        pubmed_id = idx,
                        sentence_id = sentences_idx
                        ) |>
               dplyr::mutate(broken_sentence = ifelse(sentence_id %in% just_broken_sentences_idx, TRUE, FALSE))
             
             return(df)
           }
           ) |>
  bind_rows()


print("\n \n Check the ends of sentences seem reasonable")
print("Number of papers with sentences that don't end with a full stop or question mark:")
check_2  |> pull(pubmed_id) |> unique() |> length()

print("Number of sentences that don't end with a full stop or question mark:")
check_2 |> filter(broken_sentence) |> nrow()

# Check 3: 
# Sentences don't have un-matched brackets
# "\\((?![^()]*\\))|(?<!\\([^()]*\\))\\)"
has_unmatched_parens <- function(x) {
  open  <- stringr::str_count(x, "\\(")
  close <- stringr::str_count(x, "\\)")
  open != close
}

check_3 <- purrr::keep(methods_texts, 
                       ~ any(has_unmatched_parens(.x)))

check_3 <- purrr::imap(check_3,
           function(x, idx) {
             
             sentences_idx = which(sapply(x, has_unmatched_parens))
             
             sentences = x[sentences_idx]
             
             df = data.frame(sentence = sentences,
                        pubmed_id = idx,
                        sentence_id = sentences_idx
                        ) |>
               dplyr::mutate(broken_sentence = TRUE) |>
               tibble::remove_rownames()
             
             return(df)
           }
           ) |>
  bind_rows()

print("\n \n Check that brackets seem reasonable")
print("Number of papers with sentences that have un-matched brackets:")
check_3  |> pull(pubmed_id) |> unique() |> length()

print("Number of sentences that have un-matched brackets:")
check_3 |> filter(broken_sentence) |> nrow()

# Check 4: 
# Sentences don't have un-matched quotation marks
# "\"([^\"]*\"[^\"]*)*\""
check_4_regex <- "\"([^\"]*\"[^\"]*)*\""

check_4 <- purrr::keep(methods_texts,
            ~any(grepl(check_4_regex,
                       .x)
                 )
                 )

check_4 <- purrr::imap(check_4,
           function(x, idx) {
             
             sentences_idx = grep(check_4_regex,
                                  x,
                                  value = F
                                  )
             
             sentences = x[sentences_idx]
             
             df = data.frame(sentence = sentences,
                        pubmed_id = idx,
                        sentence_id = sentences_idx
                        ) |>
               dplyr::mutate(broken_sentence = TRUE)
             
             return(df)
           }
           ) |>
  bind_rows()


print("\n \n Check that quotation marks seem reasonable")
print("Number of papers with sentences that have un-matched quotation marks:")
print(head(check_4))

  
# Check sentences that end with [a-z][0-9].
check_5_regex <- "[a-z][0-9].$|[A-Z][0-9].$"

check_5 <- purrr::keep(methods_texts,
            ~any(grepl(check_5_regex,
                       .x)
                 )
                 )


check_5 <- purrr::imap(check_5,
           function(x, idx) {
             
             sentences_idx = grep(check_5_regex,
                                  x,
                                  value = F
                                  )
             
             sentences = x[sentences_idx]
             
             df = data.frame(sentence = sentences,
                        pubmed_id = idx,
                        sentence_id = sentences_idx
                        ) |>
               dplyr::mutate(broken_sentence = TRUE)
             
             return(df)
           }
           ) |>
  bind_rows()


n_character_limit <- 100

check_5 <- purrr::keep(methods_texts,
            ~any((nchar(.x) > n_character_limit) & 
                   length(grep(.x, pattern = ".") > 1)
            )
                 )

check_5 <- purrr::imap(check_5,
           function(x, idx) {
             
             sentences_idx = which(nchar(x) > n_character_limit)
             
             sentences = x[sentences_idx]
             
             df = data.frame(sentence = sentences,
                        pubmed_id = idx,
                        sentence_id = sentences_idx
                        ) |>
               dplyr::mutate(broken_sentence = TRUE)
             
             return(df)
           }
           ) |>
  bind_rows()

split_regex <- "(?<!St\\.)(?<!Fig\\.)(?<!no\\.)(?<!vs\\.)(?<!i\\.e\\.)(?<!e\\.g\\.)(?<!Inc\\.)(?<!Co\\.)(?<=[a-z]\\.)\\s+(?=[A-Z])"


check_5 |>
  mutate(match_split = str_detect(sentence, 
                                  pattern = split_regex
                                  )
  ) |>
  filter(match_split) |>
  pull(sentence) |>
  stringr::str_split(pattern = split_regex)  

  

```

## Get cohort relevant sentences 

### Get cohort names sentences

```{r extract_cohort_sentences_methods}

# extract sentences with cohort full names (case-insensitive matching, as full names are less likely to be ambiguous)
cohort_sentences_df_p1 <- extract_cohort_sentences(methods_texts,
                                                cohort_full_names,
                                                ignore_case = TRUE
                                                )

# extract sentences with cohort abbreviation names (case-sensitive matching, as abbreviations are more likely to be ambiguous)
cohort_sentences_df_p2 = extract_cohort_sentences(methods_texts,
                                                  cohort_abbr_names,
                                                  ignore_case = FALSE
                                                  )

methods_cohort_sentences_df = 
  bind_rows(cohort_sentences_df_p1,
            cohort_sentences_df_p2
            ) |>
  distinct()


methods_cohort_sentences_df =
  methods_cohort_sentences_df |>
  group_by(pubmed_id, 
           article_id, 
           sentence_id, 
           sentence) |>
  summarise(
    COHORT = str_flatten(unique(separate_cohorts(COHORT)), 
                         collapse = "|", 
                         na.rm = T),
    has_cohort = ifelse(any(COHORT != ""), TRUE, FALSE)
  ) |>
  ungroup() 

methods_cohort_sentences_df =
  methods_cohort_sentences_df |>
  mutate(COHORT = str_remove_all(COHORT, 
                               pattern = "\\|$|^\\|"
                               )
         ) 

print("Number of methods sections containing probable cohort reference:")
methods_cohorts <- methods_cohort_sentences_df |>
  filter(COHORT != "") |>
  pull(article_id) |>
  unique()

methods_cohorts  |> length()

print("Percentage of methods sections containing probable cohort reference:")
100 * length(methods_cohorts) / length(methods_texts)


print("Number of sentences containing probable cohort reference in methods sections")
  methods_cohort_sentences_df |>
  filter(COHORT != "") |>
  select(article_id, sentence_id) |>
  distinct() |>
  nrow()

print("Number of entities in methods sections")
methods_cohort_sentences_df |>
  filter(COHORT != "") |>
  pull(COHORT) |>
  str_split("\\|") |>
  unlist() |>
  length()

print("Most common cohort names detected in methods sections:")
methods_cohort_sentences_df |>
  filter(COHORT != "") |>
  select(COHORT, article_id) |>
  tidyr::separate_rows(COHORT, sep = "\\|") |>
  distinct() |>
  group_by(COHORT) |>
  summarise(n_abstracts = n()) |>
  arrange(desc(n_abstracts)) |>
  head(5)

print("Number of cohort names only mentioned in one paper")
methods_cohort_sentences_df |>
  filter(COHORT != "") |>
  select(COHORT, article_id) |>
  tidyr::separate_rows(COHORT, sep = "\\|") |>
  distinct() |>
  group_by(COHORT) |>
  summarise(n_abstracts = n()) |>
  filter(n_abstracts == 1) |>
  nrow()

print("Number of unique cohorts referenced in methods sections (not unique cohort names, but unique strings in the COHORT column):")
detected_cohorts_methods <- methods_cohort_sentences_df |>
  filter(COHORT != "") |>
  pull(COHORT) |>
  str_split("\\|") |>
  unlist()

n_cohorts_methods <- detected_cohorts_methods |>
  unique()  |>
  length()

print(n_cohorts_methods)

print("Most common cohort names detected in methods sections by number of sentences they are mentioned in:")
data.frame(cohort = detected_cohorts_methods) |>
  group_by(cohort) |>
  summarise(n = n()) |>
  arrange(desc(n)) |>
  head(10)

```

### Get cohort-related context sentences 

```{r extract_cohort_context_sentences_methods}

masked_methods_text <- purrr::map(methods_texts,
           function(x) {
             stringr::str_replace_all(x, 
                                      stringr::regex(mask_re, ignore_case = TRUE), 
                                      "\u241F"
                                      )
           }
           )

# extract sentences with cohort context terms (case-sensitive matching, as some context terms are common words that could be ambiguous)
methods_cohort_context_df = extract_cohort_sentences(
                                            masked_methods_text,
                                            cohort_context_terms,
                                            column_name = "CONTEXT",
                                            ignore_case = FALSE
                                            )

methods_cohort_sentences_df <-
left_join(methods_cohort_sentences_df,
          methods_cohort_context_df |> select(pubmed_id, 
                                      article_id,
                                      sentence_id,
                                      CONTEXT,
                                      has_context), 
          by = c("pubmed_id",
                 "article_id", 
                 "sentence_id")
          ) 

print("Number of sentences with cohort names and/or cohort-related context:")
methods_cohort_sentences_df |> 
  mutate(has_context = ifelse(CONTEXT == "", F, T)) |>
  group_by(has_cohort, has_context) |> 
  summarise(n = n())

print("Most common cohort-related context terms detected in methods sections:")
methods_cohort_sentences_df |>
  filter(CONTEXT != "") |>
  select(CONTEXT, article_id) |>
  tidyr::separate_rows(CONTEXT, sep = "\\|") |>
  distinct() |>
  group_by(CONTEXT) |>
  summarise(n_abstracts = n()) |>
  arrange(desc(n_abstracts)) |>
  head(10)

print("Most common cohort-related context terms detected in methods sections by number of sentences they are mentioned in:")
methods_cohort_sentences_df |>
  filter(CONTEXT != "") |>
  select(CONTEXT, article_id) |>
  tidyr::separate_rows(CONTEXT, sep = "\\|") |>
  distinct() |>
  group_by(CONTEXT) |>
  summarise(n = n()) |>
  arrange(desc(n)) |>
  head(10)



```

### Get hard negatives

```{r}

json_list <- 
methods_cohort_sentences_df  |>
  filter(has_cohort) |>
  pull(sentence) |>
  as.list()

# Write to JSON file
writeLines(toJSON(json_list,
                  auto_unbox = TRUE,
                  pretty = TRUE),
           here::here("output/doccano/methods_cohort_sentences.json"))


json_list <- 
methods_cohort_sentences_df  |>
  filter(!has_cohort) |>
  pull(sentence) |>
  as.list()

# Write to JSON file
writeLines(toJSON(json_list,
                  auto_unbox = TRUE,
                  pretty = TRUE),
           here::here("output/doccano/methods_non_cohort_sentences.json"))

```

```{bash, eval = F}

source  spacyr_venv/bin/activate

python3 code/extract_text/sentence_embeddings.py \
        --label "methods" \
        --cohort_path "output/doccano/methods_cohort_sentences.json" \
        --non_cohort_path "output/doccano/methods_non_cohort_sentences.json" \
        --figure_save_path "output/doccano/methods_sentence_similarity.png"
        
```

### Make final traing set

sentences with words that indicate sample or cohort is being discussed

```{r get_sample_context_sentences_methods}

methods_hard_negatives <- jsonlite::read_json(here::here("output/doccano/hard_negatives_methods.json"),
                                               simplifyVector = TRUE)

methods_cohort_sentences_df   <- 
methods_cohort_sentences_df  |>
  dplyr::mutate(hard_negative = ifelse(sentence %in% methods_hard_negatives,
                                TRUE,
                                FALSE)
  )

print("Overlap between hard negatives and sentences with cohort names or cohort-related context in methods sections:")
methods_cohort_sentences_df |>
  filter(hard_negative == TRUE) |>
  group_by(has_cohort, has_context) |>
  summarise(n = n())


filtered_methods_cohort_sentences_df =
  methods_cohort_sentences_df |>
  filter(COHORT != "" | CONTEXT != "" | hard_negative == TRUE)

print("Number of sentences with cohort names or cohort-related context (population, ancestry, country terms):")
nrow(filtered_methods_cohort_sentences_df)

```

## Prepare data in Doccano JSON format

```{r prepare_doccano_json_methods}

# file path for intermediate json file output
json_file = here::here("output/doccano/methods_with_cohort_info.json")

# file path for final jsonl file output
jsonl_file = here::here("output/doccano/methods_with_cohort_info.jsonl")

filtered_methods_cohort_sentences_df  <- left_join(
  filtered_methods_cohort_sentences_df,
  meta_data_df,
  by = "pubmed_id"
)

# Create JSON as a list
json_list <- convert_to_doccano_json_sentence_level(
  cohort_sentences_df = filtered_methods_cohort_sentences_df 
)

# Write to JSON file
writeLines(toJSON(json_list,
                  auto_unbox = TRUE,
                  pretty = TRUE),
           json_file)


json_data <- fromJSON(json_file,
                      simplifyVector = FALSE)

# Open connection to JSONL file
con <- file(jsonl_file, "w")

# Loop over each element (object) and write as one line
for (i in seq_along(json_data)) {
  writeLines(toJSON(json_data[[i]], auto_unbox = TRUE), con)
}

# Close connection
close(con)

cat("JSONL saved to:", jsonl_file, "\n")

```

# Next steps to update: 

## Population descriptors extract / label

## Sample sizes extract / label

```{r extract_sample_sizes, eval = F}

participant_words <- c("adults", "adolescents", "children", "infants", 
                       "neonates",
                       "women", "men", "female", "male","trios", "families", "twins", "siblings", 
                        "cases", "probands", #"case subjects", #"([A-Za-z-]+)\\scases", ,
                        "controls", #"control subjects",   #"([A-Za-z-]+)\\scontrols"                            
                        "individuals", "patients",  "participants", 
                       "subjects", "samples", "donors")
                        #“([A-Za-z-]+)\\sparticipants”
                        #"\\b[0-9,]+\\s+([A-Za-z-]+)\\scases"

                        # “population controls”,
                        # "affected individuals"
                       # "study participants"
participant_phrases_regex <- paste0("\\b[0-9,]+\\s+([A-Za-z-]+)\\s", "(",
                              paste0(participant_words, collapse = "|"), 
                              ")")

participant_phrases_regex_v2 <- paste0("\\b[0-9,]+\\s+([A-Za-z-]+)\\s+([A-Za-z-]+)\\s", "(",
                                    paste0(participant_words, collapse = "|"), 
                                    ")")

participant_phrases_regex_v3 <- paste0("\\b[0-9,]+\\s+([A-Za-z-]+)\\s+([A-Za-z-]+)\\s([A-Za-z-]+)\\s", "(",
                                       paste0(participant_words, collapse = "|"), 
                                       ")")

participant_regex <- paste0("\\b[0-9,]+\\s+", "(",
                        paste0(participant_words, collapse = "|"), 
                        ")")


n_participants_regex <- "(n|N|Neffective|Neff|Ncontrol|Ncase|ncase|ncontrol|ntotal)\\s*=\\s*[0-9,]"

filtered_cohort_sentences_df |>
  filter(str_detect(sentence, participant_phrases_regex) | 
           str_detect(sentence, participant_regex) |
           str_detect(sentence, participant_phrases_regex_v2) |
           str_detect(sentence, participant_phrases_regex_v3) |
           str_detect(sentence, n_participants_regex)) |> 
  select(sentence) |> 
  distinct() |> 
  pull()

filtered_cohort_sentences_df |>
  mutate(
    participant_phrase = str_extract(sentence, participant_phrases_regex),
    participant_phrase_v2 = str_extract(sentence, participant_phrases_regex_v2),
    participant_phrase_v3 = str_extract(sentence, participant_regex),
    n_participants_phrase = str_extract(sentence, n_participants_regex)
  ) |>
  mutate(any_participant_phrase = if_else(!is.na(participant_phrase) | 
                                   !is.na(participant_phrase_v2) | 
                                   !is.na(participant_phrase_v3) | !is.na(n_participants_phrase), 
                                 TRUE, FALSE)) |>
  filter(grepl(pattern = "\\b[0-9]+", sentence)) |>
  filter(any_participant_phrase == FALSE) |>
  pull(sentence) 

filtered_methods_cohort_sentences_df |>
mutate(
  participant_phrase = str_extract(sentence, participant_phrases_regex),
  participant_phrase_v2 = str_extract(sentence, participant_phrases_regex_v2),
  participant_phrase_v3 = str_extract(sentence, participant_regex),
  n_participants_phrase = str_extract(sentence, n_participants_regex)
) |>
  mutate(any_participant_phrase = if_else(!is.na(participant_phrase) | 
                                            !is.na(participant_phrase_v2) | 
                                            !is.na(participant_phrase_v3) | !is.na(n_participants_phrase), 
                                          TRUE, FALSE)) |>
  filter(grepl(pattern = "\\b[0-9]+", sentence)) |>
  filter(any_participant_phrase == FALSE) |>
  pull(sentence) 

```


## Identify meta-analysis studies 

```{r extract_meta_analysis_studies, eval = F}
pattern <- paste(
  "meta-analy[sz]is",           # meta-analysis / meta-analyses / meta-analyze
  "meta-analy[sz]ed",
  "meta-analy[sz]ing",
  "metaanaly[sz]is",            # no hyphen
  "meta analy[sz]is",           # space
  # "pooled analysis",
  # "pooled analyses",
  # "combined analysis",
  # "combined analyses",
  # "joint analysis",
  # "mega-analysis",              # individual-participant variant
  "METAL",
  sep = "|"
)


possible_meta_abstracts <-
  purrr::keep(abstracts,
              ~ any(grepl(pattern, .x, ignore.case = TRUE))) |>
  names()

# purrr::map(abstracts,
#             ~grep(pattern, .x, ignore.case = TRUE, value = TRUE)) |>
#   purrr::list_c()

possible_meta_methods <-
  purrr::keep(methods_texts,
              ~ any(grepl(pattern, .x, ignore.case = TRUE))) |>
  names()

possible_meta <- union(possible_meta_abstracts, possible_meta_methods)

writeLines(possible_meta, here::here("output/meta_analysis_studies.txt"))

```
