---
title: "Get abstract text for GWAS Catalog studies"
author: "Isobel Beasley"
date: "2025-10-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Set up 

```{r, include = FALSE}
knitr::opts_chunk$set(engine.opts = list(bash = "-l"))
```

## Required packages

```{r, message=FALSE}

library(stringr)
library(readxl)
library(dplyr)
library(stringi)
library(xml2)
library(jsonlite)
library(tokenizers)

```

# Get PMIDs for relevant papeprs 

```{r gwas_cat_load_preprocess, message=FALSE}

# get only relevant disease studies
gwas_study_info <- data.table::fread(here::here("output/icd_map/gwas_study_gbd_causes.csv"))

gwas_study_info = gwas_study_info |>
  dplyr::rename_with(~ gsub(" ", "_", .x))

# filter out infectious diseases
gwas_study_info <- gwas_study_info |>
    dplyr::filter(!cause %in% c("HIV/AIDS",
                             "Tuberculosis",
                             "Malaria",
                             "Lower respiratory infections",
                             "Diarrhoeal diseases",
                             "Neonatal disorders",
                             "Tetanus",
                             "Diphtheria",
                             "Pertussis" ,
                             "Measles",
                             "Maternal disorders"))

gwas_study_info <-
  gwas_study_info |>
  filter(cause != "")

# all pmids
pmids <- gwas_study_info$PUBMED_ID |> unique()

# training sample pmids
training_sample <- readLines(here::here("output/gwas_cat/training_sample_pmids.txt"))

```

# Get PMCIDs

```{r load_converted_ids}

converted_ids <- data.table::fread(here::here("output/fulltexts/pmid_to_pmcid_mapping.csv"))

```

# Get grep terms 

## Cohort context terms 

```{r load_ancestry_population_terms}

cohort_context_terms <- readLines(here::here("output/gwas_cat/ancestry_population_terms.txt"))

print("Number of cohort context terms:")
length(cohort_context_terms)

```

## Cohort names 

```{r load_cohort_names}

cohort_abbr_names <- readLines(
           here::here("output/gwas_cat/cohort_abbrev_names.txt"))

cohort_full_names <- 
readLines(
           here::here("output/gwas_cat/cohort_full_names.txt"))

```

# Get cohort-relevant abstract sentences

## Convert abstract sections to sentences

```{bash, eval = F}

source spacyr_venv/bin/activate

python3 code/extract_text/spacy_obtain_sentences.py \
--input_dir output/abstracts \
--output_dir output/abstracts \
--hf_model pritamdeka/S-PubMedBert-MS-MARCO

```

## Load relevant abstracts

```{r read_abstracts_from_files}

abstract_files <- list.files(here::here("output/abstracts/"), 
                             pattern = "*.json", 
                             full.names = FALSE
                             ) |>
                   sort()

abstract_pmids = str_remove_all(abstract_files, "_sentences.json$")
abstract_pmids = abstract_pmids[abstract_pmids %in% pmids]


abstracts <- sapply(abstract_pmids,
                    function(file) {
                      json_data <- fromJSON(here::here(paste0("output/abstracts/", file, "_sentences.json")))
                      
                      abstract_lines <- json_data[json_data != ""]
                      
                      # readLines(here::here(paste0("output/abstracts/",file, ".txt")), 
                      #           warn = FALSE) |> 
                      #   paste(collapse = " ")
                    }
                    )


# check lengths of these vectors are the same
print("Check lengths of vectors are the same:")
print(paste("Length of pmids:", length(pmids)))
print(paste("Length of abstract pmids:", length(abstract_pmids)))
print(paste("Length of abstracts:", length(abstracts)))

```

###  Check: What abstracts are missing? 

```{r check_missing_abstracts, warning=FALSE, message=FALSE}

missing_abstracts = pmids[!(pmids %in% abstract_pmids)] 

print("Number of missing abstracts:")
length(missing_abstracts)

```

## Cohort name matching in abstract sentences

### Remove title sentences and small sentences that are likely to contain cohort names but are not sentences describing the cohort

```{r remove_title_and_small_sentences}

# remove title from sentences (as this is likely to contain cohort names but is not a sentence describing the cohort)

has_balanced_parens <- function(s) {
  stringr::str_count(s, fixed("(")) == stringr::str_count(s, fixed(")"))
}

get_title_sentences <- function(x, n_words_threshold = 5) {
                 
                 # get short sentences < 6 words
                 sentences <- x[stringr::str_count(x, " ") < n_words_threshold]
                 
                 # catch / inspect title sentences 
                 # upper_case <- "(?m)^\\s*[A-Z][A-Z\\s]+$"
                 
                 # get sentences that start with capital letter
                 sentences <- sentences[grepl("^(\\d+(\\.\\d+)*\\s+)?[A-Za-z]", sentences)]
                  #sentences[grepl("^[A-Z]", sentences)]
                 
                 # exclude sentences containing unmatched bracket as these are more likely to be complex sentences that are less likely to be titles or headings
                 #sentences <- sentences[!grepl("\\(|\\)", sentences)]
                 sentences <- sentences[has_balanced_parens(sentences)]
                 
                 # exclude sentences that end with a comma, as these are more likely to be incomplete sentences that are part of a larger sentence, and less likely to be titles or headings
                 sentences <- sentences[!grepl(",$|\\($", sentences)] 
                 
                 # exclude sentences that contain urls
                 sentences <- sentences[!grepl("www.|http", sentences)]
                 
                 return(sentences)
                 
}

get_non_title_sentences <- function(x) {
  
                title_sentences <- get_title_sentences(x)
                 
                keep_non_title_sentences <- x[!c(x %in% title_sentences)]
                 
                return(keep_non_title_sentences)
}

title_sentences <- purrr::map(abstracts,
                              get_title_sentences
                              ) |>
                              # ~get_title_sentences(.x, 
                              #                     n_words_threshold = 3)) |>
                   purrr::list_c() |>
                   sort() |>
                   unique()


print("Title sentences identified:")
print(title_sentences)


abstracts <-
purrr::map(abstracts,
           get_non_title_sentences        
                 )


```

### Function to extract sentences with cohort names

```{r fn_extract_cohort_sentences}

extract_cohort_sentences <- function(text_list, # list of texts to search through (e.g. abstracts, or full texts)
                                     terms, # terms to grep search for
                                     column_name = "COHORT",
                                     tokenize = FALSE,
                                     ignore_case) {
  
  #word boundary or preceded by ( [ ,
  terms_grep <- paste0("(?<=\\b|\\(|\\[)",
                            terms,
                            "(?=\\b|,|\\(|\\])")
  
  results <- lapply(seq_along(text_list), function(i) {
    
    # Split text into sentences, if not done already
    if(tokenize) {
      sentences <- tokenizers::tokenize_sentences(test_list[[i]])[[1]]
    } else {
      sentences <-  text_list[[i]]
    }
    
    # For each sentence, find all matching cohort names
    lapply(seq_along(sentences), function(s) {
      
      sentence <- sentences[s]
      
      # Identify cohort names present in this sentence
      matched_cohorts <- terms[str_detect(sentence, 
                                          regex(terms_grep, 
                                                ignore_case = ignore_case)
                                          )
                               ]
      term_df <-
      data.frame(
        article_id = i,
        sentence_id = s,
        sentence = sentence,
        stringsAsFactors = FALSE
      )
      
      # Add column with matched term names (if any)
      term_df[[column_name]] <- ifelse(length(matched_cohorts) > 0,
                                       str_flatten(unique(matched_cohorts), 
                                                   collapse = "|", 
                                                   na.rm = T),
                                       ""
      )
      
      # Add column with binary label for whether any terms were matched
      term_df[[paste0("has_", tolower(column_name))]] <- ifelse(length(matched_cohorts) > 0, 
                                                                TRUE, 
                                                                FALSE)
      
      return(term_df)
      
    }) |> bind_rows()
  })

 results <- bind_rows(results)
 
 results$pubmed_id <- names(text_list)[results$article_id]
 
 return(results) #returns a df of sentences, with labelled columns for whether they contain cohort names, and which cohort names they contain (if any)
 # id of sentence, and abstract / article
}

```

### Get cohort relevant sentences 

#### Get cohort name sentences

```{r extract_cohort_sentences_from_abstracts}

# extract sentences with cohort full names (case-insensitive matching, as full names are less likely to be ambiguous)
cohort_sentences_df_p1 <- extract_cohort_sentences(abstracts,
                                                cohort_full_names,
                                                ignore_case = TRUE
                                                )

# extract sentences with cohort abbreviation names (case-sensitive matching, as abbreviations are more likely to be ambiguous)
cohort_sentences_df_p2 = extract_cohort_sentences(abstracts,
                                                  cohort_abbr_names,
                                                  ignore_case = FALSE
                                                  )

cohort_sentences_df =
  bind_rows(cohort_sentences_df_p1,
            cohort_sentences_df_p2
            )


cohort_sentences_df = 
  cohort_sentences_df |>
  distinct()

separate_cohorts <- function(COHORT) {
  
  if (any(grepl("\\|", COHORT))) {
    return(unlist(strsplit(COHORT, "\\|")))
  } else {
    return(COHORT)
  }
  
}

cohort_sentences_df =
  cohort_sentences_df |>
  group_by(pubmed_id,
           article_id, 
           sentence_id, 
           sentence) |>
  summarise(
    COHORT = str_flatten(unique(separate_cohorts(COHORT)), 
                         collapse = "|", 
                         na.rm = T),
    has_cohort = ifelse(any(COHORT != ""), TRUE, FALSE)
  ) |>
  ungroup() 

cohort_sentences_df =
  cohort_sentences_df |>
  mutate(COHORT = str_remove_all(COHORT, 
                               pattern = "\\|$|^\\|"
                               )
         ) 

```

```{r cohort_sentence_summary}

print("Number of abstracts containing probable cohort reference")
abstracts_cohorts <- cohort_sentences_df |>
  filter(COHORT != "") |>
  pull(article_id) |>
  unique() 

n_abstracts_cohorts =  abstracts_cohorts |>
  length()

print(n_abstracts_cohorts)

print("Percentage of sampled abstracts containing probable cohort reference:")
100 * n_abstracts_cohorts / length(unique(pmids))

print("Number of sentences containing probable cohort reference")
  cohort_sentences_df |>
  filter(COHORT != "") |>
  select(article_id, sentence_id) |>
  distinct() |>
  nrow()
  
print("Number of entities")
cohort_sentences_df |>
  filter(COHORT != "") |>
  pull(COHORT) |>
  str_split("\\|") |>
  unlist() |>
  length()
  
print("Number of unique cohorts referenced (not unique cohort names, but unique strings in the COHORT column):")
detected_cohorts <- cohort_sentences_df |>
  filter(COHORT != "") |>
  pull(COHORT) |>
  str_split("\\|") |>
  unlist() 

n_cohorts <- detected_cohorts |>
  unique()  |>
  length()

print(n_cohorts)

print("Most common cohort names detected (by number of sentences they are mentioned in):")
data.frame(cohort = detected_cohorts) |>
  group_by(cohort) |>
  summarise(n = n()) |>
  arrange(desc(n)) |>
  head(10)

print("Most common cohort names detected (by number of abstracts they are mentioned in):")
cohort_sentences_df |>
  filter(COHORT != "") |>
  select(COHORT, article_id) |>
  tidyr::separate_rows(COHORT, sep = "\\|") |>
  distinct() |>
  group_by(COHORT) |>
  summarise(n_abstracts = n()) |>
  arrange(desc(n_abstracts)) |>
  head(5)
  
```

#### Get cohort-related context sentences

```{r extract_cohort_context_sentences_from_abstracts}

all_cohort_terms <- c(cohort_full_names, cohort_abbr_names)

# mask cohort names in methods text
# so that extract population descriptors etc. are not masked by cohort name matching
cohort_patterns <- 
  all_cohort_terms |>
  unique()

cohort_patterns <- cohort_patterns[order(-nchar(cohort_patterns))] # order by length, to avoid shorter patterns matching within longer patterns (e.g. "UK Biobank" and "Biobank")

cohort_patterns  <- cohort_patterns |> 
  stringr::str_escape()   # or: gsub("([][{}()+*^$|\\\\?.])", "\\\\\\1", x)

cohort_patterns <- paste0("(?<=\\b|\\(|\\[)",
                            cohort_patterns,
                            "(?=\\b|,|\\(|\\])")

mask_re <- paste0("\\b(?:",
                  paste(cohort_patterns,
                        collapse = "|"),
                  ")\\b")


masked_abstracts <- purrr::map(abstracts,
           function(x) {
             stringr::str_replace_all(x, 
                                      stringr::regex(mask_re, ignore_case = TRUE), 
                                      "\u241F"
                                      )
           }
           )

# extract sentences with cohort context terms (case-sensitive matching, as some context terms are common words that could be ambiguous)
cohort_context_df = extract_cohort_sentences(masked_abstracts,
                                            cohort_context_terms,
                                            column_name = "CONTEXT",
                                            ignore_case = FALSE
                                            )

cohort_sentences_df <-
left_join(cohort_sentences_df,
          cohort_context_df |> select(pubmed_id,
                                      article_id, 
                                      sentence_id, 
                                      CONTEXT,
                                      has_context),
          by = c("pubmed_id",
                 "article_id", 
                 "sentence_id")
          ) 

print("Number of sentences with cohort names and/or cohort-related context:")
cohort_sentences_df |> 
  group_by(has_cohort, has_context) |> 
  summarise(n = n())

filtered_cohort_sentences_df =
  cohort_sentences_df |>
  filter(COHORT != "" | CONTEXT != "")

print("Number of sentences with cohort names or cohort-related context (population, ancestry, country terms):")
nrow(filtered_cohort_sentences_df)

```


#### Get hard negatives 

These are 'semanitically similar' sentences to cohort names. 

```{r save_sentences}

json_list <- 
cohort_sentences_df |>
  filter(has_cohort) |>
  pull(sentence) |>
  as.list()

# Write to JSON file
writeLines(toJSON(json_list,
                  auto_unbox = TRUE,
                  pretty = TRUE),
           here::here("output/doccano/abstract_cohort_sentences.json"))


json_list <- 
cohort_sentences_df |>
  filter(!has_cohort) |>
  pull(sentence) |>
  as.list()

# Write to JSON file
writeLines(toJSON(json_list,
                  auto_unbox = TRUE,
                  pretty = TRUE),
           here::here("output/doccano/abstract_non_cohort_sentences.json"))

```

```{bash}

source  spacyr_venv/bin/activate

python3 code/extract_text/sentence_embeddings.py \
        --label "abstracts" \
        --cohort_path "output/doccano/abstract_cohort_sentences.json" \
        --non_cohort_path "output/doccano/abstract_non_cohort_sentences.json" \
        --figure_save_path "output/doccano/abstract_sentence_similarity.png" \
        --p 75
        
```

#### Make training set

```{r add_hard_negs_to_abstract_sentences}

abstract_hard_negatives <- jsonlite::read_json(here::here("output/doccano/hard_negatives_abstracts.json"),
                                               simplifyVector = TRUE)

cohort_sentences_df  <- 
cohort_sentences_df |>
  dplyr::mutate(hard_negative = ifelse(sentence %in% abstract_hard_negatives,
                                TRUE,
                                FALSE)
  )

print("Overlap between hard negatives and sentences with cohort names or cohort-related context:")
cohort_sentences_df |>
  filter(hard_negative == TRUE) |>
  group_by(has_cohort, has_context) |>
  summarise(n = n())

# most hard negatives overlap with context sentences! (As expected, as 
# as context sentences should also be semantically similar to cohort sentences

filtered_cohort_sentences_df <-
cohort_sentences_df |>
  filter(COHORT != "" | CONTEXT != "" | hard_negative == TRUE)

print("Size of training set:")
print("Number of sentences with cohort names or cohort-related context (population, ancestry, country terms) or hard negatives:")
nrow(filtered_cohort_sentences_df)

```

## Prepare data in Doccano JSON format

```{r fn_prepare_doccano}



convert_to_doccano_json_sentence_level <- function(cohort_sentences_df) {
  doccano_list <- vector("list", nrow(cohort_sentences_df))

  for (i in seq_len(nrow(cohort_sentences_df))) {
    current_sentence <- cohort_sentences_df$sentence[i]
    matched_cohort   <- cohort_sentences_df$COHORT[i]
    pubmed_id_val    <- cohort_sentences_df$pubmed_id[i]
    date_val         <- cohort_sentences_df$date[i]
    cohort_val       <- cohort_sentences_df$cohort[i]
    country_val      <- cohort_sentences_df$country[i]

    if (is.na(matched_cohort) || matched_cohort == "") {
      entities <- list()
    } else {
      cohorts <- if (grepl("\\|", matched_cohort)) {
        unlist(strsplit(matched_cohort, split = "\\|"))
      } else {
        matched_cohort
      }

      matches <- do.call(rbind, lapply(cohorts, function(c) {
        stringr::str_locate_all(current_sentence,
                                stringr::fixed(c, ignore_case = TRUE))[[1]]
      }))

      matches[, "start"] <- matches[, "start"] - 1  # 0-based for Doccano

      entities <- lapply(seq_len(nrow(matches)), function(k) {
        list(start_offset = matches[k, "start"],
             end_offset   = matches[k, "end"],
             label        = "COHORT")
      })
    }

    doccano_list[[i]] <- list(
      text                  = current_sentence,
      pubmed_id             = pubmed_id_val,
      date                  = date_val,
      country               = country_val,
      gwas_cat_cohort_label = cohort_val,
      label                 = entities
    )
  }

  # Convert label entries to [start, end, label] vector form
  doccano_list <- lapply(doccano_list, function(x) {
    x$label <- lapply(x$label, function(l) c(l$start_offset, l$end_offset, l$label))
    x
  })

  doccano_list
}

# old version
# convert_to_doccano_json_sentence_level <- function(cohort_sentences_df) {
#   
#   # set up json list
#   doccano_list <- list()
#   example_id <- 1
# 
#   for(current_sentence in cohort_sentences_df$sentence) {
# 
#       # Filter cohort sentences that match this sentence (safe matching)
#       df <- cohort_sentences_df |>
#              dplyr::filter(sentence == current_sentence) 
#       
#       for (i in seq_len(nrow(df))) {
#         
#         matched_cohort <- df$COHORT[i]
#         pubmed_id_val  <- df$pubmed_id[i]
#         date_val       <- df$date[i]
#         cohort_val     <- df$cohort[i]
#         country_val    <- df$country[i]
#       
#         if(matched_cohort == ""){
#         
#           doccano_list[[example_id]] <- list(
#             text = current_sentence,
#             pubmed_id = pubmed_id_val,
#             date = date_val,
#             country = country_val,
#             gwas_cat_cohort_label = cohort_val,
#             label = list()
#           )
#         
#         } else {
#         
#           # Find the location of all matches of the cohort name in the sentence
#           if(grepl("\\|", matched_cohort)) {
#              
#              # If multiple cohort names, separate
#              matched_cohort <- unlist(strsplit(matched_cohort, split = "\\|"))
#           
#              match_locations <- list()
#           
#              for(current_matched_cohort in matched_cohort){
#             
#                  matches <- str_locate_all(current_sentence,
#                                            fixed(current_matched_cohort, 
#                                                  ignore_case = TRUE)
#                                            )[[1]]
#             
#                  match_locations <- append(match_locations, list(matches))
#              }
#           
#              # Combine all match locations into a single matrix
#              matches <- do.call(rbind, match_locations)
#           
#           } else {
#           
#              matches <- str_locate_all(current_sentence,
#                                        fixed(matched_cohort, 
#                                              ignore_case = TRUE)
#                                        )[[1]]
#           }
#           
#           # Convert matches to 0-based indexing (for doccano)
#           matches[, "start"] <- matches[, "start"] - 1  
#           
#           # Turn match locations into entity list
#           entities <- list()
#           
#           for(k in seq_len(nrow(matches))) {
#             entities <- append(entities, list(list(
#               start_offset = matches[k, "start"],
#               end_offset = matches[k, "end"],
#               label = "COHORT"
#             )))
#           }
#           
#           # Create Doccano JSON entry
#           doccano_list[[example_id]] <- list(
#             text = current_sentence,
#             pubmed_id = pubmed_id_val,
#             date = date_val,
#             country = country_val,
#             gwas_cat_cohort_label = cohort_val,
#             label = entities
#           )
#         }
#       }
# 
#       example_id <- example_id + 1
#   }
#   
#   # Convert named list label format to vector format [start, end, label]
#   doccano_list <- lapply(doccano_list, function(x) {
#     x$label <- lapply(x$label, function(l) {
#       c(l$start_offset, l$end_offset, l$label)
#     })
#     x
#   })
# 
#   return(doccano_list)
# }


```

### Function to do this preparation 

```{r prepare_doccano_json_abstracts}

{
# file path for intermediate json file output
json_file = here::here("output/doccano/abstracts_with_cohort_info.json")

# file path for final jsonl file output
jsonl_file = here::here("output/doccano/abstracts_with_cohort_info.jsonl")

# Join metadata onto sentences before passing to function
meta_data_df <- data.frame(
  pubmed_id = as.character(pmids)
)

filtered_cohort_sentences_df <- left_join(
  filtered_cohort_sentences_df,
  meta_data_df,
  by = "pubmed_id"
)

# Create JSON as a list
json_list <- convert_to_doccano_json_sentence_level(
  cohort_sentences_df = filtered_cohort_sentences_df
)

# Write to JSON file
writeLines(toJSON(json_list,
                  auto_unbox = TRUE,
                  pretty = TRUE),
           json_file)


json_data <- fromJSON(json_file,
                      simplifyVector = FALSE)

# Open connection to JSONL file
con <- file(jsonl_file, "w")

# Loop over each element (object) and write as one line
for (i in seq_along(json_data)) {
  writeLines(toJSON(json_data[[i]], auto_unbox = TRUE), con)
}
  
# Close connection
close(con)

cat("JSONL saved to:", jsonl_file, "\n")
}


# training dataset saving 
{
# file path for intermediate json file output
json_file = here::here("output/doccano/training_abstracts_with_cohort_info.json")

# file path for final jsonl file output
jsonl_file = here::here("output/doccano/training_abstracts_with_cohort_info.jsonl")


filtered_training_cohort_sentences_df <-
  filtered_cohort_sentences_df |>
  filter(pubmed_id %in% training_sample)

json_list <- convert_to_doccano_json_sentence_level(
  cohort_sentences_df = filtered_training_cohort_sentences_df
)

json_data <- fromJSON(json_file,
                      simplifyVector = FALSE)
# Write to JSON file
writeLines(toJSON(json_list,
                  auto_unbox = TRUE,
                  pretty = TRUE),
           json_file)

# Open connection to JSONL file
con <- file(jsonl_file, "w")

# Loop over each element (object) and write as one line
for (i in seq_along(json_data)) {
  writeLines(toJSON(json_data[[i]], auto_unbox = TRUE), con)
}

# Close connection
close(con)

cat("JSONL saved to:", jsonl_file, "\n")
}

```

# Get cohort-revelevant methods section sentence extraction

## Convert methods sections to sentences

```{bash, eval = F}

source spacyr_venv/bin/activate

python3 code/extract_text/spacy_obtain_sentences.py \
--input_dir output/methods \
--output_dir  output/methods_sentences 

```

## Load relevant methods text

```{r load_methods_text}

methods_sections <- list.files(here::here("output/methods_sentences/"), 
                             pattern = "*.json", 
                             full.names = TRUE
                             )

methods_sections <- methods_sections[!c(methods_sections %in% 
                                          grep("failed_files|no_methods", 
                                                methods_sections, 
                                                value = TRUE)
                                      )
                                     ]

pmcids_with_methods <- methods_sections |>
  gsub(pattern = ".*/", replacement = "") |>
  gsub(pattern = "_methods_sentences\\.json$", 
       replacement = "") |>
  gsub(pattern = "_methods_main_sentences\\.json$", 
       replacement = "") |>
  gsub(pattern = "_bioc|_pdf_tei", replacement = "") 

converted_pmcids <-  converted_ids |>
  filter(pmcids %in% pmcids_with_methods) |>
  pull(PMID)

pubmeds_with_methods <- c(converted_pmcids, 
                          grep("PMC", 
                               pmcids_with_methods, 
                               value = TRUE, 
                               invert = TRUE)
                          )

print("Number of papers with methods sections extracted out of sample")
sum(pubmeds_with_methods %in% pmids) 

100 * sum(pubmeds_with_methods %in% pmids) / length(pmids)

# get corresponding files to pubmed ids 
pubmeds_to_get <- pubmeds_with_methods[pubmeds_with_methods %in% pmids]

get_file_names_df <- 
   converted_ids |>
   filter(PMID %in% pubmeds_to_get) |>
   select(-DOI) |>
   rename(pmid = PMID,
          pmcid = pmcids
          ) |>
  rowwise() |>
  
  # data.frame(pmid = pubmeds_to_get) |>
  #                    #rowwise() |>
  #                    mutate(pmcid = converted_ids |> 
  #                                    filter(PMID %in% pubmeds_to_get) |> 
  #                                       pull(pmcids)
  #                               ) |>
  mutate(file_name = ifelse(!is.na(pmcid) | pmcid != "",
                            grep(methods_sections, 
                                 pattern = pmcid, 
                                 value = TRUE),
                            grep(methods_sections,
                                 pattern = pmid, 
                                 value = TRUE)
                            )
  ) |>
  mutate(file_name = ifelse(is.na(file_name),
                            grep(methods_sections,
                                 pattern = pmid, 
                                 value = TRUE),
                            file_name
                            )) |>
  filter(!is.na(file_name)) |> 
  ungroup()

methods_texts <- sapply(get_file_names_df$file_name,
                    function(file) {
                      
                      json_data <- fromJSON(file)
                      
                      text_lines <- json_data[json_data != ""]

                                          }
                    )


names(methods_texts) <- get_file_names_df$pmid

# order methods texts by pmids
methods_texts <- methods_texts[order(names(methods_texts))]

```

### Remove title sentences

```{r remove_title_sentences}


title_sentences <- purrr::map(methods_texts,
                              get_title_sentences) |>
                   purrr::list_c() |>
                   sort() |>
                   unique()


print("Number of title sentences extracted from methods sections:")
length(title_sentences)
           

methods_texts <-
purrr::map(methods_texts,
           get_non_title_sentences        
                 )

# what small sentences are left
purrr::map(
methods_texts,
~.x[stringr::str_count(.x, " ") < 5]
) |>
  purrr::list_c()

```

### Manual checks of methods text sentences 

```{r check_methods_texts, eval = F}

# to do when more settled with methods text extraction, to check the quality of the sentences extracted from methods sections, and whether they are likely to be useful for cohort name extraction

# Check 1. 
# Do all sentences start:
# with a capital,
# Or a number 
# Or a word that is lower case, and then upper case (e.g. eQTl)
check_1_regex <- "^([A-Z]|[0-9]+|[a-z]+[A-Z])"

check_1 <- purrr::keep(methods_texts,
            ~any(!grepl(check_1_regex,
                       .x)
                 )
                 ) 

check_1 <- purrr::imap(check_1, 
           function(x, idx) {
             
             sentences_idx = grep(check_1_regex,
                                  x,
                                  invert = TRUE,
                                  value = F
                                  )
             
             just_broken_sentences_idx <- sentences_idx
             
             sentences_idx <- sort(unique(c(sentences_idx - 1,
                               sentences_idx, 
                               sentences_idx + 1)))
             
             sentences_idx <- sentences_idx[sentences_idx > 0 & sentences_idx <= length(x)]
             
             
             # idx of nearby sentences
             # sentences_idx <- sort(unique(c(sentences_idx - 2,
             #                                sentences_idx, 
             #                                sentences_idx + 2))
             #                       )
             # 
             sentences = x[sentences_idx]
             
             df = data.frame(sentence = sentences,
                        pubmed_id = idx,
                        sentence_id = sentences_idx
                        ) |>
               dplyr::mutate(broken_sentence = ifelse(sentence_id %in% just_broken_sentences_idx, TRUE, FALSE))
             
             return(df)
           }
           ) |>
  bind_rows()

print("\n \n Check the starts of sentences seem reasonable")
print("Number of papers with sentences that don't start with capital, number, or lower-upper case word:")
check_1  |> pull(pubmed_id) |> unique() |> length()

print("Number of sentences that don't start with capital, number, or lower-upper case word:")
check_1 |> filter(broken_sentence) |> nrow()

# Check 2: 
# Sentence ends with a full stop: 
# "\\.$"
check_2_regex <- "\\.$|\\?$"

check_2 <- purrr::keep(methods_texts,
            ~any(!grepl(check_2_regex,
                       .x)
                 )
                 )

check_2 <- purrr::imap(check_2,
           function(x, idx) {
             
             sentences_idx = grep(check_2_regex,
                                  x,
                                  invert = TRUE,
                                  value = F
                                  )
             
             just_broken_sentences_idx <- sentences_idx
             
             sentences_idx <- sort(unique(c(sentences_idx - 1,
                               sentences_idx, 
                               sentences_idx + 1)))
             
             sentences_idx <- sentences_idx[sentences_idx > 0 & sentences_idx <= length(x)]
             
             
             # idx of nearby sentences
             # sentences_idx <- sort(unique(c(sentences_idx - 2,
             #                                sentences_idx, 
             #                                sentences_idx + 2))
             #                       )
             # 
             sentences = x[sentences_idx]
             
             df = data.frame(sentence = sentences,
                        pubmed_id = idx,
                        sentence_id = sentences_idx
                        ) |>
               dplyr::mutate(broken_sentence = ifelse(sentence_id %in% just_broken_sentences_idx, TRUE, FALSE))
             
             return(df)
           }
           ) |>
  bind_rows()


print("\n \n Check the ends of sentences seem reasonable")
print("Number of papers with sentences that don't end with a full stop or question mark:")
check_2  |> pull(pubmed_id) |> unique() |> length()

print("Number of sentences that don't end with a full stop or question mark:")
check_2 |> filter(broken_sentence) |> nrow()

# Check 3: 
# Sentences don't have un-matched brackets
# "\\((?![^()]*\\))|(?<!\\([^()]*\\))\\)"
has_unmatched_parens <- function(x) {
  open  <- stringr::str_count(x, "\\(")
  close <- stringr::str_count(x, "\\)")
  open != close
}

check_3 <- purrr::keep(methods_texts, 
                       ~ any(has_unmatched_parens(.x)))

check_3 <- purrr::imap(check_3,
           function(x, idx) {
             
             sentences_idx = which(sapply(x, has_unmatched_parens))
             
             sentences = x[sentences_idx]
             
             df = data.frame(sentence = sentences,
                        pubmed_id = idx,
                        sentence_id = sentences_idx
                        ) |>
               dplyr::mutate(broken_sentence = TRUE) |>
               tibble::remove_rownames()
             
             return(df)
           }
           ) |>
  bind_rows()

print("\n \n Check that brackets seem reasonable")
print("Number of papers with sentences that have un-matched brackets:")
check_3  |> pull(pubmed_id) |> unique() |> length()

print("Number of sentences that have un-matched brackets:")
check_3 |> filter(broken_sentence) |> nrow()

# Check 4: 
# Sentences don't have un-matched quotation marks
# "\"([^\"]*\"[^\"]*)*\""
check_4_regex <- "\"([^\"]*\"[^\"]*)*\""

check_4 <- purrr::keep(methods_texts,
            ~any(grepl(check_4_regex,
                       .x)
                 )
                 )

check_4 <- purrr::imap(check_4,
           function(x, idx) {
             
             sentences_idx = grep(check_4_regex,
                                  x,
                                  value = F
                                  )
             
             sentences = x[sentences_idx]
             
             df = data.frame(sentence = sentences,
                        pubmed_id = idx,
                        sentence_id = sentences_idx
                        ) |>
               dplyr::mutate(broken_sentence = TRUE)
             
             return(df)
           }
           ) |>
  bind_rows()


print("\n \n Check that quotation marks seem reasonable")
print("Number of papers with sentences that have un-matched quotation marks:")
print(head(check_4))

  
# Check sentences that end with [a-z][0-9].
check_5_regex <- "[a-z][0-9].$|[A-Z][0-9].$"

check_5 <- purrr::keep(methods_texts,
            ~any(grepl(check_5_regex,
                       .x)
                 )
                 )


check_5 <- purrr::imap(check_5,
           function(x, idx) {
             
             sentences_idx = grep(check_5_regex,
                                  x,
                                  value = F
                                  )
             
             sentences = x[sentences_idx]
             
             df = data.frame(sentence = sentences,
                        pubmed_id = idx,
                        sentence_id = sentences_idx
                        ) |>
               dplyr::mutate(broken_sentence = TRUE)
             
             return(df)
           }
           ) |>
  bind_rows()


n_character_limit <- 100

check_5 <- purrr::keep(methods_texts,
            ~any((nchar(.x) > n_character_limit) & 
                   length(grep(.x, pattern = ".") > 1)
            )
                 )

check_5 <- purrr::imap(check_5,
           function(x, idx) {
             
             sentences_idx = which(nchar(x) > n_character_limit)
             
             sentences = x[sentences_idx]
             
             df = data.frame(sentence = sentences,
                        pubmed_id = idx,
                        sentence_id = sentences_idx
                        ) |>
               dplyr::mutate(broken_sentence = TRUE)
             
             return(df)
           }
           ) |>
  bind_rows()

split_regex <- "(?<!St\\.)(?<!Fig\\.)(?<!no\\.)(?<!vs\\.)(?<!i\\.e\\.)(?<!e\\.g\\.)(?<!Inc\\.)(?<!Co\\.)(?<=[a-z]\\.)\\s+(?=[A-Z])"


check_5 |>
  mutate(match_split = str_detect(sentence, 
                                  pattern = split_regex
                                  )
  ) |>
  filter(match_split) |>
  pull(sentence) |>
  stringr::str_split(pattern = split_regex)  

  

```

## Get cohort relevant sentences 

### Get cohort names sentences

```{r extract_cohort_sentences_methods}

# extract sentences with cohort full names (case-insensitive matching, as full names are less likely to be ambiguous)
cohort_sentences_df_p1 <- extract_cohort_sentences(methods_texts,
                                                cohort_full_names,
                                                ignore_case = TRUE
                                                )

# extract sentences with cohort abbreviation names (case-sensitive matching, as abbreviations are more likely to be ambiguous)
cohort_sentences_df_p2 = extract_cohort_sentences(methods_texts,
                                                  cohort_abbr_names,
                                                  ignore_case = FALSE
                                                  )

methods_cohort_sentences_df = 
  bind_rows(cohort_sentences_df_p1,
            cohort_sentences_df_p2
            ) |>
  distinct()


methods_cohort_sentences_df =
  methods_cohort_sentences_df |>
  group_by(pubmed_id, 
           article_id, 
           sentence_id, 
           sentence) |>
  summarise(
    COHORT = str_flatten(unique(separate_cohorts(COHORT)), 
                         collapse = "|", 
                         na.rm = T),
    has_cohort = ifelse(any(COHORT != ""), TRUE, FALSE)
  ) |>
  ungroup() 

methods_cohort_sentences_df =
  methods_cohort_sentences_df |>
  mutate(COHORT = str_remove_all(COHORT, 
                               pattern = "\\|$|^\\|"
                               )
         ) 

print("Number of methods sections containing probable cohort reference:")
methods_cohorts <- methods_cohort_sentences_df |>
  filter(COHORT != "") |>
  pull(article_id) |>
  unique()

methods_cohorts  |> length()

print("Percentage of methods sections containing probable cohort reference:")
100 * length(methods_cohorts) / length(methods_texts)


print("Number of sentences containing probable cohort reference in methods sections")
  methods_cohort_sentences_df |>
  filter(COHORT != "") |>
  select(article_id, sentence_id) |>
  distinct() |>
  nrow()

print("Number of entities in methods sections")
methods_cohort_sentences_df |>
  filter(COHORT != "") |>
  pull(COHORT) |>
  str_split("\\|") |>
  unlist() |>
  length()

print("Most common cohort names detected in methods sections:")
methods_cohort_sentences_df |>
  filter(COHORT != "") |>
  select(COHORT, article_id) |>
  tidyr::separate_rows(COHORT, sep = "\\|") |>
  distinct() |>
  group_by(COHORT) |>
  summarise(n_abstracts = n()) |>
  arrange(desc(n_abstracts)) |>
  head(5)

print("Number of cohort names only mentioned in one paper")
methods_cohort_sentences_df |>
  filter(COHORT != "") |>
  select(COHORT, article_id) |>
  tidyr::separate_rows(COHORT, sep = "\\|") |>
  distinct() |>
  group_by(COHORT) |>
  summarise(n_abstracts = n()) |>
  filter(n_abstracts == 1) |>
  nrow()

print("Number of unique cohorts referenced in methods sections (not unique cohort names, but unique strings in the COHORT column):")
detected_cohorts_methods <- methods_cohort_sentences_df |>
  filter(COHORT != "") |>
  pull(COHORT) |>
  str_split("\\|") |>
  unlist()

n_cohorts_methods <- detected_cohorts_methods |>
  unique()  |>
  length()

print(n_cohorts_methods)

print("Most common cohort names detected in methods sections by number of sentences they are mentioned in:")
data.frame(cohort = detected_cohorts_methods) |>
  group_by(cohort) |>
  summarise(n = n()) |>
  arrange(desc(n)) |>
  head(10)

```

### Get cohort-related context sentences 

```{r extract_cohort_context_sentences_methods}

masked_methods_text <- purrr::map(methods_texts,
           function(x) {
             stringr::str_replace_all(x, 
                                      stringr::regex(mask_re, ignore_case = TRUE), 
                                      "\u241F"
                                      )
           }
           )

# extract sentences with cohort context terms (case-sensitive matching, as some context terms are common words that could be ambiguous)
methods_cohort_context_df = extract_cohort_sentences(
                                            masked_methods_text,
                                            cohort_context_terms,
                                            column_name = "CONTEXT",
                                            ignore_case = FALSE
                                            )

methods_cohort_sentences_df <-
left_join(methods_cohort_sentences_df,
          methods_cohort_context_df |> select(pubmed_id, 
                                      article_id,
                                      sentence_id,
                                      CONTEXT,
                                      has_context), 
          by = c("pubmed_id",
                 "article_id", 
                 "sentence_id")
          ) 

print("Number of sentences with cohort names and/or cohort-related context:")
methods_cohort_sentences_df |> 
  mutate(has_context = ifelse(CONTEXT == "", F, T)) |>
  group_by(has_cohort, has_context) |> 
  summarise(n = n())

print("Most common cohort-related context terms detected in methods sections:")
methods_cohort_sentences_df |>
  filter(CONTEXT != "") |>
  select(CONTEXT, article_id) |>
  tidyr::separate_rows(CONTEXT, sep = "\\|") |>
  distinct() |>
  group_by(CONTEXT) |>
  summarise(n_abstracts = n()) |>
  arrange(desc(n_abstracts)) |>
  head(10)

print("Most common cohort-related context terms detected in methods sections by number of sentences they are mentioned in:")
methods_cohort_sentences_df |>
  filter(CONTEXT != "") |>
  select(CONTEXT, article_id) |>
  tidyr::separate_rows(CONTEXT, sep = "\\|") |>
  distinct() |>
  group_by(CONTEXT) |>
  summarise(n = n()) |>
  arrange(desc(n)) |>
  head(10)



```

### Get hard negatives

```{r}

json_list <- 
methods_cohort_sentences_df  |>
  filter(has_cohort) |>
  pull(sentence) |>
  as.list()

# Write to JSON file
writeLines(toJSON(json_list,
                  auto_unbox = TRUE,
                  pretty = TRUE),
           here::here("output/doccano/methods_cohort_sentences.json"))


json_list <- 
methods_cohort_sentences_df  |>
  filter(!has_cohort) |>
  pull(sentence) |>
  as.list()

# Write to JSON file
writeLines(toJSON(json_list,
                  auto_unbox = TRUE,
                  pretty = TRUE),
           here::here("output/doccano/methods_non_cohort_sentences.json"))

```

```{bash}

source  spacyr_venv/bin/activate

python3 code/extract_text/sentence_embeddings.py \
        --label "methods" \
        --cohort_path "output/doccano/methods_cohort_sentences.json" \
        --non_cohort_path "output/doccano/methods_non_cohort_sentences.json" \
        --figure_save_path "output/doccano/methods_sentence_similarity.png"
        
```

### Make final traing set

sentences with words that indicate sample or cohort is being discussed

```{r get_sample_context_sentences_methods}

methods_hard_negatives <- jsonlite::read_json(here::here("output/doccano/hard_negatives_methods.json"),
                                               simplifyVector = TRUE)

methods_cohort_sentences_df   <- 
methods_cohort_sentences_df  |>
  dplyr::mutate(hard_negative = ifelse(sentence %in% methods_hard_negatives,
                                TRUE,
                                FALSE)
  )

print("Overlap between hard negatives and sentences with cohort names or cohort-related context in methods sections:")
methods_cohort_sentences_df |>
  filter(hard_negative == TRUE) |>
  group_by(has_cohort, has_context) |>
  summarise(n = n())


filtered_methods_cohort_sentences_df =
  methods_cohort_sentences_df |>
  filter(COHORT != "" | CONTEXT != "" | hard_negative == TRUE)

print("Number of sentences with cohort names or cohort-related context (population, ancestry, country terms):")
nrow(filtered_methods_cohort_sentences_df)

```

## Prepare data in Doccano JSON format

```{r prepare_doccano_json_methods}

# file path for intermediate json file output
{
json_file = here::here("output/doccano/methods_with_cohort_info.json")

# file path for final jsonl file output
jsonl_file = here::here("output/doccano/methods_with_cohort_info.jsonl")

filtered_methods_cohort_sentences_df  <- left_join(
  filtered_methods_cohort_sentences_df,
  meta_data_df,
  by = "pubmed_id"
)

# Create JSON as a list
json_list <- convert_to_doccano_json_sentence_level(
  cohort_sentences_df = filtered_methods_cohort_sentences_df 
)

# Write to JSON file
writeLines(toJSON(json_list,
                  auto_unbox = TRUE,
                  pretty = TRUE),
           json_file)


json_data <- fromJSON(json_file,
                      simplifyVector = FALSE)

# Open connection to JSONL file
con <- file(jsonl_file, "w")

# Loop over each element (object) and write as one line
for (i in seq_along(json_data)) {
  writeLines(toJSON(json_data[[i]], auto_unbox = TRUE), con)
}

# Close connection
close(con)

cat("JSONL saved to:", jsonl_file, "\n")
}

{
# file path for intermediate json file output
json_file = here::here("output/doccano/training_methods_with_cohort_info.json")

# file path for final jsonl file output
jsonl_file = here::here("output/doccano/training_methods_with_cohort_info.jsonl")

filtered_training_methods_cohort_sentences_df <-
  filtered_methods_cohort_sentences_df |>
  filter(pubmed_id %in% training_sample)

json_list <- convert_to_doccano_json_sentence_level(
  cohort_sentences_df = filtered_training_methods_cohort_sentences_df
)

# Write to JSON file
writeLines(toJSON(json_list,
                  auto_unbox = TRUE,
                  pretty = TRUE),
           json_file)


json_data <- fromJSON(json_file,
                      simplifyVector = FALSE)

# Open connection to JSONL file
con <- file(jsonl_file, "w")

# Loop over each element (object) and write as one line
for (i in seq_along(json_data)) {
  writeLines(toJSON(json_data[[i]], auto_unbox = TRUE), con)
}

# Close connection
close(con)

cat("JSONL saved to:", jsonl_file, "\n")
}

```

# Next steps to update: 

## Population descriptors extract / label

## Sample sizes extract / label

```{r extract_sample_sizes, eval = F}

participant_words <- c("adults", "adolescents", "children", "infants", 
                       "neonates",
                       "women", "men", "female", "male","trios", "families", "twins", "siblings", 
                        "cases", "probands", #"case subjects", #"([A-Za-z-]+)\\scases", ,
                        "controls", #"control subjects",   #"([A-Za-z-]+)\\scontrols"                            
                        "individuals", "patients",  "participants", 
                       "subjects", "samples", "donors")
                        #“([A-Za-z-]+)\\sparticipants”
                        #"\\b[0-9,]+\\s+([A-Za-z-]+)\\scases"

                        # “population controls”,
                        # "affected individuals"
                       # "study participants"
participant_phrases_regex <- paste0("\\b[0-9,]+\\s+([A-Za-z-]+)\\s", "(",
                              paste0(participant_words, collapse = "|"), 
                              ")")

participant_phrases_regex_v2 <- paste0("\\b[0-9,]+\\s+([A-Za-z-]+)\\s+([A-Za-z-]+)\\s", "(",
                                    paste0(participant_words, collapse = "|"), 
                                    ")")

participant_phrases_regex_v3 <- paste0("\\b[0-9,]+\\s+([A-Za-z-]+)\\s+([A-Za-z-]+)\\s([A-Za-z-]+)\\s", "(",
                                       paste0(participant_words, collapse = "|"), 
                                       ")")

participant_regex <- paste0("\\b[0-9,]+\\s+", "(",
                        paste0(participant_words, collapse = "|"), 
                        ")")


n_participants_regex <- "(n|N|Neffective|Neff|Ncontrol|Ncase|ncase|ncontrol|ntotal)\\s*=\\s*[0-9,]"

filtered_cohort_sentences_df |>
  filter(str_detect(sentence, participant_phrases_regex) | 
           str_detect(sentence, participant_regex) |
           str_detect(sentence, participant_phrases_regex_v2) |
           str_detect(sentence, participant_phrases_regex_v3) |
           str_detect(sentence, n_participants_regex)) |> 
  select(sentence) |> 
  distinct() |> 
  pull()

filtered_cohort_sentences_df |>
  mutate(
    participant_phrase = str_extract(sentence, participant_phrases_regex),
    participant_phrase_v2 = str_extract(sentence, participant_phrases_regex_v2),
    participant_phrase_v3 = str_extract(sentence, participant_regex),
    n_participants_phrase = str_extract(sentence, n_participants_regex)
  ) |>
  mutate(any_participant_phrase = if_else(!is.na(participant_phrase) | 
                                   !is.na(participant_phrase_v2) | 
                                   !is.na(participant_phrase_v3) | !is.na(n_participants_phrase), 
                                 TRUE, FALSE)) |>
  filter(grepl(pattern = "\\b[0-9]+", sentence)) |>
  filter(any_participant_phrase == FALSE) |>
  pull(sentence) 

filtered_methods_cohort_sentences_df |>
mutate(
  participant_phrase = str_extract(sentence, participant_phrases_regex),
  participant_phrase_v2 = str_extract(sentence, participant_phrases_regex_v2),
  participant_phrase_v3 = str_extract(sentence, participant_regex),
  n_participants_phrase = str_extract(sentence, n_participants_regex)
) |>
  mutate(any_participant_phrase = if_else(!is.na(participant_phrase) | 
                                            !is.na(participant_phrase_v2) | 
                                            !is.na(participant_phrase_v3) | !is.na(n_participants_phrase), 
                                          TRUE, FALSE)) |>
  filter(grepl(pattern = "\\b[0-9]+", sentence)) |>
  filter(any_participant_phrase == FALSE) |>
  pull(sentence) 

```

## Identify meta-analysis studies 

```{r extract_meta_analysis_studies, eval = F}
pattern <- paste(
  "meta-analy[sz]is",           # meta-analysis / meta-analyses / meta-analyze
  "meta-analy[sz]ed",
  "meta-analy[sz]ing",
  "metaanaly[sz]is",            # no hyphen
  "meta analy[sz]is",           # space
  # "pooled analysis",
  # "pooled analyses",
  # "combined analysis",
  # "combined analyses",
  # "joint analysis",
  # "mega-analysis",              # individual-participant variant
  "METAL",
  sep = "|"
)


possible_meta_abstracts <-
  purrr::keep(abstracts,
              ~ any(grepl(pattern, .x, ignore.case = TRUE))) |>
  names()

# purrr::map(abstracts,
#             ~grep(pattern, .x, ignore.case = TRUE, value = TRUE)) |>
#   purrr::list_c()

possible_meta_methods <-
  purrr::keep(methods_texts,
              ~ any(grepl(pattern, .x, ignore.case = TRUE))) |>
  names()

possible_meta <- union(possible_meta_abstracts, possible_meta_methods)

writeLines(possible_meta, here::here("output/meta_analysis_studies.txt"))

```
