---
title: "Get abstract text for GWAS Catalog studies"
author: "Isobel Beasley"
date: "2025-10-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Set up 

## Required packages

```{r, message=FALSE}

library(stringr)
library(readxl)
library(dplyr)
library(stringi)
library(httr)
library(rentrez)
library(xml2)
library(jsonlite)
library(tokenizers)

```

# Getting required information from GWAS catalog

### Step 1: Get only disease studies, and select relevant columns

```{r gwas_cat_load_preprocess}

## Step 1: 
# get only relevant disease studies
gwas_study_info <- data.table::fread(here::here("output/icd_map/gwas_study_gbd_causes.csv"))

gwas_study_info = gwas_study_info |>
  dplyr::rename_with(~ gsub(" ", "_", .x))

# filter out infectious diseases
gwas_study_info <- gwas_study_info |>
    dplyr::filter(!cause %in% c("HIV/AIDS",
                             "Tuberculosis",
                             "Malaria",
                             "Lower respiratory infections",
                             "Diarrhoeal diseases",
                             "Neonatal disorders",
                             "Tetanus",
                             "Diphtheria",
                             "Pertussis" ,
                             "Measles",
                             "Maternal disorders"))

# group conditions into broader categories to ensure we have enough papers in each category for sampling
lancet_cause_mapping <- readxl::read_xlsx(here::here("data/icd/lancet_conditions_icd10.xlsx"),
                                     sheet = 1) |>
  dplyr::rename_with(~ gsub(" ", "_", .x))

set.seed(500)

training_sample = gwas_study_info |>
  left_join(lancet_cause_mapping |> select(cause = mapped_gbd_term,
                                           lancet_condition), 
            by = "cause",
            relationship = "many-to-many") |>
  select(PUBMED_ID,
         lancet_condition) |>
  distinct() |>
  dplyr::group_by(lancet_condition) |>
  dplyr::slice_sample(prop = 0.25) |>
  pull(PUBMED_ID) |>
  unique()

pmids <- training_sample

gwas_study_info =
  gwas_study_info |>
  filter(PUBMED_ID %in% training_sample) 

print("Number of unique pubmed ids for disease studies:")
gwas_study_info$PUBMED_ID |> unique() |> length()

```

#### How much overlap with abstracts and full-texts retrieved? 

```{r check_abstract_overlap}

converted_ids <- data.table::fread(here::here("output/fulltexts/pmid_to_pmcid_mapping.csv"))

full_text_files <-
  list.files(here::here("output/fulltexts"), 
             recursive = T, 
             pattern = "\\.html$|\\.xml$") 

full_text_files <- basename(full_text_files) |> 
  stringr::str_remove_all("\\.html$|\\.xml$") |>
  unique()

# convert pmcids to pmids 
converted_fulltext_pmcids <-
  converted_ids |>
  filter(pmcids %in% full_text_files) |>
  pull(PMID) |>
  unique() 

full_text_files  <- c(full_text_files, 
                      converted_fulltext_pmcids)

full_text_pmids <- grep("PMC", 
                        full_text_files, 
                        invert = T, 
                        value = T)


print("Number of training sample papers with full text retrieved:")
sum(training_sample %in% full_text_pmids)

print("Number of unique pubmed ids for disease studies with full text retrieved:")
gwas_study_info |>
  filter(PUBMED_ID %in% full_text_pmids) |>
  select(PUBMED_ID, cause) |>
  distinct() |>
  group_by(cause) |>
  summarise(n = n()) 

```

### Step 2: Get list of terms to use for cohort context matching

#### Source of terms: GWAS Catalog Ancestry Metadata

GWAS ancestry data.frame contains: country of recruitment, ancestry label, population descriptors

```{r gwas_cat_load_ncestry_info}

gwas_ancest_info <-  data.table::fread(here::here("data/gwas_catalog/gwas-catalog-v1.0.3.1-ancestries-r2025-07-21.tsv"),
                           sep = "\t",
                           quote = "")

gwas_ancest_info = gwas_ancest_info |>
  dplyr::rename_with(~ gsub(" ", "_", .x))

```

##### Countries 

```{r get_countries_gwas_cat, eval = F}

countries <- 
gwas_ancest_info$COUNTRY_OF_RECRUITMENT |>
  stringr::str_split("\\,") |>
  unlist() |>
  str_trim() |>
  unique() 
  
countries <- countries[countries != "NR"]
countries <- c(countries, 
               "United States", 
               "United Kingdom",
               "Korea",
               "America",
               "USA",
               "United States of America",
               "Latin America")


```

##### Population descriptors

```{r get_population_descriptors_gwas_cat, eval = F}

population_descriptors <- 
  c(gwas_ancest_info$BROAD_ANCESTRAL_CATEGORY, gwas_ancest_info$ADDITIONAL_ANCESTRY_DESCRIPTION) |>
  str_replace_all("Recruitment in North, Central, and South America, Asia, and Europe", 
                  "North America, Central America, Asia, Europe") |>
  #stringr::str_split("\\,|;|heritage:\\|") |>
  stringr::str_split("\\,|;|heritage:|") |>
  unlist() |>
  str_trim() 

# processing population descriptors to remove extra information and get more general terms to match cohort descriptions in abstracts (e.g. "European" instead of "European (non-Finnish)")
population_descriptors <- population_descriptors |>
  str_remove_all("\\.$")  |> # Escape the dot to make it literal
  str_remove_all("^and ") |> 
  str_remove_all(" \\(founder/genetic isolate\\)$|\\(founder/genetic isolate$") |>
  #str_remove_all("^[0-9+]% ") |>
  str_remove_all("^\\d+% ") |>  # \d is shorthand for [0-9]
  str_remove_all("^[‡†*]+") |> # Remove dagger symbols at start 
  str_remove_all(" - Korea Association Resource \\(KARE\\)$") |>
  str_remove_all("or unknown ancestry$|\\(middle eastern$") |>
  str_remove_all("\\(see Springelkamp 2017\\)") |>
  str_remove_all("parents & grandparents born in ") |>
  str_remove_all(" cell lines$| cell line$") |>
  str_remove_all(" cohort$") |>
  str_remove_all(" population$") |> 
  str_remove_all("^Erasmus Rucphen in |^Erasmus Rucphen") |>
  str_remove_all("^including ") |>
  str_remove_all("\\(Middle Eastern$") |>
  unique() 


# removing population descriptors that are not valid to match cohort descriptions in abstracts (e.g. "cases", "controls", "Bipolar disorder")
not_valid_pop_descriptors <- c("NR", "N.R", "Other", "other", "", "unspecified",
                               "East", "Euopre",
                               "UKB", "UKBB", "DECODE", "Controls from UKBiobank", "UKBB and deCODE", 
                               "NHAPC", "GeneID-I", " Family", 
                               "non-Hispaniic white", "Zimbabweian", "Portugese", "Europen American",
                               "(see Moffatt et al 2010)", "See Wu J. H", 
                               "et al. 2013. PMID: 23362303", "See Wu et al 23362303", "23362303",
                               "See Locke (PubMed 25673413) for BMI and Shungin (PubMed 25673412) for WHR")

disease_descriptors <- grep("cases|Bipolar|controls|Schizophrenia|disorder", 
                            population_descriptors, 
                            value = T)

population_descriptors <- population_descriptors[!(population_descriptors %in% 
                                                     c(not_valid_pop_descriptors,
                                                       disease_descriptors
                                                       )
                                                     )]

```

#### Source of terms: hancestro ontology of human ancestry terms

```{r get_hancestro_terms, eval = F}

# Get all terms in one go (if total < 500)
response <- GET("https://www.ebi.ac.uk/ols4/api/ontologies/hancestro/terms?size=1000")
data <- fromJSON(content(response, "text"), flatten = TRUE)
hancestro_terms <- data$`_embedded`$terms$label

# Remove terms with specific biobanks / datasets in brackets, as these are unlikely to be used in cohort descriptions in abstracts / texts
hancestro_terms <- str_remove_all(hancestro_terms, 
                                  " \\(SGDP\\)$| \\(1KGP\\)$| \\(HGDP\\)$| \\(GGVP\\)$")

# Remove obsolete terms
hancestro_terms <- grep("obsolete|obsolescence", 
                        hancestro_terms, 
                        value = TRUE, 
                        invert = TRUE)

# remove terms with specific regions in brackets, as these are unlikely to be used in cohort descriptions in abstracts / texts
hancestro_terms <- str_remove_all(hancestro_terms, pattern = "\\(Carmel\\)$|\\(pre1989\\)$|\\(Bergamo\\)$|\\(Negev\\)$|\\(Central\\)|\\(Caucasus\\)")

# replace underscores with spaces
hancestro_terms <- str_replace_all(hancestro_terms, 
                                   pattern = "_", 
                                   replacement = " ")

hancestro_terms <- unique(hancestro_terms)

# specific hancestro terms to exclude: 
terms_to_exclude <- c("ancestry category",
                      "ancestry status",
                      "BFO 0000006",
                      "Country",
                      "continent",
                      "continuant",
                      "curation status specification",
                      "denotator type",
                      "entity",
                      "ethnicity category",
                      "ethnicity descriptor",
                      "geographic location",
                      "geographic descriptor",
                      "geography-based population category",
                      "immaterial entity",
                      "independent continuant",
                      "!Kung",
                      "material entity",
                      "quality",
                      "reference population",
                      "region",
                      "organization",
                      "population",
                      "specifically dependent continuant",
                      "Thing",
                      "uncategorised population",
                      "undefined ancestry population")

hancestro_terms <- hancestro_terms[!c(hancestro_terms %in% terms_to_exclude)]

```

#### ? Other possible sources

Country R packages 
Population descriptors in: https://pmc.ncbi.nlm.nih.gov/articles/PMC8715140/#sec2

#### Combine sources 

```{r combine_cohort_context_terms, eval = F}

# combine terms from different sources, and add some extra terms based on looking at cohort descriptions in abstracts
cohort_context_terms <-
c(countries, hancestro_terms, population_descriptors) |> 
  unique() |>
  sort()

# terms to add: 
cohort_context_terms <- c(cohort_context_terms, 
                          "Scandinavians",
                          "Native Hawaiians")

# terms to remove ... "Qatar", "UK", "Taiwan", "Korean"
cohort_context_terms <- cohort_context_terms[!cohort_context_terms %in% c("Qatar", 
                                                                          "UK", 
                                                                          "Taiwan",
                                                                          "Korean",
                                                                          "population")]

# replace brackets with \\( & and \\)
cohort_context_terms <- cohort_context_terms |> 
  str_replace_all(pattern = "\\(", replacement = "\\\\(") |>
  str_replace_all(pattern = "\\)", replacement = "\\\\)")



cohort_context_terms <- c(cohort_context_terms, 
                          "UK\\b(?! Biobank)",
                          "UK\\b\\(?! Biobank\\)",
                          "Qatar\\b\\(?! Biobank\\)",
                          "Taiwan\\b\\(?! Biobank\\)",
                          "Korean\\b\\(?! Biobank\\)"
                          )

# sort by length of term (longest first) to match longest names first
cohort_context_terms <- cohort_context_terms[order(-nchar(cohort_context_terms))]


writeLines(cohort_context_terms, 
           here::here("output/gwas_cat/ancestry_population_terms.txt"))

```

```{r load_ancestry_population_terms}

cohort_context_terms <- readLines(here::here("output/gwas_cat/ancestry_population_terms.txt"))

print("Number of cohort context terms:")
length(cohort_context_terms)

```

### Step 3: Get cohort information

```{r gwas_cat_add_cohort_info}

gwas_study_info_cohort =
  data.table::fread(here::here("output/gwas_cohorts/gwas_cohort_name_corrected.csv"))

gwas_study_info_cohort =
  gwas_study_info_cohort |>
  dplyr::rename_with(~ gsub(" ", "_", .x))

gwas_study_info_cohort =
  gwas_study_info_cohort |>
  select(PUBMED_ID,
         COHORT) |>
  distinct()

print("Check adding cohort information has only added columns, not rows:")
print("Before adding:")
dim(gwas_study_info)

gwas_study_info =
  left_join(gwas_study_info,
            gwas_study_info_cohort,
            by = "PUBMED_ID"
  )

print("After adding:")
dim(gwas_study_info)


print("Check adding cohort information has not increaed number of unique pubmed ids:")
gwas_study_info |>
  pull(PUBMED_ID) |>
  unique() |>
  length()

print("Number of unique pubmed ids for disease studies with cohort info:")
gwas_study_info |>
  filter(COHORT != "") |>
  pull(PUBMED_ID) |>
  unique() |>
  length()

```


### Step 4: Add ancestry/population info

```{r gwas_cat_add_ancestry_info}

gwas_ancest_info =
  gwas_ancest_info |>
  select(PUBMED_ID,
         BROAD_ANCESTRAL_CATEGORY,
         COUNTRY_OF_RECRUITMENT) |>
  distinct() |>
  group_by(PUBMED_ID) |>
  summarise(
    BROAD_ANCESTRAL_CATEGORY = paste(
      unique(
        unlist(strsplit(BROAD_ANCESTRAL_CATEGORY, split = "\\|"))
      ),
      collapse = "|"
    ),
    COUNTRY_OF_RECRUITMENT = paste(
      unique(
        unlist(strsplit(COUNTRY_OF_RECRUITMENT, split = "\\|"))
      ),
      collapse = "|"
    )
  )

print("Check adding ancestry information has only added columns, not rows:")
print("Before adding:")
dim(gwas_study_info)

gwas_study_info =
  left_join(gwas_study_info,
            gwas_ancest_info,
            by = "PUBMED_ID"
  )

print("After adding:")
dim(gwas_study_info)

```

### Step 5: Extract information required to get paper text & abstracts

```{r gwas_cat_extract_info}

gwas_study_info <-
  gwas_study_info |>
  #filter(COHORT != "") |>
  select(PUBMED_ID,
         COHORT,
         YEAR,
         BROAD_ANCESTRAL_CATEGORY,
         COUNTRY_OF_RECRUITMENT) |>
  distinct() |>
  group_by(PUBMED_ID,
           YEAR,
           BROAD_ANCESTRAL_CATEGORY,
           COUNTRY_OF_RECRUITMENT) |>
  summarise(
    COHORT = paste(
      unique(
        unlist(strsplit(COHORT, split = "\\|"))
      ),
      collapse = "|"
    )
  )

gwas_study_info = 
  gwas_study_info |>
  ungroup() |>
  arrange(PUBMED_ID)

pmids = gwas_study_info$PUBMED_ID

cohort = gwas_study_info$COHORT
names(cohort) = pmids

date = gwas_study_info$YEAR
names(date) = pmids

country = gwas_study_info$COUNTRY_OF_RECRUITMENT
names(country) = pmids

ancestry = gwas_study_info$BROAD_ANCESTRAL_CATEGORY
names(ancestry) = pmids

print("Number of papers without cohort information:")
gwas_study_info |> 
  filter(COHORT == "") |>
  nrow()

print("Number of papers with cohort information:")
gwas_study_info |>
  filter(COHORT != "") |>
  nrow()

```

### Step 6: Get data dictionary of cohort names 

```{r get_cohort_names}

# this xlsx was built from looking at acrynyms / cohort names in the gwas catalog
# and finding the corresponding full names / details of cohorts

cohort_names <- readxl::read_xlsx(here::here("data/cohort/cohort_desc.xlsx"),
                                 sheet = 1) |>
  mutate(across(everything(), 
                ~stringr::str_replace_all(.x,
                                          pattern = "\u00A0",
                                          replacement = " "))
         ) 


cohort_full_names = cohort_names$full_name[!is.na(cohort_names$full_name)]
cohort_full_names <- str_trim(cohort_full_names)
cohort_full_names <- iconv(cohort_full_names, to = "UTF-8")
cohort_full_names <- gsub("[\u00A0\r\n]", " ", cohort_full_names)  # replace non-breaking spaces, CR, LF with space
cohort_full_names <- str_squish(cohort_full_names)  # trims and removes extra spaces
# sort by length of name (longest first) to match longest names first
cohort_full_names <- cohort_full_names[order(-nchar(cohort_full_names))]
cohort_full_names <- cohort_full_names[cohort_full_names != "Not Reported"]
cohort_full_names <- unique(cohort_full_names)

print("Number of unique cohort full names:")
length(cohort_full_names)


cohort_abbr_names = cohort_names$cohort[!is.na(cohort_names$cohort)]
cohort_abbr_names <- str_trim(cohort_abbr_names)
cohort_abbr_names <- iconv(cohort_abbr_names, to = "UTF-8")
cohort_abbr_names <- gsub("[\u00A0\r\n]", " ", cohort_abbr_names)  # replace non-breaking spaces, CR, LF with space
cohort_abbr_names <- str_squish(cohort_abbr_names)  # trims and removes extra spaces

# remove abbreviations that are too short
# cohort_abbr_names <- cohort_abbr_names[nchar(cohort_abbr_names) >= 4]
# small_abbr_to_keep <- c("C4D", 
#                         "BBJ", 
#                         "UKB", 
#                         "MVP", 
#                         "TWB", 
#                         "QBB",
#                         "MEC",
#                         "WHI"
#                         )

# cohort_abbr_names <- unique(c(cohort_abbr_names, 
#                               small_abbr_to_keep
#                               ))

cohort_abbr_names <- cohort_abbr_names[!str_detect(pattern = "\\?", 
                                                   cohort_abbr_names)]
# sort by length of name (longest first) to match longest names first
cohort_abbr_names <- cohort_abbr_names[order(-nchar(cohort_abbr_names))]


# add cohort names from GWAS catalog not yet added to data-dictionary
gwas_cat_cohorts = gwas_study_info_cohort$COHORT
gwas_cat_cohorts = unlist(strsplit(gwas_cat_cohorts, "\\|"))
gwas_cat_cohorts = gwas_cat_cohorts[!(gwas_cat_cohorts %in% c("", "other", "multiple"))]

cohort_abbr_names = unique(c(cohort_abbr_names,
                           gwas_cat_cohorts))

# remove small abbreviations that are likely to be false positives
cohort_abbr_names <- cohort_abbr_names[cohort_abbr_names != "DN"]
cohort_abbr_names <- cohort_abbr_names[cohort_abbr_names != "CHB"]
cohort_abbr_names <- cohort_abbr_names[cohort_abbr_names != "FG"]

print("Number of unique cohort abbreviation names:")
length(cohort_abbr_names)

```

# Get cohort-relevant abstract sentences

## Load relevant abstracts

### Convert abstract sections to sentences

```{bash, eval = F}

source spacyr_venv/bin/activate

python3 code/extract_text/spacy_obtain_sentences.py \
--input_dir output/abstracts \
--output_dir output/abstracts

```

```{r read_abstracts_from_files}

abstract_files <- list.files(here::here("output/abstracts/"), 
                             pattern = "*.json", 
                             full.names = FALSE
                             ) |>
                   sort()

abstract_pmids = str_remove_all(abstract_files, "_sentences.json$")
abstract_pmids = abstract_pmids[abstract_pmids %in% pmids]


abstracts <- sapply(abstract_pmids,
                    function(file) {
                      json_data <- fromJSON(here::here(paste0("output/abstracts/", file, "_sentences.json")))
                      
                      abstract_lines <- json_data[json_data != ""]
                      
                      # readLines(here::here(paste0("output/abstracts/",file, ".txt")), 
                      #           warn = FALSE) |> 
                      #   paste(collapse = " ")
                    }
                    )

#pmids = pmids_with_abstracts 
abstract_cohort = cohort[which(pmids %in% abstract_pmids)]
abstract_country = country[which(pmids %in% abstract_pmids)]
abstract_date = date[which(pmids %in% abstract_pmids)]
abstract_ancestry = ancestry[which(pmids %in% abstract_pmids)]


# check lengths of these vectors are the same
print("Check lengths of vectors are the same:")
print(paste("Length of pmids:", length(pmids)))
print(paste("Length of abstract pmids:", length(abstract_pmids)))
print(paste("Length of abstracts:", length(abstracts)))
print(paste("Length of cohort:", length(cohort)))
print(paste("Length of date:", length(date)))

```

###  Check: What abstracts are missing? 

```{r check_missing_abstracts, warning=FALSE, message=FALSE}

missing_abstracts = pmids[!(pmids %in% abstract_pmids)] 

print("Number of missing abstracts:")
length(missing_abstracts)

#pmids <- pmids[pmids %in% abstract_pmids]

```

## Cohort name matching in abstract sentences

### Function to extract sentences with cohort names

```{r fn_extract_cohort_sentences}

extract_cohort_sentences <- function(abstract_list, 
                                     cohort_names, 
                                     column_name = "COHORT",
                                     tokenize = FALSE,
                                     ignore_case) {
  
  cohort_names_grep <- paste0("\\b", cohort_names, "\\b") # add word boundaries to match whole words only
  
  results <- lapply(seq_along(abstract_list), function(i) {
    
    #abstract <- text_vector[i]
    # Split abstract into sentences
    if(tokenize) {
      sentences <- tokenizers::tokenize_sentences(abstract_list[[i]])[[1]]
    } else {
      sentences <-  abstract_list[[i]]
    }
    
    # For each sentence, find all matching cohort names
    lapply(seq_along(sentences), function(s) {
      
      sentence <- sentences[s]
      
      # Identify cohort names present in this sentence
      matched_cohorts <- cohort_names[str_detect(sentence, 
                                                 regex(cohort_names_grep, 
                                                       ignore_case = ignore_case))]
      cohort_df <-
      data.frame(
        article_id = i,
        sentence_id = s,
        sentence = sentence,
        has_cohort = length(matched_cohorts) > 0,
        #!!column_name = if (length(matched_cohorts) > 0) str_flatten(unique(matched_cohorts), collapse = "|", na.rm = T) else "",
        #COHORT = if (length(matched_cohorts) > 0) str_flatten(unique(matched_cohorts), collapse = "|", na.rm = T) else "",
        stringsAsFactors = FALSE
      )
      
      # Add dynamic column
      cohort_df[[column_name]] <- if (length(matched_cohorts) > 0) str_flatten(unique(matched_cohorts), collapse = "|", na.rm = T) else ""
      
      return(cohort_df)
      
    }) |> bind_rows()
  })

 results <- bind_rows(results)
 
 results$pubmed_id <- names(abstract_list)[results$article_id]
 
 return(results) #returns a df of sentences, with labelled columns for whether they contain cohort names, and which cohort names they contain (if any)
 # id of sentence, and abstract / article
}

```

### Get cohort relevant sentences 

#### Get cohort name sentences

```{r extract_cohort_sentences_from_abstracts}

# extract sentences with cohort full names (case-insensitive matching, as full names are less likely to be ambiguous)
cohort_sentences_df_p1 <- extract_cohort_sentences(abstracts,
                                                cohort_full_names,
                                                ignore_case = TRUE
                                                )

# extract sentences with cohort abbreviation names (case-sensitive matching, as abbreviations are more likely to be ambiguous)
cohort_sentences_df_p2 = extract_cohort_sentences(abstracts,
                                                  cohort_abbr_names,
                                                  ignore_case = FALSE
                                                  )

cohort_sentences_df =
  bind_rows(cohort_sentences_df_p1,
            cohort_sentences_df_p2
            )


cohort_sentences_df = 
  cohort_sentences_df |>
  distinct()

separate_cohorts <- function(COHORT) {
  
  if (any(grepl("\\|", COHORT))) {
    return(unlist(strsplit(COHORT, "\\|")))
  } else {
    return(COHORT)
  }
  
}

cohort_sentences_df =
  cohort_sentences_df |>
  group_by(article_id, sentence_id, sentence) |>
  summarise(
    COHORT = str_flatten(unique(separate_cohorts(COHORT)), 
                         collapse = "|", 
                         na.rm = T),
    has_cohort = ifelse(any(COHORT != ""), TRUE, FALSE)
  ) |>
  ungroup() 

cohort_sentences_df =
  cohort_sentences_df |>
  mutate(COHORT = str_remove_all(COHORT, 
                               pattern = "\\|$|^\\|"
                               )
         ) 

```

#### Distinguishing words between sentences with cohort names, and sentences without cohort names

```{r distinguish_cohort_sentences_abstracts}

# remove stop words and common words that are not informative for cohort context
removal_words <- c("the", "and", "of", "in", "to", "with", "a",
                   "for", "on", "by", "is", "are", "was", "were", "each", "all", "had", "have", "it",
                   "as", "from", "that", "this", "which", "be", "at", "or", "an", "then", "than",  "into",
                   "if", "not", "only", "both", "same", "after", "across", "between", "out", "up", "any", 
                   "we", "our", "us", "these", "within", "per",
                   "used", "using", "use",
                  "0.01", "0.05", "0.5", "1", "one", "2", "two", "3", "three", "iii",
                  "4", "5", "6", "p", "8",  "10", "value", "significant", "30",
                   "i", "r", "wide"
)

cohort_sentences_words <- 
cohort_sentences_df |>
  filter(has_cohort) |>
  pull(sentence) |>
  tokenizers::tokenize_words() |>
  unlist()

cohort_sentence_words_df =
data.frame(word = cohort_sentences_words) |>
  filter(!(tolower(word) %in% removal_words)) |>
  group_by(word) |>
  summarise(n_in_cohort = n()) |>
  filter(n_in_cohort > 5)

non_cohort_sentences_words <-
cohort_sentences_df |>
  filter(!has_cohort) |>
  pull(sentence) |>
  tokenizers::tokenize_words() |>
  unlist()

# sample down non-cohort sentences words
set.seed(500)
non_cohort_sentences_words_sample <- sample(non_cohort_sentences_words, 
                                              size = length(cohort_sentences_words), 
                                              replace = FALSE)

non_cohort_sentence_words_df =
data.frame(word = non_cohort_sentences_words_sample) |>
  filter(!(tolower(word) %in% removal_words)) |>
  group_by(word) |>
  summarise(n_not_cohort = n()) 


n_sentences <- cohort_sentences_df |>
  filter(COHORT != "") |>
  select(article_id, sentence_id) |>
  distinct() |>
  nrow()


print("Words more common in sentences with cohort names than sentences without cohort names:")
left_join(cohort_sentence_words_df,
          non_cohort_sentence_words_df,
          by = "word"
          ) |>
  mutate(n_not_cohort = ifelse(is.na(n_not_cohort), 0, n_not_cohort)) |>
  mutate(delta_n = n_in_cohort - n_not_cohort) |>
  mutate(recall = n_in_cohort / n_sentences) |>
  arrange(desc(delta_n)) |>
  head(20)

```

```{r cohort_sentence_summary}

print("Number of abstracts containing probable cohort reference")
abstracts_cohorts <- cohort_sentences_df |>
  filter(COHORT != "") |>
  pull(article_id) |>
  unique() 

n_abstracts_cohorts =  abstracts_cohorts |>
  length()

print(n_abstracts_cohorts)

print("Percentage of sampled abstracts containing probable cohort reference:")
100 * n_abstracts_cohorts / length(unique(pmids))

print("Number of sentences containing probable cohort reference")
  cohort_sentences_df |>
  filter(COHORT != "") |>
  select(article_id, sentence_id) |>
  distinct() |>
  nrow()
  
print("Number of entities")
cohort_sentences_df |>
  filter(COHORT != "") |>
  pull(COHORT) |>
  str_split("\\|") |>
  unlist() |>
  length()
  
print("Number of unique cohorts referenced (not unique cohort names, but unique strings in the COHORT column):")
detected_cohorts <- cohort_sentences_df |>
  filter(COHORT != "") |>
  pull(COHORT) |>
  str_split("\\|") |>
  unlist() 

n_cohorts <- detected_cohorts |>
  unique()  |>
  length()

print(n_cohorts)

print("Most common cohort names detected (by number of sentences they are mentioned in):")
data.frame(cohort = detected_cohorts) |>
  group_by(cohort) |>
  summarise(n = n()) |>
  arrange(desc(n)) |>
  head(10)

print("Most common cohort names detected (by number of abstracts they are mentioned in):")
cohort_sentences_df |>
  filter(COHORT != "") |>
  select(COHORT, article_id) |>
  tidyr::separate_rows(COHORT, sep = "\\|") |>
  distinct() |>
  group_by(COHORT) |>
  summarise(n_abstracts = n()) |>
  arrange(desc(n_abstracts)) |>
  head(5)
  
```

#### Add cohort-related context sentences

```{r extract_cohort_context_sentences_from_abstracts}

# extract sentences with cohort context terms (case-sensitive matching, as some context terms are common words that could be ambiguous)
cohort_context_df = extract_cohort_sentences(abstracts,
                                            cohort_context_terms,
                                            column_name = "CONTEXT",
                                            ignore_case = FALSE
                                            )

cohort_sentences_df <-
left_join(cohort_sentences_df,
          cohort_context_df |> select(article_id, 
                                      sentence_id, 
                                      CONTEXT),
          by = c("article_id", "sentence_id")
          ) 

print("Number of sentences with cohort names and/or cohort-related context:")
cohort_sentences_df |> 
  mutate(has_context = ifelse(CONTEXT == "", F, T)) |>
  group_by(has_cohort, has_context) |> 
  summarise(n = n())

filtered_cohort_sentences_df =
  cohort_sentences_df |>
  filter(COHORT != "" | CONTEXT != "")

print("Number of sentences with cohort names or cohort-related context (population, ancestry, country terms):")
nrow(filtered_cohort_sentences_df)

```

#### Get other sentences with words that indicate sample or cohort is being discussed

Using this to correct for the over-representation of sentences without cohort names, by identifying well-matched sentences

```{r extract_sample_context_sentences_from_abstracts}

# words that indicate sample or cohort is being discussed:
sample_terms <- c(
  # Your original list
  "ancestry", "biobank", "cases", "cohort", "controls", 
  "consortium", "consortia", "descent", "founder", 
  "enrolled", "enrollment", "ethnicity", "heritage",
  "isolate", "isolated", "individuals", "participants", 
  "patients", "population", "recruitment", "registry", 
  "sample", "study", "subjects", "volunteer",
  
  # Additional core terms
  "recruited", "enroll", "ascertained", "volunteers",
  "probands", "affected", "unaffected", "families", "twins",
  "cohorts", "populations", "samples", "subgroup", "subset",
  "subcohort", "demographic", "ethnic", "racial",
  "admixed", "admixture", "ancestral", "origin",
  "case-control", "prospective", "genotyped", "sequenced",
  "meta-analysis", "pooled", "residing", "nationwide",
  
  # Useful variants
  "enrollees", "sampling", "lineage", "self-reported",
  "self-identified", "replication", "discovery", "validation"
)

control_cohort_sentences_df <-
cohort_sentences_df |>
  filter(COHORT == "" & CONTEXT == "") |>
  filter(grepl(paste(sample_terms, collapse = "|"), 
               sentence, 
               ignore.case = TRUE)
         ) #|>

print("Number of sentences without (detected) cohort names or cohort-context but with sample-related context:")
nrow(control_cohort_sentences_df)

filtered_cohort_sentences_df =
  bind_rows(filtered_cohort_sentences_df,
            control_cohort_sentences_df
            ) |>
  distinct()

print("Percentage of sentences with cohort names vs other cohort-related context")
total_n_sentences <- nrow(filtered_cohort_sentences_df)

filtered_cohort_sentences_df |>
  group_by(has_cohort) |> 
  summarise(n = n(),
            percentage = round(100 * n/total_n_sentences, digits = 2)
  )

```

```{r get_ids_for_doccano}

abstract_pmids = filtered_cohort_sentences_df$pubmed_id |> unique() 
abstract_ids <- filtered_cohort_sentences_df$article_id |> unique()
abstract_date = abstract_date[abstract_ids]
abstract_cohort = abstract_cohort[abstract_ids]
abstract_country = abstract_country[abstract_ids]

```

## Prepare data in Doccano JSON format

```{r prepare_doccano_json_abstracts}

# file path for intermediate json file output
json_file = here::here("output/doccano/abstracts_with_cohort_info.json")

# file path for final jsonl file output
jsonl_file = here::here("output/doccano/abstracts_with_cohort_info.jsonl")

convert_to_doccano_json_sentence_level <- function(pmids,
                                                   date,
                                                   cohort,
                                                   country,
                                                   cohort_sentences_df) {
  
  # set up json list
  doccano_list <- list()
  example_id <- 1

  for(current_sentence in cohort_sentences_df$sentence) {

      # Filter cohort sentences that match this sentence (safe matching)
      df <- cohort_sentences_df |>
             dplyr::filter(sentence == current_sentence) 
      
      # article_id <- df$article_id
      # matched_cohort <- df$COHORT
      
      for (i in seq_len(nrow(df))) {
        
        matched_cohort <- df$COHORT[i]
        article_id <- df$article_id[i]
        
        #browser()
      
        if(matched_cohort == ""){
        
          doccano_list[[example_id]] <- list(
          text = current_sentence,
          pubmed_id = pmids[article_id],
          date = date[article_id],
          country = country[article_id],
          gwas_cat_cohort_label = cohort[article_id],
          label = list()
          )
        
        } else {
        
        # Find the location of all matches of the cohort name in the sentence
        if(grepl("\\|", matched_cohort)) {
           
           # If multiple cohort names, separate
           matched_cohort <- unlist(
                                    strsplit(matched_cohort, 
                                             split = "\\|")
                                    )
        
           match_locations <- list()
        
           for(current_matched_cohort in matched_cohort){
          
               matches <- str_locate_all(current_sentence,
                                         fixed(current_matched_cohort, 
                                               ignore_case = TRUE)
                                         )[[1]]
          
               match_locations <- append(match_locations, 
                                         list(matches)
                                         )
          
           }
        
           # Combine all match locations into a single matrix
           matches <- do.call(rbind, match_locations)
        
        }  else {
        
           matches <- str_locate_all(current_sentence,
                                     fixed(matched_cohort, 
                                           ignore_case = TRUE)
                                     )[[1]]
      
      }
      
      # Convert matches to 0-based indexing (for doccano)
      matches[, "start"] <- matches[, "start"] - 1  
      
      # Turn match locations into entity list
      entities <- list()
      
      for(k in seq_len(nrow(matches))) {
              entities <- append(entities, list(list(
                start_offset = matches[k, "start"],
                end_offset = matches[k, "end"],
                label = "COHORT"
              )))
      }
      
      # Create Doccano JSON entry
      doccano_list[[example_id]] <- list(
        # id = example_id,
        text = current_sentence,
        pubmed_id = pmids[article_id],
        date = date[article_id],
        country = country[article_id],
        gwas_cat_cohort_label = cohort[article_id],
        label = entities
      )
      
      }
      }

      example_id <- example_id + 1
  }
  
  
# Suppose each element of json_list has 'labels' as a list of named lists
  doccano_list <- lapply(doccano_list, function(x) {
  x$label <- lapply(x$label, function(l) {
    # convert named list to vector/list format [start, end, label]
    c(l$start_offset, l$end_offset, l$label)
  })
  x
})

  return(doccano_list)
}


# Create JSON as a list
json_list <- convert_to_doccano_json_sentence_level(pmids = abstract_pmids, #pmids,
                                                    date = abstract_date,
                                                    cohort = abstract_cohort,
                                                    country = abstract_country,
                                                    filtered_cohort_sentences_df)



# Write to JSON file
writeLines(toJSON(json_list,
                  auto_unbox = TRUE,
                  pretty = TRUE),
           json_file)


json_data <- fromJSON(json_file,
                      simplifyVector = FALSE)

# Open connection to JSONL file
con <- file(jsonl_file, "w")

# Loop over each element (object) and write as one line
for (i in seq_along(json_data)) {
  writeLines(toJSON(json_data[[i]], auto_unbox = TRUE), con)
}

# Close connection
close(con)

cat("JSONL saved to:", jsonl_file, "\n")

```

# Get cohort-revelevant methods section sentence extraction

## Convert methods sections to sentences

```{bash, eval = F}

source spacyr_venv/bin/activate

python3 code/extract_text/spacy_obtain_sentences.py \
--input_dir output/fulltexts/methods_sections \
--output_dir  output/fulltexts/methods_sentences

```

## Get relevant methods text:

```{r}

methods_sections <- list.files(here::here("output/fulltexts/methods_sentences/"), 
                             pattern = "*.json", 
                             full.names = TRUE
                             )

pmcids_with_methods <- methods_sections |>
  gsub(pattern = ".*/", replacement = "") |>
  gsub(pattern = "_methods_sentences\\.json$", replacement = "")

pubmeds_with_methods <- converted_ids |>
  filter(pmcids %in% pmcids_with_methods) |>
  pull(PMID)

pubmeds_with_methods <- c(pubmeds_with_methods, 
                          grep("PMC", pmcids_with_methods, value = TRUE, invert = TRUE)
                          )

print("Number of papers with methods sections extracted:")
sum(pubmeds_with_methods %in% pmids)

pmids_methods_to_get <-  pubmeds_with_methods[pubmeds_with_methods %in% pmids]

pmcids_methods_to_get <- converted_ids |>
  filter(PMID %in% pmids_methods_to_get) |>
  pull(pmcids)

ids_methods_to_get <- c(pmids_methods_to_get, pmcids_methods_to_get)

ids_methods_to_get <- ids_methods_to_get[ids_methods_to_get != ""]

files_to_get <-
paste0(ids_methods_to_get, 
       "_methods_sentences.json") |>
  sort()

methods_sections_to_get <- methods_sections[basename(methods_sections) %in% files_to_get]
methods_sections_to_get <- basename(methods_sections_to_get) |> 
                           gsub(pattern = "_methods_sentences\\.json$", replacement = "")

# read in methods sections for available papers
methods_texts <- sapply(methods_sections_to_get,
                    function(id) {
                      
                      file <- here::here(paste0("output/fulltexts/methods_sentences/",
                                                id, 
                                                "_methods_sentences.json"))
                      
                      json_data <- fromJSON(file)
                      
                      abstract_lines <- json_data[json_data != ""]
                      # readLines(file, 
                      #           warn = FALSE, 
                      #           encoding = "UTF-8") |>
                      # paste(collapse = " ")
                    }
                    )

# convert pmcids to pubmeds
names(methods_texts) <- sapply(names(methods_texts), function(id) {
  if(grepl("^PMC", id)) {
    pmcid <- id
    pubmed_id <- converted_ids$PMID[converted_ids$pmcids == pmcid]
    if(length(pubmed_id) > 0) {
      return(as.character(pubmed_id))
    } else {
      return(NA)
    }
  } else {
    return(id)
  }
})

# order methods texts by pmids
methods_texts <- methods_texts[order(names(methods_texts))]

methods_pmids <- names(methods_texts)
methods_cohort <- cohort[which(pmids %in% methods_pmids)]
methods_country <- country[which(pmids %in% methods_pmids)]
methods_date <- date[which(pmids %in% methods_pmids)]

```

## Extract sentences with cohort names or cohort-related context (population, ancestry, country terms) from methods sections

```{r extract_cohort_sentences_methods}

# extract sentences with cohort full names (case-insensitive matching, as full names are less likely to be ambiguous)
cohort_sentences_df_p1 <- extract_cohort_sentences(methods_texts,
                                                cohort_full_names,
                                                ignore_case = TRUE
                                                )

# extract sentences with cohort abbreviation names (case-sensitive matching, as abbreviations are more likely to be ambiguous)
cohort_sentences_df_p2 = extract_cohort_sentences(methods_texts,
                                                  cohort_abbr_names,
                                                  ignore_case = FALSE
                                                  )

methods_cohort_sentences_df = 
  bind_rows(cohort_sentences_df_p1,
            cohort_sentences_df_p2
            ) |>
  distinct()


methods_cohort_sentences_df =
  methods_cohort_sentences_df |>
  group_by(article_id, sentence_id, sentence) |>
  summarise(
    COHORT = str_flatten(unique(separate_cohorts(COHORT)), 
                         collapse = "|", 
                         na.rm = T),
    has_cohort = ifelse(any(COHORT != ""), TRUE, FALSE)
  ) |>
  ungroup() 

methods_cohort_sentences_df =
  methods_cohort_sentences_df |>
  mutate(COHORT = str_remove_all(COHORT, 
                               pattern = "\\|$|^\\|"
                               )
         ) 

print("Number of methods sections containing probable cohort reference:")
methods_cohorts <- methods_cohort_sentences_df |>
  filter(COHORT != "") |>
  pull(article_id) |>
  unique()

methods_cohorts  |> length()

print("Percentage of methods sections containing probable cohort reference:")
100 * length(methods_cohorts) / length(methods_texts)


print("Number of sentences containing probable cohort reference in methods sections")
  methods_cohort_sentences_df |>
  filter(COHORT != "") |>
  select(article_id, sentence_id) |>
  distinct() |>
  nrow()

print("Number of entities in methods sections")
methods_cohort_sentences_df |>
  filter(COHORT != "") |>
  pull(COHORT) |>
  str_split("\\|") |>
  unlist() |>
  length()

print("Most common cohort names detected in methods sections:")
methods_cohort_sentences_df |>
  filter(COHORT != "") |>
  select(COHORT, article_id) |>
  tidyr::separate_rows(COHORT, sep = "\\|") |>
  distinct() |>
  group_by(COHORT) |>
  summarise(n_abstracts = n()) |>
  arrange(desc(n_abstracts)) |>
  head(5)

print("Number of cohort names only mentioned in one paper")
methods_cohort_sentences_df |>
  filter(COHORT != "") |>
  select(COHORT, article_id) |>
  tidyr::separate_rows(COHORT, sep = "\\|") |>
  distinct() |>
  group_by(COHORT) |>
  summarise(n_abstracts = n()) |>
  filter(n_abstracts == 1) |>
  nrow()

print("Number of unique cohorts referenced in methods sections (not unique cohort names, but unique strings in the COHORT column):")
detected_cohorts_methods <- methods_cohort_sentences_df |>
  filter(COHORT != "") |>
  pull(COHORT) |>
  str_split("\\|") |>
  unlist()

n_cohorts_methods <- detected_cohorts_methods |>
  unique()  |>
  length()

print(n_cohorts_methods)

print("Most common cohort names detected in methods sections by number of sentences they are mentioned in:")
data.frame(cohort = detected_cohorts_methods) |>
  group_by(cohort) |>
  summarise(n = n()) |>
  arrange(desc(n)) |>
  head(10)


# 
# methods_cohort_sentences_df |>
#   filter(!c(article_id %in% methods_abstracts_cohorts)) |>
#   View()

```

## Distinguishing words between sentences with cohort names, and sentences without cohort names in methods sections

```{r distinguish_cohort_sentences_methods}

cohort_sentences_words <- 
cohort_sentences_df |>
  filter(has_cohort) |>
  pull(sentence) |>
  tokenizers::tokenize_words() |>
  unlist()

cohort_sentence_words_df =
data.frame(word = cohort_sentences_words) |>
  filter(!(tolower(word) %in% removal_words)) |>
  group_by(word) |>
  summarise(n_in_cohort = n()) |>
  filter(n_in_cohort > 5)

non_cohort_sentences_words <-
cohort_sentences_df |>
  filter(!has_cohort) |>
  pull(sentence) |>
  tokenizers::tokenize_words() |>
  unlist()

# sample down non-cohort sentences words
set.seed(500)
non_cohort_sentences_words_sample <- sample(non_cohort_sentences_words, 
                                              size = length(cohort_sentences_words), 
                                              replace = FALSE)

non_cohort_sentence_words_df =
data.frame(word = non_cohort_sentences_words_sample) |>
  filter(!(tolower(word) %in% removal_words)) |>
  group_by(word) |>
  summarise(n_not_cohort = n()) 


print("Words more common in sentences with cohort names than sentences without cohort names:")
left_join(cohort_sentence_words_df,
          non_cohort_sentence_words_df,
          by = "word"
          ) |>
  mutate(n_not_cohort = ifelse(is.na(n_not_cohort), 0, n_not_cohort)) |>
  mutate(delta_n = n_in_cohort - n_not_cohort) |>
  arrange(desc(delta_n)) |>
  head(20)

```

## Add cohort-related context sentences to methods sections

```{r extract_cohort_context_sentences_methods}

# extract sentences with cohort context terms (case-sensitive matching, as some context terms are common words that could be ambiguous)
cohort_context_df = extract_cohort_sentences(methods_texts,
                                            cohort_context_terms,
                                            column_name = "CONTEXT",
                                            ignore_case = FALSE
                                            )

methods_cohort_sentences_df <-
left_join(methods_cohort_sentences_df,
          cohort_context_df |> select(article_id, sentence_id, CONTEXT),
          by = c("article_id", "sentence_id")
          ) 

print("Number of sentences with cohort names and/or cohort-related context:")
methods_cohort_sentences_df |> 
  mutate(has_context = ifelse(CONTEXT == "", F, T)) |>
  group_by(has_cohort, has_context) |> 
  summarise(n = n())

filtered_cohort_sentences_df =
  methods_cohort_sentences_df |>
  filter(COHORT != "" | CONTEXT != "")

print("Number of sentences with cohort names or cohort-related context (population, ancestry, country terms):")
nrow(filtered_cohort_sentences_df)

```

## Get other sentences with words that indicate sample or cohort is being discussed

```{r get_sample_context_sentences_methods}

control_cohort_sentences_df <-
cohort_sentences_df |>
  filter(COHORT == "" & CONTEXT == "") |>
  filter(grepl(paste(sample_terms, collapse = "|"), 
               sentence, 
               ignore.case = TRUE)
         ) #|>

print("Number of sentences without (detected) cohort names or cohort-context but with sample-related context:")
nrow(control_cohort_sentences_df)

filtered_cohort_sentences_df =
  bind_rows(filtered_cohort_sentences_df,
            control_cohort_sentences_df
            ) |>
  distinct()

print("Percentage of sentences with cohort names vs other cohort-related context")
total_n_sentences <- nrow(filtered_cohort_sentences_df)

filtered_cohort_sentences_df |>
  group_by(has_cohort) |> 
  summarise(n = n(),
            percentage = round(100 * n/total_n_sentences, digits = 2)
  )

```

## Prepare data in Doccano JSON format

```{r prepare_doccano_json_methods}

# file path for intermediate json file output
json_file = here::here("output/doccano/methods_with_cohort_info.json")

# file path for final jsonl file output
jsonl_file = here::here("output/doccano/methods_with_cohort_info.jsonl")


methods_pmids <- filtered_cohort_sentences_df$pubmed_id |> unique()
article_ids <- filtered_cohort_sentences_df$article_id |> unique()

methods_cohort = methods_cohort[article_ids]
country = country[article_ids]
date = date[article_ids]
ancestry = ancestry[article_ids]

# Create JSON as a list
json_list <- convert_to_doccano_json_sentence_level(pmids = pmids,
                                                    date = date,
                                                    cohort = cohort,
                                                    country = country,
                                                    filtered_cohort_sentences_df)

# Write to JSON file
writeLines(toJSON(json_list,
                  auto_unbox = TRUE,
                  pretty = TRUE),
           json_file)


json_data <- fromJSON(json_file,
                      simplifyVector = FALSE)

# Open connection to JSONL file
con <- file(jsonl_file, "w")

# Loop over each element (object) and write as one line
for (i in seq_along(json_data)) {
  writeLines(toJSON(json_data[[i]], auto_unbox = TRUE), con)
}

# Close connection
close(con)

cat("JSONL saved to:", jsonl_file, "\n")

```


