---
title: "Selecting training and validation sets"
author: "Isobel Beasley"
date: "2026-06-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include = FALSE}
knitr::opts_chunk$set(engine.opts = list(bash = "-l"))
```

```{r, message=FALSE}

library(stringr)
library(readxl)
library(dplyr)
library(ggplot2)

```

### Get text embeddings 

```{bash get_embeddings}

source spacyr_venv/bin/activate

python3 code/text_embeddings/get_text_embeddings.py \
        -m ncbi/MedCPT-Article-Encoder \
        -t output/abstracts \
        -o output/clustering/abstracts_ \
        --gwas-csv output/icd_map/gwas_study_gbd_causes.csv 
        

python3 code/text_embeddings/get_text_embeddings.py \
        -m ncbi/MedCPT-Article-Encoder \
        -t output/methods_sentences \
        -o output/clustering/methods_ \
        --gwas-csv output/icd_map/gwas_study_gbd_causes.csv 

```


# What are the major axes of variation in embeddings for these papers?

```{r gwas_cat_load_preprocess, message=FALSE}

# get only relevant disease studies
gwas_study_info <- data.table::fread(here::here("output/icd_map/gwas_study_gbd_causes.csv"))

gwas_study_info = gwas_study_info |>
  dplyr::rename_with(~ gsub(" ", "_", .x))

# filter out infectious diseases
gwas_study_info <- gwas_study_info |>
    dplyr::filter(!cause %in% c("HIV/AIDS",
                             "Tuberculosis",
                             "Malaria",
                             "Lower respiratory infections",
                             "Diarrhoeal diseases",
                             "Neonatal disorders",
                             "Tetanus",
                             "Diphtheria",
                             "Pertussis" ,
                             "Measles",
                             "Maternal disorders"))

gwas_study_info <-
  gwas_study_info |>
  filter(cause != "")

```



```{r get_lancet_conditions}

# group conditions into broader categories to ensure we have enough papers in each category for sampling
# lancet_cause_mapping <- readxl::read_xlsx(here::here("data/icd/lancet_conditions_icd10.xlsx"),
#                                      sheet = 1) |>
#   dplyr::rename_with(~ gsub(" ", "_", .x))
# 
# gwas_study_info <- gwas_study_info |>
#   left_join(lancet_cause_mapping |> select(cause = mapped_gbd_term,
#                                            lancet_condition), 
#             by = "cause",
#             relationship = "many-to-many") 

gwas_study_info <-
gwas_study_info |>
  select(PUBMED_ID, YEAR, lancet_condition) |>
  distinct() 

gwas_study_info_grouped <- gwas_study_info |>
  group_by(PUBMED_ID) |>
  summarise(
    YEAR = paste(unique(YEAR), collapse = "|"),
    lancet_condition = paste(sort(unique(lancet_condition)), collapse = " & ")
  ) |>
  mutate(lancet_condition_group = case_when(
    str_detect(pattern = "&", lancet_condition) ~ "Multiple",
    TRUE ~ lancet_condition
  ))

gwas_study_info_grouped |>
  group_by(lancet_condition) |>
  summarise(n = n()) |>
  arrange(desc(n))

# most papers study one lancet_condition 

```

## Abstract embeddings

```{r training_set_features}

# get medcpt article embeddings for all disease studies
abstract_embeddings <- data.table::fread(here::here("output/clustering/abstracts_medcpt_embeddings.csv"))

PCA_test <- prcomp(abstract_embeddings, 
                   center = TRUE, 
                   scale. = TRUE) 

pca_table <-
data.frame(PC1 = PCA_test$x[,1], 
           PC2 = PCA_test$x[,2], 
           PC3 = PCA_test$x[,3], 
           PC4 = PCA_test$x[,4], 
           PC5 = PCA_test$x[,5], 
           PC6 = PCA_test$x[,6], 
           PC7 = PCA_test$x[,7], 
           PC8 = PCA_test$x[,8], 
           PC9 = PCA_test$x[,9], 
           PC10 = PCA_test$x[,10],
           PUBMED_ID = abstract_embeddings$PUBMED_ID) 


var_exp <- PCA_test$sdev^2 / sum(PCA_test$sdev^2)

data.frame(PC = paste0("PC", 1:length(var_exp)),
           PC_n = 1:length(var_exp),
           Variance_Explained = var_exp) |>
  filter(PC_n <= 10) |>
  mutate(PC_n = factor(PC_n, levels = 1:10)) |>
  ggplot(aes(x = PC_n, y = Variance_Explained)) +
  geom_col() +
  labs(x = "Principal Component", 
       y = "Proportion of Variance Explained", 
       title = "Variance explained by each principal component") + 
  theme_bw()

```

### Distribution of GWAS Catalog Text Embeddings By Year (Abstracts)

PC1 clearly separates / cluster GWAS Catalog abstract embeddings by year (higher PC1 score for more recent years).

```{r plot_abstract_pca_year}

pca_plot_table <- 
  pca_table |>
  left_join(gwas_study_info_grouped |> 
  select(PUBMED_ID, 
         YEAR,
         lancet_condition,
         lancet_condition_group), 
  by = "PUBMED_ID") |>
  mutate(YEAR = as.factor(YEAR)) 

n_years <- pca_plot_table$YEAR |> unique() |> length()

pca_plot_table |>
  ggplot(aes(
  x = PC1, 
  y = PC2, 
  color = YEAR)) +
  geom_point() +
  scale_color_manual(values = viridisLite::plasma(n = n_years)) +
  labs(x = paste0("PC1 (", round(100 * var_exp[1], digits = 1),
                  "%)"),
       y = paste0("PC2 (", 
                  round(100 * var_exp[2], digits = 1),
                  "%)"),
       colour = "Publication Year"
  ) +
       #title = "PCA of GWAS Catalog abstract embeddings, colored by publication year") + 
  theme_bw()


pca_plot_table |>
    ggplot(aes(x = PC2, 
               y = PC3, 
               color = YEAR)) +
    geom_point() +
    scale_color_manual(values = viridisLite::plasma(n = n_years)) +
    labs(x = paste0("PC2 (", round(100 * var_exp[2], digits = 1),
                  "%)"),
       y = paste0("PC3 (", 
                  round(100 * var_exp[3], digits = 1),
                  "%)"),
       colour = "Publication Year"
  ) + 
  theme_bw()
  
  

```

### Distribution of GWAS Catalog Text Embeddings By Lancet Condition Category (Abstracts)

PC's 2 & 3 clearly separate / cluster GWAS Catalog abstract embeddings by Lancet Condition Category. 

```{r plot_abstract_pca_lancent_condition_category}

  pca_plot_table |>
  ggplot(aes(
  x = PC1, 
  y = PC2, 
  color = lancet_condition_group)) +
  geom_point(size = 0.8) +
  scale_color_brewer(palette = "Dark2") + 
  labs(x = paste0("PC1 (", round(100 * var_exp[1], digits = 1),
                  "%)"),
       y = paste0("PC2 (", 
                  round(100 * var_exp[2], digits = 1),
                  "%)"),
       colour = "Lancet Condition Category"
  )  + 
  theme_bw()


  pca_plot_table |>
  ggplot(aes(
  x = PC2, 
  y = PC3, 
  color = lancet_condition_group)) +
  geom_point(size = 0.8) +
  scale_color_brewer(palette = "Set2") + 
  labs(x = paste0("PC2 (", round(100 * var_exp[2], digits = 1),
                  "%)"),
       y = paste0("PC3 (", 
                  round(100 * var_exp[3], digits = 1),
                  "%)"),
       colour = "Lancet Condition Category"
  )  + 
  theme_bw()

```

```{r}

library(ComplexHeatmap)

gwas_cond_comb_mat = gwas_study_info |>
  mutate(study = 1) |>
  
  tidyr::pivot_wider(names_from = lancet_condition, 
                     values_from = study, 
                     values_fill = 0) |>
  select(-PUBMED_ID, -YEAR) 

gwas_cond_comb_mat <- make_comb_mat(gwas_cond_comb_mat)

# create an upset plot
UpSet(gwas_cond_comb_mat,
      comb_order = order(comb_size(gwas_cond_comb_mat), 
                         decreasing = TRUE)
      )

papers_with_multiple_cond <- pca_plot_table |>
  filter(lancet_condition_group == "Multiple") |> 
  pull(PUBMED_ID)

pca_plot_table_ungrouped <- 
  pca_table |>
  left_join(gwas_study_info,
            by = "PUBMED_ID") |>
  mutate(YEAR = as.factor(YEAR)) 

pca_plot_table_ungrouped <-
  pca_plot_table_ungrouped |>
  mutate(N_conditions = ifelse(PUBMED_ID %in% papers_with_multiple_cond, 
                                         "Multiple",
                                         "Single")) 


pca_plot_table_ungrouped |>
    ggplot(aes(
  x = PC2, 
  y = PC3, 
  color = N_conditions)) +
  geom_point(size = 0.8) +
  scale_color_brewer(palette = "Set2") + 
  labs(x = paste0("PC2 (", round(100 * var_exp[2], digits = 1),
                  "%)"),
       y = paste0("PC3 (", 
                  round(100 * var_exp[3], digits = 1),
                  "%)"),
       colour = "Number of Lancent Categories"
  )  + 
  facet_wrap(~lancet_condition) + 
  theme_bw()


```


```{r}

  pca_plot_table |>
  ggplot(aes(
  x = PC2, 
  y = PC3, 
  color = lancet_condition_group)) +
  geom_point(size = 0.8) +
  scale_color_brewer(palette = "Set2") + 
  labs(x = paste0("PC2 (", round(100 * var_exp[2], digits = 1),
                  "%)"),
       y = paste0("PC3 (", 
                  round(100 * var_exp[3], digits = 1),
                  "%)"),
       colour = "Lancet Condition Category"
  )  + 
  theme_bw()

```

```{r}

# if PC2 > 0, PC3 > 

pca_plot_table |>
# mutate(lancet_cat = case_when(
#   grepl("^Diabetes", lancet_condition) ~ "Diabetes",
# )
mutate(multiple = ifelse(lancet_condition_group == "Multiple",
                         "Multiple",
                         "Single Category")) |>
    ggplot(aes(x = PC2, 
               y = PC3, 
               color = lancet_condition)) +
    geom_point() +
    scale_color_manual(values = viridisLite::plasma(n = 19)) +
  labs(x = paste0("PC2 (", round(100 * var_exp[2], digits = 1),
                  "%)"),
       y = paste0("PC3 (", 
                  round(100 * var_exp[3], digits = 1),
                  "%)"),
       colour = "Lancet Condition Category"
  )  + 
    facet_wrap(~multiple) + 
    theme_bw()
  
```

This analysis suggests YEAR and lancet_condition are important axes of variation in text, and therefore important axes to sample across to create a representative training sample. 

## Methods embeddings

```{r}

# get medcpt article embeddings for all disease studies
methods_embeddings <- data.table::fread(here::here("output/clustering/methods_medcpt_embeddings.csv"))

PCA_test <- prcomp(methods_embeddings, 
                   center = TRUE, 
                   scale = TRUE) 

pca_table <-
data.frame(PC1 = PCA_test$x[,1], 
           PC2 = PCA_test$x[,2], 
           PC3 = PCA_test$x[,3], 
           PC4 = PCA_test$x[,4], 
           PC5 = PCA_test$x[,5], 
           PC6 = PCA_test$x[,6], 
           PC7 = PCA_test$x[,7], 
           PC8 = PCA_test$x[,8], 
           PC9 = PCA_test$x[,9], 
           PC10 = PCA_test$x[,10],
           PUBMED_ID = methods_embeddings$PUBMED_ID) 


var_exp <- PCA_test$sdev^2 / sum(PCA_test$sdev^2)

data.frame(PC = paste0("PC", 1:length(var_exp)),
           PC_n = 1:length(var_exp),
           Variance_Explained = var_exp) |>
  filter(PC_n <= 10) |>
  mutate(PC_n = factor(PC_n, levels = 1:10)) |>
  ggplot(aes(x = PC_n, y = Variance_Explained)) +
  geom_col() +
  labs(x = "Principal Component", 
       y = "Proportion of Variance Explained", 
       title = "Variance explained by each principal component") + 
  theme_bw()

```

### Distribution of GWAS Catalog Text Embeddings By Year (Methods)

```{r plot_methods_pca_year}

pca_plot_table <- 
  pca_table |>
  left_join(gwas_study_info_grouped |> 
  select(PUBMED_ID, 
         YEAR,
         lancet_condition,
         lancet_condition_group), 
  by = "PUBMED_ID") |>
  mutate(YEAR = as.factor(YEAR)) 

n_years <- pca_plot_table$YEAR |> unique() |> length()

pca_plot_table |>
  ggplot(aes(
  x = PC1, 
  y = PC2, 
  color = YEAR)) +
  geom_point() +
  scale_color_manual(values = viridisLite::plasma(n = n_years)) +
  labs(x = paste0("PC1 (", round(100 * var_exp[1], digits = 1),
                  "%)"),
       y = paste0("PC2 (", 
                  round(100 * var_exp[2], digits = 1),
                  "%)"),
       colour = "Publication Year"
  ) +
       #title = "PCA of GWAS Catalog abstract embeddings, colored by publication year") + 
  theme_bw()

pca_plot_table |>
    ggplot(aes(x = PC2, 
               y = PC3, 
               color = YEAR)) +
    geom_point() +
    scale_color_manual(values = viridisLite::plasma(n = n_years)) +
    labs(x = paste0("PC2 (", round(100 * var_exp[2], digits = 1),
                  "%)"),
       y = paste0("PC3 (", 
                  round(100 * var_exp[3], digits = 1),
                  "%)"),
       colour = "Publication Year"
  ) + 
  theme_bw()
  



```

### Distribution of GWAS Catalog Text Embeddings By Lancet Condition Category (Methods)

```{r plot_methods_pca_lancent_condition_category}

  pca_plot_table |>
  ggplot(aes(
  x = PC1, 
  y = PC2, 
  color = lancet_condition_group)) +
  geom_point(size = 0.8) +
  scale_color_brewer(palette = "Dark2") + 
  labs(x = paste0("PC1 (", round(100 * var_exp[1], digits = 1),
                  "%)"),
       y = paste0("PC2 (", 
                  round(100 * var_exp[2], digits = 1),
                  "%)"),
       colour = "Lancet Condition Category"
  )  + 
  theme_bw()


  pca_plot_table |>
  ggplot(aes(
  x = PC2, 
  y = PC3, 
  color = lancet_condition_group)) +
  geom_point(size = 0.8) +
  scale_color_brewer(palette = "Set2") + 
  labs(x = paste0("PC2 (", round(100 * var_exp[2], digits = 1),
                  "%)"),
       y = paste0("PC3 (", 
                  round(100 * var_exp[3], digits = 1),
                  "%)"),
       colour = "Lancet Condition Category"
  )  + 
  theme_bw()

```

```{r}

papers_with_multiple_cond <- pca_plot_table |>
  filter(lancet_condition_group == "Multiple") |> 
  pull(PUBMED_ID)

pca_plot_table_ungrouped <- 
  pca_table |>
  left_join(gwas_study_info,
            by = "PUBMED_ID") |>
  mutate(YEAR = as.factor(YEAR)) 

pca_plot_table_ungrouped <-
  pca_plot_table_ungrouped |>
  mutate(N_conditions = ifelse(PUBMED_ID %in% papers_with_multiple_cond, 
                                         "Multiple",
                                         "Single")) 


pca_plot_table_ungrouped |>
    ggplot(aes(
  x = PC2, 
  y = PC3, 
  color = N_conditions)) +
  geom_point(size = 0.8) +
  scale_color_brewer(palette = "Set2") + 
  labs(x = paste0("PC2 (", round(100 * var_exp[2], digits = 1),
                  "%)"),
       y = paste0("PC3 (", 
                  round(100 * var_exp[3], digits = 1),
                  "%)"),
       colour = "Number of Lancent Categories"
  )  + 
  facet_wrap(~lancet_condition) + 
  theme_bw()


```




# Create the training set

```{r get_training_sample}

set.seed(500)

training_sample = gwas_study_info |>
  select(YEAR, 
         PUBMED_ID,
         lancet_condition) |>
  # tidyr::separate_longer_delim(lancet_condition, 
  #                              delim = " & ") |>
  distinct() |>
  dplyr::group_by(YEAR,lancet_condition) |>
  dplyr::slice_sample(prop = 0.25) |>
  pull(PUBMED_ID) |>
  unique()

# writeLines(
#   as.character(training_sample),
#   here::here("output/gwas_cat/training_sample_pmids.txt"))
# 
# training_sample <- readLines(here::here("output/gwas_cat/training_sample_pmids.txt"))

print("Number of selected training pubmed ids for disease studies:")
training_sample |> 
  unique() |> 
  length()

```

# Create the validation set

```{r}

# now let's select 44 papers for the validation set, 
# this is ~ 20% of the training set
# also stratified by 

set.seed(892)

validation_set <- sample(as.character(training_sample), 
                         size = 44)

writeLines(
  validation_set,
  here::here("output/gwas_cat/validation_set_pmids.txt"))

```
