---
title: "GWAS Trait Categorisation"
author: "Isobel Beasley"
date: "2025-08-24"
output:
  workflowr::wflow_html:
    toc: true
    toc_float: true
    number_sections: true
editor_options: 
  markdown: 
    wrap: 72
---

# Set up

```{r setup, message=FALSE, warning=FALSE}

knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE
                      )

library(data.table)
library(dplyr)
library(ggplot2)
library(stringr)

```

## Get data.frame of GWAS traits 

```{r load_gwas_study_info}

gwas_study_info <- fread(here::here("data/gwas_catalog/gwas-catalog-v1.0.3.1-studies-r2025-07-21.tsv"))

gwas_study_info = 
  gwas_study_info |>
  rename_all(~gsub(" ", "_", .x))

gwas_study_info <-
  gwas_study_info |>
  mutate(MAPPED_TRAIT = tolower(MAPPED_TRAIT),
         MAPPED_BACKGROUND_TRAIT = tolower(MAPPED_BACKGROUND_TRAIT)
         )

gwas_study_info  <-
  gwas_study_info |>
  mutate(YEAR = lubridate::year(DATE))

```

## Add mapped ontology terms for studies with unmapped traits: 

```{r deal_with_unmapped_traits}

# some traits are not mapped:
print("Before fixing, how many unmapped traits are there?")
gwas_study_info |>
  filter(is.na(MAPPED_TRAIT) | MAPPED_TRAIT == "") |>
  nrow()

# by Zoom, Anomalous atrioventricular excitation
# -> Anomalous atrioventricular excitation (disorder)
# http://snomed.info/id/17869006
unmapped_traits <-
  data.frame("DISEASE/TRAIT" = "Anomalous atrioventricular excitation (PheCode 426.4)",
             MAPPED_TRAIT = "anomalous atrioventricular excitation (disorder)",
             MAPPED_TRAIT_URI = "http://snomed.info/id/17869006",
             stringsAsFactors = FALSE
             ) |>
  rename(`DISEASE/TRAIT` = "DISEASE.TRAIT")

# by Zooma,  Pilocytic astrocytoma -> MONDO_0016691 (http://purl.obolibrary.org/obo/MONDO_0016691)
unmapped_traits <-
  unmapped_traits |>
  add_row(`DISEASE/TRAIT` = "Pilocytic astrocytoma",
             MAPPED_TRAIT = "pilocytic astrocytoma",
             MAPPED_TRAIT_URI = "http://purl.obolibrary.org/obo/MONDO_0016691"
             )

# by Zooma, Pilocytic astrocytoma and optic pathway glioma
# -> http://purl.obolibrary.org/obo/MONDO_0016167, http://purl.obolibrary.org/obo/MONDO_0016691
unmapped_traits =
  unmapped_traits |>
  add_row(`DISEASE/TRAIT` = "Pilocytic astrocytoma and optic pathway glioma",
          MAPPED_TRAIT = "optic pathway glioma, pilocytic astrocytoma",
          MAPPED_TRAIT_URI = "http://purl.obolibrary.org/obo/MONDO_0016167, http://purl.obolibrary.org/obo/MONDO_0016691"
          )

# by searching ontology lookup service:
# Leukotriene levels (480.2454_0.351) & Leukotriene levels (337.1632_0.339)
# -> Fatty Acid Measurement
# http://purl.obolibrary.org/obo/NCIT_C80157
unmapped_traits =
  unmapped_traits |>
  add_row(`DISEASE/TRAIT` = c("Leukotriene levels (480.2454_0.351)",
                               "Leukotriene levels (337.1632_0.339)"
                               ),
          MAPPED_TRAIT = c("fatty acid measurement",
                           "fatty acid measurement"
                           ),
          MAPPED_TRAIT_URI = c("http://purl.obolibrary.org/obo/NCIT_C80157",
                               "http://purl.obolibrary.org/obo/NCIT_C80157"
                               )
          )

# by searching ontology lookup service:
# X-11244 levels
# X-11255 levels 
# to be mapped to: http://www.ebi.ac.uk/efo/EFO_0004725
# metabolite measurement
unmapped_traits =
  unmapped_traits |>
  add_row(`DISEASE/TRAIT` = c("X-11244 levels",
                               "X-11255 levels"
                               ),
          MAPPED_TRAIT = c("metabolite measurement",
                           "metabolite measurement"
                           ),
          MAPPED_TRAIT_URI = c("http://www.ebi.ac.uk/efo/EFO_0004725",
                               "http://www.ebi.ac.uk/efo/EFO_0004725"
                               )
          )

# by searching ontology lookup service:
# N-acetylornithine levels, & N-acetylornithine levels in chronic kidney disease
# to be mapped to http://www.ebi.ac.uk/efo/EFO_0021538
# N-acetylornithine measurement
unmapped_traits =
  unmapped_traits |>
  add_row(`DISEASE/TRAIT` = c("N-acetylornithine levels",
                               "N-acetylornithine levels in chronic kidney disease"
                               ),
          MAPPED_TRAIT = c("n-acetylornithine measurement",
                           "n-acetylornithine measurement"
                           ),
          MAPPED_TRAIT_URI = c("http://www.ebi.ac.uk/efo/EFO_0021538",
                               "http://www.ebi.ac.uk/efo/EFO_0021538"
                               )
          )

# by searching ontology lookup service:
# Scleritis and episcleritis (PheCode 379.1)
# map to: Scleritis and episcleritis (disorder)
# http://snomed.info/id/267659002
unmapped_traits =
  unmapped_traits |>
  add_row(`DISEASE/TRAIT` = "Scleritis and episcleritis (PheCode 379.1)",
          MAPPED_TRAIT = "scleritis and episcleritis (disorder)",
          MAPPED_TRAIT_URI = "http://snomed.info/id/267659002"
          )


# Add the unmapped traits to the gwas study info df, by matching on DISEASE/TRAIT
# and making unmmaped traits are NA
gwas_study_info =
  gwas_study_info |>
  mutate(MAPPED_TRAIT = ifelse(MAPPED_TRAIT == "",
                               NA,
                               MAPPED_TRAIT))

gwas_study_info =
  gwas_study_info |>
  rows_patch(unmapped_traits,
             by = c("DISEASE/TRAIT"),
             unmatched = "ignore"
             )

# yay all mapped now
print("After fixing, how many unmapped traits remain?")
gwas_study_info |>
  filter(is.na(MAPPED_TRAIT) | MAPPED_TRAIT == "") |>
  nrow()

```

## Add DISEASE/TRAIT description for studies where this is missing

```{r deal_with_missing_disease_trait}

print("Before fixing, how many studies have missing DISEASE/TRAIT?")
gwas_study_info |> 
  filter(`DISEASE/TRAIT` == ""| is.na(`DISEASE/TRAIT`)) |>
  nrow()

# pubmed id: 35240980
gwas_study_info |>
  filter(PUBMED_ID == 35240980) |>
  select(`DISEASE/TRAIT`, 
         MAPPED_TRAIT, 
         MAPPED_BACKGROUND_TRAIT) 


# DISEASE/TRAIT == "Cognitive impairment among children born extremely preterm"
gwas_study_info =
  gwas_study_info |>
  mutate(`DISEASE/TRAIT` = ifelse(PUBMED_ID == 35240980,
                                "Cognitive impairment among children born extremely preterm",
                                `DISEASE/TRAIT`)
         )

# study accession: GCST90624363
gwas_study_info |> 
  filter(STUDY_ACCESSION == "GCST90624363") |>
  select(`DISEASE/TRAIT`, 
         MAPPED_TRAIT, 
         MAPPED_BACKGROUND_TRAIT)

# DISEASE/TRAIT == "Lyme borreliosis"
gwas_study_info =
  gwas_study_info |>
  mutate(`DISEASE/TRAIT` = ifelse(STUDY_ACCESSION == "GCST90624363",
                                "Lyme borreliosis",
                                `DISEASE/TRAIT`)
         )

# pubmed id: 38509478
# Nausea and vomiting during pregnancy
gwas_study_info |>
  filter(PUBMED_ID == 38509478) |>
  select(`DISEASE/TRAIT`, 
         MAPPED_TRAIT, 
         MAPPED_BACKGROUND_TRAIT)


gwas_study_info =
  gwas_study_info |>
  mutate(`DISEASE/TRAIT` = ifelse(PUBMED_ID == 38509478,
                                "Nausea and vomiting during pregnancy",
                                `DISEASE/TRAIT`)
         )


print("After fixing, how many studies have missing DISEASE/TRAIT?")
gwas_study_info |> 
  filter(`DISEASE/TRAIT` == ""| is.na(`DISEASE/TRAIT`)) |>
  nrow()

```

## Correcting some MAPPED_TRAIT (MAPPED to ontology terms by GWAS Catalog) 

### Re-map some traits based on DISEASE/TRAIT

#### Non-specific disease mapping

```{r fix_bad_mapping_non_specific}

# STUDY ACCESSION: GCST90043814
# MAPPED_TRAIT == "disease" 
# replace with otitis media
# EFO_0004992
  
gwas_study_info =
  gwas_study_info |>
  mutate(MAPPED_TRAIT = ifelse(STUDY_ACCESSION == "GCST90043814" & 
                                MAPPED_TRAIT == "disease",
                               "otitis media",
                               MAPPED_TRAIT)
         ) |>
    mutate(MAPPED_TRAIT_URI = ifelse(STUDY_ACCESSION == "GCST90043814" & 
                                      MAPPED_TRAIT_URI == "http://www.ebi.ac.uk/efo/EFO_0000408",
                                     "http://www.ebi.ac.uk/efo/EFO_0004992",
                                     MAPPED_TRAIT_URI)
           )

# GCST90244761
# remove MAPPED_BACKGROUND_TRAIT and MAPPED_BACKGROUND_TRAIT_URI
# (which were just 'disease')
gwas_study_info =
  gwas_study_info |>
  mutate(MAPPED_BACKGROUND_TRAIT = ifelse(STUDY_ACCESSION == "GCST90244761",
                                         "",
                                         MAPPED_BACKGROUND_TRAIT)
         ) |>
    mutate(MAPPED_BACKGROUND_TRAIT_URI = ifelse(STUDY_ACCESSION == "GCST90244761",
                                                "",
                                                MAPPED_BACKGROUND_TRAIT_URI)
           )

```

#### Pregnancy and maternal health related traits

```{r fix_bad_mapping_pregnancy}

# Pubmed id: 33239672
# Multiple consecutive miscarriage
## MAPPED_TRAIT == "fertility trait"
## MAPPED_TRAIT_URI == "http://purl.obolibrary.org/obo/OBA_VT0010464"

# What to convert to:
# make MAPPED_TRAIT -> "spontaneous abortion"
# MAPPED_TRAIT_URI -> "http://www.ebi.ac.uk/efo/EFO_1001255"

gwas_study_info =
  gwas_study_info |>
  mutate(MAPPED_TRAIT = ifelse(grepl("Multiple consecutive miscarriage", `DISEASE/TRAIT`) & 
                                MAPPED_TRAIT == "fertility trait",
                               "spontaneous abortion",
                               MAPPED_TRAIT)
         ) |>
    mutate(MAPPED_TRAIT_URI = ifelse(grepl("Multiple consecutive miscarriage", `DISEASE/TRAIT`) & 
                                      MAPPED_TRAIT_URI == "http://purl.obolibrary.org/obo/OBA_VT0010464",
                                     "http://www.ebi.ac.uk/efo/EFO_1001255",
                                     MAPPED_TRAIT_URI)
           )

# for PUBMED_ID 29018042
## DISEASE/TRAIT == "Post-term birth"
## MAPPED_TRAIT == "gestational age"
## MAPPED_TRAIT_URI == "http://www.ebi.ac.uk/efo/EFO_0005112"

# What to convert to:
# make MAPPED_TRAIT to "post term pregnancy"
# MAPPED_TRAIT_URI to "http://www.ebi.ac.uk/efo/EFO_0009681"

gwas_study_info =
  gwas_study_info |>
  mutate(MAPPED_TRAIT = ifelse(grepl("Post-term birth", `DISEASE/TRAIT`) & 
                                MAPPED_TRAIT == "gestational age",
                               "post term pregnancy",
                               MAPPED_TRAIT)
         ) |>
    mutate(MAPPED_TRAIT_URI = ifelse(grepl("Post-term birth", `DISEASE/TRAIT`) & 
                                      MAPPED_TRAIT_URI == "http://www.ebi.ac.uk/efo/EFO_0005112",
                                     "http://www.ebi.ac.uk/efo/EFO_0009681",
                                     MAPPED_TRAIT_URI)
           )

# DISEASE/TRAIT == "Spontaneous preterm birth (preterm delivery)"
# MAPPED_TRAIT == "delivery measurement"
# MAPPED_TRAIT_URI == http://www.ebi.ac.uk/efo/EFO_0006922

# make MAPPED_TRAIT to "premature birth"
# MAPPED_TRAIT_URI to http://www.ebi.ac.uk/efo/EFO_0003917
gwas_study_info =
  gwas_study_info |>
  mutate(MAPPED_TRAIT = ifelse(grepl("Spontaneous preterm birth \\(preterm delivery\\)", 
                                     `DISEASE/TRAIT`) & 
                                MAPPED_TRAIT == "delivery measurement",
                               "premature birth",
                               MAPPED_TRAIT)
         ) |>
    mutate(MAPPED_TRAIT_URI = ifelse(grepl("Spontaneous preterm birth \\(preterm delivery\\)", `DISEASE/TRAIT`) & 
                                      MAPPED_TRAIT_URI == "http://www.ebi.ac.uk/efo/EFO_0006922",
                                     "http://www.ebi.ac.uk/efo/EFO_0003917",
                                     MAPPED_TRAIT_URI)
           )

## DISEASE/TRAIT == "Sporadic miscarriage"
## MAPPED_TRAIT == "fertility trait"
## MAPPED_TRAIT_URI == "http://purl.obolibrary.org/obo/OBA_VT0010464"

# make MAPPED_TRAIT to "spontaneous abortion"
# MAPPED_TRAIT_URI to "http://www.ebi.ac.uk/efo/EFO_1001255"

gwas_study_info =
  gwas_study_info |>
  mutate(MAPPED_TRAIT = ifelse(grepl("Sporadic miscarriage", `DISEASE/TRAIT`) & 
                                MAPPED_TRAIT == "fertility trait",
                               "spontaneous abortion",
                               MAPPED_TRAIT)
         ) |>
    mutate(MAPPED_TRAIT_URI = ifelse(grepl("Sporadic miscarriage", `DISEASE/TRAIT`) & 
                                      MAPPED_TRAIT_URI == "http://purl.obolibrary.org/obo/OBA_VT0010464",
                                     "http://www.ebi.ac.uk/efo/EFO_1001255",
                                     MAPPED_TRAIT_URI)
           )
 
```

#### TB 

```{r fix_bad_mapping_tb}

# PUBMED_ID: 33661925
# `DISEASE/TRAIT contains resistance to mycobacterium tuberculosis infection
# replace MAPPED_TRAIT == "decreased susceptibility to bacterial infection"
# with "tuberculosis"
# MAPPED_TRAIT_URI from "http://www.ebi.ac.uk/efo/EFO_0008322" 
# to "http://purl.obolibrary.org/obo/MONDO_0018076"
gwas_study_info =
  gwas_study_info |>
  mutate(MAPPED_TRAIT = ifelse(PUBMED_ID == 33661925 & 
                                MAPPED_TRAIT == "decreased susceptibility to bacterial infection",
                               "tuberculosis",
                               MAPPED_TRAIT)
         ) |>
    mutate(MAPPED_TRAIT_URI = ifelse(PUBMED_ID == 33661925 & 
                                      MAPPED_TRAIT_URI == "http://www.ebi.ac.uk/efo/EFO_0008322",
                                     "http://purl.obolibrary.org/obo/MONDO_0018076",
                                     MAPPED_TRAIT_URI)
           )


# for PUBMED = 28628665
# MAPPED_TRAIT = "decreased susceptibility to bacterial infection"
# MAPPED_TRAIT_URI = "http://www.ebi.ac.uk/efo/EFO_0008322"

# replace MAPPED_TRAIT to "tuberculosis"
# MAPPED_TRAIT_URI to "http://purl.obolibrary.org/obo/MONDO_0018076"
gwas_study_info =
  gwas_study_info |>
  mutate(MAPPED_TRAIT = ifelse(PUBMED_ID == 28628665 & 
                                MAPPED_TRAIT == "tuberculin skin test reactivity measurement, decreased susceptibility to bacterial infection",
                               "tuberculosis",
                               MAPPED_TRAIT)
         ) |>
    mutate(MAPPED_TRAIT_URI = ifelse(PUBMED_ID == 28628665 & 
                                      MAPPED_TRAIT_URI == "http://www.ebi.ac.uk/efo/EFO_0008307, http://www.ebi.ac.uk/efo/EFO_0008322",
                                     "http://purl.obolibrary.org/obo/MONDO_0018076",
                                     MAPPED_TRAIT_URI)
           )


```

#### Other bad mappings

```{r fix_other_bad_mapping}

## pubmed id: 34737426
# mapped_trait = benign neoplasm
# but should be cervical carcinoma, 	EFO_0001061

gwas_study_info =
  gwas_study_info |>
  mutate(MAPPED_TRAIT = ifelse(PUBMED_ID == 34737426 & 
                                MAPPED_TRAIT == "benign neoplasm",
                               "cervical carcinoma",
                               MAPPED_TRAIT)
         ) |>
    mutate(MAPPED_TRAIT_URI = ifelse(PUBMED_ID == 34737426 & 
                                      MAPPED_TRAIT_URI == "http://purl.obolibrary.org/obo/MONDO_0021230",
                                     "http://www.ebi.ac.uk/efo/EFO_0001061",
                                     MAPPED_TRAIT_URI)
           )

# Capecitabine-induced hand-foot syndrome in breast or colorectal cancer
  gwas_study_info =
      gwas_study_info  |>
  mutate(
    MAPPED_TRAIT = 
      ifelse(grepl("Capecitabine-induced hand-foot syndrome in breast or colorectal cancer", `DISEASE/TRAIT`),
         "breast cancer, colorectal cancer, hand-foot syndrome",
         MAPPED_TRAIT
         ),
    MAPPED_TRAIT_URI = 
      ifelse(grepl("Capecitabine-induced hand-foot syndrome in breast or colorectal cancer", `DISEASE/TRAIT`),
         "http://purl.obolibrary.org/obo/MONDO_0007254, http://purl.obolibrary.org/obo/MONDO_0005575, http://purl.obolibrary.org/obo/MONDO_0700048",
         MAPPED_TRAIT_URI
         )
  )



# GCST001789
# replace MAPPED_TRAIT from "bronchopulmonary dysplasia"
# to "blood high density lipoprotein particle diameter"
# and MAPPED_TRAIT_URI to "http://purl.obolibrary.org/obo/CMO_0002692"
gwas_study_info =
  gwas_study_info |>
  mutate(MAPPED_TRAIT = ifelse(STUDY_ACCESSION == "GCST001789" & 
                                MAPPED_TRAIT == "bronchopulmonary dysplasia",
                               "blood high density lipoprotein particle diameter",
                               MAPPED_TRAIT)
         ) |>
    mutate(MAPPED_TRAIT_URI = ifelse(STUDY_ACCESSION == "GCST001789" & 
                                      MAPPED_TRAIT_URI == "http://purl.obolibrary.org/obo/MONDO_0019091",
                                     "http://purl.obolibrary.org/obo/CMO_0002692",
                                     MAPPED_TRAIT_URI)
           )
  

# Periodontal disease related phenotype
# set trait to periodontal disorder
# rather than periodontal measurement
gwas_study_info =
  gwas_study_info |>
  mutate(MAPPED_TRAIT = ifelse(grepl("Periodontal disease related phenotype", `DISEASE/TRAIT`),
                               str_replace_all(pattern = "periodontal measurement",
                                               replacement = "periodontal disorder",
                                               MAPPED_TRAIT
                                               ),
                               MAPPED_TRAIT)
         ) |>
    mutate(MAPPED_TRAIT_URI = ifelse(grepl("Periodontal disease related phenotype", `DISEASE/TRAIT`),
                               str_replace_all(pattern = "http://www.ebi.ac.uk/efo/EFO_0007780",
                                               replacement = "http://purl.obolibrary.org/obo/MONDO_0002635",
                                               MAPPED_TRAIT_URI
                                               ),
                               MAPPED_TRAIT_URI)
    )

```

### Add and correct MAPPED_BACKGROUND_TRAIT terms

```{r add_missing_background}

# for MAPPED_TRAIT contains "sars-cov-2"
# set background trait to covid-19
gwas_study_info =
  gwas_study_info |>
  mutate(MAPPED_BACKGROUND_TRAIT = ifelse(grepl("sars-cov-2", MAPPED_TRAIT),
                                         "covid-19",
                                         MAPPED_BACKGROUND_TRAIT)
         ) |>
    mutate(MAPPED_BACKGROUND_TRAIT_URI = ifelse(grepl("sars-cov-2", MAPPED_TRAIT),
                                                "http://purl.obolibrary.org/obo/MONDO_0100096",
                                                MAPPED_BACKGROUND_TRAIT_URI)
           )

# for pubmed id: 32247823
# set background trait to non-alcoholic steatohepatitis
# and background trait uri to EFO_1001249
gwas_study_info =
  gwas_study_info |>
  mutate(MAPPED_BACKGROUND_TRAIT = ifelse(PUBMED_ID == 32247823,
                                         "non-alcoholic steatohepatitis",
                                         MAPPED_BACKGROUND_TRAIT)
         ) |>
    mutate(MAPPED_BACKGROUND_TRAIT_URI = ifelse(PUBMED_ID == 32247823,
                                                "http://www.ebi.ac.uk/efo/EFO_1001249",
                                                MAPPED_BACKGROUND_TRAIT_URI)
           )

# Exploratory eye movement dysfunction in schizophrenia
# set background trait to schizophrenia
gwas_study_info =
  gwas_study_info |>
  mutate(MAPPED_BACKGROUND_TRAIT = ifelse(grepl("Exploratory eye movement dysfunction in schizophrenia", `DISEASE/TRAIT`),
                                         "schizophrenia",
                                         MAPPED_BACKGROUND_TRAIT)
         ) |>
    mutate(MAPPED_BACKGROUND_TRAIT_URI = ifelse(grepl("Exploratory eye movement dysfunction in schizophrenia", `DISEASE/TRAIT`),
                                                "http://purl.obolibrary.org/obo/MONDO_0005090",
                                                MAPPED_BACKGROUND_TRAIT_URI)
           )

# for pubmed_id: 21107309
# set background trait to schizophrenia
gwas_study_info =
  gwas_study_info |>
  mutate(MAPPED_BACKGROUND_TRAIT = ifelse(PUBMED_ID == 21107309,
                                         "schizophrenia",
                                         MAPPED_BACKGROUND_TRAIT)
         ) |>
    mutate(MAPPED_BACKGROUND_TRAIT_URI = ifelse(PUBMED_ID == 21107309,
                                                "http://purl.obolibrary.org/obo/MONDO_0005090",
                                                MAPPED_BACKGROUND_TRAIT_URI)
           )

# if `DISEASE/TRAIT` contains Adverse response to chemotherapy in breast cancer
# set MAPPED_BACKGROUND_TRAIT to breast cancer
# http://purl.obolibrary.org/obo/MONDO_0007254
gwas_study_info =
  gwas_study_info |>
  mutate(MAPPED_BACKGROUND_TRAIT = ifelse(grepl("Adverse response to chemotherapy in breast cancer", `DISEASE/TRAIT`),
                                         "breast cancer",
                                         MAPPED_BACKGROUND_TRAIT)
         ) |>
    mutate(MAPPED_BACKGROUND_TRAIT_URI = ifelse(grepl("Adverse response to chemotherapy in breast cancer", `DISEASE/TRAIT`),
                                                "http://purl.obolibrary.org/obo/MONDO_0007254",
                                                MAPPED_BACKGROUND_TRAIT_URI)
           )

# for pubmed id: 30188897
# and DISEASE/TRAIT contains "miscarriages"
# add spontaneous abortion, http://www.ebi.ac.uk/efo/EFO_1001255
# as MAPPED_BACKGROUND_TRAIT
gwas_study_info =
  gwas_study_info |>
  mutate(MAPPED_BACKGROUND_TRAIT = ifelse(PUBMED_ID == 30188897 & 
                                           grepl("miscarriage", `DISEASE/TRAIT`),
                                         "spontaneous abortion",
                                         MAPPED_BACKGROUND_TRAIT)
         ) |>
    mutate(MAPPED_BACKGROUND_TRAIT_URI = ifelse(PUBMED_ID == 30188897 & 
                                                  grepl("miscarriage", `DISEASE/TRAIT`),
                                                "http://www.ebi.ac.uk/efo/EFO_1001255",
                                                MAPPED_BACKGROUND_TRAIT_URI)
           )

# for pubmed id: 30188897
# and DISEASE/TRAIT contains "stillbirth"
# add stillbirth, http://purl.obolibrary.org/obo/NCIT_C49151
# as MAPPED_BACKGROUND_TRAIT
gwas_study_info =
  gwas_study_info |>
  mutate(MAPPED_BACKGROUND_TRAIT = ifelse(PUBMED_ID == 30188897 & 
                                           grepl("stillbirth", `DISEASE/TRAIT`),
                                         "stillbirth",
                                         MAPPED_BACKGROUND_TRAIT)
         ) |>
    mutate(MAPPED_BACKGROUND_TRAIT_URI = ifelse(PUBMED_ID == 30188897 & 
                                                  grepl("stillbirth", `DISEASE/TRAIT`),
                                                "http://purl.obolibrary.org/obo/NCIT_C49151",
                                                MAPPED_BACKGROUND_TRAIT_URI)
           )

# if MAPPED_TRAIT == "remission"
# and "Hepatitis C" is in DISEASE/TRAIT, then
# set MAPPED_BACKGROUND_TRAIT to hepatitis C virus infection
# and MAPPED_BACKGROUND_TRAIT_URI to: http://purl.obolibrary.org/obo/MONDO_0005231
gwas_study_info =
  gwas_study_info |>
  mutate(MAPPED_BACKGROUND_TRAIT = ifelse(MAPPED_TRAIT == "remission" &
                                          grepl("Hepatitis C", `DISEASE/TRAIT`, ignore.case = T),
                                         "hepatitis C virus infection",
                                         MAPPED_BACKGROUND_TRAIT)
         ) |>
    mutate(MAPPED_BACKGROUND_TRAIT_URI = ifelse(MAPPED_BACKGROUND_TRAIT == "hepatitis C virus infection",
                                                "http://purl.obolibrary.org/obo/MONDO_0005231",
                                                MAPPED_BACKGROUND_TRAIT_URI)
           )

# if MAPPED_TRAIT == "progression free survival"
# and MAPPED_BACKGROUND_TRAIT == ""
# and DISEASE/TRAIT contains epithelial ovarian cancer
gwas_study_info =
  gwas_study_info |>
  mutate(MAPPED_BACKGROUND_TRAIT = ifelse(MAPPED_TRAIT ==  "progression free survival" &
                                          MAPPED_BACKGROUND_TRAIT == "" &
                                          grepl("epithelial ovarian cancer", `DISEASE/TRAIT`, ignore.case = T),
                                         "ovarian cancer",
                                         MAPPED_BACKGROUND_TRAIT)
         ) |>
    mutate(MAPPED_BACKGROUND_TRAIT_URI = ifelse(MAPPED_BACKGROUND_TRAIT == "ovarian cancer" & 
                                                MAPPED_BACKGROUND_TRAIT_URI == "",
                                                "http://purl.obolibrary.org/obo/MONDO_0008170",
                                                MAPPED_BACKGROUND_TRAIT_URI)
           )

# if MAPPED_TRAIT == "progression free survival"
# and MAPPED_BACKGROUND_TRAIT == ""
# and DISEASE/TRAIT contains colorectal cancer 
gwas_study_info =
  gwas_study_info |>
  mutate(MAPPED_BACKGROUND_TRAIT = ifelse(MAPPED_TRAIT ==  "progression free survival" &
                                          MAPPED_BACKGROUND_TRAIT == "" &
                                          grepl("colorectal cancer|colon cancer", `DISEASE/TRAIT`, ignore.case = T),
                                         "colorectal cancer",
                                         MAPPED_BACKGROUND_TRAIT)
         ) |>
    mutate(MAPPED_BACKGROUND_TRAIT_URI = ifelse(MAPPED_BACKGROUND_TRAIT == "colorectal cancer" & 
                                                MAPPED_BACKGROUND_TRAIT_URI == "",
                                                "http://purl.obolibrary.org/obo/MONDO_0005575",
                                                MAPPED_BACKGROUND_TRAIT_URI)
           )

# if MAPPED_TRAIT == "progression free survival"
# and MAPPED_BACKGROUND_TRAIT == ""
# and DISEASE/TRAIT contains leukemia
gwas_study_info =
  gwas_study_info |>
  mutate(MAPPED_BACKGROUND_TRAIT = ifelse(MAPPED_TRAIT ==  "progression free survival" &
                                          MAPPED_BACKGROUND_TRAIT == "" &
                                          grepl("leukemia", `DISEASE/TRAIT`, ignore.case = T),
                                         "leukemia",
                                         MAPPED_BACKGROUND_TRAIT)
         ) |>
    mutate(MAPPED_BACKGROUND_TRAIT_URI = ifelse(MAPPED_BACKGROUND_TRAIT == "leukemia" & 
                                                MAPPED_BACKGROUND_TRAIT_URI == "",
                                                "http://www.ebi.ac.uk/efo/EFO_0000565",
                                                MAPPED_BACKGROUND_TRAIT_URI)
           )

# if MAPPED_TRAIT == "progression free survival"
# and MAPPED_BACKGROUND_TRAIT == ""
# and DISEASE/TRAIT contains glioma
gwas_study_info =
  gwas_study_info |>
  mutate(MAPPED_BACKGROUND_TRAIT = ifelse(MAPPED_TRAIT ==  "progression free survival" &
                                          MAPPED_BACKGROUND_TRAIT == "" &
                                          grepl("glioma", `DISEASE/TRAIT`, ignore.case = T),
                                         "glioma",
                                         MAPPED_BACKGROUND_TRAIT)
         ) |>
    mutate(MAPPED_BACKGROUND_TRAIT_URI = ifelse(MAPPED_BACKGROUND_TRAIT == "glioma" & 
                                                MAPPED_BACKGROUND_TRAIT_URI == "",
                                                "http://www.ebi.ac.uk/efo/EFO_0005543",
                                                MAPPED_BACKGROUND_TRAIT_URI)
           )

# if MAPPED_TRAIT == "progression free survival"
# and MAPPED_BACKGROUND_TRAIT == ""
# and DISEASE/TRAIT contains bladder cancer
gwas_study_info =
  gwas_study_info |>
  mutate(MAPPED_BACKGROUND_TRAIT = ifelse(MAPPED_TRAIT ==  "progression free survival" &
                                          MAPPED_BACKGROUND_TRAIT == "" &
                                          grepl("bladder cancer", `DISEASE/TRAIT`, ignore.case = T),
                                         "urinary bladder carcinoma",
                                         MAPPED_BACKGROUND_TRAIT)
         ) |>
    mutate(MAPPED_BACKGROUND_TRAIT_URI = ifelse(MAPPED_BACKGROUND_TRAIT == "urinary bladder carcinoma" & 
                                                MAPPED_BACKGROUND_TRAIT_URI == "",
                                                "http://purl.obolibrary.org/obo/MONDO_0004986",
                                                MAPPED_BACKGROUND_TRAIT_URI)
           )

# if MAPPED_TRAIT == "progression free survival"
# and MAPPED_BACKGROUND_TRAIT == ""
# and DISEASE/TRAIT contains lung adenocarcinoma
gwas_study_info =
  gwas_study_info |>
  mutate(MAPPED_BACKGROUND_TRAIT = ifelse(MAPPED_TRAIT ==  "progression free survival" &
                                          MAPPED_BACKGROUND_TRAIT == "" &
                                          grepl("lung adenocarcinoma", `DISEASE/TRAIT`, ignore.case = T),
                                         "lung adenocarcinoma",
                                         MAPPED_BACKGROUND_TRAIT)
         ) |>
    mutate(MAPPED_BACKGROUND_TRAIT_URI = ifelse(MAPPED_BACKGROUND_TRAIT == "lung adenocarcinoma" & 
                                                MAPPED_BACKGROUND_TRAIT_URI == "",
                                                "http://www.ebi.ac.uk/efo/EFO_0000571",
                                                MAPPED_BACKGROUND_TRAIT_URI)
           )

# if MAPPED_TRAIT == "survival time"
# and MAPPED_BACKGROUND_TRAIT == ""
# and DISEASE/TRAIT contains prostate cancer
gwas_study_info =
  gwas_study_info |>
  mutate(MAPPED_BACKGROUND_TRAIT = ifelse(MAPPED_TRAIT ==  "survival time" &
                                          MAPPED_BACKGROUND_TRAIT == "" &
                                          grepl("prostate cancer", `DISEASE/TRAIT`, ignore.case = T),
                                         "prostate cancer",
                                         MAPPED_BACKGROUND_TRAIT)
         ) |>
    mutate(MAPPED_BACKGROUND_TRAIT_URI = ifelse(MAPPED_BACKGROUND_TRAIT == "prostate cancer" & 
                                                MAPPED_BACKGROUND_TRAIT_URI == "",
                                                "http://purl.obolibrary.org/obo/DOID_10283",
                                                MAPPED_BACKGROUND_TRAIT_URI)
           )

# if MAPPED_TRAIT == "survival time"
# and MAPPED_BACKGROUND_TRAIT == ""
# and DISEASE/TRAIT contains breast cancer
gwas_study_info =
  gwas_study_info |>
  mutate(MAPPED_BACKGROUND_TRAIT = ifelse(MAPPED_TRAIT ==  "survival time" &
                                          MAPPED_BACKGROUND_TRAIT == "" &
                                          grepl("breast cancer", `DISEASE/TRAIT`, ignore.case = T),
                                         "breast cancer",
                                         MAPPED_BACKGROUND_TRAIT)
         ) |>
    mutate(MAPPED_BACKGROUND_TRAIT_URI = ifelse(MAPPED_BACKGROUND_TRAIT == "breast cancer" & 
                                                MAPPED_BACKGROUND_TRAIT_URI == "",
                                                "http://purl.obolibrary.org/obo/MONDO_0007254",
                                                MAPPED_BACKGROUND_TRAIT_URI)
           )

# if MAPPED_TRAIT == "survival time"
# and MAPPED_BACKGROUND_TRAIT == ""
# and DISEASE/TRAIT contains colorectal cancer
gwas_study_info =
  gwas_study_info |>
  mutate(MAPPED_BACKGROUND_TRAIT = ifelse(MAPPED_TRAIT ==  "survival time" &
                                          MAPPED_BACKGROUND_TRAIT == "" &
                                          grepl("colorectal cancer|colon cancer|rectal cancer", `DISEASE/TRAIT`, ignore.case = T),
                                         "colorectal cancer",
                                         MAPPED_BACKGROUND_TRAIT)
         ) |>
    mutate(MAPPED_BACKGROUND_TRAIT_URI = ifelse(MAPPED_BACKGROUND_TRAIT == "colorectal cancer" & 
                                                MAPPED_BACKGROUND_TRAIT_URI == "",
                                                "http://purl.obolibrary.org/obo/MONDO_0005575",
                                                MAPPED_BACKGROUND_TRAIT_URI)
           )

# if MAPPED_TRAIT == "survival time"
# and MAPPED_BACKGROUND_TRAIT == ""
# and DISEASE/TRAIT contains lung cancer
gwas_study_info =
  gwas_study_info |>
  mutate(MAPPED_BACKGROUND_TRAIT = ifelse(MAPPED_TRAIT ==  "survival time" &
                                          MAPPED_BACKGROUND_TRAIT == "" &
                                          grepl("lung cancer", `DISEASE/TRAIT`, ignore.case = T),
                                         "lung cancer",
                                         MAPPED_BACKGROUND_TRAIT)
         ) |>
    mutate(MAPPED_BACKGROUND_TRAIT_URI = ifelse(MAPPED_BACKGROUND_TRAIT == "lung cancer" & 
                                                MAPPED_BACKGROUND_TRAIT_URI == "",
                                                "http://purl.obolibrary.org/obo/MONDO_0008903",
                                                MAPPED_BACKGROUND_TRAIT_URI)
           )

# if MAPPED_TRAIT == "illness severity status"
# and MAPPED_BACKGROUND_TRAIT == ""
# and DISEASE/TRAIT contains hand, foot, and mouth disease
gwas_study_info =
  gwas_study_info |>
  mutate(MAPPED_BACKGROUND_TRAIT = ifelse(MAPPED_TRAIT ==  "illness severity status" &
                                          MAPPED_BACKGROUND_TRAIT == "" &
                                          grepl("hand, foot, and mouth disease", `DISEASE/TRAIT`, ignore.case = T),
                                         "hand, foot, and mouth disease",
                                         MAPPED_BACKGROUND_TRAIT)
         ) |>
    mutate(MAPPED_BACKGROUND_TRAIT_URI = ifelse(MAPPED_BACKGROUND_TRAIT == "hand, foot, and mouth disease" & 
                                                MAPPED_BACKGROUND_TRAIT_URI == "",
                                                "http://www.ebi.ac.uk/efo/EFO_0007294",
                                                MAPPED_BACKGROUND_TRAIT_URI)
           )

# if MAPPED_TRAIT == "response to covid-19 vaccine"
# set MAPPED_BACKGROUND_TRAIT to covid-19
gwas_study_info =
  gwas_study_info |>
  mutate(MAPPED_BACKGROUND_TRAIT = ifelse(MAPPED_TRAIT == "response to covid-19 vaccine",
                                         "covid-19",
                                         MAPPED_BACKGROUND_TRAIT)
         ) |>
    mutate(MAPPED_BACKGROUND_TRAIT_URI = ifelse(MAPPED_TRAIT == "response to covid-19 vaccine",
                                                "http://purl.obolibrary.org/obo/MONDO_0100096",
                                                MAPPED_BACKGROUND_TRAIT_URI)
           )


	
# if MAPPED_TRAIT == "response to vaccine"
# and DISEASE/TRAIT contains "smallpox" 
# set MAPPED_BACKGROUND_TRAIT to smallpox
gwas_study_info =
  gwas_study_info |>
  mutate(MAPPED_BACKGROUND_TRAIT = ifelse(MAPPED_TRAIT == "response to vaccine" &
                                           grepl("smallpox", `DISEASE/TRAIT`),
                                         "smallpox",
                                         MAPPED_BACKGROUND_TRAIT)
         ) |>
    mutate(MAPPED_BACKGROUND_TRAIT_URI = ifelse(MAPPED_TRAIT == "response to vaccine" &
                                                  grepl("smallpox", `DISEASE/TRAIT`),
                                                "http://purl.obolibrary.org/obo/DOID_8736",
                                                MAPPED_BACKGROUND_TRAIT_URI)
           )

# if MAPPED_TRAIT == "response to vaccine"
# and DISEASE/TRAIT contains "influenza"
# set MAPPED_BACKGROUND_TRAIT to influenza
gwas_study_info =
  gwas_study_info |>
  mutate(MAPPED_BACKGROUND_TRAIT = ifelse(MAPPED_TRAIT == "response to vaccine" &
                                           grepl("influenza", `DISEASE/TRAIT`),
                                         "influenza",
                                         MAPPED_BACKGROUND_TRAIT)
         ) |>
    mutate(MAPPED_BACKGROUND_TRAIT_URI = ifelse(MAPPED_TRAIT == "response to vaccine" &
                                                  grepl("influenza", `DISEASE/TRAIT`),
                                                "http://www.ebi.ac.uk/efo/EFO_0007328",
                                                MAPPED_BACKGROUND_TRAIT_URI)
           )

# if. MAPPED_TRAIT ==  "response to vaccine"
# and DISEASE/TRAIT contains "hepatitis B"
# set MAPPED_BACKGROUND_TRAIT to hepatitis B
gwas_study_info =
  gwas_study_info |>
  mutate(MAPPED_BACKGROUND_TRAIT = ifelse(MAPPED_TRAIT == "response to vaccine" &
                                           grepl("hepatitis B|Hepatitis B", `DISEASE/TRAIT`),
                                         "hepatitis b",
                                         MAPPED_BACKGROUND_TRAIT)
         ) |>
    mutate(MAPPED_BACKGROUND_TRAIT_URI = ifelse(MAPPED_TRAIT == "response to vaccine" &
                                                  grepl("hepatitis B|Hepatitis B", `DISEASE/TRAIT`),
                                                "http://purl.obolibrary.org/obo/DOID_2043",
                                                MAPPED_BACKGROUND_TRAIT_URI)
           )


# if MAPPED_TRAIT == "response to covid-19 vaccine"
# and MAPPED_BACKGROUND_TRAIT == ""
gwas_study_info =
  gwas_study_info |>
  mutate(MAPPED_BACKGROUND_TRAIT = ifelse(MAPPED_TRAIT == "response to covid-19 vaccine" &
                                           (is.na(MAPPED_BACKGROUND_TRAIT) | MAPPED_BACKGROUND_TRAIT == ""),
                                         "covid-19",
                                         MAPPED_BACKGROUND_TRAIT)
         ) |>
    mutate(MAPPED_BACKGROUND_TRAIT_URI = ifelse(MAPPED_TRAIT == "response to covid-19 vaccine" &
                                                  (is.na(MAPPED_BACKGROUND_TRAIT_URI) | MAPPED_BACKGROUND_TRAIT_URI == ""),
                                                "http://purl.obolibrary.org/obo/MONDO_0100096",
                                                MAPPED_BACKGROUND_TRAIT_URI)
           )

# if MAPPED_TRAIT == "response to vaccine"
# and MAPPED_BACKGROUND_TRAIT == ""
# and DISEASE/TRAIT contains "Immune response to smallpox"
# set MAPPED_BACKGROUND_TRAIT to smallpox

gwas_study_info =
  gwas_study_info |>
  mutate(MAPPED_BACKGROUND_TRAIT = ifelse(MAPPED_TRAIT == "response to vaccine" &
                                           (is.na(MAPPED_BACKGROUND_TRAIT) | MAPPED_BACKGROUND_TRAIT == "") &
                                           grepl("Immune response to smallpox", `DISEASE/TRAIT`),
                                         "smallpox",
                                         MAPPED_BACKGROUND_TRAIT)
         ) |>
    mutate(MAPPED_BACKGROUND_TRAIT_URI = ifelse(MAPPED_TRAIT == "response to vaccine" &
                                                  (is.na(MAPPED_BACKGROUND_TRAIT_URI) | MAPPED_BACKGROUND_TRAIT_URI == "") &
                                                  grepl("Immune response to smallpox", `DISEASE/TRAIT`),
                                                "http://purl.obolibrary.org/obo/OMIT_0013787",
                                                MAPPED_BACKGROUND_TRAIT_URI)
           )

# if MAPPED_TRAIT == "response to vaccine"
# and MAPPED_BACKGROUND_TRAIT == ""
# and DISEASE/TRAIT contains "Immune response to measles vaccine"
gwas_study_info =
  gwas_study_info |>
  mutate(MAPPED_BACKGROUND_TRAIT = ifelse(MAPPED_TRAIT == "response to vaccine" &
                                           (is.na(MAPPED_BACKGROUND_TRAIT) | MAPPED_BACKGROUND_TRAIT == "") &
                                           grepl("Immune response to measles vaccine", `DISEASE/TRAIT`),
                                         "measles",
                                         MAPPED_BACKGROUND_TRAIT)
         ) |>
    mutate(MAPPED_BACKGROUND_TRAIT_URI = ifelse(MAPPED_TRAIT == "response to vaccine" &
                                                  (is.na(MAPPED_BACKGROUND_TRAIT_URI) | MAPPED_BACKGROUND_TRAIT_URI == "") &
                                                  grepl("Immune response to measles vaccine", `DISEASE/TRAIT`),
                                                "http://purl.obolibrary.org/obo/DOID_8622",
                                                MAPPED_BACKGROUND_TRAIT_URI)
           )

```

## Replacing * for , inside a MAPPED_TERM 

To ensure splitting MAPPED_TERM column by commas would split distinct traits

```{r fix_splitting_by_commas}

# in MAPPED_BACKGROUND_TRAIT, replace commas with "*" in:
# migraine without aura, susceptibility to, 4
gwas_study_info =
gwas_study_info |>
  mutate(MAPPED_BACKGROUND_TRAIT = ifelse(grepl("migraine without aura, susceptibility to, 4", MAPPED_BACKGROUND_TRAIT),
                                         stringr::str_replace_all(MAPPED_BACKGROUND_TRAIT, 
                                                                  pattern = "migraine without aura, susceptibility to, 4", 
                                                                  "migraine without aura* susceptibility to* 4"),
                                         MAPPED_BACKGROUND_TRAIT)
         )

# in MAPPED_TRAIT, replace commas with "*" in:
# migraine without aura, susceptibility to, 4
gwas_study_info =
gwas_study_info |>
  mutate(MAPPED_TRAIT = ifelse(grepl("migraine without aura, susceptibility to, 4", MAPPED_TRAIT),
                               stringr::str_replace_all(MAPPED_TRAIT, 
                                                        pattern = "migraine without aura, susceptibility to, 4", 
                                                        "migraine without aura* susceptibility to* 4"),
                               MAPPED_TRAIT)
         )

# Other MAPPED_TRAIT fixes
gwas_study_info =
gwas_study_info |> 
  mutate(MAPPED_TRAIT = case_when(
                         # osteoarthritis, hip ... http://www.ebi.ac.uk/efo/EFO_1000786
                         grepl("EFO_1000786", MAPPED_TRAIT_URI) ~ stringr::str_replace_all(MAPPED_TRAIT, 
                                                                                           pattern = "osteoarthritis, hip", 
                                                                                           "osteoarthritis* hip"),
                         
                         # 	osteoarthritis, hand ... http://www.ebi.ac.uk/efo/EFO_1000789
                         grepl("EFO_1000789", MAPPED_TRAIT_URI) ~ stringr::str_replace_all(MAPPED_TRAIT, 
                                                                                           pattern = "osteoarthritis, hand", 
                                                                                           "osteoarthritis* hand"
                                                                                           ),
                         # osteoarthritis, knee ... http://www.ebi.ac.uk/efo/EFO_0004616
                         grepl("EFO_0004616", MAPPED_TRAIT_URI) ~ stringr::str_replace_all(MAPPED_TRAIT, 
                                                                                           pattern = "osteoarthritis, knee", 
                                                                                           "osteoarthritis* knee"),
                         
                         # 	osteoarthritis, spine ... http://www.ebi.ac.uk/efo/EFO_1000787
                         grepl("EFO_1000787", MAPPED_TRAIT_URI) ~ stringr::str_replace_all(MAPPED_TRAIT, 
                                                                                           pattern = "osteoarthritis, spine", 
                                                                                           "osteoarthritis* spine"),
                         
                         # 	Hepatitis, Alcoholic, http://www.ebi.ac.uk/efo/EFO_1001345
                         grepl("EFO_1001345", MAPPED_TRAIT_URI) ~ stringr::str_replace_all(MAPPED_TRAIT, 
                                                                                           pattern = "hepatitis, alcoholic", 
                                                                                           "hepatitis* alcoholic"),
                         
                         # psoriasis 14, pustular http://purl.obolibrary.org/obo/MONDO_0013626
                         grepl("MONDO_0013626", MAPPED_TRAIT_URI) ~ stringr::str_replace_all(MAPPED_TRAIT, 
                                                                                             pattern = "psoriasis 14, pustular", 
                                                                                             "psoriasis 14* pustular"),
                         
                         # hypertension, pregnancy-induced http://purl.obolibrary.org/obo/MONDO_0024664
                         grepl("MONDO_0024664", MAPPED_TRAIT_URI) ~ stringr::str_replace_all(MAPPED_TRAIT, 
                                                                                             pattern = "hypertension, pregnancy-induced", 
                                                                                             "hypertension* pregnancy-induced"),
                         
                         # renal agenesis, unilateral http://purl.obolibrary.org/obo/MONDO_0019636
                         grepl("MONDO_0019636", MAPPED_TRAIT_URI) ~ stringr::str_replace_all(MAPPED_TRAIT, 
                                                                                             pattern = "renal agenesis, unilateral", 
                                                                                             "renal agenesis* unilateral"),
                         
                         # 	Cholecystitis, Acute http://www.ebi.ac.uk/efo/EFO_1001289
                         grepl("EFO_1001289", MAPPED_TRAIT_URI) ~ stringr::str_replace_all(MAPPED_TRAIT, 
                                                                                           pattern = "cholecystitis, acute", 
                                                                                           "cholecystitis* acute"),
                         # 	Genital neoplasm, female http://www.ebi.ac.uk/efo/EFO_1001331
                         grepl("EFO_1001331", MAPPED_TRAIT_URI) ~ stringr::str_replace_all(MAPPED_TRAIT, 
                                                                                           pattern = "genital neoplasm, female", 
                                                                                           "genital neoplasm* female"),
                         # 	Anemia, Hemolytic, Autoimmune http://www.ebi.ac.uk/efo/EFO_1001264
                          grepl("EFO_1001264", MAPPED_TRAIT_URI) ~ stringr::str_replace_all(MAPPED_TRAIT, 
                                                                                            pattern = "anemia, hemolytic, autoimmune",
                                                                                            "anemia* hemolytic* autoimmune"),
                         grepl("EFO_1002020", MAPPED_TRAIT_URI) ~ stringr::str_replace_all(MAPPED_TRAIT, 
                                                                                          pattern = "polyarticular juvenile idiopathic arthritis, rheumatoid factor negative", 
                                                                                          "polyarticular juvenile idiopathic arthritis* rheumatoid factor negative"),
                         # http://www.ebi.ac.uk/efo/EFO_0007294, hand, foot and mouth disease, 
                         grepl("EFO_0007294", MAPPED_TRAIT_URI) ~ stringr::str_replace_all(MAPPED_TRAIT, 
                                                                                          pattern = "hand, foot and mouth disease", 
                                                                                          "hand* foot and mouth disease"),
                         # neural tube defects, susceptibility to, http://purl.obolibrary.org/obo/MONDO_0020705
                         grepl("MONDO_0020705", MAPPED_TRAIT_URI) ~ stringr::str_replace_all(MAPPED_TRAIT, 
                                                                                             pattern = "neural tube defects, susceptibility to", 
                                                                                             "neural tube defects* susceptibility to"),
                         # self-reported traits
                         grepl("EFO_0009803|EFO_0009822|EFO_0009803|EFO_0009817|EFO_0009822|EFO_0009819|EFO_0009823|EFO_0009824", MAPPED_TRAIT_URI) ~ stringr::str_replace_all(MAPPED_TRAIT, 
                                                                                          pattern = ", self-reported$", 
                                                                                          "* self-reported"),
                         # Hodgkins lymphoma, mixed cellularity http://www.ebi.ac.uk/efo/EFO_1002031
                         grepl("EFO_1002031", MAPPED_TRAIT_URI) ~ stringr::str_replace_all(MAPPED_TRAIT, 
                                                                                          pattern = "hodgkins lymphoma, mixed cellularity", 
                                                                                          "hodgkins lymphoma* mixed cellularity"),
                         # encephalopathy, acute, infection-induced, http://purl.obolibrary.org/obo/MONDO_0000166
                         grepl("MONDO_0000166", MAPPED_TRAIT_URI) ~ stringr::str_replace_all(MAPPED_TRAIT, 
                                                                                             pattern = "encephalopathy, acute, infection-induced", 
                                                                                             "encephalopathy* acute* infection-induced"),
                         # Diarrhea, Infantile http://www.ebi.ac.uk/efo/EFO_1001306
                         grepl("EFO_1001306", MAPPED_TRAIT_URI) ~ stringr::str_replace_all(MAPPED_TRAIT, 
                                                                                           pattern = "diarrhea, infantile", 
                                                                                           "diarrhea* infantile"),
                         
                         TRUE ~ MAPPED_TRAIT
                         )
  ) 


# in MAPPED_TRAIT, replace commas with "*" in:
# chromosome, telomeric region length
# fractures, ununited
# osteoarthritis, knee ... http://www.ebi.ac.uk/efo/EFO_0004616
# localized superficial swelling, mass, or lump
# cys-gly, oxidized measurement
gwas_study_info =
  gwas_study_info |>
  mutate(MAPPED_TRAIT = ifelse(grepl("chromosome, telomeric region length", MAPPED_TRAIT),
                               stringr::str_replace_all(MAPPED_TRAIT, 
                                                        pattern = "chromosome, telomeric region length", 
                                                        "chromosome* telomeric region length"),
                               MAPPED_TRAIT)
         ) |>
  mutate(MAPPED_TRAIT = ifelse(grepl("fractures, ununited", MAPPED_TRAIT),
                               stringr::str_replace_all(MAPPED_TRAIT, 
                                                        pattern = "fractures, ununited", 
                                                        "fractures* ununited"),
                               MAPPED_TRAIT)
         ) |>
  mutate(MAPPED_TRAIT =                          # osteoarthritis, knee ... http://www.ebi.ac.uk/efo/EFO_0004616
                         ifelse(grepl("EFO_0004616", MAPPED_TRAIT_URI),
                                stringr::str_replace_all(MAPPED_TRAIT, 
                                                         pattern = "osteoarthritis, knee", 
                                                         "osteoarthritis* knee"),
                                MAPPED_TRAIT)
         ) |>
  mutate(MAPPED_TRAIT = ifelse(grepl("localized superficial swelling, mass, or lump", MAPPED_TRAIT),
                               stringr::str_replace_all(MAPPED_TRAIT, 
                                                        pattern = "localized superficial swelling, mass, or lump", 
                                                        "localized superficial swelling* mass* or lump"),
                               MAPPED_TRAIT)
         ) |>
  mutate(MAPPED_TRAIT = ifelse(grepl("cys-gly, oxidized measurement", MAPPED_TRAIT),
                               stringr::str_replace_all(MAPPED_TRAIT, 
                                                        pattern = "cys-gly, oxidized measurement", 
                                                        "cys-gly* oxidized measurement"),
                               MAPPED_TRAIT)
         )

```

### Fix by the number of commas 

```{r extra_fix_splitting_by_commas}

# fixing weird terms- where comma is in the term

# count number of separating commas in 
# MAPPED_TRAIT, MAPPED_TRAIT_URI, MAPPED_BACKGROUND_TRAIT, MAPPED_BACKGROUND_TRAIT_URI
gwas_study_info =
gwas_study_info |>
  mutate(n_commas_trait = str_count(MAPPED_TRAIT, ", (?![^()]*\\))"),
                                    #", "),
         n_commas_trait_uri = str_count(MAPPED_TRAIT_URI, ", (?![^()]*\\))"),
         n_commas_bg_trait = str_count(MAPPED_BACKGROUND_TRAIT, ", (?![^()]*\\))"),
         n_commas_bg_trait_uri = str_count(MAPPED_BACKGROUND_TRAIT_URI, ", (?![^()]*\\))")
                                          # ", ")
         ) |>
  # select(contains("n_commas"), 
  #        MAPPED_TRAIT, MAPPED_TRAIT_URI, 
  #        MAPPED_BACKGROUND_TRAIT, MAPPED_BACKGROUND_TRAIT_URI
  #        ) |>
  distinct()

# if n_commas_trait >= 1, n_commas_trait_uri == 0,
# replace comma in MAPPED_TRAIT with "*"
gwas_study_info = 
  gwas_study_info |>
  mutate(MAPPED_TRAIT = ifelse(n_commas_trait >= 1 & 
                                  n_commas_trait_uri == 0,
                                stringr::str_replace_all(MAPPED_TRAIT, 
                                                         pattern = ", ", 
                                                         "* "),
                                MAPPED_TRAIT)
         )

# now that's been correct, recalculate number of commas
gwas_study_info = 
  gwas_study_info |>
    mutate(n_commas_trait = str_count(MAPPED_TRAIT, ", (?![^()]*\\))"),
                                    #", "),
         n_commas_trait_uri = str_count(MAPPED_TRAIT_URI, ", (?![^()]*\\))"),
         n_commas_bg_trait = str_count(MAPPED_BACKGROUND_TRAIT, ", (?![^()]*\\))"),
         n_commas_bg_trait_uri = str_count(MAPPED_BACKGROUND_TRAIT_URI, ", (?![^()]*\\))")
                                          # ", ")
         ) 


# check the number of commas in MAPPED_TRAIT is always equal to number of commas in MAPPED_TRAIT_URI
# and the number of commas in MAPPED_BACKGROUND_TRAIT is not equal to number of commas in MAPPED_BACKGROUND_TRAIT_URI
gwas_study_info = 
  gwas_study_info |>
  mutate(match_comma_trait = ifelse(n_commas_trait != n_commas_trait_uri,
                                   FALSE, 
                                   TRUE),
         match_comma_bg_trait = ifelse(n_commas_bg_trait != n_commas_bg_trait_uri,
                                      FALSE, 
                                      TRUE)
         )

gwas_study_info  |>
  filter(match_comma_trait == FALSE |
         match_comma_bg_trait == FALSE)

# yay! all match now

```

# Overlap ontology terms and GWAS traits

## Create data.frame of GWAS traits with one trait per row

```{r gwas_study_info_final_setup}

gwas_study_info =
  gwas_study_info |>
  select(
         `DISEASE/TRAIT`,
         PUBMED_ID,
         YEAR, 
         STUDY,
         STUDY_ACCESSION,
         contains("MAPPED")
         )

# now split by commas to get each MAPPED_TRAIT on an individual row
gwas_study_info <-
gwas_study_info |>
  tidyr::separate_longer_delim(cols = c("MAPPED_TRAIT", 
                                        "MAPPED_TRAIT_URI"
                                        ),
                               delim = stringr::regex(", (?![^()]*\\))")
                               ) |>
  tidyr::separate_longer_delim(cols = c("MAPPED_BACKGROUND_TRAIT", 
                                        "MAPPED_BACKGROUND_TRAIT_URI"
                                        ),
                               delim = stringr::regex(", (?![^()]*\\))")
  ) |>
  distinct()

# now replace '*' back to commas
gwas_study_info =
  gwas_study_info |>
  mutate(MAPPED_TRAIT = stringr::str_replace_all(MAPPED_TRAIT, 
                                                 pattern = "\\* ", 
                                                 ", "),
         MAPPED_BACKGROUND_TRAIT = stringr::str_replace_all(MAPPED_BACKGROUND_TRAIT, 
                                                          pattern = "\\* ", 
                                                          ", ")
         )


gwas_study_info <-
gwas_study_info |>
  mutate(MAPPED_TRAIT = stringr::str_trim(tolower(MAPPED_TRAIT))) |>
  mutate(MAPPED_BACKGROUND_TRAIT = stringr::str_trim(tolower(MAPPED_BACKGROUND_TRAIT))) 


```

```{r prep_mapped_trait_terms}

all_gwas_terms = gwas_study_info$MAPPED_TRAIT
all_gwas_terms = stringr::str_trim(tolower(all_gwas_terms))
all_gwas_terms = unique(all_gwas_terms)

print("Number of unique GWAS traits")
length(all_gwas_terms)

```

## Disease Overlap (How many GWAS traits fall within disease or disorder terms?)

### Combine disease terms

```{r combine_disease_terms}

efo_descendants <- readLines(here::here("output/trait_ontology/efo_0000408_descendants.txt"))

mondo_descendants <- readLines(here::here("output/trait_ontology/mondo_0700096_descendants.txt"))

ncit_descendants <- readLines(here::here("output/trait_ontology/ncit_C2991_descendants.txt"))

orphanet_descendants <- readLines(here::here("output/trait_ontology/orphanet_557493_descendants.txt"))

age_of_onset_descendants <- readLines(here::here("output/trait_ontology/oba_2020000_descendants.txt"))

disease_measurement_terms <- readLines(here::here("output/trait_ontology/efo_0001444_disease_measurement_terms.txt"))

disease_typos = c("Alzheimer disease",
                  "late-onset Alzheimers disease",
                  "Chagas cardiomyopathy",
                  "Parkinson disease",
                  "Iron deficiency anemia"
                  )

biomarker_terms <- c("cardiovascular disease biomarker measurement",
                     "cancer biomarker measurement",
                     "diabetes mellitus biomarker",
                     "osteoarthritis biomarker measurement",
                     "liver disease biomarker",
                     "alzheimer's disease biomarker measurement",
                     "iron deficiency anemia (disorder)"
                     )

other_disorders <- c(
           "Allergic disease", 
           "Churg-Strauss syndrome",
           "Iridocyclitis",
           "Phlebitis",
           "pregnancy induced alloimmunization",
           "somnambulism",
           "suicide",
           "attempted suicide",
           "suicide behaviour",
           "suicide ideation measurement",
           "suicide behaviour measurement",
           "Lewy body dementia",
           "Lewy body attribute",
           "non-Hodgkins lymphoma",
           "Ischemic Stroke",
           "Lung disease",
           "Respiratory System Disease",
           "Alzheimer disease, APOE carrier status",
           "Genital neoplasm, female",
           "HIV-associated neurocognitive disorder",
           "encephalopathy acute infection-induced",
           "anomalous atrioventricular excitation (disorder)",
           "scleritis and episcleritis (disorder)",
           "atopic march",
           "infection",
           "neural tube defects, susceptibility to",
           "migraine without aura, susceptibility to, 4",
           "hiv mother to child transmission",
           "hemolysis",
           "chromosomal aberration",
           "dna methylation",
           "gata1 gene mutation",
           "atropy",
           "premature birth",
           "growth delay",
           "reduced left ventricular ejection fraction",
           "hepatitis b",
           "vascular brain injury measurement",
           "borderline personality disorder symptom",
           "miscarriage",
           "emphysema pattern measurement",
           "emphysema imaging measurement",
           "persistent staphylococcus aureus carrier status",
           "intermittent staphylococcus aureus carrier status",
           "influenza a severity measurement",
           "pneumonia severity measurement",
           "hsv2 virologic severity measurement",
           "opioid overdose severity measurement",
           "nausea and vomiting of pregnancy severity measurement",
           "myopic maculopathy severity measurement",
           "hepatitis C virus infection"
)
 
 
 disease_status_terms <- c(
 "benign",
 "remission",
 "disease recurrence",
 "complicated disease course",
 "disease prognosis measurement",
 "mild disease course",
 "disease free survival",
 "progression free survival",
 "survival time",
 "overall survival",
 "illness severity status"
 )

family_disease_terms <- c("family history of breast cancer",
                          "family history of cancer",
                          "family history of prostate cancer",
                          "family history of upper gastrointestinal cancer",
                          "family history of uterine fibroids")

disease_terms = c(mondo_descendants,
                  efo_descendants,
                  ncit_descendants,
                  orphanet_descendants,
                  age_of_onset_descendants,
                  disease_measurement_terms,
                  family_disease_terms,
                  disease_typos,
                  biomarker_terms,
                  disease_status_terms,
                  other_disorders) |>
                  unique()


disease_terms = stringr::str_trim(tolower(disease_terms))
disease_terms = unique(disease_terms)

print("Number of ontology terms found related to disease or disorder")
length(disease_terms)

```

### Find disease terms in GWAS traits

```{r find_disease_terms_in_mapped_traits}
# Find GWAS traits that fall within disease or disorder terms
#simple_disease_terms = all_gwas_terms[all_gwas_terms %in% disease_terms]
disease_gwas <- all_gwas_terms[all_gwas_terms %in% disease_terms]

not_disease_terms = all_gwas_terms[!all_gwas_terms %in% disease_gwas]

print("Number of GWAS traits under disease or disorder terms")
length(disease_gwas)

print("Percentage of GWAS traits under disease or disorder terms")
round(100 * (length(disease_gwas)) / length(all_gwas_terms),
      digits = 1)

print("Percentage of GWAS traits not under disease or disorder terms")
round(100 * length(not_disease_terms) / length(all_gwas_terms),
      digits = 1)

not_accounted_for = not_disease_terms 

```

## Phenotype abnormality overlap

```{r find_phenotype_abnormality_terms_in_mapped_traits}

pheno_abnorm <- readLines(here::here("output/trait_ontology/hp_0000118_descendants.txt"))
pheno_abnorm = stringr::str_trim(tolower(pheno_abnorm))
pheno_abnorm = unique(pheno_abnorm)

pheno_abnorm <- c("abnormal pap smear",
                  "abnormal result of function studies",
                  "abnormal result of diagnostic imaging",
                  pheno_abnorm)

pheno_abnorm_gwas <- not_accounted_for[not_accounted_for %in% pheno_abnorm]

print("Percentage of GWAS traits under phenotype abnormality terms")
round(100 * length(pheno_abnorm_gwas) / length(all_gwas_terms),
      digits = 1)

not_accounted_for = not_accounted_for[!not_accounted_for %in% pheno_abnorm_gwas]

print("Percentage of GWAS traits not accounted for so far")
round(100 * length(not_accounted_for) / length(all_gwas_terms),
      digits = 1)

print("Number of GWAS traits not accounted for by so far")
length(not_accounted_for)

```


## Measurement Overlap (how many GWAS traits fall within measurement terms?)

### Combine measurement

```{r combine_measurement_terms}

measurement <- readLines(here::here("output/trait_ontology/efo_0001444_descendants.txt"))
total_choles <- readLines(here::here("output/trait_ontology/efo_0004574_descendants.txt"))
measurement <- c(total_choles,
                 measurement) 

measurement <- unique(measurement)
measurement <- c("cerebrospinal fluid composition attribute",
                 "blood protein amount",
                 "fatty acid measurement",
                 "obsolete_3,3',5-triiodo-l-thyronine measurement",
                 "1-(1-enyl-stearoyl)-2-linoleoyl-gpe (p-18:0/18:2), measurement",
                 "microtubule-associated protein tau",
                 measurement)

measurement = stringr::str_trim(tolower(measurement))
measurement = unique(measurement)
```

### BMI / weight terms / body fat terms

```{r bmi_weight_terms_in_gwas}

bmi_weight_terms <- grep("bmi|body mass index|weight|bmi", measurement, value = T)
bmi_weight_terms <- grep("fetal|birth|gestational", bmi_weight_terms, value = T, invert = T)
bmi_weight_terms <- c(bmi_weight_terms, 
                      "body composition measurement",
                      "body fat percentage",
                      "body fat distribution")

measurement <- measurement[!(measurement %in% bmi_weight_terms)]

bmi_weight_gwas = not_accounted_for[not_accounted_for %in% bmi_weight_terms]
print("Number of GWAS traits under BMI / weight / body fat terms")
length(bmi_weight_gwas)

print("Percentage of GWAS traits under BMI / weight / body fat terms")
round(100 * length(bmi_weight_gwas) / length(all_gwas_terms),
      digits = 1)

not_accounted_for = not_accounted_for[!not_accounted_for %in% bmi_weight_gwas]
print("Percentage of GWAS traits not accounted for by disease, disorder or BMI / weight / body fat terms")

round(100 * length(not_accounted_for) / length(all_gwas_terms),
      digits = 1)


```

### Lipid / cholesterol measurement terms

```{r lipid_cholesterol_terms_in_gwas}

lipid_cholesterol_terms <- grep("cholesterol|lipid|triglyceride|ldl|hdl|apolipoprotein", 
                                measurement, 
                                value = T)

measurement <- measurement[!(measurement %in% lipid_cholesterol_terms)]

lipid_cholesterol_gwas = not_accounted_for[not_accounted_for %in% lipid_cholesterol_terms]

print("Number of GWAS traits under lipid / cholesterol terms")
length(lipid_cholesterol_gwas)

print("Percentage of GWAS traits under lipid / cholesterol terms")
round(100 * length(lipid_cholesterol_gwas) / length(all_gwas_terms),
      digits = 1)

not_accounted_for = not_accounted_for[!not_accounted_for %in% lipid_cholesterol_gwas]
print("Percentage of GWAS traits not accounted for by disease, disorder or lipid / cholesterol terms")
round(100 * length(not_accounted_for) / length(all_gwas_terms),
      digits = 1)

```

### Brain measurement terms

```{r brain_measurement_terms_in_gwas}

brain_measurement_terms <- grep("brain|volume", 
                                measurement, 
                                value = T)

brain_measurement_terms <- grep("bone|muscle|reticulocyte|erythrocyte|expiratory|platelet|urinary|thyroid|pancreas|kidney|spleen|liver|ventricular|blood", 
                                brain_measurement_terms, 
                                value = T, 
                                invert = T)


```

### Blood pressure measurement terms

```{r blood_pressure_terms_in_gwas}

blood_pressure_terms <- grep("blood pressure", 
                                measurement, 
                                value = T)

measurement <- measurement[!(measurement %in% blood_pressure_terms)]

blood_pressure_gwas = not_accounted_for[not_accounted_for %in% blood_pressure_terms]

print("Number of GWAS traits under blood pressure terms")
length(blood_pressure_gwas)

print("Percentage of GWAS traits under blood pressure terms")
round(100 * length(blood_pressure_gwas) / length(all_gwas_terms),
      digits = 1)

not_accounted_for = not_accounted_for[!not_accounted_for %in% blood_pressure_gwas]

print("Percentage of GWAS traits not accounted for by disease, disorder or blood pressure terms")
round(100 * length(not_accounted_for) / length(all_gwas_terms),
      digits = 1)


```

### Find seropositivity terms in GWAS traits

```{r seropositivity_terms_in_gwas}

seropositivity_terms <- grep("seropositivity|antibody", measurement, value = T)
seropositivity_terms <- c(seropositivity_terms,
                          "foot-and-mouth disease virus seropositivity",
                          "bacillus phage virus seropositivity")

measurement <- measurement[!(measurement %in% seropositivity_terms)]

seropositivity_gwas = not_accounted_for[not_accounted_for %in% seropositivity_terms]

print("Number of GWAS traits under seropositivity terms")
length(seropositivity_gwas)

print("Percentage of GWAS traits under seropositivity terms")
round(100 * length(seropositivity_gwas) / length(all_gwas_terms),
      digits = 1)

not_accounted_for = not_accounted_for[!not_accounted_for %in% seropositivity_gwas]
print("Percentage of GWAS traits not accounted for by disease, disorder or seropositivity terms")
round(100 * length(not_accounted_for) / length(all_gwas_terms),
      digits = 1)

```

### General measurement terms 

```{r general_measurement_terms_in_gwas}

behavior_measurement <- c(
                    "smoking",
                    "alcohol consumption",
                    "alcoholic beverage consumption",
                    "alcohol exposure",
                    "behavior",
                    "farm exposure",
                    "tobacco",
                    "cannabis",
                    "physical activity",
                    "cognitive function",
                    "pack-years",
                    "coffee",
                    "opioid",
                    "environment",
                    "exercise"
                    )

behavior_measurement< grep(paste0(behavior_measurement, 
                                  collapse = "|"), 
                            measurement,
                            value = T
                           )

measurement <- measurement[!(measurement %in% behavior_measurement)]

measurement <- grep("emphysema|eye colour|lifestyle", 
                    measurement, 
                    value = T, 
                    invert = T)

measurement_gwas <- not_accounted_for[not_accounted_for %in% measurement]

print("Number of GWAS traits under measurement terms")
length(measurement_gwas)

print("Percentage of GWAS traits under measurement terms")
round(100 * length(measurement_gwas) / length(all_gwas_terms),
      digits = 1)

not_accounted_for = not_accounted_for[!not_accounted_for %in% measurement_gwas]

print("Percentage of GWAS traits not accounted for by disease, disorder or measurement terms")
round(100 * length(not_accounted_for) / length(all_gwas_terms),
      digits = 1)

print("Number of GWAS traits not accounted for by disease, disorder or measurement terms")
length(not_accounted_for)
```

## Response to stimulus

### Combine response terms

```{r combine_response_terms}

go_response = readLines(here::here("output/trait_ontology/go_0050896_descendants.txt"))
efo_response <- readLines(here::here("output/trait_ontology/efo_go_0050896_descendants.txt"))

response <- c(go_response,
              efo_response,
              "response to stimulus")

response <- unique(response)

response = stringr::str_trim(tolower(response))
response = unique(response)

```

```{r response_terms_in_gwas}

response_gwas <- not_accounted_for[not_accounted_for %in% response]

#additional_response <- not_accounted_for[not_accounted_for %in% response]
#response_gwas  = c(response_gwas, additional_response) |> unique()

print("Percentage of GWAS traits under response terms")
round(100 * length(response_gwas) / length(all_gwas_terms),
      digits = 1)

not_accounted_for = not_accounted_for[!not_accounted_for %in% response_gwas]

print("Percentage of GWAS traits not accounted for by disease, measurement or response terms")
round(100 * length(not_accounted_for) / length(all_gwas_terms),
      digits = 1)

print("Number of GWAS traits not accounted for by disease, measurement or response terms")
length(not_accounted_for)

```

## Mental process

```{r mental_process_terms_in_gwas}

mental <- readLines(here::here("output/trait_ontology/efo_0004323_descendants.txt"))
mental = stringr::str_trim(tolower(mental))
mental <- unique(mental)
mental <- c(mental,
          "memory performance",
          "visual memory process attribute",
          "verbal memory measurement",
          "executive function measurement",
          "cognitive function measurement"
          )

mental_gwas = not_accounted_for[not_accounted_for %in% mental]

print("Percentage of GWAS traits under mental process terms")
round(100 * length(mental_gwas) / length(all_gwas_terms),
      digits = 1)

not_accounted_for = not_accounted_for[!not_accounted_for %in% mental_gwas]

print("Percentage of GWAS traits not accounted for thus far")
round(100 * length(not_accounted_for) / length(all_gwas_terms),
      digits = 1)

print("Number of GWAS traits not accounted for thus far")
length(not_accounted_for)

```

## Behavior

```{r find_behavior_terms_in_gwas}

behavior <- readLines(here::here("output/trait_ontology/go_0007610_descendants.txt"))
behavior = stringr::str_trim(tolower(behavior))
behavior <- unique(behavior)
behavior <- c(behavior,
              behavior_measurement,
              "physical activity")

behavior_gwas = not_accounted_for[not_accounted_for %in% behavior]

print("Percentage of GWAS traits under behavouir terms")
round(100 * length(behavior_gwas) / length(all_gwas_terms),
      digits = 1)

not_accounted_for = not_accounted_for[!not_accounted_for %in% behavior_gwas]

print("Percentage of GWAS traits not accounted for so far")
round(100 * length(not_accounted_for) / length(all_gwas_terms),
      digits = 1)

print("Number of GWAS traits not accounted for so far")
length(not_accounted_for)

```

## Injury

```{r find_injury_terms_in_gwas}

injury <- readLines(here::here("output/trait_ontology/efo_0000546_descendants.txt"))
injury = stringr::str_trim(tolower(injury))
injury <- c(injury,
            "fall")

injury_gwas = not_accounted_for[not_accounted_for %in% injury]

print("Percentage of GWAS traits under injury terms")
round(100 * length(injury_gwas) / length(all_gwas_terms),
      digits = 1)

not_accounted_for = not_accounted_for[!not_accounted_for %in% injury_gwas]

print("Percentage of GWAS traits not accounted for so far")
round(100 * length(not_accounted_for) / length(all_gwas_terms),
      digits = 1)

print("Number of GWAS traits not accounted for so far")
length(not_accounted_for)

```

## Phenotype

```{r find_phenotype_terms_in_gwas}

phenotype <- readLines(here::here("output/trait_ontology/efo_0000651_descendants.txt"))
phenotype = stringr::str_trim(tolower(phenotype))
phenotype <- unique(c(phenotype,
                      "aging",
                      "biological sex",
                      "comparative body size at age 10, self-reported",
                      "complex trait",                      
                      "eye colour measurement",
                      "strand of hair color",
                      "high altitude adaptation",
                      "multiple gestation",
                      "normal",
                      "personality trait",
                      "skin pigmentation",
                      "personality",
                      "growth delay",
                      "sensory perception of taste",
                      "sensory perception of bitter taste",
                      "sensory perception of sweet taste",
                      "sensory perception of smell",
                      "sensory perception of sound",
                      "size",
                      "skin aging",
                      "sexual dimorphism",
                      "voice quality trait")
)

phenotype_gwas = not_accounted_for[not_accounted_for %in% phenotype]

print("Percentage of GWAS traits under phenotype terms")
round(100 * length(phenotype_gwas) / length(all_gwas_terms),
      digits = 1)

not_accounted_for = not_accounted_for[!not_accounted_for %in% phenotype_gwas]

print("Percentage of GWAS traits not accounted for so far")
round(100 * length(not_accounted_for) / length(all_gwas_terms),
      digits = 1)

print("Number of GWAS traits not accounted for so far")
length(not_accounted_for)

```

## Medical procedure

```{r find_medical_procedure_terms_in_gwas}

medical_procedure <- readLines(here::here("output/trait_ontology/efo_0002571_descendants.txt"))
surgical_procedure <- readLines(here::here("output/trait_ontology/maxo_0000004_descendants.txt"))
clinical_history <- c("clinical history",
                      "encounter with health service",
                      "encounter with health service for adjustment and management of implanted device",
                      "encounter with health service related to reproduction")



medical_procedure = stringr::str_trim(tolower(medical_procedure))
medical_procedure = unique(c(medical_procedure, 
                             surgical_procedure,
                             clinical_history,
                             "braces",
                             "vaccination",
                             "hormone replacement therapy",
                             "cognitive behavioural therapy",
                             "organ extraction",
                             "gastric bypass",
                             "medical procedure",
                             "number of treatments or medications taken, self-reported",
                             "treatment",
                             "test result",
                             "hospitalisation",
                             "clinical treatment")
                           )



medical_procedure_gwas = not_accounted_for[not_accounted_for %in% medical_procedure]

print("Percentage of GWAS traits under medical procedure terms")
round(100 * length(medical_procedure_gwas) / length(all_gwas_terms),
      digits = 1)

not_accounted_for = not_accounted_for[!not_accounted_for %in% medical_procedure_gwas]
print("Percentage of GWAS traits not accounted for so far")
round(100 * length(not_accounted_for) / length(all_gwas_terms),
      digits = 1)

print("Number of GWAS traits not accounted for so far")
length(not_accounted_for)

```

## Environmental factors

```{r find_environmental_factor_terms_in_gwas}

enviro_factors <- c(
"diet measurement",
"economic and social preference",
"educational attainment",
"encounter with health service related to socioeconomic and psychosocial circumstances" ,
"energy intake",
"environmental factor",
"family relationship",
"household income",
"income",
"lifestyle measurement",
"risk factor",
"self reported educational attainment",
"social deprivation",
"social risk factor",
"socioeconomic status",
"townsend deprivation index"
)


enviro_factors_gwas = not_accounted_for[not_accounted_for %in% enviro_factors]

print("Percentage of GWAS traits under environmental factor terms")
round(100 * length(enviro_factors_gwas) / length(all_gwas_terms),
      digits = 1)

not_accounted_for = not_accounted_for[!not_accounted_for %in% enviro_factors_gwas]

print("Percentage of GWAS traits not accounted for so far")
round(100 * length(not_accounted_for) / length(all_gwas_terms),
      digits = 1)

print("Number of GWAS traits not accounted for so far")
length(not_accounted_for)


```

## Other

### Biological process 

```{r find_biological_process_terms_in_gwas}

bio_process <- c("pregnancy",
                 "puberty",
                 "menopause",
                 "ovulation",
                 "positive regulation of ovulation")

bio_process = stringr::str_trim(tolower(bio_process))

bio_process_gwas = not_accounted_for[not_accounted_for %in% bio_process]

print("Percentage of GWAS traits under biological process terms")
round(100 * length(bio_process_gwas) / length(all_gwas_terms),
      digits = 1)

not_accounted_for = not_accounted_for[!not_accounted_for %in% bio_process_gwas]
print("Percentage of GWAS traits not accounted for so far")

round(100 * length(not_accounted_for) / length(all_gwas_terms),
      digits = 1)

print("Number of GWAS traits not accounted for so far")
length(not_accounted_for)


```


### Traditional medicine constitutional types

```{r find_traditional_medicine_constitution_terms_in_gwas}

tm_constitution <- c("yu-zhi constitution type",
                      "sasang constitutional medicine",
                      "sasang constitutional medicine type",
                      "hepatonia constitution type",
                      "pulmotonia constitution type",
                      "tae-yang",
                      "tae-eum",
                      "so-eum",
                      "so-yang"
                      )

tm_constitution = stringr::str_trim(tolower(tm_constitution))

tm_constitution_gwas = not_accounted_for[not_accounted_for %in% tm_constitution]

print("Percentage of GWAS traits under traditional medicine constitutional type terms")
round(100 * length(tm_constitution_gwas) / length(all_gwas_terms),
      digits = 1)

not_accounted_for = not_accounted_for[!not_accounted_for %in% tm_constitution_gwas]

print("Percentage of GWAS traits not accounted for so far")
round(100 * length(not_accounted_for) / length(all_gwas_terms),
      digits = 1)

print("Number of GWAS traits not accounted for so far")
length(not_accounted_for)
      
```

### Ancestry-info

```{r find_ancestry_terms_in_gwas}

ancestry_gwas <- "latin or admixed american ancestry"

not_accounted_for = not_accounted_for[!not_accounted_for %in% ancestry_gwas]

```

### Cell lines 

```{r find_cell_line_terms_in_gwas}

cell_line_gwas <- "gm11992"

not_accounted_for = not_accounted_for[!not_accounted_for %in% cell_line_gwas]

```



# Add Categories to GWAS Info

## Add disease & phenotypic abnormality terms to GWAS study info dataset

```{r disease_terms}

disease_or_disorder <- c(disease_gwas,
                          pheno_abnorm_gwas,
                          seropositivity_gwas
                          )


disease_progress_measure <- 
gwas_study_info |> 
filter(MAPPED_TRAIT == "disease prognosis measurement") |>
pull(STUDY_ACCESSION) |>
unique()

# for all GWAS Catalog studies with trait, "disease prognosis measurement"
# the actual disease is captured / recorded in trait, so we can remove disease prognosis measurement and safely capture all disease studies
gwas_study_info |>
filter(STUDY_ACCESSION %in% disease_progress_measure) |>
select(STUDY_ACCESSION, MAPPED_TRAIT, MAPPED_BACKGROUND_TRAIT)

# similarly for: 
# disease free survival
# complicated disease course
# mild disease course
# remission
# progression free survival
# illness severity status
unneeded_disease_progress_terms <-
  c("disease prognosis measurement",
"complicated disease course",
"mild disease course",
"remission",
"disease free survival",
"survival time",
"overall survival",
"progression free survival",
"illness severity status"
)

# accessory eyelid
# dna methylation
# tube feeding
# widow's peak
# gata1 gene mutation
other_not_disease_terms <- 
  c(
"anti-drug antibody measurement",
"accessory eyelid",
"dna methylation",
"tube feeding",
"widow's peak",
"gata1 gene mutation"
  )


disease_or_disorder <-
  disease_or_disorder[!(disease_or_disorder %in% 
                        c(unneeded_disease_progress_terms,
                          other_not_disease_terms
                        )
                        )]

gwas_study_info <- 
  gwas_study_info |> 
  #dplyr::rowwise() |>
  dplyr::mutate(
    disease_terms = 
      ifelse(MAPPED_TRAIT %in% disease_or_disorder,
             MAPPED_TRAIT,
             NA)
  )



```

## Map GWAS traits to high-level categories

```{r map_gwas_traits_to_categories}

gwas_study_info = 
gwas_study_info |>
  dplyr::mutate(MAPPED_TRAIT_CATEGORY = dplyr::case_when(is.na(MAPPED_TRAIT) ~ NA,
                                                         MAPPED_TRAIT == "" ~ NA,
                                                         tolower(MAPPED_TRAIT) %in% disease_or_disorder ~ "Disease/Disorder",
                                                         #tolower(MAPPED_TRAIT) %in% pheno_abnorm_gwas ~ "Phenotypic Abnormality",
                                                         tolower(MAPPED_TRAIT) %in% seropositivity_gwas ~ "Seropositivity",
                                                         tolower(MAPPED_TRAIT) %in% bmi_weight_gwas ~ "BMI/Weight/Body Fat Measurement",
                                                         tolower(MAPPED_TRAIT) %in% lipid_cholesterol_gwas ~ "Lipid/Cholesterol Measurement",
                                                         tolower(MAPPED_TRAIT) %in% brain_measurement_terms ~ "Brain Measurement",
                                                         tolower(MAPPED_TRAIT) %in% blood_pressure_gwas ~ "Blood Pressure Measurement",
                                                         tolower(MAPPED_TRAIT) %in% measurement_gwas ~ "Measurement",
                                                         tolower(MAPPED_TRAIT) %in% response_gwas ~ "Response",
                                                         tolower(MAPPED_TRAIT) %in% mental_gwas ~ "Mental Process",
                                                         tolower(MAPPED_TRAIT) %in% behavior_gwas ~ "Behavior",
                                                         tolower(MAPPED_TRAIT) %in% injury_gwas ~ "Injury",
                                                         tolower(MAPPED_TRAIT) %in% phenotype_gwas ~ "Phenotype",
                                                         tolower(MAPPED_TRAIT) %in% medical_procedure_gwas ~ "Medical Procedure",
                                                         tolower(MAPPED_TRAIT) %in% enviro_factors_gwas ~ "Environmental Factor",
                                                         
                                                          TRUE ~ "Other"
                                                          )
                )

```

# Background traits

```{r background_traits}

gwas_background <- gwas_study_info$MAPPED_BACKGROUND_TRAIT

gwas_background = stringr::str_trim(tolower(gwas_background))
gwas_background <- unique(gwas_background)
gwas_background <- gwas_background[gwas_background != ""]


print("Number of unique background GWAS traits")
length(gwas_background)

```

## Overlap with disease/disorder traits

```{r background_disease_terms}

disease_gwas = gwas_background[gwas_background %in% disease_terms]

print("Number of background GWAS traits under disease or disorder terms")
length(disease_gwas)

print("Percentage of background GWAS traits under disease or disorder terms")
round(100 * length(disease_gwas) / length(gwas_background),
      digits = 1)

not_accounted_for = gwas_background[!gwas_background %in% disease_gwas]

print("Percentage of background GWAS traits not under disease or disorder terms")
round(100 * length(not_accounted_for) / length(gwas_background),
      digits = 1)

```

## Phenotype abnormality overlap

```{r background_phenotype_abnormality_terms}

pheno_abnorm_gwas  = pheno_abnorm_gwas

print("Percentage of background GWAS traits under phenotype abnormality terms")
round(100 * length(pheno_abnorm_gwas) / length(gwas_background),
      digits = 1)

not_accounted_for = not_accounted_for[!not_accounted_for %in% pheno_abnorm_gwas]

print("Percentage of background GWAS traits not accounted for so far")
round(100 * length(not_accounted_for) / length(gwas_background),
      digits = 1)

print("Number of background GWAS traits not accounted for so far")
length(not_accounted_for)

```

## Add disease & phenotypic abnormality terms to GWAS study info dataset

```{r background_trait_disease_terms}

disease_or_disorder <- c(disease_gwas,
                          pheno_abnorm_gwas
  )

gwas_study_info <- 
  gwas_study_info |>
  #rowwise() |>
  dplyr::mutate(
    background_disease_terms = 
      ifelse(MAPPED_BACKGROUND_TRAIT %in% disease_or_disorder,
             MAPPED_BACKGROUND_TRAIT,
             NA)
  ) |>
  ungroup()

```

## Measurement traits

```{r background_measurement_terms}

measurement_gwas  = measurement[measurement %in% not_accounted_for]

not_accounted_for = not_accounted_for[!not_accounted_for %in% measurement_gwas]

print("Percentage of background GWAS traits not accounted for by disease, disorder or measurement terms")
round(100 * length(not_accounted_for) / length(gwas_background),
      digits = 1)

print("Number of background GWAS traits not accounted for by disease, disorder or measurement terms")
length(not_accounted_for)


```

## Response traits

```{r background_response_terms}

response_gwas  = response[response %in% not_accounted_for]

not_accounted_for = not_accounted_for[!not_accounted_for %in% response_gwas]

print("Percentage of background GWAS traits under response terms")
round(100 * length(response_gwas) / length(gwas_background),
      digits = 1)

print("Number of background GWAS traits under response terms")
length(response_gwas)

print("Number of background GWAS traits not accounted for by disease, measurement or response terms")
length(not_accounted_for)

print("Percentage of background GWAS traits not accounted for by disease, measurement or response terms")
round(100 * length(not_accounted_for) / length(gwas_background),
      digits = 1)


```

## Medical procedure traits

```{r background_medical_procedure_terms}

medical_procedure_gwas = medical_procedure[medical_procedure %in% not_accounted_for]

not_accounted_for = not_accounted_for[!not_accounted_for %in% medical_procedure_gwas]

print("Percentage of background GWAS traits not accounted for by disease, measurement, response or medical procedure terms")
round(100 * length(not_accounted_for) / length(gwas_background),
      digits = 1)

print("Number of background GWAS traits not accounted for by disease, measurement, response or medical procedure terms")
length(not_accounted_for)

```

## Background trait categories 

```{r map_background_traits_to_categories}

gwas_study_info = 
gwas_study_info |>
  dplyr::mutate(BACKGROUND_TRAIT_CATEGORY = 
                   dplyr::case_when(
                                      MAPPED_BACKGROUND_TRAIT == "" ~ NA,
                                      is.na(MAPPED_BACKGROUND_TRAIT) ~ NA,
                                      stringr::str_trim(tolower(MAPPED_BACKGROUND_TRAIT)) %in% disease_or_disorder ~ "Disease/Disorder",
                                      #stringr::str_trim(tolower(MAPPED_BACKGROUND_TRAIT)) %in% pheno_abnorm_gwas ~ "Phenotypic Abnormality",
                                      stringr::str_trim(tolower(MAPPED_BACKGROUND_TRAIT)) %in% measurement_gwas ~ "Measurement",
                                      stringr::str_trim(tolower(MAPPED_BACKGROUND_TRAIT)) %in% response_gwas ~ "Response",
                                      stringr::str_trim(tolower(MAPPED_BACKGROUND_TRAIT)) %in% medical_procedure_gwas ~ "Medical Procedure",
                                      TRUE ~ "Other")
                )

```


# Summary of number of disease studies (and studies of each kind of trait)

```{r summary_of_study_types}

gwas_study_info |>
  group_by(MAPPED_TRAIT_CATEGORY, BACKGROUND_TRAIT_CATEGORY) |>
  summarise(n_studies = n()) |> 
  arrange(desc(n_studies))

gwas_study_info = 
gwas_study_info |>
  #dplyr::rowwise() |>
  dplyr::mutate(DISEASE_STUDY = 
                   case_when(MAPPED_TRAIT_CATEGORY == "Disease/Disorder" | 
                             MAPPED_TRAIT_CATEGORY == "Phenotypic Abnormality" |  
                             MAPPED_TRAIT_CATEGORY == "Seropositivity" |
                             BACKGROUND_TRAIT_CATEGORY == "Disease/Disorder" | 
                             BACKGROUND_TRAIT_CATEGORY == "Phenotypic Abnormality" ~ T,
                             T ~ F )
                ) |>
  dplyr::ungroup() 

print("Number of studies of each kind of trait")
gwas_study_info |> 
  group_by(DISEASE_STUDY, 
           MAPPED_TRAIT_CATEGORY, 
           BACKGROUND_TRAIT_CATEGORY) |> 
  summarise(n = n()) 

# Number of papers with at least one disease study
print("Number of papers with at least one disease study")
gwas_study_info |> 
  group_by(PUBMED_ID) |> 
  summarise(DISEASE_STUDY = any(DISEASE_STUDY == T)) |> 
  group_by(DISEASE_STUDY) |> 
  summarise(n = n())
# ~ 60% of papers have at least one disease study

gwas_study_info |> 
  group_by(PUBMED_ID, YEAR) |> 
  summarise(DISEASE_STUDY = any(DISEASE_STUDY == T)) |> 
  group_by(YEAR) |>
  summarise(n_disease_studies = sum(DISEASE_STUDY == T),
            n_total_studies = n()) |>
  mutate(percentage_disease_studies = 100 * n_disease_studies / n_total_studies) |>
  ggplot(aes(x= YEAR, 
             y= percentage_disease_studies)) +
  geom_line() +
  geom_point() +
  labs(title = "Percentage of papers with at least one disease GWAS over time",
       x = "Year",
       y = "Percentage of papers with at least one disease GWAS") +
  theme_bw() + 
  lims(y = c(0,100))

```

# Check for disease terms in traits of studies that don't have a disease term mapped to either the trait or background trait - to see if there are any disease studies that we missed

```{r check_for_disease_terms_in_non_disease_studies}

not_disease_pubmeds <- gwas_study_info  |> 
                       group_by(PUBMED_ID) |> 
                       summarise(DISEASE_STUDY = any(DISEASE_STUDY)) |> 
                       filter(DISEASE_STUDY == F) |> 
                       pull(PUBMED_ID) |>
                       unique()

print("Number of papers with no disease study")
length(not_disease_pubmeds)

```

## Cancer

```{r cancer_terms_in_non_disease_studies}

gwas_study_info |> 
  filter(PUBMED_ID %in% not_disease_pubmeds) |> 
  filter(grepl("breast cancer", `DISEASE/TRAIT`, ignore.case = T)) |>
  select(PUBMED_ID, `DISEASE/TRAIT`, STUDY) |>
  distinct()

# for these studies, add breast cancer to the disease terms column, and disease/disorder to the trait category column, and then we can re-run the summary of number of studies of each kind

add_missed_breast_cancer <-
gwas_study_info |>
    filter(PUBMED_ID %in% not_disease_pubmeds) |> 
    filter(grepl("breast cancer", `DISEASE/TRAIT`, ignore.case = T)) |>
    mutate(MAPPED_TRAIT_CATEGORY = "Disease/Disorder",
           disease_terms = "breast carcinoma",
           DISEASE_STUDY = T)

# colorectal cancer
add_missed_colorectal_cancer <-
gwas_study_info |>
    filter(PUBMED_ID %in% not_disease_pubmeds) |> 
    filter(grepl("colorectal cancer", `DISEASE/TRAIT`, ignore.case = T)) |>
    mutate(MAPPED_TRAIT_CATEGORY = "Disease/Disorder",
           disease_terms = "colorectal cancer",
           DISEASE_STUDY = T) 

# colon cancer
add_missed_colon_cancer <-
gwas_study_info |>
    filter(PUBMED_ID %in% not_disease_pubmeds) |> 
    filter(grepl("colon cancer", `DISEASE/TRAIT`, ignore.case = T)) |>
    mutate(MAPPED_TRAIT_CATEGORY = "Disease/Disorder",
           disease_terms = "colorectal adenocarcinoma",
           DISEASE_STUDY = T) 

# small-cell lung cancer
add_small_cell_lung_carcinoma <-
gwas_study_info |> 
    filter(PUBMED_ID %in% not_disease_pubmeds) |> 
    filter(grepl("small-cell lung cancer", `DISEASE/TRAIT`, ignore.case = T)) |>
    mutate(MAPPED_TRAIT_CATEGORY = "Disease/Disorder",
           disease_terms = "small cell lung carcinoma",
           DISEASE_STUDY = T) 
  
# lung cancer
add_lung_cancer <-
  gwas_study_info |> 
    filter(PUBMED_ID %in% not_disease_pubmeds) |> 
    filter(grepl("lung cancer", `DISEASE/TRAIT`, ignore.case = T)) |>
    mutate(MAPPED_TRAIT_CATEGORY = "Disease/Disorder",
           disease_terms = "lung carcinoma",
           DISEASE_STUDY = T) 

# head and neck cancer
add_head_neck_cancer <-
  gwas_study_info |> 
    filter(PUBMED_ID %in% not_disease_pubmeds) |> 
    filter(grepl("head and neck cancer", `DISEASE/TRAIT`, ignore.case = T)) |>
    mutate(MAPPED_TRAIT_CATEGORY = "Disease/Disorder",
           disease_terms = "head and neck malignant neoplasia",
           DISEASE_STUDY = T) 

gwas_study_info <- 
rows_append(gwas_study_info,
            rbind(add_missed_breast_cancer,
              add_missed_colorectal_cancer,
              add_small_cell_lung_carcinoma,
              add_lung_cancer,
              add_head_neck_cancer
              )
            )
  

not_disease_pubmeds <- gwas_study_info  |> 
                       group_by(PUBMED_ID) |> 
                       summarise(DISEASE_STUDY = any(DISEASE_STUDY)) |> 
                       filter(DISEASE_STUDY == F) |> 
                       pull(PUBMED_ID) |>
                       unique()

print("Papers mentioning cancer in DISEASE/TRAIT that are still not categorized as disease studies after adding missed cancer studies")
gwas_study_info |> 
    filter(PUBMED_ID %in% not_disease_pubmeds) |> 
    filter(grepl("cancer", 
                 `DISEASE/TRAIT`, 
                 ignore.case = T)) |>
    select(PUBMED_ID, `DISEASE/TRAIT`) |>
    distinct()

gwas_study_info |> 
    filter(PUBMED_ID %in% not_disease_pubmeds) |> 
    filter(grepl("cancer", 
                 STUDY, 
                 ignore.case = T)) |>
    select(PUBMED_ID, STUDY) |>
    distinct() |> 
    head()

print("Number of papers with no disease study after adding missed cancer studies")
length(not_disease_pubmeds)

```

## Diabetes 

```{r diabetes_terms_in_non_disease_studies}

# type 2 diabetes
add_missed_type_2_diabetes <-
gwas_study_info |>
    filter(PUBMED_ID %in% not_disease_pubmeds) |> 
    filter(grepl("type 2 diabetes|\\bT2D\\b", `DISEASE/TRAIT`, ignore.case = T)) |>
    mutate(MAPPED_TRAIT_CATEGORY = "Disease/Disorder",
           disease_terms = "type 2 diabetes mellitus",
           DISEASE_STUDY = T) 

# type 1 diabetes
add_missed_type_1_diabetes <-
gwas_study_info |>
    filter(PUBMED_ID %in% not_disease_pubmeds) |> 
    filter(grepl("type 1 diabetes", `DISEASE/TRAIT`, ignore.case = T)) |>
    mutate(MAPPED_TRAIT_CATEGORY = "Disease/Disorder",
           disease_terms = "type 1 diabetes mellitus",
           DISEASE_STUDY = T) 


gwas_study_info <-
rows_append(gwas_study_info,
            rbind(add_missed_type_2_diabetes,
              add_missed_type_1_diabetes
              )
            )

not_disease_pubmeds <- gwas_study_info  |> 
                       group_by(PUBMED_ID) |> 
                       summarise(DISEASE_STUDY = any(DISEASE_STUDY)) |> 
                       filter(DISEASE_STUDY == F) |> 
                       pull(PUBMED_ID) |>
                       unique()

print("Papers mentioning diabetes in DISEASE/TRAIT that are still not categorized as disease studies after adding missed diabetes studies")
gwas_study_info |> 
    filter(PUBMED_ID %in% not_disease_pubmeds) |> 
    filter(grepl("diabetes", 
                 `DISEASE/TRAIT`, 
                 ignore.case = T)) |>
    select(PUBMED_ID, `DISEASE/TRAIT`) |>
    distinct()

gwas_study_info |> 
    filter(PUBMED_ID %in% not_disease_pubmeds) |> 
    filter(grepl("\\bdiabetes", 
                 STUDY,
                 ignore.case = T)) |>
    select(PUBMED_ID, STUDY) |>
    distinct() |>
    head()

# diabetes add 
add_missed_diabetes <-
gwas_study_info |>
    filter(PUBMED_ID %in% not_disease_pubmeds) |> 
    filter(grepl("diabetes", `DISEASE/TRAIT`, ignore.case = T)) |>
    mutate(MAPPED_TRAIT_CATEGORY = "Disease/Disorder",
           disease_terms = "diabetes mellitus",
           DISEASE_STUDY = T)

gwas_study_info <-
rows_append(gwas_study_info,
            add_missed_diabetes
            )

not_disease_pubmeds <- gwas_study_info  |> 
                       group_by(PUBMED_ID) |> 
                       summarise(DISEASE_STUDY = any(DISEASE_STUDY)) |> 
                       filter(DISEASE_STUDY == F) |> 
                       pull(PUBMED_ID) |>
                       unique()

print("Number of papers with no disease study after adding missed diabetes studies")
length(not_disease_pubmeds)


```

## COPD, and other chronic lung/airway diseases

```{r lung_terms_in_non_disease_studies}

add_missed_copd <-
gwas_study_info |>
    filter(PUBMED_ID %in% not_disease_pubmeds) |> 
    filter(grepl("chronic osbtructive pulmonary disease|COPD|chronic obstructive pulmonary disease", `DISEASE/TRAIT`, ignore.case = T)) |>
    mutate(MAPPED_TRAIT_CATEGORY = "Disease/Disorder",
           disease_terms = "chronic obstructive pulmonary disease",
           DISEASE_STUDY = T)


gwas_study_info |>
    filter(PUBMED_ID %in% not_disease_pubmeds) |> 
    filter(grepl("asthma", `DISEASE/TRAIT`, ignore.case = T))  |>
    select(PUBMED_ID, `DISEASE/TRAIT`) |>
    distinct()


add_missed_asthma <-
gwas_study_info |>
    filter(PUBMED_ID %in% not_disease_pubmeds) |> 
    filter(grepl("asthma", `DISEASE/TRAIT`, ignore.case = T)) |>
    mutate(MAPPED_TRAIT_CATEGORY = "Disease/Disorder",
           disease_terms = "asthma",
           DISEASE_STUDY = T)

# interstitial lung disease
add_missed_ild <-
gwas_study_info |>
    filter(PUBMED_ID %in% not_disease_pubmeds) |> 
    filter(grepl("interstitial lung disease", `DISEASE/TRAIT`, ignore.case = T)) |>
    mutate(MAPPED_TRAIT_CATEGORY = "Disease/Disorder",
           disease_terms = "interstitial lung disease",
           DISEASE_STUDY = T)

gwas_study_info <-
rows_append(gwas_study_info,
            rbind(add_missed_copd,
              add_missed_asthma,
              add_missed_ild
              )
            )
not_disease_pubmeds <- gwas_study_info  |> 
                       group_by(PUBMED_ID) |> 
                       summarise(DISEASE_STUDY = any(DISEASE_STUDY)) |> 
                       filter(DISEASE_STUDY == F) |> 
                       pull(PUBMED_ID) |>
                       unique()

print("Papers mentioning COPD in DISEASE/TRAIT that are still not categorized as disease studies after adding missed COPD studies")
gwas_study_info |> 
    filter(PUBMED_ID %in% not_disease_pubmeds) |> 
    filter(grepl("COPD|chronic obstructive pulmonary disease|lung|asthma", 
                 `DISEASE/TRAIT`, 
                 ignore.case = T)) |>
    select(PUBMED_ID, `DISEASE/TRAIT`) |>
    distinct()

gwas_study_info |>
    filter(PUBMED_ID %in% not_disease_pubmeds) |> 
    filter(grepl("COPD|chronic obstructive pulmonary disease|lung|asthma", 
                 STUDY, 
                 ignore.case = T)) |>
    select(PUBMED_ID, STUDY) |>
    distinct() |>
    head()

print("Number of papers with no disease study after adding missed chronic lung/airway studies")
length(not_disease_pubmeds)


```

## Psychiatric/Neurological 

```{r psychiatric_neurological_terms_in_non_disease_studies}

gwas_study_info |>
    filter(PUBMED_ID %in% not_disease_pubmeds) |> 
    filter(grepl("bipolar disorder", `DISEASE/TRAIT`, ignore.case = T)) |>
    select(PUBMED_ID, `DISEASE/TRAIT`, STUDY) |>
    distinct()

# bipolar disorder
add_missed_bipolar_disorder <-
gwas_study_info |>
    filter(PUBMED_ID %in% not_disease_pubmeds) |> 
    filter(grepl("bipolar disorder", `DISEASE/TRAIT`, ignore.case = T)) |>
    mutate(MAPPED_TRAIT_CATEGORY = "Disease/Disorder",
           disease_terms = "bipolar disorder",
           DISEASE_STUDY = T)


# major depressive disorder
gwas_study_info |>
    filter(PUBMED_ID %in% not_disease_pubmeds) |> 
    filter(grepl("major depressive disorder", `DISEASE/TRAIT`, ignore.case = T)) |>
    select(PUBMED_ID, `DISEASE/TRAIT`, STUDY) |>
    distinct()

add_missed_major_depressive_disorder <-
gwas_study_info |>
    filter(PUBMED_ID %in% not_disease_pubmeds) |> 
    filter(grepl("major depressive disorder", `DISEASE/TRAIT`, ignore.case = T)) |>
    mutate(MAPPED_TRAIT_CATEGORY = "Disease/Disorder",
           disease_terms = "major depressive disorder",
           DISEASE_STUDY = T)


# ADHD
gwas_study_info |>
    filter(PUBMED_ID %in% not_disease_pubmeds) |> 
    filter(grepl("attention deficit hyperactivity disorder|ADHD|attention-deficit/hyperactivity disorder", 
                 `DISEASE/TRAIT`, 
                 ignore.case = T)) |>
    select(PUBMED_ID, `DISEASE/TRAIT`, STUDY) |>
    distinct()

add_missed_adhd <-
gwas_study_info |>
    filter(PUBMED_ID %in% not_disease_pubmeds) |> 
    filter(grepl("attention deficit hyperactivity disorder|ADHD|attention-deficit/hyperactivity disorder", 
                 `DISEASE/TRAIT`, 
                 ignore.case = T)) |>
    mutate(MAPPED_TRAIT_CATEGORY = "Disease/Disorder",
           disease_terms = "attention deficit hyperactivity disorder",
           DISEASE_STUDY = T)

# schizophrenia
gwas_study_info |>
    filter(PUBMED_ID %in% not_disease_pubmeds) |> 
    filter(grepl("schizophrenia", `DISEASE/TRAIT`, ignore.case = T)) |>
    select(PUBMED_ID, `DISEASE/TRAIT`, STUDY) |>
    distinct()

add_missed_schizophrenia <-
gwas_study_info |>
    filter(PUBMED_ID %in% not_disease_pubmeds) |> 
    filter(grepl("schizophrenia", `DISEASE/TRAIT`, ignore.case = T)) |>
    mutate(MAPPED_TRAIT_CATEGORY = "Disease/Disorder",
           disease_terms = "schizophrenia",
           DISEASE_STUDY = T)

# Alzheimer's disease
gwas_study_info |>
    filter(PUBMED_ID %in% not_disease_pubmeds) |> 
    filter(grepl("Alzheimer", `DISEASE/TRAIT`, ignore.case = T)) |>
    select(PUBMED_ID, `DISEASE/TRAIT`, STUDY) |>
    distinct()

add_missed_alzheimers <-
gwas_study_info |>
    filter(PUBMED_ID %in% not_disease_pubmeds) |> 
    filter(grepl("Alzheimer", `DISEASE/TRAIT`, ignore.case = T)) |>
    mutate(MAPPED_TRAIT_CATEGORY = "Disease/Disorder",
           disease_terms = "Alzheimer's disease",
           DISEASE_STUDY = T)


# dementia
gwas_study_info |>
    filter(PUBMED_ID %in% not_disease_pubmeds) |> 
    filter(grepl("dementia", 
                 `DISEASE/TRAIT`, 
                 ignore.case = T)) |>
    select(PUBMED_ID, `DISEASE/TRAIT`) |>
    distinct()

gwas_study_info <-
rows_append(gwas_study_info,
            rbind(add_missed_bipolar_disorder,
              add_missed_major_depressive_disorder,
              add_missed_adhd,
              add_missed_schizophrenia,
              add_missed_alzheimers
              )
            )

not_disease_pubmeds <- gwas_study_info  |> 
                       group_by(PUBMED_ID) |> 
                       summarise(DISEASE_STUDY = any(DISEASE_STUDY)) |> 
                       filter(DISEASE_STUDY == F) |> 
                       pull(PUBMED_ID) |>
                       unique()

print("Papers mentioning psychiatric or neurological diseases in DISEASE/TRAIT that are still not categorized as disease studies after adding missed psychiatric/neurological disease studies")
gwas_study_info |> 
    filter(PUBMED_ID %in% not_disease_pubmeds) |> 
    filter(grepl("bipolar disorder|major depressive disorder|attention deficit hyperactivity disorder|ADHD|schizophrenia|Alzheimer|dementia", 
                 `DISEASE/TRAIT`, 
                 ignore.case = T)) |>
    select(PUBMED_ID, `DISEASE/TRAIT`) |>
    distinct()

gwas_study_info |>
    filter(PUBMED_ID %in% not_disease_pubmeds) |> 
    filter(grepl("bipolar disorder|major depressive disorder|attention deficit hyperactivity disorder|ADHD|schizophrenia|Alzheimer|dementia", 
                 STUDY, 
                 ignore.case = T)) |>
    select(PUBMED_ID, STUDY) |>
    distinct()

print("Number of papers with no disease study after adding missed psychiatric/neurological disease studies")
length(not_disease_pubmeds)


```

## Pubmed ID 35760791

```{r hereditary hemorrhagic telangiectasia}

gwas_study_info |> 
  filter(PUBMED_ID =="35760791") |>
  select(PUBMED_ID, `DISEASE/TRAIT`) |>
    distinct()

add_missed_hereditary_hemorrhagic_telangiectasia <-
gwas_study_info |>
    filter(PUBMED_ID %in% not_disease_pubmeds) |> 
    filter(grepl("hereditary hemorrhagic telangiectasia", 
                 `DISEASE/TRAIT`, 
                 ignore.case = T)) |>
    mutate(MAPPED_TRAIT_CATEGORY = "Disease/Disorder",
           disease_terms = "hereditary hemorrhagic telangiectasia",
           DISEASE_STUDY = T)


add_missed_marfan_syndrome <-
gwas_study_info |>
    filter(PUBMED_ID %in% not_disease_pubmeds) |> 
    filter(grepl("marfan syndrome", `DISEASE/TRAIT`, ignore.case = T)) |>
    mutate(MAPPED_TRAIT_CATEGORY = "Disease/Disorder",
           disease_terms = "marfan syndrome",
           DISEASE_STUDY = T)

add_missed_alport_syndrome <-
gwas_study_info |>
    filter(PUBMED_ID %in% not_disease_pubmeds) |> 
    filter(grepl("alport syndrome", `DISEASE/TRAIT`, ignore.case = T)) |>
    mutate(MAPPED_TRAIT_CATEGORY = "Disease/Disorder",
           disease_terms = "alport syndrome",
           DISEASE_STUDY = T)

# autosomal dominant polycystic kidney disease
add_missed_adpkd <-
gwas_study_info |>
    filter(PUBMED_ID %in% not_disease_pubmeds) |> 
    filter(grepl("autosomal dominant polycystic kidney disease|ADPKD", `DISEASE/TRAIT`, ignore.case = T)) |>
    mutate(MAPPED_TRAIT_CATEGORY = "Disease/Disorder",
           disease_terms = "autosomal dominant polycystic kidney disease",
           DISEASE_STUDY = T)

# alpha-1-antitrypsin deficiency
add_missed_a1at_deficiency <-
gwas_study_info |>
    filter(PUBMED_ID %in% not_disease_pubmeds) |> 
    filter(grepl("alpha-1-antitrypsin deficiency", `DISEASE/TRAIT`, ignore.case = T)) |>
    mutate(MAPPED_TRAIT_CATEGORY = "Disease/Disorder",
           disease_terms = "alpha-1-antitrypsin deficiency",
           DISEASE_STUDY = T)


gwas_study_info <-
rows_append(gwas_study_info,
            rbind(
              add_missed_hereditary_hemorrhagic_telangiectasia,
              add_missed_marfan_syndrome,
              add_missed_alport_syndrome,
              add_missed_adpkd,
              add_missed_a1at_deficiency
              )
            )

not_disease_pubmeds <- gwas_study_info  |> 
                       group_by(PUBMED_ID) |> 
                       summarise(DISEASE_STUDY = any(DISEASE_STUDY)) |> 
                       filter(DISEASE_STUDY == F) |> 
                       pull(PUBMED_ID) |>
                       unique()

print("Number of papers with no disease study after adding missed hereditary hemorrhagic telangiectasia study")
length(not_disease_pubmeds)



```

## Infectious disease

```{r infectious_disease_terms_in_non_disease_studies}
gwas_study_info |>
    filter(PUBMED_ID %in% not_disease_pubmeds) |> 
    filter(grepl("HIV|human immunodeficiency virus|AIDS|acquired immunodeficiency syndrome", 
                 `DISEASE/TRAIT`, 
                 ignore.case = T)) |>
    select(PUBMED_ID, `DISEASE/TRAIT`) |>
    distinct()

add_hiv_1_infection <-
gwas_study_info |>
    filter(PUBMED_ID %in% not_disease_pubmeds) |> 
    filter(grepl("HIV-1|human immunodeficiency virus 1", 
                 `DISEASE/TRAIT`, 
                 ignore.case = T)) |>
    mutate(MAPPED_TRAIT_CATEGORY = "Disease/Disorder",
           disease_terms = "hiv-1 infection",
           DISEASE_STUDY = T)

# hepatitis c virus infection
add_hep_c_virus_infection <-
gwas_study_info |>
    filter(PUBMED_ID %in% not_disease_pubmeds) |> 
    filter(grepl("hepatitis c virus infection", 
                 `DISEASE/TRAIT`, 
                 ignore.case = T)) |>
    mutate(MAPPED_TRAIT_CATEGORY = "Disease/Disorder",
           disease_terms = "hepatitis c virus infection",
           DISEASE_STUDY = T)

gwas_study_info <-
rows_append(gwas_study_info,
            rbind(add_hiv_1_infection,
              add_hep_c_virus_infection
              )
            )

not_disease_pubmeds <- gwas_study_info  |> 
                       group_by(PUBMED_ID) |> 
                       summarise(DISEASE_STUDY = any(DISEASE_STUDY)) |> 
                       filter(DISEASE_STUDY == F) |> 
                       pull(PUBMED_ID) |>
                       unique()

print("Papers mentioning infectious diseases in DISEASE/TRAIT that are still not categorized as disease studies after adding missed infectious disease studies")
gwas_study_info |> 
    filter(PUBMED_ID %in% not_disease_pubmeds) |> 
    filter(grepl("HIV|human immunodeficiency virus|AIDS|acquired immunodeficiency syndrome|hepatitis c virus infection", 
                 `DISEASE/TRAIT`, 
                 ignore.case = T)) |>
    select(PUBMED_ID, `DISEASE/TRAIT`) |>
    distinct()

gwas_study_info |>
    filter(PUBMED_ID %in% not_disease_pubmeds) |> 
    filter(grepl("HIV|human immunodeficiency virus|AIDS|acquired immunodeficiency syndrome|hepatitis c virus infection", 
                 STUDY, 
                 ignore.case = T)) |>
    select(PUBMED_ID, STUDY) |>
    distinct() |> 
    head()

print("Number of papers with no disease study after adding missed infectious disease studies")
length(not_disease_pubmeds)

```

## Autoimmune/Inflammatory: 

```{r autoimmune_inflammatory_terms_in_non_disease_studies}

gwas_study_info |>
    filter(PUBMED_ID %in% not_disease_pubmeds) |> 
    filter(grepl("inflammatory bowel disease|IBD", 
                 `DISEASE/TRAIT`, 
                 ignore.case = T)) |>
    select(PUBMED_ID, `DISEASE/TRAIT`) |>
    distinct()

# inflammatory bowel disease
add_missed_ibd <-
gwas_study_info |>
    filter(PUBMED_ID %in% not_disease_pubmeds) |> 
    filter(grepl("inflammatory bowel disease|IBD", `DISEASE/TRAIT`, ignore.case = T)) |>
    mutate(MAPPED_TRAIT_CATEGORY = "Disease/Disorder",
           disease_terms = "inflammatory bowel disease",
           DISEASE_STUDY = T)

# rheumatoid arthritis
gwas_study_info |>
    filter(PUBMED_ID %in% not_disease_pubmeds) |> 
    filter(grepl("in rheumatoid arthritis\\b", `DISEASE/TRAIT`, ignore.case = T)) |>
    select(PUBMED_ID, `DISEASE/TRAIT`) |>
    distinct()

add_missed_ra <-
gwas_study_info |>
    filter(PUBMED_ID %in% not_disease_pubmeds) |> 
    filter(grepl("in rheumatoid arthritis\\b", `DISEASE/TRAIT`, ignore.case = T)) |>
    mutate(MAPPED_TRAIT_CATEGORY = "Disease/Disorder",
           disease_terms = "rheumatoid arthritis",
           DISEASE_STUDY = T)

# juvenile idiopathic arthritis
gwas_study_info |>
    filter(PUBMED_ID %in% not_disease_pubmeds) |> 
    filter(grepl("juvenile idiopathic arthritis", 
                 `DISEASE/TRAIT`, 
                 ignore.case = T)) |>
    select(PUBMED_ID, `DISEASE/TRAIT`) |>
    distinct()

add_missed_jia <-
gwas_study_info |>
    filter(PUBMED_ID %in% not_disease_pubmeds) |> 
    filter(grepl("juvenile idiopathic arthritis", `DISEASE/TRAIT`, ignore.case = T)) |>
    mutate(MAPPED_TRAIT_CATEGORY = "Disease/Disorder",
           disease_terms = "juvenile idiopathic arthritis",
           DISEASE_STUDY = T)

gwas_study_info <-
rows_append(gwas_study_info,
            rbind(add_missed_ibd,
              add_missed_ra,
              add_missed_jia
              )
            )

not_disease_pubmeds <- gwas_study_info  |> 
                       group_by(PUBMED_ID) |> 
                       summarise(DISEASE_STUDY = any(DISEASE_STUDY)) |> 
                       filter(DISEASE_STUDY == F) |> 
                       pull(PUBMED_ID) |>
                       unique()

print("Papers mentioning autoimmune/inflammatory diseases in DISEASE/TRAIT that are still not categorized as disease studies after adding missed autoimmune/inflammatory disease studies")
gwas_study_info |> 
    filter(PUBMED_ID %in% not_disease_pubmeds) |> 
    filter(grepl("inflammatory bowel disease|IBD|rheumatoid arthritis|juvenile idiopathic arthritis", 
                 `DISEASE/TRAIT`, 
                 ignore.case = T)) |>
    select(PUBMED_ID, `DISEASE/TRAIT`) |>
    distinct()

gwas_study_info |>
    filter(PUBMED_ID %in% not_disease_pubmeds) |> 
    filter(grepl("inflammatory bowel disease|IBD|rheumatoid arthritis|juvenile idiopathic arthritis", 
                 STUDY, 
                 ignore.case = T)) |>
    select(PUBMED_ID, STUDY) |>
    distinct()

print("Number of papers with no disease study after adding missed autoimmune/inflammatory disease studies")
length(not_disease_pubmeds)

```

## Cardiovascular disease

```{r cardiovascular_terms_in_non_disease_studies}

gwas_study_info |>
    filter(PUBMED_ID %in% not_disease_pubmeds) |> 
    filter(grepl("cardiovascular disease|coronary artery disease|myocardial infarction|heart attack|heart disease",
                 `DISEASE/TRAIT`, 
                 ignore.case = T)
           ) |>
    select(PUBMED_ID, `DISEASE/TRAIT`) |>
  distinct()

add_missed_coronary_artery_disease <-
gwas_study_info |>
    filter(PUBMED_ID %in% not_disease_pubmeds) |> 
    filter(grepl("stable coronary artery disease", `DISEASE/TRAIT`, ignore.case = T)) |>
    mutate(MAPPED_TRAIT_CATEGORY = "Disease/Disorder",
           disease_terms = "coronary artery disease",
           DISEASE_STUDY = T)

gwas_study_info <-
rows_append(gwas_study_info,
            add_missed_coronary_artery_disease
            )

not_disease_pubmeds <- gwas_study_info  |> 
                       group_by(PUBMED_ID) |> 
                       summarise(DISEASE_STUDY = any(DISEASE_STUDY)) |> 
                       filter(DISEASE_STUDY == F) |> 
                       pull(PUBMED_ID) |>
                       unique()

print("Papers mentioning cardiovascular diseases in DISEASE/TRAIT that are still not categorized as disease studies after adding missed cardiovascular disease studies")
gwas_study_info |> 
    filter(PUBMED_ID %in% not_disease_pubmeds) |> 
    filter(grepl("cardiovascular disease|coronary artery disease|myocardial infarction|heart attack|heart disease", 
                 `DISEASE/TRAIT`, 
                 ignore.case = T)) |>
    select(PUBMED_ID, `DISEASE/TRAIT`) |>
    distinct()

gwas_study_info |>
    filter(PUBMED_ID %in% not_disease_pubmeds) |> 
    filter(grepl("cardiovascular disease|coronary artery disease|myocardial infarction|heart attack|heart disease", 
                 STUDY, 
                 ignore.case = T)) |>
    select(PUBMED_ID, STUDY) |>
    distinct()

print("Number of papers with no disease study after adding missed cardiovascular disease studies")
length(not_disease_pubmeds)

```


# Creating disease labels column of just disease or phenotype abnormality terms for each study - so that we can see what diseases are being studied

## Make disease label column - combining disease terms from both mapped trait and background trait

```{r combined_disease_terms}

combined_disease_terms = function(MAPPED_TRAIT_1, MAPPED_TRAIT_2){
  
  
  #MAPPED_TRAIT_1 = stringr::str_split(MAPPED_TRAIT_1, ", ") |> unlist()
  #MAPPED_TRAIT_2  = stringr::str_split(MAPPED_TRAIT_2, ", ") |> unlist()
  
  all_mapped_disease_terms = 
    c(MAPPED_TRAIT_1, MAPPED_TRAIT_2) |>
    unique()
  
  combined_mapped_disease_terms = str_flatten(all_mapped_disease_terms, 
                                         collapse = "; ",
                                         na.rm = T
                                         )
  
  return(combined_mapped_disease_terms)
  
}


gwas_study_info <- 
  gwas_study_info |>
  tidyr::unite(col = "all_disease_terms",
               c("disease_terms", "background_disease_terms"),
               remove = F,
               sep = "; ", 
               na.rm = TRUE)
  #dplyr::rowwise() |>
  # dplyr::mutate(all_disease_terms = 
  #               case_when(is.na(background_disease_terms) & is.na(disease_terms) ~ NA,
  #                         is.na(background_disease_terms) & !is.na(disease_terms) ~ disease_terms,
  #                         !is.na(background_disease_terms) & is.na(disease_terms) ~ background_disease_terms,
  #                         !is.na(background_disease_terms) & !is.na(disease_terms) ~
  #                           combined_disease_terms(disease_terms,
  #                                                  background_disease_terms,
  #                                                  )) 
  # 
  # ) |>
  # dplyr::ungroup()

# correct commas
gwas_study_info =
  gwas_study_info |>
    mutate(
    all_disease_terms =
      str_replace(
        all_disease_terms,
        "^osteoarthritis,\\s*(.+)$",
        "osteoarthritis of \\1"
      )
  )

gwas_study_info =
  gwas_study_info |>
  mutate(all_disease_terms = 
           str_replace_all(
             pattern = "hodgkins lymphoma, mixed cellularity",
             replacement = "hodgkins lymphoma",
             string = all_disease_terms)
  )

gwas_study_info =
  gwas_study_info |>
  mutate(all_disease_terms = 
           str_replace_all(
             pattern = "hypertension, pregnancy-induced",
             replacement = "pregnancy-induced hypertension",
             string = all_disease_terms)
  ) |>
    mutate(all_disease_terms = 
           str_replace_all(
             pattern = "renal agenesis, unilateral",
             replacement = "unilateral renal agenesis",
             string = all_disease_terms)
  ) |>
       mutate(all_disease_terms = 
           str_replace_all(
             pattern = "diarrhea, infantile",
             replacement = "infantile diarrhea",
             string = all_disease_terms)
  ) |>
         mutate(all_disease_terms = 
           str_replace_all(
             pattern = "fractures, ununited",
             replacement = "ununited fractures",
             string = all_disease_terms)
         ) |>
           mutate(all_disease_terms = 
           str_replace_all(
             pattern = "cholecystitis, acute",
             replacement = "acute cholecystitis",
             string = all_disease_terms)
         )  |>
           mutate(all_disease_terms = 
           str_replace_all(
             pattern = "hepatitis, alcoholic",
             replacement = "alcoholic hepatitis",
             string = all_disease_terms)
         ) 
  
gwas_study_info =
  gwas_study_info |>
  mutate(all_disease_terms = 
           str_replace_all(
             pattern = "encephalopathy, acute, infection-induced",
             replacement = "infectious encephalitis",
             string = all_disease_terms
           )
         )

gwas_study_info =
  gwas_study_info |>
  mutate(all_disease_terms = 
           str_replace_all(
             pattern = "genital neoplasm, female",
             replacement = "female genital neoplasm",
             string = all_disease_terms
           )
         )

gwas_study_info =
  gwas_study_info |>
  mutate(all_disease_terms = 
           str_replace_all(
             pattern = "psoriasis 14, pustular",
             replacement = "pustular psoriasis 14",
             string = all_disease_terms
           )
         )

gwas_study_info =
  gwas_study_info |>
  mutate(all_disease_terms = 
           str_replace_all(
             pattern = "hand, foot and mouth disease",
             replacement = "hand foot and mouth disease",
             string = all_disease_terms
           )
         )

gwas_study_info =
  gwas_study_info |>
  mutate(all_disease_terms = 
           str_replace_all(
             pattern = "anemia, hemolytic, autoimmune",
             replacement = "autoimmune hemolytic anemia",
             string = all_disease_terms
           )
         )

gwas_study_info =
  gwas_study_info |>
  mutate(all_disease_terms = 
           str_replace_all(
             pattern = "polyarticular juvenile idiopathic arthritis, rheumatoid factor negative",
             replacement = "rheumatoid factor-negative polyarticular juvenile idiopathic arthritis",
             string = all_disease_terms
           )
         )

  
gwas_study_info =
  gwas_study_info |>
  mutate(all_disease_terms = 
           str_replace_all(
             pattern = "neural tube defects, susceptibility to",
             replacement = "neural tube defects",
             string = all_disease_terms
           )) 

gwas_study_info  = 
gwas_study_info |> 
    mutate(all_disease_terms = 
           str_replace_all(
             pattern = "migraine without aura, susceptibility to, 4",
             replacement = "migraine without aura",
             string = all_disease_terms
           )) 

gwas_study_info  = 
gwas_study_info |> 
    mutate(all_disease_terms = 
           str_replace_all(
             pattern = "genital neoplasm, female",
             replacement = "female genital neoplasm",
             string = all_disease_terms
           )) 

print("Number of studies with bad commas")
gwas_study_info |> filter(grepl(",", all_disease_terms))

gwas_study_info =
  gwas_study_info |>
  mutate(all_disease_terms = str_replace_all(
    pattern = "; ",
    replacement = ", ",
    string = all_disease_terms
  ))
  
```

## Minor fixes of trait categorisation and returning traits 

```{r}

# What studies are disease studies but have no collected disease terms?
gwas_study_info |> 
  filter(DISEASE_STUDY == T) |> 
  filter(all_disease_terms == "")  |> 
  select(PUBMED_ID) |> 
  distinct() |>
  nrow()

gwas_study_info |> 
  filter(DISEASE_STUDY == T) |> 
  filter(all_disease_terms == "")  |> 
  select(MAPPED_TRAIT, MAPPED_TRAIT_CATEGORY) |> 
  distinct() |>
  head()

```

### Fix bug where MAPPED_TRAIT/BACKGROUND_MAPPED_TRAIT is empty string but TRAIT_CATEGORY is listed as disease/phenotypic abnormality

```{r fix_empty_mapped_trait_but_disease_category}

gwas_study_info = gwas_study_info |>
  #rowwise() |>
  mutate(MAPPED_TRAIT_CATEGORY = ifelse(MAPPED_TRAIT == "",
                                        "Other",
                                        MAPPED_TRAIT_CATEGORY)) |>
  mutate(BACKGROUND_TRAIT_CATEGORY = ifelse(MAPPED_BACKGROUND_TRAIT == "",
                                        "Other",
                                        BACKGROUND_TRAIT_CATEGORY)) 

```


### Recalculate disease study flag (is disease study or not?)

now that I have corrected any mistakes in categorization and added some missing disease terms

```{r recalculate_disease_study_flag}

gwas_study_info = 
gwas_study_info |>
  #dplyr::rowwise() |>
  dplyr::mutate(DISEASE_STUDY = 
                   ifelse(MAPPED_TRAIT_CATEGORY == "Disease/Disorder" | 
                          MAPPED_TRAIT_CATEGORY == "Phenotypic Abnormality" |  
                          BACKGROUND_TRAIT_CATEGORY == "Disease/Disorder" | 
                          BACKGROUND_TRAIT_CATEGORY == "Phenotypic Abnormality",
                          T, F )
                ) |>
  dplyr::ungroup() 

gwas_study_info |> 
  filter(DISEASE_STUDY == T) |>
    group_by(MAPPED_TRAIT_CATEGORY, 
           BACKGROUND_TRAIT_CATEGORY) |>
    summarise(n = n()) |>
  arrange(desc(n))


# gwas_study_info |> 
#   filter(DISEASE_STUDY == T) |> 
#   filter(all_disease_terms == "" | is.na(all_disease_terms))  |> 
#   select(MAPPED_TRAIT, MAPPED_TRAIT_CATEGORY) |> 
#   distinct() |>
#   nrow()

print("Number of studies of each kind of trait")
gwas_study_info |> 
  group_by(DISEASE_STUDY, 
           MAPPED_TRAIT_CATEGORY, 
           BACKGROUND_TRAIT_CATEGORY) |> 
  summarise(n = n()) |>
  arrange(desc(n))

print("Number of papers with at least one disease study")
gwas_study_info |> 
  group_by(PUBMED_ID) |> 
  summarise(DISEASE_STUDY = any(DISEASE_STUDY == T)) |> 
  group_by(DISEASE_STUDY) |> 
  summarise(n = n())

```

# Saving:

```{r}

data.table::fwrite(gwas_study_info,
                  here::here("output/gwas_cat/gwas_study_info_trait_cat.csv"), 
                  sep = ",")

```
